### 2016-12-06 articles/cycle/alps-cycle.tex
ALPCYC: init Alps glacial cycle paper.
Add title and authors, and Makefile..

\documentclass{article}

\usepackage{authblk}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[pdftex]{xcolor}

\definecolor{darkblue}{cmyk}{0.9,0.3,0.0,0.0}
\definecolor{darkgreen}{cmyk}{0.8,0.0,1.0,0.0}
\definecolor{darkred}{cmyk}{0.1,0.9,0.8,0.0}

\newcommand{\idea}[1]{\textcolor{darkgreen}{\emph{[\textbf{IDEA:} #1]}}}
\newcommand{\note}[1]{\textcolor{darkblue}{\emph{[\textbf{NOTE:} #1]}}}
\newcommand{\todo}[1]{\textcolor{darkred}{\emph{[\textbf{TODO:} #1]}}}
\newcommand{\aref}[0]{\textcolor{darkblue}{\textbf{[REF.]}}}

\title{Modelling the last glacial cycle in the Alps}

\author[1]{Julien Seguinot%
           \thanks{Correspondence to seguinot@vaw.baug.ethz.ch}}
\author[1]{Guillaume Jouvet}
\author[1]{Matthias Huss}
\author[1]{Martin Funk}
\author[2]{Frank Preusser}

\affil[1]{Laboratory of Hydraulics, Hydrology and Glaciology,
          ETH Zürich, Switzerland}
\affil[2]{Institute of Earth and Environmental Sciences,
          University of Freiburg, Germany}

% ======================================================================
\begin{document}
% ======================================================================

\maketitle

\begin{abstract}
\end{abstract}

% ======================================================================
\end{document}
% ======================================================================

### 2016-12-06 articles/cycle/alps-cycle.tex
ALPCYC: add SGM 2016 abstract.
For once it corresponds exactly to what I have presented.

@@
 \begin{abstract}
+
+    The European Alps, cradle of pioneer glacial studies, are one of the the
+    regions where geological markers of past glaciations are most abundant and
+    well- studied. Such conditions make the region ideal for testing numerical
+    glacier models based on approximated ice flow physics against field-based
+    reconstructions, and vice-versa.
+
+    Here, we use the Parallel Ice Sheet Model (PISM) to model the entire last
+    glacial cycle (120–0 ka) in the Alps, with a horizontal resolution of 1 km.
+    Climate forcing is derived using present-day climate data from WorldClim,
+    the ERA- Interim reanalysis, and time-dependent temperature offsets from
+    the EPICA ice core record in Antarctica, which allows reproduce the two
+    major glaciation events documented in the geological record during marine
+    oxygen isotope stages 4 (69–62 ka) and 2 (34–18 ka).
+
+    Despite the low variability of this Antarctic-based climate forcing, our
+    simulation depict a highly dynamic ice cap, showing that alpine glaciers
+    may have advanced many times over the foreland during MIS 5 and 3.
+    Cumulative basal sliding, a proxy for glacial erosion, is modelled to be
+    highest in the deep valleys of the western Alps. However, relict uplift
+    rates at the end of the simulation are most important in the eastern Alps
+    where a thick ice cover persists for most of the modelled glacial cycle.
+    Finally, the Last Glacial Maximum advance, often considered synchronous, is
+    here modelled as a time-transgressive event, with some glacier lobes
+    reaching their maximum as early as 27 ka, and some as late as 21 ka, which
+    may explain some of the variability in published cosmogenic dates.
+
 \end{abstract}

### 2017-01-17 figures/alpcyc_lr_footprints.py
Add LGM outline on lo-res footprints plot.

@@ -57,6 +57,7 @@ for i, rec in enumerate(records):
         ut.pl.add_corner_tag('MIS %d' % (2+2*j), ax=ax, va='bottom')
         ut.pl.add_subfig_label('(%s)' % list('abcdef')[i+3*j], ax=ax)
         ut.pl.draw_natural_earth(ax)
+        ut.pl.draw_lgm_outline(ax, c='k')
 
 # save
 fig.savefig('alpcyc_lr_footprints')

### 2017-01-17 figures/util/pl.py

@@ -249,12 +249,12 @@ def draw_natural_earth(ax=None):
     ax.add_feature(graticules)
 
 
-def draw_lgm_outline(ax=None):
+def draw_lgm_outline(ax=None, c='#e31a1c'):
     """Add Ehlers et al. hole-filled LGM outline."""
     ax = ax or iplt.gca()
     shp = cshp.Reader('../data/native/lgm_alpen_holefilled.shp')
     ax.add_geometries(shp.geometries(), ll, lw=0.5, alpha=0.75,
-                      edgecolor='#e31a1c', facecolor='none', zorder=0)
+                      edgecolor=c, facecolor='none', zorder=0)
     del shp
 
 

### 2017-01-17 articles/cycle/alps-cycle.tex
ALPCYC: add tentative outline.
This could work for a traditional paper but not for Nature's short format.
I have to check their instructions once again.
Update abstract and title to EGU 2017 version.

--- 2016-12-06 articles/cycle/alps-cycle.tex
@@
 
-\title{Modelling the last glacial cycle in the Alps}
+\title{Modelling last glacial cycle ice dynamics in the Alps}
 
@@
 \begin{abstract}
+
+    The European Alps, cradle of pioneer glacial studies, are one of the
+    regions where geological markers of past glaciations are most abundant and
+    well-studied. Such conditions make the region ideal for testing numerical
+    glacier models based on approximated ice flow physics against field-based
+    reconstructions, and vice-versa.
+
+    Here, we use the Parallel Ice Sheet Model (PISM) to model the entire last
+    glacial cycle (120--0\,ka) in the Alps, with a horizontal resolution of
+    1\,km. Climate forcing is derived using present-day climate data from
+    WorldClim and the ERA-Interim reanalysis, and time-dependent temperature
+    offsets from multiple paleo-climate proxies, among which only the EPICA ice
+    core record yields glacial extent during marine oxygen isotope stages~4
+    (69--62\,ka) and~2 (34--18\,ka) in agreement to geological reconstructions.
+
+    Despite the low variability of this Antarctic-based climate forcing, our
+    simulation depicts a highly dynamic ice cap, showing that alpine glaciers
+    may have advanced many times over the foreland during the last glacial
+    cycle. Cumulative basal sliding, a proxy for glacial erosion, is modelled
+    to be highest in the deep valleys of the western Alps. Finally, the Last
+    Glacial Maximum advance, often considered synchronous, is here modelled as
+    a time-transgressive event, with some glacier lobes reaching their maximum
+    as early as 27\,ka, and some as late as 21\,ka. Modelled ice thickness is
+    about 900\,m higher than observed trimline elevations, yet our simulation
+    predicts little erosion at high elevation due to cold ice conditions.
+
 \end{abstract}
+
+\section{Introduction}
+
+\section{Model set-up}
+\subsection{Overview}
+\subsection{Ice rheology}
+\subsection{Basal sliding}
+\subsection{Bedrock deformation}
+\subsection{Surface mass balance}
+\subsection{Climate forcing}
+    Fig. -- GRIP and EPICA time series
+\subsection{Computing cost}
+    Fig. -- Scalability
+    Fig. -- Time stamps
+
+\section{Results}
+\subsection{Ice cap evolution}
+    Fig. -- 21 ka snapshot and time series
+\subsection{Ice thickness}
+    Fig. -- Trimlines
+\subsection{Timing of the LGM}
+    Fig. -- Timing of the LGM
+\subsection{Total sliding}
+    Fig. -- Total sliding
+\subsection{Uplift rates}
+    Fig. -- Uplift rates
+
+\section{Conclusion}
 

### 2017-01-18 articles/cycle/alps-cycle.tex
Update outline to Nature format and new abstract.
Add selected figures into manuscript.
Nature format is limited to four figures. I have many more... Can't quite
settle on which to include.

* 21 ka snapshot -> for sure
* Timing of the LGM -> for sure
* Trimlines -> harder to understand, easier to critic
* Erosion -> slightly off-topic, but interesting to many

Maybe another figure clearly showing the many glacier advances, like the Rhine
profile or an improved number of glaciations map, is still needed. I guess I
should work and see if I can improve these two.

@@
 \usepackage[pdftex]{xcolor}
+\usepackage[pdftex]{graphicx}
+\graphicspath{{../../figures/}}
 
@@
 
-\section{Introduction}
+\section{Main}
+\subsection{Introduction}
+\subsection{Ice cap evolution}
+    Fig.~\ref{fig:lgmvel} -- Snapshot at 21\,ka and volume time series.
+\subsection{Timing of the LGM}
+    Fig.~\ref{fig:timing} -- Timing of the LGM and area time series.
+\subsection{Ice thickness}
+    Fig.~\ref{fig:trimlines} -- Ice thickness comparison to trimlines.
+\subsection{Erosion potential}
+    Fig.~\ref{fig:erosion} -- Integrated erosion potential.
 
-\section{Model set-up}
+\section{Methods}
 \subsection{Overview}
+    Tab. -- Model parameters
 \subsection{Ice rheology}
@@
 \subsection{Climate forcing}
-    Fig. -- GRIP and EPICA time series
-\subsection{Computing cost}
-    Fig. -- Scalability
-    Fig. -- Time stamps
-
-\section{Results}
-\subsection{Ice cap evolution}
-    Fig. -- 21 ka snapshot and time series
-\subsection{Ice thickness}
-    Fig. -- Trimlines
-\subsection{Timing of the LGM}
-    Fig. -- Timing of the LGM
-\subsection{Total sliding}
-    Fig. -- Total sliding
-\subsection{Uplift rates}
-    Fig. -- Uplift rates
+    Fig.~\ref{fig:inputs} -- Climate and geothermal model inputs.\\
+    Tab. -- Records\\
+    Fig.~\ref{fig:timeseries} -- Low-resolution time series.\\
+    Fig.~\ref{fig:footprints} -- Low-resolution ice cover.
 
 \section{Conclusion}
+
+% ----------------------------------------------------------------------
+% References
+% ----------------------------------------------------------------------
+
+\bibliographystyle{abbrvnat}
+\bibliography{../../references/references}
+
+% ----------------------------------------------------------------------
+% Figures
+% ----------------------------------------------------------------------
+
+    \begin{figure}
+      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
+      \caption{Model output at 21\,ka.}
+      \label{fig:lgmvel}
+    \end{figure}
+
+    \begin{figure}
+      \centerline{\includegraphics{alpcyc_hr_timing}}
+      \caption{Timing of the LGM and area time series.}
+      \label{fig:timing}
+    \end{figure}
+
+    \begin{figure}
+      \centerline{\includegraphics{alpcyc_hr_trimlines}}
+      \caption{Ice thickness comparison to trimlines.}
+      \label{fig:trimlines}
+    \end{figure}
+
+    \begin{figure}
+      \centerline{\includegraphics{alpcyc_hr_erosion}}
+      \caption{Integrated erosion potential.}
+      \label{fig:erosion}
+    \end{figure}
+
+% ----------------------------------------------------------------------
+% Supplementary tables
+\setcounter{table}{0}
+\renewcommand{\thetable}{S\arabic{table}}%
+% ----------------------------------------------------------------------
+
+% ----------------------------------------------------------------------
+% Supplementary figures
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}%
+% ----------------------------------------------------------------------
+
+    \begin{figure}
+      \centerline{\includegraphics{alpcyc_hr_inputs}}
+      \caption{Climate and geothermal model inputs.}
+      \label{fig:inputs}
+    \end{figure}
+
+    \begin{figure}
+      \centerline{\includegraphics{alpcyc_lr_timeseries}}
+      \caption{Low-resolution time series.}
+      \label{fig:timeseries}
+    \end{figure}
+
+    \begin{figure}
+      \centerline{\includegraphics{alpcyc_lr_footprints}}
+      \caption{Low-resolution ice cover.}
+      \label{fig:footprints}
+    \end{figure}
 

### 2017-01-23 figures/alpcyc_hr_profiles.py
Rename figure 'profile' to 'profiles'.
There will be 1 -- 3 more profiles to come.

@@ -1,4 +1,5 @@
 #!/usr/bin/env python2
+
 # coding: utf-8
 
 import util as ut
@@ -74,4 +75,4 @@ ax.yaxis.tick_right()
 ax.grid(axis='y')
 
 # save
-fig.savefig('alpcyc_hr_profile')
+fig.savefig('alpcyc_hr_profiles')

### 2017-01-23 articles/cycle/alps-cycle.tex
Rework outline and fig. list after disc. Matthias.

@@
 \subsection{Introduction}
-\subsection{Ice cap evolution}
-    Fig.~\ref{fig:lgmvel} -- Snapshot at 21\,ka and volume time series.
-\subsection{Timing of the LGM}
-    Fig.~\ref{fig:timing} -- Timing of the LGM and area time series.
-\subsection{Ice thickness}
-    Fig.~\ref{fig:trimlines} -- Ice thickness comparison to trimlines.
+\subsection{Glacier dynamics}
+    Fig.~\ref{fig:lgmvel} -- Snapshot at 21\,ka and volume time series.\\
+    Fig.~\ref{fig:timing} -- Timing of the LGM and area time series.\\
+    Fig.~\ref{fig:profiles} -- Individual glacier extent profiles.
 \subsection{Erosion potential}
@@
 \section{Methods}
-\subsection{Overview}
-    Tab. -- Model parameters
-\subsection{Ice rheology}
-\subsection{Basal sliding}
-\subsection{Bedrock deformation}
-\subsection{Surface mass balance}
-\subsection{Climate forcing}
+\subsection{Ice sheet model set-up}
+    Tab. -- Model parameters\\
     Fig.~\ref{fig:inputs} -- Climate and geothermal model inputs.\\
+\subsection{Palaeo-climate forcing}
     Tab. -- Records\\
@@
     \begin{figure}
-      \centerline{\includegraphics{alpcyc_hr_trimlines}}
-      \caption{Ice thickness comparison to trimlines.}
-      \label{fig:trimlines}
+      \centerline{\includegraphics{alpcyc_hr_profiles}}
+      \caption{Individual glacier extent profiles.}
+      \label{fig:profiles}
     \end{figure}

### 2017-01-23 figures/alpcyc_hr_erosion.py
Fix power 2.02 (was 2.00) in erosion figure.

@@ -24,7 +24,7 @@ c = nc.variables['velbase_mag'][:]
 kg = 2.7e-7  # m^{1-l} a^{l-1} 
 l = 2.02  # unitless
 c = np.ma.masked_where(thk < 1.0, c)
-erate = (kg*c**2)
+erate = (kg*c**l)
 erosion = erate.sum(axis=0)
 
 # set levels, colors and hatches

### 2017-01-23 articles/cycle/alps-cycle.tex
Add main text figure captions.

@@
 \usepackage[pdftex]{graphicx}
+\usepackage[authoryear,round]{natbib}
+
 \graphicspath{{../../figures/}}
@@
           University of Freiburg, Germany}
+
+\newcommand{\chem}[1]{\ensuremath{\mathrm{#1}}}
+\newcommand{\unit}[1]{\ensuremath{\mathrm{#1}}}
 
@@
       \centerline{\includegraphics{alpcyc_hr_lgmvel}}
-      \caption{Model output at 21\,ka.}
+      \caption{%
+        \textbf{(a)} Modelled bedrock topography (grey) ice surface topography
+        (200\,m contours) and ice surface velocity (blue) in the Alps
+        21~thousand years (ka) before present. Modelled Last Glacial Maximum
+        (LGM) ice extent (dashed orange line) and geomorphological
+        reconstruction \citep[solid red line,][]{Ehlers.etal.2011}. The
+        background map consists of depressed SRTM \citep{Jarvis.etal.2008}
+        topography and Natural Earth Data \citep{Patterson.Kelso.2015}.
+        \textbf{(b)} Temperature offset time-series from the EPICA ice core
+        used as palaeo-climate forcing for the ice flow model \citep[black
+        curve,][]{Jouzel.etal.2007}, and modelled total ice volume through the
+        last glacial cycle (120--0\,ka), expressed in meters of sea level
+        equivalent (m~s.l.e., blue curve). Gray fields indicate Marine
+        Oxygen Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
+        a~global compilation of benthic \chem{\delta^{18}O} records
+        \citep{Lisiecki.Raymo.2005}.}
       \label{fig:lgmvel}
@@
       \centerline{\includegraphics{alpcyc_hr_timing}}
-      \caption{Timing of the LGM and area time series.}
+      \caption{%
+        \textbf{(a)} Timing of the Last Glacial Maximum (LGM) given by the
+        modelled age (colour mapping) and value (200 m contours) of maximum
+        surface elevation throughout the entire simulation.
+        \textbf{(b)} Temperature offset time-series from the EPICA ice core
+        used as palaeo-climate forcing for the ice flow model (black curve),
+        and modelled glaciated area around the LGM (coloured curve). The LGM
+        is here modelled as a time-transgressive event.}
       \label{fig:timing}
@@
       \centerline{\includegraphics{alpcyc_hr_profiles}}
-      \caption{Individual glacier extent profiles.}
+      \caption{%
+        Modelled extent of glaciation along selected profiles for the Lyon,
+        Solothurn, Rhine and Ivrea glaciers.
+        \todo{Implement extraction of ice thickness along a more detailed
+              profile, and add the other glaciers.}}
       \label{fig:profiles}
@@
       \centerline{\includegraphics{alpcyc_hr_erosion}}
-      \caption{Integrated erosion potential.}
+      \caption{%
+        \textbf{(a)} Modelled total erosion integrated in time over the entire
+        simulation (120--0\,ka) is highest in the deep valleys and cirques of
+        the western Alps.
+        \textbf{(b)} Temperature offset time-series from the EPICA ice core
+        used as palaeo-climate forcing for the ice flow model (black curve),
+        and modelled erosion rate integrated in space over the entire
+        ice-covered area. The local erosion rate, $\dot{e}$, is computed from
+        the sliding velocity, $\vec{v}_{\mathrm{b}}$, through
+        $\dot{e} = K_g \cdot |\vec{v}_{\mathrm{b}}|^{l}$, with
+        $l = 2.02$ and $K_g = 2.7 \cdot 10^{-7}\,m^{1-l}\,a^{l-1}$
+        \citep{Herman.etal.2015}.}
       \label{fig:erosion}

### 2017-01-23 articles/cycle/alps-cycle.tex
Add supplemental figure captions.

@@
 \subsection{Ice sheet model set-up}
-    Tab. -- Model parameters\\
+    Table~\ref{tab:params} -- Model parameters\\
     Fig.~\ref{fig:inputs} -- Climate and geothermal model inputs.\\
 \subsection{Palaeo-climate forcing}
-    Tab. -- Records\\
+    Table~\ref{tab:records} -- Palaeo-climate records.\\
     Fig.~\ref{fig:timeseries} -- Low-resolution time series.\\
     Fig.~\ref{fig:footprints} -- Low-resolution ice cover.
-
-\section{Conclusion}
 
@@
       \centerline{\includegraphics{alpcyc_hr_inputs}}
-      \caption{Climate and geothermal model inputs.}
+      \caption{%
+        \textbf{(a)} July mean near-surface air temperature and
+        \textbf{(b)} January precipitation from WorldClim
+        \citep[1960--1990]{Hijmans.etal.2005}, and
+        \textbf{(c)} Mordern July standard deviation of daily mean temperature
+        from the ERA-Interim \citep[1979--2012]{Dee.etal.2011} from the
+        reference monthly climatology used to force the surface mass balance
+        (PDD) component of the ice sheet model.
+        \textbf{(d)} Geothermal heat flow from applying the similarity method
+        to multiple geophysical proxies \citep{Goutorbe.etal.2011} used as a
+        boundary condition to the bedrock thermal model 3\,km below the
+        ice-bedrock interface.}
       \label{fig:inputs}
@@
       \centerline{\includegraphics{alpcyc_lr_timeseries}}
-      \caption{Low-resolution time series.}
+      \caption{%
+        \textbf{(a)} Temperature offset time-series from ice core and ocean
+        records (Table~\ref{tab:records}) used as palaeo-climate forcing for
+        the ice sheet model.
+        \textbf{(a)} Modelled total ice volume through the last 120\,ka,
+        expressed in meters of sea level equivalent (m~s.l.e.). Gray fields
+        indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2 and
+        MIS~4 according to a~global compilation of benthic \chem{\delta^{18}O}
+        records \citep{Lisiecki.Raymo.2005}.}
       \label{fig:timeseries}
@@
       \centerline{\includegraphics{alpcyc_lr_footprints}}
-      \caption{Low-resolution ice cover.}
+      \caption{%
+        \textbf{(a--c)} Cumulative extent of modelled ice cover during MIS~2
+        (29--14\,ka) using temperature time-series scaling factors
+        (Table~\ref{tab:records}) adjusted to obtain model results in agreement
+        with the Last Glacial Maximum (LGM) geomorphological reconstruction
+        \citep[solid red line,][]{Ehlers.etal.2011}.
+        \textbf{(d--f)} Cumulative extent of modelled ice cover during MIS~4
+        (71--57\,ka). Only the simulation driven by the EPICA temperature
+        time-series yields reasonable MIS~4 ice cover.}
       \label{fig:footprints}

### 2017-01-23 articles/cycle/alps-cycle.tex
Add palaeo-temperature records table.

@@
 
+\usepackage{amsmath}
 \usepackage{authblk}
+\usepackage{booktabs}
 \usepackage[T1]{fontenc}
@@
 
+% Common units
+\newcommand{\e}[1]{\ensuremath{\times 10^{#1}}}
 \newcommand{\chem}[1]{\ensuremath{\mathrm{#1}}}
 \newcommand{\unit}[1]{\ensuremath{\mathrm{#1}}}
+\newcommand{\degree}[0]{\ensuremath{^{\circ}}}
+\newcommand{\degC}[0]{\unit{{\degree}C}}
 
@@
 % Figures
+\clearpage
 % ----------------------------------------------------------------------
@@
         $\dot{e} = K_g \cdot |\vec{v}_{\mathrm{b}}|^{l}$, with
-        $l = 2.02$ and $K_g = 2.7 \cdot 10^{-7}\,m^{1-l}\,a^{l-1}$
+        $l = 2.02$ and $K_g = 2.7\e{-7}\,m^{1-l}\,a^{l-1}$
         \citep{Herman.etal.2015}.}
@@
 % Supplementary tables
+\clearpage
 \setcounter{table}{0}
@@
 
+    \begin{table*}
+      \caption{%
+        Palaeo-temperature proxy records and scaling factors yielding
+        temperature offset time-series used to force the ice sheet model
+        through the last glacial cycle (Fig.~\ref{fig:timeseries}). $f$
+        corresponds to the scaling factor adopted to yield Last Glacial Maximum
+        ice limits in the vicinity of mapped end moraines
+        (Fig.~\ref{fig:footprints}a), and $[{\Delta}T_{\textrm{TS}}]_{32}^{22}$
+        refers to the resulting mean temperature anomaly during the period 32
+        to~22\,\unit{ka} after scaling.}
+      \label{tab:records}
+      %\scalebox{.850}[.850]
+      \centerline{\begin{tabular}{lccccccl}
+        \toprule
+        Record    & Latitude & Longitude & Elev. (m~a.s.l.)
+                  & Proxy & $f$ & $[{\Delta}\text{TS}]_{32}^{22}$ (K)
+                  & Reference\\
+        \midrule
+        GRIP      &  72{\degree}35$^{\prime}$\,N   % 72.58 (decimal)
+                  &  37{\degree}38$^{\prime}$\,W   % 37.64 (decimal)
+                  & 3238
+                  & \chem{\delta^{18}O}
+                  & 0.46 & $-$7.6  % -16.4126 (before scaling)
+                  & \citet{Dansgaard.etal.1993} \\
+        EPICA     &  75{\degree}06$^{\prime}$\,S   % 75.1
+                  & 123{\degree}21$^{\prime}$\,E   % 123.35
+                  & 3233
+                  & \chem{\delta^{18}O}
+                  & 0.46 & $-$9.5  % -9.2055
+                  & \citet{Jouzel.etal.2007} \\
+        MD01-2444 &  37{\degree}34$^{\prime}$\,N   % 37.561
+                  &  10{\degree}04$^{\prime}$\,W   % -10.142
+                  & $-$2637
+                  & \chem{U^{K'}_{37}}
+                  & 1.84 & $-$8.0  % -4.345625
+                  & \citet{Martrat.etal.2007} \\
+        \bottomrule
+      \end{tabular}}
+    \end{table*}
+
 % ----------------------------------------------------------------------
 % Supplementary figures
+\clearpage
 \setcounter{figure}{0}

### 2017-01-23 figures/alpcyc_lr_records.py
Fix a small shift in coordinates in record map.

@@ -25,8 +25,8 @@ ax.plot(x, y, 'k', lw=1.0, transform=ut.pl.utm, zorder=3)
 records = ['GRIP', 'EPICA', 'MD01-2444']
 colors = ['darkblue', 'darkred', 'darkgreen']
 colors = [ut.pl.palette[c] for c in colors]
-lats = np.array([72.58, -75.1, 37.565])
-lons = np.array([-37.64, 123.35, -10.134])
+lats = np.array([72.58, -75.1, 37.561])
+lons = np.array([-37.64, 123.35, -10.142])
 locs = ['uc', 'uc', 'lc']
 x, y, _ = ax.projection.transform_points(ut.pl.ll, lons, lats).T
 ax.scatter(lons, lats, c=colors, linewidths=0.25, transform=ut.pl.ll)

### 2017-01-24 articles/cycle/alps-cycle.tex
Add parameter table from Cordillera cycle paper.

@@
       \caption{%
+        Parameter values used in the ice sheet model.}
+      \label{tab:params}
+      \noindent\small\makebox[\textwidth]
+      {\begin{tabular}{llrll}
+        \toprule
+
+        Not.    & Name & Value & Unit & Source \\
+
+        \midrule
+        \multicolumn{2}{l}{{Ice rheology}} \\
+        \midrule
+
+        $\rho$  & Ice density
+                & 910
+                & \unit{kg\,m^{-3}}
+                & \citet{Aschwanden.etal.2012} \\
+
+        $g$     & Standard gravity
+                & 9.81
+                & \unit{m\,s^{-2}}
+                & \citet{Aschwanden.etal.2012} \\
+
+        $n$     & Glen exponent
+                & 3
+                & --
+                & \citet{Cuffey.Paterson.2010} \\
+
+        $A_{\mathrm{c}}$   & Ice hardness coefficient cold
+                & $3.61\e{-13}$
+                & \unit{Pa^{-3}\,s^{-1}}
+                & \citet{Paterson.Budd.1982} \\
+
+        $A_{\mathrm{w}}$   & Ice hardness coefficient warm
+                & $1.73\e3$
+                & \unit{Pa^{-3}\,s^{-1}}
+                & \citet{Paterson.Budd.1982} \\
+
+        $Q_{\mathrm{c}}$   & Flow law activation energy cold
+                & $6.0\e4$
+                & \unit{J\,mol^{-1}}
+                & \citet{Paterson.Budd.1982} \\
+
+        $Q_{\mathrm{w}}$   & Flow law activation energy warm
+                & $13.9\e4$
+                & \unit{J\,mol^{-1}}
+                & \citet{Paterson.Budd.1982} \\
+
+        $T_{\mathrm{c}}$   & Flow law critical temperature
+                & 263.15
+                & \unit{K}
+                & \citet{Paterson.Budd.1982} \\
+
+        $f$     & Flow law water fraction coeff.
+                & 181.25
+                & --
+                & \citet{Lliboutry.Duval.1985} \\
+
+        $R$     & Ideal gas constant
+                & 8.31441
+                & \unit{J\,mol^{-1}\,K^{-1}}
+                & -- \\
+
+        $\beta$ & Clapeyron constant
+                & $7.9\e{-8}$
+                & \unit{K\,Pa^{-1}}
+                & \citet{Luthi.etal.2002} \\
+
+        $c_{\mathrm{i}}$   & Ice specific heat capacity
+                & 2009
+                & \unit{J\,kg^{-1}\,K^{-1}}
+                & \citet{Aschwanden.etal.2012} \\
+
+        $c_{\mathrm{w}}$   & Water specific heat capacity
+                & 4170
+                & \unit{J\,kg^{-1}\,K^{-1}}
+                & \citet{Aschwanden.etal.2012} \\
+
+        $k$     & Ice thermal conductivity
+                & 2.10
+                & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
+                & \citet{Aschwanden.etal.2012} \\
+
+        $L$     & Water latent heat of fusion
+                & $3.34\e5$
+                & \unit{J\,kg^{-1}\,K^{-1}}
+                & \citet{Aschwanden.etal.2012} \\
+
+        \midrule
+        \multicolumn{2}{l}{{Basal sliding}} \\
+        \midrule
+
+        $q$     & Pseudo-plastic sliding exponent
+                & 0.25
+                & --
+                & \citet{Aschwanden.etal.2013} \\
+
+        $v_{\text{th}}$& Pseudo-plastic threshold velocity
+                & 100.0
+                & \unit{m\,a^{-1}}
+                & \citet{Aschwanden.etal.2013} \\
+
+        $c_0$   & Till cohesion
+                & 0.0
+                & Pa
+                & \citet{Tulaczyk.etal.2000} \\
+
+        $e_0$   & Till reference void ratio
+                & 0.69
+                & --
+                & \citet{Tulaczyk.etal.2000} \\
+
+        $C_{\mathrm{c}}$   & Till compressibility coefficient
+                & 0.12
+                & --
+                & \citet{Tulaczyk.etal.2000} \\
+
+        $\delta$& Minimum effective pressure ratio
+                & 0.02
+                & --
+                & \citet{Bueler.Pelt.2015} \\
+
+        $W_{\text{max}}$ & Maximal till water thickness
+                & 2.0
+                & m
+                & \citet{Bueler.Pelt.2015} \\
+
+        $b_0$   & Altitude of max. friction angle
+                & 0
+                & m
+                & -- \\
+
+        $b_1$   & Altitude of min. friction angle
+                & 200
+                & m
+                & \citet{Clague.1981} \\
+
+        $\phi_0$& Minimum friction angle
+                & 15
+                & \degree
+                & -- \\
+
+        $\phi_1$& Maximum friction angle
+                & 45
+                & \degree
+                & -- \\
+
+        \midrule
+        \multicolumn{2}{l}{{Bedrock and lithosphere}} \\
+        \midrule
+
+        $q_{\mathrm{G}}$   & Geothermal heat flux
+                & 70.0
+                & \unit{mW\,m^{-2}}
+                & -- \\
+
+        $\rho_{\mathrm{b}}$& Bedrock density
+                & 3300
+                & \unit{kg\,m^{-3}}
+                & -- \\
+
+        $c_{\mathrm{b}}$   & Bedrock specific heat capacity
+                & 1000
+                & \unit{J\,kg^{-1}\,K^{-1}}
+                & -- \\
+
+        $k_{\mathrm{b}}$   & Bedrock thermal conductivity
+                & 3.0
+                & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
+                & -- \\
+
+        $\nu_{\mathrm{m}}$ & Astenosphere viscosity
+                & $1\e{19}$
+                & \unit{Pa\,s}
+                & \citet{James.etal.2009} \\
+
+        $\rho_{\mathrm{l}}$& Lithosphere density
+                & 3300
+                & \unit{kg\,m^{-3}}
+                & \citet{Lingle.Clark.1985} \\
+
+        $D$     & Lithosphere flexural rigidity
+                & $5.0\e{24}$
+                & \unit{N}
+                & \citet{Lingle.Clark.1985} \\
+
+        \midrule
+        \multicolumn{2}{l}{{Surface and atmosphere}} \\
+        \midrule
+
+        $T_{\mathrm{s}}$   & Temperature of snow precipitation
+                & 273.15
+                & \unit{K}
+                & -- \\
+
+        $T_{\mathrm{r}}$   & Temperature of rain precipitation
+                & 275.15
+                & \unit{K}
+                & -- \\
+
+        $F_{\mathrm{s}}$   & Degree-day factor for snow
+                & $3.04\e{-3}$
+                & \unit{m\,K^{-1}\,day^{-1}}
+                & \citet{Shea.etal.2009} \\
+
+        $F_{\mathrm{i}}$   & Degree-day factor for ice
+                & $4.59\e{-3}$
+                & \unit{m\,K^{-1}\,day^{-1}}
+                & \citet{Shea.etal.2009} \\
+
+        $\gamma$& Air temperature lapse rate
+                & $6\e{-3}$
+                & \unit{K\,m{-1}}
+                & -- \\
+
+        \bottomrule
+      \end{tabular}}
+    \end{table*}
+
+    \begin{table*}
+      \caption{%
         Palaeo-temperature proxy records and scaling factors yielding
@@
       \label{tab:records}
-      %\scalebox{.850}[.850]
-      \centerline{\begin{tabular}{lccccccl}
+      \noindent\small\makebox[\textwidth]
+      {\begin{tabular}{lccccccl}
         \toprule
+
         Record    & Latitude & Longitude & Elev. (m~a.s.l.)
@@
                   & Reference\\
+
         \midrule
+
         GRIP      &  72{\degree}35$^{\prime}$\,N   % 72.58 (decimal)
@@
                   & \citet{Dansgaard.etal.1993} \\
+
         EPICA     &  75{\degree}06$^{\prime}$\,S   % 75.1
@@
                   & \citet{Jouzel.etal.2007} \\
+
         MD01-2444 &  37{\degree}34$^{\prime}$\,N   % 37.561
@@
                   & \citet{Martrat.etal.2007} \\
+
         \bottomrule

### 2017-01-24 articles/cycle/alps-cycle.tex
Update parameter table to alpine set-up.

@@
       \caption{%
-        Parameter values used in the ice sheet model.}
+        Parameter values used in the ice sheet model.
+        \idea{If I need to re-run all simulations, we could rethink bedrock,
+              lithosphere and PDD parameters.}}
       \label{tab:params}
@@
         $A_{\mathrm{c}}$   & Ice hardness coefficient cold
-                & $3.61\e{-13}$
+                & $2.847\e{-13}$
                 & \unit{Pa^{-3}\,s^{-1}}
-                & \citet{Paterson.Budd.1982} \\
+                & \citet{Cuffey.Paterson.2010} \\
 
         $A_{\mathrm{w}}$   & Ice hardness coefficient warm
-                & $1.73\e3$
+                & $2.356\e{-2}$
                 & \unit{Pa^{-3}\,s^{-1}}
-                & \citet{Paterson.Budd.1982} \\
+                & \citet{Cuffey.Paterson.2010} \\
 
@@
                 & \unit{J\,mol^{-1}}
-                & \citet{Paterson.Budd.1982} \\
+                & \citet{Cuffey.Paterson.2010} \\
 
         $Q_{\mathrm{w}}$   & Flow law activation energy warm
-                & $13.9\e4$
+                & $11.5\e4$
                 & \unit{J\,mol^{-1}}
-                & \citet{Paterson.Budd.1982} \\
+                & \citet{Cuffey.Paterson.2010} \\
+
+        $E_{\text{SIA}}$   & SIA enhancement factor
+                & 5
+                & --
+                & \citet{Cuffey.Paterson.2010} \\
+
+        $E_{\text{SSA}}$   & SSA enhancement factor
+                & 1
+                & --
+                & \citet{Cuffey.Paterson.2010} \\
 
@@
         $v_{\text{th}}$& Pseudo-plastic threshold velocity
-                & 100.0
+                & 100
                 & \unit{m\,a^{-1}}
@@
         $c_0$   & Till cohesion
-                & 0.0
+                & 0
                 & Pa
@@
 
-        $W_{\text{max}}$ & Maximal till water thickness
-                & 2.0
+        $\phi$  & Till friction angle
+                & 30
+                & \degree
+                & \citet{Cuffey.Paterson.2010} \\
+
+        $W_{\text{max}}$ & Maximum till water thickness
+                & 2
                 & m
                 & \citet{Bueler.Pelt.2015} \\
-
-        $b_0$   & Altitude of max. friction angle
-                & 0
-                & m
-                & -- \\
-
-        $b_1$   & Altitude of min. friction angle
-                & 200
-                & m
-                & \citet{Clague.1981} \\
-
-        $\phi_0$& Minimum friction angle
-                & 15
-                & \degree
-                & -- \\
-
-        $\phi_1$& Maximum friction angle
-                & 45
-                & \degree
-                & -- \\
 
@@
         \midrule
-
-        $q_{\mathrm{G}}$   & Geothermal heat flux
-                & 70.0
-                & \unit{mW\,m^{-2}}
-                & -- \\
 
@@
         $k_{\mathrm{b}}$   & Bedrock thermal conductivity
-                & 3.0
+                & 3
                 & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
@@
         $D$     & Lithosphere flexural rigidity
-                & $5.0\e{24}$
+                & $5\e{24}$
                 & \unit{N}
@@
         $F_{\mathrm{s}}$   & Degree-day factor for snow
-                & $3.04\e{-3}$
+                & $3.297\e{-3}$
                 & \unit{m\,K^{-1}\,day^{-1}}
-                & \citet{Shea.etal.2009} \\
+                & \citet{Huybrechts.1998} \\
 
         $F_{\mathrm{i}}$   & Degree-day factor for ice
-                & $4.59\e{-3}$
+                & $8.791\e{-3}$
                 & \unit{m\,K^{-1}\,day^{-1}}
-                & \citet{Shea.etal.2009} \\
+                & \citet{Huybrechts.1998} \\
+
+        $R$     & Refreezing fraction
+                & 0.0
+                & --
+                & -- \\
 

### 2017-01-25 articles/cycle/alps-cycle.tex
Develop main text (one sentence per paragraph).

@@
 
+% ----------------------------------------------------------------------
 \section{Main}
+% ----------------------------------------------------------------------
+
 \subsection{Introduction}
+
+    \begin{itemize}
+    \item The European Alps are the cradle of pioneer glacial studies.
+    \item Today, their glacial history is better understood than any other.
+    \item The LGM extent is very well known (Fig.~\ref{fig:lgmvel}a).
+    \item But modelling was difficult due to the steep topography.
+    \item Here, we us PISM to model the last glacial cycle (120--0\,ka).
+    \end{itemize}
+
 \subsection{Glacier dynamics}
@@
     Fig.~\ref{fig:profiles} -- Individual glacier extent profiles.
+
+    \begin{itemize}
+    \item Rapid variations of total ice volume (Fig.~\ref{fig:lgmvel}b).
+    \item Two major glaciations during MIS 4 and 2 (Fig.~\ref{fig:lgmvel}b).
+    \item Maximum stages flow pattern is complex (Fig.~\ref{fig:lgmvel}a).
+    \item Maximum extent is farily well reproduced (Fig.~\ref{fig:lgmvel}a).
+    \item Maximum extent is time-transgressive (Fig.~\ref{fig:timing}).
+    \item Glaciers have different response times (Fig.~\ref{fig:timing}).
+    \item Some glaciers advance many times (Fig.~\ref{fig:profiles}).
+    \end{itemize}
+
 \subsection{Erosion potential}
@@
 
+    \begin{itemize}
+    \item We calculate erosion after \citet{Herman.etal.2015}
+    \item More erosion in valleys of western Alps (Fig.~\ref{fig:erosion}a).
+    \item Erosion is constant during 110--15\,ka (Fig.~\ref{fig:erosion}b).
+    \end{itemize}
+
+% ----------------------------------------------------------------------
 \section{Methods}
+% ----------------------------------------------------------------------
+
 \subsection{Ice sheet model set-up}

### 2017-01-25 articles/cycle/alps-cycle.tex
Develop methods (one sentence per paragraph).

@@
     Fig.~\ref{fig:inputs} -- Climate and geothermal model inputs.\\
+
+    \begin{itemize}
+    \item Ice dynamics are approximated by SSA+SIA.
+    \item Ice temperature follows an enthalpy scheme.
+    \item Bedrock temperature computed until 3\,km depth
+    \item Temperature conditioned by geothermal flux (Fig.~\ref{fig:inputs}d).
+    \item Basal sliding follwos a pseudo-plastic law.
+    \item Yield stress is determined by a Mohr-Coulomb criterion.
+    \item Till friction angle is constant.
+    \item Till effective pressure relates to basal melt.
+    \item Surface mass balance is computed by a PDD model.
+    \item Modern T and P forcing from WorldClim (Fig.~\ref{fig:inputs}a and~b).
+    \item Modern PDD SD forcing from ERA-Interim (Fig.~\ref{fig:inputs}c).
+    \item SRTM topography with glaciers removed \citep{Huss.Farinotti.2012}.
+    \end{itemize}
+
 \subsection{Palaeo-climate forcing}
@@
     Fig.~\ref{fig:footprints} -- Low-resolution ice cover.
+
+    \begin{itemize}
+    \item We conducted a 5/,km-resolution sensitivity study.
+    \item We used three palaeo-temperature records (Table~\ref{tab:records}).
+    \item All records scaled to fit the LGM extent (Table~\ref{tab:records}).
+    \item All yield glaciation during MIS~4 and 2 (Fig.~\ref{fig:timeseries}).
+    \item But otherwise very different results (Fig.~\ref{fig:timeseries}).
+    \item Two yield two large MIS~4 ice extent (Fig.~\ref{fig:footprints}).
+    \item Only EPICA yield realistic MIS~4 (Fig.~\ref{fig:footprints}).
+    \end{itemize}
 

### 2017-01-25 articles/cycle/alps-cycle.tex
Add notes and ideas for co-autors.
Also fix wrong EPICA scaling factore in records table.

@@
 \begin{abstract}
+
+    \note{For the Nature Geoscience format this abstract will need to be
+          converted to an ``introductory paragraph''.}
 
@@
     \begin{itemize}
-    \item We conducted a 5/,km-resolution sensitivity study.
+    \item We conducted a 5\,km-resolution sensitivity study.
     \item We used three palaeo-temperature records (Table~\ref{tab:records}).
@@
     \end{itemize}
+
+    \note{An important assumption in the method is that the MIS~4 glaciation
+          was less (or equally) extensive than the MIS~2, i.e. that the LGM
+          outline from \citet{Ehlers.etal.2011} dates indeed from the LGM
+          (MIS~2). A different assumption could be that this outline represents
+          the most extensive glaciation in the last 120\,ka. I we allow the
+          MIS~2 glaciation to be a bit smaller than MIS~4, simulations driven
+          by GRIP and MD01-2444 also give reasonable results.}
 
@@
         and modelled glaciated area around the LGM (coloured curve). The LGM
-        is here modelled as a time-transgressive event.}
+        is here modelled as a time-transgressive event.
+        \note{The Nature Geoscience format is limited to four figures. Now we
+              indeed have four figures, albeit large ones.
+              Figure~\ref{fig:timing} will interest the dating people but is
+              not crucial for the paper. My plan is to include in the main text
+              for the first submission and be prepared that the editor may want
+              to move it into supplementary material.}}
       \label{fig:timing}
@@
         \todo{Implement extraction of ice thickness along a more detailed
-              profile, and add the other glaciers.}}
+              profile, and add the other glaciers.}
+        \idea{We can also move this as panel b on Fig.~\ref{fig:timing}, with
+              profiles drawn on panel a. But then we should probably limit it
+              to two glaciers instead of four.}}
       \label{fig:profiles}
@@
         refers to the resulting mean temperature anomaly during the period 32
-        to~22\,\unit{ka} after scaling.}
+        to~22\,\unit{ka} after scaling.
+        \note{The EPICA scaling factor is close to one, and that ended up
+              yielding too much ice in the high-resolution simulation
+              (Fig.~\ref{fig:lgmvel}), though not so at low resotion
+              (Fig.~\ref{fig:footprints}b). If I re-run the whole thing, I will
+              probably fix it to one (i.e. use EPICA temperature anomalies
+              without applying any scaling).}}
       \label{tab:records}
@@
                   & \chem{\delta^{18}O}
-                  & 0.46 & $-$9.5  % -9.2055
+                  & 1.03 & $-$9.5  % -9.2055
                   & \citet{Jouzel.etal.2007} \\
@@
         boundary condition to the bedrock thermal model 3\,km below the
-        ice-bedrock interface.}
+        ice-bedrock interface.
+        \idea{Since this will be in the supplement, I can also add additional
+              information such as winter temperature, summer precipitation,
+              winter PDD SD, and/or basal topography with modern glaciers
+              removed, e.g. in the Bernese Alps.}}
       \label{fig:inputs}

### 2017-01-31 articles/cycle/alps-cycle.tex
Add ideas from Frank and Matthias in the text.

@@
 
+    \idea{Matthias: I think it is very important to emphasize the knowledge gap
+          that you fill in with your work in the Intro. (1) it is incompletely
+          known how many advances occurred during the Last glacial cycle, (2)
+          what drove the different response of the individual lobes (you show
+          that it is glacier response time, i.e. ice dynamics), (3) how the
+          temporal dynamics of subglacial erosion was structured.}
+
 \subsection{Glacier dynamics}
@@
 
+    \idea{Matthias: The selection of figures in the main text and SOM is good.
+          I find Figure 2 quite difficult to understand at first glance. I
+          would combine it (already before submission) with Fig. 3. the Figure
+          can then focus both on the temporal advance/retreat patterns of the 4
+          lobes (four panels on the left side of figure) and the spatial
+          differences in max. ice elevation (1 panel on right side of figure,
+          with centerlines indicated). I would have the one with trimline
+          elevation again in the Supplementary.}
+
 \subsection{Erosion potential}
     Fig.~\ref{fig:erosion} -- Integrated erosion potential.
+
+    \idea{Frank: I would keep this topic for another paper.}
 

### 2017-07-04 articles/cycle/alps-cycle.tex
Rework outline into standard (not Nature) format.
Giving up on Nature and the like. I prefer to promote a somewhat ethical
open-access publisher instead.

@@
 % ----------------------------------------------------------------------
-\section{Main}
+\section{Introduction}
 % ----------------------------------------------------------------------
-
-\subsection{Introduction}
 
@@
 
-\subsection{Glacier dynamics}
+% ----------------------------------------------------------------------
+\section{Ice sheet model set-up}
+% ----------------------------------------------------------------------
+
+    Table~\ref{tab:params} -- Model parameters\\
+    Fig.~\ref{fig:inputs} -- Climate and geothermal model inputs.\\
+
+    \begin{itemize}
+    \item Ice dynamics are approximated by SSA+SIA.
+    \item Ice temperature follows an enthalpy scheme.
+    \item Bedrock temperature computed until 3\,km depth
+    \item Temperature conditioned by geothermal flux (Fig.~\ref{fig:inputs}d).
+    \item Basal sliding follwos a pseudo-plastic law.
+    \item Yield stress is determined by a Mohr-Coulomb criterion.
+    \item Till friction angle is constant.
+    \item Till effective pressure relates to basal melt.
+    \item Surface mass balance is computed by a PDD model.
+    \item Modern T and P forcing from WorldClim (Fig.~\ref{fig:inputs}a and~b).
+    \item Modern PDD SD forcing from ERA-Interim (Fig.~\ref{fig:inputs}c).
+    \item SRTM topography with glaciers removed \citep{Huss.Farinotti.2012}.
+    \end{itemize}
+
+% ----------------------------------------------------------------------
+\section{Palaeo-climate forcing}
+% ----------------------------------------------------------------------
+
+    Table~\ref{tab:records} -- Palaeo-climate records.\\
+    Fig.~\ref{fig:timeseries} -- Low-resolution time series.\\
+    Fig.~\ref{fig:footprints} -- Low-resolution ice cover.
+
+    \begin{itemize}
+    \item We conducted a 5\,km-resolution sensitivity study.
+    \item We used three palaeo-temperature records (Table~\ref{tab:records}).
+    \item All records scaled to fit the LGM extent (Table~\ref{tab:records}).
+    \item All yield glaciation during MIS~4 and 2 (Fig.~\ref{fig:timeseries}).
+    \item But otherwise very different results (Fig.~\ref{fig:timeseries}).
+    \item Two yield two large MIS~4 ice extent (Fig.~\ref{fig:footprints}).
+    \item Only EPICA yield realistic MIS~4 (Fig.~\ref{fig:footprints}).
+    \end{itemize}
+
+    \note{An important assumption in the method is that the MIS~4 glaciation
+          was less (or equally) extensive than the MIS~2, i.e. that the LGM
+          outline from \citet{Ehlers.etal.2011} dates indeed from the LGM
+          (MIS~2). A different assumption could be that this outline represents
+          the most extensive glaciation in the last 120\,ka. I we allow the
+          MIS~2 glaciation to be a bit smaller than MIS~4, simulations driven
+          by GRIP and MD01-2444 also give reasonable results.}
+
+% ----------------------------------------------------------------------
+\section{Glacier dynamics}
+% ----------------------------------------------------------------------
+
     Fig.~\ref{fig:lgmvel} -- Snapshot at 21\,ka and volume time series.\\
@@
 
-\subsection{Erosion potential}
+% ----------------------------------------------------------------------
+\section{Erosion potential}
+% ----------------------------------------------------------------------
+
     Fig.~\ref{fig:erosion} -- Integrated erosion potential.
@@
 % ----------------------------------------------------------------------
-\section{Methods}
+\section{Conclusions}
 % ----------------------------------------------------------------------
 
-\subsection{Ice sheet model set-up}
-    Table~\ref{tab:params} -- Model parameters\\
-    Fig.~\ref{fig:inputs} -- Climate and geothermal model inputs.\\
-
     \begin{itemize}
-    \item Ice dynamics are approximated by SSA+SIA.
-    \item Ice temperature follows an enthalpy scheme.
-    \item Bedrock temperature computed until 3\,km depth
-    \item Temperature conditioned by geothermal flux (Fig.~\ref{fig:inputs}d).
-    \item Basal sliding follwos a pseudo-plastic law.
-    \item Yield stress is determined by a Mohr-Coulomb criterion.
-    \item Till friction angle is constant.
-    \item Till effective pressure relates to basal melt.
-    \item Surface mass balance is computed by a PDD model.
-    \item Modern T and P forcing from WorldClim (Fig.~\ref{fig:inputs}a and~b).
-    \item Modern PDD SD forcing from ERA-Interim (Fig.~\ref{fig:inputs}c).
-    \item SRTM topography with glaciers removed \citep{Huss.Farinotti.2012}.
+    \item Summary of the previous stuff.
     \end{itemize}
-
-\subsection{Palaeo-climate forcing}
-    Table~\ref{tab:records} -- Palaeo-climate records.\\
-    Fig.~\ref{fig:timeseries} -- Low-resolution time series.\\
-    Fig.~\ref{fig:footprints} -- Low-resolution ice cover.
-
-    \begin{itemize}
-    \item We conducted a 5\,km-resolution sensitivity study.
-    \item We used three palaeo-temperature records (Table~\ref{tab:records}).
-    \item All records scaled to fit the LGM extent (Table~\ref{tab:records}).
-    \item All yield glaciation during MIS~4 and 2 (Fig.~\ref{fig:timeseries}).
-    \item But otherwise very different results (Fig.~\ref{fig:timeseries}).
-    \item Two yield two large MIS~4 ice extent (Fig.~\ref{fig:footprints}).
-    \item Only EPICA yield realistic MIS~4 (Fig.~\ref{fig:footprints}).
-    \end{itemize}
-
-    \note{An important assumption in the method is that the MIS~4 glaciation
-          was less (or equally) extensive than the MIS~2, i.e. that the LGM
-          outline from \citet{Ehlers.etal.2011} dates indeed from the LGM
-          (MIS~2). A different assumption could be that this outline represents
-          the most extensive glaciation in the last 120\,ka. I we allow the
-          MIS~2 glaciation to be a bit smaller than MIS~4, simulations driven
-          by GRIP and MD01-2444 also give reasonable results.}
 
@@
 % ----------------------------------------------------------------------
+
+    \begin{figure}
+      \centerline{\includegraphics{alpcyc_hr_inputs}}
+      \caption{%
+        \textbf{(a)} July mean near-surface air temperature and
+        \textbf{(b)} January precipitation from WorldClim
+        \citep[1960--1990]{Hijmans.etal.2005}, and
+        \textbf{(c)} Mordern July standard deviation of daily mean temperature
+        from the ERA-Interim \citep[1979--2012]{Dee.etal.2011} from the
+        reference monthly climatology used to force the surface mass balance
+        (PDD) component of the ice sheet model.
+        \textbf{(d)} Geothermal heat flow from applying the similarity method
+        to multiple geophysical proxies \citep{Goutorbe.etal.2011} used as a
+        boundary condition to the bedrock thermal model 3\,km below the
+        ice-bedrock interface.
+        \idea{Since this will be in the supplement, I can also add additional
+              information such as winter temperature, summer precipitation,
+              winter PDD SD, and/or basal topography with modern glaciers
+              removed, e.g. in the Bernese Alps.}}
+      \label{fig:inputs}
+    \end{figure}
+
+    \begin{figure}
+      \centerline{\includegraphics{alpcyc_lr_timeseries}}
+      \caption{%
+        \textbf{(a)} Temperature offset time-series from ice core and ocean
+        records (Table~\ref{tab:records}) used as palaeo-climate forcing for
+        the ice sheet model.
+        \textbf{(a)} Modelled total ice volume through the last 120\,ka,
+        expressed in meters of sea level equivalent (m~s.l.e.). Gray fields
+        indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2 and
+        MIS~4 according to a~global compilation of benthic \chem{\delta^{18}O}
+        records \citep{Lisiecki.Raymo.2005}.}
+      \label{fig:timeseries}
+    \end{figure}
+
+    \begin{figure}
+      \centerline{\includegraphics{alpcyc_lr_footprints}}
+      \caption{%
+        \textbf{(a--c)} Cumulative extent of modelled ice cover during MIS~2
+        (29--14\,ka) using temperature time-series scaling factors
+        (Table~\ref{tab:records}) adjusted to obtain model results in agreement
+        with the Last Glacial Maximum (LGM) geomorphological reconstruction
+        \citep[solid red line,][]{Ehlers.etal.2011}.
+        \textbf{(d--f)} Cumulative extent of modelled ice cover during MIS~4
+        (71--57\,ka). Only the simulation driven by the EPICA temperature
+        time-series yields reasonable MIS~4 ice cover.}
+      \label{fig:footprints}
+    \end{figure}
 
@@
 % ----------------------------------------------------------------------
-% Supplementary tables
+% Tables
 \clearpage
-\setcounter{table}{0}
-\renewcommand{\thetable}{S\arabic{table}}%
 % ----------------------------------------------------------------------
@@
                 & $5\e{24}$
-                & \unit{N}
+                & \unit{N\,m}
                 & \citet{Lingle.Clark.1985} \\
@@
 
-% ----------------------------------------------------------------------
-% Supplementary figures
-\clearpage
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}%
-% ----------------------------------------------------------------------
-
-    \begin{figure}
-      \centerline{\includegraphics{alpcyc_hr_inputs}}
-      \caption{%
-        \textbf{(a)} July mean near-surface air temperature and
-        \textbf{(b)} January precipitation from WorldClim
-        \citep[1960--1990]{Hijmans.etal.2005}, and
-        \textbf{(c)} Mordern July standard deviation of daily mean temperature
-        from the ERA-Interim \citep[1979--2012]{Dee.etal.2011} from the
-        reference monthly climatology used to force the surface mass balance
-        (PDD) component of the ice sheet model.
-        \textbf{(d)} Geothermal heat flow from applying the similarity method
-        to multiple geophysical proxies \citep{Goutorbe.etal.2011} used as a
-        boundary condition to the bedrock thermal model 3\,km below the
-        ice-bedrock interface.
-        \idea{Since this will be in the supplement, I can also add additional
-              information such as winter temperature, summer precipitation,
-              winter PDD SD, and/or basal topography with modern glaciers
-              removed, e.g. in the Bernese Alps.}}
-      \label{fig:inputs}
-    \end{figure}
-
-    \begin{figure}
-      \centerline{\includegraphics{alpcyc_lr_timeseries}}
-      \caption{%
-        \textbf{(a)} Temperature offset time-series from ice core and ocean
-        records (Table~\ref{tab:records}) used as palaeo-climate forcing for
-        the ice sheet model.
-        \textbf{(a)} Modelled total ice volume through the last 120\,ka,
-        expressed in meters of sea level equivalent (m~s.l.e.). Gray fields
-        indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2 and
-        MIS~4 according to a~global compilation of benthic \chem{\delta^{18}O}
-        records \citep{Lisiecki.Raymo.2005}.}
-      \label{fig:timeseries}
-    \end{figure}
-
-    \begin{figure}
-      \centerline{\includegraphics{alpcyc_lr_footprints}}
-      \caption{%
-        \textbf{(a--c)} Cumulative extent of modelled ice cover during MIS~2
-        (29--14\,ka) using temperature time-series scaling factors
-        (Table~\ref{tab:records}) adjusted to obtain model results in agreement
-        with the Last Glacial Maximum (LGM) geomorphological reconstruction
-        \citep[solid red line,][]{Ehlers.etal.2011}.
-        \textbf{(d--f)} Cumulative extent of modelled ice cover during MIS~4
-        (71--57\,ka). Only the simulation driven by the EPICA temperature
-        time-series yields reasonable MIS~4 ice cover.}
-      \label{fig:footprints}
-    \end{figure}
-
 % ======================================================================

### 2017-07-04 articles/cycle/alps-cycle.tex
Update notes and todos to reflect new plans.

@@
 \begin{abstract}
-
-    \note{For the Nature Geoscience format this abstract will need to be
-          converted to an ``introductory paragraph''.}
 
@@
           (MIS~2). A different assumption could be that this outline represents
-          the most extensive glaciation in the last 120\,ka. I we allow the
+          the most extensive glaciation in the last 120\,ka. If we allow the
           MIS~2 glaciation to be a bit smaller than MIS~4, simulations driven
@@
     \idea{Matthias: The selection of figures in the main text and SOM is good.
-          I find Figure 2 quite difficult to understand at first glance. I
-          would combine it (already before submission) with Fig. 3. the Figure
+          I find Figure 5 quite difficult to understand at first glance. I
+          would combine it (already before submission) with Fig. 6. the Figure
           can then focus both on the temporal advance/retreat patterns of the 4
@@
           elevation again in the Supplementary.}
+
+    \idea{Add trimline figure in the main text now.}
 
@@
         ice-bedrock interface.
-        \idea{Since this will be in the supplement, I can also add additional
-              information such as winter temperature, summer precipitation,
-              winter PDD SD, and/or basal topography with modern glaciers
-              removed, e.g. in the Bernese Alps.}}
+        \todo{Fix geothermal plot.}}
       \label{fig:inputs}
@@
         MIS~4 according to a~global compilation of benthic \chem{\delta^{18}O}
-        records \citep{Lisiecki.Raymo.2005}.}
+        records \citep{Lisiecki.Raymo.2005}.
+        \todo{Add temporary subfigure labels. Refine temperature offsets.}}
       \label{fig:timeseries}
@@
         (71--57\,ka). Only the simulation driven by the EPICA temperature
-        time-series yields reasonable MIS~4 ice cover.}
+        time-series yields reasonable MIS~4 ice cover.
+        \todo{Refine temperature offsets. Try to make reduced precipitation
+              outlines more visible.}}
       \label{fig:footprints}
@@
         a~global compilation of benthic \chem{\delta^{18}O} records
-        \citep{Lisiecki.Raymo.2005}.}
+        \citep{Lisiecki.Raymo.2005}.
+        \note{I expect the simulation to complete in late July.}}
       \label{fig:lgmvel}
@@
         is here modelled as a time-transgressive event.
-        \note{The Nature Geoscience format is limited to four figures. Now we
-              indeed have four figures, albeit large ones.
-              Figure~\ref{fig:timing} will interest the dating people but is
-              not crucial for the paper. My plan is to include in the main text
-              for the first submission and be prepared that the editor may want
-              to move it into supplementary material.}}
+        \note{I expect the simulation to complete in late July.}}
       \label{fig:timing}
@@
               profiles drawn on panel a. But then we should probably limit it
-              to two glaciers instead of four.}}
+              to two glaciers instead of four.}
+        \note{I expect the simulation to complete in late July.}}
       \label{fig:profiles}
@@
         $l = 2.02$ and $K_g = 2.7\e{-7}\,m^{1-l}\,a^{l-1}$
-        \citep{Herman.etal.2015}.}
+        \citep{Herman.etal.2015}.
+        \note{I expect the simulation to complete in late July.}}
       \label{fig:erosion}
@@
         Parameter values used in the ice sheet model.
-        \idea{If I need to re-run all simulations, we could rethink bedrock,
-              lithosphere and PDD parameters.}}
+        \todo{Update parameters to new configuration.}}
       \label{tab:params}
@@
         to~22\,\unit{ka} after scaling.
-        \note{The EPICA scaling factor is close to one, and that ended up
-              yielding too much ice in the high-resolution simulation
-              (Fig.~\ref{fig:lgmvel}), though not so at low resotion
-              (Fig.~\ref{fig:footprints}b). If I re-run the whole thing, I will
-              probably fix it to one (i.e. use EPICA temperature anomalies
-              without applying any scaling).}}
+        \todo{Correct and refine temperature offsets. Add reduced precipitation
+              runs.}}
       \label{tab:records}

### 2017-07-04 figures/alpcyc_lr_footprints.py
Use 2 km resolution in lores plots (preliminary).
Temperature offsets are preliminary. Simulations in progress.

@@ -15,7 +15,7 @@ fig, grid = ut.pl.subplots_mm(figsize=(170.0, 80.0), projection=ut.pl.utm,
                               hspace=2.5, wspace=2.5)
 
 # add boot topo  # FIXME move to util
-nc = ut.io.load('input/boot/alps-srtm+thk+gou11simi-5km.nc')
+nc = ut.io.load('input/boot/alps-srtm+thk+gou11simi-2km.nc')
 for ax in grid.flat:
     ax.set_extent(ut.pl.regions['alps'], crs=ax.projection)
     ax.set_rasterization_zorder(2.5)
@@ -37,7 +37,7 @@ for i, rec in enumerate(ut.alpcyc_records):
 
     # load extra output
     dtfile = '%s3222cool%04d' % (rec.replace('-', '').lower(), round(dt*100))
-    nc = ut.io.load('output/e9d2d1f/alps-wcnn-5km/%s+%s/'
+    nc = ut.io.load('output/e9d2d1f/alps-wcnn-2km/%s+%s/'
                     'y???????-extra.nc' % (dtfile, conf))
     x = nc.variables['x'][:]
     y = nc.variables['y'][:]

### 2017-07-04 figures/alpcyc_lr_stackprints.py

@@ -42,7 +42,7 @@ for i, rec in enumerate(ut.alpcyc_records):
 
             # load extra file
             dtfile = '%s3222cool%04d' % (rec.replace('-', '').lower(), round(dt*100))
-            nc = ut.io.load('output/e9d2d1f/alps-wcnn-5km/%s+%s/'
+            nc = ut.io.load('output/e9d2d1f/alps-wcnn-2km/%s+%s/'
                             'y???????-extra.nc' % (dtfile, conf))
             x = nc.variables['x'][:]
             y = nc.variables['y'][:]

### 2017-07-04 figures/alpcyc_lr_timeseries.py

@@ -25,7 +25,7 @@ for i, rec in enumerate(ut.alpcyc_records):
     ax1.plot(age, dt, c=c, alpha=0.75)
 
     # load output time series
-    nc = ut.io.load('output/e9d2d1f/alps-wcnn-5km/%s+%s/'
+    nc = ut.io.load('output/e9d2d1f/alps-wcnn-2km/%s+%s/'
                     'y???????-ts.nc' % (dtfile, conf))
     age = -nc.variables['time'][:]/(1e3*365*24*60*60)
     vol = nc.variables['slvol'][:]

### 2017-07-04 articles/cycle/alps-cycle.tex
Add reduced precip runs in Table 2 (preliminary).
Temperature offsets need to be refined.

@@
         records \citep{Lisiecki.Raymo.2005}.
-        \todo{Add temporary subfigure labels. Refine temperature offsets.}}
+        \todo{Refine temperature offsets.}}
       \label{fig:timeseries}
@@
         to~22\,\unit{ka} after scaling.
-        \todo{Correct and refine temperature offsets. Add reduced precipitation
-              runs.}}
+        \todo{Refine temperature offsets.}
+        \idea{Move scaling factors in separate table or transpose.}}
       \label{tab:records}
@@
                   & \chem{\delta^{18}O}
-                  & 0.46 & $-$7.6  % -16.4126 (before scaling)
+                  & 0.50 & $-$8.2  % -16.4126 (before scaling)
                   & \citet{Dansgaard.etal.1993} \\
+
+        GRIP, $\Delta P$ & " & " & " & "
+                  & 0.63 & $-$10.4  % -16.4126 (before scaling)
+                  & " \\
 
@@
                   & \chem{\delta^{18}O}
-                  & 1.03 & $-$9.5  % -9.2055
+                  & 1.05 & $-$9.7  % -9.2055
                   & \citet{Jouzel.etal.2007} \\
+
+        EPICA, $\Delta P$ & " & " & " & "
+                  & 1.33 & $-$12.2  % -9.2055
+                  & " \\
 
@@
                   & \chem{U^{K'}_{37}}
-                  & 1.84 & $-$8.0  % -4.345625
+                  & 1.82 & $-$7.9  % -4.345625
                   & \citet{Martrat.etal.2007} \\
+
+        MD01-2444, $\Delta P$ & " & " & " & "
+                  & 2.46 & $-$10.7  % -4.345625
+                  & " \\
 

### 2017-07-04 articles/cycle/alps-cycle.tex
Try to improve the record table using multirow.

@@
 \usepackage{booktabs}
+\usepackage{multirow}
 \usepackage[T1]{fontenc}
@@
 
-        GRIP      &  72{\degree}35$^{\prime}$\,N   % 72.58 (decimal)
-                  &  37{\degree}38$^{\prime}$\,W   % 37.64 (decimal)
-                  & 3238
-                  & \chem{\delta^{18}O}
+        GRIP      & \multirow{2}{*}{ 72{\degree}35$^{\prime}$\,N}   % 72.58 (decimal)
+                  & \multirow{2}{*}{ 37{\degree}38$^{\prime}$\,W}   % 37.64 (decimal)
+                  & \multirow{2}{*}{3238}
+                  & \multirow{2}{*}{\chem{\delta^{18}O}}
                   & 0.50 & $-$8.2  % -16.4126 (before scaling)
-                  & \citet{Dansgaard.etal.1993} \\
+                  & \multirow{2}{*}{\citet{Dansgaard.etal.1993}} \\
 
-        GRIP, $\Delta P$ & " & " & " & "
-                  & 0.63 & $-$10.4  % -16.4126 (before scaling)
-                  & " \\
+        GRIP, $\Delta P$ &&&&& 0.63 & $-$10.4 \\
 
-        EPICA     &  75{\degree}06$^{\prime}$\,S   % 75.1
-                  & 123{\degree}21$^{\prime}$\,E   % 123.35
-                  & 3233
-                  & \chem{\delta^{18}O}
+        EPICA     & \multirow{2}{*}{ 75{\degree}06$^{\prime}$\,S}   % 75.1
+                  & \multirow{2}{*}{123{\degree}21$^{\prime}$\,E}   % 123.35
+                  & \multirow{2}{*}{3233}
+                  & \multirow{2}{*}{\chem{\delta^{18}O}}
                   & 1.05 & $-$9.7  % -9.2055
-                  & \citet{Jouzel.etal.2007} \\
+                  & \multirow{2}{*}{\citet{Jouzel.etal.2007}} \\
 
-        EPICA, $\Delta P$ & " & " & " & "
-                  & 1.33 & $-$12.2  % -9.2055
-                  & " \\
+        EPICA, $\Delta P$ &&&&& 1.33 & $-$12.2 \\
 
-        MD01-2444 &  37{\degree}34$^{\prime}$\,N   % 37.561
-                  &  10{\degree}04$^{\prime}$\,W   % -10.142
-                  & $-$2637
-                  & \chem{U^{K'}_{37}}
+        MD01-2444 & \multirow{2}{*}{ 37{\degree}34$^{\prime}$\,N}   % 37.561
+                  & \multirow{2}{*}{ 10{\degree}04$^{\prime}$\,W}   % -10.142
+                  & \multirow{2}{*}{$-$2637}
+                  & \multirow{2}{*}{\chem{U^{K'}_{37}}}
                   & 1.82 & $-$7.9  % -4.345625
-                  & \citet{Martrat.etal.2007} \\
+                  & \multirow{2}{*}{\citet{Martrat.etal.2007}} \\
 
-        MD01-2444, $\Delta P$ & " & " & " & "
-                  & 2.46 & $-$10.7  % -4.345625
-                  & " \\
+        MD01-2444, $\Delta P$ &&&&& 2.46 & $-$10.7 \\
 
@@
       \end{tabular}}
+
     \end{table*}

### 2017-07-06 figures/alpcyc_hr_inputs.py
Fix geothermal input plot.

@@ -26,7 +26,7 @@ cax = cgrid[3]
 levs = range(55, 96, 5)
 cmap = ut.pl.get_cmap('PuOr_r', len(levs)-1)
 cols = cmap(range(len(levs)-1))
-cs = nc.contourf('bheatflx', ax, levels=levs, colors=cols, alpha=0.75)
+cs = nc.contourf('bheatflx', ax, levels=levs, colors=cols, thkth=-1, alpha=0.75)
 cb = fig.colorbar(cs, cax, orientation='vertical', ticks=levs[1::2])
 cb.set_label(r'Geothermal flux ($mW\,m^{-2}$)')
 nc.close()

### 2017-07-06 articles/cycle/alps-cycle.tex
Test transposed record table, leave as comment.
Pull in methods from Cordillera cycle paper.
This will need a serious clean-up.

@@
 
+\usepackage{doi}
 \usepackage{amsmath}
@@
 
+\section{Model setup}
+\label{sec:model}
+
+\subsection{Overview}
+\label{sec:overview}%
+
+      The simulations presented here were run using the Parallel Ice Sheet
+      Model (PISM, version~0.7.2), an open source, finite
+      difference, shallow ice sheet model \citep{PISM-authors.2015}. The
+      model requires input on basal topography, sea level, geothermal heat
+      flux and climate forcing. It computes the evolution of ice extent and
+      thickness over time, the thermal and dynamic states of the ice sheet,
+      and the associated lithospheric response.
+
+      Basal topography is bilinearly interpolated from the ETOPO1 combined
+      topography and bathymetry dataset with a~resolution of 1\,arc-min
+      \citep{Amante.Eakins.2009} to the model grids. It responds to ice load
+      following a~bedrock deformation model that includes local isostasy,
+      elastic lithosphere flexure and viscous astenosphere deformation in an
+      infinite half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
+      A~relatively low viscosity value of $\nu_{\mathrm{m}} =
+      1\times10^{19}$\,\unit{Pa\,s} is used for the astenosphere
+      (Table~\ref{tab:params}) in accordance with the results from regional
+      glacial isostatic adjustment modelling at the northern Cascadia
+      subduction zone \citep{James.etal.2009}. Sea level is lowered as
+      a~function of time based on the Spectral Mapping Project
+      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale.
+
+      Ice deformation
+      follows temperature and water-content dependent creep
+      (Sect.~\ref{sec:icedyn}). Default parameter values are given in
+      Table~\ref{tab:params}. Basal sliding follows
+      a~pseudo-plastic law where the yield stress accounts for till
+      deconsolidation under high water pressure (Sect.~\ref{sec:sliding}).
+      Ice shelf calving is simulated using an ice thickness threshold and
+      principle components of the strain rate tensor
+      (Sect.~\ref{sec:calving}). Surface mass balance is
+      computed using a~positive degree-day (PDD) model
+      (Sect.~\ref{sec:surface}). Climate forcing is provided by a~monthly
+      climatology averaged from 1979 to 2000 from the North American
+      Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, perturbed
+      by time-dependent offsets and lapse-rate temperature corrections
+      (Sect.~\ref{sec:atm}, Table~\ref{tab:records}). A~sensitivity study
+      to some of the most influential ice rheology (Sect~\ref{sec:icedyn})
+      and basal sliding (Sect.~\ref{sec:sliding}) parameters is performed
+      (Table~\ref{tab:sens_params}).
+
+      Each simulation starts from assumed ice-free conditions at
+      120\,\unit{ka}, and runs to the present.
+      Our modelling domain of 1\,500 by 3\,000\,\unit{km} encompasses
+      the entire area covered by the Cordilleran ice sheet at the LGM
+      (Fig.~\ref{fig:locmap}). The simulations were run on two distinct
+      grids, using a~lower horizontal resolution of 10\,\unit{km}, and
+      a~higher horizontal resolution of 5\,\unit{km}.
+
+\subsection{Ice rheology}
+\label{sec:icedyn}
+
+      Ice sheet dynamics are typically modelled using a~combination of
+      internal deformation and basal sliding. PISM is a~shallow ice sheet
+      model, which implies that the balance of stresses is approximated
+      based on their predominant components. The Shallow Shelf Approximation
+      (SSA) is combined with the Shallow Ice Approximation
+      (SIA) by adding velocity solutions of the two approximations
+      \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}. Although this
+      heuristic approach implies errors in the transition zone where gravitational
+      stresses intervene both in the SIA and SSA velocity computation, this
+      hybrid scheme is computationally much more efficient than a fully
+      three-dimensional model with which the simulations presented here
+      would not be feasible.
+
+      Ice deformation is governed by the constitutive law for ice
+      \citep{Glen.1952, Nye.1953},
+%
+\begin{align}
+&\label{eqn:glenslaw}
+&\vec{\dot{\epsilon}} = A\,\tau_{\mathrm{e}}^{n-1}\,\vec{\tau} \,.
+\end{align}
+%
+      where $\vec{\dot{\epsilon}}$ is the strain-rate tensor,
+      $\vec{\tau}$ the deviatoric stress tensor, and $\tau_{\mathrm{e}}$ the
+      effective stress defined in our case by ${\tau_{\mathrm{e}}}^2 = \frac{1}{2}
+      \mathrm{tr}(\vec{\tau}^2)$. The ice softness coefficient, $A$, depends
+      on ice temperature, $T$, pressure, $p$, and water content, $\omega$,
+      through a~piece-wise Arrhenius-type law,
+%
+\begin{align}
+&\label{eqn:softness}
+&A = E\cdot
+\begin{cases}
+A_{\mathrm{c}} \,e^\frac{-Q_{\mathrm{c}}}{RT_{\text{pa}}}            & \text{if}\ T_{\text{pa}}  <  T_{\mathrm{c}} \,, \\
+A_{\mathrm{w}} (1+f\omega)\,e^\frac{-Q_{\mathrm{w}}}{RT_{\text{pa}}} & \text{if}\ T_{\text{pa}} \ge T_{\mathrm{c}} \,,
+\end{cases}
+\end{align}
+%
+      where $T_{\text{pa}}$ is the pressure-adjusted ice temperature
+      calculated using the Clapeyron relation, ${T_{\text{pa}} = T - \beta
+      p}$. $R=8.31441$\,\unit{J\,mol^{-1}\,K^{-1}} is the ideal gas
+      constant, and $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
+      $Q_{\mathrm{w}}$, are constant parameters corresponding to values
+      measured below and above a~critical temperature threshold
+      $T_{\mathrm{c}}=-10$\,\unit{{\degree}C}
+      \citep[p.~72]{Paterson.Budd.1982,Cuffey.Paterson.2010}. The water
+      fraction, $\omega$, is capped at a~maximum value of 0.01, above which
+      no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
+      Greve.1997}. Finally, $E$ is a~non-dimensional enhancement factor
+      which can take different values, $E_{\text{SIA}}$, in the SIA and
+      $E_{\text{SSA}}$, in the SSA.
+
+      In all our simulations, we set constant the power-law exponent, $n=3$,
+      according to \citet[p.~55--57]{Cuffey.Paterson.2010}, the Clapeyron
+      constant, $\beta=7.9\times 10^{-8}$\,\unit{K\,Pa^{-1}}, according to
+      \citet{Luthi.etal.2002}, the water fraction coefficient, $f=181.25$,
+      according to \citet{Lliboutry.Duval.1985}, and the SSA enhancement
+      factor, $E_{\text{SSA}}=1$, according to
+      \citet[p.~77]{Cuffey.Paterson.2010}. These fixed parameter values are
+      summarized in Table~\ref{tab:params}.
+
+      On the other hand, we test the model sensitivity
+      (Sect.~\ref{sec:sens}) to different values for the two creep
+      parameters, $A_{\mathrm{c}}$ and $A_{\mathrm{w}}$, the two activation
+      energies, $Q_{\mathrm{c}}$ and $Q_{\mathrm{w}}$, and the SIA
+      enhancement factor, $E_{\text{SIA}}$, as follows.
+%
+\begin{itemize}
+  \item Our \emph{default} configuration used in the control run of the
+    sensitivity study and all other simulations in the paper includes
+    rheological parameters,
+    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
+    $Q_{\mathrm{w}}$, derived from \citet{Paterson.Budd.1982} and given in
+    \citet[Eqn.~5]{Bueler.Brown.2009}, and $E_{\text{SIA}}=1$.
+  \item Our \emph{hard ice} configuration includes rheological parameters,
+    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
+    $Q_{\mathrm{w}}$, derived from \citet[p.~72 and
+    76]{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=1$, which correspond to
+    a~stiffer rheology than that used in the control run.
+  \item Our \emph{soft ice} configuration includes rheological parameters
+    from \citet{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=5$, the
+    recommended value for ice age polar ice
+    \citep[p.~77]{Cuffey.Paterson.2010}.
+\end{itemize}
+%
+      An additional simulation using the ice rheology from
+      \citet{Cuffey.Paterson.2010} and $E_{\text{SIA}}=2$, the recommended
+      value for Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010} has
+      been performed, but is not presented here, since its results were very
+      similar to that of our default run.
+
+      Actual parameter values for $A_{\mathrm{c}}$, $A_{\mathrm{w}}$,
+      $Q_{\mathrm{c}}$, $Q_{\mathrm{w}}$ and $E_{\text{SIA}}$ used in our
+      simulations are given in Table~\ref{tab:sens_params}, while the effect
+      of the three different parametrisations on temperature-dependent ice
+      softness, $A$, is illustrated in Fig.~\ref{fig:sens_plot_rheo}.
+
+      Surface air
+      temperature derived from the climate forcing (Sect.~\ref{sec:atm})
+      provides the upper boundary condition to the ice enthalpy
+      model. Temperature is computed in the ice and in the bedrock to
+      a~depth of 3\,\unit{km} below the ice-bedrock interface, where it is
+      conditioned by a~lower boundary geothermal heat flux of
+      $q_{\mathrm{G}}=70$\,\unit{mW\,m^{-2}}. Although this uniform value
+      does
+      not account for the high spatial geothermal variability in the region
+      \citep{Blackwell.Richards.2004}, it is, on average, representative of
+      available heat flow measurements. In the low-resolution simulations,
+      the vertical grid consists of 31~temperature layers in the bedrock and
+      up to 51~enthalpy layers in the ice sheet, corresponding to a~vertical
+      resolution of 100\,\unit{m}. The high-resolution simulations use
+      61~bedrock layers and up to 101~ice layers with a~vertical resolution
+      of 50\,\unit{m}.
+
+\subsection{Basal sliding}
+\label{sec:sliding}
+
+      A~pseudo-plastic sliding law,
+%
+\begin{align}
+&\label{eqn:pseudoplastic}
+    \vec{\tau}_{\mathrm{b}} = -\tau_{\mathrm{c}} \frac{\vec{v}_{\mathrm{b}}}{{v_{\text{th}}}^q\,|\vec{v}_{\mathrm{b}}|^{1-q}} \,,
+\end{align}
+%
+      relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to
+      the sliding velocity, $\vec{v}_{\mathrm{b}}$.
+      The yield stress, $\tau_{\mathrm{c}}$, is modelled using the
+      Mohr--Coulomb criterion,
+%
+\begin{align}
+&\tau_{\mathrm{c}} = c_0 + N\,\tan{\phi} \,,
+\end{align}
+%
+      where cohesion, $c_0$, is assumed to be zero. Effective pressure, $N$,
+      is related to the ice overburden stress, $P_0=\rho gh$, and the
+      modelled amount of
+      subglacial water, using a~formula derived from laboratory experiments
+      with till extracted from the base of Ice Stream B in West Antarctica
+      \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
+%
+\begin{align}
+&\label{eqn:ntil}
+&N = \delta P_0 \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,,
+\end{align}
+%
+      where $\delta$ sets the minimum ratio between the effective and
+      overburden pressures, $e_0$ is a~reference void ratio and
+      $C_{\mathrm{c}}$ is the till compressibility coefficient
+      \citep{Tulaczyk.etal.2000}. The amount of water at the base, $W$,
+      varies from zero to $W_{\text{max}}$, a~threshold above which
+      additional melt water is assumed to drain off instantaneously.
+
+      In all our simulations, we set constant the pseudo-plastic sliding
+      exponent, $q=0.25$, and the threshold velocity,
+      $v_{\text{th}}=100$\,\unit{m\,a^{-1}}, according to values used by
+      \citet{Aschwanden.etal.2013}, the till cohesion, $c_0=0$, whose
+      measured values are consistently negligible
+      \citep[p.~268]{Tulaczyk.etal.2000, Cuffey.Paterson.2010}, the till
+      reference void ratio, $e_0=0.69$, and the till compressibility
+      coefficient, $C_{\mathrm{c}}=0.12$, according to the only measurements
+      available to our knowledge, published by \citet{Tulaczyk.etal.2000}.
+      These fixed parameter values are summarized in Table~\ref{tab:params}.
+
+      We also use a~constant spatial distribution of the till friction
+      angle, $\phi$, whose values vary from 15 to 45\unit{\degree} as
+      a~piecewise-linear function of modern bed elevation, with the lowest
+      value occuring below the modern sea level (0\,\unit{m} above sea level,
+      \unit{m\,a.s.l.}) and the highest value occuring above the generalised
+      elevation of the highest shorelines
+      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}. This range of
+      values span over the range of measured values for glacial till of 18
+      to 40\unit{\degree} \citep[p.~268]{Cuffey.Paterson.2010}. It accounts
+      for frictional basal conditions associated with discontinuous till
+      cover at high elevations, and for a~weakening of till associated with
+      the presence of marine sediments (cf. \citealp{Martin.etal.2011};
+      \citealp[Supplement]{Aschwanden.etal.2013};
+      \citealp{PISM-authors.2015}).
+
+      An additional simulation with a~constant till friction angle,
+      $\phi=30$\unit{\degree}, corresponding to the average value in
+      \citet[p.~268]{Cuffey.Paterson.2010}, has been performed, but is not
+      presented here, since the induced variability was small.
+
+      On the other hand, we test the model sensitivity
+      (Sect.~\ref{sec:sens}) to different values for the minimum ratio
+      between the effective and overburden pressures, $\delta$, and the
+      maximum water height in the till, $W_{\text{max}}$, as follows.
+%
+\begin{itemize}
+  \item Our \emph{default} configuration used in the control run of the
+    sensitivity study and all other simulations in the paper includes
+    $\delta=0.02$ and $W_{\text{max}}=2$\,\unit{m} as in
+    \citet{Bueler.Pelt.2015}.
+  \item Our \emph{soft bed} configuration use $\delta=0.01$ and
+    $W_{\text{max}}=1$\,\unit{m}.
+  \item Our \emph{hard bed} configuration use $\delta=0.05$ and
+    $W_{\text{max}}=5$\,\unit{m}.
+\end{itemize}
+%
+      The effect of the three different parametrisations on the effective
+      pressure on the till, $N$, in response to water content, $W$, is
+      illustrated in Fig.~\ref{fig:sens_plot_ntil}. All parameter choices
+      are listed in Table~\ref{tab:sens_params}.
+
+\subsection{Ice shelf calving}
+\label{sec:calving}
+
+      Ice shelf calving is computed using a~double criterion. First,
+      a~physically-realistic calving flux is computed based on eigenvalues
+      of the horizontal strain rate tensor \citep{Winkelmann.etal.2011,
+      Levermann.etal.2012}. This allows floating ice to advance in confined
+      embayments, but prevents the formation of extensive ice shelves in the
+      open ocean. Second, floating ice thinner than 50\,\unit{m} is
+      systematically calved off. A~subgrid scheme by
+      \citet{Albrecht.etal.2011} allows for a~continuous migration of the
+      calving front. This formulation of calving has been applied to the
+      Antarctic ice sheet and has shown to produce a~realistic calving front
+      positon for many of the present-day ice shelves
+      \citep{Martin.etal.2011}.
+
+\subsection{Surface mass balance}
+\label{sec:surface}
+
+      Ice surface accumulation and ablation are computed from monthly mean
+      near-surface air temperature, $T_{\mathrm{m}}$, monthly standard
+      deviation of near-surface air temperature, $\sigma$, and monthly
+      precipitation, $P_{\mathrm{m}}$, using a~temperature-index model
+      \citep[e.g.,][]{Hock.2003}. Accumulation is equal to precipitation
+      when air temperatures are below 0\,\unit{{\degree}C}, and decreases to
+      zero linearly with temperatures between 0 and 2\,\unit{{\degree}C}.
+      Ablation is computed from PDD, defined as an integral of temperatures
+      above 0\,\unit{{\degree}C} in one year.
+
+      The PDD computation accounts for stochastic temperature variations by
+      assuming a~normal temperature distribution of standard deviation
+      $\sigma$ around the expected value $T_{\mathrm{m}}$. It is expressed
+      by an error-function formulation \citep{Calov.Greve.2005},
+\begin{align}
+&\label{eqn:calovgreve}
+    {\text{PDD}} = \int_{t_1}^{t_2} \mathrm{d}t
+        \left[\frac{\sigma}{\sqrt{2\pi}}
+                \exp\left({-\frac{T_{\mathrm{m}}^2}{2\sigma^2}}\right)
+              + \frac{T_{\mathrm{m}}}{2} \, {\text{erfc}}
+                \left(-\frac{T_{\mathrm{m}}}{\sqrt{2}\sigma}\right)\right] \,,
+\end{align}
+      which is numerically approximated using week-long sub-intervals. In
+      order to account for the effects of spatial and seasonal variations of
+      temperature variability \citep{Seguinot.2013}, $\sigma$ is computed
+      from NARR daily temperature values from 1979 to 2000
+      \citep{Mesinger.etal.2006}, including variability associated with the
+      seasonal cycle, and bilinearly interpolated to the model grids
+      (Fig.~\ref{fig:atm}). Degree-day factors for snow and
+      ice melt are derived from mass-balance measurements on contemporary
+      glaciers from the Coast Mountains and Rocky Mountains in British
+      Columbia \citep[Table~\ref{tab:params};][]{Shea.etal.2009}.
+
+\subsection{Climate forcing}
+\label{sec:atm}%
+
+      Climate forcing driving ice sheet simulations consists of
+      a~present-day monthly climatology, $\{T_{\mathrm{m}0},
+      P_{\mathrm{m}0}\}$, where temperatures are modified by offset time
+      series, ${\Delta}T_{\text{TS}}$, and lapse-rate corrections,
+      ${\Delta}T_{\text{LR}}$:
+\begin{align}
+&T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) + {\Delta}T_{\text{TS}}(t)
+                    + {\Delta}T_{\text{LR}}(t, x, y) \,, \\
+&    P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \,.
+\end{align}
+      The present-day monthly climatology was bilinearly interpolated from
+      near-surface air
+      temperature and precipitation rate fields from the NARR, averaged from
+      1979 to 2000. Modern climate of the North American Cordillera is
+      characterised by strong geographic variations in temperature
+      seasonality, timing of the maximum annual precipitation, and daily
+      temperature variability (Fig.~\ref{fig:atm}). Although the ability
+      of the NARR to reproduce the steep climatic gradients is limited by its
+      spatial resolution of 32\,\unit{km} \citep{Jarosch.etal.2012}, it has
+      been tested against observational data in our previous sensitivity
+      study and identified as yielding a closer fit between the modelled LGM
+      extent of the Cordilleran ice sheet and the geological evidence than
+      other atmospheric reanalyses \citep{Seguinot.etal.2014}.
+
+      Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived
+      from palaeo-temperature proxy records from the Greenland Ice Core
+      Project \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice
+      Core Project \citep[NGRIP,][]{Andersen.etal.2004}, the European
+      Project for Ice Coring in Antarctica \citep[EPICA,][]
+      {Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and
+      Ocean Drilling Program (ODP) sites 1012 and 1020, both located off the
+      coast of California \citep{Herbert.etal.2001}. Palaeo-temperature
+      anomalies from the GRIP and NGRIP records were calculated from oxygen
+      isotope (\chem{\delta^{18}O}) measurements using a~quadratic equation
+      \citep{Johnsen.etal.1995},
+\begin{align}
+{\Delta}T_{\text{TS}}(t) ={~}&-11.88 [\chem{\delta^{18}O}(t)
+                                -\chem{\delta^{18}O}(0)] \nonumber \\
+                        &-0.1925[\chem{\delta^{18}O}(t)^2
+                                 -\chem{\delta^{18}O}(0)^2] \,,
+\end{align}
+      while temperature reconstructions from Antarctic and oceanic cores
+      were provided as such. For each proxy record used and each of the
+      parameter setup used in the sensitivity tests, palaeo-temperature
+      anomalies were scaled linearly (Tables~\ref{tab:records}
+      and~\ref{tab:sens_params}) in order to simulate comparable ice extents
+      at the LGM (Table~\ref{tab:extrema}) and realistic outlines
+      (Fig.~\ref{fig:lr_maps}).
+
+      Finally, lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are computed
+      as a~function of ice surface elevation, $s$, using the NARR surface
+      geopotential height invariant field as a~reference topography,
+      $b_{\text{ref}}$:
+\begin{align}
+{\Delta}T_{\text{LR}}(t, x, y) &= -\gamma [s(t, x, y)-b_{\text{ref}}] \\
+                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{\text{ref}}],
+\end{align}
+      thus accounting for the evolution of ice thickness, ${h=s-b}$, on the
+      one hand, and for differences between the basal topography of the ice
+      flow model, $b$, and the NARR reference topography, $b_{\text{ref}}$,
+      on the other hand. All simulations use an annual temperature lapse
+      rate of $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper, we
+      refer to different model runs by the name of the proxy record used for
+      the palaeo-temperature forcing.
+
 % ----------------------------------------------------------------------
@@
         boundary condition to the bedrock thermal model 3\,km below the
-        ice-bedrock interface.
-        \todo{Fix geothermal plot.}}
+        ice-bedrock interface.}
       \label{fig:inputs}
@@
         to~22\,\unit{ka} after scaling.
-        \todo{Refine temperature offsets.}
-        \idea{Move scaling factors in separate table or transpose.}}
+        \todo{Refine temperature offsets.}}
       \label{tab:records}
@@
 
-        Record    & Latitude & Longitude & Elev. (m~a.s.l.)
+        Forcing   & Latitude & Longitude & Elev. (m~a.s.l.)
                   & Proxy & $f$ & $[{\Delta}\text{TS}]_{32}^{22}$ (K)
@@
 
+      %\noindent\small\makebox[\textwidth]
+      %{\begin{tabular}{llrlrlr}
+      %  \toprule
+      %
+      %  Record    & \multicolumn{2}{c}{GRIP}
+      %            & \multicolumn{2}{c}{EPICA}
+      %            & \multicolumn{2}{c}{MD01-2444} \\
+      %
+      %  Precip.   & cst. & $\Delta P$
+      %            & cst. & $\Delta P$
+      %            & cst. & $\Delta P$ \\
+      %
+      %  \midrule
+      %
+      %  Latitude  & \multicolumn{2}{c}{ 72{\degree}35$^{\prime}$\,N}   % 72.58 (decimal)
+      %            & \multicolumn{2}{c}{ 75{\degree}06$^{\prime}$\,S}   % 75.1
+      %            & \multicolumn{2}{c}{ 37{\degree}34$^{\prime}$\,N} \\  % 37.561
+      %
+      %  Longitude & \multicolumn{2}{c}{ 37{\degree}38$^{\prime}$\,W}   % 37.64 (decimal)
+      %            & \multicolumn{2}{c}{123{\degree}21$^{\prime}$\,E}   % 123.35
+      %            & \multicolumn{2}{c}{ 10{\degree}04$^{\prime}$\,W} \\  % -10.142
+      %
+      %  Elevation & \multicolumn{2}{c}{3238\,m~a.s.l.}
+      %            & \multicolumn{2}{c}{3233\,m~a.s.l.}
+      %            & \multicolumn{2}{c}{$-$2637\,m~a.s.l.} \\
+      %
+      %  Proxy     & \multicolumn{2}{c}{\chem{\delta^{18}O}}
+      %            & \multicolumn{2}{c}{\chem{\delta^{18}O}}
+      %            & \multicolumn{2}{c}{\chem{U^{K'}_{37}}} \\
+      %
+      %  $f$       & 0.50 & 0.63
+      %            & 1.05 & 1.33
+      %            & 1.82 & 2.46 \\
+      %
+      %  $[{\Delta}\text{TS}]_{32}^{22}$
+      %            & $-$8.2\,K & $-$10.4\,K     % -16.4126 (before scaling)
+      %            & $-$9.7\,K & $-$12.2\,K     % -9.2055
+      %            & $-$7.9\,K & $-$10.7\,K \\  % -4.345625
+      %
+      %  Reference & \multicolumn{2}{c}{\citet{Dansgaard.etal.1993}}
+      %            & \multicolumn{2}{c}{\citet{Jouzel.etal.2007}}
+      %            & \multicolumn{2}{c}{\citet{Martrat.etal.2007}} \\
+      %
+      %  \bottomrule
+      %\end{tabular}}
+
     \end{table*}

### 2017-07-06 articles/cycle/alps-cycle.tex
Remove calving methods.

@@
       deconsolidation under high water pressure (Sect.~\ref{sec:sliding}).
-      Ice shelf calving is simulated using an ice thickness threshold and
-      principle components of the strain rate tensor
-      (Sect.~\ref{sec:calving}). Surface mass balance is
+      Surface mass balance is
       computed using a~positive degree-day (PDD) model
@@
       are listed in Table~\ref{tab:sens_params}.
-
-\subsection{Ice shelf calving}
-\label{sec:calving}
-
-      Ice shelf calving is computed using a~double criterion. First,
-      a~physically-realistic calving flux is computed based on eigenvalues
-      of the horizontal strain rate tensor \citep{Winkelmann.etal.2011,
-      Levermann.etal.2012}. This allows floating ice to advance in confined
-      embayments, but prevents the formation of extensive ice shelves in the
-      open ocean. Second, floating ice thinner than 50\,\unit{m} is
-      systematically calved off. A~subgrid scheme by
-      \citet{Albrecht.etal.2011} allows for a~continuous migration of the
-      calving front. This formulation of calving has been applied to the
-      Antarctic ice sheet and has shown to produce a~realistic calving front
-      positon for many of the present-day ice shelves
-      \citep{Martin.etal.2011}.
 

### 2017-07-06 articles/cycle/alps-cycle.tex
Update parameter table to alpcyc4 config.
Add subsection on basal topography.

@@
 
-\section{Model setup}
-\label{sec:model}
-
 \subsection{Overview}
@@
 
-      The simulations presented here were run using the Parallel Ice Sheet
-      Model (PISM, version~0.7.2), an open source, finite
-      difference, shallow ice sheet model \citep{PISM-authors.2015}. The
-      model requires input on basal topography, sea level, geothermal heat
+    We use the Parallel Ice Sheet
+    Model (PISM, development version~e9d2d1f), an open source, finite
+    difference, shallow ice sheet model \citep{PISM-authors.2017}. The
+    model requires input on basal topography, geothermal heat
     flux and climate forcing. It computes the evolution of ice extent and
@@
     and the associated lithospheric response.
-
-      Basal topography is bilinearly interpolated from the ETOPO1 combined
-      topography and bathymetry dataset with a~resolution of 1\,arc-min
-      \citep{Amante.Eakins.2009} to the model grids. It responds to ice load
-      following a~bedrock deformation model that includes local isostasy,
-      elastic lithosphere flexure and viscous astenosphere deformation in an
-      infinite half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.
-      A~relatively low viscosity value of $\nu_{\mathrm{m}} =
-      1\times10^{19}$\,\unit{Pa\,s} is used for the astenosphere
-      (Table~\ref{tab:params}) in accordance with the results from regional
-      glacial isostatic adjustment modelling at the northern Cascadia
-      subduction zone \citep{James.etal.2009}. Sea level is lowered as
-      a~function of time based on the Spectral Mapping Project
-      \citep[SPECMAP,][]{Imbrie.etal.1989} time scale.
 
@@
       are listed in Table~\ref{tab:sens_params}.
+
+\subsection{Basal topography}
+\label{sec:bedrock}
+
+    The initial basal topography is bilinearly interpolated from the
+    hole-filled, Shuttle Radar Topography Mission (SRTM) data with a~resolution
+    of 30\,arc-sec \citep{Jarvis.etal.2008}. These data include post-glacial
+    sediment fills and lakes surface topography. However, they were corrected
+    for an estimation of present-day ice thickness according from modern
+    glacier outlines, surface topography and simplified ice physics
+    \citep{Huss.Farinotti.2012}.
+
+    Basal topography responds to ice load following a~bedrock deformation model
+    that includes local isostasy, elastic lithosphere flexure and viscous
+    astenosphere deformation in an infinite half-space
+    \citep{Lingle.Clark.1985,Bueler.etal.2007}. The astenosphere viscosity,
+    $\nu_{\mathrm{m}}=2.2\times10^{20}$\,\unit{Pa\,s}, the astenosphere
+    density, $\rho_{\mathrm{m}}=3300$\,\unit{kg\,m^{-3}}, and the lithosphere
+    elastic rigidity, $D=1.389\e{24}$\,\unit{N\,m}, were set according to
+    results from glacial isostatic adjustment modelling of deglacial rebound in
+    the Alps most closely reproducing observed modern uplift rates
+    \citep[Table~\ref{tab:params};][Supplementary Fig.~7]{Mey.etal.2016}. The
+    latter was computed as $D=YE^3/12(1-\nu)$, where $Y=100$\,GPa is Young's
+    modulus, $\nu=0.25$ is the Poisson ratio, and $E=50$\,km is the average
+    effective elastic thickness of the lithosphere
+    \citep[Table~\ref{tab:params};][]{Mey.etal.2016}.
 
@@
         $E_{\text{SIA}}$   & SIA enhancement factor
-                & 5
+                & 2
                 & --
@@
         $\nu_{\mathrm{m}}$ & Astenosphere viscosity
-                & $1\e{19}$
+                & $2.2\e{20}$
                 & \unit{Pa\,s}
-                & \citet{James.etal.2009} \\
-
-        $\rho_{\mathrm{l}}$& Lithosphere density
+                & \citet{Mey.etal.2016} \\
+
+        $\rho_{\mathrm{m}}$& Astenosphere density
                 & 3300
                 & \unit{kg\,m^{-3}}
-                & \citet{Lingle.Clark.1985} \\
+                & \citet{Mey.etal.2016} \\
 
         $D$     & Lithosphere flexural rigidity
-                & $5\e{24}$
+                & $1.389\e{24}$
                 & \unit{N\,m}
-                & \citet{Lingle.Clark.1985} \\
+                & \citet{Mey.etal.2016} \\
 
@@
                 & -- \\
+
+        $\omega$& Precipitation factor
+                & 0.0704
+                & --
+                & \citet{Huybrechts.2002} \\
 

### 2017-07-07 articles/cycle/alps-cycle.tex
Update web references.
Rework methods overview.
I am not convinced this is useful.

@@
 
-    We use the Parallel Ice Sheet
-    Model (PISM, development version~e9d2d1f), an open source, finite
-    difference, shallow ice sheet model \citep{PISM-authors.2017}. The
-    model requires input on basal topography, geothermal heat
-    flux and climate forcing. It computes the evolution of ice extent and
-    thickness over time, the thermal and dynamic states of the ice sheet,
-    and the associated lithospheric response.
-
-      Ice deformation
-      follows temperature and water-content dependent creep
-      (Sect.~\ref{sec:icedyn}). Default parameter values are given in
-      Table~\ref{tab:params}. Basal sliding follows
-      a~pseudo-plastic law where the yield stress accounts for till
-      deconsolidation under high water pressure (Sect.~\ref{sec:sliding}).
-      Surface mass balance is
-      computed using a~positive degree-day (PDD) model
-      (Sect.~\ref{sec:surface}). Climate forcing is provided by a~monthly
-      climatology averaged from 1979 to 2000 from the North American
-      Regional Reanalysis \citep[NARR,][]{Mesinger.etal.2006}, perturbed
-      by time-dependent offsets and lapse-rate temperature corrections
-      (Sect.~\ref{sec:atm}, Table~\ref{tab:records}). A~sensitivity study
-      to some of the most influential ice rheology (Sect~\ref{sec:icedyn})
-      and basal sliding (Sect.~\ref{sec:sliding}) parameters is performed
-      (Table~\ref{tab:sens_params}).
-
-      Each simulation starts from assumed ice-free conditions at
-      120\,\unit{ka}, and runs to the present.
-      Our modelling domain of 1\,500 by 3\,000\,\unit{km} encompasses
-      the entire area covered by the Cordilleran ice sheet at the LGM
-      (Fig.~\ref{fig:locmap}). The simulations were run on two distinct
-      grids, using a~lower horizontal resolution of 10\,\unit{km}, and
-      a~higher horizontal resolution of 5\,\unit{km}.
+    We use the Parallel Ice Sheet Model (PISM, development version~e9d2d1f), an
+    open source, finite difference, shallow ice sheet model
+    \citep{PISM-authors.2017}. The model requires input on basal
+    topography, geothermal heat flux and climate forcing. It computes the
+    evolution of ice extent and thickness over time, the thermal and dynamic
+    states of the ice sheet, and the associated lithospheric response. The
+    set-up used here is largely based on that used in previous studies of the
+    former Cordilleran ice sheet \citep{Seguinot.2014, Seguinot.etal.2014,
+    Seguinot.etal.2016}.
+
+    \idea{Remove next paragraph?}
+
+    Ice deformation follows temperature and water-content dependent creep
+    (Sect.~\ref{sec:icedyn}). Basal sliding follows a~pseudo-plastic law where
+    the yield stress accounts for till deconsolidation under high water
+    pressure (Sect.~\ref{sec:sliding}). Bedrock topography is deflected
+    under the ice load (Sect.~\ref{sec:bedrock}). Surface mass balance is
+    computed using a~positive degree-day (PDD) model (Sect.~\ref{sec:surface}).
+    Climate forcing is provided by a~monthly climatology from interpolated
+    observational data \citep[WorldClim;][]{Hijmans.etal.2005} and the European
+    Centre for Medium-Range Weather Forecasts Reanalysis Interim
+    \citep[ERA-Interim;][]{Dee.etal.2011}, perturbed by time-dependent
+    temperature offsets, temperature lapse-rate corrections, and in some cases,
+    time-dependent paleo-precipitation reductions (Sect.~\ref{sec:atm},
+    Table~\ref{tab:records}).
+
+    \idea{Remove next paragraph?}
+
+    Each simulation starts from assumed present-day ice thickness and
+    equilibrium temperature distribution at 120\,\unit{ka}, and runs to the
+    present. Our modelling domain of 900 by 600\,\unit{km} encompasses the
+    entire Alpine range (Fig.~\ref{fig:inputs}). The simulations were run on
+    two distinct grids, using a~lower horizontal resolution of 2\,\unit{km},
+    and a~higher horizontal resolution of 1\,\unit{km}.
 
@@
       \citealp[Supplement]{Aschwanden.etal.2013};
-      \citealp{PISM-authors.2015}).
+      \citealp{PISM-authors.2017}).
 
@@
         background map consists of depressed SRTM \citep{Jarvis.etal.2008}
-        topography and Natural Earth Data \citep{Patterson.Kelso.2015}.
+        topography and Natural Earth Data \citep{Patterson.Kelso.2017}.
         \textbf{(b)} Temperature offset time-series from the EPICA ice core

### 2017-07-07 articles/cycle/alps-cycle.tex
Rework ice rheology subsection.

@@
 
+    \idea{In the following, I repeat all parameters in the text and table. Is
+          this necessary? Equations are the same as in my Cordillera paper.
+          Should I remove them?}
+
 \subsection{Ice rheology}
@@
 
-      Ice sheet dynamics are typically modelled using a~combination of
-      internal deformation and basal sliding. PISM is a~shallow ice sheet
-      model, which implies that the balance of stresses is approximated
-      based on their predominant components. The Shallow Shelf Approximation
-      (SSA) is combined with the Shallow Ice Approximation
-      (SIA) by adding velocity solutions of the two approximations
-      \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}. Although this
-      heuristic approach implies errors in the transition zone where gravitational
-      stresses intervene both in the SIA and SSA velocity computation, this
-      hybrid scheme is computationally much more efficient than a fully
-      three-dimensional model with which the simulations presented here
+    Ice sheet dynamics are typically modelled using a~combination of internal
+    deformation and basal sliding. PISM is a~shallow ice sheet model, which
+    implies that the balance of stresses is approximated based on their
+    predominant components. The Shallow Shelf Approximation (SSA) is combined
+    with the Shallow Ice Approximation (SIA) by adding velocity solutions of
+    the two approximations \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}.
+    Although this heuristic approach implies errors in the transition zone
+    where gravitational stresses intervene both in the SIA and SSA velocity
+    computation, this hybrid scheme is computationally much more efficient than
+    a fully three-dimensional model with which the simulations presented here
     would not be feasible.
@@
 %
-      where $\vec{\dot{\epsilon}}$ is the strain-rate tensor,
-      $\vec{\tau}$ the deviatoric stress tensor, and $\tau_{\mathrm{e}}$ the
-      effective stress defined in our case by ${\tau_{\mathrm{e}}}^2 = \frac{1}{2}
-      \mathrm{tr}(\vec{\tau}^2)$. The ice softness coefficient, $A$, depends
-      on ice temperature, $T$, pressure, $p$, and water content, $\omega$,
-      through a~piece-wise Arrhenius-type law,
+    where $\vec{\dot{\epsilon}}$ is the strain-rate tensor, $\vec{\tau}$ the
+    deviatoric stress tensor, and $\tau_{\mathrm{e}}$ the effective stress
+    defined in our case by
+    ${\tau_{\mathrm{e}}}^2=\frac{1}{2}\mathrm{tr}(\vec{\tau}^2)$. The ice
+    softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$,
+    and water content, $\omega$, through a~piece-wise Arrhenius-type law,
 %
@@
 %
-      where $T_{\text{pa}}$ is the pressure-adjusted ice temperature
-      calculated using the Clapeyron relation, ${T_{\text{pa}} = T - \beta
-      p}$. $R=8.31441$\,\unit{J\,mol^{-1}\,K^{-1}} is the ideal gas
-      constant, and $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
-      $Q_{\mathrm{w}}$, are constant parameters corresponding to values
-      measured below and above a~critical temperature threshold
-      $T_{\mathrm{c}}=-10$\,\unit{{\degree}C}
-      \citep[p.~72]{Paterson.Budd.1982,Cuffey.Paterson.2010}. The water
-      fraction, $\omega$, is capped at a~maximum value of 0.01, above which
-      no measurements are available \citep[Eq.~5.7]{Lliboutry.Duval.1985,
-      Greve.1997}. Finally, $E$ is a~non-dimensional enhancement factor
-      which can take different values, $E_{\text{SIA}}$, in the SIA and
-      $E_{\text{SSA}}$, in the SSA.
+    where $T_{\text{pa}}$ is the pressure-adjusted ice temperature calculated
+    using the Clapeyron relation, $T_{\text{pa}}=T-{\beta}p$.
+    $R=8.31441$\,\unit{J\,mol^{-1}\,K^{-1}} is the ideal gas constant, and
+    $A_{\mathrm{c}}=2.847\e{-13}$\,\unit{Pa^{-3}\,s^{-1}},
+    $A_{\mathrm{w}}=2.356\e{-2}$\,\unit{Pa^{-3}\,s^{-1}},
+    $Q_{\mathrm{c}}=6.0\e4$\,\unit{J\,mol^{-1}}, and
+    $Q_{\mathrm{w}}=11.5\e4$\,\unit{J\,mol^{-1}} are constant parameters
+    corresponding to values measured below and above a~critical temperature
+    threshold $T_{\mathrm{c}}=-10$\,\unit{{\degree}C}
+    \citep[p.~72]{Cuffey.Paterson.2010}. The water fraction, $\omega$, is
+    capped at a~maximum value of 0.01, above which no measurements are
+    available \citep[Eq.~5.7]{Lliboutry.Duval.1985, Greve.1997}. Finally, $E$
+    is a~non-dimensional enhancement factor which can take different values,
+    $E_{\text{SIA}}=2$, in the SIA and $E_{\text{SSA}}=1$, in the SSA, as
+    recommended for Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010}.
 
@@
     constant, $\beta=7.9\times 10^{-8}$\,\unit{K\,Pa^{-1}}, according to
-      \citet{Luthi.etal.2002}, the water fraction coefficient, $f=181.25$,
-      according to \citet{Lliboutry.Duval.1985}, and the SSA enhancement
-      factor, $E_{\text{SSA}}=1$, according to
-      \citet[p.~77]{Cuffey.Paterson.2010}. These fixed parameter values are
+    \citet{Luthi.etal.2002}, and the water fraction coefficient, $f=181.25$,
+    according to \citet{Lliboutry.Duval.1985}. These fixed parameter values are
     summarized in Table~\ref{tab:params}.
 
-      On the other hand, we test the model sensitivity
-      (Sect.~\ref{sec:sens}) to different values for the two creep
-      parameters, $A_{\mathrm{c}}$ and $A_{\mathrm{w}}$, the two activation
-      energies, $Q_{\mathrm{c}}$ and $Q_{\mathrm{w}}$, and the SIA
-      enhancement factor, $E_{\text{SIA}}$, as follows.
-%
-\begin{itemize}
-  \item Our \emph{default} configuration used in the control run of the
-    sensitivity study and all other simulations in the paper includes
-    rheological parameters,
-    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
-    $Q_{\mathrm{w}}$, derived from \citet{Paterson.Budd.1982} and given in
-    \citet[Eqn.~5]{Bueler.Brown.2009}, and $E_{\text{SIA}}=1$.
-  \item Our \emph{hard ice} configuration includes rheological parameters,
-    $A_{\mathrm{c}}$, $A_{\mathrm{w}}$, $Q_{\mathrm{c}}$ and
-    $Q_{\mathrm{w}}$, derived from \citet[p.~72 and
-    76]{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=1$, which correspond to
-    a~stiffer rheology than that used in the control run.
-  \item Our \emph{soft ice} configuration includes rheological parameters
-    from \citet{Cuffey.Paterson.2010}, and $E_{\text{SIA}}=5$, the
-    recommended value for ice age polar ice
-    \citep[p.~77]{Cuffey.Paterson.2010}.
-\end{itemize}
-%
-      An additional simulation using the ice rheology from
-      \citet{Cuffey.Paterson.2010} and $E_{\text{SIA}}=2$, the recommended
-      value for Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010} has
-      been performed, but is not presented here, since its results were very
-      similar to that of our default run.
-
-      Actual parameter values for $A_{\mathrm{c}}$, $A_{\mathrm{w}}$,
-      $Q_{\mathrm{c}}$, $Q_{\mathrm{w}}$ and $E_{\text{SIA}}$ used in our
-      simulations are given in Table~\ref{tab:sens_params}, while the effect
-      of the three different parametrisations on temperature-dependent ice
-      softness, $A$, is illustrated in Fig.~\ref{fig:sens_plot_rheo}.
-
-      Surface air
-      temperature derived from the climate forcing (Sect.~\ref{sec:atm})
-      provides the upper boundary condition to the ice enthalpy
-      model. Temperature is computed in the ice and in the bedrock to
+    Surface air temperature derived from the climate forcing
+    (Sect.~\ref{sec:atm}) provides the upper boundary condition to the ice
+    enthalpy model. Temperature is computed in the ice and in the bedrock to
     a~depth of 3\,\unit{km} below the ice-bedrock interface, where it is
-      conditioned by a~lower boundary geothermal heat flux of
-      $q_{\mathrm{G}}=70$\,\unit{mW\,m^{-2}}. Although this uniform value
-      does
-      not account for the high spatial geothermal variability in the region
-      \citep{Blackwell.Richards.2004}, it is, on average, representative of
-      available heat flow measurements. In the low-resolution simulations,
-      the vertical grid consists of 31~temperature layers in the bedrock and
-      up to 51~enthalpy layers in the ice sheet, corresponding to a~vertical
-      resolution of 100\,\unit{m}. The high-resolution simulations use
-      61~bedrock layers and up to 101~ice layers with a~vertical resolution
-      of 50\,\unit{m}.
+    conditioned by a~lower boundary geothermal heat flux estimate from multiple
+    geothermal proxies \citep[similarity method]{Goutorbe.etal.2011}. In the
+    2\,km-resolution simulations, the vertical grid consists of 31~temperature
+    layers in the bedrock and up to 126~enthalpy layers in the ice,
+    corresponding to vertical resolutions of 100 and 40\,\unit{m},
+    respectively. The 1\,km-resolution simulation uses 61~bedrock layers and up
+    to 251~ice layers with a~vertical resolutions of 50 and 20\,\unit{m}.
 

### 2017-07-07 articles/cycle/alps-cycle.tex
Rework basal sliding subsection.

@@
 %
-      relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to
-      the sliding velocity, $\vec{v}_{\mathrm{b}}$.
-      The yield stress, $\tau_{\mathrm{c}}$, is modelled using the
-      Mohr--Coulomb criterion,
+    relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to the
+    sliding velocity, $\vec{v}_{\mathrm{b}}$. The pseudo-plastic sliding
+    exponent, $q=0.25$, and the threshold velocity,
+    $v_{\text{th}}=100$\,\unit{m\,a^{-1}}, are set to values successfully used
+    to model the Greenland ice sheet \citet{Aschwanden.etal.2013} The yield
+    stress, $\tau_{\mathrm{c}}$, is modelled using the Mohr--Coulomb criterion,
 %
@@
 %
-      where cohesion, $c_0$, is assumed to be zero. Effective pressure, $N$,
-      is related to the ice overburden stress, $P_0=\rho gh$, and the
-      modelled amount of
-      subglacial water, using a~formula derived from laboratory experiments
-      with till extracted from the base of Ice Stream B in West Antarctica
-      \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
+    where the till cohesion, $c_0=0$, was consistently measured to be
+    negligible \citep[p.~268]{Tulaczyk.etal.2000, Cuffey.Paterson.2010}. We use
+    a~constant till friction angle, $\phi=30$\unit{\degree}, corresponding to
+    the average of values presented in \citet[p.~268]{Cuffey.Paterson.2010}.
+
+    Effective pressure, $N$, is related to the ice overburden stress, $P_0=\rho
+    gh$, and the modelled amount of subglacial water, using a~formula derived
+    from laboratory experiments with till extracted from the base of Ice Stream
+    B in West Antarctica \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
 %
@@
 %
-      where $\delta$ sets the minimum ratio between the effective and
-      overburden pressures, $e_0$ is a~reference void ratio and
-      $C_{\mathrm{c}}$ is the till compressibility coefficient
-      \citep{Tulaczyk.etal.2000}. The amount of water at the base, $W$,
-      varies from zero to $W_{\text{max}}$, a~threshold above which
-      additional melt water is assumed to drain off instantaneously.
-
-      In all our simulations, we set constant the pseudo-plastic sliding
-      exponent, $q=0.25$, and the threshold velocity,
-      $v_{\text{th}}=100$\,\unit{m\,a^{-1}}, according to values used by
-      \citet{Aschwanden.etal.2013}, the till cohesion, $c_0=0$, whose
-      measured values are consistently negligible
-      \citep[p.~268]{Tulaczyk.etal.2000, Cuffey.Paterson.2010}, the till
-      reference void ratio, $e_0=0.69$, and the till compressibility
-      coefficient, $C_{\mathrm{c}}=0.12$, according to the only measurements
-      available to our knowledge, published by \citet{Tulaczyk.etal.2000}.
-      These fixed parameter values are summarized in Table~\ref{tab:params}.
-
-      We also use a~constant spatial distribution of the till friction
-      angle, $\phi$, whose values vary from 15 to 45\unit{\degree} as
-      a~piecewise-linear function of modern bed elevation, with the lowest
-      value occuring below the modern sea level (0\,\unit{m} above sea level,
-      \unit{m\,a.s.l.}) and the highest value occuring above the generalised
-      elevation of the highest shorelines
-      \citep[200\,\unit{m\,a.s.l.},][Fig.~5]{Clague.1981}. This range of
-      values span over the range of measured values for glacial till of 18
-      to 40\unit{\degree} \citep[p.~268]{Cuffey.Paterson.2010}. It accounts
-      for frictional basal conditions associated with discontinuous till
-      cover at high elevations, and for a~weakening of till associated with
-      the presence of marine sediments (cf. \citealp{Martin.etal.2011};
-      \citealp[Supplement]{Aschwanden.etal.2013};
-      \citealp{PISM-authors.2017}).
-
-      An additional simulation with a~constant till friction angle,
-      $\phi=30$\unit{\degree}, corresponding to the average value in
-      \citet[p.~268]{Cuffey.Paterson.2010}, has been performed, but is not
-      presented here, since the induced variability was small.
-
-      On the other hand, we test the model sensitivity
-      (Sect.~\ref{sec:sens}) to different values for the minimum ratio
-      between the effective and overburden pressures, $\delta$, and the
-      maximum water height in the till, $W_{\text{max}}$, as follows.
-%
-\begin{itemize}
-  \item Our \emph{default} configuration used in the control run of the
-    sensitivity study and all other simulations in the paper includes
-    $\delta=0.02$ and $W_{\text{max}}=2$\,\unit{m} as in
-    \citet{Bueler.Pelt.2015}.
-  \item Our \emph{soft bed} configuration use $\delta=0.01$ and
-    $W_{\text{max}}=1$\,\unit{m}.
-  \item Our \emph{hard bed} configuration use $\delta=0.05$ and
-    $W_{\text{max}}=5$\,\unit{m}.
-\end{itemize}
-%
-      The effect of the three different parametrisations on the effective
-      pressure on the till, $N$, in response to water content, $W$, is
-      illustrated in Fig.~\ref{fig:sens_plot_ntil}. All parameter choices
-      are listed in Table~\ref{tab:sens_params}.
+    where $\delta=0.02$ sets the minimum ratio between the effective and
+    overburden pressures. Parameter values for the till reference void ratio,
+    $e_0=0.69$, and the till compressibility coefficient,
+    $C_{\mathrm{c}}=0.12$, were set to the only measurements available to our
+    knowledge \citep{Tulaczyk.etal.2000}. The amount of water at the base, $W$,
+    varies from zero to $W_{\text{max}}=2$\,m, a~threshold above which
+    additional melt water is assumed to drain off instantaneously. These fixed
+    parameter values are summarized in Table~\ref{tab:params}.
 

### 2017-07-11 articles/cycle/alps-cycle.tex
Add coloured hyperlinks.
Refine temperature offsets in text and figures.

@@
 \definecolor{darkred}{cmyk}{0.1,0.9,0.8,0.0}
+\definecolor{darkorange}{cmyk}{0.0,0.5,1.0,0.0}
+\definecolor{darkpurple}{cmyk}{0.6,0.7,0.0,0.0}
+\definecolor{darkbrown}{cmyk}{0.23,0.73,0.98,0.12}
 
@@
 \newcommand{\aref}[0]{\textcolor{darkblue}{\textbf{[REF.]}}}
+
+\hypersetup{colorlinks, linkcolor=darkorange,
+            urlcolor=darkbrown, citecolor=darkpurple}
 
@@
         MIS~4 according to a~global compilation of benthic \chem{\delta^{18}O}
-        records \citep{Lisiecki.Raymo.2005}.
-        \todo{Refine temperature offsets.}}
+        records \citep{Lisiecki.Raymo.2005}.}
       \label{fig:timeseries}
@@
         time-series yields reasonable MIS~4 ice cover.
-        \todo{Refine temperature offsets. Try to make reduced precipitation
-              outlines more visible.}}
+        \todo{Try to make reduced precipitation outlines more visible.}}
       \label{fig:footprints}
@@
         refers to the resulting mean temperature anomaly during the period 32
-        to~22\,\unit{ka} after scaling.
-        \todo{Refine temperature offsets.}}
+        to~22\,\unit{ka} after scaling.}
       \label{tab:records}
@@
                   & \multirow{2}{*}{\chem{U^{K'}_{37}}}
-                  & 1.82 & $-$7.9  % -4.345625
+                  & 1.84 & $-$8.0  % -4.345625
                   & \multirow{2}{*}{\citet{Martrat.etal.2007}} \\
 
-        MD01-2444, $\Delta P$ &&&&& 2.46 & $-$10.7 \\
+        MD01-2444, $\Delta P$ &&&&& 2.44 & $-$10.6 \\
 
@@
       %            & 1.05 & 1.33
-      %            & 1.82 & 2.46 \\
+      %            & 1.84 & 2.44 \\
       %
@@
       %            & $-$9.7\,K & $-$12.2\,K     % -9.2055
-      %            & $-$7.9\,K & $-$10.7\,K \\  % -4.345625
+      %            & $-$8.0\,K & $-$10.6\,K \\  % -4.345625
       %

### 2017-07-11 figures/util/__init__.py

@@ -16,8 +16,8 @@ alpcyc_offsets = [7.0, 9.0, 8.8, 11.3, 7.3, 9.7]  # total area till1545 5km
 alpcyc_offsets = [7.0, 9.0, 8.8, 11.4, 7.3, 9.8]  # total area till1545 2km
 alpcyc_offsets = [8.3, 10.5, 9.8, 12.3, 8.2, 10.7]  # Rhine area till1545 5km
 alpcyc_offsets = [8.2, 10.4, 9.5, 12.1, 8.1, 10.5]  # Rhine area phi30 5km
-alpcyc_offsets = [8.2, 10.4, 9.6, 12.2, 8.1, 10.7]  # Rhine area phi30 2km (preliminary)
-#              = [8.2, 10.4, 9.7, 12.2, 7.9, 10.7]  # Rhine area phi30 2km (estimate)
+alpcyc_offsets = [8.2, 10.4, 9.7, 12.2, 8.0, 10.6]  # Rhine area phi30 2km
+#                 val   pre  val   val  val   pre   # (validated/preliminary)
 alpcyc_colours = [pl.palette[tone+hue] for hue in ['blue', 'red', 'green']
                                        for tone in ['dark', 'light']]
 alpcyc_colmaps = ['Blues_r']*2 + ['Reds_r']*2 + ['Greens_r']*2

### 2017-07-13 articles/cycle/alps-cycle.tex
Rework surface mass balance subsection.

@@
     Ice surface accumulation and ablation are computed from monthly mean
-      near-surface air temperature, $T_{\mathrm{m}}$, monthly standard
-      deviation of near-surface air temperature, $\sigma$, and monthly
-      precipitation, $P_{\mathrm{m}}$, using a~temperature-index model
-      \citep[e.g.,][]{Hock.2003}. Accumulation is equal to precipitation
-      when air temperatures are below 0\,\unit{{\degree}C}, and decreases to
-      zero linearly with temperatures between 0 and 2\,\unit{{\degree}C}.
-      Ablation is computed from PDD, defined as an integral of temperatures
-      above 0\,\unit{{\degree}C} in one year.
+    near-surface air temperature, $T_{\mathrm{m}}$, monthly standard deviation
+    of near-surface air temperature, $\sigma$, and monthly precipitation,
+    $P_{\mathrm{m}}$, using a~temperature-index model
+    \citep[e.g.,][]{Hock.2003}. Accumulation is equal to precipitation when air
+    temperatures are below 0\,\unit{{\degree}C}, and decreases to zero linearly
+    with temperatures between 0 and 2\,\unit{{\degree}C}. Ablation is computed
+    from PDD, defined as an integral of temperatures above 0\,\unit{{\degree}C}
+    in one year.
 
     The PDD computation accounts for stochastic temperature variations by
-      assuming a~normal temperature distribution of standard deviation
-      $\sigma$ around the expected value $T_{\mathrm{m}}$. It is expressed
-      by an error-function formulation \citep{Calov.Greve.2005},
+    assuming a~normal temperature distribution of standard deviation $\sigma$
+    around the expected value $T_{\mathrm{m}}$. It is expressed by an
+    error-function formulation \citep{Calov.Greve.2005},
+%
 \begin{align}
@@
 \end{align}
+%
     which is numerically approximated using week-long sub-intervals. In
@@
     temperature variability \citep{Seguinot.2013}, $\sigma$ is computed
-      from NARR daily temperature values from 1979 to 2000
+    from ERA-Interim daily temperature values from 1979 to 2012
     \citep{Mesinger.etal.2006}, including variability associated with the
-      seasonal cycle, and bilinearly interpolated to the model grids
-      (Fig.~\ref{fig:atm}). Degree-day factors for snow and
-      ice melt are derived from mass-balance measurements on contemporary
-      glaciers from the Coast Mountains and Rocky Mountains in British
-      Columbia \citep[Table~\ref{tab:params};][]{Shea.etal.2009}.
+    seasonal cycle \citep{Seguinot.2013}, and bilinearly interpolated to the
+    model grids (Fig.~\ref{fig:inputs}). Degree-day factors for snow and ice
+    melt are set to values used in the European Ice Sheet Modelling INiTiative
+    \citep[Table~\ref{tab:params}; EISMINT,][]{Huybrechts.1998}.
 

### 2017-07-13 articles/cycle/alps-cycle.tex
Rework climate forcing subsection.

@@
 
-      Climate forcing driving ice sheet simulations consists of
-      a~present-day monthly climatology, $\{T_{\mathrm{m}0},
-      P_{\mathrm{m}0}\}$, where temperatures are modified by offset time
-      series, ${\Delta}T_{\text{TS}}$, and lapse-rate corrections,
-      ${\Delta}T_{\text{LR}}$:
+    Climate forcing driving ice sheet simulations consists of a~present-day
+    monthly climatology, $\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, where
+    temperatures are modified by offset time series, ${\Delta}T_{\text{TS}}$,
+    and lapse-rate corrections, ${\Delta}T_{\text{LR}}$:
+%
 \begin{align}
-&T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) + {\Delta}T_{\text{TS}}(t)
+T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) + {\Delta}T_{\text{TS}}(t)
                     + {\Delta}T_{\text{LR}}(t, x, y) \,, \\
-&    P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \,.
 \end{align}
+%
     The present-day monthly climatology was bilinearly interpolated from
-      near-surface air
-      temperature and precipitation rate fields from the NARR, averaged from
-      1979 to 2000. Modern climate of the North American Cordillera is
-      characterised by strong geographic variations in temperature
-      seasonality, timing of the maximum annual precipitation, and daily
-      temperature variability (Fig.~\ref{fig:atm}). Although the ability
-      of the NARR to reproduce the steep climatic gradients is limited by its
-      spatial resolution of 32\,\unit{km} \citep{Jarosch.etal.2012}, it has
-      been tested against observational data in our previous sensitivity
-      study and identified as yielding a closer fit between the modelled LGM
-      extent of the Cordilleran ice sheet and the geological evidence than
-      other atmospheric reanalyses \citep{Seguinot.etal.2014}.
+    near-surface air temperature and precipitation rate fields from
+    \citep[WorldClim;][]{Hijmans.etal.2005}, representative of the period 1960
+    to 1990. Modern climate of the European Alps is characterised by a
+    latitudinal gradient in summer air temperatures, and a longitudinal
+    gradient in winter precipitation (Fig.~\ref{fig:inputs}). WorldClim data
+    were selected as an input to the ice sheet model because they incorporate
+    observations from the dense weather station network of central Europe.
+    Besides, last glacial cycle alpine glaciers did not extend over marine
+    areas where WorldClim data is missing. Finally, WorldClim data were
+    previously used as climate forcing for PISM to model the LGM extent of the
+    former Cordilleran ice sheet in good agreement with geological evidence
+    along the southern margin \citep{Seguinot.etal.2014} were weather station
+    density is lower than in the Alps.
 
-      Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived
-      from palaeo-temperature proxy records from the Greenland Ice Core
-      Project \citep[GRIP,][]{Dansgaard.etal.1993}, the North Greenland Ice
-      Core Project \citep[NGRIP,][]{Andersen.etal.2004}, the European
-      Project for Ice Coring in Antarctica \citep[EPICA,][]
-      {Jouzel.etal.2007}, the Vostok ice core \citep{Petit.etal.1999}, and
-      Ocean Drilling Program (ODP) sites 1012 and 1020, both located off the
-      coast of California \citep{Herbert.etal.2001}. Palaeo-temperature
-      anomalies from the GRIP and NGRIP records were calculated from oxygen
-      isotope (\chem{\delta^{18}O}) measurements using a~quadratic equation
-      \citep{Johnsen.etal.1995},
+    Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived from
+    palaeo-temperature proxy records from the Greenland Ice Core Project
+    \citep[GRIP,][]{Dansgaard.etal.1993}, the European Project for Ice Coring
+    in Antarctica \citep[EPICA,][] {Jouzel.etal.2007}, and an oceanic sediment
+    core from the Iberian margin \citep[MD01-2444,][]{Martrat.etal.2007}.
+    Palaeo-temperature anomalies from the GRIP record were calculated from
+    oxygen isotope (\chem{\delta^{18}O}) measurements using a~quadratic
+    equation \citep{Johnsen.etal.1995},
+%
 \begin{align}
@@
 \end{align}
-      while temperature reconstructions from Antarctic and oceanic cores
-      were provided as such. For each proxy record used and each of the
-      parameter setup used in the sensitivity tests, palaeo-temperature
-      anomalies were scaled linearly (Tables~\ref{tab:records}
-      and~\ref{tab:sens_params}) in order to simulate comparable ice extents
-      at the LGM (Table~\ref{tab:extrema}) and realistic outlines
-      (Fig.~\ref{fig:lr_maps}).
+%
+    while temperature reconstructions from the EPICA and MD01-2444 records were
+    provided as such. For each proxy record used and each of the parameter
+    setup used in the sensitivity tests, palaeo-temperature anomalies are
+    scaled linearly (Table~\ref{tab:records}) so that the cumulative glaciated
+    area of the Rhine glacier piedmont lobe during Oxygen Marine Isotope Stage
+    (MIS)~2 (29--14\,ka) is modelled comparably to the reconstructions
+    (Fig.~\ref{fig:footprints}).
 
-      Finally, lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are computed
-      as a~function of ice surface elevation, $s$, using the NARR surface
-      geopotential height invariant field as a~reference topography,
-      $b_{\text{ref}}$:
+    The temperature lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are
+    computed as a~function of ice surface elevation, $s$, using the SRTM
+    topography shipped with WorldClimas a~reference, $b_{\text{ref}}$:
+%
 \begin{align}
@@
 \end{align}
-      thus accounting for the evolution of ice thickness, ${h=s-b}$, on the
-      one hand, and for differences between the basal topography of the ice
-      flow model, $b$, and the NARR reference topography, $b_{\text{ref}}$,
-      on the other hand. All simulations use an annual temperature lapse
-      rate of $\gamma = 6\,\unit{K\,km^{-1}}$. In the rest of this paper, we
-      refer to different model runs by the name of the proxy record used for
-      the palaeo-temperature forcing.
+%
+    thus accounting for the evolution of ice thickness, ${h=s-b}$, on the one
+    hand, and for differences between the basal topography of the ice flow
+    model, $b$, and the NARR reference topography, $b_{\text{ref}}$, on the
+    other hand. All simulations use an annual temperature lapse rate of $\gamma
+    = 6\,\unit{K\,km^{-1}}$.
+
+    Finally, in some simulations, precipitation was reduced with temperature in
+    order to simulate the potential rarification of atmospheric moisture in
+    colder climates. This was done using an empirical relationship derived from
+    observed accumulation rates and oxygen isotopes concentrations in the GRIP
+    ice core \citep{Dahl-Jensen.etal.1993},
+%
+\begin{align}
+P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \cdot
+                          \exp[\zeta{\Delta}T_{\text{LR}}(t, x, y)] \,.
+\end{align}
+%
+    with $\zeta=0.169/2.4=0.0704$ \citep{Huybrechts.2002}. This simple
+    relationship likely does not reflect the complexity of atmospheric
+    circulation changes that governed moisture availability over the Alps
+    during the last glacial cycle. Thus other simulations use constant
+    precipitation, corresponding to $\zeta=0$. In the rest of this paper, we
+    refer to different model runs by the name of the proxy record used for the
+    palaeo-temperature forcing. Model runs using paleo-precipitation
+    corrections are identified by appending ${\Delta}P$ to their name.
 
@@
 
-        $\omega$& Precipitation factor
+        $\zeta$ & Precipitation factor
                 & 0.0704

### 2017-07-13 articles/cycle/alps-cycle.tex
Subdivide paleo-climate forcing section.
Rework to make LaTeX file look pretty.

@@
 
-    Table~\ref{tab:params} -- Model parameters\\
+    Table~\ref{tab:params} -- Model parameters.\\
     Fig.~\ref{fig:inputs} -- Climate and geothermal model inputs.\\
 
-    \begin{itemize}
-    \item Ice dynamics are approximated by SSA+SIA.
-    \item Ice temperature follows an enthalpy scheme.
-    \item Bedrock temperature computed until 3\,km depth
-    \item Temperature conditioned by geothermal flux (Fig.~\ref{fig:inputs}d).
-    \item Basal sliding follwos a pseudo-plastic law.
-    \item Yield stress is determined by a Mohr-Coulomb criterion.
-    \item Till friction angle is constant.
-    \item Till effective pressure relates to basal melt.
-    \item Surface mass balance is computed by a PDD model.
-    \item Modern T and P forcing from WorldClim (Fig.~\ref{fig:inputs}a and~b).
-    \item Modern PDD SD forcing from ERA-Interim (Fig.~\ref{fig:inputs}c).
-    \item SRTM topography with glaciers removed \citep{Huss.Farinotti.2012}.
-    \end{itemize}
-
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 \subsection{Overview}
-\label{sec:overview}%
+\label{sec:overview}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 
@@
 
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 \subsection{Ice rheology}
 \label{sec:icedyn}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 
@@
     \begin{align}
-&\label{eqn:glenslaw}
-&\vec{\dot{\epsilon}} = A\,\tau_{\mathrm{e}}^{n-1}\,\vec{\tau} \,.
+      \vec{\dot{\epsilon}} = A\,\tau_{\mathrm{e}}^{n-1}\,\vec{\tau} \,.
     \end{align}
@@
     \begin{align}
-&\label{eqn:softness}
     &A = E\cdot
       \begin{cases}
-A_{\mathrm{c}} \,e^\frac{-Q_{\mathrm{c}}}{RT_{\text{pa}}}            & \text{if}\ T_{\text{pa}}  <  T_{\mathrm{c}} \,, \\
-A_{\mathrm{w}} (1+f\omega)\,e^\frac{-Q_{\mathrm{w}}}{RT_{\text{pa}}} & \text{if}\ T_{\text{pa}} \ge T_{\mathrm{c}} \,,
+        A_{\mathrm{c}} \,e^\frac{-Q_{\mathrm{c}}}{RT_{\text{pa}}}
+            & \text{if}\ T_   {\text{pa}}  <  T_{\mathrm{c}} \,, \\
+        A_{\mathrm{w}} (1+f\omega)\,e^\frac{-Q_{\mathrm{w}}}{RT_{\text{pa}}}
+            & \text{if}\ T_   {\text{pa}} \ge T_{\mathrm{c}} \,,
       \end{cases}
@@
 
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 \subsection{Basal sliding}
 \label{sec:sliding}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 
@@
     \begin{align}
-&\label{eqn:pseudoplastic}
-    \vec{\tau}_{\mathrm{b}} = -\tau_{\mathrm{c}} \frac{\vec{v}_{\mathrm{b}}}{{v_{\text{th}}}^q\,|\vec{v}_{\mathrm{b}}|^{1-q}} \,,
+      \vec{\tau}_{\mathrm{b}} = -\tau_{\mathrm{c}}
+        \frac{\vec{v}_{\mathrm{b}}}
+             {{v_{\text  {th}}}^q\,|\vec{v}_{\mathrm{b}}|^{1-q}} \,,
     \end{align}
@@
     \begin{align}
-&\tau_{\mathrm{c}} = c_0 + N\,\tan{\phi} \,,
+      \tau_{\mathrm{c}} = c_0 + N\,\tan{\phi} \,,
     \end{align}
@@
     \begin{align}
-&\label{eqn:ntil}
-&N = \delta P_0 \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,,
+      N = \delta P_0 \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,,
     \end{align}
@@
 
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 \subsection{Basal topography}
 \label{sec:bedrock}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 
@@
 
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 \subsection{Surface mass balance}
 \label{sec:surface}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 
@@
     \begin{align}
-&\label{eqn:calovgreve}
       {\text{PDD}} = \int_{t_1}^{t_2} \mathrm{d}t
@@
 
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 \subsection{Climate forcing}
-\label{sec:atm}%
+\label{sec:atm}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \idea{move paleo-climate forcing info to the next section.}
 
@@
     \begin{align}
-T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) + {\Delta}T_{\text{TS}}(t)
-                    + {\Delta}T_{\text{LR}}(t, x, y) \,, \\
+      T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) +
+                                {\Delta}T_{\text{TS}}(t) +
+                                {\Delta}T_{\text{LR}}(t, x, y) \,, \\
     \end{align}
@@
                                     -\chem{\delta^{18}O}(0)] \nonumber \\
-                        &-0.1925[\chem{\delta^{18}O}(t)^2
+                                   &-0.1925 [\chem{\delta^{18}O}(t)^2
                                     -\chem{\delta^{18}O}(0)^2] \,,
@@
       {\Delta}T_{\text{LR}}(t, x, y) &= -\gamma [s(t, x, y)-b_{\text{ref}}] \\
-                            &= -\gamma [h(t, x, y)+b(t, x, y)-b_{\text{ref}}],
+                                     &= -\gamma [h(t, x, y)+
+                                                 b(t, x, y)-b_{\text{ref}}],
     \end{align}
@@
     model, $b$, and the NARR reference topography, $b_{\text{ref}}$, on the
-    other hand. All simulations use an annual temperature lapse rate of $\gamma
-    = 6\,\unit{K\,km^{-1}}$.
+    other hand. All simulations use an annual temperature lapse rate of
+    $\gamma=6\,\unit{K\,km^{-1}}$.
 
@@
 
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Palaeo-temperature forcing}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
     \begin{itemize}
@@
     \item All records scaled to fit the LGM extent (Table~\ref{tab:records}).
+    \end{itemize}
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Palaeo-precipitation forcing}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{itemize}
+    \item Some simulations use reduced precipitation (Table~\ref{tab:records}).
+    \end{itemize}
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Sensitivity of ice volume evolution}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{itemize}
     \item All yield glaciation during MIS~4 and 2 (Fig.~\ref{fig:timeseries}).
     \item But otherwise very different results (Fig.~\ref{fig:timeseries}).
+    \end{itemize}
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Sensitivity of glaciated area}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{itemize}
     \item Two yield two large MIS~4 ice extent (Fig.~\ref{fig:footprints}).

### 2017-07-13 articles/cycle/alps-cycle.tex
Move palaeo forcing methods to dedicated section.

@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Climate forcing}
+\subsection{Reference climate forcing}
 \label{sec:atm}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    \idea{move paleo-climate forcing info to the next section.}
 
@@
                                 {\Delta}T_{\text{TS}}(t) +
-                                {\Delta}T_{\text{LR}}(t, x, y) \,, \\
+                                {\Delta}T_{\text{LR}}(t, x, y) \,,
     \end{align}
@@
     density is lower than in the Alps.
+
+    The temperature lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are
+    computed as a~function of ice surface elevation, $s$, using the SRTM
+    topography shipped with WorldClimas a~reference, $b_{\text{ref}}$:
+    %
+    \begin{align}
+      {\Delta}T_{\text{LR}}(t, x, y) &= -\gamma [s(t, x, y)-b_{\text{ref}}] \\
+                                     &= -\gamma [h(t, x, y)+
+                                                 b(t, x, y)-b_{\text{ref}}],
+    \end{align}
+    %
+    thus accounting for the evolution of ice thickness, ${h=s-b}$, on the one
+    hand, and for differences between the basal topography of the ice flow
+    model, $b$, and the NARR reference topography, $b_{\text{ref}}$, on the
+    other hand. All simulations use an annual temperature lapse rate of
+    $\gamma=6\,\unit{K\,km^{-1}}$.
+
+% ----------------------------------------------------------------------
+\section{Palaeo-climate forcing}
+% ----------------------------------------------------------------------
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Palaeo-temperature forcing}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 
@@
 
-    The temperature lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are
-    computed as a~function of ice surface elevation, $s$, using the SRTM
-    topography shipped with WorldClimas a~reference, $b_{\text{ref}}$:
-    %
-    \begin{align}
-      {\Delta}T_{\text{LR}}(t, x, y) &= -\gamma [s(t, x, y)-b_{\text{ref}}] \\
-                                     &= -\gamma [h(t, x, y)+
-                                                 b(t, x, y)-b_{\text{ref}}],
-    \end{align}
-    %
-    thus accounting for the evolution of ice thickness, ${h=s-b}$, on the one
-    hand, and for differences between the basal topography of the ice flow
-    model, $b$, and the NARR reference topography, $b_{\text{ref}}$, on the
-    other hand. All simulations use an annual temperature lapse rate of
-    $\gamma=6\,\unit{K\,km^{-1}}$.
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Palaeo-precipitation forcing}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 
@@
     corrections are identified by appending ${\Delta}P$ to their name.
-
-% ----------------------------------------------------------------------
-\section{Palaeo-climate forcing}
-% ----------------------------------------------------------------------
-
-    Table~\ref{tab:records} -- Palaeo-climate records.\\
-    Fig.~\ref{fig:timeseries} -- Low-resolution time series.\\
-    Fig.~\ref{fig:footprints} -- Low-resolution ice cover.
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Palaeo-temperature forcing}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    \begin{itemize}
-    \item We conducted a 5\,km-resolution sensitivity study.
-    \item We used three palaeo-temperature records (Table~\ref{tab:records}).
-    \item All records scaled to fit the LGM extent (Table~\ref{tab:records}).
-    \end{itemize}
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Palaeo-precipitation forcing}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    \begin{itemize}
-    \item Some simulations use reduced precipitation (Table~\ref{tab:records}).
-    \end{itemize}
 

### 2017-07-13 articles/cycle/alps-cycle.tex
Add introductory paragraph to palaeo-clim section.

@@
 
+    Table~\ref{tab:records} -- Palaeo-climate records.\\
+    Fig.~\ref{fig:timeseries} -- Low-resolution time series.\\
+    Fig.~\ref{fig:footprints} -- Low-resolution ice cover.\\
+
+    In this section, we analyze the model sensitivity to palaeo-climate forcing
+    through the last glacial cycle, using three palaeo-temperature records
+    (Sect.~\ref{sec:paltemp}) and two parametrizations of palaeo-precipitation
+    (Sect.~\ref{sec:palprec}), in terms of modelled evolution of total ice
+    volume (Sect.~\ref{sec:timeseries}) and glaciated area during MIS~2 and~4
+    (Sect.~\ref{sec:footprints}).
+
+    These simulations use an horizontal resolution of 2\,km. The vertical grid
+    consists of 31~temperature layers in the bedrock and up to 126~enthalpy
+    layers in the ice, corresponding to vertical resolutions of 100 and
+    40\,\unit{m}, respectively.
+
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 \subsection{Palaeo-temperature forcing}
+\label{sec:paltemp}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
@@
 \subsection{Palaeo-precipitation forcing}
+\label{sec:palprec}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
@@
 \subsection{Sensitivity of ice volume evolution}
+\label{sec:timeseries}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
@@
 \subsection{Sensitivity of glaciated area}
+\label{sec:footprints}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -

### 2017-07-13 articles/cycle/alps-cycle.tex
Introduce psi notation for paleo-precipitation.

@@
     Climate forcing driving ice sheet simulations consists of a~present-day
-    monthly climatology, $\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, where
-    temperatures are modified by offset time series, ${\Delta}T_{\text{TS}}$,
-    and lapse-rate corrections, ${\Delta}T_{\text{LR}}$:
+    monthly climatology, $\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, modified by
+    temperature lapse-rate corrections, ${\Delta}T_{\text{LR}}$, temperature
+    offset time series, ${\Delta}T_{\text{TS}}$, and time-dependent
+    palaeo-precipitation corrections, $\Psi_{\text{PP}}$:
     %
     \begin{align}
-      T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) +
-                                {\Delta}T_{\text{TS}}(t) +
-                                {\Delta}T_{\text{LR}}(t, x, y) \,,
+      &T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) +
+                                 {\Delta}T_{\text{LR}}(t, x, y) +
+                                 {\Delta}T_{\text{TS}}(t) \,, \\
+      &P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \cdot
+                                 {\Psi}_{\text{PP}}(t) \,, \\
     \end{align}
@@
     \begin{align}
-      P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \cdot
-                                \exp[\zeta{\Delta}T_{\text{LR}}(t, x, y)] \,.
+      {\Psi}_{\text{PP}}(t) = \exp[\psi{\Delta}T_{\text{TS}}(t)] \,.
     \end{align}
     %
-    with $\zeta=0.169/2.4=0.0704$ \citep{Huybrechts.2002}. This simple
+    with $\psi=0.169/2.4=0.0704$ \citep{Huybrechts.2002}. This simple
     relationship likely does not reflect the complexity of atmospheric
@@
     during the last glacial cycle. Thus other simulations use constant
-    precipitation, corresponding to $\zeta=0$. In the rest of this paper, we
+    precipitation, corresponding to $\psi=0$. In the rest of this paper, we
     refer to different model runs by the name of the proxy record used for the
     palaeo-temperature forcing. Model runs using paleo-precipitation
-    corrections are identified by appending ${\Delta}P$ to their name.
+    corrections are labelled PP.
 
@@
 
-        $\zeta$ & Precipitation factor
+        $\psi$  & Precipitation factor
                 & 0.0704

### 2017-07-13 figures/util/__init__.py
Use PP instead of Delta P in figure labels.

@@ -21,6 +21,6 @@ alpcyc_offsets = [8.2, 10.4, 9.7, 12.2, 8.0, 10.6]  # Rhine area phi30 2km
 alpcyc_colours = [pl.palette[tone+hue] for hue in ['blue', 'red', 'green']
                                        for tone in ['dark', 'light']]
 alpcyc_colmaps = ['Blues_r']*2 + ['Reds_r']*2 + ['Greens_r']*2
-alpcyc_clabels = [rec + ('pp' in conf)*', $\Delta P$' for (rec, conf) in
+alpcyc_clabels = [rec + ('pp' in conf)*', PP' for (rec, conf) in
                   zip(alpcyc_records, alpcyc_configs)]
 alpcyc_bestrun = 'output/e9d2d1f/alps-wcnn-1km/epica3222cool1220+alpcyc4+pp/'

### 2017-07-18 articles/cycle/alps-cycle.tex
Add tentative introduction, a big mess.

@@
 \section{Introduction}
-% ----------------------------------------------------------------------
-
-    \begin{itemize}
-    \item The European Alps are the cradle of pioneer glacial studies.
-    \item Today, their glacial history is better understood than any other.
-    \item The LGM extent is very well known (Fig.~\ref{fig:lgmvel}a).
-    \item But modelling was difficult due to the steep topography.
-    \item Here, we us PISM to model the last glacial cycle (120--0\,ka).
-    \end{itemize}
+\label{sec:intro}
+% ----------------------------------------------------------------------
+
+    Glaciers move by a combination of meltwater-induced sliding at the base
+    \citep[\S532]{Saussure.1779}, and viscous deformation within the ice body
+    \citep{Forbes.1846b}. As glaciers flow and slide across their bed, they
+    transport rock debris and erode the landscape, thereby leaving
+    geomorphological traces of their former presence.
+
+    For nearly 300~years, montane people and early explorers of the European
+    Alps learned to read the geomorphological imprint left by glaciers in the
+    landscape, and to understand that glaciers had once been more extensive
+    than today \citep[e.g.,][p.~21]{Windham.Martel.1744}. In the mid-nineteenth
+    century, more systematic studies of glacial features showed that alpine
+    glaciers extended well below their contemporary margins \citep{Venetz.1821}
+    and even onto the alpine foreland \citep{Charpentier.1841}, yielding the
+    idea that, under colder temperatures, expansive ice sheets had once covered
+    much of Europe and North America \citep{Agassiz.1840}.
+
+    However, this glacial theory did not gain general acceptance until the
+    discovery and exploration of the two present-day ice sheets on Earth, the
+    Greenland and Antarctic ice sheets, provided a modern analogue for the
+    proposed European and North American ice sheets. Although it was long
+    unclear whether there had been a single or multiple glaciations, this
+    controversy ended with the large-scale mapping of two distinct moraine
+    systems in North America \citep{Chamberlin.1894}. In the European Alps, the
+    systematic classification of the glaciofluvial terraces on the northern
+    foreland indicated that landscape, that there had been at least four major
+    glaciations in the Alps \citep{Penck.Bruckner.1909}. More recent studies
+    of glaciofluvial stratigraphy in the foreland indicate up to eight
+    glaciations \citep{Ivy-Ochs.etal.2008, Preusser.etal.2011}.
+
+    More recently, palaeoclimate records extracted from deep sea sediments and
+    ice cores have provided a much more detailed picture of the Earth
+    environmental history \citep[e.g.,][]{Emiliani.1955,
+    Shackleton.Opdyke.1973, Dansgaard.etal.1993, Augustin.etal.2004},
+    indicating neither one, nor four, but rather tens of glacial cycles. During
+    the last 800000 years (800\,ka), glacial and interglacial periods have
+    succeeded each other with a 100\,ka periodicity \citep{Hays.etal.1976,
+    Augustin.etal.2004}. However, much of the traces left by glaciers on the
+    landscape date from the most recent of these cycles, the \emph{last glacial
+    cycle}.
+
+    Nevertheless, the landform record is sparse in time and, most often,
+    spatially incomplete. Palaeo-ice sheets did not leave a continuous imprint
+    on the lanscape, and much of this evidence has been overprinted by
+    subsequent glacier re-advances and other geomorphological processes
+    \citep{Kleman.1994, Kleman.etal.2006, Kleman.etal.2010}. The landscape
+    imprint of previous glaciations has been largely overprinted by the more
+    recent ones; therefore most of the glacial features currently left on the
+    foreland present a record of the last major glaciation of the Alps, dating
+    from the Last Glacial Maximum (LGM).
+
+    Because this global signal is largely governed by the North American and
+    Eurasian ice sheet complexes, it is unclear wether glacier advances and
+    retreats in the Alps were in pac(71--57\,ka)e with global sea-level
+    fluctuations. The timing of the LGM in the Alps (Ivy-Ochs et al., 2008;
+    Monegato et al., 2007) is in good agreement with the maximum expansion of
+    continental ice sheets recorded by the Marine Oxygen Isotope Stage (MIS) 2
+    between 29 000 and 14 000 years before present (BP) (Lisiecki and Raymo,
+    2005). There are nonetheless regional variations between different piedmont
+    lobes, that are of the same order of magnitude with the dating
+    uncertainties. Because recent glaciation overprinted older ones, and
+    because dating uncertainties typically increase with age, dated
+    reconstructions are strongly biased towards earlier glaciations
+    \citep{Heyman.etal.2011}, and thus much of the glacial records dates to the
+    Last Glacial Maximum (LGM). The Alps have been studied in more detail than
+    any other glacial region. There is evidence for glaciation during MIS 4
+    (71--57\,ka) but potential foreland glaciations during the rest of the last
+    glacial cycle are still discussed.
+
+    Here, we use the Parallel Ice Sheet Model
+    \citep[PISM,][]{PISM-authors.2017}, a numerical ice sheet model that
+    approximates glacier sliding and deformation (Sect.~\ref{sec:model}), to
+    model alpine glacier dynamics through the last glacial cycle (120--0\,ka),
+    a period for which palaeo-temperature proxies are available, albeit non
+    regional. We test the model sensitivity to multiple palaeo-climate forcing
+    (Sect.~\ref{sec:climate}), and then explore the modelled glacier dynamics
+    at high resolution for the optimal forcing (Sect.~\ref{sec:results}), and
+    the resulting glacial erosion pattern (Sect.~\ref{sec:erosion}).
 
@@
 \section{Ice sheet model set-up}
+\label{sec:model}
 % ----------------------------------------------------------------------
@@
 \section{Palaeo-climate forcing}
+\label{sec:climate}
 % ----------------------------------------------------------------------
@@
 \section{Glacier dynamics}
+\label{sec:results}
 % ----------------------------------------------------------------------
@@
 \section{Erosion potential}
+\label{sec:erosion}
 % ----------------------------------------------------------------------

### 2017-07-19 articles/cycle/alps-cycle.tex
Add palaeo-climate sensitivity first shot.

@@
 
-    \begin{itemize}
-    \item All yield glaciation during MIS~4 and 2 (Fig.~\ref{fig:timeseries}).
-    \item But otherwise very different results (Fig.~\ref{fig:timeseries}).
-    \end{itemize}
+    For the three palaeo-temperature records and the two palaeo-precipitation
+    parametrizations used, the model yield significant ice volume build-up
+    during MIS~4 and 2 (Fig.~\ref{fig:timeseries}). All simulations also
+    yield important glaciations during MIS~5 and 3, but timing and amplitude
+    varies a lot between the different forcing. All simulations overestimate
+    ice cover during the Younger Dryas. This might be due partly to the 2\,km
+    resolution, except for GRIP which largely overestimates Younger Dryas ice
+    cover.
+
+    Even more importantly, all simulations yield very strong ice volume
+    variability which is the result of multiple glacier advance and retreats
+    over the foreland, many more than two. Palaeo-precipitation reductions
+    partly results in a smoother modelled ice volume time series and less
+    extreme variations, but variability remains high. Ice volume variability
+    is the smallest for the EPICA palaeo-temperature record, which has the
+    least temperature variability.
 
@@
 
-    \begin{itemize}
-    \item Two yield two large MIS~4 ice extent (Fig.~\ref{fig:footprints}).
-    \item Only EPICA yield realistic MIS~4 (Fig.~\ref{fig:footprints}).
-    \end{itemize}
+    To select an optimal climate forcing, we look at the best known alpine
+    glaciations during MIS~2 and 4. During MIS~2, all forcing yield ice extent
+    of the same order of magnitude (Fig.~\ref{fig:footprints}). This is
+    because the temperature records have been scaled to resul in similar
+    ice covered area for the Rhine Glacier piedmont lobe. Looking at ice cover
+    on other parts, all simulations tend to overestimate ice extent in the
+    eastern part of the model domain, and to underestimate it in the western
+    part, as compared to the reconstructed LGM margin. Regarding the western
+    part, there are discussions though, as if the Lyon Lobe actually dates from
+    MIS~2 or from an older glaciation. Regarding the eastern part, the MIS~2
+    glaciation has been mapped in detail and dated (?), so that there is no
+    doubt that our model overestimate ice cover in this region. This could
+    indicate that during LGM the east-west precipitation gradient over the Alps
+    was lower than today, or that temperature depression in the east was lower
+    than in the west.
+
+    There are spatial differences though: MD01-2444, and to a greater extent
+    GRIP, forcings, tend to overestimate ice cover in all peripherical ranges:
+    the Vosges, Black Forest, Bavarian Forest, Dinaric Alps and Pennine. This
+    is because these records have a higher temperature variability so that
+    after scaling they comport brief periods of cold climate, not long enough
+    to develop the alpine ice cap, but long enough to build up ice on the
+    peripherical ranges.
+
+    The extent of ice cover during MIS~4 shows more sensitivity to the choice
+    of palaeo-climate forcing. Using the GRIP and MD01-2444, glaciers extend
+    well beyound the reach of documented moraines and glacial erratic terrain.
+    Because this extent corresponds to a short-term cold climate event,
+    palaeo-precipitation reductions help to greatly reduce this excessive
+    modelled ice cover, yet in both cases, modelled ice extent and volume
+    during MIS~4 remains higher than during MIS~2. EPICA forcing, on the other
+    hand, yield a MIS~4 glaciation that is only slightly less expensive than
+    the MIS~2, with only little sensitivity of the glaciated area to
+    palaeo-precipitation reductions.
+
+    Based on the above consideration on timing of the LGM ice extent and
+    ice extent during MIS~4, we chose EPICA as our optimal temperature record
+    for the rest of this paper. As a conservative approach in regard to quick
+    ice volume fluctuations, we choose to include palaeo-precipitation
+    correction in the following 1\,km-resolution simulation.
 

### 2017-07-21 articles/cycle/alps-cycle.tex
Subdivide results section.

@@
 
+    In this section, we compare the model output to geological evidence from
+    the last glacial cycle, in terms of LGM extent (Sect.~\ref{sec:extent}),
+    ice flow patterns (Sect.~\ref{sec:flow}), ice thickness
+    (Sect.~\ref{sec:thickness}), timing (Sect.~\ref{sec:timing}), and the
+    number of glaciations (Sect.~\ref{sec:glaciations}).
+
+    This simulation is forced by the optimal EPICA palaeo-temperature record
+    (Sect.~\ref{sec:paltemp}) and includes palaeo-precipitation reduction
+    (Sect.~\ref{sec:palprec}). It uses an horizontal resolution of 1\,km. The
+    vertical grid consists of 61~temperature layers in the bedrock and up to
+    251~enthalpy layers in the ice, corresponding to vertical resolutions of 50
+    and 20\,\unit{m}, respectively.
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{LGM extent}
+\label{sec:extent}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
     \begin{itemize}
@@
     \item Two major glaciations during MIS 4 and 2 (Fig.~\ref{fig:lgmvel}b).
+    \item Maximum extent is farily well reproduced (Fig.~\ref{fig:lgmvel}a).
+    \end{itemize}
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{LGM flow pattern}
+\label{sec:flow}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{itemize}
     \item Maximum stages flow pattern is complex (Fig.~\ref{fig:lgmvel}a).
-    \item Maximum extent is farily well reproduced (Fig.~\ref{fig:lgmvel}a).
+    \end{itemize}
+
+    \idea{Add trimline figure in the main text now.}
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{LGM ice thickness}
+\label{sec:thickness}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{itemize}
+    \item Ice surface is well above the trimlines (Fig.~\ref{fig:lgmvel}a).
+    \end{itemize}
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Timing of the LGM}
+\label{sec:timing}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{itemize}
     \item Maximum extent is time-transgressive (Fig.~\ref{fig:timing}).
     \item Glaciers have different response times (Fig.~\ref{fig:timing}).
+    \end{itemize}
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Number of glaciations}
+\label{sec:glaciations}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{itemize}
     \item Some glaciers advance many times (Fig.~\ref{fig:profiles}).
@@
           elevation again in the Supplementary.}
-
-    \idea{Add trimline figure in the main text now.}
 

### 2017-08-05 figures/alpcyc_hr_trimlines.py
Remove math symbols in trimline figure.
They are not needed now that I removed the equation.

@@ -96,8 +96,8 @@ ax.scatter(xt, yt, c=at, cmap=cmap, norm=norm, s=4**2, alpha=0.75)
 
 # draw scatter plot
 sc = scax.scatter(zt, ht, c=at, cmap=cmap, norm=norm, alpha=0.75)
-scax.set_xlabel('observed trimline elevation $z_t$ (m)')
-scax.set_ylabel('modelled max ice thickness $h_t$ (m)', labelpad=2)
+scax.set_xlabel('observed trimline elevation (m)')
+scax.set_ylabel('modelled maximum ice thickness (m)', labelpad=2)
 
 # add colorbar
 cb = fig.colorbar(sc, cax)

### 2017-08-05 articles/cycle/alps-cycle.tex
Add trimline figure caption and update notes.

@@
         a~global compilation of benthic \chem{\delta^{18}O} records
-        \citep{Lisiecki.Raymo.2005}.
-        \note{I expect the simulation to complete in late July.}}
+        \citep{Lisiecki.Raymo.2005}.}
       \label{fig:lgmvel}
@@
         \textbf{(a)} Timing of the Last Glacial Maximum (LGM) given by the
-        modelled age (colour mapping) and value (200 m contours) of maximum
+        modelled age (colour mapping) and value (200\,m contours) of maximum
         surface elevation throughout the entire simulation.
@@
         and modelled glaciated area around the LGM (coloured curve). The LGM
-        is here modelled as a time-transgressive event.
-        \note{I expect the simulation to complete in late July.}}
+        is here modelled as a time-transgressive event.}
       \label{fig:timing}
+    \end{figure}
+
+    \begin{figure}
+      \centerline{\includegraphics{alpcyc_hr_trimlines}}
+      \caption{%
+        \textbf{(a)} Modelled age of maximum ice thickness (colour) and maximum
+        ice surface elevation (200\,m contours) above trimline locations
+        \citep[Table~1]{Kelly.etal.2004} in the upper Rhone valley.
+        \textbf{(b)} Modelled maximum ice thickness against observed trimline
+        elevations \citep[Table~1]{Kelly.etal.2004}.
+        \textbf{(c)} Histogram of modelled maximum ice thickness above trimline
+        locations (500\,m bands). The average maximum ice thickness above the
+        trimline is 979\,m.}
+      \label{fig:trimlines}
     \end{figure}
@@
               profiles drawn on panel a. But then we should probably limit it
-              to two glaciers instead of four.}
-        \note{I expect the simulation to complete in late July.}}
+              to two glaciers instead of four.}}
       \label{fig:profiles}
@@
         $l = 2.02$ and $K_g = 2.7\e{-7}\,m^{1-l}\,a^{l-1}$
-        \citep{Herman.etal.2015}.
-        \note{I expect the simulation to complete in late July.}}
+        \citep{Herman.etal.2015}.}
       \label{fig:erosion}

### 2017-08-08 articles/cycle/alps-cycle.tex
Rework last section's outline.

@@
     Fig.~\ref{fig:timing} -- Timing of the LGM and area time series.\\
+    Fig.~\ref{fig:trimlines} -- LGM ice thickness compared to trimlines.\\
     Fig.~\ref{fig:profiles} -- Individual glacier extent profiles.
@@
     In this section, we compare the model output to geological evidence from
-    the last glacial cycle, in terms of LGM extent (Sect.~\ref{sec:extent}),
-    ice flow patterns (Sect.~\ref{sec:flow}), ice thickness
-    (Sect.~\ref{sec:thickness}), timing (Sect.~\ref{sec:timing}), and the
-    number of glaciations (Sect.~\ref{sec:glaciations}).
+    the last glacial cycle, in terms of maximum extent and ice flow patterns
+    (Sect.~\ref{sec:flow}), timing of the Last Glacial Maximum
+    (Sect.~\ref{sec:timing}), ice thickness (Sect.~\ref{sec:thickness}), and
+    glacial cycle dynamics (Sect.~\ref{sec:glaciations}).
 
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{LGM extent}
-\label{sec:extent}
+\subsection{Maximum configuration}
+\label{sec:flow}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
@@
     \begin{itemize}
-    \item Rapid variations of total ice volume (Fig.~\ref{fig:lgmvel}b).
-    \item Two major glaciations during MIS 4 and 2 (Fig.~\ref{fig:lgmvel}b).
+    \item Maximum extent is reached at 24.57 ka (Fig.~\ref{fig:lgmvel}b).
+    \item Some glaciers are not at their maximum (Fig.~\ref{fig:lgmvel}a).
     \item Maximum extent is farily well reproduced (Fig.~\ref{fig:lgmvel}a).
+    \item Peripheral ice caps well reproduced (Fig.~\ref{fig:lgmvel}a).
+    \item Too much ice in the east (Fig.~\ref{fig:lgmvel}a).
+    \item Too little ice in the west (Fig.~\ref{fig:lgmvel}a).
+    \item Maximum stages flow pattern is complex (Fig.~\ref{fig:lgmvel}a).
+    \item Flow goes sometimes across topography (Fig.~\ref{fig:lgmvel}a).
+    \item Transfluences (Montgenèvre, col de l'Échelle, Simplon, Albrun, Gries,
+                        (Brünig, Majola, Tagliamento)
     \end{itemize}
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{LGM flow pattern}
-\label{sec:flow}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    \begin{itemize}
-    \item Maximum stages flow pattern is complex (Fig.~\ref{fig:lgmvel}a).
-    \end{itemize}
-
-    \idea{Add trimline figure in the main text now.}
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{LGM ice thickness}
-\label{sec:thickness}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    \begin{itemize}
-    \item Ice surface is well above the trimlines (Fig.~\ref{fig:lgmvel}a).
-    \end{itemize}
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Timing of the LGM}
+\subsection{Timing of the Last Glacial Maximum}
 \label{sec:timing}
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Number of glaciations}
+\subsection{Ice thickness and trimlines}
+\label{sec:thickness}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{itemize}
+    \item Ice surface is well above the trimlines (Fig.~\ref{fig:lgmvel}a).
+    \end{itemize}
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Glacial cycle dynamics}
 \label{sec:glaciations}
@@
     \begin{itemize}
+    \item Rapid variations of total ice volume (Fig.~\ref{fig:lgmvel}b).
+    \item Two major glaciations during MIS 4 and 2 (Fig.~\ref{fig:lgmvel}b).
     \item Some glaciers advance many times (Fig.~\ref{fig:profiles}).
@@
         Solothurn, Rhine and Ivrea glaciers.
-        \todo{Implement extraction of ice thickness along a more detailed
-              profile, and add the other glaciers.}
         \idea{We can also move this as panel b on Fig.~\ref{fig:timing}, with

### 2017-08-08 articles/cycle/alps-cycle.tex
Update figure captions and notes.

@@
         \textbf{(c)} Mordern July standard deviation of daily mean temperature
-        from the ERA-Interim \citep[1979--2012]{Dee.etal.2011} from the
-        reference monthly climatology used to force the surface mass balance
-        (PDD) component of the ice sheet model.
+        from the ERA-Interim \citep[1979--2012]{Dee.etal.2011, Seguinot.2013}
+        from the reference monthly climatology used to force the surface mass
+        balance (PDD) component of the ice sheet model.
         \textbf{(d)} Geothermal heat flow from applying the similarity method
@@
         the ice sheet model.
-        \textbf{(a)} Modelled total ice volume through the last 120\,ka,
-        expressed in meters of sea level equivalent (m~s.l.e.). Gray fields
-        indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2 and
-        MIS~4 according to a~global compilation of benthic \chem{\delta^{18}O}
-        records \citep{Lisiecki.Raymo.2005}.}
+        \textbf{(b)} Modelled total ice volume through the last 120~thousand
+        years (ka) expressed in meters of sea level equivalent (m~s.l.e.). Gray
+        fields indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2
+        and MIS~4 according to a~global compilation of benthic
+        \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}.}
       \label{fig:timeseries}
@@
         \textbf{(a--c)} Cumulative extent of modelled ice cover during MIS~2
-        (29--14\,ka) using temperature time-series scaling factors
-        (Table~\ref{tab:records}) adjusted to obtain model results in agreement
-        with the Last Glacial Maximum (LGM) geomorphological reconstruction
-        \citep[solid red line,][]{Ehlers.etal.2011}.
+        (29--14\,ka), with (light colours) and without (dark colours)
+        palaeo-precipitation corrections, and using temperature time-series
+        scaling factors (Table~\ref{tab:records}) adjusted to model the area
+        glaciated by the Rhine glacier piedmont lobe (black rectangle) in
+        agreement with the Last Glacial Maximum (LGM) geomorphological
+        reconstruction \citep[solid red line,][]{Ehlers.etal.2011}.
         \textbf{(d--f)} Cumulative extent of modelled ice cover during MIS~4
@@
         time-series yields reasonable MIS~4 ice cover.
-        \todo{Try to make reduced precipitation outlines more visible.}}
+        \todo{Add Rhine lobe scaling domain.}}
       \label{fig:footprints}
@@
       \caption{%
-        \textbf{(a)} Modelled bedrock topography (grey) ice surface topography
-        (200\,m contours) and ice surface velocity (blue) in the Alps
-        21~thousand years (ka) before present. Modelled Last Glacial Maximum
+        \textbf{(a)} Modelled bedrock topography (grey), ice surface topography
+        (200\,m contours), and ice surface velocity (blue) in the Alps
+        24.57\,ka before present, corresponding to the maximum modelled ice
+        cover. Modelled Last Glacial Maximum
         (LGM) ice extent (dashed orange line) and geomorphological
@@
       \caption{%
-        \textbf{(a)} Timing of the Last Glacial Maximum (LGM) given by the
+        \textbf{(a)} Timing of the LGM given by the
         modelled age (colour mapping) and value (200\,m contours) of maximum
@@
         used as palaeo-climate forcing for the ice flow model (black curve),
-        and modelled glaciated area around the LGM (coloured curve). The LGM
+        and modelled glaciated area during the LGM (coloured curve). The LGM
         is here modelled as a time-transgressive event.}
@@
         locations (500\,m bands). The average maximum ice thickness above the
-        trimline is 979\,m.}
+        trimline is 979\,m.
+        \todo{Move map below scatter plot and add cold ice areas or so. Use max
+              surface elevation age as in Fig.~\ref{fig:timing}.}}
       \label{fig:trimlines}
@@
       \caption{%
-        Modelled extent of glaciation along selected profiles for the Lyon,
-        Solothurn, Rhine and Ivrea glaciers.
-        \idea{We can also move this as panel b on Fig.~\ref{fig:timing}, with
-              profiles drawn on panel a. But then we should probably limit it
-              to two glaciers instead of four.}}
+        Modelled glacier extent and thickess along selected profiles for the
+        Rhine, Rhone, Ivrea and Isère glaciers.
+        \todo{Add accent on Isère, show ice thickness.}}
       \label{fig:profiles}

### 2017-08-08 articles/cycle/alps-cycle.tex
Remove erosion section and figure.

@@
     (Sect.~\ref{sec:climate}), and then explore the modelled glacier dynamics
-    at high resolution for the optimal forcing (Sect.~\ref{sec:results}), and
-    the resulting glacial erosion pattern (Sect.~\ref{sec:erosion}).
+    at high resolution for the optimal forcing (Sect.~\ref{sec:results}).
 
@@
           elevation again in the Supplementary.}
-
-% ----------------------------------------------------------------------
-\section{Erosion potential}
-\label{sec:erosion}
-% ----------------------------------------------------------------------
-
-    Fig.~\ref{fig:erosion} -- Integrated erosion potential.
-
-    \idea{Frank: I would keep this topic for another paper.}
-
-    \begin{itemize}
-    \item We calculate erosion after \citet{Herman.etal.2015}
-    \item More erosion in valleys of western Alps (Fig.~\ref{fig:erosion}a).
-    \item Erosion is constant during 110--15\,ka (Fig.~\ref{fig:erosion}b).
-    \end{itemize}
 
@@
 
-    \begin{figure}
-      \centerline{\includegraphics{alpcyc_hr_erosion}}
-      \caption{%
-        \textbf{(a)} Modelled total erosion integrated in time over the entire
-        simulation (120--0\,ka) is highest in the deep valleys and cirques of
-        the western Alps.
-        \textbf{(b)} Temperature offset time-series from the EPICA ice core
-        used as palaeo-climate forcing for the ice flow model (black curve),
-        and modelled erosion rate integrated in space over the entire
-        ice-covered area. The local erosion rate, $\dot{e}$, is computed from
-        the sliding velocity, $\vec{v}_{\mathrm{b}}$, through
-        $\dot{e} = K_g \cdot |\vec{v}_{\mathrm{b}}|^{l}$, with
-        $l = 2.02$ and $K_g = 2.7\e{-7}\,m^{1-l}\,a^{l-1}$
-        \citep{Herman.etal.2015}.}
-      \label{fig:erosion}
-    \end{figure}
-
 % ----------------------------------------------------------------------
@@
       \caption{%
-        Parameter values used in the ice sheet model.
-        \todo{Update parameters to new configuration.}}
+        Parameter values used in the ice sheet model.}
       \label{tab:params}

### 2017-08-08 figures/alpcyc_lr_footprints.py
Add Rhine scaling domain on footprints figure.

@@ -22,6 +22,14 @@ for ax in grid.flat:
     im = nc.imshow('topg', ax=ax, vmin=0e3, vmax=3e3, cmap='Greys', zorder=-1)
 nc.close()
 
+# add Rhine lobe scaling domain  # FIXME move to util
+w, e, s, n = ut.pl.regions['rhlobe']
+x = [w, e, e, w, w]
+y = [s, s, n, n, s]
+for ax in grid[0]:
+    ax.plot(x, y, c='k', lw=0.5)
+
+
 # for each record
 for i, rec in enumerate(ut.alpcyc_records):
     label = ut.alpcyc_clabels[i]

### 2017-08-08 figures/alpcyc_hr_trimlines.py
Move map below scatter plot on trimline figure.

@@ -5,12 +5,12 @@ import util as ut
 import numpy as np
 
 # initialize figure (one column)
-figw, figh = 85.0, 90.0
+figw, figh = 85.0, 115.0
 fig = ut.pl.figure(figsize=(figw/25.4, figh/25.4))
-ax = fig.add_axes([10.0/figw, 57.5/figh, 45.0/figw, 30.0/figh], projection=ut.pl.utm)
-cax = fig.add_axes([57.5/figw, 57.5/figh, 5.0/figw, 30.0/figh])
-scax = fig.add_axes([10.0/figw, 7.5/figh, 45.0/figw, 47.5/figh])
-hsax = fig.add_axes([57.5/figw, 7.5/figh, 25.0/figw, 47.5/figh], sharey=scax)
+ax = fig.add_axes([2.5/figw, 2.5/figh, 80.0/figw, 80.0*2/3/figh], projection=ut.pl.utm)
+cax = fig.add_axes([12.5/figw, (80.0*2/3-2.5)/figh, 30.0/figw, 2.5/figh])
+scax = fig.add_axes([10.0/figw, 65.0/figh, 45.0/figw, 47.5/figh])
+hsax = fig.add_axes([57.5/figw, 65.0/figh, 25.0/figw, 47.5/figh], sharey=scax)
 
 ## initialize figure (full width)
 #figw, figh = 170.0, 60.0
@@ -25,9 +25,9 @@ ax.set_rasterization_zorder(2.5)
 ax.set_extent(ut.pl.regions['valais'], crs=ax.projection)
 
 # add subfigure labels
-ut.pl.add_subfig_label('(a)', ax=ax)
-ut.pl.add_subfig_label('(b)', ax=scax)
-ut.pl.add_subfig_label('(c)', ax=hsax)
+ut.pl.add_subfig_label('(a)', ax=scax)
+ut.pl.add_subfig_label('(b)', ax=hsax)
+ut.pl.add_subfig_label('(c)', ax=ax)
 
 
 # Input data
@@ -100,7 +100,7 @@ scax.set_xlabel('observed trimline elevation (m)')
 scax.set_ylabel('modelled maximum ice thickness (m)', labelpad=2)
 
 # add colorbar
-cb = fig.colorbar(sc, cax)
+cb = fig.colorbar(sc, cax, orientation='horizontal')
 cb.set_label(r'age (ka)')
 
 

### 2017-08-08 articles/cycle/alps-cycle.tex
Add Rhine scaling domain on footprints figure.
Move map below scatter plot on trimline figure.
Fix Isère label on profile plot.
Tried plotting several contours for ice thickness but that is too much info.

@@
         (71--57\,ka). Only the simulation driven by the EPICA temperature
-        time-series yields reasonable MIS~4 ice cover.
-        \todo{Add Rhine lobe scaling domain.}}
+        time-series yields reasonable MIS~4 ice cover.}
       \label{fig:footprints}
@@
         trimline is 979\,m.
-        \todo{Move map below scatter plot and add cold ice areas or so. Use max
+        \todo{Add cold ice areas or so. Use max
               surface elevation age as in Fig.~\ref{fig:timing}.}}
@@
         Modelled glacier extent and thickess along selected profiles for the
-        Rhine, Rhone, Ivrea and Isère glaciers.
-        \todo{Add accent on Isère, show ice thickness.}}
+        Rhine, Rhone, Ivrea and Isère glaciers.}
       \label{fig:profiles}

### 2017-08-08 figures/alpcyc_hr_profiles.py

@@ -8,6 +8,7 @@ import cartopy.io.shapereader as shpreader
 
 # parameters
 regions = ['rhine', 'rhone', 'ivrea', 'isere']
+labels = ['Rhine', 'Rhone', 'Ivrea', u'Isère']
 colors = [ut.pl.palette['dark'+hue] for hue in ['blue', 'green', 'red', 'orange']]
 
 # initialize figure
@@ -30,6 +31,7 @@ nc.close()
 for i, reg in enumerate(regions):
     c = colors[i]
     ax = grid[i]
+    label = labels[i]
 
     # Map axes
     # --------
@@ -86,7 +88,7 @@ for i, reg in enumerate(regions):
     # set axes properties
     tsax.set_xlim(120.0, 0.0)
     tsax.set_xlabel('model age (ka)', labelpad=-20)
-    tsax.set_ylabel('%s glacier (km)' % reg.capitalize())
+    tsax.set_ylabel('%s glacier (km)' % label)
     tsax.xaxis.set_visible(i==len(regions)-1)
     tsax.yaxis.set_label_position("right")
     tsax.yaxis.tick_right()

### 2017-08-08 figures/alpcyc_hr_trimlines.py
Add warm ice cover less than 1ka on trimlines fig.

@@ -49,8 +49,11 @@ nc = ut.io.load(filepath)
 x = nc.variables['x'][:]
 y = nc.variables['y'][:]
 age = -nc.variables['time'][:]/(1e3*365.0*24*60*60)
-thk = nc.variables['thk'][:]
-srf = nc.variables['usurf'][:]
+mis = (age < 29.0) * (age >= 14.0)
+age = age[mis]
+thk = nc.variables['thk'][mis]
+srf = nc.variables['usurf'][mis]
+tpa = nc.variables['temppabase'][mis]
 nc.close()
 
 # compute max thickness and its age
@@ -58,6 +61,12 @@ lgmage = age[thk.argmax(axis=0)]
 maxthk = thk.max(axis=0)
 maxsrf = srf.max(axis=0)
 mask = (thk < 1.0).prod(axis=0)
+maxsrf = np.ma.masked_where(mask, maxsrf)
+
+# compute duration of warm-based coved
+dt = age[0] - age[1]
+warm = ((thk >= 1.0)*(tpa >= -1e-3)).sum(axis=0)*dt
+warm = np.ma.masked_where(mask, warm)
 
 # get model elevation at trimline locations
 i = np.argmin(abs(xt[:, None] - x), axis=1)
@@ -66,8 +75,8 @@ ht = maxthk[j, i]
 at = lgmage[j, i]
 
 
-# Map axes
-# --------
+# Scatter axes
+# ------------
 
 # set contour levels and colors
 mpl = ut.pl.iplt.matplotlib
@@ -76,32 +85,10 @@ cmap = ut.pl.get_cmap('Paired', 12)
 cols = cmap(range(12))[:len(levs)+1]
 cmap, norm = mpl.colors.from_levels_and_colors(levs, cols, extend='both')
 
-# plot boot topo
-nc = ut.io.load('input/boot/alps-srtm+thk+gou11simi-1km.nc')
-im = nc.imshow('topg', ax, vmin=0.0, vmax=3e3, cmap='Greys', zorder=-1)
-boot = nc.variables['topg'][:]
-nc.close()
-
-# add ice mask and contour levels
-ax.contourf(x, y, mask, levels=[-0.5, 0.5], colors='w', alpha=0.75)
-ax.contour(x, y, maxsrf, ut.pl.inlevs, colors='0.25', linewidths=0.1)
-ax.contour(x, y, maxsrf, ut.pl.utlevs, colors='0.25', linewidths=0.25)
-
[…]

### 2017-08-08 articles/cycle/alps-cycle.tex
Add warm ice cover less than 1ka on trimlines fig.
Use max elev. age in trimline fig, update caption.

@@
       \caption{%
-        \textbf{(a)} Modelled age of maximum ice thickness (colour) and maximum
+        \textbf{(a)} Modelled maximum ice thickness against observed trimline
+        elevations \citep[Table~1]{Kelly.etal.2004}.
+        \textbf{(b)} Histogram of modelled maximum ice thickness above trimline
+        locations (500\,m bands). The average maximum ice thickness above the
+        trimline is 979\,m.}
+        \textbf{(c)} Modelled age of maximum ice thickness (colour) and maximum
         ice surface elevation (200\,m contours) above trimline locations
-        \citep[Table~1]{Kelly.etal.2004} in the upper Rhone valley.
-        \textbf{(b)} Modelled maximum ice thickness against observed trimline
-        elevations \citep[Table~1]{Kelly.etal.2004}.
-        \textbf{(c)} Histogram of modelled maximum ice thickness above trimline
-        locations (500\,m bands). The average maximum ice thickness above the
-        trimline is 979\,m.
-        \todo{Add cold ice areas or so. Use max
-              surface elevation age as in Fig.~\ref{fig:timing}.}}
+        \citep[Table~1]{Kelly.etal.2004} in the upper Rhone valley. Hatches
+        mark glaciated areas with less than 1\,ka temperate ice cover during
+        MIS~2 (29--14\,ka).
       \label{fig:trimlines}

### 2017-08-08 figures/alpcyc_hr_trimlines.py

@@ -57,7 +57,7 @@ tpa = nc.variables['temppabase'][mis]
 nc.close()
 
 # compute max thickness and its age
-lgmage = age[thk.argmax(axis=0)]
+lgmage = age[srf.argmax(axis=0)]
 maxthk = thk.max(axis=0)
 maxsrf = srf.max(axis=0)
 mask = (thk < 1.0).prod(axis=0)

### 2017-08-18 figures/alpcyc_hr_trimlines.py
Add lines for +/- one std on trimlines figure.

@@ -108,10 +108,15 @@ hsax.yaxis.tick_right()
 # highlight mean thickness
 zavg = zt.mean()
 davg = dt.mean()
+dstd = dt.std()
 zz = [zt.min(), zt.max()]
 scax.plot(zz, zz+davg, c='0.25')
+scax.plot(zz, zz+davg-dstd, c='0.25', dashes=(2, 1), lw=0.5)
+scax.plot(zz, zz+davg+dstd, c='0.25', dashes=(2, 1), lw=0.5)
 scax.axhline(zavg+davg, c='0.25', dashes=(2, 1), lw=0.5)
-hsax.axhline(davg, c='0.25', dashes=(2, 1), lw=0.5)
+hsax.axhline(davg, c='0.25')
+hsax.axhline(davg-dstd, c='0.25', dashes=(2, 1), lw=0.5)
+hsax.axhline(davg+dstd, c='0.25', dashes=(2, 1), lw=0.5)
 hsax.text(2.0, davg+25.0, '%.0f m' % davg, color='0.25')
 
 # align axes bounds

### 2017-08-18 articles/cycle/alps-cycle.tex
Update figure captions.

--- 2017-08-08 articles/cycle/alps-cycle.tex
@@
       \caption{%
-        \textbf{(a)} Timing of the LGM given by the
-        modelled age (colour mapping) and value (200\,m contours) of maximum
-        surface elevation throughout the entire simulation.
+        \textbf{(a)} Timing of the LGM given by the modelled age of maximum ice
+        thickness throughout the entire simulation (colour mapping) and
+        corresponding, ice surface elevation (200\,m contours).
         \textbf{(b)} Temperature offset time-series from the EPICA ice core
         used as palaeo-climate forcing for the ice flow model (black curve),
-        and modelled glaciated area during the LGM (coloured curve). The LGM
+        and modelled glacierized area during the LGM (coloured curve). The LGM
         is here modelled as a time-transgressive event.}
@@
       \caption{%
-        \textbf{(a)} Modelled age of maximum ice thickness (colour) and maximum
-        ice surface elevation (200\,m contours) above trimline locations
-        \citep[Table~1]{Kelly.etal.2004} in the upper Rhone valley.
-        \textbf{(b)} Modelled maximum ice thickness against observed trimline
-        elevations \citep[Table~1]{Kelly.etal.2004}.
-        \textbf{(c)} Histogram of modelled maximum ice thickness above trimline
-        locations (500\,m bands). The average maximum ice thickness above the
-        trimline is 979\,m.
-        \todo{Add cold ice areas or so. Use max
-              surface elevation age as in Fig.~\ref{fig:timing}.}}
+        \textbf{(a)} Comparison of modelled ice surface elevation at the LGM
+        (time-transgressive, corresponding to maximum ice thickness,
+        Fig.~\ref{fig:timing}), compensated for bedrock deformation, against
+        observed trimline elevations \citep[Table~1]{Kelly.etal.2004}. Model
+        variables were bilinearly interpolated to the trimline locations.
+        \textbf{(b)} Histogram of differences between modelled LGM ice surface
+        elevation and trimline elevations (500\,m bands). The average
+        difference is 861\,m.}
+        \textbf{(c)} Modelled age of maximum ice thickness at the trimline
+        locations \citep[Table~1]{Kelly.etal.2004} in the upper Rhone valley
+        (colour) and LGM ice surface elevation (200\,m contours). Hatches
+        mark areas modelled to have experienced less than 1\,ka temperate ice
+        cover during MIS~2 (29--14\,ka).
       \label{fig:trimlines}
@@
       \caption{%
-        Modelled glacier extent and thickess along selected profiles for the
-        Rhine, Rhone, Ivrea and Isère glaciers.}
+        \textbf{(a, c, e, g)} Approximate glacier flowlines drawn by hand
+        rougly following selected glacial valley centerlines.
+        \textbf{(b, d, f, h)} Evolution of modelled glacier extent in time,
+        bilinearly interpolated along the corresponding profiles, showing
+        numerous cycles of advance and retreat over the last glacial cycle.
+        Isolated patches indicate periodic surges from tributary glaciers.}
       \label{fig:profiles}

### 2017-08-23 articles/cycle/alps-cycle.tex
Draft discusion of maximum extent.

@@
     In this section, we compare the model output to geological evidence from
-    the last glacial cycle, in terms of maximum extent and ice flow patterns
+    the last glacial cycle, in terms of Last Glacial Maximum extent
+    (Sect.~\ref{sec:extent}), ice flow patterns
     (Sect.~\ref{sec:flow}), timing of the Last Glacial Maximum
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Maximum configuration}
+\subsection{Last Glacial Maximum ice extent}
+\label{sec:extent}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    The modelled total ice volume reaches a maximum of 122.5\e{3}\,\unit{km^2},
+    or 300.4\,mm of sea-level equivalent, at 24.56\,ka
+    (Fig.~\ref{fig:lgmvel}b). A maximum glacierized area of
+    163.3\e{3}\,\unit{km^2} is attained shortly aftewards at 24.57\,ka.
+    (Fig.~\ref{fig:lgmvel}b). Although the modelled timing of the LGM varies
+    accross the mountain range (Sect.~\ref{sec:timing}), at 24.57\,ka nearly
+    all glaciers extend to within a few km from their modelled maximum stage
+    (Fig.~\ref{fig:lgmvel}a, dashed orange line).
+
+    Although the palaeo-climate forcing was adapted (Sect.~\label{sec:climate})
+    to model this maximum configuration in broad agreement with the geological
+    reconstruction of the LGM extent (Fig.~\ref{fig:lgmvel}a, solid red line),
+    local discrepancies remain. Most notably, reconstructed ice extent in the
+    north-western Alps, including the Rhone Glacier complex, the Jura ice cap,
+    and the Lyon Lobe, is largely missing from our model results. On the other
+    hand, the model yields excessive glaciation in the eastern Alps, where the
+    Palten, Drau and Sava glaciers are modelled to extend tens of kilometers
+    beyond the mapped moraines. These discrepancies might indicate that the
+    LGM climate was characterized by an east-west gradient in temperature or
+    precipitation anomalies relative to present not included in the climate
+    forcing, or perhaps a east-west gradient in variables not accounted for
+    by the PDD model, such as cloud cover or dust deposition.
+
+    On a more regional level, the modelled maximum extent of the Ticino, Iseo,
+    Garda and Piave glaciers in the central southern Alps is constrained
+    several kilometres within the mapped LGM moraines. This appears to result
+    from the paleo-precipitation reduction used to force the model
+    (Fig.~\ref{footprints}b), which is likely unrealistic in at least this part
+    of the model domain. On the other hand, the Durance glacier in the
+    south-western Alps is modelled to extend several kilometers beyond the
+    mapped limit, indicating that the LGM temperature depression was likely
+    dampened by Mediterranean climate in this part of the model domain.
+
+    The model reproduces peripheral ice caps where documented by geological
+    evidence on the Vercors, Chartreuse and Bauge prealpine reliefs,
+    the Jura Mountains, the Vosges Mountain, the Black Forest, the Bohemian
+    Forest and the Dinaric Alps. An independent ice cap also covers the
+    Hochwart massif during most of the simulation, yet it is engulfed by
+    alpine glaciers during the LGM to become a peripheral ice dome. The LGM
+    extent of peripheral ice caps is underestimated in the Vosges and Jura
+    mountains, while it is overestimated for the Vercors. There is a conformity
+    between the discrepancies between modelled and reconstructed LGM extent
+    for the main alpine glacier complex vs peripheral ice caps, which indicates
+    a climatic cause rather than an ice dynamics cause.
+
+    \todo{add location names on Fig.~\ref{fig:lgmvel}.}
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Ice flow patterns}
 \label{sec:flow}
@@
     \begin{itemize}
-    \item Maximum extent is reached at 24.57 ka (Fig.~\ref{fig:lgmvel}b).
-    \item Some glaciers are not at their maximum (Fig.~\ref{fig:lgmvel}a).
-    \item Maximum extent is farily well reproduced (Fig.~\ref{fig:lgmvel}a).
-    \item Peripheral ice caps well reproduced (Fig.~\ref{fig:lgmvel}a).
-    \item Too much ice in the east (Fig.~\ref{fig:lgmvel}a).
-    \item Too little ice in the west (Fig.~\ref{fig:lgmvel}a).
     \item Maximum stages flow pattern is complex (Fig.~\ref{fig:lgmvel}a).

### 2017-09-12 data/postprocess_alpcyc_bestrun.py
Add LGM velocity postprocessing, update contours.

@@ -28,6 +28,7 @@ thk = da.from_array(nc.variables['thk'], chunks=chunks)
 srf = da.from_array(nc.variables['usurf'], chunks=chunks)
 tpa = da.from_array(nc.variables['temppabase'], chunks=chunks)
 cba = da.from_array(nc.variables['velbase_mag'], chunks=chunks)
+csu = da.from_array(nc.variables['velsurf_mag'], chunks=chunks)
 dx = x[1] - x[0]
 dy = y[1] - y[0]
 dt = age[0] - age[1]
@@ -39,6 +40,13 @@ icy = (thk >= 1.0)
 cols, rows = thk.shape[1:]
 jj = np.arange(cols)[:, None]
 kk = np.arange(rows)[None, :]
+
+# extract time slices
+a = 24570.0
+lgmidx = abs(age-a).argmin()
+lgmicethk = thk[lgmidx].compute()
+lgmicesrf = srf[lgmidx].compute()
+lgmsrfvel = csu[lgmidx].compute()
 
 # compute aggregated variables (apparently dask fails to perform parallel
 # multiplication on two netCDF variables with posixio assertion failed,
@@ -90,6 +98,9 @@ print "applying masks..."
 deglacage = np.ma.masked_where(~footprint+modernice, deglacage)
 duration = np.ma.masked_where(~footprint, duration)
 erosion = np.ma.masked_where(~footprint, erosion)
+lgmicethk = np.ma.masked_where(lgmicethk<1.0, lgmicethk)
+lgmicesrf = np.ma.masked_where(lgmicethk<1.0, lgmicesrf)
+lgmsrfvel = np.ma.masked_where(lgmicethk<1.0, lgmsrfvel)
 maxicesrf = np.ma.masked_where(~footprint, maxicesrf)
 maxicethk = np.ma.masked_where(~footprint, maxicethk)
 maxsrfage = np.ma.masked_where(~footprint, maxsrfage)
@@ -106,12 +117,15 @@ ut.make_gtif_shp(x, y, deglacage, prefix+'deglacage', range(0, 30001, 1000))
 ut.make_gtif_shp(x, y, duration, prefix+'duration', range(0, 120001, 10000))
 ut.make_gtif_shp(x, y, erosion, prefix+'erosion', [0.1, 1, 10, 100, 1000])
[…]

### 2017-09-12 data/retrieve-external.sh
Add QGIS project file.

@@ -46,6 +46,7 @@ done
 gdalbuildvrt srtm.vrt srtm_??_??.tif
 
 # SRTM UTM 32 entire Alps, 1500x1000 km, 1000 m
+# FIXME: use geotiff format
 gdalwarp -s_srs EPSG:4326 -t_srs EPSG:32632 -r bilinear \
          -te 0 4500000 1500000 5500000 -tr 1000 1000 \
          -srcnodata -32768 -dstnodata -32768 \

### 2017-09-12 figures/alpcyc_hr_lgmvel.py
Add commented prints of max volume and area.
I used these values in the main text.

@@ -49,7 +49,7 @@ vol = nc.variables['slvol'][:]
 nc.close()
 
 # print age of max volume
-#print age[vol.argmax()]
+#print age[vol.argmax()], vol.max()
 
 # plot time series
 tsax=tsax.twinx()

### 2017-09-12 figures/alpcyc_hr_timing.py

@@ -44,7 +44,7 @@ area = nc.variables['area_glacierized'][:]*1e-9
 nc.close()
 
 # print age of max area
-#print age[area.argmax()]
+#print age[area.argmax()], area.max()
 
 # plot time series
 twax = tsax.twinx()

### 2017-09-12 articles/cycle/alps-cycle.tex
Draft discussion of flow patterns.

@@
 
-    \begin{itemize}
-    \item Maximum stages flow pattern is complex (Fig.~\ref{fig:lgmvel}a).
-    \item Flow goes sometimes across topography (Fig.~\ref{fig:lgmvel}a).
-    \item Transfluences (Montgenèvre, col de l'Échelle, Simplon, Albrun, Gries,
-                        (Brünig, Majola, Tagliamento)
-    \end{itemize}
+    The modelled flow pattern at 24.56\,ka is complex (Fig.~\ref{fig:lgmvel}a).
+    The modelled fast-flow regions generally occur along the main river
+    valleys, while ice domes and ice divides are predominantly located over
+    major reliefs . Nevertheless, the model results depict occasional ice flow
+    across mountain passes (transfluences) and ice divide above topographic
+    lows.
+
+    Major transfluences occur for instance across col du Mont-Cenis,
+    Simplonpass, Brünigpass in the western Alps, and over Fernpass,
+    Gailbergsattel and Stranigeralpe in the Eastern Alps
+    (Fig.~\ref{fig:lgmvel}a). Finally, self-sustained ice domes characteristic
+    of ice caps are modelled in two locations over Flüelapass and Ötztal
+    (Fig.~\ref{fig:lgmvel}a).
+
+    In the Eastern Alps, modelled transfluences and crosswise divides are
+    generally incompatible with geological reconstructions, indicating that
+    climate deterioration was overestimated in the model input. In the Western
+    Alps, however, transfluences generally occur where they have been
+    documented by geologic evidence. These model results depict the LGM Alpine
+    ice complex as an intermediate between ice fields and ice caps, bearing
+    characteristics of both and perhaps no modern analogue.
+
+    \todo{look for references, add location names on Fig.~\ref{fig:lgmvel}.}
 

### 2017-09-12 articles/cycle/alps-cycle.tex
Draft discussion of timing of the LGM.

@@
 
-    The modelled flow pattern at 24.56\,ka is complex (Fig.~\ref{fig:lgmvel}a).
+    The modelled flow pattern at 24.57\,ka is complex (Fig.~\ref{fig:lgmvel}a).
     The modelled fast-flow regions generally occur along the main river
@@
 
-    \begin{itemize}
-    \item Maximum extent is time-transgressive (Fig.~\ref{fig:timing}).
-    \item Glaciers have different response times (Fig.~\ref{fig:timing}).
-    \end{itemize}
+    The timing of the LGM has been documented by radiocarbon and cosmogenic
+    isotope dating techniques at multiple glacier lobes around the Alps.
+    These data indicate that Alpine glaciers reached their maximum extent
+    about synchronously with the Laurentide ice sheet around 25\,ka, but
+    glacier lobes stayed in the foreland until as late as 20\,ka, a time
+    around which the Fennoscandian ice sheet reached is maximum, while the
+    Laurentide ice sheet was already in decay.
+
+    In our simulation, the maximum areal cover is reached at 24.57\,ka
+    (Fig.~\ref{fig:lgmvel}a), yet individual glacier lobes reach their
+    maximum extent at different ages (Fig.~\ref{fig:timing}). For instance, the
+    Dora Riparia, Ivrea, and Tagliamento glaciers reach an early maximum before
+    27\,ka, the Durance, Rhone, Inn, Enns, Ticino, and Garda glaciers reach
+    their maximum thickness in phase with the global Alpine areal maximum
+    around 25\,ka, while the Isère, Como and Iseo glaciers reach a late maximum
+    after 24\,ka (Fig.~\ref{fig:timing}). Remarkably, peripheral ice caps on
+    the Vercors, Jura Mountains, Black Forest and Hochschwab reach their
+    maximum extent even later, locally after 21\,ka.
+
+    It is important to notice that these differences in timing of the LGM occur
+    in the model results despite the simple climate forcing used consisting of
+    homogeneous anomalies. These discrepancies between different glacier lobes
+    are thus not related to climate, but are a pure product of modelled
+    inherent glacier dynamics. They occur because of differences in the
+    subglacial topography, in particular catchment sizes and bed roughness.
+
+    PISM is a polythermal ice sheet model, i.e. it computes the distribution
+    of englacial temperatures resulting from heat advection, vertical
+    diffusion in ice and rock, and strain heat release, in three dimensions.
+    For large Alpine glaciers flowing in deep valleys, such as the Rhone,
+    Rhine and Garda glaciers, several thousands of years are needed to attain
+    a thermodynamical equilibrium between cold ice advection from the high
+    accumulation areas, upward diffusion of geothermal heat, and heat release
+    from strong basal shear strain. Temperature evolution is, in part, limited
+    by the slow warming of subglacial bedrock, that has been previously cooled
+    downed by subfreezing air temperature before glacier advance.
+
+    For such glaciers, this thermodynamical equilibrium is typically not yet
+    attained around 25\,ka. This is why several Alpine lobes overshoot their
+    balanced extent before thinning and receding towards the mountains as they
+    warm towards thermodynamical equilibrium.
+
+    Our results indicate that even for a relatively small ice sheet like that
+    found in the Alps, differences in timing on the order of the millenium can
+    result from complex glacier dynamics. Further differences in timing of the
+    LGM caused by heterogeneous climatic anomalies, such as precipitation
+    anomaly gradients, not included in our model set-up, may have
+    counterbalanced or emphasize those presented here.
+
+    \todo{look for references, add location names on Fig.~\ref{fig:lgmvel}.}
 

### 2017-09-13 articles/cycle/alps-cycle.tex
Draft trimlines discussion.

@@
 
-    \begin{itemize}
-    \item Ice surface is well above the trimlines (Fig.~\ref{fig:lgmvel}a).
-    \end{itemize}
+    In several places across the Alps, the maximum elevation reached by the
+    ice surface during the LGM has been reconstructed from trimlines, the upper
+    limit of glacial erosion observed in the landscape. This include the entire
+    western Alps, the Rhone Glacier catchment, the Rhine Glacier catchment,
+    south-eastern Switzerland and all of Austria. These reconstructions are
+    based on the assumption that glacial ice was significantly erosive up to
+    the or almost up to the surface, which was then located at or slightly
+    above the elevation of the trimlines. This assumption is supported by the
+    fact that there exist a regional consistence in trimlines elevations, which
+    are generally located along a surface sloping in the assumed direction of
+    flow. However, in other glaciated mountain regions, including several
+    sites in the Scandinavian Mountains, the Cairgorns in Scotland, and high
+    plateaus in the Canadian Cordillera, that cold-based ice commonly occured
+    at high elevation and preserved the pre-glacial landscape.
+
+    In the upper Rhone valley, the maximum ice surface elevation reached
+    during MIS~2 in our simulation, corrected for bedrock deformation, is
+    consistently modelled to have occured several hundred metres above the
+    observed trimline elevations (Fig.~\ref{fig:trimlines}a, b), with a mean
+    difference of 861\,m (xx\,\%, Fig.~\ref{fig:trimlines}b). This result is
+    somewhat depedent on uncertain basal sliding and ice rheological parameters
+    to which the model sensitivity was not tested here. However, a previous
+    sensitivity study conducted with PISM and a very similar model set-up on
+    the Cordillera ice sheet show that varying ice rheological parameters
+    resulted relative errors of 25\,\unit{\%} with regard to ice rheological
+    parameters and 14\,\unit{\%} regarding basal sliding parameters
+    \citep[Fig.~7]{Seguinot.etal.2016}.
+
+    In addition, our model results depict a polythermal LGM Alpine ice cover.
+    Due to sub-freezing temperatures applied in the climate forcing of the
+    model, the entire Alpine ice field is capped by an upper layer of cold ice.
+    In major Alpine valleys, such as the upper Rhone Valley, important ice
+    thickness and strain heating contribute to form a layer of temperate ice
+    near the glacier base, allowing basal melt, sliding and potentially
+    erosion. On the other hand, on the highest mountains ice cover is to thin
+    and too static to form temperate ice, resulting in cold ice down to the
+    bed, preventing potential erosion.
+
+    In the upper Rhone valley, the geographical location of observed trimlines
+    show remarkable agreement with the areal extent of regions that are
+    modelled to have experienced warm-based for less than 1\,ka (7\,\%) of the
+    MIS~2. Although this result calls for more detailed comparisons across the
+    entire Alpine range and specific sensitivity studies to relevant basal
+    sliding and ice rheological parameters, the presence of an upper layer of
+    cold ice during the LGM, already found in some places in a much warmer
+    climate today, is inevitable. Thus we want to challenge the assumption that
+    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
+    ice cover and call for a more accurate estimation of the thickness of the
+    upper layer cold ice in the Alps.
+
+    \todo{look for references.}
 
@@
         elevation and trimline elevations (500\,m bands). The average
-        difference is 861\,m.}
+        difference is 861\,m.
         \textbf{(c)} Modelled age of maximum ice thickness at the trimline
@@
         cover during MIS~2 (29--14\,ka).
+        \todo{add relative mean difference.}}
       \label{fig:trimlines}

### 2017-09-13 articles/cycle/alps-cycle.tex
Fix two bugs in internal references.

@@
 
-    Although the palaeo-climate forcing was adapted (Sect.~\label{sec:climate})
+    Although the palaeo-climate forcing was adapted (Sect.~\ref{sec:climate})
     to model this maximum configuration in broad agreement with the geological
@@
     from the paleo-precipitation reduction used to force the model
-    (Fig.~\ref{footprints}b), which is likely unrealistic in at least this part
-    of the model domain. On the other hand, the Durance glacier in the
+    (Fig.~\ref{fig:footprints}b), which is likely unrealistic in at least this
+    part of the model domain. On the other hand, the Durance glacier in the
     south-western Alps is modelled to extend several kilometers beyond the

### 2017-09-13 articles/cycle/alps-cycle.tex
Draft discussion of glacial cycle dynamics.

@@
 
-    \begin{itemize}
-    \item Rapid variations of total ice volume (Fig.~\ref{fig:lgmvel}b).
-    \item Two major glaciations during MIS 4 and 2 (Fig.~\ref{fig:lgmvel}b).
-    \item Some glaciers advance many times (Fig.~\ref{fig:profiles}).
-    \end{itemize}
+    As previously mentioned (Sect.~\ref{sec:timeseries}), independently of the
+    palaeo-temperature (GRIP, EPICA, or MD01-2444) and palaeo-precipitation
+    (with or without corrections) applied, all simulations presented here
+    result in a high time variability in the total modelled ice volume
+    (Fig.~\ref{fig:timeseries}). Unsurprisingly, this result is confirmed by
+    the 1-km resolution run, using the optimal EPICA palaeo-temperature record
+    (Sect.~\ref{sec:paltemp}) and conservatively includes palaeo-precipitation
+    reductions (Sect.~\ref{sec:palprec}), which also results in strong total
+    ice volume variability throughout the last glacial cycle
+    (Fig.~\ref{fig:lgmvel}b).
+
+    Two major glaciations occur during MIS~4 and 2 (Fig.~\ref{fig:lgmvel}b).
+    However, several minor glaciations occur during MIS~5 and 3, as well as a
+    minor late-glacial readvance at the onset of MIS~1
+    (Fig.~\ref{fig:lgmvel}b). These episodes are the result of synchronous
+    advances several Alpine glaciers well into the major valleys and sometimes
+    even onto the foreland (Fig.~\ref{fig:profiles}), Supplementary animation).
+
+    For instance the Rhine Glacier (Fig.~\ref{fig:profiles}a) reaches a lenght
+    exceeding 130\,km, roughly corresponding to location of the last major
+    Alpine reliefs and the beginning of the foothills, 6~times during the
+    simulation, sometimes retreating almost to its present day frontal position
+    (Fig.~\ref{fig:profiles}b). The Rhone Glacier, fed by several high-altitude
+    accumulation areas, advanced 8\,times onto the location of modern Lake
+    Geneva (Fig.~\ref{fig:profiles}c, d). The Ivrea Glacier, characterised by
+    a very steep catchment, and mutliple tributaries, show even a higher
+    variability and reaches close to its maximum position 6\,times throughout
+    the simulation. Finally, the Isère Glacier, with its complex system of
+    confluences and diffluences, needs longer time to build up and reaches the
+    foreland only twice during the major glaciations of MIS~4 and 2.
+
+    Despite the low temperature variability of the palaeo-climate forcing, and
+    reduced precipitation dampening glacier response, our simulation depict
+    the Alpine ice complex as highly dynamic, with dynamic of individual
+    glaciers controlled by variations in subglacial topography and catchment
+    geometry.
 

### 2017-09-13 articles/cycle/alps-cycle.tex
Draft conclusions, almost there!

@@
 
+    This study consists in the application of the numerical ice sheet model
+    PISM to ice dynamics of the last glacial cycle in the Alps, using a model
+    set-up based on that previously developed and validated for the Cordilleran
+    ice sheet (Sect.~\ref{sec:model}).
+
+    Using three different palaeo-temperature forcing records (GRIP, EPICA, and
+    MD01-2444, Sect.~\ref{sec:paltemp}), scaled to model the Rhine Glacier
+    piedmont lobe in agreement with the mapped LGM ice margin, and two
+    different palaeo-precipitation parametrizations (with and without
+    precipitation reductions, Sect.~\ref{sec:palprec}), we find that only the
+    EPICA palaeo-temperature record yields model results in agreement with
+    geological findings, in the sense that:
+
     \begin{itemize}
-    \item Summary of the previous stuff.
+      \item The EPICA palaeo-temperature forcing yields maximum glacierized
+            area around 25\,ka, followed by a standstill of major piedmont lobe
+            in the forelands until about 18\,ka, both compatible with much of
+            the dating results, whereas the GRIP forcing results in early
+            deglaciation of the foreland complete by 21\,ka, and the MD01-2444
+            forcing results in a late LGM glaciation around 16\,ka
+            (Sect.~\ref{sec:timeseries}).
+      \item The EPICA palaeo-temperature forcing yields cumulative ice extent
+            compatible with geological evidence during MIS~4 and 2, whereas
+            both GRIP and MD01-2444 forcings result in MIS~4 glaciation well
+            beyong the mapped LGM ice limits  (Sect.~\ref{sec:footprints}).
     \end{itemize}
+
+    We then use this optimal palaeo-temperature forcing and, as a conservative
+    approach, palaeo-precipitation reductions, to force a 1-km resolution
+    simulation of the last glacial cycle in the Alps. A more detailed analysis
+    of its output lead us to the following conclusions.
+
+    \begin{itemize}
+      \item Ice cover is generally underestimated in the Western Alps and
+            overestimated in the Eastern Alps, indicating that east-west
+            gradients in temperature or precipitation change, absent from our
+            model forcing, probably controlled the LGM extent of ice cover in
+            the Alps (Sect.~\ref{sec:extent}).
+      \item The assymetric nature of LGM ice extent north and south of the Alps
+            can be explained by the transient nature of the LGM extent without
+            involving north-south gradients in temperature and precipitation
+            change (Sect.~\ref{sec:extent}).
+      \item The LGM ice flow pattern in the Alps is largely controlled by
+            subglacial topography, but several transfluences and two
+            self-sustained ice domes occured (Sect.~\ref{sec:flow}).
+      \item The modelled LGM is a transient stage in which glaciers are not in
+            equilibrium with climate. Timing potentially varies across the
+            range (Sect.~\ref{sec:timing}).
+      \item Ice thickness during the LGM is modelled to be much higher than in
+            reconstructions. In average modeleld surface elevation is 861\,m
+            above the Rhone Glacier trimlines, which may instead indicate an
+            englacial thermal boundary (Sect.~\ref{sec:thickness}).
+      \item Aline glaciers were potentially very dyamic. They responded quickly
+            to climate fluctuation and some of them potentially advanced many
+            times over the foreland during the last glacial cycle
+            (Sect.~\ref{sec:glaciations})
+    \end{itemize}
+
+    These results are nevertheless limited by the simplicity of the climate
+    forcing and surface mass balance model applied, uncertain ice physical
+    model parameters. Through more detailed sensitivity studies, targeted to
+    specific aspects of the last glacial cycle ice dynamics in the Alps, future
+    modelling studies will certainly be able to quantify uncertainties
+    associated with some of the above statements. However, we also hope that
+    these conclusion will also serve as a basis for future studies of glacial
+    geology in the Alps, and call for a more systematic aggregation and
+    homogeneisation of glacial geological data to form a basis for model
+    validation across the entire Alpine range.
 

### 2017-09-13 articles/cycle/alps-cycle.tex
Add author contributions and acknowledgements.

@@
 % ----------------------------------------------------------------------
+% Acknowledgements
+% ----------------------------------------------------------------------
+
+\subsection*{Author contributions}
+
+    J.~Seguinot designed the study, ran the simulations, and wrote most of
+    the manuscript. All authors contributed to the text.
+
+\subsection*{Acknowledgements}
+
+    This manuscript builds largely upon a previous study on the Cordilleran Ice
+    Sheet which is part of J.~Seguinot's PhD thesis. The experimental design
+    used here, as well as the general layout of this study, are in important
+    proportions the work of Irina Rogozhina, Arjen~P. Stroeven, Martin Margold and
+    Johan Kleman without whom this study would not have happened. As always, we
+    are very thankful to Constantine Khroulev, Ed Bueler, and Andy Aschwanden
+    for providing constant help and development with PISM, in particular with
+    recent issues with the computation of ice temperature (Github issue
+    no.~371) and with the computation of bedrock deformation (Github issues
+    no.~370 and 377). We are equally thankful to Giovanni Monegato and Marc
+    Lüether for very insightful discussions of preliminary model results during
+    the European Geoscience Union General Assembly~2017.
+
+    This work was supported by the Swiss National Foundation grant
+    no.~200021-153179/1 to M.~Funk.
+    Computer resources were provided by the Swiss National Supercomputing
+    Centre (CSCS) allocation no.~s573 to J.~Seguinot.
+
+% ----------------------------------------------------------------------
 % References

### 2017-09-13 articles/cycle/alps-cycle.tex
Use Copernicus class. First draft complete!

@@
-\documentclass{article}
-
-\usepackage{doi}
-\usepackage{amsmath}
-\usepackage{authblk}
-\usepackage{booktabs}
-\usepackage{multirow}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage[pdftex]{xcolor}
-\usepackage[pdftex]{graphicx}
-\usepackage[authoryear,round]{natbib}
+\documentclass[tc, manuscript]{copernicus}
 
@@
 
-\author[1]{Julien Seguinot%
-           \thanks{Correspondence to seguinot@vaw.baug.ethz.ch}}
-\author[1]{Guillaume Jouvet}
-\author[1]{Matthias Huss}
-\author[1]{Martin Funk}
-\author[2]{Frank Preusser}
+\Author[1]{Julien}{Seguinot}
+\Author[1]{Guillaume}{Jouvet}
+\Author[1]{Matthias}{Huss}
+\Author[1]{Martin}{Funk}
+\Author[2]{Frank}{Preusser}
 
@@
 
-% Common units
-\newcommand{\e}[1]{\ensuremath{\times 10^{#1}}}
-\newcommand{\chem}[1]{\ensuremath{\mathrm{#1}}}
-\newcommand{\unit}[1]{\ensuremath{\mathrm{#1}}}
-\newcommand{\degree}[0]{\ensuremath{^{\circ}}}
-\newcommand{\degC}[0]{\unit{{\degree}C}}
+\runningtitle{Modelling last glacial cycle ice dynamics in the Alps}
+\runningauthor{J.~Seguinot et al.}
+\correspondence{J.~Seguinot (seguinot@vaw.baug.ethz.ch)}
+
+\received{}
+\pubdiscuss{}
+\revised{}
+\accepted{}
+\published{}
 
@@
 % ======================================================================
+
+\firstpage{1}
 
@@
 
-% ----------------------------------------------------------------------
-\section{Introduction}
+%\copyrightstatement{}
+
+% ----------------------------------------------------------------------
+\introduction
 \label{sec:intro}
@@
     $R=8.31441$\,\unit{J\,mol^{-1}\,K^{-1}} is the ideal gas constant, and
-    $A_{\mathrm{c}}=2.847\e{-13}$\,\unit{Pa^{-3}\,s^{-1}},
-    $A_{\mathrm{w}}=2.356\e{-2}$\,\unit{Pa^{-3}\,s^{-1}},
-    $Q_{\mathrm{c}}=6.0\e4$\,\unit{J\,mol^{-1}}, and
-    $Q_{\mathrm{w}}=11.5\e4$\,\unit{J\,mol^{-1}} are constant parameters
+    $A_{\mathrm{c}}=2.847 \times 10^{-13}$\,\unit{Pa^{-3}\,s^{-1}},
+    $A_{\mathrm{w}}=2.356 \times 10^{-2}$\,\unit{Pa^{-3}\,s^{-1}},
+    $Q_{\mathrm{c}}=6.0 \times 10^4$\,\unit{J\,mol^{-1}}, and
+    $Q_{\mathrm{w}}=11.5 \times 10^4$\,\unit{J\,mol^{-1}} are constant parameters
     corresponding to values measured below and above a~critical temperature
@@
     density, $\rho_{\mathrm{m}}=3300$\,\unit{kg\,m^{-3}}, and the lithosphere
-    elastic rigidity, $D=1.389\e{24}$\,\unit{N\,m}, were set according to
+    elastic rigidity, $D=1.389 \times 10^{24}$\,\unit{N\,m}, were set according to
     results from glacial isostatic adjustment modelling of deglacial rebound in
@@
 
-    The modelled total ice volume reaches a maximum of 122.5\e{3}\,\unit{km^2},
+    The modelled total ice volume reaches a maximum of $122.5 \times 10^{3}$\,\unit{km^2},
     or 300.4\,mm of sea-level equivalent, at 24.56\,ka
     (Fig.~\ref{fig:lgmvel}b). A maximum glacierized area of
-    163.3\e{3}\,\unit{km^2} is attained shortly aftewards at 24.57\,ka.
+    $163.3 \times 10^{3}$\,\unit{km^2} is attained shortly aftewards at 24.57\,ka.
     (Fig.~\ref{fig:lgmvel}b). Although the modelled timing of the LGM varies
@@
 % ----------------------------------------------------------------------
-\section{Conclusions}
+\conclusions
 % ----------------------------------------------------------------------
@@
 
-\subsection*{Author contributions}
-
+\codedataavailability{%
+    PISM is available as open-source software at http://pism-docs.org/.
+    Please contact the corresponding author for model output data.
+    \idea{Should we make model output available somewhere?}}
+
+\authorcontribution{%
     J.~Seguinot designed the study, ran the simulations, and wrote most of
-    the manuscript. All authors contributed to the text.
-
-\subsection*{Acknowledgements}
+    the manuscript. All authors contributed to the text.}
+
+\competinginterests{%
+    We have no conflict of interest in publishing these results.}
+
+\begin{acknowledgements}
 
@@
 
+\end{acknowledgements}
+
 % ----------------------------------------------------------------------
@@
 
-\bibliographystyle{abbrvnat}
+\bibliographystyle{copernicus}
 \bibliography{../../references/references}
@@
 
-    \begin{figure}
+    \begin{figure*}
       \centerline{\includegraphics{alpcyc_hr_inputs}}
@@
       \label{fig:inputs}
-    \end{figure}
-
-    \begin{figure}
+    \end{figure*}
+
+    \begin{figure*}
       \centerline{\includegraphics{alpcyc_lr_timeseries}}
@@
       \label{fig:timeseries}
-    \end{figure}
-
-    \begin{figure}
+    \end{figure*}
+
+    \begin{figure*}
       \centerline{\includegraphics{alpcyc_lr_footprints}}
@@
       \label{fig:footprints}
-    \end{figure}
-
-    \begin{figure}
+    \end{figure*}
+
+    \begin{figure*}
       \centerline{\includegraphics{alpcyc_hr_lgmvel}}
@@
       \label{fig:lgmvel}
-    \end{figure}
-
-    \begin{figure}
+    \end{figure*}
+
+    \begin{figure*}
       \centerline{\includegraphics{alpcyc_hr_timing}}
@@
       \label{fig:timing}
-    \end{figure}
+    \end{figure*}
 
@@
 
-    \begin{figure}
+    \begin{figure*}
       \centerline{\includegraphics{alpcyc_hr_profiles}}
@@
       \label{fig:profiles}
-    \end{figure}
+    \end{figure*}
 
@@
       {\begin{tabular}{llrll}
-        \toprule
+        \tophline
 
@@
 
-        \midrule
+        \middlehline
         \multicolumn{2}{l}{{Ice rheology}} \\
-        \midrule
+        \middlehline
 
@@
         $A_{\mathrm{c}}$   & Ice hardness coefficient cold
-                & $2.847\e{-13}$
+                & $2.847 \times 10^{-13}$
                 & \unit{Pa^{-3}\,s^{-1}}
@@
         $A_{\mathrm{w}}$   & Ice hardness coefficient warm
-                & $2.356\e{-2}$
+                & $2.356 \times 10^{-2}$
                 & \unit{Pa^{-3}\,s^{-1}}
@@
         $Q_{\mathrm{c}}$   & Flow law activation energy cold
-                & $6.0\e4$
+                & $6.0 \times 10^4$
                 & \unit{J\,mol^{-1}}
@@
         $Q_{\mathrm{w}}$   & Flow law activation energy warm
-                & $11.5\e4$
+                & $11.5 \times 10^4$
                 & \unit{J\,mol^{-1}}
@@
         $\beta$ & Clapeyron constant
-                & $7.9\e{-8}$
+                & $7.9 \times 10^{-8}$
                 & \unit{K\,Pa^{-1}}
@@
         $L$     & Water latent heat of fusion
-                & $3.34\e5$
+                & $3.34\times10^5$
                 & \unit{J\,kg^{-1}\,K^{-1}}
@@
 
-        \midrule
+        \middlehline
         \multicolumn{2}{l}{{Basal sliding}} \\
-        \midrule
+        \middlehline
 
@@
 
-        \midrule
+        \middlehline
         \multicolumn{2}{l}{{Bedrock and lithosphere}} \\
-        \midrule
+        \middlehline
 
@@
         $\nu_{\mathrm{m}}$ & Astenosphere viscosity
-                & $2.2\e{20}$
+                & $2.2 \times 10^{20}$
                 & \unit{Pa\,s}
@@
         $D$     & Lithosphere flexural rigidity
-                & $1.389\e{24}$
+                & $1.389 \times 10^{24}$
                 & \unit{N\,m}
@@
 
-        \midrule
+        \middlehline
         \multicolumn{2}{l}{{Surface and atmosphere}} \\
-        \midrule
+        \middlehline
 
@@
         $F_{\mathrm{s}}$   & Degree-day factor for snow
-                & $3.297\e{-3}$
+                & $3.297 \times 10^{-3}$
                 & \unit{m\,K^{-1}\,day^{-1}}
@@
         $F_{\mathrm{i}}$   & Degree-day factor for ice
-                & $8.791\e{-3}$
+                & $8.791 \times 10^{-3}$
                 & \unit{m\,K^{-1}\,day^{-1}}
@@
         $\gamma$& Air temperature lapse rate
-                & $6\e{-3}$
+                & $6 \times 10^{-3}$
                 & \unit{K\,m{-1}}
@@
 
-        \bottomrule
+        \bottomhline
       \end{tabular}}
@@
       {\begin{tabular}{lccccccl}
-        \toprule
+        \tophline
 
@@
 
-        \midrule
+        \middlehline
 
@@
       %{\begin{tabular}{llrlrlr}
-      %  \toprule
+      %  \tophline
       %
@@
       %
-      %  \midrule
+      %  \middlehline
       %
@@
       %
-      %  \bottomrule
+      %  \bottomhline
       %\end{tabular}}

### 2017-09-15 articles/cycle/alps-cycle.tex
Add references on ice extent.

@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    The LGM extent of alpine glaciers has been mapped with varying level of
+    across the mountain range \citep{Penck.Bruckner.1909, Jackli.1962,
+    Bini.etal.2009, Coutterand.2010, Bavec.Verbic.2011,
+    Buoncristiani.Campy.2011, Husen.2011}. These maps have been compiled
+    into an reconstruction covering the entire Alps \citep{Ehlers.etal.2011},
+    reproduced here (Fig.~4, red line) for comparison against model results.
 

### 2017-09-15 articles/cycle/alps-cycle.tex
Add references on flow patterns.

@@
 
+    The LGM Alpine ice flow pattern was traditionally described as that of a
+    network of interconnected valley glaciers that was primarily controlled by
+    subglacial topography. However, the geomorphology shows that ice was thick
+    enough to flow accross high mountain passes throughout the mountain range
+    \citep[e.g.,][]{Onde.1938, Penck.Bruckner.1909, Jackli.1962, Husen.1985,
+    Coutterand.2010, Kelly.etal.2004, Husen.2011}, and even perhaps form
+    self-sustained ice domes \citep{Florineth.1998, Florineth.Schluchter.1998,
+    Kelly.etal.2004, Bini.etal.2009}.
+
     The modelled flow pattern at 24.57\,ka is complex (Fig.~\ref{fig:lgmvel}a).
@@
     (Fig.~\ref{fig:lgmvel}a).
+
+    Transfluence across col de Montgenèvre was previously questioned
+    \citep[Fig.~2]{Cossart.etal.2012}. Transfluence across col de Mont Cenis
+    was previoysly described \citep[Fig.~3.18]{Onde.1938, Coutterand.2010}.
+    Transfluences across Fernpass and Seefeld are known from the geomorphology
+    \citep[Fig.~2.4]{Penck.Bruckner.1909, Husen.2011}. Transfluence across
+    Brunigpass was already marked in the maps \citep{Jackli.1962}. There is
+    numerous evidence for transfluence across Simplon Pass
+    \citep{Kelly.etal.2004}. Ice crossed the Pyhrn Pass \citep{Husen.2004}. Ice
+    flew across the Gailberg and Kreuzberg saddles \citep{Husen.1985}. The
+    Tagliamento catchment is usually assumed to not have received tranfluences
+    from the Drau catchment \citep{Monegato.etal.2007}, but this would not be
+    incompatible with reconstructed ice surface elevations in this area
+    \citep{Husen.1987}.
 
@@
 
-    \todo{look for references, add location names on Fig.~\ref{fig:lgmvel}.}
+    \todo{add location names on Fig.~\ref{fig:lgmvel}.}
+
+%    \citep{Jouvet.etal.2017}
 

### 2017-09-18 articles/cycle/alps-cycle.tex
Add references on timing.

@@
     The timing of the LGM has been documented by radiocarbon and cosmogenic
-    isotope dating techniques at multiple glacier lobes around the Alps.
-    These data indicate that Alpine glaciers reached their maximum extent
-    about synchronously with the Laurentide ice sheet around 25\,ka, but
-    glacier lobes stayed in the foreland until as late as 20\,ka, a time
-    around which the Fennoscandian ice sheet reached is maximum, while the
-    Laurentide ice sheet was already in decay.
+    isotope dating techniques at multiple locations around the Alps (cf.
+    \citealp[Fig.~5]{Wirsig.etal.2016} for a review and the more recent
+    publications by \citealp{Monegato.etal.2017, Federici.etal.2017}). These
+    data indicate that Alpine glaciers reached their maximum extent between 20
+    and 26\,ka (calibrated \chem{^{14}C} or \chem{^{10}Be} ages), but glacier
+    lobes stayed in the foreland until as late as 22 to 17\,ka, when they
+    experienced rapid retreat synchronously to lowering of the ice surface in
+    the mountains (\citealp[Fig.~5]{Wirsig.etal.2016},
+    \citealp[Fig.~3]{Monegato.etal.2017}).
+
+    \note{Please edit or complete with more references if needed}
 
@@
     the Vercors, Jura Mountains, Black Forest and Hochschwab reach their
-    maximum extent even later, locally after 21\,ka.
+    maximum extent even later, locally after 21\,ka (Fig.~\ref{fig:timing}).
 
@@
 
-    \todo{look for references, add location names on Fig.~\ref{fig:lgmvel}.}
+    \todo{add location names on Fig.~\ref{fig:lgmvel}.}
 

### 2017-09-18 articles/cycle/alps-cycle.tex
Add references on trimlines.

@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    Following the general assumption that trimlines, which mark the transition
+    between the steep frost-shattered ridges and the more gentle glacially eroded
+    valley troughs, represent the maximum elevation of the LGM ice surface,
+    the LGM ice thickness in the Alps been reconstructed in several areas
+    \citep{Husen.1987, Florineth.1998, Florineth.Schluchter.1998,
+    Kelly.etal.2004, Bini.etal.2009, Coutterand.2010, Cossart.etal.2012}.
+    However, this assumption is challenged by geological evidence from
+    Scandinavia \citep[e.g.,]{Kleman.1994, Kleman.Borgstrom.1994,
+    Stroeven.etal.2002}, the British Isles \citep[e.g.,]{Fabel.etal.2012}, and
+    North America \citep[e.g.,]{Kleman.etal.2010}, that pre-glacial landscapes
+    located well above the trimlines have been glaciated and preserved under
+    under cold-based ice, sometimes for several glacial cycles.
 

### 2017-09-18 articles/cycle/alps-cycle.tex
Add references on glacial cycle dynamics.

@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    Glacial history of the Alps prior to the LGM remains poorly constrained.
+    Although the four glaciations model \citep{Penck.Bruckner.1909} has long
+    been used, it is now known that glaciers advanced onto the foreland many
+    more times \citep{Schluchter.1991, Preusser.etal.2011}. Sparse geological
+    data indicate that the last glacial cycle may have comprised two or even
+    three periods of glacier growth and decay \citep{Preusser.2004,Ivy-Ochs.etal.2008}.
+    Luminescence dating from two sites in the central northern foreland
+    indicate an early last glacial advance of Alpine glaciers onto the near
+    foreland around $107\pm9$ to $101\pm5$\,ka \citep{Preusser.etal.2003,
+    Preusser.Schluchter.2004}. Evidence for a MIS~4 glaciation is equally
+    sparse, and its timing remain uncertain \citep[e.g.,]{Preusser.etal.2003,
+    Link.Preusser.2006}.
 

### 2017-09-19 articles/cycle/alps-cycle.tex
Begin to rephrase, starting with abstract.

@@
     offsets from multiple paleo-climate proxies, among which only the EPICA ice
-    core record yields glacial extent during marine oxygen isotope stages~4
-    (69--62\,ka) and~2 (34--18\,ka) in agreement to geological reconstructions.
+    core record yields glaciation during marine oxygen isotope stages~4
+    (69--62\,ka) and~2 (34--18\,ka) spatially and temporally consistent with
+    the geological reconstructions.
 
@@
     may have advanced many times over the foreland during the last glacial
-    cycle. Cumulative basal sliding, a proxy for glacial erosion, is modelled
-    to be highest in the deep valleys of the western Alps. Finally, the Last
+    cycle. Ice flow patterns during peak glaciation are largely governed by
+    subglacial topography but include occasional transfluences and
+    self-sustained ice domes. Finally, the Last
     Glacial Maximum advance, often considered synchronous, is here modelled as

### 2017-09-19 articles/cycle/alps-cycle.tex
Rephrase introduction and fill logical gaps.
The five bullet points (extent, flow, timing, thickness, dynamics) should form
the red line of the paper. They correspond to subsections in the discussion,
and bullet points in the conclusions.

@@
 
-    Glaciers move by a combination of meltwater-induced sliding at the base
-    \citep[\S532]{Saussure.1779}, and viscous deformation within the ice body
-    \citep{Forbes.1846b}. As glaciers flow and slide across their bed, they
-    transport rock debris and erode the landscape, thereby leaving
-    geomorphological traces of their former presence.
-
     For nearly 300~years, montane people and early explorers of the European
@@
     landscape, and to understand that glaciers had once been more extensive
-    than today \citep[e.g.,][p.~21]{Windham.Martel.1744}. In the mid-nineteenth
+    than today \citep[e.g.,][p.~21]{Windham.Martel.1744}. Contemporaneously,
+    it was also observed in the Alps that glaciers move by a combination of
+    meltwater-induced \emph{sliding} at the base \citep[\S532]{Saussure.1779},
+    and viscous \emph{deformation} within the ice body \citep{Forbes.1846b}. As
+    glaciers flow and slide across their bed, they transport rock debris and
+    erode the landscape, thereby leaving geomorphological traces of their
+    former presence. In the mid-nineteenth
     century, more systematic studies of glacial features showed that alpine
@@
     proposed European and North American ice sheets. Although it was long
-    unclear whether there had been a single or multiple glaciations, this
-    controversy ended with the large-scale mapping of two distinct moraine
+    unclear whether there had been a single or multiple \emph{glaciations},
+    this controversy ended with the large-scale mapping of two distinct moraine
     systems in North America \citep{Chamberlin.1894}. In the European Alps, the
     systematic classification of the glaciofluvial terraces on the northern
-    foreland indicated that landscape, that there had been at least four major
+    foreland indicated that there had been at least four major
     glaciations in the Alps \citep{Penck.Bruckner.1909}. More recent studies
@@
     Shackleton.Opdyke.1973, Dansgaard.etal.1993, Augustin.etal.2004},
-    indicating neither one, nor four, but rather tens of glacial cycles. During
-    the last 800000 years (800\,ka), glacial and interglacial periods have
-    succeeded each other with a 100\,ka periodicity \citep{Hays.etal.1976,
-    Augustin.etal.2004}. However, much of the traces left by glaciers on the
-    landscape date from the most recent of these cycles, the \emph{last glacial
-    cycle}.
-
-    Nevertheless, the landform record is sparse in time and, most often,
+    indicating neither one, nor four, but rather tens of \emph{glacial cycles}.
+    During the last 800000 years (800\,ka), glacial and interglacial periods
+    have succeeded each other with a 100\,ka periodicity \citep{Hays.etal.1976,
+    Augustin.etal.2004}. Nevertheless, this global signal is largely governed
+    by the North American and Eurasian ice sheet complexes. It is thus unclear
+    wether glacier advances and retreats in the Alps were in pace with global
+    sea-level fluctuations.
+
+    Besides, the landform record is typically sparse in time and, most often,
     spatially incomplete. Palaeo-ice sheets did not leave a continuous imprint
@@
     subsequent glacier re-advances and other geomorphological processes
-    \citep{Kleman.1994, Kleman.etal.2006, Kleman.etal.2010}. The landscape
-    imprint of previous glaciations has been largely overprinted by the more
-    recent ones; therefore most of the glacial features currently left on the
-    foreland present a record of the last major glaciation of the Alps, dating
-    from the Last Glacial Maximum (LGM).
-
-    Because this global signal is largely governed by the North American and
-    Eurasian ice sheet complexes, it is unclear wether glacier advances and
-    retreats in the Alps were in pac(71--57\,ka)e with global sea-level
-    fluctuations. The timing of the LGM in the Alps (Ivy-Ochs et al., 2008;
-    Monegato et al., 2007) is in good agreement with the maximum expansion of
-    continental ice sheets recorded by the Marine Oxygen Isotope Stage (MIS) 2
-    between 29 000 and 14 000 years before present (BP) (Lisiecki and Raymo,
-    2005). There are nonetheless regional variations between different piedmont
-    lobes, that are of the same order of magnitude with the dating
-    uncertainties. Because recent glaciation overprinted older ones, and
-    because dating uncertainties typically increase with age, dated
-    reconstructions are strongly biased towards earlier glaciations
-    \citep{Heyman.etal.2011}, and thus much of the glacial records dates to the
-    Last Glacial Maximum (LGM). The Alps have been studied in more detail than
-    any other glacial region. There is evidence for glaciation during MIS 4
-    (71--57\,ka) but potential foreland glaciations during the rest of the last
-    glacial cycle are still discussed.
+    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2006, Kleman.etal.2010}. Dating
+    uncertainties typically increase with age, such that dated reconstructions
+    are strongly biased towards earlier glaciations \citep{Heyman.etal.2011}.
+    The European Alps have been studied in more detail than any other glacial
+    region, but they are no exception to these rules. Although sparse
+    geological traces indicate that the last glacial cycle may have comprised
+    two or three cycles of glacier growth and decay \citep{Preusser.2004,
+    Ivy-Ochs.etal.2008}, most glacial features currently left on the foreland
+    present a record of the last major glaciation of the Alps, dating from the
+    \emph{Last Glacial Maximum} (LGM).
+
+    The spatial extent and thickness of Alpine glaciers during the LGM, an
+    integrated footprint of thousands of years of climate history and glacier
+    dynamics, have been reconstructed from moraines and trimlines across the
+    mountain range \citep[e.g.,][]{Bini.etal.2009, Coutterand.2010,
+    Husen.2011}. Although no evidence for cold-based glaciation has been found
+    in the Alps to date, the assumption that trimlines, the upper limit of
+    glacial erosion, represent the maximum ice surface elevation, which has
+    been repeatedly invalidated in other glaciated regions of the globe
+    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012}. Ice flow
+    patterns were primarily controlled by subglacial topography, but there is
+    evidence for flow accross major mountain passes
+    \citep[e.g.,][]{Coutterand.2010, Kelly.etal.2004, Husen.2011}, and
+    self-sustained ice domes \citep{Bini.etal.2009}. Finally, the timing of the
+    LGM in the Alps \citep{Ivy-Ochs.etal.2008, Monegato.etal.2017} is in good
+    agreement with the maximum expansion of continental ice sheets recorded by
+    the Marine Oxygen Isotope Stage (MIS) 2 (29--14\,ka)
+    \citep{Lisiecki.Raymo.2005}. Regional variation between different piedmont
+    lobes exist \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear wether they
+    relate to climate or glacier dynamics \citep{Monegato.etal.2017} or
+    uncertainties in the dating methods.
+
+    Although the glacial history of the European Alps has been studied for
+    nearly three hundred years, it thus remains incompletely known
+    \begin{itemize}
+      \item what climate evolution lead to the known maximum ice limits,
+      \item to which extent ice flow was controlled by subglacial topography,
+      \item what drove the different response of the individual lobes,
+      \item how far above the trimline was the ice surface located, and
+      \item how many advances occurred during the last glacial cycle.
+    \end{itemize}
 
@@
     a period for which palaeo-temperature proxies are available, albeit non
-    regional. We test the model sensitivity to multiple palaeo-climate forcing
+    regional. In an attempt to analyse the long-standing questions outline
+    above, we test the model sensitivity to multiple palaeo-climate forcing
     (Sect.~\ref{sec:climate}), and then explore the modelled glacier dynamics
     at high resolution for the optimal forcing (Sect.~\ref{sec:results}).
-
-    \idea{Matthias: I think it is very important to emphasize the knowledge gap
-          that you fill in with your work in the Intro. (1) it is incompletely
-          known how many advances occurred during the Last glacial cycle, (2)
-          what drove the different response of the individual lobes (you show
-          that it is glacier response time, i.e. ice dynamics), (3) how the
-          temporal dynamics of subglacial erosion was structured.}
 
@@
     The LGM extent of alpine glaciers has been mapped with varying level of
-    across the mountain range \citep{Penck.Bruckner.1909, Jackli.1962,
+    detail across the mountain range \citep{Penck.Bruckner.1909, Jackli.1962,
     Bini.etal.2009, Coutterand.2010, Bavec.Verbic.2011,
@@
     However, this assumption is challenged by geological evidence from
-    Scandinavia \citep[e.g.,]{Kleman.1994, Kleman.Borgstrom.1994,
-    Stroeven.etal.2002}, the British Isles \citep[e.g.,]{Fabel.etal.2012}, and
-    North America \citep[e.g.,]{Kleman.etal.2010}, that pre-glacial landscapes
+    Scandinavia \citep[e.g.,][]{Kleman.1994, Kleman.Borgstrom.1994,
+    Stroeven.etal.2002}, the British Isles \citep[e.g.,][]{Fabel.etal.2012}, and
+    North America \citep[e.g.,][]{Kleman.etal.2010}, that pre-glacial landscapes
     located well above the trimlines have been glaciated and preserved under
@@
     Preusser.Schluchter.2004}. Evidence for a MIS~4 glaciation is equally
-    sparse, and its timing remain uncertain \citep[e.g.,]{Preusser.etal.2003,
+    sparse, and its timing remain uncertain \citep[e.g.,][]{Preusser.etal.2003,
     Link.Preusser.2006}.

### 2017-09-19 articles/cycle/alps-cycle.tex
Wrap lists of figs and tables in temporary notes.

@@
 
+    \note{This section contains\\
         Table~\ref{tab:params} -- Model parameters.\\
-    Fig.~\ref{fig:inputs} -- Climate and geothermal model inputs.\\
+        Fig.~\ref{fig:inputs} -- Climate and geothermal model inputs.}
 
@@
 
+    \note{This section contains\\
       Table~\ref{tab:records} -- Palaeo-climate records.\\
       Fig.~\ref{fig:timeseries} -- Low-resolution time series.\\
-    Fig.~\ref{fig:footprints} -- Low-resolution ice cover.\\
+      Fig.~\ref{fig:footprints} -- Low-resolution ice cover.}
 
@@
 
+    \note{This section contains\\
         Fig.~\ref{fig:lgmvel} -- Snapshot at 21\,ka and volume time series.\\
@@
         Fig.~\ref{fig:trimlines} -- LGM ice thickness compared to trimlines.\\
-    Fig.~\ref{fig:profiles} -- Individual glacier extent profiles.
+        Fig.~\ref{fig:profiles} -- Individual glacier extent profiles.}
 

### 2017-09-20 articles/cycle/alps-cycle.tex
Rephrase methods, mention erroneous rigidity.

@@
     states of the ice sheet, and the associated lithospheric response. The
-    set-up used here is largely based on that used in previous studies of the
-    former Cordilleran ice sheet \citep{Seguinot.2014, Seguinot.etal.2014,
-    Seguinot.etal.2016}.
-
-    \idea{Remove next paragraph?}
+    model set-up used here was previously developed and tested on the former
+    Cordilleran ice sheet \citep{Seguinot.2014, Seguinot.etal.2014,
+    Seguinot.etal.2016} and subsequently adapted for steady climate
+    \citep{Becker.etal.2016} and regional \citep{Jouvet.etal.2017a}
+    applications to the European Alps.
 
@@
     Centre for Medium-Range Weather Forecasts Reanalysis Interim
-    \citep[ERA-Interim;][]{Dee.etal.2011}, perturbed by time-dependent
-    temperature offsets, temperature lapse-rate corrections, and in some cases,
-    time-dependent paleo-precipitation reductions (Sect.~\ref{sec:atm},
-    Table~\ref{tab:records}).
-
-    \idea{Remove next paragraph?}
+    \citep[ERA-Interim;][]{Dee.etal.2011}, perturbed by temperature lapse-rate
+    corrections (Sect.~\ref{sec:atm}), time-dependent temperature offsets, and
+    in some cases, time-dependent paleo-precipitation reductions
+    (Sect.~\ref{sec:climate}).
 
@@
     entire Alpine range (Fig.~\ref{fig:inputs}). The simulations were run on
-    two distinct grids, using a~lower horizontal resolution of 2\,\unit{km},
-    and a~higher horizontal resolution of 1\,\unit{km}.
-
-    \idea{In the following, I repeat all parameters in the text and table. Is
-          this necessary? Equations are the same as in my Cordillera paper.
-          Should I remove them?}
+    two distinct grids, using horizontal resolutions of 2 and 1\,\unit{km}.
+    All parameter values are summarized in Table~\ref{tab:params}.
 
@@
     where gravitational stresses intervene both in the SIA and SSA velocity
-    computation, this hybrid scheme is computationally much more efficient than
-    a fully three-dimensional model with which the simulations presented here
-    would not be feasible.
+    computation, it was shown to effectively approximate complex ice field flow
+    dynamics over mountainous topographies similar to that of the European Alps
+    \citep{Golledge.etal.2012, Ziemen.etal.2016}. Beside, this hybrid scheme is
+    computationally much more efficient than a fully three-dimensional model
+    with which the simulations of the scale presented here would not be
+    feasible.
 
@@
     defined in our case by
-    ${\tau_{\mathrm{e}}}^2=\frac{1}{2}\mathrm{tr}(\vec{\tau}^2)$. The ice
+    ${\tau_{\mathrm{e}}}^2=\frac{1}{2}\mathrm{tr}(\vec{\tau}^2)$. We set the
+    power-law exponent, $n=3$, according to
+    \citet[p.~55--57]{Cuffey.Paterson.2010}. The ice
     softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$,
@@
     threshold $T_{\mathrm{c}}=-10$\,\unit{{\degree}C}
-    \citep[p.~72]{Cuffey.Paterson.2010}. The water fraction, $\omega$, is
+    \citep[p.~72]{Cuffey.Paterson.2010}. We set the Clapeyron constant,
+    $\beta=7.9\times 10^{-8}$\,\unit{K\,Pa^{-1}}, according to
+    \citet{Luthi.etal.2002}, and the water fraction coefficient, $f=181.25$,
+    according to \citet{Lliboutry.Duval.1985}. The water fraction, $\omega$, is
     capped at a~maximum value of 0.01, above which no measurements are
     available \citep[Eq.~5.7]{Lliboutry.Duval.1985, Greve.1997}. Finally, $E$
-    is a~non-dimensional enhancement factor which can take different values,
+    is a~non-dimensional enhancement factor which take the values,
     $E_{\text{SIA}}=2$, in the SIA and $E_{\text{SSA}}=1$, in the SSA, as
@@
 
-    In all our simulations, we set constant the power-law exponent, $n=3$,
-    according to \citet[p.~55--57]{Cuffey.Paterson.2010}, the Clapeyron
-    constant, $\beta=7.9\times 10^{-8}$\,\unit{K\,Pa^{-1}}, according to
-    \citet{Luthi.etal.2002}, and the water fraction coefficient, $f=181.25$,
-    according to \citet{Lliboutry.Duval.1985}. These fixed parameter values are
-    summarized in Table~\ref{tab:params}.
-
     Surface air temperature derived from the climate forcing
-    (Sect.~\ref{sec:atm}) provides the upper boundary condition to the ice
-    enthalpy model. Temperature is computed in the ice and in the bedrock to
-    a~depth of 3\,\unit{km} below the ice-bedrock interface, where it is
+    (Sects.~\ref{sec:atm}, ~\ref{sec:paltemp}) provides the upper boundary
+    condition to the ice
+    enthalpy model. Temperature is computed in the ice and in the bedrock down
+    to a~depth of 3\,\unit{km} below the glacier base, where it is
     conditioned by a~lower boundary geothermal heat flux estimate from multiple
-    geothermal proxies \citep[similarity method]{Goutorbe.etal.2011}. In the
-    2\,km-resolution simulations, the vertical grid consists of 31~temperature
-    layers in the bedrock and up to 126~enthalpy layers in the ice,
-    corresponding to vertical resolutions of 100 and 40\,\unit{m},
-    respectively. The 1\,km-resolution simulation uses 61~bedrock layers and up
-    to 251~ice layers with a~vertical resolutions of 50 and 20\,\unit{m}.
+    geothermal proxies \citep[similarity method]{Goutorbe.etal.2011}.
 
@@
     $v_{\text{th}}=100$\,\unit{m\,a^{-1}}, are set to values successfully used
-    to model the Greenland ice sheet \citet{Aschwanden.etal.2013} The yield
+    to model the Greenland ice sheet \citep{Aschwanden.etal.2013}. The yield
     stress, $\tau_{\mathrm{c}}$, is modelled using the Mohr--Coulomb criterion,
@@
     %
-    where the till cohesion, $c_0=0$, was consistently measured to be
+    where the till cohesion, $c_0=0$, was oftentimes measured to be
     negligible \citep[p.~268]{Tulaczyk.etal.2000, Cuffey.Paterson.2010}. We use
@@
     varies from zero to $W_{\text{max}}=2$\,m, a~threshold above which
-    additional melt water is assumed to drain off instantaneously. These fixed
-    parameter values are summarized in Table~\ref{tab:params}.
+    additional melt water is assumed to drain off instantaneously.
 
@@
     the Alps most closely reproducing observed modern uplift rates
-    \citep[Table~\ref{tab:params};][Supplementary Fig.~7]{Mey.etal.2016}. The
-    latter was computed as $D=YE^3/12(1-\nu)$, where $Y=100$\,GPa is Young's
-    modulus, $\nu=0.25$ is the Poisson ratio, and $E=50$\,km is the average
-    effective elastic thickness of the lithosphere
-    \citep[Table~\ref{tab:params};][]{Mey.etal.2016}.
+    \citep[Supplementary Fig.~7]{Mey.etal.2016}. Unfortunately, the latter was
+    computed using the erroneous formula, ${D=YE^3/12(1-\nu)}$
+    \citep{Mey.etal.2016}. The correct formula is ${D=YE^3/12(1-\nu^2)}$.
+    Using the Young's modulus, $Y=100$\,GPa, and the Poisson ratio, $\nu=0.25$
+    \citep{Mey.etal.2016}, our value for lithosphere rigidity thus correspond
+    to an effective elastic thickness of the lithosphere of $E=52.415$\,km,
+    instead of the 50\,km value used by \citet{Mey.etal.2016}.
+
+    \note{This is embarrassing, but I realised too late that the formula in
+          \citet{Mey.etal.2016} was wrong.}
 
@@
     The PDD computation accounts for stochastic temperature variations by
-    assuming a~normal temperature distribution of standard deviation $\sigma$
-    around the expected value $T_{\mathrm{m}}$. It is expressed by an
+    assuming a~normal temperature distribution of standard deviation, $\sigma$,
+    around the expected value, $T_{\mathrm{m}}$. It is expressed by an
     error-function formulation \citep{Calov.Greve.2005},
@@
     near-surface air temperature and precipitation rate fields from
-    \citep[WorldClim;][]{Hijmans.etal.2005}, representative of the period 1960
-    to 1990. Modern climate of the European Alps is characterised by a
-    latitudinal gradient in summer air temperatures, and a longitudinal
+    WorldClim \citep{Hijmans.etal.2005}, representative of the period~1960
+    to~1990. Modern climate of the European Alps is characterised by a
+    north-south gradient in summer air temperatures, and an east-west
     gradient in winter precipitation (Fig.~\ref{fig:inputs}). WorldClim data
     were selected as an input to the ice sheet model because they incorporate
-    observations from the dense weather station network of central Europe.
-    Besides, last glacial cycle alpine glaciers did not extend over marine
-    areas where WorldClim data is missing. Finally, WorldClim data were
+    observations from the dense weather station network of central Europe
+    \citep[Fig.~1]{Hijmans.etal.2005}. Besides, WorldClim data were
     previously used as climate forcing for PISM to model the LGM extent of the
@@
     along the southern margin \citep{Seguinot.etal.2014} were weather station
-    density is lower than in the Alps.
+    density is lower than in the Alps. Finally, the last glacial cycle alpine
+    glaciers did not extend over marine areas where WorldClim data is missing.
+    Although low station density in the Arctic and missing data on marine
+    surfaces were major setbacks against accurate representation of Cordilleran
+    climate \citep{Seguinot.etal.2014}, these concerns are irrelevant to the
+    European Alps.
 
@@
     computed as a~function of ice surface elevation, $s$, using the SRTM
-    topography shipped with WorldClimas a~reference, $b_{\text{ref}}$:
+    topography shipped with WorldClim as a~reference, $b_{\text{ref}}$:
     %

### 2017-09-20 figures/alpcyc_lr_footprints.py
Move scaling domain plot function to util.

@@ -19,13 +19,6 @@ for ax in grid.flat:
     ax.set_extent(ut.pl.regions['alps'], crs=ax.projection)
     ax.set_rasterization_zorder(2.5)
 
-# add Rhine lobe scaling domain  # FIXME move to util
-w, e, s, n = ut.pl.regions['rhlobe']
-x = [w, e, e, w, w]
-y = [s, s, n, n, s]
-for ax in grid[0]:
-    ax.plot(x, y, c='k', lw=0.5)
-
 # for each record
 for i, rec in enumerate(ut.alpcyc_records):
     label = ut.alpcyc_clabels[i]
@@ -38,6 +31,7 @@ for i, rec in enumerate(ut.alpcyc_records):
         grid[0, i/2].set_title(rec, fontweight='bold', color=c)
         ut.pl.add_subfig_label('(%s)' % list('abc')[i/2], ax=grid[0, i/2])
         ut.pl.add_subfig_label('(%s)' % list('def')[i/2], ax=grid[1, i/2])
+        ut.pl.draw_scaling_domain(grid[0, i/2])
 
     # load extra output
     dtfile = '%s3222cool%04d' % (rec.replace('-', '').lower(), round(dt*100))

### 2017-09-20 figures/util/pl.py

@@ -368,6 +368,13 @@ def draw_boot_topo(ax=None, res='1km'):
     return im
 
 
+def draw_scaling_domain(ax=None):
+    """Add Rhine lobe scaling domain."""
+    w, e, s, n = ut.pl.regions['rhlobe']
+    x = [w, e, e, w, w]
+    y = [s, s, n, n, s]
+    ax.plot(x, y, c='k', lw=0.5)
+
 # Timeseries elements
 # -------------------
 

### 2017-09-20 articles/cycle/alps-cycle.tex
Reorder and reference spatial inputs figure.

@@
     conditioned by a~lower boundary geothermal heat flux estimate from multiple
-    geothermal proxies \citep[similarity method]{Goutorbe.etal.2011}.
+    geothermal proxies (\citealp[similarity method]{Goutorbe.etal.2011};
+    Fig.~\ref{fig:inputs}a).
 
@@
     seasonal cycle \citep{Seguinot.2013}, and bilinearly interpolated to the
-    model grids (Fig.~\ref{fig:inputs}). Degree-day factors for snow and ice
+    model grids (Fig.~\ref{fig:inputs}b). Degree-day factors for snow and ice
     melt are set to values used in the European Ice Sheet Modelling INiTiative
@@
     to~1990. Modern climate of the European Alps is characterised by a
-    north-south gradient in summer air temperatures, and an east-west
-    gradient in winter precipitation (Fig.~\ref{fig:inputs}). WorldClim data
+    north-south gradient in summer air temperatures (Fig.~\ref{fig:inputs}c),
+    and an east-west gradient in winter precipitation (Fig.~\ref{fig:inputs}d).
+    WorldClim data
     were selected as an input to the ice sheet model because they incorporate
@@
     setup used in the sensitivity tests, palaeo-temperature anomalies are
-    scaled linearly (Table~\ref{tab:records}) so that the cumulative glaciated
-    area of the Rhine glacier piedmont lobe during Oxygen Marine Isotope Stage
-    (MIS)~2 (29--14\,ka) is modelled comparably to the reconstructions
-    (Fig.~\ref{fig:footprints}).
+    scaled linearly (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}a) so
+    that the cumulative glaciated area of the Rhine glacier piedmont lobe
+    during Oxygen Marine Isotope Stage (MIS)~2 (29--14\,ka) is modelled
+    comparably to the reconstructions (Fig.~\ref{fig:footprints}a, black
+    rectangle).
 
@@
     Finally, in some simulations, precipitation was reduced with temperature in
-    order to simulate the potential rarification of atmospheric moisture in
+    order to simulate the potential rarefaction of atmospheric moisture in
     colder climates. This was done using an empirical relationship derived from
@@
       \caption{%
-        \textbf{(a)} July mean near-surface air temperature and
-        \textbf{(b)} January precipitation from WorldClim
-        \citep[1960--1990]{Hijmans.etal.2005}, and
-        \textbf{(c)} Mordern July standard deviation of daily mean temperature
+        \textbf{(a)} Geothermal heat flow from applying the similarity method
+        to multiple geophysical proxies \citep{Goutorbe.etal.2011} used as a
+        boundary condition to the bedrock thermal model 3\,km below the
+        ice-bedrock interface.
+        \textbf{(b)} Mordern July standard deviation of daily mean temperature
         from the ERA-Interim \citep[1979--2012]{Dee.etal.2011, Seguinot.2013}
@@
         balance (PDD) component of the ice sheet model.
-        \textbf{(d)} Geothermal heat flow from applying the similarity method
-        to multiple geophysical proxies \citep{Goutorbe.etal.2011} used as a
-        boundary condition to the bedrock thermal model 3\,km below the
-        ice-bedrock interface.}
+        \textbf{(c)} Modern July mean near-surface air temperature and
+        \textbf{(d)} January precipitation from WorldClim
+        \citep[1960--1990]{Hijmans.etal.2005}.}
       \label{fig:inputs}

### 2017-09-20 figures/alpcyc_hr_inputs.py

@@ -19,10 +19,10 @@ for i, ax in enumerate(grid.flat):
     ut.pl.draw_natural_earth(ax)
     ut.pl.add_subfig_label('(%s)' % list('abcd')[i], ax)
 
-# plot boot geoflux on last panel
+# plot boot geoflux on first panel
 nc = ut.io.load('input/boot/alps-srtm+thk+gou11simi-1km.nc')
-ax = grid.flat[3]
-cax = cgrid[3]
+ax = grid.flat[0]
+cax = cgrid[0]
 levs = range(55, 96, 5)
 cmap = ut.pl.get_cmap('PuOr_r', len(levs)-1)
 cols = cmap(range(len(levs)-1))
@@ -50,8 +50,8 @@ nc.close()
 
 # plot July temperature
 #print 'July temp min %.1f, max %.1f' % (temp.min(), temp[3:-3, 3:-3].max())
-ax = grid.flat[0]
-cax = cgrid[0]
+ax = grid.flat[2]
+cax = cgrid[2]
 levs = range(-5, 26, 5)
 cmap = ut.pl.get_cmap('RdBu_r', len(levs)-1)
 cols = cmap(range(len(levs)))
@@ -61,8 +61,8 @@ cb.set_label(u'July temperature (°C)')
 
 # plot January precipitation
 #print 'Jan. prec min %.1f, max %.1f' % (prec.min(), prec.max())
-ax = grid.flat[1]
-cax = cgrid[1]
+ax = grid.flat[3]
+cax = cgrid[3]
 levs = range(0, 301, 50)
 cmap = ut.pl.get_cmap('Greens', len(levs)-1)
 cols = cmap(range(len(levs)))
@@ -72,8 +72,8 @@ cb.set_label(r'January precipitation (mm)')
 
 # plot July standard deviation
 #print 'July s.d. min %.1f, max %.1f' % (sd.min(), sd.max())
-ax = grid.flat[2]
-cax = cgrid[2]
+ax = grid.flat[1]
+cax = cgrid[1]
 levs = [1.7, 2.0, 2.3, 2.6, 2.9, 3.2, 3.5]
 cmap = ut.pl.get_cmap('Reds', len(levs)-1)
 cols = cmap(range(len(levs)-1))

### 2017-09-20 articles/cycle/alps-cycle.tex
Rephrase palaeo-climate section.
I just realised that an important point about LGM timing is missing.

@@
     Despite the low variability of this Antarctic-based climate forcing, our
-    simulation depicts a highly dynamic ice cap, showing that alpine glaciers
+    simulation depicts a highly dynamic ice sheet, showing that alpine glaciers
     may have advanced many times over the foreland during the last glacial
@@
     parametrizations used, the model yield significant ice volume build-up
-    during MIS~4 and 2 (Fig.~\ref{fig:timeseries}). All simulations also
-    yield important glaciations during MIS~5 and 3, but timing and amplitude
-    varies a lot between the different forcing. All simulations overestimate
-    ice cover during the Younger Dryas. This might be due partly to the 2\,km
-    resolution, except for GRIP which largely overestimates Younger Dryas ice
-    cover.
+    during MIS~4 and 2 (Fig.~\ref{fig:timeseries}b), corresponding to documented
+    glaciation periods in the Alps \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
+    All simulations also yield important glaciations during MIS~5 and 3, but
+    their timing and amplitude varies significantly depending on the climate
+    forcing used. All six simulations overestimate
+    ice cover during the Younger Dryas \citep[cf. e.g.,][]{Ivy-Ochs.etal.2009}.
+    This is partly because the 2\,km resolution is too coarse to resolve
+    Younger Dryas Alpine glaciers, but the large total ice volume
+    overestimation resulting from the GRIP temperature forcing
+    (Fig.~\ref{fig:timeseries}b, blue curves) can not be explained by
+    resolution alone.
 
     Even more importantly, all simulations yield very strong ice volume
-    variability which is the result of multiple glacier advance and retreats
-    over the foreland, many more than two. Palaeo-precipitation reductions
-    partly results in a smoother modelled ice volume time series and less
-    extreme variations, but variability remains high. Ice volume variability
-    is the smallest for the EPICA palaeo-temperature record, which has the
-    least temperature variability.
+    variability, indicative of many more than two or three cycles of glacier
+    advance and retreats onto the foreland (Fig.~\ref{fig:timeseries}b).
+    Palaeo-precipitation reductions dampen some of the small-scale variability,
+    but they have little effect on the millennial scale that characterise those
+    cycles (Fig.~\ref{fig:timeseries}b, light colour curves). Finally, the
+    EPICA palaeo-temperature record has least variability the among records used
+    (Fig.~\ref{fig:timeseries}a, red curves). This results in fewer variability
+    in the total ice volume, and more restrictive glaciations during MIS~5
+    and~3, (Fig.~\ref{fig:timeseries}b, red curves).
+
+    \todo{comment on timing of the LGM}
 
@@
 
-    To select an optimal climate forcing, we look at the best known alpine
-    glaciations during MIS~2 and 4. During MIS~2, all forcing yield ice extent
-    of the same order of magnitude (Fig.~\ref{fig:footprints}). This is
-    because the temperature records have been scaled to resul in similar
-    ice covered area for the Rhine Glacier piedmont lobe. Looking at ice cover
-    on other parts, all simulations tend to overestimate ice extent in the
-    eastern part of the model domain, and to underestimate it in the western
-    part, as compared to the reconstructed LGM margin. Regarding the western
-    part, there are discussions though, as if the Lyon Lobe actually dates from
-    MIS~2 or from an older glaciation. Regarding the eastern part, the MIS~2
-    glaciation has been mapped in detail and dated (?), so that there is no
-    doubt that our model overestimate ice cover in this region. This could
-    indicate that during LGM the east-west precipitation gradient over the Alps
-    was lower than today, or that temperature depression in the east was lower
-    than in the west.
-
-    There are spatial differences though: MD01-2444, and to a greater extent
-    GRIP, forcings, tend to overestimate ice cover in all peripherical ranges:
-    the Vosges, Black Forest, Bavarian Forest, Dinaric Alps and Pennine. This
-    is because these records have a higher temperature variability so that
-    after scaling they comport brief periods of cold climate, not long enough
-    to develop the alpine ice cap, but long enough to build up ice on the
-    peripherical ranges.
-
-    The extent of ice cover during MIS~4 shows more sensitivity to the choice
-    of palaeo-climate forcing. Using the GRIP and MD01-2444, glaciers extend
-    well beyound the reach of documented moraines and glacial erratic terrain.
-    Because this extent corresponds to a short-term cold climate event,
-    palaeo-precipitation reductions help to greatly reduce this excessive
-    modelled ice cover, yet in both cases, modelled ice extent and volume
-    during MIS~4 remains higher than during MIS~2. EPICA forcing, on the other
-    hand, yield a MIS~4 glaciation that is only slightly less expensive than
-    the MIS~2, with only little sensitivity of the glaciated area to
-    palaeo-precipitation reductions.
-
-    Based on the above consideration on timing of the LGM ice extent and
-    ice extent during MIS~4, we chose EPICA as our optimal temperature record
-    for the rest of this paper. As a conservative approach in regard to quick
-    ice volume fluctuations, we choose to include palaeo-precipitation
-    correction in the following 1\,km-resolution simulation.
-
-    \note{An important assumption in the method is that the MIS~4 glaciation
-          was less (or equally) extensive than the MIS~2, i.e. that the LGM
-          outline from \citet{Ehlers.etal.2011} dates indeed from the LGM
-          (MIS~2). A different assumption could be that this outline represents
-          the most extensive glaciation in the last 120\,ka. If we allow the
-          MIS~2 glaciation to be a bit smaller than MIS~4, simulations driven
-          by GRIP and MD01-2444 also give reasonable results.}
+    During MIS~2, all six simulations yield comparative cumulative glaciated
+    areas (Fig.~\ref{fig:footprints}a). This result is inherent to the
+    palaeo-climate forcing approach used, which involved a linear scaling of
+    each palaeo-temperature anomaly record to a geomorphological
+    reconstruction of the area glaciated by the Rhine Glacier piedmont lobe
+    (Sect.~\ref{sec:paltemp}, Fig.~\ref{fig:footprints}a, black rectangle).
+    Concerning other areas, all simulations underestimate glaciated area in the
+    western part of the model domain, and overestimate it in the western
+    part, relative to the geomorphological ice limits
+    (Fig.~\ref{fig:footprints}a). This result might indicate that the LGM
+    temperature depression, here considered homogeneous, was actually lower in
+    the Eastern Alps and in the Western Alps. Alternatively, the east-west
+    gradient in precipitation existing today (Fig.~\ref{fig:inputs}d) was
+    perhaps enhanced during the LGM, as indicated by pollen reconstruction
+    \citep{Wu.etal.2007} and positive degree-day modelling of central European
+    palaeo-glaciers \citep{Heyman.etal.2013}.
+
+    Beside this general pattern, there exist differences in glaciated area
+    depending on the palaeo-climate forcing used. The MD01-2444, and to a
+    greater extent, the GRIP, palaeo-temperature records, tend to overestimate
+    MIS~2 ice cover on all peripheral ranges that surround the Alps
+    (Fig.~\ref{fig:footprints}a and c). This is particularly true when
+    precipitation corrections are applied (bright colours). In fact, both
+    palaeo-temperature records have a higher temperature variability and
+    contain brief spells of cold temperatures, which are too short to develop a
+    fully grown Alpine ice sheet, but long enough to build up ice cover on a
+    smaller scale on these peripheral ranges (Fig.~\ref{fig:timeseries}). On
+    the other hand, peripheral glaciation modelled in the EPICA simulation
+    (Fig.~\ref{fig:footprints}b) is in relatively good agreement with the
+    geomorphological reconstructions.
+
+    The modelled extent of glaciation during MIS~4 depicts a more pronounced
+    sensitivity to the choice of palaeo-climate forcing used. Using the GRIP
+    and MD01-2444 palaeo-temperature records, Alpine glaciers are modelled to
+    extend well beyond the reach of documented LGM end-moraines and glacial
+    erratic terrain (Fig.~\ref{fig:footprints}d and f). Because, in both cases,
+    the modelled glaciation extent corresponds to brief cold spells in the
+    palaeo-temperature forcing (Fig.~\ref{fig:timeseries}a),
+    palaeo-precipitation reductions greatly reduce the excessive modelled ice
+    cover (Figs.~\ref{fig:timeseries}b and \ref{fig:footprints}d and f, light
+    colours). However, with or without palaeo-precipitation corrections,
+    modelled ice extents (Fig.~\ref{fig:footprints}d and f) and volumes
+    (Fig.~\ref{fig:timeseries}b) are modelled to be higher during MIS~4 than
+    during MIS~2. This is in contradiction with geological evidence
+    \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
+
+    The EPICA palaeo-temperature forcing, on the other hand, yields a MIS~4
+    glaciation that is only slightly less expensive than the MIS~2 glaciation
+    (Fig.~\ref{fig:footprints}e). The sensitivity of the glaciated area to
+    palaeo-precipitation reductions is mostly limited to the southern terminal
+    lobes of central-Alpine glaciers, which are slightly less extensive during
+    MIS~2 and 4 (Fig.~\ref{fig:footprints}b and e).
+
+    Based on the above considerations on timing of the LGM ice extent and ice
+    extent during MIS~4, we chose EPICA as our optimal palaeo-temperature
+    record for a more detailed and higher-resolution comparison of modelled
+    glacier dynamics to available geological evidence
+    (Sect.~\ref{sec:results}). As a conservative approach in regard to the
+    rapid ice volume fluctuations, we choose to include palaeo-precipitation
+    correction in the following comparison.
 
@@
     Due to sub-freezing temperatures applied in the climate forcing of the
-    model, the entire Alpine ice field is capped by an upper layer of cold ice.
+    model, the entire Alpine ice sheet is capped by an upper layer of cold ice.
     In major Alpine valleys, such as the upper Rhone Valley, important ice

### 2017-09-20 articles/cycle/alps-cycle.tex
Rescale parameter table to fit page.
Add sensitivity of LGM timing to climate forcing.

@@
 
-    Even more importantly, all simulations yield very strong ice volume
+    Geological data from the best documented Alpine piedmont lobes indicate
+    that their maximum extension during MIS~2 occurred at around 25.5--24.0\,ka
+    \citep{Monegato.etal.2017}, after which Alpine glaciers remained or
+    potentially re-advanced to within close reach of this maximum extent, until
+    as late as 22 to 17\,ka \citep[Fig.~5]{Wirsig.etal.2016}. This is in very
+    good agreement with the result of the EPICA simulations, which yield an
+    early maximum ice volume at 25.2/24.6 (without/with palaeo-precipitation
+    reductions) followed by a standstill until 17.3\,ka
+    (Fig.~\ref{fig:timeseries}b, blue curves). The GRIP palaeo-temperature
+    forcing, on the other hand, yield two distinct total ice volume maxima
+    at 27.3/27.0 and 21.8/21.7\,ka followed by an early deglaciation of the
+    foreland by 21.4\,ka. The MD01-2444 palaeo-temperature forcing yields
+    a late LGM peak ice volume at 16.5/15.7\,ka several thousand years after
+    the dated geological evidence.
+
+    Finally, all simulations yield very strong ice volume
     variability, indicative of many more than two or three cycles of glacier
@@
     but they have little effect on the millennial scale that characterise those
-    cycles (Fig.~\ref{fig:timeseries}b, light colour curves). Finally, the
-    EPICA palaeo-temperature record has least variability the among records used
-    (Fig.~\ref{fig:timeseries}a, red curves). This results in fewer variability
-    in the total ice volume, and more restrictive glaciations during MIS~5
-    and~3, (Fig.~\ref{fig:timeseries}b, red curves).
-
-    \todo{comment on timing of the LGM}
+    cycles (Fig.~\ref{fig:timeseries}b, light colour curves). On the opposite
+    EPICA palaeo-temperature record has least variability the among records
+    used (Fig.~\ref{fig:timeseries}a, red curves). This results in fewer
+    variability in the total ice volume, and more restrictive glaciations
+    during MIS~5 and~3, (Fig.~\ref{fig:timeseries}b, red curves).
 
@@
     \begin{itemize}
-      \item The EPICA palaeo-temperature forcing yields maximum glacierized
-            area around 25\,ka, followed by a standstill of major piedmont lobe
-            in the forelands until about 18\,ka, both compatible with much of
-            the dating results, whereas the GRIP forcing results in early
-            deglaciation of the foreland complete by 21\,ka, and the MD01-2444
-            forcing results in a late LGM glaciation around 16\,ka
+      \item The EPICA palaeo-temperature forcing yields maximum ice volume at
+            25.2/24.6\,ka (without/with palaeo-precipitation reduction),
+            followed by a standstill of major piedmont lobe in the forelands
+            until 17.3\,ka, both compatible with much of the dating results,
+            whereas the GRIP forcing results in early deglaciation of the
+            foreland complete by 21.4\,ka, and the MD01-2444 forcing results in
+            a late LGM glaciation peaking at 16.5/15.7\,ka
             (Sect.~\ref{sec:timeseries}).
@@
       \label{tab:params}
-      \noindent\small\makebox[\textwidth]
+      \scalebox{0.8}
       {\begin{tabular}{llrll}
@@
       \label{tab:records}
-      \noindent\small\makebox[\textwidth]
       {\begin{tabular}{lccccccl}

### 2017-09-25 articles/cycle/alps-cycle.tex
Rephrase glacier dynamics section.

@@
     temperature depression, here considered homogeneous, was actually lower in
-    the Eastern Alps and in the Western Alps. Alternatively, the east-west
+    the Eastern Alps and in the Western Alps, as previously shown by positive
+    degree-day modelling of the central European palaeo-ice caps
+    \citep{Heyman.etal.2013}. Alternatively, the east-west
     gradient in precipitation existing today (Fig.~\ref{fig:inputs}d) was
     perhaps enhanced during the LGM, as indicated by pollen reconstruction
-    \citep{Wu.etal.2007} and positive degree-day modelling of central European
-    palaeo-glaciers \citep{Heyman.etal.2013}.
+    \citep{Wu.etal.2007}. Finally, the LGM extent of Alpine glaciers might have
+    been affected by east-west gradient in variables not accounted for
+    by the PDD model, such as cloud cover or dust deposition.
 
@@
     detail across the mountain range \citep{Penck.Bruckner.1909, Jackli.1962,
-    Bini.etal.2009, Coutterand.2010, Bavec.Verbic.2011,
+    Husen.1987, Bini.etal.2009, Coutterand.2010, Bavec.Verbic.2011,
     Buoncristiani.Campy.2011, Husen.2011}. These maps have been compiled
-    into an reconstruction covering the entire Alps \citep{Ehlers.etal.2011},
-    reproduced here (Fig.~4, red line) for comparison against model results.
-
-    The modelled total ice volume reaches a maximum of $122.5 \times 10^{3}$\,\unit{km^2},
-    or 300.4\,mm of sea-level equivalent, at 24.56\,ka
-    (Fig.~\ref{fig:lgmvel}b). A maximum glacierized area of
-    $163.3 \times 10^{3}$\,\unit{km^2} is attained shortly aftewards at 24.57\,ka.
+    into a reconstruction, covering the entire Alps \citep{Ehlers.etal.2011},
+    and reproduced here (Fig.~4, red line) for comparison against model
+    results.
+
+    The modelled total ice volume reaches a maximum of $122.5 \times
+    10^{3}$\,\unit{km^3}, or 300.4\,mm of sea-level equivalent, at 24.56\,ka
+    (Fig.~\ref{fig:lgmvel}b). A maximum glacierized area of $163.3 \times
+    10^{3}$\,\unit{km^2} is attained shortly aftewards at 24.57\,ka
     (Fig.~\ref{fig:lgmvel}b). Although the modelled timing of the LGM varies
     accross the mountain range (Sect.~\ref{sec:timing}), at 24.57\,ka nearly
-    all glaciers extend to within a few km from their modelled maximum stage
-    (Fig.~\ref{fig:lgmvel}a, dashed orange line).
-
-    Although the palaeo-climate forcing was adapted (Sect.~\ref{sec:climate})
+    all outlet glaciers extend to within a few km from their modelled maximum
+    stage (Fig.~\ref{fig:lgmvel}a, dashed orange line).
+
+    The palaeo-climate forcing was adapted (Sect.~\ref{sec:paltemp})
     to model this maximum configuration in broad agreement with the geological
     reconstruction of the LGM extent (Fig.~\ref{fig:lgmvel}a, solid red line),
-    local discrepancies remain. Most notably, reconstructed ice extent in the
+    yet local discrepancies remain. As already outlined for low resolution
+    runs (Sect.~\ref{sec:footprints}), the extent of glaciation in the
     north-western Alps, including the Rhone Glacier complex, the Jura ice cap,
-    and the Lyon Lobe, is largely missing from our model results. On the other
-    hand, the model yields excessive glaciation in the eastern Alps, where the
-    Palten, Drau and Sava glaciers are modelled to extend tens of kilometers
-    beyond the mapped moraines. These discrepancies might indicate that the
-    LGM climate was characterized by an east-west gradient in temperature or
-    precipitation anomalies relative to present not included in the climate
-    forcing, or perhaps a east-west gradient in variables not accounted for
-    by the PDD model, such as cloud cover or dust deposition.
+    and the Lyon Lobe, is underestimated in the model results
+    (Fig.~\ref{fig:lgmvel}a). On the other hand, the model yields excessive ice
+    cover in the eastern Alps, where the Palten, Drau and Sava glaciers are
+    modelled to extend tens of km beyond the mapped end-moraines
+    (Fig.~\ref{fig:lgmvel}a). As previously discussed
+    (Sect.~\ref{sec:footprints}), these discrepancies might indicate that the
+    LGM climate was characterized by an east-west gradient in temperature
+    \citep[cf.][]{Heyman.etal.2013} or precipitation
+    \citep[cf.][]{Wu.etal.2007} anomalies relative to present, or an east-west
+    gradient in variables not accounted for by the PDD model.
 
@@
     Garda and Piave glaciers in the central southern Alps is constrained
-    several kilometres within the mapped LGM moraines. This appears to result
+    several kilometres within the mapped LGM moraines (Fig.~\ref{fig:lgmvel}a).
+    This appears to result
     from the paleo-precipitation reduction used to force the model
@@
     south-western Alps is modelled to extend several kilometers beyond the
-    mapped limit, indicating that the LGM temperature depression was likely
+    mapped limit (Fig.~\ref{fig:lgmvel}a), indicating that the LGM temperature
+    depression was likely
     dampened by Mediterranean climate in this part of the model domain.
@@
     the Jura Mountains, the Vosges Mountain, the Black Forest, the Bohemian
-    Forest and the Dinaric Alps. An independent ice cap also covers the
-    Hochwart massif during most of the simulation, yet it is engulfed by
-    alpine glaciers during the LGM to become a peripheral ice dome. The LGM
+    Forest and the Dinaric Alps (Fig.~\ref{fig:lgmvel}a). An independent ice
+    cap also covers the Hochwart massif during most of the simulation, yet it
+    is engulfed by alpine glaciers during the LGM to become a peripheral ice
+    dome \citep[Fig.~\ref{fig:lgmvel}a; cf.][Fig.~2.5]{Husen.2011}. The LGM
     extent of peripheral ice caps is underestimated in the Vosges and Jura
-    mountains, while it is overestimated for the Vercors. There is a conformity
-    between the discrepancies between modelled and reconstructed LGM extent
-    for the main alpine glacier complex vs peripheral ice caps, which indicates
-    a climatic cause rather than an ice dynamics cause.
-
-    \todo{add location names on Fig.~\ref{fig:lgmvel}.}
+    mountains, while it is overestimated for the Vercors
+    \citep[Fig.~\ref{fig:lgmvel}a; cf.][Figs.~4.28, 4.32, and
+    4.33]{Coutterand.2010}. Thus, there exists a regional conformity between
+    the model-data discrepancies obtained for the main Alpine ice sheet and
+    those obtained for peripheral ice caps. Because peripheral ice cap have
+    very different glacier dynamics than the main ice sheet, this conformity
+    most likely indicates a climatic cause, rather than an ice dynamics cause,
+    for these discrepancies.
+
+    \todo{add location names on Fig.~\ref{fig:lgmvel}: Rhone Glacier, Jura ice
+          cap, and the Lyon Lobe, Palten, Drau and Sava glaciers, Ticino, Iseo,
+          Garda and Piave glaciers, Durance glacier, Vercors, Chartreuse,
+          Bauge, Jura, Vosges, Black Forest, Bohemian Forest, Dinaric Alps,
+          Hochwart.}
 
@@
     The LGM Alpine ice flow pattern was traditionally described as that of a
-    network of interconnected valley glaciers that was primarily controlled by
+    network of interconnected valley glaciers, primarily controlled by
     subglacial topography. However, the geomorphology shows that ice was thick
@@
     \citep[e.g.,][]{Onde.1938, Penck.Bruckner.1909, Jackli.1962, Husen.1985,
-    Coutterand.2010, Kelly.etal.2004, Husen.2011}, and even perhaps form
+    Coutterand.2010, Kelly.etal.2004, Husen.2011}, and even to perhaps form
     self-sustained ice domes \citep{Florineth.1998, Florineth.Schluchter.1998,
@@
 
-    The modelled flow pattern at 24.57\,ka is complex (Fig.~\ref{fig:lgmvel}a).
-    The modelled fast-flow regions generally occur along the main river
-    valleys, while ice domes and ice divides are predominantly located over
-    major reliefs . Nevertheless, the model results depict occasional ice flow
-    across mountain passes (transfluences) and ice divide above topographic
-    lows.
-
-    Major transfluences occur for instance across col du Mont-Cenis,
-    Simplonpass, Brünigpass in the western Alps, and over Fernpass,
-    Gailbergsattel and Stranigeralpe in the Eastern Alps
-    (Fig.~\ref{fig:lgmvel}a). Finally, self-sustained ice domes characteristic
-    of ice caps are modelled in two locations over Flüelapass and Ötztal
-    (Fig.~\ref{fig:lgmvel}a).
-
-    Transfluence across col de Montgenèvre was previously questioned
-    \citep[Fig.~2]{Cossart.etal.2012}. Transfluence across col de Mont Cenis
-    was previoysly described \citep[Fig.~3.18]{Onde.1938, Coutterand.2010}.
-    Transfluences across Fernpass and Seefeld are known from the geomorphology
-    \citep[Fig.~2.4]{Penck.Bruckner.1909, Husen.2011}. Transfluence across
-    Brunigpass was already marked in the maps \citep{Jackli.1962}. There is
-    numerous evidence for transfluence across Simplon Pass
-    \citep{Kelly.etal.2004}. Ice crossed the Pyhrn Pass \citep{Husen.2004}. Ice
-    flew across the Gailberg and Kreuzberg saddles \citep{Husen.1985}. The
-    Tagliamento catchment is usually assumed to not have received tranfluences
-    from the Drau catchment \citep{Monegato.etal.2007}, but this would not be
-    incompatible with reconstructed ice surface elevations in this area
-    \citep{Husen.1987}.
-
-    In the Eastern Alps, modelled transfluences and crosswise divides are
-    generally incompatible with geological reconstructions, indicating that
-    climate deterioration was overestimated in the model input. In the Western
-    Alps, however, transfluences generally occur where they have been
-    documented by geologic evidence. These model results depict the LGM Alpine
-    ice complex as an intermediate between ice fields and ice caps, bearing
-    characteristics of both and perhaps no modern analogue.
-
-    \todo{add location names on Fig.~\ref{fig:lgmvel}.}
-
-%    \citep{Jouvet.etal.2017}
+    The modelled flow pattern at 24.57\,ka is complex (Fig.~\ref{fig:lgmvel}a,
+    blue colour mapping). Fast-flow regions generally occur along the main
+    river valleys, while ice domes and ice divides are predominantly located
+    over major reliefs (Fig.~\ref{fig:lgmvel}a). Nevertheless, the model
+    results depict occasional ice flow across the modern water divides, i.e.
+    transfluences.
+
+    In the Western Alps, major transfluences occur for instance across Col de
+    Montgenèvre, Col du Mont-Cenis, Simplon Pass and Brünig Pass
+    (Fig.~\ref{fig:lgmvel}a). Although a transfluence across col de Montgenèvre
+    was previously questioned \citep[Fig.~2]{Cossart.etal.2012}, evidence for
+    southerly ice flow across Mont-Cenis has been recognized
+    \citep[Fig.~3.18]{Onde.1938, Coutterand.2010}. Similarly, tranfluences have
+    been previously identified from the geomorphology across Simplon Pass
+    \citep{Kelly.etal.2004} and Brünig Pass \citep{Jackli.1962}.
+
+    In the Eastern Alps, major transfluences are modelled to have occur for
+    instance across at Fern Pass, the Seefeld Saddle, the Gailberg Saddle, the
+    Kreuxberg Saddle, the Straniger Alp, and the Pyhrn Pass
+    (Fig.~\ref{fig:lgmvel}a). Transfluences across Fern Pass and the Seefeld
+    Saddle are known from the geomorphology
+    \citep[Fig.~2.4]{Penck.Bruckner.1909, Husen.2011}. It is also known that
+    ice flew across the Gailberg and Kreuzberg saddles \citep{Husen.1985}, and
+    the Pyhrn Pass \citep[Fig.~\ref{fig:lgmvel}a; cf.][Fig.~2.5]{Husen.2011}.
+    On the other hand, no transfluence has previously been documented at the
+    Straniger Alp. The Tagliamento catchment is usually assumed to not have
+    received tranfluences from the Drau catchment \citep{Monegato.etal.2007},
+    but this would not be incompatible with reconstructed ice surface
+    elevations in this area \citep{Husen.1987}.
+
+    Finally, self-sustained ice domes characteristic of ice caps are modelled
+    in two locations over Flüelapass and Ötztal (Fig.~\ref{fig:lgmvel}a).
+    Except for too extensive ice cover in the easternmost part of the model
+    domain (Sect.~\ref{sec:extent}), there is generally a good agreement
+    between transfluences observed in the geomorphology and the model results.
+    Both depict the LGM Alpine ice flow pattern as an intermediate between that
+    of a topography-controlled ice field, and that of a self-sustained ice cap.
+
+    \todo{add location names on Fig.~\ref{fig:lgmvel}: Mont-Cenis, Simplonpass,
+          Brünigpass, Fernpass, Seefeld, Gailbergsattel and Stranigeralpe,
+          Tagliamento, Drau, Flüelapass, Ötztal.}
 
@@
     data indicate that Alpine glaciers reached their maximum extent between 20
-    and 26\,ka (calibrated \chem{^{14}C} or \chem{^{10}Be} ages), but glacier
-    lobes stayed in the foreland until as late as 22 to 17\,ka, when they
-    experienced rapid retreat synchronously to lowering of the ice surface in
-    the mountains (\citealp[Fig.~5]{Wirsig.etal.2016},
+    and 26\,ka (calibrated \chem{^{14}C} and \chem{^{10}Be} ages), but also
+    that terminal lobes stayed in the foreland until as late as 22 to 17\,ka,
+    when they experienced rapid retreat synchronously to lowering of the ice
+    surface in the mountains (\citealp[Fig.~5]{Wirsig.etal.2016},
     \citealp[Fig.~3]{Monegato.etal.2017}).
@@
 
-    It is important to notice that these differences in timing of the LGM occur
-    in the model results despite the simple climate forcing used consisting of
-    homogeneous anomalies. These discrepancies between different glacier lobes
-    are thus not related to climate, but are a pure product of modelled
-    inherent glacier dynamics. They occur because of differences in the
-    subglacial topography, in particular catchment sizes and bed roughness.
+    These differences in timing of the LGM occur in the model results despite
+    the homogeneous temperature and precipitation anomalies supplied as
+    palaeo-climate forcing used. The LGM timing differences modelled here are
+    thus not related to climate, but are inherent to modelled glacier dynamics.
+    They result from differences in the subglacial topography, in particular
+    catchment sizes and bed roughness, of the different outlet glaciers.
 
@@
     For large Alpine glaciers flowing in deep valleys, such as the Rhone,
-    Rhine and Garda glaciers, several thousands of years are needed to attain
+    Rhine and Garda glaciers, several thousand model years are needed to attain
     a thermodynamical equilibrium between cold ice advection from the high
@@
     downed by subfreezing air temperature before glacier advance.
-
-    For such glaciers, this thermodynamical equilibrium is typically not yet
+    For larger glaciers, this thermodynamical equilibrium is typically not yet
     attained around 25\,ka. This is why several Alpine lobes overshoot their
@@
     Our results indicate that even for a relatively small ice sheet like that
-    found in the Alps, differences in timing on the order of the millenium can
-    result from complex glacier dynamics. Further differences in timing of the
-    LGM caused by heterogeneous climatic anomalies, such as precipitation
+    found in the Alps, ka-scale differences in timing of the LGM can
+    result from complex glacier dynamics. Further differences in LGM timing
+    caused by heterogeneous climatic anomalies, such as precipitation
     anomaly gradients, not included in our model set-up, may have
-    counterbalanced or emphasize those presented here.
-
-    \todo{add location names on Fig.~\ref{fig:lgmvel}.}
+    counterbalanced or emphasize those modelled here.
+
+    \todo{add location names on Figs.~\ref{fig:lgmvel} and \ref{fig:timing}:
+    Dora Riparia, Ivrea, and Tagliamento, Durance, Rhone, Inn, Enns, Ticino,
+    and Garda, Isère, Como and Iseo, Vercors, Jura Mountains, Black Forest and
+    Hochschwab.}
 
@@
     valley troughs, represent the maximum elevation of the LGM ice surface,
-    the LGM ice thickness in the Alps been reconstructed in several areas
+    maximum ice thickness in the Alps been reconstructed in several areas
     \citep{Husen.1987, Florineth.1998, Florineth.Schluchter.1998,
@@
     However, this assumption is challenged by geological evidence from
-    Scandinavia \citep[e.g.,][]{Kleman.1994, Kleman.Borgstrom.1994,
-    Stroeven.etal.2002}, the British Isles \citep[e.g.,][]{Fabel.etal.2012}, and
+    Scandinavia \citep[e.g.,][]{Kleman.1994, Kleman.Borgstrom.1994},
+    the British Isles \citep[e.g.,][]{Fabel.etal.2012}, and
     North America \citep[e.g.,][]{Kleman.etal.2010}, that pre-glacial landscapes
     located well above the trimlines have been glaciated and preserved under
-    under cold-based ice, sometimes for several glacial cycles.
-
-    In several places across the Alps, the maximum elevation reached by the
-    ice surface during the LGM has been reconstructed from trimlines, the upper
-    limit of glacial erosion observed in the landscape. This include the entire
-    western Alps, the Rhone Glacier catchment, the Rhine Glacier catchment,
-    south-eastern Switzerland and all of Austria. These reconstructions are
-    based on the assumption that glacial ice was significantly erosive up to
-    the or almost up to the surface, which was then located at or slightly
-    above the elevation of the trimlines. This assumption is supported by the
-    fact that there exist a regional consistence in trimlines elevations, which
-    are generally located along a surface sloping in the assumed direction of
-    flow. However, in other glaciated mountain regions, including several
-    sites in the Scandinavian Mountains, the Cairgorns in Scotland, and high
-    plateaus in the Canadian Cordillera, that cold-based ice commonly occured
-    at high elevation and preserved the pre-glacial landscape.
+    under cold-based ice, sometimes for several glacial cycles
+    \citep{Stroeven.etal.2002}.
 
@@
     consistently modelled to have occured several hundred metres above the
-    observed trimline elevations (Fig.~\ref{fig:trimlines}a, b), with a mean
+    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
     difference of 861\,m (xx\,\%, Fig.~\ref{fig:trimlines}b). This result is
@@
     the Cordillera ice sheet show that varying ice rheological parameters
-    resulted relative errors of 25\,\unit{\%} with regard to ice rheological
-    parameters and 14\,\unit{\%} regarding basal sliding parameters
-    \citep[Fig.~7]{Seguinot.etal.2016}.
-
-    In addition, our model results depict a polythermal LGM Alpine ice cover.
+    resulted in relative errors in modelled total ice volume of 25\,\unit{\%}
+    with regard to ice rheological parameters and 14\,\unit{\%} regarding basal
+    sliding parameters \citep[Fig.~7]{Seguinot.etal.2016}.
+
+    Beside, our model results depict a polythermal LGM Alpine ice cover.
     Due to sub-freezing temperatures applied in the climate forcing of the
@@
     near the glacier base, allowing basal melt, sliding and potentially
-    erosion. On the other hand, on the highest mountains ice cover is to thin
-    and too static to form temperate ice, resulting in cold ice down to the
-    bed, preventing potential erosion.
+    erosion (Fig.~\ref{fig:trimlines}c, white areas). On the other hand, on the
+    highest mountains ice cover is to thin and too static to form temperate
+    ice, resulting in cold ice down to the bed, preventing potential erosion
+    (Fig.~\ref{fig:trimlines}c, hatched areas).
 
@@
     modelled to have experienced warm-based for less than 1\,ka (7\,\%) of the
-    MIS~2. Although this result calls for more detailed comparisons across the
+    MIS~2 (Fig.~\ref{fig:trimlines}c). Although this result calls for more
+    detailed comparisons across the
     entire Alpine range and specific sensitivity studies to relevant basal
@@
     cold ice during the LGM, already found in some places in a much warmer
-    climate today, is inevitable. Thus we want to challenge the assumption that
+    climate today \todo{ref.}, is inevitable. Thus we want to challenge the assumption that
     Alpine trimlines mark the maximum upper ice surface elevation of the LGM
@@
 
-    \todo{look for references.}
+    \todo{compute relative difference.}
 
@@
     result in a high time variability in the total modelled ice volume
-    (Fig.~\ref{fig:timeseries}). Unsurprisingly, this result is confirmed by
-    the 1-km resolution run, using the optimal EPICA palaeo-temperature record
-    (Sect.~\ref{sec:paltemp}) and conservatively includes palaeo-precipitation
-    reductions (Sect.~\ref{sec:palprec}), which also results in strong total
-    ice volume variability throughout the last glacial cycle
-    (Fig.~\ref{fig:lgmvel}b).
+    (Fig.~\ref{fig:timeseries}). Using the optimal EPICA palaeo-temperature
+    record with least variability (Sect.~\ref{sec:paltemp}) and conservatively
+    including palaeo-precipitation reductions (Sect.~\ref{sec:palprec}), the
+    1-km resolution simulation also results in strong total ice volume
+    variability throughout the last glacial cycle (Fig.~\ref{fig:lgmvel}b).
 
@@
     (Fig.~\ref{fig:lgmvel}b). These episodes are the result of synchronous
-    advances several Alpine glaciers well into the major valleys and sometimes
+    advances of several Alpine glaciers well into the major valleys and sometimes
     even onto the foreland (Fig.~\ref{fig:profiles}), Supplementary animation).
@@
     reduced precipitation dampening glacier response, our simulation depict
-    the Alpine ice complex as highly dynamic, with dynamic of individual
-    glaciers controlled by variations in subglacial topography and catchment
-    geometry.
-
-    \idea{Matthias: The selection of figures in the main text and SOM is good.
-          I find Figure 5 quite difficult to understand at first glance. I
-          would combine it (already before submission) with Fig. 6. the Figure
-          can then focus both on the temporal advance/retreat patterns of the 4
-          lobes (four panels on the left side of figure) and the spatial
-          differences in max. ice elevation (1 panel on right side of figure,
-          with centerlines indicated). I would have the one with trimline
-          elevation again in the Supplementary.}
+    the Alpine ice complex as highly dynamic, with many more than two or three
+    \citep[cf.][]{Preusser.2004,Ivy-Ochs.etal.2008} glaciations, and regional
+    glacier dynamic controlled by variations in subglacial topography and
+    catchment geometry.
 

### 2017-09-25 articles/cycle/alps-cycle.tex
Add paragraph on north-south symetry.

@@
     gradient in variables not accounted for by the PDD model.
+
+    On the other hand, while atmospheric circulation models
+    \citep{Strandberg.etal.2011, Ludwig.etal.2016} and palaeo-climate proxies
+    \citep{Luetscher.etal.2015} both support differential precipitation changes
+    north and south of the Alps, our model results show no obvious north-south
+    bias as compared to the mapped LGM margins (Fig.~\ref{fig:lgmvel}a) despite
+    the homogeneous temperature and precipitation anomalies applied relative to
+    present. Although this may appear as a contradiction with previous,
+    constant-climate modelling results \citep{Becker.etal.2016}, it is the
+    result of introducing a time-dependent palaeo-temperature forcing. Constant
+    climate forcing systematically resulted in extraneous
+    \citep[Fig.~3]{Becker.etal.2016} and premature
+    \citep[Fig.~4]{Becker.etal.2016} glaciation on the north slope of the Alps.
+    Using time-dependent palaeo-temperature forcing, progressive climate
+    cooling over several ka allow for warmer temperatures, closer to
+    the thermodynamic equilibrium, within the ice and the bedrock. This results
+    in thinner glaciers and limits overshoot of the equilibrium state
+    (discussed further in Sect.~\ref{sec:timing}).
 

### 2017-09-25 articles/cycle/alps-cycle.tex
Add reference for present day cold ice.

@@
     sliding and ice rheological parameters, the presence of an upper layer of
-    cold ice during the LGM, already found in some places in a much warmer
-    climate today \todo{ref.}, is inevitable. Thus we want to challenge the assumption that
+    cold ice during the LGM, already found at high altitude in the much warmer
+    climate of today \citep{Suter.etal.2001}, is inevitable.
+    Thus we want to challenge the assumption that
     Alpine trimlines mark the maximum upper ice surface elevation of the LGM

### 2017-09-25 articles/cycle/alps-cycle.tex
Rephrase conclusions.

@@
     ice sheet (Sect.~\ref{sec:model}).
-
     Using three different palaeo-temperature forcing records (GRIP, EPICA, and
@@
             model forcing, probably controlled the LGM extent of ice cover in
-            the Alps (Sect.~\ref{sec:extent}).
-      \item The assymetric nature of LGM ice extent north and south of the Alps
+            the Alps. The asymmetric extent north and south of the Alps
             can be explained by the transient nature of the LGM extent without
@@
             self-sustained ice domes occured (Sect.~\ref{sec:flow}).
-      \item The modelled LGM is a transient stage in which glaciers are not in
-            equilibrium with climate. Timing potentially varies across the
-            range (Sect.~\ref{sec:timing}).
+      \item The LGM was a transient stage in which glaciers were not in
+            equilibrium with climate. Timing potentially varied across the
+            range due to inherent glacier dynamics (Sect.~\ref{sec:timing}).
       \item Ice thickness during the LGM is modelled to be much higher than in
-            reconstructions. In average modeleld surface elevation is 861\,m
+            reconstructions. In average, modelled surface elevation is 861\,m
             above the Rhone Glacier trimlines, which may instead indicate an
             englacial thermal boundary (Sect.~\ref{sec:thickness}).
-      \item Aline glaciers were potentially very dyamic. They responded quickly
-            to climate fluctuation and some of them potentially advanced many
+      \item Alpine glaciers were potentially very dynamic. They quickly
+            responded to climate fluctuation and some potentially advanced many
             times over the foreland during the last glacial cycle
-            (Sect.~\ref{sec:glaciations})
+            (Sect.~\ref{sec:glaciations}).
     \end{itemize}
@@
     These results are nevertheless limited by the simplicity of the climate
-    forcing and surface mass balance model applied, uncertain ice physical
+    forcing and surface mass balance model applied, and uncertain ice physical
     model parameters. Through more detailed sensitivity studies, targeted to
@@
     modelling studies will certainly be able to quantify uncertainties
-    associated with some of the above statements. However, we also hope that
-    these conclusion will also serve as a basis for future studies of glacial
+    associated with some of the above statements. Nevertheless, we hope that
+    these conclusions will also serve as a basis for future studies of glacial
     geology in the Alps, and call for a more systematic aggregation and
-    homogeneisation of glacial geological data to form a basis for model
+    homogenization of glacial geological data to form a basis for model
     validation across the entire Alpine range.

### 2017-09-25 articles/cycle/alps-cycle.tex
Rephrase acknowledgements. Second draft complete!

@@
     J.~Seguinot designed the study, ran the simulations, and wrote most of
-    the manuscript. All authors contributed to the text.}
+    the manuscript. M.~Huss provided modern ice thickness data. All authors
+    contributed to the text. The idea for this study stems in part from an
+    excursion organised by F.~Preusser in the central Alps in 2012.}
 
@@
 
-    This manuscript builds largely upon a previous study on the Cordilleran Ice
+    This manuscript is partly based on a previous study on the Cordilleran Ice
     Sheet which is part of J.~Seguinot's PhD thesis. The experimental design
@@
     no.~370 and 377). We are equally thankful to Giovanni Monegato and Marc
-    Lüether for very insightful discussions of preliminary model results during
+    Luetscher for very insightful discussions of preliminary model results during
     the European Geoscience Union General Assembly~2017.

### 2017-09-26 figures/alpcyc_hr_lgmvel.py
Add glacier names on lgm and timing figures.

@@ -36,6 +36,7 @@ cb.set_label(r'surface velocity ($m\,a^{-1}$)')
 ut.pl.draw_natural_earth(ax)
 ut.pl.draw_lgm_outline(ax)
 ut.pl.draw_footprint(ax)
+ut.pl.draw_glacier_names(ax)
 ut.pl.add_corner_tag('%.2f ka' % a, ax)
 
 # Time series

### 2017-09-26 figures/alpcyc_hr_timing.py

@@ -27,6 +27,7 @@ cs = ax.contourf(maxthkage/1e3, levs, extent=extent, colors=cols, extend='both',
 ut.pl.draw_boot_topo(ax)
 ut.pl.draw_envelope(ax)
 ut.pl.draw_natural_earth(ax)
+ut.pl.draw_glacier_names(ax)
 
 # add colorbar
 cb = fig.colorbar(cs, cax)

### 2017-09-26 figures/util/pl.py

@@ -359,6 +359,19 @@ def draw_trimlines(ax=None, c=palette['darkblue'], s=4**2, alpha=0.75):
                transform=ut.pl.swiss)
 
 
+def draw_glacier_names(ax=None):
+    """Add glacier lobes and ice cap names."""
+    shp = cshp.Reader('../data/native/alpcyc_glacier_names.shp')
+    for rec in shp.records():
+        name = rec.attributes['name'].decode('utf-8').replace(' ', '\n')
+        sort = rec.attributes['type']
+        lon = rec.geometry.x
+        lat = rec.geometry.y
+        x, y = ax.projection.transform_point(lon, lat, src_crs=ut.pl.ll)
+        style = ('italic' if sort == 'cap' else 'normal')
+        ax.text(x, y, name, fontsize=6, style=style, ha='center', va='center')
+
+
 def draw_boot_topo(ax=None, res='1km'):
     """Add bootstrapping topography image."""
     ax = ax or iplt.gca()

### 2017-09-26 figures/alpcyc_hr_lgmvel.py
Add ice domes and transfluences on lgm figure.

@@ -37,6 +37,8 @@ ut.pl.draw_natural_earth(ax)
 ut.pl.draw_lgm_outline(ax)
 ut.pl.draw_footprint(ax)
 ut.pl.draw_glacier_names(ax)
+ut.pl.draw_ice_domes(ax)
+ut.pl.draw_major_transfluences(ax)
 ut.pl.add_corner_tag('%.2f ka' % a, ax)
 
 # Time series

### 2017-09-26 figures/util/pl.py

@@ -372,6 +372,55 @@ def draw_glacier_names(ax=None):
         ax.text(x, y, name, fontsize=6, style=style, ha='center', va='center')
 
 
+def draw_ice_domes(ax=None, textoffset=4):
+    """Add ice domes."""
+    shp = cshp.Reader('../data/native/alpcyc_ice_domes.shp')
+    for rec in shp.records():
+        name = rec.attributes['name'].decode('utf-8')
+        lon = rec.geometry.x
+        lat = rec.geometry.y
+        x, y = ax.projection.transform_point(lon, lat, src_crs=ut.pl.ll)
+        ax.plot(x, y, 'k^')
+        ax.annotate(name, xy=(x, y), xytext=(0, textoffset), style='italic',
+                    textcoords='offset points', ha='center', va='bottom')
+
+
+def draw_major_transfluences(ax=None, textoffset=4):
+    """Add major transfluences."""
+    locations = {u'Col de Montgenèvre': 'lc',
+                 u'Col du Mont-Cenis': 'uc',
+                 u'Simplon Pass': 'uc',
+                 u'Brünig Pass': 'uc',
+                 u'Fern Pass': 'cl',
+                 u'Seefeld Saddle': 'cr',
+                 u'Gailberg Saddle': 'cl',
+                 u'Kreuzberg Saddle': 'cr',
+                 u'Straniger Alp': 'cr',
+                 u'Pyhrn Pass': 'lc'}
+    shp = cshp.Reader('../data/native/alpcyc_transfluences.shp')
+    for rec in shp.records():
+        name = rec.attributes['name'].decode('utf-8')
+        if name in locations:
+            lon = rec.geometry.x
+            lat = rec.geometry.y
+            x, y = ax.projection.transform_point(lon, lat, src_crs=ut.pl.ll)
+            xloc = locations[name][1]
+            yloc = locations[name][0]
+            dx = {'c': 0, 'l': -1, 'r': 1}[xloc]*textoffset
+            dy = {'c': 0, 'l': -1, 'u': 1}[yloc]*textoffset
+            ha = {'c': 'center', 'l': 'right', 'r': 'left'}[xloc]
+            va = {'c': 'center', 'l': 'top', 'u': 'bottom'}[yloc]
+            name = name.replace('Col de ', '')
+            name = name.replace('Col du ', '')
+            name = name.replace(' Alp', '')
+            name = name.replace(' Pass', '')
[…]

### 2017-09-26 articles/cycle/alps-cycle.tex
Update caption, names in text and remove todos.

@@
     (Fig.~\ref{fig:lgmvel}a). On the other hand, the model yields excessive ice
-    cover in the eastern Alps, where the Palten, Drau and Sava glaciers are
+    cover in the eastern Alps, where the Mur, Drava and Sava glaciers are
     modelled to extend tens of km beyond the mapped end-moraines
@@
 
-    \todo{add location names on Fig.~\ref{fig:lgmvel}: Rhone Glacier, Jura ice
-          cap, and the Lyon Lobe, Palten, Drau and Sava glaciers, Ticino, Iseo,
-          Garda and Piave glaciers, Durance glacier, Vercors, Chartreuse,
-          Bauge, Jura, Vosges, Black Forest, Bohemian Forest, Dinaric Alps,
-          Hochwart.}
-
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
@@
     Straniger Alp. The Tagliamento catchment is usually assumed to not have
-    received tranfluences from the Drau catchment \citep{Monegato.etal.2007},
+    received tranfluences from the Drava catchment \citep{Monegato.etal.2007},
     but this would not be incompatible with reconstructed ice surface
@@
     of a topography-controlled ice field, and that of a self-sustained ice cap.
-
-    \todo{add location names on Fig.~\ref{fig:lgmvel}: Mont-Cenis, Simplonpass,
-          Brünigpass, Fernpass, Seefeld, Gailbergsattel and Stranigeralpe,
-          Tagliamento, Drau, Flüelapass, Ötztal.}
 
@@
     counterbalanced or emphasize those modelled here.
-
-    \todo{add location names on Figs.~\ref{fig:lgmvel} and \ref{fig:timing}:
-    Dora Riparia, Ivrea, and Tagliamento, Durance, Rhone, Inn, Enns, Ticino,
-    and Garda, Isère, Como and Iseo, Vercors, Jura Mountains, Black Forest and
-    Hochschwab.}
 
@@
         24.57\,ka before present, corresponding to the maximum modelled ice
-        cover. Modelled Last Glacial Maximum
+        cover. Modelled self-sustained ice domes (triangles) and major
+        transfluences (crosses). Modelled Last Glacial Maximum
         (LGM) ice extent (dashed orange line) and geomorphological

### 2017-09-26 articles/cycle/alps-cycle.tex
Add mean obs and modelled elev on trimline plot.
I just realised that relative errors to altitude make no sense.

@@
     observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
-    difference of 861\,m (xx\,\%, Fig.~\ref{fig:trimlines}b). This result is
+    difference of 861\,m (Fig.~\ref{fig:trimlines}b). This result is
     somewhat depedent on uncertain basal sliding and ice rheological parameters
@@
     upper layer cold ice in the Alps.
-
-    \todo{compute relative difference.}
 
@@
         mark areas modelled to have experienced less than 1\,ka temperate ice
-        cover during MIS~2 (29--14\,ka).
-        \todo{add relative mean difference.}}
+        cover during MIS~2 (29--14\,ka).}
       \label{fig:trimlines}

### 2017-09-26 figures/alpcyc_hr_trimlines.py

@@ -113,7 +113,11 @@ zz = [zt.min(), zt.max()]
 scax.plot(zz, zz+davg, c='0.25')
 scax.plot(zz, zz+davg-dstd, c='0.25', dashes=(2, 1), lw=0.5)
 scax.plot(zz, zz+davg+dstd, c='0.25', dashes=(2, 1), lw=0.5)
+scax.axvline(zavg, c='0.25', dashes=(2, 1), lw=0.5)
 scax.axhline(zavg+davg, c='0.25', dashes=(2, 1), lw=0.5)
+scax.text(zavg-25.0, (bt+ht).min(), '%.0f m' % zavg, color='0.25',
+          rotation=90, rotation_mode='anchor')
+scax.text(zz[0], zavg+davg+25.0, '%.0f m' % (zavg+davg), color='0.25')
 hsax.axhline(davg, c='0.25')
 hsax.axhline(davg-dstd, c='0.25', dashes=(2, 1), lw=0.5)
 hsax.axhline(davg+dstd, c='0.25', dashes=(2, 1), lw=0.5)

### 2017-09-29 articles/cycle/alps-cycle.tex
Reread entire manuscript and polish style.

@@
     and even onto the alpine foreland \citep{Charpentier.1841}, yielding the
-    idea that, under colder temperatures, expansive ice sheets had once covered
-    much of Europe and North America \citep{Agassiz.1840}.
+    idea that, under colder temperatures, expansive \emph{ice sheets} had once
+    covered much of Europe and North America \citep{Agassiz.1840}.
 
@@
     systematic classification of the glaciofluvial terraces on the northern
-    foreland indicated that there had been at least four major
+    foreland later indicated that there had been at least four major
     glaciations in the Alps \citep{Penck.Bruckner.1909}. More recent studies
@@
 
-    More recently, palaeoclimate records extracted from deep sea sediments and
+    From the mid-twentieth
+    century, palaeoclimate records extracted from deep sea sediments and
     ice cores have provided a much more detailed picture of the Earth
@@
     Shackleton.Opdyke.1973, Dansgaard.etal.1993, Augustin.etal.2004},
-    indicating neither one, nor four, but rather tens of \emph{glacial cycles}.
-    During the last 800000 years (800\,ka), glacial and interglacial periods
+    indicating neither one, nor four, but rather tens of glacial and
+    interglacial periods. During the last 800000 years (800\,ka), these
+    \emph{glacial cycles}
     have succeeded each other with a 100\,ka periodicity \citep{Hays.etal.1976,
@@
     by the North American and Eurasian ice sheet complexes. It is thus unclear
-    wether glacier advances and retreats in the Alps were in pace with global
+    whether glacier advances and retreats in the Alps were in pace with global
     sea-level fluctuations.
@@
     mountain range \citep[e.g.,][]{Bini.etal.2009, Coutterand.2010,
-    Husen.2011}. Although no evidence for cold-based glaciation has been found
-    in the Alps to date, the assumption that trimlines, the upper limit of
-    glacial erosion, represent the maximum ice surface elevation, which has
+    Husen.2011}. The assumption that trimlines, the upper limit of glacial
+    erosion, represent the maximum ice surface elevation, has
     been repeatedly invalidated in other glaciated regions of the globe
-    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012}. Ice flow
+    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012}. Yet no
+    evidence for cold-based glaciation has been reported in the Alps to date.
+    Alpine ice flow
     patterns were primarily controlled by subglacial topography, but there is
@@
     agreement with the maximum expansion of continental ice sheets recorded by
-    the Marine Oxygen Isotope Stage (MIS) 2 (29--14\,ka)
-    \citep{Lisiecki.Raymo.2005}. Regional variation between different piedmont
+    the Marine Oxygen Isotope Stage (MIS) 2
+    \citep[29--14\,ka;][]{Lisiecki.Raymo.2005}. Regional variation between
+    different piedmont
     lobes exist \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear wether they
@@
     Centre for Medium-Range Weather Forecasts Reanalysis Interim
-    \citep[ERA-Interim;][]{Dee.etal.2011}, perturbed by temperature lapse-rate
+    \citep[ERA-Interim;][]{Dee.etal.2011} over the period 1979--2012,
+    perturbed by temperature lapse-rate
     corrections (Sect.~\ref{sec:atm}), time-dependent temperature offsets, and
@@
     computationally much more efficient than a fully three-dimensional model
-    with which the simulations of the scale presented here would not be
-    feasible.
+    with which simulations of the scale presented here would not be feasible.
 
@@
     with temperatures between 0 and 2\,\unit{{\degree}C}. Ablation is computed
-    from PDD, defined as an integral of temperatures above 0\,\unit{{\degree}C}
-    in one year.
+    from PDD, the integral of temperatures above 0\,\unit{{\degree}C}.
 
@@
     palaeo-temperature proxy records from the Greenland Ice Core Project
-    \citep[GRIP,][]{Dansgaard.etal.1993}, the European Project for Ice Coring
-    in Antarctica \citep[EPICA,][] {Jouzel.etal.2007}, and an oceanic sediment
-    core from the Iberian margin \citep[MD01-2444,][]{Martrat.etal.2007}.
-    Palaeo-temperature anomalies from the GRIP record were calculated from
+    \citep[GRIP;][]{Dansgaard.etal.1993}, the European Project for Ice Coring
+    in Antarctica \citep[EPICA;][] {Jouzel.etal.2007}, and an oceanic sediment
+    core from the Iberian margin \citep[MD01-2444;][]{Martrat.etal.2007}.
+    Palaeo-temperature anomalies from the GRIP record are calculated from
     oxygen isotope (\chem{\delta^{18}O}) measurements using a~quadratic
@@
     %
-    while temperature reconstructions from the EPICA and MD01-2444 records were
+    while temperature reconstructions from the EPICA and MD01-2444 records are
     provided as such. For each proxy record used and each of the parameter
@@
     circulation changes that governed moisture availability over the Alps
-    during the last glacial cycle. Thus other simulations use constant
+    during the last glacial cycle, of which little is known outside from the
+    LGM \citep[cf.][]{Wu.etal.2007, Strandberg.etal.2011, Ludwig.etal.2016}.
+    Thus another set of simulations presented here use constant
     precipitation, corresponding to $\psi=0$. In the rest of this paper, we
@@
     For the three palaeo-temperature records and the two palaeo-precipitation
-    parametrizations used, the model yield significant ice volume build-up
+    parametrizations used, the model yields significant ice volume build-up
     during MIS~4 and 2 (Fig.~\ref{fig:timeseries}b), corresponding to documented
@@
     areas (Fig.~\ref{fig:footprints}a). This result is inherent to the
-    palaeo-climate forcing approach used, which involved a linear scaling of
+    palaeo-climate forcing approach used, which involves a linear scaling of
     each palaeo-temperature anomaly record to a geomorphological
@@
     (Sect.~\ref{sec:paltemp}, Fig.~\ref{fig:footprints}a, black rectangle).
-    Concerning other areas, all simulations underestimate glaciated area in the
-    western part of the model domain, and overestimate it in the western
+    Outside this benchmark, all simulations underestimate glaciated area in the
+    western part of the model domain, and overestimate it in the eastern
     part, relative to the geomorphological ice limits
@@
     gradient in precipitation existing today (Fig.~\ref{fig:inputs}d) was
-    perhaps enhanced during the LGM, as indicated by pollen reconstruction
+    perhaps enhanced during the LGM, as indicated by pollen reconstructions
     \citep{Wu.etal.2007}. Finally, the LGM extent of Alpine glaciers might have
@@
     colours). However, with or without palaeo-precipitation corrections,
+    the GRIP and MD01-2444 forcing yield
     modelled ice extents (Fig.~\ref{fig:footprints}d and f) and volumes
-    (Fig.~\ref{fig:timeseries}b) are modelled to be higher during MIS~4 than
+    (Fig.~\ref{fig:timeseries}b) higher during MIS~4 than
     during MIS~2. This is in contradiction with geological evidence
@@
 
-    Based on the above considerations on timing of the LGM ice extent and ice
+    Based on the above considerations on timing of the LGM during MIS~2, and
+    modelled ice
     extent during MIS~4, we chose EPICA as our optimal palaeo-temperature
@@
     rapid ice volume fluctuations, we choose to include palaeo-precipitation
-    correction in the following comparison.
+    corrections in the following comparison.
 
@@
     This simulation is forced by the optimal EPICA palaeo-temperature record
-    (Sect.~\ref{sec:paltemp}) and includes palaeo-precipitation reduction
+    (Sect.~\ref{sec:paltemp}) and includes palaeo-precipitation reductions
     (Sect.~\ref{sec:palprec}). It uses an horizontal resolution of 1\,km. The
@@
     into a reconstruction, covering the entire Alps \citep{Ehlers.etal.2011},
-    and reproduced here (Fig.~4, red line) for comparison against model
-    results.
+    and reproduced here (Fig.~\ref{fig:lgmvel}, red line) for comparison
+    against model results.
 
@@
     cover in the eastern Alps, where the Mur, Drava and Sava glaciers are
-    modelled to extend tens of km beyond the mapped end-moraines
+    modelled to extend tens of km beyond the mapped ice limits
     (Fig.~\ref{fig:lgmvel}a). As previously discussed
@@
     mountains, while it is overestimated for the Vercors
-    \citep[Fig.~\ref{fig:lgmvel}a; cf.][Figs.~4.28, 4.32, and
-    4.33]{Coutterand.2010}. Thus, there exists a regional conformity between
+    \citep[Fig.~\ref{fig:lgmvel}a; cf.][Figs.~4.28, 4.32, and 4.33,
+    p.~322--321]{Coutterand.2010}. Thus, there exists a regional conformity
+    between
     the model-data discrepancies obtained for the main Alpine ice sheet and
@@
     \citep[e.g.,][]{Onde.1938, Penck.Bruckner.1909, Jackli.1962, Husen.1985,
-    Coutterand.2010, Kelly.etal.2004, Husen.2011}, and even to perhaps form
+    Coutterand.2010, Kelly.etal.2004, Husen.2011}, and even perhaps to form
     self-sustained ice domes \citep{Florineth.1998, Florineth.Schluchter.1998,
@@
     southerly ice flow across Mont-Cenis has been recognized
-    \citep[Fig.~3.18]{Onde.1938, Coutterand.2010}. Similarly, tranfluences have
-    been previously identified from the geomorphology across Simplon Pass
-    \citep{Kelly.etal.2004} and Brünig Pass \citep{Jackli.1962}.
+    \citep[Fig.~3.18, p.~284]{Onde.1938, Coutterand.2010}. Similarly,
+    tranfluences have been previously identified from the geomorphology across
+    Simplon Pass \citep{Kelly.etal.2004} and Brünig Pass \citep{Jackli.1962}.
 
@@
 
+    \note{A lot more could be said here but I want to keep a balance with
+          other parts of the discussion. Susan and I are currently working on a
+          more detailed comparison of modelled transfluences against documented
+          evidence across the Alps for a poster at SGM in November.}
+
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
@@
     when they experienced rapid retreat synchronously to lowering of the ice
-    surface in the mountains (\citealp[Fig.~5]{Wirsig.etal.2016},
+    surface in the mountains (\citealp[Fig.~5]{Wirsig.etal.2016};
     \citealp[Fig.~3]{Monegato.etal.2017}).
-
-    \note{Please edit or complete with more references if needed}
 
@@
     Dora Riparia, Ivrea, and Tagliamento glaciers reach an early maximum before
-    27\,ka, the Durance, Rhone, Inn, Enns, Ticino, and Garda glaciers reach
+    27\,ka; the Durance, Rhone, Inn, Enns, Ticino, and Garda glaciers reach
     their maximum thickness in phase with the global Alpine areal maximum
-    around 25\,ka, while the Isère, Como and Iseo glaciers reach a late maximum
+    around 25\,ka; while the Isère, Como and Iseo glaciers reach a late maximum
     after 24\,ka (Fig.~\ref{fig:timing}). Remarkably, peripheral ice caps on
     the Vercors, Jura Mountains, Black Forest and Hochschwab reach their
-    maximum extent even later, locally after 21\,ka (Fig.~\ref{fig:timing}).
+    maximum extent even later and locally after 21\,ka (Fig.~\ref{fig:timing}).
 
@@
     the homogeneous temperature and precipitation anomalies supplied as
-    palaeo-climate forcing used. The LGM timing differences modelled here are
+    palaeo-climate forcing. The LGM timing differences modelled here are
     thus not related to climate, but are inherent to modelled glacier dynamics.
@@
     Our results indicate that even for a relatively small ice sheet like that
-    found in the Alps, ka-scale differences in timing of the LGM can
+    found in the Alps, the LGM glacier extent corresponds to a transient
+    stage, while ka-scale differences in its timing can
     result from complex glacier dynamics. Further differences in LGM timing
@@
     under cold-based ice, sometimes for several glacial cycles
-    \citep{Stroeven.etal.2002}.
+    \citep{Stroeven.etal.2002}. Thus, the trimline could also mark the
+    maximum elevation of the transition to cold-based ice, or a late-glacial
+    ice surface elevation \citep[Fig.~1, p.~403]{Coutterand.2010}.
 
@@
     ice cover and call for a more accurate estimation of the thickness of the
-    upper layer cold ice in the Alps.
+    upper layer of cold ice in the Alps.
 
@@
     advances of several Alpine glaciers well into the major valleys and sometimes
-    even onto the foreland (Fig.~\ref{fig:profiles}), Supplementary animation).
+    even onto the foreland (Fig.~\ref{fig:profiles}, Supplementary animation).
 
@@
     Alpine reliefs and the beginning of the foothills, 6~times during the
-    simulation, sometimes retreating almost to its present day frontal position
+    simulation, and sometimes retreats almost completely between advances
     (Fig.~\ref{fig:profiles}b). The Rhone Glacier, fed by several high-altitude
     accumulation areas, advanced 8\,times onto the location of modern Lake
-    Geneva (Fig.~\ref{fig:profiles}c, d). The Ivrea Glacier, characterised by
-    a very steep catchment, and mutliple tributaries, show even a higher
+    Geneva (Fig.~\ref{fig:profiles}c and d). The Ivrea Glacier, characterised
+    by a very steep catchment, and mutliple tributaries, show even a higher
     variability and reaches close to its maximum position 6\,times throughout
-    the simulation. Finally, the Isère Glacier, with its complex system of
-    confluences and diffluences, needs longer time to build up and reaches the
-    foreland only twice during the major glaciations of MIS~4 and 2.
+    the simulation (Fig.~\ref{fig:profiles}e and f). Finally, the Isère
+    Glacier, with its complex system of confluences and diffluences, needs
+    longer time to build up and reaches the foreland only twice during the
+    major glaciations of MIS~4 and 2 (Fig.~\ref{fig:profiles}g and h).
 
@@
             change (Sect.~\ref{sec:extent}).
-      \item The LGM ice flow pattern in the Alps is largely controlled by
+      \item The LGM ice flow pattern in the Alps was largely controlled by
             subglacial topography, but several transfluences and two
-            self-sustained ice domes occured (Sect.~\ref{sec:flow}).
+            self-sustained ice domes may have occurred (Sect.~\ref{sec:flow}).
       \item The LGM was a transient stage in which glaciers were not in
@@
     These results are nevertheless limited by the simplicity of the climate
-    forcing and surface mass balance model applied, and uncertain ice physical
-    model parameters. Through more detailed sensitivity studies, targeted to
+    forcing and surface mass balance model applied, and uncertainties in ice
+    flow physics. Through more detailed sensitivity studies, targeted to
     specific aspects of the last glacial cycle ice dynamics in the Alps, future
@@
         \textbf{(b)} Mordern July standard deviation of daily mean temperature
-        from the ERA-Interim \citep[1979--2012]{Dee.etal.2011, Seguinot.2013}
+        from the ERA-Interim \citep[1979--2012;][]{Dee.etal.2011, Seguinot.2013}
         from the reference monthly climatology used to force the surface mass
@@
         \textbf{(d)} January precipitation from WorldClim
-        \citep[1960--1990]{Hijmans.etal.2005}.}
+        \citep[1960--1990;][]{Hijmans.etal.2005}.}
       \label{fig:inputs}
@@
         \textbf{(a--c)} Cumulative extent of modelled ice cover during MIS~2
-        (29--14\,ka), with (light colours) and without (dark colours)
+        (29--14\,ka), without (dark colours) and with (light colours)
         palaeo-precipitation corrections, and using temperature time-series
@@
         bilinearly interpolated along the corresponding profiles, showing
-        numerous cycles of advance and retreat over the last glacial cycle.
+        numerous cycles of advance and retreat over the last glacial cycle
+        modulated by subglacial topography and catchment geometry.
         Isolated patches indicate periodic surges from tributary glaciers.}
@@
 
-        GRIP, $\Delta P$ &&&&& 0.63 & $-$10.4 \\
+        GRIP, PP &&&&& 0.63 & $-$10.4 \\
 
@@
 
-        EPICA, $\Delta P$ &&&&& 1.33 & $-$12.2 \\
+        EPICA, PP &&&&& 1.33 & $-$12.2 \\
 
@@
 
-        MD01-2444, $\Delta P$ &&&&& 2.44 & $-$10.6 \\
+        MD01-2444, PP &&&&& 2.44 & $-$10.6 \\
 
@@
       %
-      %  Precip.   & cst. & $\Delta P$
-      %            & cst. & $\Delta P$
-      %            & cst. & $\Delta P$ \\
+      %  Precip.   & cst. & PP
+      %            & cst. & PP
+      %            & cst. & PP \\
       %

### 2017-10-20 articles/cycle/alps-cycle.tex
Add Susan in authors list.
Add references to new papers in TCD.

@@
 \Author[1]{Martin}{Funk}
-\Author[2]{Frank}{Preusser}
+\Author[2]{Susan}{Ivy-Ochs}
+\Author[3]{Frank}{Preusser}
 
@@
           ETH Zürich, Switzerland}
-\affil[2]{Institute of Earth and Environmental Sciences,
+\affil[2]{Laboratory of Ion Beam Physics, ETH Zürich, Switzerland}
+\affil[3]{Institute of Earth and Environmental Sciences,
           University of Freiburg, Germany}
@@
     Seguinot.etal.2016} and subsequently adapted for steady climate
-    \citep{Becker.etal.2016} and regional \citep{Jouvet.etal.2017a}
-    applications to the European Alps.
+    \citep{Becker.etal.2016} and regional \citep{Jouvet.etal.2017a,
+    Cohen.etal.2017} applications to the European Alps.
 
@@
     cold ice during the LGM, already found at high altitude in the much warmer
-    climate of today \citep{Suter.etal.2001}, is inevitable.
-    Thus we want to challenge the assumption that
+    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
+    inevitable. Thus we want to challenge the assumption that
     Alpine trimlines mark the maximum upper ice surface elevation of the LGM

### 2017-11-03 articles/cycle/alps-cycle.tex
Apply minor corrections from Guillaume.
I have left some apart to think what to do.

@@
     subglacial topography but include occasional transfluences and
-    self-sustained ice domes. Finally, the Last
+    self-sustained ice domes. Modelled maximum ice surface is about 900\,m higher
+    than observed trimline elevations, yet our simulation predicts little
+    erosion at high elevation due to cold ice conditions. Finally, the Last
     Glacial Maximum advance, often considered synchronous, is here modelled as
     a time-transgressive event, with some glacier lobes reaching their maximum
-    as early as 27\,ka, and some as late as 21\,ka. Modelled ice thickness is
-    about 900\,m higher than observed trimline elevations, yet our simulation
-    predicts little erosion at high elevation due to cold ice conditions.
+    as early as 27\,ka, and some as late as 21\,ka.
 
@@
     \emph{glacial cycles}
-    have succeeded each other with a 100\,ka periodicity \citep{Hays.etal.1976,
+    have succeeded each other with a periodicity about 100\,ka \citep{Hays.etal.1976,
     Augustin.etal.2004}. Nevertheless, this global signal is largely governed
@@
     spatially incomplete. Palaeo-ice sheets did not leave a continuous imprint
-    on the lanscape, and much of this evidence has been overprinted by
+    on the landscape, and much of this evidence has been overprinted by
     subsequent glacier re-advances and other geomorphological processes
@@
     layers in the ice, corresponding to vertical resolutions of 100 and
-    40\,\unit{m}, respectively.
+    40\,\unit{m}, respectively. Simulations ran for \todo{how long?}
+    using 36~processors.
 
@@
     their timing and amplitude varies significantly depending on the climate
-    forcing used. All six simulations overestimate
-    ice cover during the Younger Dryas \citep[cf. e.g.,][]{Ivy-Ochs.etal.2009}.
+    forcing used. All six simulations overestimate the late-glacial re-advance
+    during the Younger Dryas
+    \citep[\todo{timing}, cf. e.g.,][]{Ivy-Ochs.etal.2009}.
     This is partly because the 2\,km resolution is too coarse to resolve
@@
     overestimation resulting from the GRIP temperature forcing
-    (Fig.~\ref{fig:timeseries}b, blue curves) can not be explained by
-    resolution alone.
+    (Fig.~\ref{fig:timeseries}b, blue curves) certainly indicates that the
+    vigorous cooling experienced in Greenland is certainly not representative
+    of the Alps.
 
@@
     (Fig.~\ref{fig:timeseries}b, blue curves). The GRIP palaeo-temperature
-    forcing, on the other hand, yield two distinct total ice volume maxima
+    forcing, on the other hand, yields two distinct total ice volume maxima
     at 27.3/27.0 and 21.8/21.7\,ka followed by an early deglaciation of the
@@
     smaller scale on these peripheral ranges (Fig.~\ref{fig:timeseries}). On
-    the other hand, peripheral glaciation modelled in the EPICA simulation
-    (Fig.~\ref{fig:footprints}b) is in relatively good agreement with the
-    geomorphological reconstructions.
+    the other hand, peripheral glaciation modelled using the EPICA record
+    (Fig.~\ref{fig:footprints}b), which has less temperature variability,
+    is in relatively good agreement with the geomorphological reconstructions.
 
@@
     \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
-
     The EPICA palaeo-temperature forcing, on the other hand, yields a MIS~4
-    glaciation that is only slightly less expensive than the MIS~2 glaciation
+    glaciation that is only slightly less expansive than the MIS~2 glaciation
     (Fig.~\ref{fig:footprints}e). The sensitivity of the glaciated area to
@@
     modelled ice
-    extent during MIS~4, we chose EPICA as our optimal palaeo-temperature
+    extent during MIS~4, we select EPICA as our optimal palaeo-temperature
     record for a more detailed and higher-resolution comparison of modelled
@@
     251~enthalpy layers in the ice, corresponding to vertical resolutions of 50
-    and 20\,\unit{m}, respectively.
+    and 20\,\unit{m}, respectively. This simulation ran for 33~days using
+    576~processors.
 
@@
     The modelled flow pattern at 24.57\,ka is complex (Fig.~\ref{fig:lgmvel}a,
-    blue colour mapping). Fast-flow regions generally occur along the main
-    river valleys, while ice domes and ice divides are predominantly located
-    over major reliefs (Fig.~\ref{fig:lgmvel}a). Nevertheless, the model
-    results depict occasional ice flow across the modern water divides, i.e.
-    transfluences.
+    blue colour mapping). Ice velocities of several hundred metres per year,
+    characterising basal sliding, generally occur along the main river valleys,
+    while ice domes and ice divides are predominantly located over major
+    reliefs areas, where ice moves only by a few metres per year,
+    characterising internal deformation without basal sliding
+    (Fig.~\ref{fig:lgmvel}a). Nevertheless, the model results depict exceptions
+    to this general pattern as occasional ice flow across the modern water
+    divides, i.e. transfluences.
 
@@
     They result from differences in the subglacial topography, in particular
-    catchment sizes and bed roughness, of the different outlet glaciers.
-
-    PISM is a polythermal ice sheet model, i.e. it computes the distribution
-    of englacial temperatures resulting from heat advection, vertical
-    diffusion in ice and rock, and strain heat release, in three dimensions.
+    catchment sizes and hypsometry, of the different outlet glaciers.
+
     For large Alpine glaciers flowing in deep valleys, such as the Rhone,
@@
     balanced extent before thinning and receding towards the mountains as they
-    warm towards thermodynamical equilibrium.
+    warm towards thermodynamical equilibrium (supplementary animation).
 
@@
     found in the Alps, the LGM glacier extent corresponds to a transient
-    stage, while ka-scale differences in its timing can
+    stage, while ka-scale inherent differences in its timing can
     result from complex glacier dynamics. Further differences in LGM timing
@@
     observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
-    difference of 861\,m (Fig.~\ref{fig:trimlines}b). This result is
-    somewhat depedent on uncertain basal sliding and ice rheological parameters
+    difference of 861\,m (Fig.~\ref{fig:trimlines}b), wich is significant in
+    respect to the modelled regional ice thickness. This result is
+    somewhat dependent on uncertain basal sliding and ice rheological parameters
     to which the model sensitivity was not tested here. However, a previous
@@
     erosion (Fig.~\ref{fig:trimlines}c, white areas). On the other hand, on the
-    highest mountains ice cover is to thin and too static to form temperate
+    highest mountains ice cover is too thin and too static to form temperate
     ice, resulting in cold ice down to the bed, preventing potential erosion
@@
     climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
-    inevitable. Thus we want to challenge the assumption that
+    inevitable. In this context, the model results challenge the assumption that
     Alpine trimlines mark the maximum upper ice surface elevation of the LGM
@@
             both GRIP and MD01-2444 forcings result in MIS~4 glaciation well
-            beyong the mapped LGM ice limits  (Sect.~\ref{sec:footprints}).
+            beyond the mapped LGM ice limits  (Sect.~\ref{sec:footprints}).
     \end{itemize}
@@
       \item The LGM was a transient stage in which glaciers were not in
-            equilibrium with climate. Timing potentially varied across the
+            equilibrium with climate. Its timing potentially varied across the
             range due to inherent glacier dynamics (Sect.~\ref{sec:timing}).
@@
         and MIS~4 according to a~global compilation of benthic
-        \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}.}
+        \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}. The simulation
+        driven by the EPICA temperature yields smaller ice volume variability
+        and a more realistic timing of deglaciation.}
       \label{fig:timeseries}
@@
         \textbf{(a, c, e, g)} Approximate glacier flowlines drawn by hand
-        rougly following selected glacial valley centerlines.
+        rougly following selected glacial valley centerlines for the Rhine,
+        Rhone, Ivrea and Isère glaciers.
         \textbf{(b, d, f, h)} Evolution of modelled glacier extent in time,

### 2017-11-03 articles/cycle/alps-cycle.tex
Apply corrections from Martin.

@@
     well-studied. Such conditions make the region ideal for testing numerical
-    glacier models based on approximated ice flow physics against field-based
+    glacier models based on simplified ice flow physics against field-based
     reconstructions, and vice-versa.
@@
     WorldClim and the ERA-Interim reanalysis, and time-dependent temperature
-    offsets from multiple paleo-climate proxies, among which only the EPICA ice
+    offsets from multiple palaeo-climate proxies, among which only the EPICA ice
     core record yields glaciation during marine oxygen isotope stages~4
@@
     corrections (Sect.~\ref{sec:atm}), time-dependent temperature offsets, and
-    in some cases, time-dependent paleo-precipitation reductions
+    in some cases, time-dependent palaeo-precipitation reductions
     (Sect.~\ref{sec:climate}).
@@
     former Cordilleran ice sheet in good agreement with geological evidence
-    along the southern margin \citep{Seguinot.etal.2014} were weather station
+    along the southern margin \citep{Seguinot.etal.2014} where weather station
     density is lower than in the Alps. Finally, the last glacial cycle alpine
@@
     refer to different model runs by the name of the proxy record used for the
-    palaeo-temperature forcing. Model runs using paleo-precipitation
+    palaeo-temperature forcing. Model runs using palaeo-precipitation
     corrections are labelled PP.
@@
     cycles (Fig.~\ref{fig:timeseries}b, light colour curves). On the opposite
-    EPICA palaeo-temperature record has least variability the among records
+    EPICA palaeo-temperature record has least variability among the records
     used (Fig.~\ref{fig:timeseries}a, red curves). This results in fewer
@@
     This appears to result
-    from the paleo-precipitation reduction used to force the model
+    from the palaeo-precipitation reduction used to force the model
     (Fig.~\ref{fig:footprints}b), which is likely unrealistic in at least this
@@
     North America \citep[e.g.,][]{Kleman.etal.2010}, that pre-glacial landscapes
-    located well above the trimlines have been glaciated and preserved under
+    located well above the trimlines have been glaciated and preserved
     under cold-based ice, sometimes for several glacial cycles
@@
     \citep[cf.][]{Preusser.2004,Ivy-Ochs.etal.2008} glaciations, and regional
-    glacier dynamic controlled by variations in subglacial topography and
-    catchment geometry.
+    glacier dynamic controlled by spatial variations in subglacial topography
+    and catchment geometry.
 
@@
         difference is 861\,m.
-        \textbf{(c)} Modelled age of maximum ice thickness at the trimline
+        \textbf{(c)} Modelled age of maximum ice thickness at the observed trimline
         locations \citep[Table~1]{Kelly.etal.2004} in the upper Rhone valley

### 2017-11-04 articles/cycle/alps-cycle.tex
Apply minor corrections from Matthias.

@@
     the geological reconstructions.
-
     Despite the low variability of this Antarctic-based climate forcing, our
@@
     subglacial topography but include occasional transfluences and
-    self-sustained ice domes. Modelled maximum ice surface is about 900\,m higher
-    than observed trimline elevations, yet our simulation predicts little
-    erosion at high elevation due to cold ice conditions. Finally, the Last
+    self-sustained ice domes. Modelled maximum ice surface is about 900\,m
+    higher than observed trimline elevations in the interior of the Alps, yet
+    our simulation predicts little erosion at high elevation due to cold-based
+    ice. Finally, the Last
     Glacial Maximum advance, often considered synchronous, is here modelled as
@@
     century, more systematic studies of glacial features showed that alpine
-    glaciers extended well below their contemporary margins \citep{Venetz.1821}
+    glaciers extended well below their current margins \citep{Venetz.1821}
     and even onto the alpine foreland \citep{Charpentier.1841}, yielding the
@@
     proposed European and North American ice sheets. Although it was long
-    unclear whether there had been a single or multiple \emph{glaciations},
+    unclear whether there had been a single or multiple glaciations,
     this controversy ended with the large-scale mapping of two distinct moraine
@@
     indicating neither one, nor four, but rather tens of glacial and
-    interglacial periods. During the last 800000 years (800\,ka), these
+    interglacial periods. During the last 800\,000 years (800\,ka), these
     \emph{glacial cycles}
@@
     uncertainties typically increase with age, such that dated reconstructions
-    are strongly biased towards earlier glaciations \citep{Heyman.etal.2011}.
+    are strongly biased towards later glaciations \citep{Heyman.etal.2011}.
     The European Alps have been studied in more detail than any other glacial
@@
     nearly three hundred years, it thus remains incompletely known
-    \begin{itemize}
-      \item what climate evolution lead to the known maximum ice limits,
-      \item to which extent ice flow was controlled by subglacial topography,
-      \item what drove the different response of the individual lobes,
-      \item how far above the trimline was the ice surface located, and
-      \item how many advances occurred during the last glacial cycle.
-    \end{itemize}
+      (1) what climate evolution lead to the known maximum ice limits,
+      (2) to which extent ice flow was controlled by subglacial topography,
+      (3) what drove the different response of the individual lobes,
+      (4) how far above the trimline was the ice surface located, and
+      (5) how many advances occurred during the last glacial cycle.
 
@@
     Each simulation starts from assumed present-day ice thickness and
-    equilibrium temperature distribution at 120\,\unit{ka}, and runs to the
+    equilibrium ice and bedrock temperature at 120\,\unit{ka}, and runs to the
     present. Our modelling domain of 900 by 600\,\unit{km} encompasses the
@@
     dynamics over mountainous topographies similar to that of the European Alps
-    \citep{Golledge.etal.2012, Ziemen.etal.2016}. Beside, this hybrid scheme is
+    \citep{Golledge.etal.2012, Ziemen.etal.2016}. This hybrid scheme is
     computationally much more efficient than a fully three-dimensional model
@@
     sediment fills and lakes surface topography. However, they were corrected
-    for an estimation of present-day ice thickness according from modern
+    for an estimation of present-day ice thickness based on modern
     glacier outlines, surface topography and simplified ice physics
@@
     glaciers did not extend over marine areas where WorldClim data is missing.
-    Although low station density in the Arctic and missing data on marine
-    surfaces were major setbacks against accurate representation of Cordilleran
-    climate \citep{Seguinot.etal.2014}, these concerns are irrelevant to the
-    European Alps.
 
@@
     hand, and for differences between the basal topography of the ice flow
-    model, $b$, and the NARR reference topography, $b_{\text{ref}}$, on the
-    other hand. All simulations use an annual temperature lapse rate of
+    model, $b$, and the WorldClim reference topography, $b_{\text{ref}}$, on
+    the other hand. All simulations use an annual temperature lapse rate of
     $\gamma=6\,\unit{K\,km^{-1}}$.
@@
 
-    These simulations use an horizontal resolution of 2\,km. The vertical grid
+    These simulations use a horizontal resolution of 2\,km. The vertical grid
     consists of 31~temperature layers in the bedrock and up to 126~enthalpy
@@
     provided as such. For each proxy record used and each of the parameter
-    setup used in the sensitivity tests, palaeo-temperature anomalies are
+    set-ups used in the sensitivity tests, palaeo-temperature anomalies are
     scaled linearly (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}a) so
-    that the cumulative glaciated area of the Rhine glacier piedmont lobe
-    during Oxygen Marine Isotope Stage (MIS)~2 (29--14\,ka) is modelled
-    comparably to the reconstructions (Fig.~\ref{fig:footprints}a, black
-    rectangle).
+    that the modelled cumulative glaciated area of the Rhine glacier piedmont
+    lobe during Oxygen Marine Isotope Stage (MIS)~2 (29--14\,ka) is consistent
+    with reconstructions (Fig.~\ref{fig:footprints}a, black rectangle).
 
@@
 
-    Finally, in some simulations, precipitation was reduced with temperature in
+    Finally, in some simulations, precipitation was reduced with air temperature in
     order to simulate the potential rarefaction of atmospheric moisture in
-    colder climates. This was done using an empirical relationship derived from
+    colder climates \todo{ref.}. This was done using an empirical relationship derived from
     observed accumulation rates and oxygen isotopes concentrations in the GRIP
@@
     \begin{align}
-      {\Psi}_{\text{PP}}(t) = \exp[\psi{\Delta}T_{\text{TS}}(t)] \,.
+      {\Psi}_{\text{PP}}(t) = \exp[\psi{\Delta}T_{\text{TS}}(t)] \,,
     \end{align}
@@
     LGM \citep[cf.][]{Wu.etal.2007, Strandberg.etal.2011, Ludwig.etal.2016}.
-    Thus another set of simulations presented here use constant
+    Thus, another set of simulations presented here use constant
     precipitation, corresponding to $\psi=0$. In the rest of this paper, we
@@
     \citep[\todo{timing}, cf. e.g.,][]{Ivy-Ochs.etal.2009}.
-    This is partly because the 2\,km resolution is too coarse to resolve
-    Younger Dryas Alpine glaciers, but the large total ice volume
+    This is partly because the 2\,km resolution used in these simulations is
+    too coarse to resolve
+    Younger Dryas Alpine glaciers. However, the large total ice volume
     overestimation resulting from the GRIP temperature forcing
     (Fig.~\ref{fig:timeseries}b, blue curves) certainly indicates that the
-    vigorous cooling experienced in Greenland is certainly not representative
-    of the Alps.
+    vigorous cooling experienced in Greenland is not representative of the
+    Alps.
 
@@
     (Fig.~\ref{fig:timeseries}b, blue curves). The GRIP palaeo-temperature
-    forcing, on the other hand, yields two distinct total ice volume maxima
+    forcing, on the other hand, yield two distinct total ice volume maxima
     at 27.3/27.0 and 21.8/21.7\,ka followed by an early deglaciation of the
@@
     Palaeo-precipitation reductions dampen some of the small-scale variability,
-    but they have little effect on the millennial scale that characterise those
-    cycles (Fig.~\ref{fig:timeseries}b, light colour curves). On the opposite
-    EPICA palaeo-temperature record has least variability among the records
-    used (Fig.~\ref{fig:timeseries}a, red curves). This results in fewer
+    but they have little effect on the millennial scales that characterise those
+    cycles (Fig.~\ref{fig:timeseries}b, light colour curves). On the opposite,
+    the EPICA palaeo-temperature record has least variability among the records
+    used (Fig.~\ref{fig:timeseries}a, red curves). This results in smaller
     variability in the total ice volume, and more restrictive glaciations
@@
     temperature depression, here considered homogeneous, was actually lower in
-    the Eastern Alps and in the Western Alps, as previously shown by positive
+    the Eastern Alps than in the Western Alps, as previously shown by positive
     degree-day modelling of the central European palaeo-ice caps
@@
 
-    Beside this general pattern, there exist differences in glaciated area
+    Besides this general pattern, there exist differences in glaciated area
     depending on the palaeo-climate forcing used. The MD01-2444, and to a
-    greater extent, the GRIP, palaeo-temperature records, tend to overestimate
+    greater extent, the GRIP palaeo-temperature records, tend to overestimate
     MIS~2 ice cover on all peripheral ranges that surround the Alps
-    (Fig.~\ref{fig:footprints}a and c). This is particularly true when
+    (Fig.~\ref{fig:footprints}a and c). This is particularly true when no
     precipitation corrections are applied (bright colours). In fact, both
@@
     rapid ice volume fluctuations, we choose to include palaeo-precipitation
-    corrections in the following comparison.
-
-% ----------------------------------------------------------------------
-\section{Glacier dynamics}
+    corrections in the following higher-resolution run.
+
+% ----------------------------------------------------------------------
+\section{Results and discussion}
 \label{sec:results}
@@
     (Fig.~\ref{fig:lgmvel}b). A maximum glacierized area of $163.3 \times
-    10^{3}$\,\unit{km^2} is attained shortly aftewards at 24.57\,ka
+    10^{3}$\,\unit{km^2} is attained shortly afterwards at 24.57\,ka
     (Fig.~\ref{fig:lgmvel}b). Although the modelled timing of the LGM varies
     accross the mountain range (Sect.~\ref{sec:timing}), at 24.57\,ka nearly
-    all outlet glaciers extend to within a few km from their modelled maximum
-    stage (Fig.~\ref{fig:lgmvel}a, dashed orange line).
+    all outlet glaciers extend to within a few kilometres from their modelled
+    maximum stage (Fig.~\ref{fig:lgmvel}a, dashed orange line).
 
@@
     modelled to extend tens of km beyond the mapped ice limits
-    (Fig.~\ref{fig:lgmvel}a). As previously discussed
-    (Sect.~\ref{sec:footprints}), these discrepancies might indicate that the
+    (Fig.~\ref{fig:lgmvel}a). As previously discussed, these discrepancies
+    might indicate that the
     LGM climate was characterized by an east-west gradient in temperature
@@
     \citep[Fig.~4]{Becker.etal.2016} glaciation on the north slope of the Alps.
-    Using time-dependent palaeo-temperature forcing, progressive climate
-    cooling over several ka allow for warmer temperatures, closer to
+    Using time-dependent palaeo-temperature forcing, progressive atmospheric
+    cooling over several thousand years allow for warmer temperatures, closer to
     the thermodynamic equilibrium, within the ice and the bedrock. This results
@@
 
-    On a more regional level, the modelled maximum extent of the Ticino, Iseo,
-    Garda and Piave glaciers in the central southern Alps is constrained
+    On a more regional level, the maximum extent of the Ticino, Iseo,
+    Garda and Piave glaciers in the central southern Alps is modelled
     several kilometres within the mapped LGM moraines (Fig.~\ref{fig:lgmvel}a).
@@
     evidence on the Vercors, Chartreuse and Bauge prealpine reliefs,
-    the Jura Mountains, the Vosges Mountain, the Black Forest, the Bohemian
+    the Jura Mountains, the Vosges Mountains, the Black Forest, the Bohemian
     Forest and the Dinaric Alps (Fig.~\ref{fig:lgmvel}a). An independent ice
@@
     extent of peripheral ice caps is underestimated in the Vosges and Jura
-    mountains, while it is overestimated for the Vercors
+    mountains, but it is overestimated for the more meridional Vercors Massif
     \citep[Fig.~\ref{fig:lgmvel}a; cf.][Figs.~4.28, 4.32, and 4.33,
     p.~322--321]{Coutterand.2010}. Thus, there exists a regional conformity
-    between
-    the model-data discrepancies obtained for the main Alpine ice sheet and
-    those obtained for peripheral ice caps. Because peripheral ice cap have
+    between the model-data discrepancies obtained for the main Alpine ice sheet
+    and those obtained for peripheral ice caps, including too extensive
+    modelled ice cover to the East and extreme South-West, and too restrictive
+    modelled ice cover to the North-West. Because peripheral ice cap have
     very different glacier dynamics than the main ice sheet, this conformity
@@
     network of interconnected valley glaciers, primarily controlled by
-    subglacial topography. However, the geomorphology shows that ice was thick
+    subglacial topography. However, geomorphology shows that ice was thick
     enough to flow accross high mountain passes throughout the mountain range
@@
     Montgenèvre, Col du Mont-Cenis, Simplon Pass and Brünig Pass
-    (Fig.~\ref{fig:lgmvel}a). Although a transfluence across col de Montgenèvre
+    (Fig.~\ref{fig:lgmvel}a). Although a transfluence across Col de Montgenèvre
     was previously questioned \citep[Fig.~2]{Cossart.etal.2012}, evidence for
-    southerly ice flow across Mont-Cenis has been recognized
+    southerly ice flow across Col du Mont-Cenis has been recognized
     \citep[Fig.~3.18, p.~284]{Onde.1938, Coutterand.2010}. Similarly,
@@
     In the Eastern Alps, major transfluences are modelled to have occur for
-    instance across at Fern Pass, the Seefeld Saddle, the Gailberg Saddle, the
-    Kreuxberg Saddle, the Straniger Alp, and the Pyhrn Pass
+    instance across Fern Pass, the Seefeld Saddle, the Gailberg Saddle, the
+    Kreuxberg Saddle, the Straniger Alp, and Pyhrn Pass
     (Fig.~\ref{fig:lgmvel}a). Transfluences across Fern Pass and the Seefeld
@@
     ice flew across the Gailberg and Kreuzberg saddles \citep{Husen.1985}, and
-    the Pyhrn Pass \citep[Fig.~\ref{fig:lgmvel}a; cf.][Fig.~2.5]{Husen.2011}.
+    Pyhrn Pass \citep[Fig.~\ref{fig:lgmvel}a; cf.][Fig.~2.5]{Husen.2011}.
     On the other hand, no transfluence has previously been documented at the
@@
     Finally, self-sustained ice domes characteristic of ice caps are modelled
-    in two locations over Flüelapass and Ötztal (Fig.~\ref{fig:lgmvel}a).
+    in two locations over Flüela Pass and Ötztal (Fig.~\ref{fig:lgmvel}a).
     Except for too extensive ice cover in the easternmost part of the model
@@
     isotope dating techniques at multiple locations around the Alps (cf.
-    \citealp[Fig.~5]{Wirsig.etal.2016} for a review and the more recent
-    publications by \citealp{Monegato.etal.2017, Federici.etal.2017}). These
+    \citealp[Fig.~5]{Wirsig.etal.2016}, for a review and the more recent
+    publications by \citealp{Monegato.etal.2017} and
+    \citealp{Federici.etal.2017}). These
     data indicate that Alpine glaciers reached their maximum extent between 20
@@
     27\,ka; the Durance, Rhone, Inn, Enns, Ticino, and Garda glaciers reach
-    their maximum thickness in phase with the global Alpine areal maximum
+    their maximum thickness in phase with the overall Alpine areal maximum
     around 25\,ka; while the Isère, Como and Iseo glaciers reach a late maximum
@@
     Rhine and Garda glaciers, several thousand model years are needed to attain
-    a thermodynamical equilibrium between cold ice advection from the high
+    a thermodynamical equilibrium between cold-ice advection from the high
     accumulation areas, upward diffusion of geothermal heat, and heat release
@@
     balanced extent before thinning and receding towards the mountains as they
-    warm towards thermodynamical equilibrium (supplementary animation).
+    warm towards thermodynamical equilibrium (supplementary animation). On the
+    opposite, peripheral ice caps with basal topography restricted to high
+    elevations experience very low shear strain and virtually no basal sliding.
+    They advance regularly on a frozen bed during the entire cold period,
+    resulting in a late maximum stage (Fig.~\ref{fig:timing}).
 
@@
     anomaly gradients, not included in our model set-up, may have
-    counterbalanced or emphasize those modelled here.
+    counterbalanced or emphasized those modelled here.
 
@@
     valley troughs, represent the maximum elevation of the LGM ice surface,
-    maximum ice thickness in the Alps been reconstructed in several areas
+    maximum ice thickness in the Alps has been reconstructed in several areas
     \citep{Husen.1987, Florineth.1998, Florineth.Schluchter.1998,
@@
     \citep{Stroeven.etal.2002}. Thus, the trimline could also mark the
-    maximum elevation of the transition to cold-based ice, or a late-glacial
-    ice surface elevation \citep[Fig.~1, p.~403]{Coutterand.2010}.
+    maximum elevation of the transition from temperate to cold-based ice, or a
+    late-glacial ice surface elevation \citep[Fig.~1, p.~403]{Coutterand.2010}.
 
@@
     erosion (Fig.~\ref{fig:trimlines}c, white areas). On the other hand, on the
-    highest mountains ice cover is too thin and too static to form temperate
+    highest mountains, ice cover is too thin and too static to form temperate
     ice, resulting in cold ice down to the bed, preventing potential erosion
@@
     show remarkable agreement with the areal extent of regions that are
-    modelled to have experienced warm-based for less than 1\,ka (7\,\%) of the
+    modelled to have experienced warm-based ice for less than 1\,ka (7\,\%) of the
     MIS~2 (Fig.~\ref{fig:trimlines}c). Although this result calls for more
@@
     climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
-    inevitable. In this context, the model results challenge the assumption that
+    inevitable. In this context, our results challenge the assumption that
     Alpine trimlines mark the maximum upper ice surface elevation of the LGM
@@
     Preusser.Schluchter.2004}. Evidence for a MIS~4 glaciation is equally
-    sparse, and its timing remain uncertain \citep[e.g.,][]{Preusser.etal.2003,
+    sparse, and its timing remains uncertain \citep[e.g.,][]{Preusser.etal.2003,
     Link.Preusser.2006}.
@@
     (with or without corrections) applied, all simulations presented here
-    result in a high time variability in the total modelled ice volume
+    result in a high temporal variability in the total modelled ice volume
     (Fig.~\ref{fig:timeseries}). Using the optimal EPICA palaeo-temperature
@@
     including palaeo-precipitation reductions (Sect.~\ref{sec:palprec}), the
-    1-km resolution simulation also results in strong total ice volume
+    1\,km resolution simulation also results in strong total ice volume
     variability throughout the last glacial cycle (Fig.~\ref{fig:lgmvel}b).
@@
 
-    For instance the Rhine Glacier (Fig.~\ref{fig:profiles}a) reaches a lenght
-    exceeding 130\,km, roughly corresponding to location of the last major
-    Alpine reliefs and the beginning of the foothills, 6~times during the
+    For instance the Rhine Glacier (Fig.~\ref{fig:profiles}a) reaches a length
+    exceeding 130\,km, roughly corresponding to the location of the last major
+    Alpine reliefs and the beginning of the foothills, six times during the
     simulation, and sometimes retreats almost completely between advances
     (Fig.~\ref{fig:profiles}b). The Rhone Glacier, fed by several high-altitude
-    accumulation areas, advanced 8\,times onto the location of modern Lake
+    accumulation areas, advanced eight times onto of modern Lake
     Geneva (Fig.~\ref{fig:profiles}c and d). The Ivrea Glacier, characterised
-    by a very steep catchment, and mutliple tributaries, show even a higher
-    variability and reaches close to its maximum position 6\,times throughout
+    by a very steep catchment, and mutliple tributaries, shows even a higher
+    variability and reaches close to its maximum position six times throughout
     the simulation (Fig.~\ref{fig:profiles}e and f). Finally, the Isère
@@
     Despite the low temperature variability of the palaeo-climate forcing, and
-    reduced precipitation dampening glacier response, our simulation depict
+    reduced precipitation dampening glacier response, our simulation depicts
     the Alpine ice complex as highly dynamic, with many more than two or three
     \citep[cf.][]{Preusser.2004,Ivy-Ochs.etal.2008} glaciations, and regional
-    glacier dynamic controlled by spatial variations in subglacial topography
+    glacier dynamics controlled by spatial variations in subglacial topography
     and catchment geometry.
@@
     Using three different palaeo-temperature forcing records (GRIP, EPICA, and
-    MD01-2444, Sect.~\ref{sec:paltemp}), scaled to model the Rhine Glacier
+    MD01-2444, Sect.~\ref{sec:paltemp}), scaled to reproduce the Rhine Glacier
     piedmont lobe in agreement with the mapped LGM ice margin, and two
@@
     \begin{itemize}
-      \item Ice cover is generally underestimated in the Western Alps and
-            overestimated in the Eastern Alps, indicating that east-west
+      \item Ice cover is generally underestimated in the north-western Alps and
+            overestimated in the eastern and south-western Alps, indicating
+            that east-west
             gradients in temperature or precipitation change, absent from our
@@
     used here, as well as the general layout of this study, are in important
-    proportions the work of Irina Rogozhina, Arjen~P. Stroeven, Martin Margold and
-    Johan Kleman without whom this study would not have happened. As always, we
+    proportions the work of Irina Rogozhina, Arjen~P. Stroeven, Martin Margold
+    and Johan Kleman. We
     are very thankful to Constantine Khroulev, Ed Bueler, and Andy Aschwanden
@@
     the European Geoscience Union General Assembly~2017.
-
     This work was supported by the Swiss National Foundation grant
@@
         ice-bedrock interface.
-        \textbf{(b)} Mordern July standard deviation of daily mean temperature
-        from the ERA-Interim \citep[1979--2012;][]{Dee.etal.2011, Seguinot.2013}
+        \textbf{(b)} Modern July standard deviation \citep{Seguinot.2013} of
+        daily mean temperature from the ERA-Interim \citep[1979--2012;][]{Dee.etal.2011}
         from the reference monthly climatology used to force the surface mass
         balance (PDD) component of the ice sheet model.
-        \textbf{(c)} Modern July mean near-surface air temperature and
+        \textbf{(c)} Modern July mean near-surface air temperature, and
         \textbf{(d)} January precipitation from WorldClim
@@
         \textbf{(b)} Modelled total ice volume through the last 120~thousand
-        years (ka) expressed in meters of sea level equivalent (m~s.l.e.). Gray
+        years (ka) expressed in meters of sea-level equivalent (m~s.l.e.). Gray
         fields indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2
@@
         24.57\,ka before present, corresponding to the maximum modelled ice
-        cover. Modelled self-sustained ice domes (triangles) and major
-        transfluences (crosses). Modelled Last Glacial Maximum
+        cover. Modelled self-sustained ice domes (triangles) and some major
+        transfluences (crosses) are shown. Modelled Last Glacial Maximum
         (LGM) ice extent (dashed orange line) and geomorphological
@@
         Fig.~\ref{fig:timing}), compensated for bedrock deformation, against
-        observed trimline elevations \citep[Table~1]{Kelly.etal.2004}. Model
-        variables were bilinearly interpolated to the trimline locations.
+        observed trimline elevations \citep[Table~1]{Kelly.etal.2004}. Colours
+        show the age of maximum ice thickness (as in Fig.~\ref{fig:timing}).
+        Model variables were bilinearly interpolated to the trimline locations.
         \textbf{(b)} Histogram of differences between modelled LGM ice surface
@@
         difference is 861\,m.
-        \textbf{(c)} Modelled age of maximum ice thickness at the observed trimline
-        locations \citep[Table~1]{Kelly.etal.2004} in the upper Rhone valley
+        \textbf{(c)} Observed upper Rhone valley trimline locations
+        \citep[Table~1]{Kelly.etal.2004}, modelled age of maximum ice thickness
         (colour) and LGM ice surface elevation (200\,m contours). Hatches
@@
       \caption{%
-        \textbf{(a, c, e, g)} Approximate glacier flowlines drawn by hand
-        rougly following selected glacial valley centerlines for the Rhine,
-        Rhone, Ivrea and Isère glaciers.
+        \textbf{(a, c, e, g)} Profile lines roughly following valley
+        centerlines for the Rhine, Rhone, Ivrea and Isère glaciers.
         \textbf{(b, d, f, h)} Evolution of modelled glacier extent in time,

### 2017-11-04 articles/cycle/alps-cycle.tex
Apply minor corrections from Frank.

@@
     glaciations in the Alps \citep{Penck.Bruckner.1909}. More recent studies
-    of glaciofluvial stratigraphy in the foreland indicate up to eight
-    glaciations \citep{Ivy-Ochs.etal.2008, Preusser.etal.2011}.
+    of glaciofluvial stratigraphy in the foreland indicate up to 15 glaciations
+    \citep{Schluchter.1988, Ivy-Ochs.etal.2008, Preusser.etal.2011}.
 
     From the mid-twentieth
-    century, palaeoclimate records extracted from deep sea sediments and
+    century, palaeo-climate records extracted from deep sea sediments and
     ice cores have provided a much more detailed picture of the Earth
@@
     Shackleton.Opdyke.1973, Dansgaard.etal.1993, Augustin.etal.2004},
-    indicating neither one, nor four, but rather tens of glacial and
+    indicating tens of glacial and
     interglacial periods. During the last 800\,000 years (800\,ka), these
@@
     (Fig.~\ref{fig:timeseries}b) higher during MIS~4 than
-    during MIS~2. This is in contradiction with geological evidence
+    during MIS~2, which is not supported by geological evidence
     \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
@@
     Preusser.Schluchter.2004}. Evidence for a MIS~4 glaciation is equally
-    sparse, and its timing remains uncertain \citep[e.g.,][]{Preusser.etal.2003,
-    Link.Preusser.2006}.
+    sparse, and its timing remains uncertain
+    \citep[e.g.,][]{Preusser.etal.2007, Link.Preusser.2006}.
 

### 2017-11-04 articles/cycle/alps-cycle.tex
Apply minor corrections from Susan.

@@
     Despite the low variability of this Antarctic-based climate forcing, our
-    simulation depicts a highly dynamic ice sheet, showing that alpine glaciers
+    simulation depicts a highly dynamic ice sheet, showing that Alpine glaciers
     may have advanced many times over the foreland during the last glacial
@@
     former presence. In the mid-nineteenth
-    century, more systematic studies of glacial features showed that alpine
+    century, more systematic studies of glacial features showed that Alpine
     glaciers extended well below their current margins \citep{Venetz.1821}
-    and even onto the alpine foreland \citep{Charpentier.1841}, yielding the
+    and even onto the Alpine foreland \citep{Charpentier.1841}, yielding the
     idea that, under colder temperatures, expansive \emph{ice sheets} had once
@@
     are strongly biased towards later glaciations \citep{Heyman.etal.2011}.
-    The European Alps have been studied in more detail than any other glacial
-    region, but they are no exception to these rules. Although sparse
+    In the European Alps, although sparse
     geological traces indicate that the last glacial cycle may have comprised
@@
     different piedmont
-    lobes exist \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear wether they
+    lobes exists \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear whether it
     relate to climate or glacier dynamics \citep{Monegato.etal.2017} or
@@
     approximates glacier sliding and deformation (Sect.~\ref{sec:model}), to
-    model alpine glacier dynamics through the last glacial cycle (120--0\,ka),
+    model Alpine glacier dynamics through the last glacial cycle (120--0\,ka),
     a period for which palaeo-temperature proxies are available, albeit non
-    regional. In an attempt to analyse the long-standing questions outline
+    regional. In an attempt to analyse the longstanding questions outlined
     above, we test the model sensitivity to multiple palaeo-climate forcing
@@
     that includes local isostasy, elastic lithosphere flexure and viscous
-    astenosphere deformation in an infinite half-space
-    \citep{Lingle.Clark.1985,Bueler.etal.2007}. The astenosphere viscosity,
-    $\nu_{\mathrm{m}}=2.2\times10^{20}$\,\unit{Pa\,s}, the astenosphere
+    asthenosphere deformation in an infinite half-space
+    \citep{Lingle.Clark.1985,Bueler.etal.2007}. The asthenosphere viscosity,
+    $\nu_{\mathrm{m}}=2.2\times10^{20}$\,\unit{Pa\,s}, the asthenosphere
     density, $\rho_{\mathrm{m}}=3300$\,\unit{kg\,m^{-3}}, and the lithosphere
@@
     along the southern margin \citep{Seguinot.etal.2014} where weather station
-    density is lower than in the Alps. Finally, the last glacial cycle alpine
+    density is lower than in the Alps. Finally, the last glacial cycle Alpine
     glaciers did not extend over marine areas where WorldClim data is missing.
@@
     scaled linearly (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}a) so
-    that the modelled cumulative glaciated area of the Rhine glacier piedmont
-    lobe during Oxygen Marine Isotope Stage (MIS)~2 (29--14\,ka) is consistent
-    with reconstructions (Fig.~\ref{fig:footprints}a, black rectangle).
+    that the modelled cumulative glaciated area during Oxygen Marine Isotope
+    Stage (MIS)~2 (29--14\,ka), within a rectangular region covering the Rhine
+    glacier piedmont lobe (Fig.~\ref{fig:footprints}a, black rectangle) is
+    consistent with the glaciated area of the geological reconstruction
+    \citep{Ehlers.etal.2011}.
 
@@
     circulation changes that governed moisture availability over the Alps
-    during the last glacial cycle, of which little is known outside from the
+    during the last glacial cycle, of which little is known apart from the
     LGM \citep[cf.][]{Wu.etal.2007, Strandberg.etal.2011, Ludwig.etal.2016}.
@@
     but they have little effect on the millennial scales that characterise those
-    cycles (Fig.~\ref{fig:timeseries}b, light colour curves). On the opposite,
+    cycles (Fig.~\ref{fig:timeseries}b, light colour curves). In contrast,
     the EPICA palaeo-temperature record has least variability among the records
@@
     and MD01-2444 palaeo-temperature records, Alpine glaciers are modelled to
-    extend well beyond the reach of documented LGM end-moraines and glacial
-    erratic terrain (Fig.~\ref{fig:footprints}d and f). Because, in both cases,
+    extend well beyond the reach of documented LGM end-moraines
+    (Fig.~\ref{fig:footprints}d and f). Because, in both cases,
     the modelled glaciation extent corresponds to brief cold spells in the
@@
     modelled ice extents (Fig.~\ref{fig:footprints}d and f) and volumes
-    (Fig.~\ref{fig:timeseries}b) higher during MIS~4 than
+    (Fig.~\ref{fig:timeseries}b) considerably larger during MIS~4 than
     during MIS~2, which is not supported by geological evidence
@@
     lobes of central-Alpine glaciers, which are slightly less extensive during
-    MIS~2 and 4 (Fig.~\ref{fig:footprints}b and e).
+    MIS~2 and 4 under reduced precipitation (Fig.~\ref{fig:footprints}b and e).
 
@@
 
-    The LGM extent of alpine glaciers has been mapped with varying level of
+    The LGM extent of Alpine glaciers has been mapped with varying level of
     detail across the mountain range \citep{Penck.Bruckner.1909, Jackli.1962,
-    Husen.1987, Bini.etal.2009, Coutterand.2010, Bavec.Verbic.2011,
-    Buoncristiani.Campy.2011, Husen.2011}. These maps have been compiled
+    Husen.1987, Bini.etal.2009, Husen.2011, Coutterand.2010, Bavec.Verbic.2011,
+    Buoncristiani.Campy.2011}. These maps have been compiled
     into a reconstruction, covering the entire Alps \citep{Ehlers.etal.2011},
@@
 
-    The modelled total ice volume reaches a maximum of $122.5 \times
-    10^{3}$\,\unit{km^3}, or 300.4\,mm of sea-level equivalent, at 24.56\,ka
-    (Fig.~\ref{fig:lgmvel}b). A maximum glacierized area of $163.3 \times
+    The modelled total ice volume reaches a maximum of $123 \times
+    10^{3}$\,\unit{km^3}, or 300\,mm of sea-level equivalent, at 24.56\,ka
+    (Fig.~\ref{fig:lgmvel}b). A maximum glacierized area of $163 \times
     10^{3}$\,\unit{km^2} is attained shortly afterwards at 24.57\,ka
@@
     \citep[cf.][]{Wu.etal.2007} anomalies relative to present, or an east-west
-    gradient in variables not accounted for by the PDD model.
+    gradient in variables that are not accounted for by the PDD model.
 
@@
     The model reproduces peripheral ice caps where documented by geological
-    evidence on the Vercors, Chartreuse and Bauge prealpine reliefs,
+    evidence on the Vercors, Chartreuse and Bauge Prealpine reliefs,
     the Jura Mountains, the Vosges Mountains, the Black Forest, the Bohemian
@@
     cap also covers the Hochwart massif during most of the simulation, yet it
-    is engulfed by alpine glaciers during the LGM to become a peripheral ice
+    is engulfed by Alpine glaciers during the LGM to become a peripheral ice
     dome \citep[Fig.~\ref{fig:lgmvel}a; cf.][Fig.~2.5]{Husen.2011}. The LGM
@@
     modelled ice cover to the East and extreme South-West, and too restrictive
-    modelled ice cover to the North-West. Because peripheral ice cap have
+    modelled ice cover to the North-West. Because peripheral ice caps have
     very different glacier dynamics than the main ice sheet, this conformity
@@
     \citep[e.g.,][]{Onde.1938, Penck.Bruckner.1909, Jackli.1962, Husen.1985,
-    Coutterand.2010, Kelly.etal.2004, Husen.2011}, and even perhaps to form
+    Husen.2011, Kelly.etal.2004, Coutterand.2010}, and even perhaps to form
     self-sustained ice domes \citep{Florineth.1998, Florineth.Schluchter.1998,
@@
     tranfluences have been previously identified from the geomorphology across
-    Simplon Pass \citep{Kelly.etal.2004} and Brünig Pass \citep{Jackli.1962}.
+    Simplon Pass \citep{Kelly.etal.2004} and from the distribution of glacial
+    erratics across Brünig Pass \citep{Jackli.1962}.
 
@@
     \citealp{Federici.etal.2017}). These
-    data indicate that Alpine glaciers reached their maximum extent between 20
-    and 26\,ka (calibrated \chem{^{14}C} and \chem{^{10}Be} ages), but also
+    data indicate that Alpine glaciers reached their maximum extent between 26
+    and 20\,ka (calibrated \chem{^{14}C} and \chem{^{10}Be} ages), but also
     that terminal lobes stayed in the foreland until as late as 22 to 17\,ka,
@@
     maximum extent at different ages (Fig.~\ref{fig:timing}). For instance, the
-    Dora Riparia, Ivrea, and Tagliamento glaciers reach an early maximum before
+    Dora Riparia, Dora Baltea, and Tagliamento glaciers reach an early maximum before
     27\,ka; the Durance, Rhone, Inn, Enns, Ticino, and Garda glaciers reach
@@
     anomaly gradients, not included in our model set-up, may have
-    counterbalanced or emphasized those modelled here.
+    counterbalanced or enhanced those modelled here.
 
@@
 
-    In the upper Rhone valley, the maximum ice surface elevation reached
+    In the upper Rhone Valley, the maximum ice surface elevation reached
     during MIS~2 in our simulation, corrected for bedrock deformation, is
@@
 
-    Beside, our model results depict a polythermal LGM Alpine ice cover.
+    Our model results depict a polythermal LGM Alpine ice cover.
     Due to sub-freezing temperatures applied in the climate forcing of the
@@
     accumulation areas, advanced eight times onto of modern Lake
-    Geneva (Fig.~\ref{fig:profiles}c and d). The Ivrea Glacier, characterised
+    Geneva (Fig.~\ref{fig:profiles}c and d). The Dora Baltea Glacier, characterised
     by a very steep catchment, and mutliple tributaries, shows even a higher
@@
     \citep[cf.][]{Preusser.2004,Ivy-Ochs.etal.2008} glaciations, and regional
-    glacier dynamics controlled by spatial variations in subglacial topography
-    and catchment geometry.
+    glacier dynamics controlled by spatial variations in catchment size and
+    hypsometry.
 
@@
         a~global compilation of benthic \chem{\delta^{18}O} records
-        \citep{Lisiecki.Raymo.2005}.}
+        \citep{Lisiecki.Raymo.2005}.
+        \todo{Rename Ivrea to Dora Baltea.}}
       \label{fig:lgmvel}
@@
         and modelled glacierized area during the LGM (coloured curve). The LGM
-        is here modelled as a time-transgressive event.}
+        is here modelled as a time-transgressive event.
+        \todo{Rename Ivrea to Dora Baltea.}}
       \label{fig:timing}
@@
         \textbf{(a, c, e, g)} Profile lines roughly following valley
-        centerlines for the Rhine, Rhone, Ivrea and Isère glaciers.
+        centerlines for the Rhine, Rhone, Dora Baltea and Isère glaciers.
         \textbf{(b, d, f, h)} Evolution of modelled glacier extent in time,
@@
         modulated by subglacial topography and catchment geometry.
-        Isolated patches indicate periodic surges from tributary glaciers.}
+        Isolated patches indicate periodic surges from tributary glaciers.
+        \todo{Rename Ivrea to Dora Baltea.}}
       \label{fig:profiles}

### 2017-11-04 articles/cycle/alps-cycle.tex
Add discussion of Dansgaard-Oeschger events.

@@
     Palaeo-precipitation reductions dampen some of the small-scale variability,
-    but they have little effect on the millennial scales that characterise those
-    cycles (Fig.~\ref{fig:timeseries}b, light colour curves). In contrast,
+    but they have little effect on the millennial scales that characterise
+    those cycles (Fig.~\ref{fig:timeseries}b, light colour curves). In
+    particular, strong millenial variability recorded in GRIP
+    \chem{\delta^{18}O} (Dansgaard-Oeschger events) have large repercussions
+    on the modelled Alpine ice volume, including around six glaciations of LGM
+    magnitude (Fig.~\ref{fig:timeseries}b, blue curves), which is not supported
+    by geologic evidence \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
+    Dansgaard-Oeschger events, however, have been recorded in Central European
+    lake sediments \citep{Wohlfarth.etal.2008} and cave speleothems
+    \citep{Spotl.Mangini.2002, Luetscher.etal.2015}. Our results indicate that
+    their magnitude in Central Europe was likely lower than that recorded in
+    GRIP \chem{\delta^{18}O}, which either implies smaller associated
+    palaeo-temperature fluctuations in Central Europe than Greenland, or
+    temperature-independant controls on GRIP \chem{\delta^{18}O}
+    \todo{ref}. In contrast,
     the EPICA palaeo-temperature record has least variability among the records
@@
     glacier dynamics controlled by spatial variations in catchment size and
-    hypsometry.
+    hypsometry. Importantly, Dansgaard-Oeschger events, recorded in Central
+    Europe \citep{Spotl.Mangini.2002, Wohlfarth.etal.2008,
+    Luetscher.etal.2015} but absent from the EPICA palaeo-temperature forcing,
+    may have induced a more dynamic glacier response than that modelled here.
 
@@
       \item Alpine glaciers were potentially very dynamic. They quickly
-            responded to climate fluctuation and some potentially advanced many
-            times over the foreland during the last glacial cycle
+            responded to climate fluctuations and some potentially advanced
+            many times over the foreland during the last glacial cycle
             (Sect.~\ref{sec:glaciations}).
@@
     forcing and surface mass balance model applied, and uncertainties in ice
-    flow physics. Through more detailed sensitivity studies, targeted to
+    flow physics. For instance, additional climate variability over Central
+    Europe may have caused additional glaciations to those modelled here.
+    Through more detailed sensitivity studies, targeted to
     specific aspects of the last glacial cycle ice dynamics in the Alps, future

### 2017-11-05 articles/cycle/alps-cycle.tex
Remove notes to co-authors.

@@
 
-    \note{This section contains\\
-        Table~\ref{tab:params} -- Model parameters.\\
-        Fig.~\ref{fig:inputs} -- Climate and geothermal model inputs.}
-
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
@@
 
-    \note{This is embarrassing, but I realised too late that the formula in
-          \citet{Mey.etal.2016} was wrong.}
-
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
@@
 % ----------------------------------------------------------------------
-
-    \note{This section contains\\
-      Table~\ref{tab:records} -- Palaeo-climate records.\\
-      Fig.~\ref{fig:timeseries} -- Low-resolution time series.\\
-      Fig.~\ref{fig:footprints} -- Low-resolution ice cover.}
 
@@
 
-    \note{This section contains\\
-        Fig.~\ref{fig:lgmvel} -- Snapshot at 21\,ka and volume time series.\\
-        Fig.~\ref{fig:timing} -- Timing of the LGM and area time series.\\
-        Fig.~\ref{fig:trimlines} -- LGM ice thickness compared to trimlines.\\
-        Fig.~\ref{fig:profiles} -- Individual glacier extent profiles.}
-
     In this section, we compare the model output to geological evidence from
@@
     of a topography-controlled ice field, and that of a self-sustained ice cap.
-
-    \note{A lot more could be said here but I want to keep a balance with
-          other parts of the discussion. Susan and I are currently working on a
-          more detailed comparison of modelled transfluences against documented
-          evidence across the Alps for a poster at SGM in November.}
 

### 2017-11-05 articles/cycle/alps-cycle.tex
Move comment on rigidity error into footnote.

@@
     the Alps most closely reproducing observed modern uplift rates
-    \citep[Supplementary Fig.~7]{Mey.etal.2016}. Unfortunately, the latter was
-    computed using the erroneous formula, ${D=YE^3/12(1-\nu)}$
+    \citep[Supplementary Fig.~7]{Mey.etal.2016}.\footnote{Lithosphere rigidity
+    was computed using the erroneous formula, ${D=YE^3/12(1-\nu)}$, in
     \citep{Mey.etal.2016}. The correct formula is ${D=YE^3/12(1-\nu^2)}$.
     Using the Young's modulus, $Y=100$\,GPa, and the Poisson ratio, $\nu=0.25$
-    \citep{Mey.etal.2016}, our value for lithosphere rigidity thus correspond
+    \citep{Mey.etal.2016}, our value for lithosphere rigidity thus corresponds
     to an effective elastic thickness of the lithosphere of $E=52.415$\,km,
-    instead of the 50\,km value used by \citet{Mey.etal.2016}.
+    instead of the 50\,km value used by \citet{Mey.etal.2016}.}
 

### 2017-11-05 articles/cycle/alps-cycle.tex
Remove parameter values from methods text.

@@
     defined in our case by
-    ${\tau_{\mathrm{e}}}^2=\frac{1}{2}\mathrm{tr}(\vec{\tau}^2)$. We set the
-    power-law exponent, $n=3$, according to
-    \citet[p.~55--57]{Cuffey.Paterson.2010}. The ice
+    ${\tau_{\mathrm{e}}}^2=\frac{1}{2}\mathrm{tr}(\vec{\tau}^2)$. The ice
     softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$,
@@
     using the Clapeyron relation, $T_{\text{pa}}=T-{\beta}p$.
-    $R=8.31441$\,\unit{J\,mol^{-1}\,K^{-1}} is the ideal gas constant, and
-    $A_{\mathrm{c}}=2.847 \times 10^{-13}$\,\unit{Pa^{-3}\,s^{-1}},
-    $A_{\mathrm{w}}=2.356 \times 10^{-2}$\,\unit{Pa^{-3}\,s^{-1}},
-    $Q_{\mathrm{c}}=6.0 \times 10^4$\,\unit{J\,mol^{-1}}, and
-    $Q_{\mathrm{w}}=11.5 \times 10^4$\,\unit{J\,mol^{-1}} are constant parameters
+    $R$ is the ideal gas constant, and $A_{\mathrm{c}}$, $A_{\mathrm{w}}$,
+    $Q_{\mathrm{c}}$, and $Q_{\mathrm{w}}$ are constant parameters
     corresponding to values measured below and above a~critical temperature
-    threshold $T_{\mathrm{c}}=-10$\,\unit{{\degree}C}
-    \citep[p.~72]{Cuffey.Paterson.2010}. We set the Clapeyron constant,
-    $\beta=7.9\times 10^{-8}$\,\unit{K\,Pa^{-1}}, according to
-    \citet{Luthi.etal.2002}, and the water fraction coefficient, $f=181.25$,
-    according to \citet{Lliboutry.Duval.1985}. The water fraction, $\omega$, is
+    threshold $T_{\mathrm{c}}$. The Clapeyron constant,
+    $\beta$, the water fraction coefficient, $f$,
+    The water fraction, $\omega$, is
     capped at a~maximum value of 0.01, above which no measurements are
     available \citep[Eq.~5.7]{Lliboutry.Duval.1985, Greve.1997}. Finally, $E$
-    is a~non-dimensional enhancement factor which take the values,
-    $E_{\text{SIA}}=2$, in the SIA and $E_{\text{SSA}}=1$, in the SSA, as
-    recommended for Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010}.
+    is an SIA enhancement factor which reflects the observed rheology of
+    Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010}.
 
@@
     relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to the
-    sliding velocity, $\vec{v}_{\mathrm{b}}$. The pseudo-plastic sliding
-    exponent, $q=0.25$, and the threshold velocity,
-    $v_{\text{th}}=100$\,\unit{m\,a^{-1}}, are set to values successfully used
-    to model the Greenland ice sheet \citep{Aschwanden.etal.2013}. The yield
+    sliding velocity, $\vec{v}_{\mathrm{b}}$. The yield
     stress, $\tau_{\mathrm{c}}$, is modelled using the Mohr--Coulomb criterion,
@@
     \begin{align}
-      \tau_{\mathrm{c}} = c_0 + N\,\tan{\phi} \,,
+      \tau_{\mathrm{c}} = c_0 + N\,\tan{\phi} \,.
     \end{align}
     %
-    where the till cohesion, $c_0=0$, was oftentimes measured to be
-    negligible \citep[p.~268]{Tulaczyk.etal.2000, Cuffey.Paterson.2010}. We use
-    a~constant till friction angle, $\phi=30$\unit{\degree}, corresponding to
-    the average of values presented in \citet[p.~268]{Cuffey.Paterson.2010}.
 
@@
     \begin{align}
-      N = \delta P_0 \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,,
+      N = \delta P_0 \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,.
     \end{align}
     %
-    where $\delta=0.02$ sets the minimum ratio between the effective and
-    overburden pressures. Parameter values for the till reference void ratio,
-    $e_0=0.69$, and the till compressibility coefficient,
-    $C_{\mathrm{c}}=0.12$, were set to the only measurements available to our
-    knowledge \citep{Tulaczyk.etal.2000}. The amount of water at the base, $W$,
-    varies from zero to $W_{\text{max}}=2$\,m, a~threshold above which
+    The amount of water at the base, $W$,
+    varies from zero to $W_{\text{max}}$, a~threshold above which
     additional melt water is assumed to drain off instantaneously.
@@
     \citep{Lingle.Clark.1985,Bueler.etal.2007}. The asthenosphere viscosity,
-    $\nu_{\mathrm{m}}=2.2\times10^{20}$\,\unit{Pa\,s}, the asthenosphere
-    density, $\rho_{\mathrm{m}}=3300$\,\unit{kg\,m^{-3}}, and the lithosphere
-    elastic rigidity, $D=1.389 \times 10^{24}$\,\unit{N\,m}, were set according to
+    $\nu_{\mathrm{m}}$, the asthenosphere density, $\rho_{\mathrm{m}}$, and the
+    lithosphere elastic rigidity, $D$, were set according to
     results from glacial isostatic adjustment modelling of deglacial rebound in
@@
                 & --
-                & \citet{Cuffey.Paterson.2010} \\
+                & \citet[p.~55--57]{Cuffey.Paterson.2010} \\
 
@@
                 & \unit{Pa^{-3}\,s^{-1}}
-                & \citet{Cuffey.Paterson.2010} \\
+                & \citet[p.~72]{Cuffey.Paterson.2010} \\
 
@@
                 & \unit{J\,mol^{-1}}
-                & \citet{Cuffey.Paterson.2010} \\
+                & \citet[p.~72]{Cuffey.Paterson.2010} \\
 
@@
                 & --
-                & \citet{Cuffey.Paterson.2010} \\
+                & \citet[p.~77]{Cuffey.Paterson.2010} \\
 
@@
                 & \degree
-                & \citet{Cuffey.Paterson.2010} \\
+                & \citet[p.~268]{Cuffey.Paterson.2010} \\
 

### 2017-11-05 articles/cycle/alps-cycle.tex
Remove equations and cut down methods.

@@
     Ice deformation is governed by the constitutive law for ice
-    \citep{Glen.1952, Nye.1953},
-    %
-    \begin{align}
-      \vec{\dot{\epsilon}} = A\,\tau_{\mathrm{e}}^{n-1}\,\vec{\tau} \,.
-    \end{align}
-    %
-    where $\vec{\dot{\epsilon}}$ is the strain-rate tensor, $\vec{\tau}$ the
-    deviatoric stress tensor, and $\tau_{\mathrm{e}}$ the effective stress
-    defined in our case by
-    ${\tau_{\mathrm{e}}}^2=\frac{1}{2}\mathrm{tr}(\vec{\tau}^2)$. The ice
-    softness coefficient, $A$, depends on ice temperature, $T$, pressure, $p$,
-    and water content, $\omega$, through a~piece-wise Arrhenius-type law,
-    %
-    \begin{align}
-    &A = E\cdot
-      \begin{cases}
-        A_{\mathrm{c}} \,e^\frac{-Q_{\mathrm{c}}}{RT_{\text{pa}}}
-            & \text{if}\ T_   {\text{pa}}  <  T_{\mathrm{c}} \,, \\
-        A_{\mathrm{w}} (1+f\omega)\,e^\frac{-Q_{\mathrm{w}}}{RT_{\text{pa}}}
-            & \text{if}\ T_   {\text{pa}} \ge T_{\mathrm{c}} \,,
-      \end{cases}
-    \end{align}
-    %
-    where $T_{\text{pa}}$ is the pressure-adjusted ice temperature calculated
-    using the Clapeyron relation, $T_{\text{pa}}=T-{\beta}p$.
-    $R$ is the ideal gas constant, and $A_{\mathrm{c}}$, $A_{\mathrm{w}}$,
-    $Q_{\mathrm{c}}$, and $Q_{\mathrm{w}}$ are constant parameters
-    corresponding to values measured below and above a~critical temperature
-    threshold $T_{\mathrm{c}}$. The Clapeyron constant,
-    $\beta$, the water fraction coefficient, $f$,
-    The water fraction, $\omega$, is
+    \citep{Glen.1952, Nye.1953}. Ice softness depends on ice temperature,
+    pressure, and water content, through an enthalpy scheme
+    \citep[Table~\ref{tab:params};]{Aschwanden.etal.2012} and piece-wise
+    Arrhenius-type rheology representative of Holocene polar ice
+    \citep[Table~\ref{tab:params};][p.~77]{Cuffey.Paterson.2010}.
+    The englacial water fraction is
     capped at a~maximum value of 0.01, above which no measurements are
-    available \citep[Eq.~5.7]{Lliboutry.Duval.1985, Greve.1997}. Finally, $E$
-    is an SIA enhancement factor which reflects the observed rheology of
-    Holocene polar ice \citep[p.~77]{Cuffey.Paterson.2010}.
+    available \citep[Eq.~5.7]{Lliboutry.Duval.1985, Greve.1997}.
 
@@
 
-    A~pseudo-plastic sliding law,
-    %
-    \begin{align}
-      \vec{\tau}_{\mathrm{b}} = -\tau_{\mathrm{c}}
-        \frac{\vec{v}_{\mathrm{b}}}
-             {{v_{\text  {th}}}^q\,|\vec{v}_{\mathrm{b}}|^{1-q}} \,,
-    \end{align}
-    %
-    relates the bed-parallel shear stresses, $\vec{\tau}_{\mathrm{b}}$, to the
-    sliding velocity, $\vec{v}_{\mathrm{b}}$. The yield
-    stress, $\tau_{\mathrm{c}}$, is modelled using the Mohr--Coulomb criterion,
-    %
-    \begin{align}
-      \tau_{\mathrm{c}} = c_0 + N\,\tan{\phi} \,.
-    \end{align}
-    %
-
-    Effective pressure, $N$, is related to the ice overburden stress, $P_0=\rho
-    gh$, and the modelled amount of subglacial water, using a~formula derived
-    from laboratory experiments with till extracted from the base of Ice Stream
-    B in West Antarctica \citep{Tulaczyk.etal.2000, Bueler.Pelt.2015},
-    %
-    \begin{align}
-      N = \delta P_0 \, 10^{(e_0/C_{\mathrm{c}}) (1 - (W/W_{\text{max}}))} \,.
-    \end{align}
-    %
-    The amount of water at the base, $W$,
-    varies from zero to $W_{\text{max}}$, a~threshold above which
-    additional melt water is assumed to drain off instantaneously.
+    A~pseudo-plastic sliding law relates the bed-parallel shear stresses to the
+    sliding velocity. The yield stress is modelled using the Mohr--Coulomb
+    criterion using a constant basal friction angle (Table~\ref{tab:params}).
+    Effective pressure is related to the ice overburden stress and the modelled
+    amount of subglacial water, using a~formula derived from laboratory
+    experiments with till extracted from the base of Ice Stream B in West
+    Antarctica \citep[Table~\ref{tab:params};][]{Tulaczyk.etal.2000,
+    Bueler.Pelt.2015}. When the till becomes saturated, additional melt water
+    is assumed to drain off instantaneously.
 
@@
     asthenosphere deformation in an infinite half-space
-    \citep{Lingle.Clark.1985,Bueler.etal.2007}. The asthenosphere viscosity,
-    $\nu_{\mathrm{m}}$, the asthenosphere density, $\rho_{\mathrm{m}}$, and the
-    lithosphere elastic rigidity, $D$, were set according to
+    \citep{Lingle.Clark.1985, Bueler.etal.2007}. Model parameters were set
+    according to
     results from glacial isostatic adjustment modelling of deglacial rebound in
@@
     around the expected value, $T_{\mathrm{m}}$. It is expressed by an
-    error-function formulation \citep{Calov.Greve.2005},
-    %
-    \begin{align}
-      {\text{PDD}} = \int_{t_1}^{t_2} \mathrm{d}t
-        \left[\frac{\sigma}{\sqrt{2\pi}}
-                \exp\left({-\frac{T_{\mathrm{m}}^2}{2\sigma^2}}\right)
-              + \frac{T_{\mathrm{m}}}{2} \, {\text{erfc}}
-                \left(-\frac{T_{\mathrm{m}}}{\sqrt{2}\sigma}\right)\right] \,,
-    \end{align}
-    %
+    error-function formulation \citep{Calov.Greve.2005}
     which is numerically approximated using week-long sub-intervals. In
@@
       &P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \cdot
-                                 {\Psi}_{\text{PP}}(t) \,, \\
+                                 {\Psi}_{\text{PP}}(t) \,,
     \end{align}
@@
     \citealp[Fig.~5]{Wirsig.etal.2016}, for a review and the more recent
-    publications by \citealp{Monegato.etal.2017} and
-    \citealp{Federici.etal.2017}). These
+    publications by \citealp{Monegato.etal.2017, Federici.etal.2017}). These
     data indicate that Alpine glaciers reached their maximum extent between 26
@@
       \caption{%
-        Parameter values used in the ice sheet model.}
+        Parameter values used in the ice sheet model. Symbols refer to
+        equations used by \citet{Seguinot.2014, Seguinot.etal.2016}.}
       \label{tab:params}

### 2017-11-05 articles/cycle/alps-cycle.tex
Address one more small comment from Martin.

@@
 
-    Finally, in some simulations, precipitation was reduced with air temperature in
+    Finally, in some simulations (hereafter labelled PP), precipitation was
+    reduced with air temperature in
     order to simulate the potential rarefaction of atmospheric moisture in
@@
     refer to different model runs by the name of the proxy record used for the
-    palaeo-temperature forcing. Model runs using palaeo-precipitation
-    corrections are labelled PP.
+    palaeo-temperature forcing.
 

### 2017-11-05 articles/cycle/alps-cycle.tex
Add references about MIS 4 from Susan.

@@
     \citep{Mey.etal.2016}, our value for lithosphere rigidity thus corresponds
-    to an effective elastic thickness of the lithosphere of $E=52.415$\,km,
+    to an effective elastic thickness of the lithosphere of $E=52.4$\,km,
     instead of the 50\,km value used by \citet{Mey.etal.2016}.}
@@
     during MIS~2, which is not supported by geological evidence
-    \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
+    \citep{Preusser.2004, Ivy-Ochs.etal.2008, Husen.Reitner.2011,
+    Barrett.etal.2017, Haldimann.etal.2017}. Gravel deposits from the Rhine
+    Glacier indicate an early last glacial cycle foreland glaciation, but the
+    extent reached is less than that in MIS~2 \citep{Keller.Krayss.2010}.
+    Dropstones in lake sediments indicate a late MIS~4 advance restricted to
+    the upper Inn Valley \citep{Barrett.etal.2017}. For the Reuss Glacier, no
+    evidence for a pre-LGM glacier advance during the last glacial cycle is
+    indicated \citep{Haldimann.etal.2017}.
+
     The EPICA palaeo-temperature forcing, on the other hand, yields a MIS~4
@@
     The LGM extent of Alpine glaciers has been mapped with varying level of
-    detail across the mountain range \citep{Penck.Bruckner.1909, Jackli.1962,
-    Husen.1987, Bini.etal.2009, Husen.2011, Coutterand.2010, Bavec.Verbic.2011,
-    Buoncristiani.Campy.2011}. These maps have been compiled
+    detail across the Western \citep{Coutterand.2010,
+    Buoncristiani.Campy.2011}, Central \citep{Jackli.1962, Bini.etal.2009}
+    \todo{ref. Hantke}, and Eastern Alps \citep{Penck.Bruckner.1909,
+    Husen.1987, Husen.2011, Bavec.Verbic.2011} \todo{ref. German Geologische
+    Übersichtskarte 1:200,000}. Some of these maps have been compiled
     into a reconstruction, covering the entire Alps \citep{Ehlers.etal.2011},
@@
     data indicate that the last glacial cycle may have comprised two or even
-    three periods of glacier growth and decay \citep{Preusser.2004,Ivy-Ochs.etal.2008}.
+    three periods of glacier growth and decay \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
     Luminescence dating from two sites in the central northern foreland
@@
     Preusser.Schluchter.2004}. Evidence for a MIS~4 glaciation is equally
-    sparse, and its timing remains uncertain
-    \citep[e.g.,][]{Preusser.etal.2007, Link.Preusser.2006}.
+    sparse \citep[Sect.~\ref{sec:footprints};][]{Keller.Krayss.2010,
+    Barrett.etal.2017, Haldimann.etal.2017}, and its timing remains uncertain
+    \citep[e.g.,][]{Link.Preusser.2006, Preusser.etal.2007}.
 

### 2017-11-05 articles/cycle/alps-cycle.tex
Turn large maps sideways.
Make a few more previously ignore fixes.

@@
     Shackleton.Opdyke.1973, Dansgaard.etal.1993, Augustin.etal.2004},
-    indicating tens of glacial and
-    interglacial periods. During the last 800\,000 years (800\,ka), these
-    \emph{glacial cycles}
+    indicating tens of glacial and interglacial periods. During the last
+    800\,000 years before present (800\,ka), these \emph{glacial cycles}
     have succeeded each other with a periodicity about 100\,ka \citep{Hays.etal.1976,
@@
     The palaeo-climate forcing was adapted (Sect.~\ref{sec:paltemp})
-    to model this maximum configuration in broad agreement with the geological
+    to model the maximum configuration of the Rhine Glacier in broad agreement
+    with the geological
     reconstruction of the LGM extent (Fig.~\ref{fig:lgmvel}a, solid red line),
-    yet local discrepancies remain. As already outlined for low resolution
+    yet discrepancies remain elsewhere. As already outlined for low resolution
     runs (Sect.~\ref{sec:footprints}), the extent of glaciation in the
@@
     Our results indicate that even for a relatively small ice sheet like that
-    found in the Alps, the LGM glacier extent corresponds to a transient
-    stage, while ka-scale inherent differences in its timing can
-    result from complex glacier dynamics. Further differences in LGM timing
-    caused by heterogeneous climatic anomalies, such as precipitation
-    anomaly gradients, not included in our model set-up, may have
-    counterbalanced or enhanced those modelled here.
+    found in the Alps, the LGM glacier extent corresponds to a transient stage,
+    while millennial scale, spatial differences in its timing can result not
+    only from climate variability but also from complex glacier dynamics.
+    Heterogeneous climatic anomalies, such as temperature or precipitation
+    anomaly gradients, not included in our model set-up, have perhaps induced
+    further spatial differences in the timing of the LGM Alpine ice sheet,
+    which may have counterbalanced or enhanced those modelled here.
 
@@
 
-    \begin{figure*}
+    \begin{sidewaysfigure*}
       \centerline{\includegraphics{alpcyc_hr_lgmvel}}
@@
         \citep{Lisiecki.Raymo.2005}.
-        \todo{Rename Ivrea to Dora Baltea.}}
+        \todo{Enlarge to full width, rename Ivrea to Dora Baltea.}}
       \label{fig:lgmvel}
-    \end{figure*}
-
-    \begin{figure*}
+    \end{sidewaysfigure*}
+
+    \begin{sidewaysfigure*}
       \centerline{\includegraphics{alpcyc_hr_timing}}
@@
         is here modelled as a time-transgressive event.
-        \todo{Rename Ivrea to Dora Baltea.}}
+        \todo{Enlarge to full width, rename Ivrea to Dora Baltea.}}
       \label{fig:timing}
-    \end{figure*}
+    \end{sidewaysfigure*}
 

### 2017-11-05 articles/cycle/alps-cycle.tex
Add a bunch of todo notes regarding figures.

@@
     meltwater-induced \emph{sliding} at the base \citep[\S532]{Saussure.1779},
-    and viscous \emph{deformation} within the ice body \citep{Forbes.1846b}. As
+    and viscous \emph{deformation} within the ice body \citep{Forbes.1846b}.
+    \todo{check ref. history from Guillaume}. As
     glaciers flow and slide across their bed, they transport rock debris and
@@
         \textbf{(d)} January precipitation from WorldClim
-        \citep[1960--1990;][]{Hijmans.etal.2005}.}
+        \citep[1960--1990;][]{Hijmans.etal.2005}.
+        \todo{larger sub-fig. labels in all figs.}}
       \label{fig:inputs}
@@
         driven by the EPICA temperature yields smaller ice volume variability
-        and a more realistic timing of deglaciation.}
+        and a more realistic timing of deglaciation.
+        \todo{move (b) label from behind legend.}}
       \label{fig:timeseries}
@@
         (71--57\,ka). Only the simulation driven by the EPICA temperature
-        time-series yields reasonable MIS~4 ice cover.}
+        time-series yields reasonable MIS~4 ice cover.
+        \todo{Try to hatch MIS~4}}
       \label{fig:footprints}
@@
         \citep{Lisiecki.Raymo.2005}.
-        \todo{Enlarge to full width, rename Ivrea to Dora Baltea.}}
+        \todo{Enlarge to full width, make geological outline and transfluences
+              more visible, rename Ivrea to Dora Baltea.}}
       \label{fig:lgmvel}
@@
         Isolated patches indicate periodic surges from tributary glaciers.
-        \todo{Rename Ivrea to Dora Baltea.}}
+        \todo{Move glacier names into timeseries panels, change y-labels to
+              glacier length, add new profiles, rename Ivrea to Dora Baltea.}}
       \label{fig:profiles}

### 2017-11-06 articles/cycle/alps-cycle.tex
Complete rereading. What a long text...

@@
     meltwater-induced \emph{sliding} at the base \citep[\S532]{Saussure.1779},
-    and viscous \emph{deformation} within the ice body \citep{Forbes.1846b}.
+    and viscous \emph{deformation} within the ice body \citep{Forbes.1846b}
     \todo{check ref. history from Guillaume}. As
@@
     Seguinot.etal.2016} and subsequently adapted for steady climate
-    \citep{Becker.etal.2016} and regional \citep{Jouvet.etal.2017a,
-    Cohen.etal.2017} applications to the European Alps.
+    \citep{Becker.etal.2016} and regional \citep{Jouvet.etal.2017a}
+    applications to the European Alps.
 
@@
     Centre for Medium-Range Weather Forecasts Reanalysis Interim
-    \citep[ERA-Interim;][]{Dee.etal.2011} over the period 1979--2012,
-    perturbed by temperature lapse-rate
+    \citep[ERA-Interim;][]{Dee.etal.2011}, perturbed by temperature lapse-rate
     corrections (Sect.~\ref{sec:atm}), time-dependent temperature offsets, and
@@
     Each simulation starts from assumed present-day ice thickness and
-    equilibrium ice and bedrock temperature at 120\,\unit{ka}, and runs to the
-    present. Our modelling domain of 900 by 600\,\unit{km} encompasses the
+    equilibrium ice and bedrock temperature at 120\,ka, and runs to the
+    present. Our modelling domain of 900 by 600\,km encompasses the
     entire Alpine range (Fig.~\ref{fig:inputs}). The simulations were run on
-    two distinct grids, using horizontal resolutions of 2 and 1\,\unit{km}.
+    two distinct grids, using horizontal resolutions of 2 and 1\,km.
     All parameter values are summarized in Table~\ref{tab:params}.
@@
     computationally much more efficient than a fully three-dimensional model
+    \citep[e.g.,][]{Cohen.etal.2017}
     with which simulations of the scale presented here would not be feasible.
@@
     pressure, and water content, through an enthalpy scheme
-    \citep[Table~\ref{tab:params};]{Aschwanden.etal.2012} and piece-wise
+    \citep[Table~\ref{tab:params};][]{Aschwanden.etal.2012} and a piece-wise
     Arrhenius-type rheology representative of Holocene polar ice
@@
     enthalpy model. Temperature is computed in the ice and in the bedrock down
-    to a~depth of 3\,\unit{km} below the glacier base, where it is
+    to a~depth of 3\,km below the glacier base, where it is
     conditioned by a~lower boundary geothermal heat flux estimate from multiple
@@
     was computed using the erroneous formula, ${D=YE^3/12(1-\nu)}$, in
-    \citep{Mey.etal.2016}. The correct formula is ${D=YE^3/12(1-\nu^2)}$.
+    \citet{Mey.etal.2016}. The correct formula is ${D=YE^3/12(1-\nu^2)}$.
     Using the Young's modulus, $Y=100$\,GPa, and the Poisson ratio, $\nu=0.25$
@@
     layers in the ice, corresponding to vertical resolutions of 100 and
-    40\,\unit{m}, respectively. Simulations ran for \todo{how long?}
+    40\,m, respectively. Simulations ran for \todo{how long?}
     using 36~processors.
@@
     scaled linearly (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}a) so
-    that the modelled cumulative glaciated area during Oxygen Marine Isotope
-    Stage (MIS)~2 (29--14\,ka), within a rectangular region covering the Rhine
+    that the modelled cumulative glaciated area during MIS~2 (29--14\,ka)
+    and within a rectangular region covering the Rhine
     glacier piedmont lobe (Fig.~\ref{fig:footprints}a, black rectangle) is
@@
     Thus, another set of simulations presented here use constant
-    precipitation, corresponding to $\psi=0$. In the rest of this paper, we
-    refer to different model runs by the name of the proxy record used for the
-    palaeo-temperature forcing.
+    precipitation, corresponding to $\psi=0$.
 
@@
     reconstruction of the area glaciated by the Rhine Glacier piedmont lobe
-    (Sect.~\ref{sec:paltemp}, Fig.~\ref{fig:footprints}a, black rectangle).
+    (Sect.~\ref{sec:paltemp}; Fig.~\ref{fig:footprints}a, black rectangle).
     Outside this benchmark, all simulations underestimate glaciated area in the
@@
     (Sect.~\ref{sec:paltemp}) and includes palaeo-precipitation reductions
-    (Sect.~\ref{sec:palprec}). It uses an horizontal resolution of 1\,km. The
+    (Sect.~\ref{sec:palprec}). It uses a horizontal resolution of 1\,km. The
     vertical grid consists of 61~temperature layers in the bedrock and up to
     251~enthalpy layers in the ice, corresponding to vertical resolutions of 50
-    and 20\,\unit{m}, respectively. This simulation ran for 33~days using
+    and 20\,m, respectively. This simulation ran for 33~days using
     576~processors.
@@
     10^{3}$\,\unit{km^2} is attained shortly afterwards at 24.57\,ka
-    (Fig.~\ref{fig:lgmvel}b). Although the modelled timing of the LGM varies
+    (Fig.~\ref{fig:lgmvel}a). Although the modelled timing of the LGM varies
     accross the mountain range (Sect.~\ref{sec:timing}), at 24.57\,ka nearly
@@
     present. Although this may appear as a contradiction with previous,
-    constant-climate modelling results \citep{Becker.etal.2016}, it is the
-    result of introducing a time-dependent palaeo-temperature forcing. Constant
+    constant-climate modelling results \citep{Becker.etal.2016}, it follows
+    from introducing a time-dependent palaeo-temperature forcing. Constant
     climate forcing systematically resulted in extraneous
@@
     ice flew across the Gailberg and Kreuzberg saddles \citep{Husen.1985}, and
-    Pyhrn Pass \citep[Fig.~\ref{fig:lgmvel}a; cf.][Fig.~2.5]{Husen.2011}.
+    Pyhrn Pass \citep[Fig.~2.5]{Husen.2011}.
     On the other hand, no transfluence has previously been documented at the
@@
     from strong basal shear strain. Temperature evolution is, in part, limited
-    by the slow warming of subglacial bedrock, that has been previously cooled
-    downed by subfreezing air temperature before glacier advance.
+    by the slow warming of the subglacial bedrock, that has been previously
+    cooled downed by subfreezing air temperature before glacier advance.
     For larger glaciers, this thermodynamical equilibrium is typically not yet
@@
     late-glacial ice surface elevation \citep[Fig.~1, p.~403]{Coutterand.2010}.
+    However, no evidence for cold-based glaciation has yet been reported in the
+    Alps.
 
@@
     sensitivity study conducted with PISM and a very similar model set-up on
-    the Cordillera ice sheet show that varying ice rheological parameters
+    the Cordilleran ice sheet show that varying ice rheological parameters
     resulted in relative errors in modelled total ice volume of 25\,\unit{\%}
@@
     highest mountains, ice cover is too thin and too static to form temperate
-    ice, resulting in cold ice down to the bed, preventing potential erosion
+    ice, resulting in cold ice down to the bed and preventing potential erosion
     (Fig.~\ref{fig:trimlines}c, hatched areas).
@@
     In the upper Rhone valley, the geographical location of observed trimlines
-    show remarkable agreement with the areal extent of regions that are
+    show a remarkable agreement with the areal extent of regions that are
     modelled to have experienced warm-based ice for less than 1\,ka (7\,\%) of the
@@
     entire Alpine range and specific sensitivity studies to relevant basal
-    sliding and ice rheological parameters, the presence of an upper layer of
+    sliding and ice rheological parameters, and to the uncertain subglacial
+    topography the presence of an upper layer of
     cold ice during the LGM, already found at high altitude in the much warmer
@@
     advances of several Alpine glaciers well into the major valleys and sometimes
-    even onto the foreland (Fig.~\ref{fig:profiles}, Supplementary animation).
+    even onto the foreland (Fig.~\ref{fig:profiles}; supplementary animation).
 
@@
     Alpine reliefs and the beginning of the foothills, six times during the
-    simulation, and sometimes retreats almost completely between advances
+    simulation, and sometimes retreats almost completely between two advances
     (Fig.~\ref{fig:profiles}b). The Rhone Glacier, fed by several high-altitude
-    accumulation areas, advanced eight times onto of modern Lake
+    accumulation areas, advanced eight times onto modern Lake
     Geneva (Fig.~\ref{fig:profiles}c and d). The Dora Baltea Glacier, characterised
@@
     Using three different palaeo-temperature forcing records (GRIP, EPICA, and
-    MD01-2444, Sect.~\ref{sec:paltemp}), scaled to reproduce the Rhine Glacier
+    MD01-2444; Sect.~\ref{sec:paltemp}), scaled to reproduce the Rhine Glacier
     piedmont lobe in agreement with the mapped LGM ice margin, and two
     different palaeo-precipitation parametrizations (with and without
-    precipitation reductions, Sect.~\ref{sec:palprec}), we find that only the
+    precipitation reductions; Sect.~\ref{sec:palprec}), we find that only the
     EPICA palaeo-temperature record yields model results in agreement with
@@
             25.2/24.6\,ka (without/with palaeo-precipitation reduction),
-            followed by a standstill of major piedmont lobe in the forelands
+            followed by a standstill of major piedmont lobes in the forelands
             until 17.3\,ka, both compatible with much of the dating results,
@@
             compatible with geological evidence during MIS~4 and 2, whereas
-            both GRIP and MD01-2444 forcings result in MIS~4 glaciation well
+            both GRIP and MD01-2444 forcing result in MIS~4 glaciation well
             beyond the mapped LGM ice limits (Sect.~\ref{sec:footprints}).
@@
 \codedataavailability{%
-    PISM is available as open-source software at http://pism-docs.org/.
-    Please contact the corresponding author for model output data.
-    \idea{Should we make model output available somewhere?}}
+    PISM is available as open-source software at http://pism-docs.org/. Upon
+    publication, we will be happy to make aggregated model output available
+    through The Cryosphere. Please contact the corresponding author for mode
+    detailed model output data.}
 
@@
     the manuscript. M.~Huss provided modern ice thickness data. All authors
-    contributed to the text. The idea for this study stems in part from an
-    excursion organised by F.~Preusser in the central Alps in 2012.}
+    contributed to interpreting results and improving the text. The idea for
+    this study stems in part from an excursion organised by F.~Preusser in the
+    central Alps in 2012.}
 
@@
     Sheet which is part of J.~Seguinot's PhD thesis. The experimental design
-    used here, as well as the general layout of this study, are in important
-    proportions the work of Irina Rogozhina, Arjen~P. Stroeven, Martin Margold
-    and Johan Kleman. We
-    are very thankful to Constantine Khroulev, Ed Bueler, and Andy Aschwanden
-    for providing constant help and development with PISM, in particular with
-    recent issues with the computation of ice temperature (Github issue
-    no.~371) and with the computation of bedrock deformation (Github issues
-    no.~370 and 377). We are equally thankful to Giovanni Monegato and Marc
-    Luetscher for very insightful discussions of preliminary model results during
-    the European Geoscience Union General Assembly~2017.
-    This work was supported by the Swiss National Foundation grant
-    no.~200021-153179/1 to M.~Funk.
-    Computer resources were provided by the Swiss National Supercomputing
-    Centre (CSCS) allocation no.~s573 to J.~Seguinot.
+    used here, as well as the general layout of this study, are largely
+    inspired by Irina Rogozhina and Arjen~P. Stroeven. We are very thankful to
+    Constantine Khroulev, Ed Bueler, and Andy Aschwanden for providing constant
+    help and development with PISM, in particular with recent issues with the
+    computation of ice temperature (Github issue no.~371) and with the
+    computation of bedrock deformation (Github issues no.~370 and 377). We are
+    equally thankful to Giovanni Monegato and Marc Luetscher for very
+    insightful discussions of preliminary model results during the European
+    Geoscience Union General Assembly~2017. This work was supported by the
+    Swiss National Foundation grant no.~200021-153179/1 to M.~Funk. Computer
+    resources were provided by the Swiss National Supercomputing Centre (CSCS)
+    allocation no.~s573 to J.~Seguinot.
 
@@
                 & 263.15
-                & \unit{K}
+                & K
                 & \citet{Paterson.Budd.1982} \\
@@
                 & 273.15
-                & \unit{K}
+                & K
                 & -- \\
@@
                 & 275.15
-                & \unit{K}
+                & K
                 & -- \\
@@
         refers to the resulting mean temperature anomaly during the period 32
-        to~22\,\unit{ka} after scaling.}
+        to~22\,ka after scaling.}
       \label{tab:records}

### 2017-11-14 figures/alpcyc_hr_profiles.py
Rename Ivrea to Dora Baltea in figures.

@@ -8,7 +8,7 @@ import cartopy.io.shapereader as shpreader
 
 # parameters
 regions = ['rhine', 'rhone', 'ivrea', 'isere', 'inn', 'taglia']
-labels = ['Rhine', 'Rhone', 'Ivrea', u'Isère', 'Inn', 'Tagliamento']
+labels = ['Rhine', 'Rhone', 'Dora Baltea', u'Isère', 'Inn', 'Tagliamento']
 colors = ['blue', 'green', 'red', 'orange', 'purple', 'brown']
 colors = [ut.pl.palette['dark'+hue] for hue in colors]
 

### 2017-11-14 articles/cycle/alps-cycle.tex
Update todo notes in manuscript.

@@
     meltwater-induced \emph{sliding} at the base \citep[\S532]{Saussure.1779},
-    and viscous \emph{deformation} within the ice body \citep{Forbes.1846b}
-    \todo{check ref. history from Guillaume}. As
+    and viscous \emph{deformation} within the ice body \citep{Forbes.1846b}. As
     glaciers flow and slide across their bed, they transport rock debris and
@@
     order to simulate the potential rarefaction of atmospheric moisture in
-    colder climates \todo{ref.}. This was done using an empirical relationship derived from
+    colder climates. This was done using an empirical relationship derived from
     observed accumulation rates and oxygen isotopes concentrations in the GRIP
@@
     palaeo-temperature fluctuations in Central Europe than Greenland, or
-    temperature-independant controls on GRIP \chem{\delta^{18}O}
-    \todo{ref}. In contrast,
+    temperature-independant controls on GRIP \chem{\delta^{18}O}. In contrast,
     the EPICA palaeo-temperature record has least variability among the records
@@
         \textbf{(d)} January precipitation from WorldClim
-        \citep[1960--1990;][]{Hijmans.etal.2005}.
-        \todo{larger sub-fig. labels in all figs.}}
+        \citep[1960--1990;][]{Hijmans.etal.2005}.}
       \label{fig:inputs}
@@
         driven by the EPICA temperature yields smaller ice volume variability
-        and a more realistic timing of deglaciation.
-        \todo{move (b) label from behind legend.}}
+        and a more realistic timing of deglaciation.}
       \label{fig:timeseries}
@@
         (71--57\,ka). Only the simulation driven by the EPICA temperature
-        time-series yields reasonable MIS~4 ice cover.
-        \todo{Try to hatch MIS~4}}
+        time-series yields reasonable MIS~4 ice cover.}
       \label{fig:footprints}
@@
         a~global compilation of benthic \chem{\delta^{18}O} records
-        \citep{Lisiecki.Raymo.2005}.
-        \todo{Enlarge to full width, make geological outline and transfluences
-              more visible, rename Ivrea to Dora Baltea.}}
+        \citep{Lisiecki.Raymo.2005}.}
       \label{fig:lgmvel}
@@
         and modelled glacierized area during the LGM (coloured curve). The LGM
-        is here modelled as a time-transgressive event.
-        \todo{Enlarge to full width, rename Ivrea to Dora Baltea.}}
+        is here modelled as a time-transgressive event.}
       \label{fig:timing}
@@
         modulated by subglacial topography and catchment geometry.
-        Isolated patches indicate periodic surges from tributary glaciers.
-        \todo{Move glacier names into timeseries panels, change y-labels to
-              glacier length, add new profiles, rename Ivrea to Dora Baltea.}}
+        Isolated patches indicate periodic surges from tributary glaciers.}
       \label{fig:profiles}

### 2017-11-14 articles/cycle/alps-cycle.tex
Add lo-res compute time and Younger Dryas timing.
Add references to Hantke and German maps.

@@
     layers in the ice, corresponding to vertical resolutions of 100 and
-    40\,m, respectively. Simulations ran for \todo{how long?}
-    using 36~processors.
+    40\,m, respectively. Simulations ran for two to four days
+    using 144~processors.
 
@@
     during the Younger Dryas
-    \citep[\todo{timing}, cf. e.g.,][]{Ivy-Ochs.etal.2009}.
+    \citep[12.9--11.7\,ka; cf. e.g.,][]{Ivy-Ochs.etal.2009}.
     This is partly because the 2\,km resolution used in these simulations is
@@
     detail across the Western \citep{Coutterand.2010,
-    Buoncristiani.Campy.2011}, Central \citep{Jackli.1962, Bini.etal.2009}
-    \todo{ref. Hantke}, and Eastern Alps \citep{Penck.Bruckner.1909,
-    Husen.1987, Husen.2011, Bavec.Verbic.2011} \todo{ref. German Geologische
-    Übersichtskarte 1:200,000}. Some of these maps have been compiled
+    Buoncristiani.Campy.2011}, Central \citep{Jackli.1962, Bini.etal.2009,
+    Hantke.2011}, and Eastern Alps \citep{Penck.Bruckner.1909, Husen.1987,
+    BGR.2007, Husen.2011, Bavec.Verbic.2011}.
+    Some of these maps have been compiled
     into a reconstruction, covering the entire Alps \citep{Ehlers.etal.2011},

### 2017-11-14 figures/alpcyc_hr_inputs.py
Fix subfig label order in inputs plot.

@@ -56,7 +56,7 @@ for i, ax in enumerate(grid.flat):
     ax.set_extent(ut.pl.regions['alps'], crs=ax.projection)
     im = ut.pl.draw_boot_topo(ax)
     ut.pl.draw_natural_earth(ax)
-    ut.pl.add_subfig_label('(%s)' % list('abcdefghi')[i], ax)
+    ut.pl.add_subfig_label('(%s)' % list('abcdfhegi')[i], ax)
 
 # add boot topo colorbar
 ticks = range(0, 3001, 1000)

### 2017-11-14 figures/alpcyc_lr_footprints.py
Fix two small bugs in footprints plot.

@@ -29,6 +29,10 @@ for i, rec in enumerate(ut.alpcyc_records):
     ax = grid[0, i/2]
     pp = 'pp' in conf
 
+    # add scaling domain and outline on top panel only
+    ut.pl.draw_scaling_domain(ax)
+    ut.pl.draw_lgm_outline(ax, c='k')
+
     # set title
     ax.text(0.4+0.1*pp, 1.05, label, color=c, fontweight='bold',
             ha=('left' if pp else 'right'), transform=ax.transAxes)
@@ -53,10 +57,6 @@ for i, rec in enumerate(ut.alpcyc_records):
         ut.pl.add_subfig_label('(%s)' % list('abcdef')[i/2+j], ax=ax)
         ut.pl.draw_boot_topo(ax, res='2km')
         ut.pl.draw_natural_earth(ax)
-        ut.pl.draw_lgm_outline(ax, c='k')
-
-    # draw scaling domain
-    ut.pl.draw_scaling_domain(ax)
 
 # save
 fig.savefig('alpcyc_lr_footprints')

### 2017-11-14 articles/cycle/alps-cycle.tex
Update text for new inputs plot.

@@
     hole-filled, Shuttle Radar Topography Mission (SRTM) data with a~resolution
-    of 30\,arc-sec \citep{Jarvis.etal.2008}. These data include post-glacial
+    of 30\,arc-sec \citep[Fig.~\ref{fig:inputs}b;][]{Jarvis.etal.2008}. These
+    data include post-glacial
     sediment fills and lakes surface topography. However, they were corrected
@@
     glacier outlines, surface topography and simplified ice physics
-    \citep{Huss.Farinotti.2012}.
+    \citep[Fig.~\ref{fig:inputs}c;][]{Huss.Farinotti.2012}.
 
@@
     seasonal cycle \citep{Seguinot.2013}, and bilinearly interpolated to the
-    model grids (Fig.~\ref{fig:inputs}b). Degree-day factors for snow and ice
+    model grids (Fig.~\ref{fig:inputs}d and e). Degree-day factors for snow and ice
     melt are set to values used in the European Ice Sheet Modelling INiTiative
@@
     to~1990. Modern climate of the European Alps is characterised by a
-    north-south gradient in summer air temperatures (Fig.~\ref{fig:inputs}c),
-    and an east-west gradient in winter precipitation (Fig.~\ref{fig:inputs}d).
-    WorldClim data
+    north-south gradient in winter and summer air temperatures
+    (Fig.~\ref{fig:inputs}f and g), and an east-west gradient in winter
+    precipitation (Fig.~\ref{fig:inputs}h) which is reversed in summer
+    (Fig.~\ref{fig:inputs}h). WorldClim data
     were selected as an input to the ice sheet model because they incorporate
@@
     degree-day modelling of the central European palaeo-ice caps
-    \citep{Heyman.etal.2013}. Alternatively, the east-west
-    gradient in precipitation existing today (Fig.~\ref{fig:inputs}d) was
+    \citep{Heyman.etal.2013}. Alternatively, the east-west gradient in winter
+    precipitation existing today (Fig.~\ref{fig:inputs}h) was
     perhaps enhanced during the LGM, as indicated by pollen reconstructions
@@
         ice-bedrock interface.
-        \textbf{(b)} Modern July standard deviation \citep{Seguinot.2013} of
-        daily mean temperature from the ERA-Interim \citep[1979--2012;][]{Dee.etal.2011}
-        from the reference monthly climatology used to force the surface mass
-        balance (PDD) component of the ice sheet model.
-        \textbf{(c)} Modern July mean near-surface air temperature, and
-        \textbf{(d)} January precipitation from WorldClim
+        \textbf{(b)} Initial basal topography from SRTM
+        \citep{Jarvis.etal.2008} and geomorphological reconstruction of Last
+        Glacial Maximum Alpine glacier extent \citep[solid red
+        line,][]{Ehlers.etal.2011}.
+        \textbf{(c)} Extract from the estimated of present-day ice thickness
+        \citep{Huss.Farinotti.2012} substracted from the SRTM topography.
+        \textbf{(d)} Modern January and \textbf{(e)} July standard deviation
+        \citep{Seguinot.2013} of daily mean temperature from the ERA-Interim
+        \citep[1979--2012;][]{Dee.etal.2011} from the reference monthly
+        climatology used to force the surface mass balance (PDD) component of
+        the ice sheet model.
+        \textbf{(f)} Modern January and \textbf{(g)} July mean near-surface air
+        temperature, and \textbf{(h, i)} precipitation from WorldClim
         \citep[1960--1990;][]{Hijmans.etal.2005}.}
@@
         cover. Modelled self-sustained ice domes (triangles) and some major
-        transfluences (crosses) are shown. Modelled Last Glacial Maximum
-        (LGM) ice extent (dashed orange line) and geomorphological
+        transfluences (crosses) are shown. Modelled LGM
+        ice extent (dashed orange line) and geomorphological
         reconstruction \citep[solid red line,][]{Ehlers.etal.2011}. The

### 2017-11-14 articles/cycle/alps-cycle.tex
Rescale landscape figures for manuscript mode.
Update text for new profiles plot.

@@
     (Fig.~\ref{fig:profiles}b). The Rhone Glacier, fed by several high-altitude
-    accumulation areas, advanced eight times onto modern Lake
-    Geneva (Fig.~\ref{fig:profiles}c and d). The Dora Baltea Glacier, characterised
+    accumulation areas, advanced eight times onto modern Lake Geneva
+    (Fig.~\ref{fig:profiles}c and d). The Dora Baltea Glacier, characterised
     by a very steep catchment, and mutliple tributaries, shows even a higher
     variability and reaches close to its maximum position six times throughout
-    the simulation (Fig.~\ref{fig:profiles}e and f). Finally, the Isère
-    Glacier, with its complex system of confluences and diffluences, needs
-    longer time to build up and reaches the foreland only twice during the
-    major glaciations of MIS~4 and 2 (Fig.~\ref{fig:profiles}g and h).
+    the simulation (Fig.~\ref{fig:profiles}e and f). The Isère and Inn
+    glaciers, with their complex system of confluences and diffluences, need
+    longer time to build up and reach the foreland only two and three times
+    (Fig.~\ref{fig:profiles}g--j). Finally, the Tagliamento Glacier, distant
+    from the major ice-dispersal centres of the inner Alps, is absent for most
+    of the modelled glacial cycle (Fig.~\ref{fig:profiles}k and l)
 
@@
     \begin{sidewaysfigure*}
-      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
+      % scalebox is not necessary in production mode
+      \centerline{\scalebox{0.9}{\includegraphics{alpcyc_hr_lgmvel}}}
       \caption{%
@@
     \begin{sidewaysfigure*}
-      \centerline{\includegraphics{alpcyc_hr_timing}}
+      % scalebox is not necessary in production mode
+      \centerline{\scalebox{0.9}{\includegraphics{alpcyc_hr_timing}}}
       \caption{%
@@
       \caption{%
-        \textbf{(a, c, e, g)} Profile lines roughly following valley
-        centerlines for the Rhine, Rhone, Dora Baltea and Isère glaciers.
-        \textbf{(b, d, f, h)} Evolution of modelled glacier extent in time,
-        bilinearly interpolated along the corresponding profiles, showing
+        \textbf{(a, c, e, g, i, k)} Profile lines roughly following valley
+        centerlines for the Rhine, Rhone, Dora Baltea and Isère, Inn and
+        Tagliamento glaciers.
+        \textbf{(b, d, f, h, j, l)} Evolution of modelled glacier extent in
+        time, bilinearly interpolated along the corresponding profiles, showing
         numerous cycles of advance and retreat over the last glacial cycle

### 2017-11-15 articles/cycle/alps-cycle.tex
Make two small fixes. First revision complete.

@@
     variability and reaches close to its maximum position six times throughout
-    the simulation (Fig.~\ref{fig:profiles}e and f). The Isère and Inn
+    the simulation (Fig.~\ref{fig:profiles}e and f). The Isère
+    (Fig.~\ref{fig:profiles}g and h) and Inn (Fig.~\ref{fig:profiles}i and j)
     glaciers, with their complex system of confluences and diffluences, need
@@
     forcing and surface mass balance model applied, and uncertainties in ice
-    flow physics. For instance, additional climate variability over Central
-    Europe may have caused additional glaciations to those modelled here.
+    flow physics. In particular, additional palaeo-climate variability over
+    Central Europe may have caused more glaciations than modelled here.
     Through more detailed sensitivity studies, targeted to

### 2017-11-20 articles/cycle/alps-cycle.tex
Add reference to Haeberli and Schluchter (1987).

@@
     been repeatedly invalidated in other glaciated regions of the globe
-    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012}. Yet no
-    evidence for cold-based glaciation has been reported in the Alps to date.
-    Alpine ice flow
+    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012}. Although
+    no evidence for cold-based glaciation has been reported in the Alps to
+    date, it is supported by numerical modelling
+    \citep[Fig.~3]{Haeberli.Schluchter.1987}. Alpine ice flow
     patterns were primarily controlled by subglacial topography, but there is
@@
     climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
-    inevitable. In this context, our results challenge the assumption that
+    inevitable \citep[Fig.~3]{Haeberli.Schluchter.1987}.
+    In this context, our results challenge the assumption that
     Alpine trimlines mark the maximum upper ice surface elevation of the LGM

### 2017-11-20 articles/cycle/alps-cycle.tex
Implement new comments from Matthias.

@@
     entire Alpine range (Fig.~\ref{fig:inputs}). The simulations were run on
-    two distinct grids, using horizontal resolutions of 2 and 1\,km.
+    two distinct grids, using horizontal resolutions of 2 and 1\,km,
+    respectively.
     All parameter values are summarized in Table~\ref{tab:params}.
@@
     Using the Young's modulus, $Y=100$\,GPa, and the Poisson ratio, $\nu=0.25$
-    \citep{Mey.etal.2016}, our value for lithosphere rigidity thus corresponds
-    to an effective elastic thickness of the lithosphere of $E=52.4$\,km,
-    instead of the 50\,km value used by \citet{Mey.etal.2016}.}
+    \citep{Mey.etal.2016}, our simulations effectively use an elastic thickness
+    of the lithosphere of $E=52.4$\,km, instead of the 50\,km value used by
+    \citet{Mey.etal.2016}, introducing in a small change in the lenght scale of
+    bedrock deformation well within uncertainties \citep{Mey.etal.2016}.}
 
@@
     those cycles (Fig.~\ref{fig:timeseries}b, light colour curves). In
-    particular, strong millenial variability recorded in GRIP
+    particular, strong millennial variability recorded in GRIP
     \chem{\delta^{18}O} (Dansgaard-Oeschger events) have large repercussions
@@
     palaeo-temperature fluctuations in Central Europe than Greenland, or
-    temperature-independant controls on GRIP \chem{\delta^{18}O}. In contrast,
+    temperature-independent controls on GRIP \chem{\delta^{18}O}. In contrast,
     the EPICA palaeo-temperature record has least variability among the records
@@
     the other hand, peripheral glaciation modelled using the EPICA record
-    (Fig.~\ref{fig:footprints}b), which has less temperature variability,
+    (Fig.~\ref{fig:footprints}b), which has smaller temperature variability,
     is in relatively good agreement with the geomorphological reconstructions.
@@
     evidence for a pre-LGM glacier advance during the last glacial cycle is
-    indicated \citep{Haldimann.etal.2017}.
+    available \citep{Haldimann.etal.2017}.
 
@@
     (Fig.~\ref{fig:lgmvel}a). On the other hand, the model yields excessive ice
-    cover in the eastern Alps, where the Mur, Drava and Sava glaciers are
+    cover in the eastern Alps, where the Mur, Drava and Sava Glaciers are
     modelled to extend tens of km beyond the mapped ice limits
@@
     On a more regional level, the maximum extent of the Ticino, Iseo,
-    Garda and Piave glaciers in the central southern Alps is modelled
+    Garda and Piave Glaciers in the central southern Alps is modelled
     several kilometres within the mapped LGM moraines (Fig.~\ref{fig:lgmvel}a).
@@
     blue colour mapping). Ice velocities of several hundred metres per year,
-    characterising basal sliding, generally occur along the main river valleys,
+    characteristic for basal sliding, generally occur along the main river valleys,
     while ice domes and ice divides are predominantly located over major
     reliefs areas, where ice moves only by a few metres per year,
-    characterising internal deformation without basal sliding
+    characteristic for internal deformation without basal sliding
     (Fig.~\ref{fig:lgmvel}a). Nevertheless, the model results depict exceptions
@@
     maximum extent at different ages (Fig.~\ref{fig:timing}). For instance, the
-    Dora Riparia, Dora Baltea, and Tagliamento glaciers reach an early maximum before
-    27\,ka; the Durance, Rhone, Inn, Enns, Ticino, and Garda glaciers reach
+    Dora Riparia, Dora Baltea, and Tagliamento Glaciers reach an early maximum before
+    27\,ka; the Durance, Rhone, Inn, Enns, Ticino, and Garda Glaciers reach
     their maximum thickness in phase with the overall Alpine areal maximum
-    around 25\,ka; while the Isère, Como and Iseo glaciers reach a late maximum
+    around 25\,ka; while the Isère, Como and Iseo Glaciers reach a late maximum
     after 24\,ka (Fig.~\ref{fig:timing}). Remarkably, peripheral ice caps on
@@
     For large Alpine glaciers flowing in deep valleys, such as the Rhone,
-    Rhine and Garda glaciers, several thousand model years are needed to attain
+    Rhine and Garda Glaciers, several thousand model years are needed to attain
     a thermodynamical equilibrium between cold-ice advection from the high
@@
     found in the Alps, the LGM glacier extent corresponds to a transient stage,
-    while millennial scale, spatial differences in its timing can result not
+    while millennial-scale spatial differences in its timing can result not
     only from climate variability but also from complex glacier dynamics.
@@
     glaciers, with their complex system of confluences and diffluences, need
-    longer time to build up and reach the foreland only two and three times
+    longer time to build up and reach the foreland only two or three times
     (Fig.~\ref{fig:profiles}g--j). Finally, the Tagliamento Glacier, distant
@@
     Luetscher.etal.2015} but absent from the EPICA palaeo-temperature forcing,
-    may have induced a more dynamic glacier response than that modelled here.
+    may have induced an even more dynamic glacier response than that modelled
+    here.
 
@@
         centerlines for the Rhine, Rhone, Dora Baltea and Isère, Inn and
-        Tagliamento glaciers.
+        Tagliamento Glaciers.
         \textbf{(b, d, f, h, j, l)} Evolution of modelled glacier extent in

### 2017-11-20 articles/cycle/alps-cycle.tex
Add texcount magic comments.
Add minor comments from Guillaume.

@@
 
+%TC: break Abstract
 \begin{abstract}
@@
     Here, we use the Parallel Ice Sheet Model (PISM) to model the entire last
-    glacial cycle (120--0\,ka) in the Alps, with a horizontal resolution of
+    glacial cycle (120--0\,ka) in the Alps, using 576~processors and a
+    horizontal resolution of
     1\,km. Climate forcing is derived using present-day climate data from
@@
 % ----------------------------------------------------------------------
+%TC: break Introduction
 \introduction
@@
     open source, finite difference, shallow ice sheet model
-    \citep{PISM-authors.2017}. The model requires input on basal
-    topography, geothermal heat flux and climate forcing. It computes the
+    \citep{PISM-authors.2017}. The model requires input on initial bedrock and
+    glacier topographies, geothermal heat flux and climate forcing. It computes
+    the
     evolution of ice extent and thickness over time, the thermal and dynamic
@@
     Centre for Medium-Range Weather Forecasts Reanalysis Interim
-    \citep[ERA-Interim;][]{Dee.etal.2011}, perturbed by temperature lapse-rate
+    \citep[ERA-Interim;][]{Dee.etal.2011}, amended with temperature lapse-rate
     corrections (Sect.~\ref{sec:atm}), time-dependent temperature offsets, and
@@
 % ----------------------------------------------------------------------
+%TC: break Conclusions
 \conclusions
@@
 % ----------------------------------------------------------------------
+%TC: break Acknowledgements
 % Acknowledgements
@@
 % ----------------------------------------------------------------------
+%TC: break References
 % References
@@
 % ----------------------------------------------------------------------
+%TC: break Figures
 % Figures

### 2017-11-21 figures/util/pl.py
Convert single map figures to Nature dimensions.

@@ -98,7 +98,7 @@ def cut_ts_axes(ax, tsw=2/3., tsh=1/3.):
     poly = iplt.Polygon(zip(x, y), **polykw)
     rect = iplt.Rectangle((1-tsw, 0.0), tsw, tsh, **rectkw)
     tsax = fig.add_axes([pos.x1-tsw*(pos.x1-pos.x0)+12.5/figw, 10.0/figh,
-                         tsw*(pos.x1-pos.x0)-20.0/figw,
+                         tsw*(pos.x1-pos.x0)-25.0/figw,
                          tsh*(pos.y1-pos.y0)-15.0/figh])
     tsax.add_patch(poly)
     tsax.add_patch(rect)
@@ -106,7 +106,6 @@ def cut_ts_axes(ax, tsw=2/3., tsh=1/3.):
 
 
 def cut_ts_sc_axes(ax, tsw=2/3., tsh=1/3., scw=1/5., sch=1/3.):
-    # scw=45.0/225.0, sch=45.0/150.0):
     """Cut timeseries and scatter plot insets into main axes."""
     fig = ax.figure
     pos = ax.get_position()
@@ -124,7 +123,7 @@ def cut_ts_sc_axes(ax, tsw=2/3., tsh=1/3., scw=1/5., sch=1/3.):
                          scw*(pos.x1-pos.x0)-10.0/figw,
                          sch*(pos.y1-pos.y0)-10.0/figh])
     tsax = fig.add_axes([pos.x1-tsw*(pos.x1-pos.x0)+12.5/figw, 10.0/figh,
-                         tsw*(pos.x1-pos.x0)-20.0/figw,
+                         tsw*(pos.x1-pos.x0)-25.0/figw,
                          tsh*(pos.y1-pos.y0)-15.0/figh])
     scax.add_patch(screct)
     tsax.add_patch(poly)
@@ -166,22 +165,22 @@ def subplots_ts(nrows=1, ncols=1, mode='column', labels=True):
 
 def subplots_cax(extent='alps'):
     """Init figure with unique subplot and colorbar inset."""
-    figw, figh = 230.0, 155.0
+    figw, figh = 183.0, 123.0
     fig, ax = iplt.subplots_mm(figsize=(figw, figh), projection=utm,
-                               gridspec_kw=dict(left=2.5, right=2.5,
-                                                bottom=2.5, top=2.5))
[…]

### 2017-11-21 articles/cycle/alps-cycle.tex
Embed word counts in Alps cycle manuscript.
Move methods and climate forcing after main.
Mask sections planned for removal.
Restore sideways figures orientation to portrait.
Move supplementary figures and tables apart.
Combine timing, trimlines and profiles figures.

@@
 
+\usepackage{framed}
+\usepackage{verbatim}
+
 \graphicspath{{../../figures/}}
 
+\definecolor{gray}{gray}{0.75}
 \definecolor{darkblue}{cmyk}{0.9,0.3,0.0,0.0}
@@
 \newcommand{\aref}[0]{\textcolor{darkblue}{\textbf{[REF.]}}}
+\newenvironment{mask}{\color{gray}}{\color{black}}
+%TC:newcounter mask
+%TC:envir mask [] mask
 
@@
 % ======================================================================
+
+% Texcount template
+%TC:newtemplate
+%TC:template Word counts (text (masked), headers, other):\n
+%TC:template  {SUB?* {title}: {text} ({mask}), {headerword}, {otherword}\n?SUB}
+%TC:template Total: {text} ({mask}), {headerword}, {otherword}
+
+\begin{framed}
+\small\verbatiminput{alps-cycle.sum}
+\end{framed}
 
@@
 % ----------------------------------------------------------------------
-%TC: break Introduction
-\introduction
+\section{Main}
 \label{sec:intro}
@@
 
+    \begin{mask}
+
     However, this glacial theory did not gain general acceptance until the
@@
     \emph{Last Glacial Maximum} (LGM).
+
+    \end{mask}
 
@@
 
-% ----------------------------------------------------------------------
-\section{Ice sheet model set-up}
-\label{sec:model}
-% ----------------------------------------------------------------------
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Overview}
-\label{sec:overview}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    We use the Parallel Ice Sheet Model (PISM, development version~e9d2d1f), an
-    open source, finite difference, shallow ice sheet model
-    \citep{PISM-authors.2017}. The model requires input on initial bedrock and
-    glacier topographies, geothermal heat flux and climate forcing. It computes
-    the
-    evolution of ice extent and thickness over time, the thermal and dynamic
-    states of the ice sheet, and the associated lithospheric response. The
-    model set-up used here was previously developed and tested on the former
-    Cordilleran ice sheet \citep{Seguinot.2014, Seguinot.etal.2014,
-    Seguinot.etal.2016} and subsequently adapted for steady climate
-    \citep{Becker.etal.2016} and regional \citep{Jouvet.etal.2017a}
-    applications to the European Alps.
-
-    Ice deformation follows temperature and water-content dependent creep
-    (Sect.~\ref{sec:icedyn}). Basal sliding follows a~pseudo-plastic law where
-    the yield stress accounts for till deconsolidation under high water
-    pressure (Sect.~\ref{sec:sliding}). Bedrock topography is deflected
-    under the ice load (Sect.~\ref{sec:bedrock}). Surface mass balance is
-    computed using a~positive degree-day (PDD) model (Sect.~\ref{sec:surface}).
-    Climate forcing is provided by a~monthly climatology from interpolated
-    observational data \citep[WorldClim;][]{Hijmans.etal.2005} and the European
-    Centre for Medium-Range Weather Forecasts Reanalysis Interim
-    \citep[ERA-Interim;][]{Dee.etal.2011}, amended with temperature lapse-rate
-    corrections (Sect.~\ref{sec:atm}), time-dependent temperature offsets, and
-    in some cases, time-dependent palaeo-precipitation reductions
-    (Sect.~\ref{sec:climate}).
-
-    Each simulation starts from assumed present-day ice thickness and
-    equilibrium ice and bedrock temperature at 120\,ka, and runs to the
-    present. Our modelling domain of 900 by 600\,km encompasses the
-    entire Alpine range (Fig.~\ref{fig:inputs}). The simulations were run on
-    two distinct grids, using horizontal resolutions of 2 and 1\,km,
-    respectively.
-    All parameter values are summarized in Table~\ref{tab:params}.
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Ice rheology}
-\label{sec:icedyn}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    Ice sheet dynamics are typically modelled using a~combination of internal
-    deformation and basal sliding. PISM is a~shallow ice sheet model, which
-    implies that the balance of stresses is approximated based on their
-    predominant components. The Shallow Shelf Approximation (SSA) is combined
-    with the Shallow Ice Approximation (SIA) by adding velocity solutions of
-    the two approximations \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}.
-    Although this heuristic approach implies errors in the transition zone
-    where gravitational stresses intervene both in the SIA and SSA velocity
-    computation, it was shown to effectively approximate complex ice field flow
-    dynamics over mountainous topographies similar to that of the European Alps
-    \citep{Golledge.etal.2012, Ziemen.etal.2016}. This hybrid scheme is
-    computationally much more efficient than a fully three-dimensional model
-    \citep[e.g.,][]{Cohen.etal.2017}
-    with which simulations of the scale presented here would not be feasible.
-
-    Ice deformation is governed by the constitutive law for ice
-    \citep{Glen.1952, Nye.1953}. Ice softness depends on ice temperature,
-    pressure, and water content, through an enthalpy scheme
-    \citep[Table~\ref{tab:params};][]{Aschwanden.etal.2012} and a piece-wise
-    Arrhenius-type rheology representative of Holocene polar ice
-    \citep[Table~\ref{tab:params};][p.~77]{Cuffey.Paterson.2010}.
-    The englacial water fraction is
-    capped at a~maximum value of 0.01, above which no measurements are
-    available \citep[Eq.~5.7]{Lliboutry.Duval.1985, Greve.1997}.
-
-    Surface air temperature derived from the climate forcing
-    (Sects.~\ref{sec:atm}, ~\ref{sec:paltemp}) provides the upper boundary
-    condition to the ice
-    enthalpy model. Temperature is computed in the ice and in the bedrock down
-    to a~depth of 3\,km below the glacier base, where it is
-    conditioned by a~lower boundary geothermal heat flux estimate from multiple
-    geothermal proxies (\citealp[similarity method]{Goutorbe.etal.2011};
-    Fig.~\ref{fig:inputs}a).
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Basal sliding}
-\label{sec:sliding}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    A~pseudo-plastic sliding law relates the bed-parallel shear stresses to the
-    sliding velocity. The yield stress is modelled using the Mohr--Coulomb
-    criterion using a constant basal friction angle (Table~\ref{tab:params}).
-    Effective pressure is related to the ice overburden stress and the modelled
-    amount of subglacial water, using a~formula derived from laboratory
-    experiments with till extracted from the base of Ice Stream B in West
-    Antarctica \citep[Table~\ref{tab:params};][]{Tulaczyk.etal.2000,
-    Bueler.Pelt.2015}. When the till becomes saturated, additional melt water
-    is assumed to drain off instantaneously.
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Basal topography}
-\label{sec:bedrock}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    The initial basal topography is bilinearly interpolated from the
-    hole-filled, Shuttle Radar Topography Mission (SRTM) data with a~resolution
-    of 30\,arc-sec \citep[Fig.~\ref{fig:inputs}b;][]{Jarvis.etal.2008}. These
-    data include post-glacial
-    sediment fills and lakes surface topography. However, they were corrected
-    for an estimation of present-day ice thickness based on modern
-    glacier outlines, surface topography and simplified ice physics
-    \citep[Fig.~\ref{fig:inputs}c;][]{Huss.Farinotti.2012}.
-
-    Basal topography responds to ice load following a~bedrock deformation model
-    that includes local isostasy, elastic lithosphere flexure and viscous
-    asthenosphere deformation in an infinite half-space
-    \citep{Lingle.Clark.1985, Bueler.etal.2007}. Model parameters were set
-    according to
-    results from glacial isostatic adjustment modelling of deglacial rebound in
-    the Alps most closely reproducing observed modern uplift rates
-    \citep[Supplementary Fig.~7]{Mey.etal.2016}.\footnote{Lithosphere rigidity
-    was computed using the erroneous formula, ${D=YE^3/12(1-\nu)}$, in
-    \citet{Mey.etal.2016}. The correct formula is ${D=YE^3/12(1-\nu^2)}$.
-    Using the Young's modulus, $Y=100$\,GPa, and the Poisson ratio, $\nu=0.25$
-    \citep{Mey.etal.2016}, our simulations effectively use an elastic thickness
-    of the lithosphere of $E=52.4$\,km, instead of the 50\,km value used by
-    \citet{Mey.etal.2016}, introducing in a small change in the lenght scale of
-    bedrock deformation well within uncertainties \citep{Mey.etal.2016}.}
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Surface mass balance}
-\label{sec:surface}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    Ice surface accumulation and ablation are computed from monthly mean
-    near-surface air temperature, $T_{\mathrm{m}}$, monthly standard deviation
-    of near-surface air temperature, $\sigma$, and monthly precipitation,
-    $P_{\mathrm{m}}$, using a~temperature-index model
-    \citep[e.g.,][]{Hock.2003}. Accumulation is equal to precipitation when air
-    temperatures are below 0\,\unit{{\degree}C}, and decreases to zero linearly
-    with temperatures between 0 and 2\,\unit{{\degree}C}. Ablation is computed
-    from PDD, the integral of temperatures above 0\,\unit{{\degree}C}.
-
-    The PDD computation accounts for stochastic temperature variations by
-    assuming a~normal temperature distribution of standard deviation, $\sigma$,
-    around the expected value, $T_{\mathrm{m}}$. It is expressed by an
-    error-function formulation \citep{Calov.Greve.2005}
-    which is numerically approximated using week-long sub-intervals. In
-    order to account for the effects of spatial and seasonal variations of
-    temperature variability \citep{Seguinot.2013}, $\sigma$ is computed
-    from ERA-Interim daily temperature values from 1979 to 2012
-    \citep{Mesinger.etal.2006}, including variability associated with the
-    seasonal cycle \citep{Seguinot.2013}, and bilinearly interpolated to the
-    model grids (Fig.~\ref{fig:inputs}d and e). Degree-day factors for snow and ice
-    melt are set to values used in the European Ice Sheet Modelling INiTiative
-    \citep[Table~\ref{tab:params}; EISMINT,][]{Huybrechts.1998}.
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Reference climate forcing}
-\label{sec:atm}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    Climate forcing driving ice sheet simulations consists of a~present-day
-    monthly climatology, $\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, modified by
-    temperature lapse-rate corrections, ${\Delta}T_{\text{LR}}$, temperature
-    offset time series, ${\Delta}T_{\text{TS}}$, and time-dependent
-    palaeo-precipitation corrections, $\Psi_{\text{PP}}$:
-    %
-    \begin{align}
-      &T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) +
-                                 {\Delta}T_{\text{LR}}(t, x, y) +
-                                 {\Delta}T_{\text{TS}}(t) \,, \\
-      &P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \cdot
-                                 {\Psi}_{\text{PP}}(t) \,,
-    \end{align}
-    %
-    The present-day monthly climatology was bilinearly interpolated from
-    near-surface air temperature and precipitation rate fields from
-    WorldClim \citep{Hijmans.etal.2005}, representative of the period~1960
-    to~1990. Modern climate of the European Alps is characterised by a
-    north-south gradient in winter and summer air temperatures
-    (Fig.~\ref{fig:inputs}f and g), and an east-west gradient in winter
-    precipitation (Fig.~\ref{fig:inputs}h) which is reversed in summer
-    (Fig.~\ref{fig:inputs}h). WorldClim data
-    were selected as an input to the ice sheet model because they incorporate
-    observations from the dense weather station network of central Europe
-    \citep[Fig.~1]{Hijmans.etal.2005}. Besides, WorldClim data were
-    previously used as climate forcing for PISM to model the LGM extent of the
-    former Cordilleran ice sheet in good agreement with geological evidence
-    along the southern margin \citep{Seguinot.etal.2014} where weather station
-    density is lower than in the Alps. Finally, the last glacial cycle Alpine
-    glaciers did not extend over marine areas where WorldClim data is missing.
-
-    The temperature lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are
-    computed as a~function of ice surface elevation, $s$, using the SRTM
-    topography shipped with WorldClim as a~reference, $b_{\text{ref}}$:
-    %
-    \begin{align}
-      {\Delta}T_{\text{LR}}(t, x, y) &= -\gamma [s(t, x, y)-b_{\text{ref}}] \\
-                                     &= -\gamma [h(t, x, y)+
-                                                 b(t, x, y)-b_{\text{ref}}],
-    \end{align}
-    %
-    thus accounting for the evolution of ice thickness, ${h=s-b}$, on the one
-    hand, and for differences between the basal topography of the ice flow
-    model, $b$, and the WorldClim reference topography, $b_{\text{ref}}$, on
-    the other hand. All simulations use an annual temperature lapse rate of
-    $\gamma=6\,\unit{K\,km^{-1}}$.
-
-% ----------------------------------------------------------------------
-\section{Palaeo-climate forcing}
-\label{sec:climate}
-% ----------------------------------------------------------------------
-
-    In this section, we analyze the model sensitivity to palaeo-climate forcing
-    through the last glacial cycle, using three palaeo-temperature records
-    (Sect.~\ref{sec:paltemp}) and two parametrizations of palaeo-precipitation
-    (Sect.~\ref{sec:palprec}), in terms of modelled evolution of total ice
-    volume (Sect.~\ref{sec:timeseries}) and glaciated area during MIS~2 and~4
-    (Sect.~\ref{sec:footprints}).
-
-    These simulations use a horizontal resolution of 2\,km. The vertical grid
-    consists of 31~temperature layers in the bedrock and up to 126~enthalpy
-    layers in the ice, corresponding to vertical resolutions of 100 and
-    40\,m, respectively. Simulations ran for two to four days
-    using 144~processors.
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Palaeo-temperature forcing}
-\label{sec:paltemp}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived from
-    palaeo-temperature proxy records from the Greenland Ice Core Project
-    \citep[GRIP;][]{Dansgaard.etal.1993}, the European Project for Ice Coring
-    in Antarctica \citep[EPICA;][] {Jouzel.etal.2007}, and an oceanic sediment
-    core from the Iberian margin \citep[MD01-2444;][]{Martrat.etal.2007}.
-    Palaeo-temperature anomalies from the GRIP record are calculated from
-    oxygen isotope (\chem{\delta^{18}O}) measurements using a~quadratic
-    equation \citep{Johnsen.etal.1995},
-    %
-    \begin{align}
-      {\Delta}T_{\text{TS}}(t) ={~}&-11.88 [\chem{\delta^{18}O}(t)
-                                    -\chem{\delta^{18}O}(0)] \nonumber \\
-                                   &-0.1925 [\chem{\delta^{18}O}(t)^2
-                                    -\chem{\delta^{18}O}(0)^2] \,,
-    \end{align}
-    %
-    while temperature reconstructions from the EPICA and MD01-2444 records are
-    provided as such. For each proxy record used and each of the parameter
-    set-ups used in the sensitivity tests, palaeo-temperature anomalies are
-    scaled linearly (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}a) so
-    that the modelled cumulative glaciated area during MIS~2 (29--14\,ka)
-    and within a rectangular region covering the Rhine
-    glacier piedmont lobe (Fig.~\ref{fig:footprints}a, black rectangle) is
-    consistent with the glaciated area of the geological reconstruction
-    \citep{Ehlers.etal.2011}.
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Palaeo-precipitation forcing}
-\label{sec:palprec}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    Finally, in some simulations (hereafter labelled PP), precipitation was
-    reduced with air temperature in
-    order to simulate the potential rarefaction of atmospheric moisture in
-    colder climates. This was done using an empirical relationship derived from
-    observed accumulation rates and oxygen isotopes concentrations in the GRIP
-    ice core \citep{Dahl-Jensen.etal.1993},
-    %
-    \begin{align}
-      {\Psi}_{\text{PP}}(t) = \exp[\psi{\Delta}T_{\text{TS}}(t)] \,,
-    \end{align}
-    %
-    with $\psi=0.169/2.4=0.0704$ \citep{Huybrechts.2002}. This simple
-    relationship likely does not reflect the complexity of atmospheric
-    circulation changes that governed moisture availability over the Alps
-    during the last glacial cycle, of which little is known apart from the
-    LGM \citep[cf.][]{Wu.etal.2007, Strandberg.etal.2011, Ludwig.etal.2016}.
-    Thus, another set of simulations presented here use constant
-    precipitation, corresponding to $\psi=0$.
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Sensitivity of ice volume evolution}
-\label{sec:timeseries}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    For the three palaeo-temperature records and the two palaeo-precipitation
-    parametrizations used, the model yields significant ice volume build-up
-    during MIS~4 and 2 (Fig.~\ref{fig:timeseries}b), corresponding to documented
-    glaciation periods in the Alps \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
-    All simulations also yield important glaciations during MIS~5 and 3, but
-    their timing and amplitude varies significantly depending on the climate
-    forcing used. All six simulations overestimate the late-glacial re-advance
-    during the Younger Dryas
-    \citep[12.9--11.7\,ka; cf. e.g.,][]{Ivy-Ochs.etal.2009}.
-    This is partly because the 2\,km resolution used in these simulations is
-    too coarse to resolve
-    Younger Dryas Alpine glaciers. However, the large total ice volume
-    overestimation resulting from the GRIP temperature forcing
-    (Fig.~\ref{fig:timeseries}b, blue curves) certainly indicates that the
-    vigorous cooling experienced in Greenland is not representative of the
-    Alps.
-
-    Geological data from the best documented Alpine piedmont lobes indicate
-    that their maximum extension during MIS~2 occurred at around 25.5--24.0\,ka
-    \citep{Monegato.etal.2017}, after which Alpine glaciers remained or
-    potentially re-advanced to within close reach of this maximum extent, until
-    as late as 22 to 17\,ka \citep[Fig.~5]{Wirsig.etal.2016}. This is in very
-    good agreement with the result of the EPICA simulations, which yield an
-    early maximum ice volume at 25.2/24.6 (without/with palaeo-precipitation
-    reductions) followed by a standstill until 17.3\,ka
-    (Fig.~\ref{fig:timeseries}b, blue curves). The GRIP palaeo-temperature
-    forcing, on the other hand, yields two distinct total ice volume maxima
-    at 27.3/27.0 and 21.8/21.7\,ka followed by an early deglaciation of the
-    foreland by 21.4\,ka. The MD01-2444 palaeo-temperature forcing yields
-    a late LGM peak ice volume at 16.5/15.7\,ka several thousand years after
-    the dated geological evidence.
-
-    Finally, all simulations yield very strong ice volume
-    variability, indicative of many more than two or three cycles of glacier
-    advance and retreats onto the foreland (Fig.~\ref{fig:timeseries}b).
-    Palaeo-precipitation reductions dampen some of the small-scale variability,
-    but they have little effect on the millennial scales that characterise
-    those cycles (Fig.~\ref{fig:timeseries}b, light colour curves). In
-    particular, strong millennial variability recorded in GRIP
-    \chem{\delta^{18}O} (Dansgaard-Oeschger events) have large repercussions
-    on the modelled Alpine ice volume, including around six glaciations of LGM
-    magnitude (Fig.~\ref{fig:timeseries}b, blue curves), which is not supported
-    by geologic evidence \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
-    Dansgaard-Oeschger events, however, have been recorded in Central European
-    lake sediments \citep{Wohlfarth.etal.2008} and cave speleothems
-    \citep{Spotl.Mangini.2002, Luetscher.etal.2015}. Our results indicate that
-    their magnitude in Central Europe was likely lower than that recorded in
-    GRIP \chem{\delta^{18}O}, which either implies smaller associated
-    palaeo-temperature fluctuations in Central Europe than Greenland, or
-    temperature-independent controls on GRIP \chem{\delta^{18}O}. In contrast,
-    the EPICA palaeo-temperature record has least variability among the records
-    used (Fig.~\ref{fig:timeseries}a, red curves). This results in smaller
-    variability in the total ice volume, and more restrictive glaciations
-    during MIS~5 and~3, (Fig.~\ref{fig:timeseries}b, red curves).
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Sensitivity of glaciated area}
-\label{sec:footprints}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    During MIS~2, all six simulations yield comparative cumulative glaciated
-    areas (Fig.~\ref{fig:footprints}a). This result is inherent to the
-    palaeo-climate forcing approach used, which involves a linear scaling of
-    each palaeo-temperature anomaly record to a geomorphological
-    reconstruction of the area glaciated by the Rhine Glacier piedmont lobe
-    (Sect.~\ref{sec:paltemp}; Fig.~\ref{fig:footprints}a, black rectangle).
-    Outside this benchmark, all simulations underestimate glaciated area in the
-    western part of the model domain, and overestimate it in the eastern
-    part, relative to the geomorphological ice limits
-    (Fig.~\ref{fig:footprints}a). This result might indicate that the LGM
-    temperature depression, here considered homogeneous, was actually lower in
-    the Eastern Alps than in the Western Alps, as previously shown by positive
-    degree-day modelling of the central European palaeo-ice caps
-    \citep{Heyman.etal.2013}. Alternatively, the east-west gradient in winter
-    precipitation existing today (Fig.~\ref{fig:inputs}h) was
-    perhaps enhanced during the LGM, as indicated by pollen reconstructions
-    \citep{Wu.etal.2007}. Finally, the LGM extent of Alpine glaciers might have
-    been affected by east-west gradient in variables not accounted for
-    by the PDD model, such as cloud cover or dust deposition.
-
-    Besides this general pattern, there exist differences in glaciated area
-    depending on the palaeo-climate forcing used. The MD01-2444, and to a
-    greater extent, the GRIP palaeo-temperature records, tend to overestimate
-    MIS~2 ice cover on all peripheral ranges that surround the Alps
-    (Fig.~\ref{fig:footprints}a and c). This is particularly true when no
-    precipitation corrections are applied (bright colours). In fact, both
-    palaeo-temperature records have a higher temperature variability and
-    contain brief spells of cold temperatures, which are too short to develop a
-    fully grown Alpine ice sheet, but long enough to build up ice cover on a
-    smaller scale on these peripheral ranges (Fig.~\ref{fig:timeseries}). On
-    the other hand, peripheral glaciation modelled using the EPICA record
-    (Fig.~\ref{fig:footprints}b), which has smaller temperature variability,
-    is in relatively good agreement with the geomorphological reconstructions.
-
-    The modelled extent of glaciation during MIS~4 depicts a more pronounced
-    sensitivity to the choice of palaeo-climate forcing used. Using the GRIP
-    and MD01-2444 palaeo-temperature records, Alpine glaciers are modelled to
-    extend well beyond the reach of documented LGM end-moraines
-    (Fig.~\ref{fig:footprints}d and f). Because, in both cases,
-    the modelled glaciation extent corresponds to brief cold spells in the
-    palaeo-temperature forcing (Fig.~\ref{fig:timeseries}a),
-    palaeo-precipitation reductions greatly reduce the excessive modelled ice
-    cover (Figs.~\ref{fig:timeseries}b and \ref{fig:footprints}d and f, light
-    colours). However, with or without palaeo-precipitation corrections,
-    the GRIP and MD01-2444 forcing yield
-    modelled ice extents (Fig.~\ref{fig:footprints}d and f) and volumes
-    (Fig.~\ref{fig:timeseries}b) considerably larger during MIS~4 than
-    during MIS~2, which is not supported by geological evidence
-    \citep{Preusser.2004, Ivy-Ochs.etal.2008, Husen.Reitner.2011,
-    Barrett.etal.2017, Haldimann.etal.2017}. Gravel deposits from the Rhine
-    Glacier indicate an early last glacial cycle foreland glaciation, but the
-    extent reached is less than that in MIS~2 \citep{Keller.Krayss.2010}.
-    Dropstones in lake sediments indicate a late MIS~4 advance restricted to
-    the upper Inn Valley \citep{Barrett.etal.2017}. For the Reuss Glacier, no
-    evidence for a pre-LGM glacier advance during the last glacial cycle is
-    available \citep{Haldimann.etal.2017}.
-
-    The EPICA palaeo-temperature forcing, on the other hand, yields a MIS~4
-    glaciation that is only slightly less expansive than the MIS~2 glaciation
-    (Fig.~\ref{fig:footprints}e). The sensitivity of the glaciated area to
-    palaeo-precipitation reductions is mostly limited to the southern terminal
-    lobes of central-Alpine glaciers, which are slightly less extensive during
-    MIS~2 and 4 under reduced precipitation (Fig.~\ref{fig:footprints}b and e).
-
-    Based on the above considerations on timing of the LGM during MIS~2, and
-    modelled ice
-    extent during MIS~4, we select EPICA as our optimal palaeo-temperature
-    record for a more detailed and higher-resolution comparison of modelled
-    glacier dynamics to available geological evidence
-    (Sect.~\ref{sec:results}). As a conservative approach in regard to the
-    rapid ice volume fluctuations, we choose to include palaeo-precipitation
-    corrections in the following higher-resolution run.
-
-% ----------------------------------------------------------------------
-\section{Results and discussion}
+    \begin{mask}
+
+% ----------------------------------------------------------------------
+\subsection{Results and discussion}
 \label{sec:results}
@@
     576~processors.
+
+    \end{mask}
 
@@
 
+    \begin{mask}
+
     The LGM Alpine ice flow pattern was traditionally described as that of a
@@
     Kelly.etal.2004, Bini.etal.2009}.
+
+    \end{mask}
 
@@
 
+    \begin{mask}
+
     The timing of the LGM has been documented by radiocarbon and cosmogenic
@@
     \citealp[Fig.~3]{Monegato.etal.2017}).
+
+    \end{mask}
 
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{mask}
 
@@
 
+    \end{mask}
+
     In the upper Rhone Valley, the maximum ice surface elevation reached
@@
 
+    \begin{mask}
+
     Glacial history of the Alps prior to the LGM remains poorly constrained.
@@
     \citep[e.g.,][]{Link.Preusser.2006, Preusser.etal.2007}.
+
+    \end{mask}
 
@@
 
-% ----------------------------------------------------------------------
-%TC: break Conclusions
-\conclusions
+    \begin{mask}
+
+% ----------------------------------------------------------------------
+\subsection{conclusions}
 % ----------------------------------------------------------------------
@@
     \end{itemize}
+
+    \end{mask}
 
@@
 % ----------------------------------------------------------------------
-%TC: break Acknowledgements
+\section{Methods}
+\label{sec:model}
+% ----------------------------------------------------------------------
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Overview}
+\label{sec:overview}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    We use the Parallel Ice Sheet Model (PISM, development version~e9d2d1f), an
+    open source, finite difference, shallow ice sheet model
+    \citep{PISM-authors.2017}. The model requires input on initial bedrock and
+    glacier topographies, geothermal heat flux and climate forcing. It computes
+    the
+    evolution of ice extent and thickness over time, the thermal and dynamic
+    states of the ice sheet, and the associated lithospheric response. The
+    model set-up used here was previously developed and tested on the former
+    Cordilleran ice sheet \citep{Seguinot.2014, Seguinot.etal.2014,
+    Seguinot.etal.2016} and subsequently adapted for steady climate
+    \citep{Becker.etal.2016} and regional \citep{Jouvet.etal.2017a}
+    applications to the European Alps.
+
+    Ice deformation follows temperature and water-content dependent creep
+    (Sect.~\ref{sec:icedyn}). Basal sliding follows a~pseudo-plastic law where
+    the yield stress accounts for till deconsolidation under high water
+    pressure (Sect.~\ref{sec:sliding}). Bedrock topography is deflected
+    under the ice load (Sect.~\ref{sec:bedrock}). Surface mass balance is
+    computed using a~positive degree-day (PDD) model (Sect.~\ref{sec:surface}).
+    Climate forcing is provided by a~monthly climatology from interpolated
+    observational data \citep[WorldClim;][]{Hijmans.etal.2005} and the European
+    Centre for Medium-Range Weather Forecasts Reanalysis Interim
+    \citep[ERA-Interim;][]{Dee.etal.2011}, amended with temperature lapse-rate
+    corrections (Sect.~\ref{sec:atm}), time-dependent temperature offsets, and
+    in some cases, time-dependent palaeo-precipitation reductions
+    (Sect.~\ref{sec:climate}).
+
+    Each simulation starts from assumed present-day ice thickness and
+    equilibrium ice and bedrock temperature at 120\,ka, and runs to the
+    present. Our modelling domain of 900 by 600\,km encompasses the
+    entire Alpine range (Fig.~\ref{fig:inputs}). The simulations were run on
+    two distinct grids, using horizontal resolutions of 2 and 1\,km,
+    respectively.
+    All parameter values are summarized in Table~\ref{tab:params}.
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Ice rheology}
+\label{sec:icedyn}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    Ice sheet dynamics are typically modelled using a~combination of internal
+    deformation and basal sliding. PISM is a~shallow ice sheet model, which
+    implies that the balance of stresses is approximated based on their
+    predominant components. The Shallow Shelf Approximation (SSA) is combined
+    with the Shallow Ice Approximation (SIA) by adding velocity solutions of
+    the two approximations \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}.
+    Although this heuristic approach implies errors in the transition zone
+    where gravitational stresses intervene both in the SIA and SSA velocity
+    computation, it was shown to effectively approximate complex ice field flow
+    dynamics over mountainous topographies similar to that of the European Alps
+    \citep{Golledge.etal.2012, Ziemen.etal.2016}. This hybrid scheme is
+    computationally much more efficient than a fully three-dimensional model
+    \citep[e.g.,][]{Cohen.etal.2017}
+    with which simulations of the scale presented here would not be feasible.
+
+    Ice deformation is governed by the constitutive law for ice
+    \citep{Glen.1952, Nye.1953}. Ice softness depends on ice temperature,
+    pressure, and water content, through an enthalpy scheme
+    \citep[Table~\ref{tab:params};][]{Aschwanden.etal.2012} and a piece-wise
+    Arrhenius-type rheology representative of Holocene polar ice
+    \citep[Table~\ref{tab:params};][p.~77]{Cuffey.Paterson.2010}.
+    The englacial water fraction is
+    capped at a~maximum value of 0.01, above which no measurements are
+    available \citep[Eq.~5.7]{Lliboutry.Duval.1985, Greve.1997}.
+
+    Surface air temperature derived from the climate forcing
+    (Sects.~\ref{sec:atm}, ~\ref{sec:paltemp}) provides the upper boundary
+    condition to the ice
+    enthalpy model. Temperature is computed in the ice and in the bedrock down
+    to a~depth of 3\,km below the glacier base, where it is
+    conditioned by a~lower boundary geothermal heat flux estimate from multiple
+    geothermal proxies (\citealp[similarity method]{Goutorbe.etal.2011};
+    Fig.~\ref{fig:inputs}a).
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Basal sliding}
+\label{sec:sliding}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    A~pseudo-plastic sliding law relates the bed-parallel shear stresses to the
+    sliding velocity. The yield stress is modelled using the Mohr--Coulomb
+    criterion using a constant basal friction angle (Table~\ref{tab:params}).
+    Effective pressure is related to the ice overburden stress and the modelled
+    amount of subglacial water, using a~formula derived from laboratory
+    experiments with till extracted from the base of Ice Stream B in West
+    Antarctica \citep[Table~\ref{tab:params};][]{Tulaczyk.etal.2000,
+    Bueler.Pelt.2015}. When the till becomes saturated, additional melt water
+    is assumed to drain off instantaneously.
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Basal topography}
+\label{sec:bedrock}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    The initial basal topography is bilinearly interpolated from the
+    hole-filled, Shuttle Radar Topography Mission (SRTM) data with a~resolution
+    of 30\,arc-sec \citep[Fig.~\ref{fig:inputs}b;][]{Jarvis.etal.2008}. These
+    data include post-glacial
+    sediment fills and lakes surface topography. However, they were corrected
+    for an estimation of present-day ice thickness based on modern
+    glacier outlines, surface topography and simplified ice physics
+    \citep[Fig.~\ref{fig:inputs}c;][]{Huss.Farinotti.2012}.
+
+    Basal topography responds to ice load following a~bedrock deformation model
+    that includes local isostasy, elastic lithosphere flexure and viscous
+    asthenosphere deformation in an infinite half-space
+    \citep{Lingle.Clark.1985, Bueler.etal.2007}. Model parameters were set
+    according to
+    results from glacial isostatic adjustment modelling of deglacial rebound in
+    the Alps most closely reproducing observed modern uplift rates
+    \citep[Supplementary Fig.~7]{Mey.etal.2016}.\footnote{Lithosphere rigidity
+    was computed using the erroneous formula, ${D=YE^3/12(1-\nu)}$, in
+    \citet{Mey.etal.2016}. The correct formula is ${D=YE^3/12(1-\nu^2)}$.
+    Using the Young's modulus, $Y=100$\,GPa, and the Poisson ratio, $\nu=0.25$
+    \citep{Mey.etal.2016}, our simulations effectively use an elastic thickness
+    of the lithosphere of $E=52.4$\,km, instead of the 50\,km value used by
+    \citet{Mey.etal.2016}, introducing in a small change in the lenght scale of
+    bedrock deformation well within uncertainties \citep{Mey.etal.2016}.}
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Surface mass balance}
+\label{sec:surface}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    Ice surface accumulation and ablation are computed from monthly mean
+    near-surface air temperature, $T_{\mathrm{m}}$, monthly standard deviation
+    of near-surface air temperature, $\sigma$, and monthly precipitation,
+    $P_{\mathrm{m}}$, using a~temperature-index model
+    \citep[e.g.,][]{Hock.2003}. Accumulation is equal to precipitation when air
+    temperatures are below 0\,\unit{{\degree}C}, and decreases to zero linearly
+    with temperatures between 0 and 2\,\unit{{\degree}C}. Ablation is computed
+    from PDD, the integral of temperatures above 0\,\unit{{\degree}C}.
+
+    The PDD computation accounts for stochastic temperature variations by
+    assuming a~normal temperature distribution of standard deviation, $\sigma$,
+    around the expected value, $T_{\mathrm{m}}$. It is expressed by an
+    error-function formulation \citep{Calov.Greve.2005}
+    which is numerically approximated using week-long sub-intervals. In
+    order to account for the effects of spatial and seasonal variations of
+    temperature variability \citep{Seguinot.2013}, $\sigma$ is computed
+    from ERA-Interim daily temperature values from 1979 to 2012
+    \citep{Mesinger.etal.2006}, including variability associated with the
+    seasonal cycle \citep{Seguinot.2013}, and bilinearly interpolated to the
+    model grids (Fig.~\ref{fig:inputs}d and e). Degree-day factors for snow and ice
+    melt are set to values used in the European Ice Sheet Modelling INiTiative
+    \citep[Table~\ref{tab:params}; EISMINT,][]{Huybrechts.1998}.
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Reference climate forcing}
+\label{sec:atm}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    Climate forcing driving ice sheet simulations consists of a~present-day
+    monthly climatology, $\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, modified by
+    temperature lapse-rate corrections, ${\Delta}T_{\text{LR}}$, temperature
+    offset time series, ${\Delta}T_{\text{TS}}$, and time-dependent
+    palaeo-precipitation corrections, $\Psi_{\text{PP}}$:
+    %
+    \begin{align}
+      &T_{\mathrm{m}}(t, x, y) = T_{\mathrm{m}0}(x, y) +
+                                 {\Delta}T_{\text{LR}}(t, x, y) +
+                                 {\Delta}T_{\text{TS}}(t) \,, \\
+      &P_{\mathrm{m}}(t, x, y) = P_{\mathrm{m}0}(x, y) \cdot
+                                 {\Psi}_{\text{PP}}(t) \,,
+    \end{align}
+    %
+    The present-day monthly climatology was bilinearly interpolated from
+    near-surface air temperature and precipitation rate fields from
+    WorldClim \citep{Hijmans.etal.2005}, representative of the period~1960
+    to~1990. Modern climate of the European Alps is characterised by a
+    north-south gradient in winter and summer air temperatures
+    (Fig.~\ref{fig:inputs}f and g), and an east-west gradient in winter
+    precipitation (Fig.~\ref{fig:inputs}h) which is reversed in summer
+    (Fig.~\ref{fig:inputs}h). WorldClim data
+    were selected as an input to the ice sheet model because they incorporate
+    observations from the dense weather station network of central Europe
+    \citep[Fig.~1]{Hijmans.etal.2005}. Besides, WorldClim data were
+    previously used as climate forcing for PISM to model the LGM extent of the
+    former Cordilleran ice sheet in good agreement with geological evidence
+    along the southern margin \citep{Seguinot.etal.2014} where weather station
+    density is lower than in the Alps. Finally, the last glacial cycle Alpine
+    glaciers did not extend over marine areas where WorldClim data is missing.
+
+    The temperature lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are
+    computed as a~function of ice surface elevation, $s$, using the SRTM
+    topography shipped with WorldClim as a~reference, $b_{\text{ref}}$:
+    %
+    \begin{align}
+      {\Delta}T_{\text{LR}}(t, x, y) &= -\gamma [s(t, x, y)-b_{\text{ref}}] \\
+                                     &= -\gamma [h(t, x, y)+
+                                                 b(t, x, y)-b_{\text{ref}}],
+    \end{align}
+    %
+    thus accounting for the evolution of ice thickness, ${h=s-b}$, on the one
+    hand, and for differences between the basal topography of the ice flow
+    model, $b$, and the WorldClim reference topography, $b_{\text{ref}}$, on
+    the other hand. All simulations use an annual temperature lapse rate of
+    $\gamma=6\,\unit{K\,km^{-1}}$.
+
+% ----------------------------------------------------------------------
+\section{Supplementary}
+\label{sec:climate}
+% ----------------------------------------------------------------------
+
+    In this section, we analyze the model sensitivity to palaeo-climate forcing
+    through the last glacial cycle, using three palaeo-temperature records
+    (Sect.~\ref{sec:paltemp}) and two parametrizations of palaeo-precipitation
+    (Sect.~\ref{sec:palprec}), in terms of modelled evolution of total ice
+    volume (Sect.~\ref{sec:timeseries}) and glaciated area during MIS~2 and~4
+    (Sect.~\ref{sec:footprints}).
+
+    These simulations use a horizontal resolution of 2\,km. The vertical grid
+    consists of 31~temperature layers in the bedrock and up to 126~enthalpy
+    layers in the ice, corresponding to vertical resolutions of 100 and
+    40\,m, respectively. Simulations ran for two to four days
+    using 144~processors.
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Palaeo-temperature forcing}
+\label{sec:paltemp}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived from
+    palaeo-temperature proxy records from the Greenland Ice Core Project
+    \citep[GRIP;][]{Dansgaard.etal.1993}, the European Project for Ice Coring
+    in Antarctica \citep[EPICA;][] {Jouzel.etal.2007}, and an oceanic sediment
+    core from the Iberian margin \citep[MD01-2444;][]{Martrat.etal.2007}.
+    Palaeo-temperature anomalies from the GRIP record are calculated from
+    oxygen isotope (\chem{\delta^{18}O}) measurements using a~quadratic
+    equation \citep{Johnsen.etal.1995},
+    %
+    \begin{align}
+      {\Delta}T_{\text{TS}}(t) ={~}&-11.88 [\chem{\delta^{18}O}(t)
+                                    -\chem{\delta^{18}O}(0)] \nonumber \\
+                                   &-0.1925 [\chem{\delta^{18}O}(t)^2
+                                    -\chem{\delta^{18}O}(0)^2] \,,
+    \end{align}
+    %
+    while temperature reconstructions from the EPICA and MD01-2444 records are
+    provided as such. For each proxy record used and each of the parameter
+    set-ups used in the sensitivity tests, palaeo-temperature anomalies are
+    scaled linearly (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}a) so
+    that the modelled cumulative glaciated area during MIS~2 (29--14\,ka)
+    and within a rectangular region covering the Rhine
+    glacier piedmont lobe (Fig.~\ref{fig:footprints}a, black rectangle) is
+    consistent with the glaciated area of the geological reconstruction
+    \citep{Ehlers.etal.2011}.
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Palaeo-precipitation forcing}
+\label{sec:palprec}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    Finally, in some simulations (hereafter labelled PP), precipitation was
+    reduced with air temperature in
+    order to simulate the potential rarefaction of atmospheric moisture in
+    colder climates. This was done using an empirical relationship derived from
+    observed accumulation rates and oxygen isotopes concentrations in the GRIP
+    ice core \citep{Dahl-Jensen.etal.1993},
+    %
+    \begin{align}
+      {\Psi}_{\text{PP}}(t) = \exp[\psi{\Delta}T_{\text{TS}}(t)] \,,
+    \end{align}
+    %
+    with $\psi=0.169/2.4=0.0704$ \citep{Huybrechts.2002}. This simple
+    relationship likely does not reflect the complexity of atmospheric
+    circulation changes that governed moisture availability over the Alps
+    during the last glacial cycle, of which little is known apart from the
+    LGM \citep[cf.][]{Wu.etal.2007, Strandberg.etal.2011, Ludwig.etal.2016}.
+    Thus, another set of simulations presented here use constant
+    precipitation, corresponding to $\psi=0$.
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Sensitivity of ice volume evolution}
+\label{sec:timeseries}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    For the three palaeo-temperature records and the two palaeo-precipitation
+    parametrizations used, the model yields significant ice volume build-up
+    during MIS~4 and 2 (Fig.~\ref{fig:timeseries}b), corresponding to documented
+    glaciation periods in the Alps \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
+    All simulations also yield important glaciations during MIS~5 and 3, but
+    their timing and amplitude varies significantly depending on the climate
+    forcing used. All six simulations overestimate the late-glacial re-advance
+    during the Younger Dryas
+    \citep[12.9--11.7\,ka; cf. e.g.,][]{Ivy-Ochs.etal.2009}.
+    This is partly because the 2\,km resolution used in these simulations is
+    too coarse to resolve
+    Younger Dryas Alpine glaciers. However, the large total ice volume
+    overestimation resulting from the GRIP temperature forcing
+    (Fig.~\ref{fig:timeseries}b, blue curves) certainly indicates that the
+    vigorous cooling experienced in Greenland is not representative of the
+    Alps.
+
+    Geological data from the best documented Alpine piedmont lobes indicate
+    that their maximum extension during MIS~2 occurred at around 25.5--24.0\,ka
+    \citep{Monegato.etal.2017}, after which Alpine glaciers remained or
+    potentially re-advanced to within close reach of this maximum extent, until
+    as late as 22 to 17\,ka \citep[Fig.~5]{Wirsig.etal.2016}. This is in very
+    good agreement with the result of the EPICA simulations, which yield an
+    early maximum ice volume at 25.2/24.6 (without/with palaeo-precipitation
+    reductions) followed by a standstill until 17.3\,ka
+    (Fig.~\ref{fig:timeseries}b, blue curves). The GRIP palaeo-temperature
+    forcing, on the other hand, yields two distinct total ice volume maxima
+    at 27.3/27.0 and 21.8/21.7\,ka followed by an early deglaciation of the
+    foreland by 21.4\,ka. The MD01-2444 palaeo-temperature forcing yields
+    a late LGM peak ice volume at 16.5/15.7\,ka several thousand years after
+    the dated geological evidence.
+
+    Finally, all simulations yield very strong ice volume
+    variability, indicative of many more than two or three cycles of glacier
+    advance and retreats onto the foreland (Fig.~\ref{fig:timeseries}b).
+    Palaeo-precipitation reductions dampen some of the small-scale variability,
+    but they have little effect on the millennial scales that characterise
+    those cycles (Fig.~\ref{fig:timeseries}b, light colour curves). In
+    particular, strong millennial variability recorded in GRIP
+    \chem{\delta^{18}O} (Dansgaard-Oeschger events) have large repercussions
+    on the modelled Alpine ice volume, including around six glaciations of LGM
+    magnitude (Fig.~\ref{fig:timeseries}b, blue curves), which is not supported
+    by geologic evidence \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
+    Dansgaard-Oeschger events, however, have been recorded in Central European
+    lake sediments \citep{Wohlfarth.etal.2008} and cave speleothems
+    \citep{Spotl.Mangini.2002, Luetscher.etal.2015}. Our results indicate that
+    their magnitude in Central Europe was likely lower than that recorded in
+    GRIP \chem{\delta^{18}O}, which either implies smaller associated
+    palaeo-temperature fluctuations in Central Europe than Greenland, or
+    temperature-independent controls on GRIP \chem{\delta^{18}O}. In contrast,
+    the EPICA palaeo-temperature record has least variability among the records
+    used (Fig.~\ref{fig:timeseries}a, red curves). This results in smaller
+    variability in the total ice volume, and more restrictive glaciations
+    during MIS~5 and~3, (Fig.~\ref{fig:timeseries}b, red curves).
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Sensitivity of glaciated area}
+\label{sec:footprints}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    During MIS~2, all six simulations yield comparative cumulative glaciated
+    areas (Fig.~\ref{fig:footprints}a). This result is inherent to the
+    palaeo-climate forcing approach used, which involves a linear scaling of
+    each palaeo-temperature anomaly record to a geomorphological
+    reconstruction of the area glaciated by the Rhine Glacier piedmont lobe
+    (Sect.~\ref{sec:paltemp}; Fig.~\ref{fig:footprints}a, black rectangle).
+    Outside this benchmark, all simulations underestimate glaciated area in the
+    western part of the model domain, and overestimate it in the eastern
+    part, relative to the geomorphological ice limits
+    (Fig.~\ref{fig:footprints}a). This result might indicate that the LGM
+    temperature depression, here considered homogeneous, was actually lower in
+    the Eastern Alps than in the Western Alps, as previously shown by positive
+    degree-day modelling of the central European palaeo-ice caps
+    \citep{Heyman.etal.2013}. Alternatively, the east-west gradient in winter
+    precipitation existing today (Fig.~\ref{fig:inputs}h) was
+    perhaps enhanced during the LGM, as indicated by pollen reconstructions
+    \citep{Wu.etal.2007}. Finally, the LGM extent of Alpine glaciers might have
+    been affected by east-west gradient in variables not accounted for
+    by the PDD model, such as cloud cover or dust deposition.
+
+    Besides this general pattern, there exist differences in glaciated area
+    depending on the palaeo-climate forcing used. The MD01-2444, and to a
+    greater extent, the GRIP palaeo-temperature records, tend to overestimate
+    MIS~2 ice cover on all peripheral ranges that surround the Alps
+    (Fig.~\ref{fig:footprints}a and c). This is particularly true when no
+    precipitation corrections are applied (bright colours). In fact, both
+    palaeo-temperature records have a higher temperature variability and
+    contain brief spells of cold temperatures, which are too short to develop a
+    fully grown Alpine ice sheet, but long enough to build up ice cover on a
+    smaller scale on these peripheral ranges (Fig.~\ref{fig:timeseries}). On
+    the other hand, peripheral glaciation modelled using the EPICA record
+    (Fig.~\ref{fig:footprints}b), which has smaller temperature variability,
+    is in relatively good agreement with the geomorphological reconstructions.
+
+    The modelled extent of glaciation during MIS~4 depicts a more pronounced
+    sensitivity to the choice of palaeo-climate forcing used. Using the GRIP
+    and MD01-2444 palaeo-temperature records, Alpine glaciers are modelled to
+    extend well beyond the reach of documented LGM end-moraines
+    (Fig.~\ref{fig:footprints}d and f). Because, in both cases,
+    the modelled glaciation extent corresponds to brief cold spells in the
+    palaeo-temperature forcing (Fig.~\ref{fig:timeseries}a),
+    palaeo-precipitation reductions greatly reduce the excessive modelled ice
+    cover (Figs.~\ref{fig:timeseries}b and \ref{fig:footprints}d and f, light
+    colours). However, with or without palaeo-precipitation corrections,
+    the GRIP and MD01-2444 forcing yield
+    modelled ice extents (Fig.~\ref{fig:footprints}d and f) and volumes
+    (Fig.~\ref{fig:timeseries}b) considerably larger during MIS~4 than
+    during MIS~2, which is not supported by geological evidence
+    \citep{Preusser.2004, Ivy-Ochs.etal.2008, Husen.Reitner.2011,
+    Barrett.etal.2017, Haldimann.etal.2017}. Gravel deposits from the Rhine
+    Glacier indicate an early last glacial cycle foreland glaciation, but the
+    extent reached is less than that in MIS~2 \citep{Keller.Krayss.2010}.
+    Dropstones in lake sediments indicate a late MIS~4 advance restricted to
+    the upper Inn Valley \citep{Barrett.etal.2017}. For the Reuss Glacier, no
+    evidence for a pre-LGM glacier advance during the last glacial cycle is
+    available \citep{Haldimann.etal.2017}.
+
+    The EPICA palaeo-temperature forcing, on the other hand, yields a MIS~4
+    glaciation that is only slightly less expansive than the MIS~2 glaciation
+    (Fig.~\ref{fig:footprints}e). The sensitivity of the glaciated area to
+    palaeo-precipitation reductions is mostly limited to the southern terminal
+    lobes of central-Alpine glaciers, which are slightly less extensive during
+    MIS~2 and 4 under reduced precipitation (Fig.~\ref{fig:footprints}b and e).
+
+    Based on the above considerations on timing of the LGM during MIS~2, and
+    modelled ice
+    extent during MIS~4, we select EPICA as our optimal palaeo-temperature
+    record for a more detailed and higher-resolution comparison of modelled
+    glacier dynamics to available geological evidence
+    (Sect.~\ref{sec:results}). As a conservative approach in regard to the
+    rapid ice volume fluctuations, we choose to include palaeo-precipitation
+    corrections in the following higher-resolution run.
+
+%TC:break Acknowledgements
+% ----------------------------------------------------------------------
 % Acknowledgements
@@
 
-% ----------------------------------------------------------------------
-%TC: break References
+%TC:break References
+% ----------------------------------------------------------------------
 % References
@@
 
-% ----------------------------------------------------------------------
-%TC: break Figures
+%TC:break Figures
+% ----------------------------------------------------------------------
 % Figures
 \clearpage
+% ----------------------------------------------------------------------
+
+    \begin{figure*}
+      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
+      \caption{%
+        \textbf{(a)} Modelled bedrock topography (grey), ice surface topography
+        (200\,m contours), and ice surface velocity (blue) in the Alps
+        24.57\,ka before present, corresponding to the maximum modelled ice
+        cover. Modelled self-sustained ice domes (triangles) and some major
+        transfluences (crosses) are shown. Modelled LGM
+        ice extent (dashed orange line) and geomorphological
+        reconstruction \citep[solid red line,][]{Ehlers.etal.2011}. The
+        background map consists of depressed SRTM \citep{Jarvis.etal.2008}
+        topography and Natural Earth Data \citep{Patterson.Kelso.2017}.
+        \textbf{(b)} Temperature offset time-series from the EPICA ice core
+        used as palaeo-climate forcing for the ice flow model \citep[black
+        curve,][]{Jouzel.etal.2007}, and modelled total ice volume through the
+        last glacial cycle (120--0\,ka), expressed in meters of sea level
+        equivalent (m~s.l.e., blue curve). Gray fields indicate Marine
+        Oxygen Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
+        a~global compilation of benthic \chem{\delta^{18}O} records
+        \citep{Lisiecki.Raymo.2005}.}
+      \label{fig:lgmvel}
+    \end{figure*}
+
+    \begin{figure*}
+      \centerline{\includegraphics{alpcyc_hr_dynamics}}
+      \caption{%
+        \textbf{(a)} Timing of the LGM given by the modelled age of maximum ice
+        thickness throughout the entire simulation (colour mapping) and
+        corresponding, ice surface elevation (200\,m contours). Observed upper
+        Rhone valley trimline locations \citep[black dots;][Table~1]
+        {Kelly.etal.2004} and areas modelled to have experienced
+        less than 1\,ka temperate ice cover during MIS~2 (29--14\,ka, hatches).
+        The LGM is here modelled as a time-transgressive event. Profile lines
+        roughly following valley centerlines for the Rhine and Inn Glaciers.
+        \textbf{(b)} Comparison of modelled ice surface elevation at the LGM
+        (time-transgressive, corresponding to maximum ice thickness,
+        Fig.~\ref{fig:timing}), compensated for bedrock deformation, against
+        observed trimline elevations \citep[Table~1]{Kelly.etal.2004}. Colours
+        show the age of maximum ice thickness (as in Fig.~\ref{fig:timing}).
+        Model variables were bilinearly interpolated to the trimline locations.
+        \textbf{(c)} Evolution of modelled glacier extent in
+        time, bilinearly interpolated along the Rhine and Inn profiles, showing
+        numerous cycles of advance and retreat over the last glacial cycle
+        modulated by subglacial topography and catchment geometry.
+        Isolated patches indicate periodic surges from tributary glaciers.}
+      \label{fig:timing}
+      \label{fig:trimlines}
+      \label{fig:profiles}
+    \end{figure*}
+
+%TC:break Supplementary figures
+% ----------------------------------------------------------------------
+% Supplementary figures
+\clearpage
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}%
 % ----------------------------------------------------------------------
@@
 
-    \begin{sidewaysfigure*}
-      % scalebox is not necessary in production mode
-      \centerline{\scalebox{0.9}{\includegraphics{alpcyc_hr_lgmvel}}}
-      \caption{%
-        \textbf{(a)} Modelled bedrock topography (grey), ice surface topography
-        (200\,m contours), and ice surface velocity (blue) in the Alps
-        24.57\,ka before present, corresponding to the maximum modelled ice
-        cover. Modelled self-sustained ice domes (triangles) and some major
-        transfluences (crosses) are shown. Modelled LGM
-        ice extent (dashed orange line) and geomorphological
-        reconstruction \citep[solid red line,][]{Ehlers.etal.2011}. The
-        background map consists of depressed SRTM \citep{Jarvis.etal.2008}
-        topography and Natural Earth Data \citep{Patterson.Kelso.2017}.
-        \textbf{(b)} Temperature offset time-series from the EPICA ice core
-        used as palaeo-climate forcing for the ice flow model \citep[black
-        curve,][]{Jouzel.etal.2007}, and modelled total ice volume through the
-        last glacial cycle (120--0\,ka), expressed in meters of sea level
-        equivalent (m~s.l.e., blue curve). Gray fields indicate Marine
-        Oxygen Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
-        a~global compilation of benthic \chem{\delta^{18}O} records
-        \citep{Lisiecki.Raymo.2005}.}
-      \label{fig:lgmvel}
-    \end{sidewaysfigure*}
-
-    \begin{sidewaysfigure*}
-      % scalebox is not necessary in production mode
-      \centerline{\scalebox{0.9}{\includegraphics{alpcyc_hr_timing}}}
-      \caption{%
-        \textbf{(a)} Timing of the LGM given by the modelled age of maximum ice
-        thickness throughout the entire simulation (colour mapping) and
-        corresponding, ice surface elevation (200\,m contours).
-        \textbf{(b)} Temperature offset time-series from the EPICA ice core
-        used as palaeo-climate forcing for the ice flow model (black curve),
-        and modelled glacierized area during the LGM (coloured curve). The LGM
-        is here modelled as a time-transgressive event.}
-      \label{fig:timing}
-    \end{sidewaysfigure*}
-
-    \begin{figure}
-      \centerline{\includegraphics{alpcyc_hr_trimlines}}
-      \caption{%
-        \textbf{(a)} Comparison of modelled ice surface elevation at the LGM
-        (time-transgressive, corresponding to maximum ice thickness,
-        Fig.~\ref{fig:timing}), compensated for bedrock deformation, against
-        observed trimline elevations \citep[Table~1]{Kelly.etal.2004}. Colours
-        show the age of maximum ice thickness (as in Fig.~\ref{fig:timing}).
-        Model variables were bilinearly interpolated to the trimline locations.
-        \textbf{(b)} Histogram of differences between modelled LGM ice surface
-        elevation and trimline elevations (500\,m bands). The average
-        difference is 861\,m.
-        \textbf{(c)} Observed upper Rhone valley trimline locations
-        \citep[Table~1]{Kelly.etal.2004}, modelled age of maximum ice thickness
-        (colour) and LGM ice surface elevation (200\,m contours). Hatches
-        mark areas modelled to have experienced less than 1\,ka temperate ice
-        cover during MIS~2 (29--14\,ka).}
-      \label{fig:trimlines}
-    \end{figure}
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpcyc_hr_profiles}}
-      \caption{%
-        \textbf{(a, c, e, g, i, k)} Profile lines roughly following valley
-        centerlines for the Rhine, Rhone, Dora Baltea and Isère, Inn and
-        Tagliamento Glaciers.
-        \textbf{(b, d, f, h, j, l)} Evolution of modelled glacier extent in
-        time, bilinearly interpolated along the corresponding profiles, showing
-        numerous cycles of advance and retreat over the last glacial cycle
-        modulated by subglacial topography and catchment geometry.
-        Isolated patches indicate periodic surges from tributary glaciers.}
-      \label{fig:profiles}
-    \end{figure*}
-
-% ----------------------------------------------------------------------
-% Tables
+%TC:break Supplementary tables
+% ----------------------------------------------------------------------
+% Supplementary tables
 \clearpage
+\setcounter{table}{0}
+\renewcommand{\thetable}{S\arabic{table}}%
 % ----------------------------------------------------------------------

### 2017-12-23 figures/util/pl.py
Convert other figs to 177 mm and shrink margins.

@@ -97,8 +97,8 @@ def cut_ts_axes(ax, tsw=2/3., tsh=1/3.):
     rectkw = dict(ec='w', fc='w', zorder=-1, **commkw)
     poly = iplt.Polygon(zip(x, y), **polykw)
     rect = iplt.Rectangle((1-tsw, 0.0), tsw, tsh, **rectkw)
-    tsax = fig.add_axes([pos.x1-tsw*(pos.x1-pos.x0)+12.5/figw, 10.0/figh,
-                         tsw*(pos.x1-pos.x0)-25.0/figw,
+    tsax = fig.add_axes([pos.x1-tsw*(pos.x1-pos.x0)+12.0/figw, 9.0/figh,
+                         tsw*(pos.x1-pos.x0)-24.0/figw,
                          tsh*(pos.y1-pos.y0)-15.0/figh])
     tsax.add_patch(poly)
     tsax.add_patch(rect)
@@ -119,11 +119,11 @@ def cut_ts_sc_axes(ax, tsw=2/3., tsh=1/3., scw=1/5., sch=1/3.):
     poly = iplt.Polygon(zip(x, y), **polykw)
     screct = iplt.Rectangle((0.0, 1-sch), scw, sch, **rectkw)
     tsrect = iplt.Rectangle((1-tsw, 0.0), tsw, tsh, **rectkw)
-    scax = fig.add_axes([pos.x0+7.5/figw, pos.y1-sch*(pos.y1-pos.y0)+10.0/figh,
-                         scw*(pos.x1-pos.x0)-10.0/figw,
-                         sch*(pos.y1-pos.y0)-10.0/figh])
-    tsax = fig.add_axes([pos.x1-tsw*(pos.x1-pos.x0)+12.5/figw, 10.0/figh,
-                         tsw*(pos.x1-pos.x0)-25.0/figw,
+    scax = fig.add_axes([pos.x0+7.5/figw, pos.y1-sch*(pos.y1-pos.y0)+9.0/figh,
+                         scw*(pos.x1-pos.x0)-9.0/figw,
+                         sch*(pos.y1-pos.y0)-9.0/figh])
+    tsax = fig.add_axes([pos.x1-tsw*(pos.x1-pos.x0)+12.0/figw, 9.0/figh,
+                         tsw*(pos.x1-pos.x0)-24.0/figw,
                          tsh*(pos.y1-pos.y0)-15.0/figh])
     scax.add_patch(screct)
     tsax.add_patch(poly)
[…]

### 2017-12-24 figures/util/pl.py
Move cut axes polygons to main axes.

@@ -92,16 +92,16 @@ def cut_ts_axes(ax, tsw=2/3., tsh=1/3.):
     ax.outline_patch.set_ec('none')
     x = [0.0, 1-tsw, 1-tsw, 1.0, 1.0, 0.0, 0.0]
     y = [0.0, 0.0, tsh, tsh, 1.0, 1.0, 0.0]
-    commkw = dict(clip_on=False, transform=ax.transAxes)
-    polykw = dict(ec='k', fc='none', zorder=3, **commkw)
-    rectkw = dict(ec='w', fc='w', zorder=-1, **commkw)
+    commkw = dict(clip_on=False, transform=ax.transAxes, zorder=3)
+    polykw = dict(ec='k', fc='none', **commkw)
+    rectkw = dict(ec='w', fc='w', **commkw)
     poly = iplt.Polygon(zip(x, y), **polykw)
     rect = iplt.Rectangle((1-tsw, 0.0), tsw, tsh, **rectkw)
     tsax = fig.add_axes([pos.x1-tsw*(pos.x1-pos.x0)+12.0/figw, 9.0/figh,
                          tsw*(pos.x1-pos.x0)-24.0/figw,
                          tsh*(pos.y1-pos.y0)-15.0/figh])
-    tsax.add_patch(poly)
-    tsax.add_patch(rect)
+    ax.add_patch(rect)
+    ax.add_patch(poly)
     return tsax
 
 
@@ -113,9 +113,9 @@ def cut_ts_sc_axes(ax, tsw=2/3., tsh=1/3., scw=1/5., sch=1/3.):
     ax.outline_patch.set_ec('none')
     x = [0.0, 1-tsw, 1-tsw, 1.0, 1.0, scw, scw, 0.0, 0.0]
     y = [0.0, 0.0, tsh, tsh, 1.0, 1.0, 1-sch, 1-sch, 0.0]
-    commkw = dict(clip_on=False, transform=ax.transAxes)
-    polykw = dict(ec='k', fc='none', zorder=3, **commkw)
-    rectkw = dict(ec='w', fc='w', zorder=-1, **commkw)
+    commkw = dict(clip_on=False, transform=ax.transAxes, zorder=3)
+    polykw = dict(ec='k', fc='none', **commkw)
+    rectkw = dict(ec='w', fc='w', **commkw)
     poly = iplt.Polygon(zip(x, y), **polykw)
[…]

### 2017-12-27 figures/alpcyc_lr_footprints.py
Add model domain drawing function.

@@ -21,7 +21,7 @@ for i, rec in enumerate(ut.alpcyc_records):
     pp = 'pp' in conf
 
     # add scaling domain and outline on top panel only
-    ut.pl.draw_scaling_domain(ax)
+    ut.pl.draw_model_domain(ax, extent='rhlobe')
     ut.pl.draw_lgm_outline(ax, c='k')
 
     # set title

### 2017-12-27 figures/alpcyc_lr_records.py

@@ -18,10 +18,7 @@ ax.gridlines(color='0.5', linestyle='-', linewidth=0.1)
 ax.coastlines(edgecolor='k', lw=0.25)
 
 # add modelling domain
-w, e, s, n = 150e3, 1050e3, 4800e3, 5475e3
-x = [w, w, e, e, w]
-y = [s, n, n, s, s]
-ax.plot(x, y, 'k', lw=1.0, transform=ut.pl.utm, zorder=3)
+ut.pl.draw_model_domain(ax=ax, extent='alps')
 
 # add record locations
 records = ['GRIP', 'EPICA', 'MD01-2444']

### 2017-12-27 figures/util/pl.py

@@ -746,9 +746,9 @@ def draw_boot_topo(ax=None, res='1km'):
     return im
 
 
-def draw_scaling_domain(ax=None):
+def draw_model_domain(ax=None, extent='alps'):
     """Add Rhine lobe scaling domain."""
-    w, e, s, n = regions['rhlobe']
+    w, e, s, n = regions[extent]
     x = [w, e, e, w, w]
     y = [s, s, n, n, s]
     ax.plot(x, y, c='k', lw=0.5)

### 2017-12-27 figures/alpcyc_hr_trimlines.py
Add place names on trimlines figure.

@@ -118,6 +118,15 @@ ut.pl.draw_boot_topo(ax)
 ut.pl.draw_envelope(ax, levels=[0e3, 3e3], colors='w')
 ut.pl.draw_natural_earth(ax)
 
+# add text labels
+kw = dict(ha='center', va='center', transform=ut.pl.ll)
+ax.text(6.865, 45.834, 'Mont\nBlanc', **kw)
+ax.text(7.659, 45.976, 'Matter-\nhorn', **kw)
+ax.text(7.867, 45.937, 'Monte\nRosa', **kw)
+ax.text(7.963, 46.537, 'Jungfrau', **kw)
+#ax.text(8.126, 46.537, 'Finsternaarhorn', **kw)
+ax.text(7.367, 46.233, 'Rhone', rotation=30, **kw)
+
 # draw trimlines
 sc = ax.scatter(xt, yt, c=at, cmap=cmap, norm=norm, s=4**2, alpha=0.75)
 

### 2017-12-27 figures/alpcyc_hr_deglacage.py
Add deglaciation age figure.

@@ -1,0 +1,38 @@
+#!/usr/bin/env python2
+# coding: utf-8
+
+import util as ut
+import numpy as np
+import matplotlib.pyplot as plt
+
+# initialize figure
+fig, ax, cax = ut.pl.subplots_cax(extent='west')
+
+
+# Map axes
+# --------
+
+# read postprocessed data
+deglacage, extent = ut.io.load_postproc_gtif(ut.alpcyc_bestrun, 'deglacage')
+
+# set contour levels, colors and hatches
+levs = range(0, 25, 2)
+cmap = plt.get_cmap('Spectral', len(levs)+1)
+cols = cmap(range(len(levs)+1))
+
+# plot
+cs = ax.contourf(deglacage/1e3, levs, extent=extent, colors=cols, extend='both', alpha=0.75)
+ax.contour(deglacage.mask, [0.5], extent=extent, colors='k', linewidths=0.5)
+
+# add cartopy vectors
+ut.pl.draw_boot_topo(ax)
+ut.pl.draw_natural_earth(ax)
+ut.pl.draw_lgm_outline(ax)
+ut.pl.draw_footprint(ax)
+
+# add colorbar
+cb = ut.pl.add_colorbar(cs, cax)
+cb.set_label(r'age of glaciation (ka)')
+
+# save figure
+ut.pl.savefig()

### 2017-12-27 articles/cycle/alps-cycle.tex
Make up a temporary new attractive title.
Restore sideways figures orientation to portrait.
Add corrections from Susan.

--- 2017-11-20 articles/cycle/alps-cycle.tex
@@
     (69--62\,ka) and~2 (34--18\,ka) spatially and temporally consistent with
-    the geological reconstructions.
+    the geological reconstructions, while the other records used result in
+    excessive early glacial cycle ice cover and a late Last Glacial Maximum.
     Despite the low variability of this Antarctic-based climate forcing, our
@@
     subglacial topography but include occasional transfluences and
-    self-sustained ice domes. Modelled maximum ice surface is about 900\,m
-    higher than observed trimline elevations in the interior of the Alps, yet
+    self-sustained ice domes. Modelled maximum ice surface is 861\,m
+    higher than observed trimline elevations in the upper Rhone Valley, yet
     our simulation predicts little erosion at high elevation due to cold-based
@@
     glaciations in the Alps \citep{Penck.Bruckner.1909}. More recent studies
-    of glaciofluvial stratigraphy in the foreland indicate up to 15 glaciations
-    \citep{Schluchter.1988, Ivy-Ochs.etal.2008, Preusser.etal.2011}.
+    of glaciofluvial stratigraphy in the foreland indicate at least 15
+    glaciations \citep{Schluchter.1988, Ivy-Ochs.etal.2008,
+    Preusser.etal.2011}.
 
@@
     century, palaeo-climate records extracted from deep sea sediments and
-    ice cores have provided a much more detailed picture of the Earth
+    ice cores have provided a much more detailed picture of the Earth's
     environmental history \citep[e.g.,][]{Emiliani.1955,
@@
     sea-level fluctuations.
-
     Besides, the landform record is typically sparse in time and, most often,
@@
     dynamics, have been reconstructed from moraines and trimlines across the
-    mountain range \citep[e.g.,][]{Bini.etal.2009, Coutterand.2010,
-    Husen.2011}. The assumption that trimlines, the upper limit of glacial
+    mountain range \citep[e.g.,][]{Penck.Bruckner.1909, Castiglioni.1940,
+    Husen.1987, Bini.etal.2009, Coutterand.2010}. The assumption that
+    trimlines, the upper limit of glacial
     erosion, represent the maximum ice surface elevation, has
@@
     different piedmont
-    lobes exists \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear whether it
-    relate to climate or glacier dynamics \citep{Monegato.etal.2017} or
+    lobes exists \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear whether
+    this relates to climate or glacier dynamics \citep{Monegato.etal.2017} or
     uncertainties in the dating methods.
@@
     (Sect.~\ref{sec:icedyn}). Basal sliding follows a~pseudo-plastic law where
-    the yield stress accounts for till deconsolidation under high water
+    the yield stress accounts for till dilatation under high water
     pressure (Sect.~\ref{sec:sliding}). Bedrock topography is deflected
@@
     (Fig.~\ref{fig:timeseries}b, blue curves) certainly indicates that the
-    vigorous cooling experienced in Greenland is not representative of the
-    Alps.
+    vigorous cooling experienced during the Younger Dryas in Greenland is not
+    representative of the Alps.
 
@@
     potentially re-advanced to within close reach of this maximum extent, until
-    as late as 22 to 17\,ka \citep[Fig.~5]{Wirsig.etal.2016}. This is in very
+    as late as 22 to 17\,ka \citep[Fig.~5]{Ivy-Ochs.2015, Wirsig.etal.2016}.
+    This is in very
     good agreement with the result of the EPICA simulations, which yield an
@@
     (Fig.~\ref{fig:footprints}a). This result might indicate that the LGM
-    temperature depression, here considered homogeneous, was actually lower in
+    temperature depression, here taken as homogeneous, was actually lower in
     the Eastern Alps than in the Western Alps, as previously shown by positive
@@
     Dropstones in lake sediments indicate a late MIS~4 advance restricted to
-    the upper Inn Valley \citep{Barrett.etal.2017}. For the Reuss Glacier, no
+    the upper Inn Valley \citep{Barrett.etal.2017}. For the Linth Glacier, no
     evidence for a pre-LGM glacier advance during the last glacial cycle is
@@
     palaeo-precipitation reductions is mostly limited to the southern terminal
-    lobes of central-Alpine glaciers, which are slightly less extensive during
-    MIS~2 and 4 under reduced precipitation (Fig.~\ref{fig:footprints}b and e).
+    lobes of central-Alpine glaciers, where reduced precipitation results in a
+    slightly lesser extent during both MIS~2 and 4 (Fig.~\ref{fig:footprints}b
+    and e).
 
@@
     The LGM extent of Alpine glaciers has been mapped with varying level of
-    detail across the Western \citep{Coutterand.2010,
-    Buoncristiani.Campy.2011}, Central \citep{Jackli.1962, Bini.etal.2009,
-    Hantke.2011}, and Eastern Alps \citep{Penck.Bruckner.1909, Husen.1987,
-    BGR.2007, Husen.2011, Bavec.Verbic.2011}.
-    Some of these maps have been compiled
+    detail across the Western \citep{Coutterand.2010, Buoncristiani.Campy.2011,
+    Jackli.1962, Bini.etal.2009, Hantke.2011}, and Eastern Alps
+    \citep{Penck.Bruckner.1909, Castiglioni.1940, Husen.1987, BGR.2007,
+    Husen.2011, Bavec.Verbic.2011}. Some of these maps have been compiled
     into a reconstruction, covering the entire Alps \citep{Ehlers.etal.2011},
@@
     and the Lyon Lobe, is underestimated in the model results
-    (Fig.~\ref{fig:lgmvel}a). On the other hand, the model yields excessive ice
+    \citep[Fig.~\ref{fig:lgmvel}a, cf][]{Bini.etal.2009, Coutterand.2010}.
+    On the other hand, the model yields excessive ice
     cover in the eastern Alps, where the Mur, Drava and Sava Glaciers are
     modelled to extend tens of km beyond the mapped ice limits
-    (Fig.~\ref{fig:lgmvel}a). As previously discussed, these discrepancies
-    might indicate that the
+    \citep[Fig.~\ref{fig:lgmvel}a, cf][]{Husen.1987, Bavec.Verbic.2011}.
+    As previously discussed, these discrepancies might indicate that the
     LGM climate was characterized by an east-west gradient in temperature
@@
     constant-climate modelling results \citep{Becker.etal.2016}, it follows
-    from introducing a time-dependent palaeo-temperature forcing. Constant
-    climate forcing systematically resulted in extraneous
+    from introducing a time-dependent palaeo-temperature forcing. Without
+    introducing differential precipitation change north and south of the Alps,
+    constant climate forcing systematically resulted in extraneous
     \citep[Fig.~3]{Becker.etal.2016} and premature
@@
 
-    On a more regional level, the maximum extent of the Ticino, Iseo,
-    Garda and Piave Glaciers in the central southern Alps is modelled
+    On a more regional level, the maximum extent of the Ticino, Oglio,
+    Adige and Piave Glaciers in the central southern Alps is modelled
     several kilometres within the mapped LGM moraines (Fig.~\ref{fig:lgmvel}a).
@@
     \citep[Fig.~2.4]{Penck.Bruckner.1909, Husen.2011}. It is also known that
-    ice flew across the Gailberg and Kreuzberg saddles \citep{Husen.1985}, and
-    Pyhrn Pass \citep[Fig.~2.5]{Husen.2011}.
+    ice flowed across the Gailberg and Kreuzberg saddles \citep{Husen.1985},
+    and Pyhrn Pass \citep[Fig.~2.5]{Husen.2011}.
     On the other hand, no transfluence has previously been documented at the
@@
     in two locations over Flüela Pass and Ötztal (Fig.~\ref{fig:lgmvel}a).
+    \todo{Compare to Florineth.}
     Except for too extensive ice cover in the easternmost part of the model
@@
     isotope dating techniques at multiple locations around the Alps (cf.
-    \citealp[Fig.~5]{Wirsig.etal.2016}, for a review and the more recent
+    \citealp{Ivy-Ochs.2015};
+    \citealp[Fig.~5]{Wirsig.etal.2016} for reviews; and the more recent
     publications by \citealp{Monegato.etal.2017, Federici.etal.2017}). These
@@
     and 20\,ka (calibrated \chem{^{14}C} and \chem{^{10}Be} ages), but also
-    that terminal lobes stayed in the foreland until as late as 22 to 17\,ka,
+    that terminal lobes stayed in the foreland until 22\,ka to 17\,ka,
     when they experienced rapid retreat synchronously to lowering of the ice
-    surface in the mountains (\citealp[Fig.~5]{Wirsig.etal.2016};
+    surface in the mountains (\citealp{Ivy-Ochs.2015}; \citealp[Fig.~5]{Wirsig.etal.2016};
     \citealp[Fig.~3]{Monegato.etal.2017}).
@@
     Dora Riparia, Dora Baltea, and Tagliamento Glaciers reach an early maximum before
-    27\,ka; the Durance, Rhone, Inn, Enns, Ticino, and Garda Glaciers reach
+    27\,ka; the Durance, Rhone, Inn, Enns, Ticino, and Adige Glaciers reach
     their maximum thickness in phase with the overall Alpine areal maximum
-    around 25\,ka; while the Isère, Como and Iseo Glaciers reach a late maximum
+    around 25\,ka; while the Isère, Adda and Oglio Glaciers reach a late maximum
     after 24\,ka (Fig.~\ref{fig:timing}). Remarkably, peripheral ice caps on
@@
     For large Alpine glaciers flowing in deep valleys, such as the Rhone,
-    Rhine and Garda Glaciers, several thousand model years are needed to attain
+    Rhine and Adige Glaciers, several thousand model years are needed to attain
     a thermodynamical equilibrium between cold-ice advection from the high
@@
     balanced extent before thinning and receding towards the mountains as they
-    warm towards thermodynamical equilibrium (supplementary animation). On the
-    opposite, peripheral ice caps with basal topography restricted to high
+    warm towards thermodynamical equilibrium (supplementary animation). In
+    contrast, peripheral ice caps with basal topography restricted to high
     elevations experience very low shear strain and virtually no basal sliding.
@@
     Following the general assumption that trimlines, which mark the transition
-    between the steep frost-shattered ridges and the more gentle glacially eroded
+    between the steep frost-shattered ridges and the more gentle glacially sculpted
     valley troughs, represent the maximum elevation of the LGM ice surface,
@@
     Kelly.etal.2004, Bini.etal.2009, Coutterand.2010, Cossart.etal.2012}.
-    However, this assumption is challenged by geological evidence from
+    However, this assumption is challenged by geomorphological and cosmogenic
+    nuclide dating evidence from
     Scandinavia \citep[e.g.,][]{Kleman.1994, Kleman.Borgstrom.1994},
@@
     late-glacial ice surface elevation \citep[Fig.~1, p.~403]{Coutterand.2010}.
-    However, no evidence for cold-based glaciation has yet been reported in the
-    Alps.
+    However, no such evidence has been reported in the Alps.
 
@@
 
-    In the upper Rhone valley, the geographical location of observed trimlines
+    In the upper Rhone Valley, the geographical location of observed trimlines
     show a remarkable agreement with the areal extent of regions that are
@@
     Although the four glaciations model \citep{Penck.Bruckner.1909} has long
-    been used, it is now known that glaciers advanced onto the foreland many
-    more times \citep{Schluchter.1991, Preusser.etal.2011}. Sparse geological
+    been used, it is now known that glaciers advanced onto the foreland at
+    least 15 times since the beginning of the Quaternary at 2.58\,Ma
+    \citep{Schluchter.1991, Preusser.etal.2011}. Sparse geological
     data indicate that the last glacial cycle may have comprised two or even
@@
 
-    As previously mentioned (Sect.~\ref{sec:timeseries}), independently of the
+    As previously mentioned (Sect.~\ref{sec:timeseries}), independent of the
     palaeo-temperature (GRIP, EPICA, or MD01-2444) and palaeo-precipitation
@@
     variability throughout the last glacial cycle (Fig.~\ref{fig:lgmvel}b).
-
     Two major glaciations occur during MIS~4 and 2 (Fig.~\ref{fig:lgmvel}b).
@@
     (Fig.~\ref{fig:profiles}g and h) and Inn (Fig.~\ref{fig:profiles}i and j)
-    glaciers, with their complex system of confluences and diffluences, need
+    Glaciers, with their complex system of confluences and diffluences, need
     longer time to build up and reach the foreland only two or three times
@@
             self-sustained ice domes may have occurred (Sect.~\ref{sec:flow}).
-      \item The LGM was a transient stage in which glaciers were not in
-            equilibrium with climate. Its timing potentially varied across the
+      \item The LGM (maximum) extent was a transient stage in which glaciers
+            were out of balance with the
+            contemporary climate. Its timing potentially varied across the
             range due to inherent glacier dynamics (Sect.~\ref{sec:timing}).
@@
     flow physics. In particular, additional palaeo-climate variability over
-    Central Europe may have caused more glaciations than modelled here.
+    the European Alps may have caused more glaciations than modelled here.
     Through more detailed sensitivity studies, targeted to
@@
 
-    \begin{sidewaysfigure*}
-      % scalebox is not necessary in production mode
-      \centerline{\scalebox{0.9}{\includegraphics{alpcyc_hr_lgmvel}}}
+    \begin{figure*}
+      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
       \caption{%
@@
       \label{fig:lgmvel}
-    \end{sidewaysfigure*}
-
-    \begin{sidewaysfigure*}
-      % scalebox is not necessary in production mode
-      \centerline{\scalebox{0.9}{\includegraphics{alpcyc_hr_timing}}}
+    \end{figure*}
+
+    \begin{figure*}
+      \centerline{\includegraphics{alpcyc_hr_timing}}
       \caption{%
@@
       \label{fig:timing}
-    \end{sidewaysfigure*}
+    \end{figure*}
 
@@
         Fig.~\ref{fig:timing}), compensated for bedrock deformation, against
-        observed trimline elevations \citep[Table~1]{Kelly.etal.2004}. Colours
+        observed trimline elevations for the upper Rhone Glacier
+        \citep[Table~1]{Kelly.etal.2004}. Colours
         show the age of maximum ice thickness (as in Fig.~\ref{fig:timing}).
@@
         difference is 861\,m.
-        \textbf{(c)} Observed upper Rhone valley trimline locations
+        \textbf{(c)} Observed upper Rhone Valley trimline locations
         \citep[Table~1]{Kelly.etal.2004}, modelled age of maximum ice thickness
         (colour) and LGM ice surface elevation (200\,m contours). Hatches
-        mark areas modelled to have experienced less than 1\,ka temperate ice
-        cover during MIS~2 (29--14\,ka).}
+        mark areas modelled to have been glaciated but to have experienced
+        temperate ice for less than 1\,ka during MIS~2 (29--14\,ka).}
       \label{fig:trimlines}

### 2017-12-27 articles/cycle/alps-cycle.tex
Add discussion of ice domes.

@@
     in two locations over Flüela Pass and Ötztal (Fig.~\ref{fig:lgmvel}a).
-    \todo{Compare to Florineth.}
+    These modelled ice domes exhibit radial ice surface slopes and radial flow
+    velocities crosscutting the underlying topographic troughs and ridges.
+    On the one hand, ice-dispersal centres previously labelled as ice domes
+    in the upper Rhone and Rhine Valley \citep{Florineth.Schluchter.1998,
+    Kelly.etal.2004, Bini.etal.2009} feature as mere ice divides in our
+    simulation. On the other hand, the modelled ice dome over Flüela Pass
+    overlaps with the previously mapped Engadine ice dome
+    \citep{Florineth.1998}, although its centre is modelled ca.~20\,km north of
+    that indicated by geomorphology \citep[Fig.~3]{Florineth.1998}.
+
     Except for too extensive ice cover in the easternmost part of the model

### 2017-12-27 articles/cycle/alps-cycle.tex
Fix parameter table overflow in manuscript mode.
Adress a small comment by Martin and Guillaume.

@@
     the two approximations \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}.
-    Although this heuristic approach implies errors in the transition zone
-    where gravitational stresses intervene both in the SIA and SSA velocity
-    computation, it was shown to effectively approximate complex ice field flow
+    Although this heuristic approach is inaccurate in the transition zone
+    where both vertical shear and longitudinal stretch are significant,
+    it was shown to effectively approximate complex ice field flow
     dynamics over mountainous topographies similar to that of the European Alps
@@
       \label{tab:params}
-      \scalebox{0.8}
+      \scalebox{0.75}
       {\begin{tabular}{llrll}

### 2018-02-15 articles/cycle/alps-cycle.tex
Move figures up into the text.
Apply editor corrections for TCD publication.

@@
 
+\interfootnotelinepenalty=10000  % avoid footnote page break
+
 \hypersetup{colorlinks, linkcolor=darkorange,
@@
     Here, we use the Parallel Ice Sheet Model (PISM) to model the entire last
-    glacial cycle (120--0\,ka) in the Alps, using 576~processors and a
-    horizontal resolution of
-    1\,km. Climate forcing is derived using present-day climate data from
+    glacial cycle (120--0\,ka) in the Alps, using horizontal resolutions of 2
+    and 1\,km and up to 576~processors.
+    Climate forcing is derived using present-day climate data from
     WorldClim and the ERA-Interim reanalysis, and time-dependent temperature
@@
     Shackleton.Opdyke.1973, Dansgaard.etal.1993, Augustin.etal.2004},
-    indicating tens of glacial and interglacial periods. During the last
-    800\,000 years before present (800\,ka), these \emph{glacial cycles}
+    indicating several tens of glacial and interglacial periods. During the
+    last 800\,ka before present, these \emph{glacial cycles}
     have succeeded each other with a periodicity about 100\,ka \citep{Hays.etal.1976,
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{figure*}[t]
+      \centerline{\includegraphics{alpcyc_hr_inputs}}
+      \caption{%
+        \textbf{(a)} Geothermal heat flow from applying the similarity method
+        to multiple geophysical proxies \citep{Goutorbe.etal.2011} used as a
+        boundary condition to the bedrock thermal model 3\,km below the
+        ice-bedrock interface.
+        \textbf{(b)} Initial basal topography from SRTM
+        \citep{Jarvis.etal.2008} and geomorphological reconstruction of Last
+        Glacial Maximum Alpine glacier extent \citep[solid red
+        line,][]{Ehlers.etal.2011}.
+        \textbf{(c)} Extract from the estimated of present-day ice thickness
+        \citep{Huss.Farinotti.2012} substracted from the SRTM topography and
+        aggregated to a 1\,km horizontal resolution.
+        \textbf{(d)} Modern January and \textbf{(e)} July standard deviation
+        \citep{Seguinot.2013} of daily mean temperature from the ERA-Interim
+        \citep[1979--2012;][]{Dee.etal.2011} from the reference monthly
+        climatology used to force the surface mass balance (PDD) component of
+        the ice sheet model.
+        \textbf{(f)} Modern January and \textbf{(g)} July mean near-surface air
+        temperature, and \textbf{(h, i)} precipitation from WorldClim
+        \citep[1960--1990;][]{Hijmans.etal.2005}.}
+      \label{fig:inputs}
+    \end{figure*}
 
@@
 
+    \begin{table*}[p]
+      \caption{%
+        Parameter values used in the ice sheet model. Symbols refer to
+        equations used by \citet{Seguinot.2014, Seguinot.etal.2016}.}
+      \label{tab:params}
+      \scalebox{0.75}
+      {\begin{tabular}{llrll}
+        \tophline
+
+        Not.    & Name & Value & Unit & Source \\
+
+        \middlehline
+        \multicolumn{2}{l}{{Ice rheology}} \\
+        \middlehline
+
+        $\rho$  & Ice density
+                & 910
+                & \unit{kg\,m^{-3}}
+                & \citet{Aschwanden.etal.2012} \\
+
+        $g$     & Standard gravity
+                & 9.81
+                & \unit{m\,s^{-2}}
+                & \citet{Aschwanden.etal.2012} \\
+
+        $n$     & Glen exponent
+                & 3
+                & --
+                & \citet[p.~55--57]{Cuffey.Paterson.2010} \\
+
+        $A_{\mathrm{c}}$   & Ice hardness coefficient cold
+                & $2.847 \times 10^{-13}$
+                & \unit{Pa^{-3}\,s^{-1}}
+                & \citet[p.~72]{Cuffey.Paterson.2010} \\
+
+        $A_{\mathrm{w}}$   & Ice hardness coefficient warm
+                & $2.356 \times 10^{-2}$
+                & \unit{Pa^{-3}\,s^{-1}}
+                & \citet[p.~72]{Cuffey.Paterson.2010} \\
+
+        $Q_{\mathrm{c}}$   & Flow law activation energy cold
+                & $6.0 \times 10^4$
+                & \unit{J\,mol^{-1}}
+                & \citet[p.~72]{Cuffey.Paterson.2010} \\
+
+        $Q_{\mathrm{w}}$   & Flow law activation energy warm
+                & $11.5 \times 10^4$
+                & \unit{J\,mol^{-1}}
+                & \citet[p.~72]{Cuffey.Paterson.2010} \\
+
+        $E_{\text{SIA}}$   & SIA enhancement factor
+                & 2
+                & --
+                & \citet[p.~77]{Cuffey.Paterson.2010} \\
+
+        $E_{\text{SSA}}$   & SSA enhancement factor
+                & 1
+                & --
+                & \citet[p.~77]{Cuffey.Paterson.2010} \\
+
+        $T_{\mathrm{c}}$   & Flow law critical temperature
+                & 263.15
+                & K
+                & \citet{Paterson.Budd.1982} \\
+
+        $f$     & Flow law water fraction coeff.
+                & 181.25
+                & --
+                & \citet{Lliboutry.Duval.1985} \\
+
+        $R$     & Ideal gas constant
+                & 8.31441
+                & \unit{J\,mol^{-1}\,K^{-1}}
+                & -- \\
+
+        $\beta$ & Clapeyron constant
+                & $7.9 \times 10^{-8}$
+                & \unit{K\,Pa^{-1}}
+                & \citet{Luthi.etal.2002} \\
+
+        $c_{\mathrm{i}}$   & Ice specific heat capacity
+                & 2009
+                & \unit{J\,kg^{-1}\,K^{-1}}
+                & \citet{Aschwanden.etal.2012} \\
+
+        $c_{\mathrm{w}}$   & Water specific heat capacity
+                & 4170
+                & \unit{J\,kg^{-1}\,K^{-1}}
+                & \citet{Aschwanden.etal.2012} \\
+
+        $k$     & Ice thermal conductivity
+                & 2.10
+                & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
+                & \citet{Aschwanden.etal.2012} \\
+
+        $L$     & Water latent heat of fusion
+                & $3.34\times10^5$
+                & \unit{J\,kg^{-1}\,K^{-1}}
+                & \citet{Aschwanden.etal.2012} \\
+
+        \middlehline
+        \multicolumn{2}{l}{{Basal sliding}} \\
+        \middlehline
+
+        $q$     & Pseudo-plastic sliding exponent
+                & 0.25
+                & --
+                & \citet{Aschwanden.etal.2013} \\
+
+        $v_{\text{th}}$& Pseudo-plastic threshold velocity
+                & 100
+                & \unit{m\,a^{-1}}
+                & \citet{Aschwanden.etal.2013} \\
+
+        $c_0$   & Till cohesion
+                & 0
+                & Pa
+                & \citet{Tulaczyk.etal.2000} \\
+
+        $e_0$   & Till reference void ratio
+                & 0.69
+                & --
+                & \citet{Tulaczyk.etal.2000} \\
+
+        $C_{\mathrm{c}}$   & Till compressibility coefficient
+                & 0.12
+                & --
+                & \citet{Tulaczyk.etal.2000} \\
+
+        $\delta$& Minimum effective pressure ratio
+                & 0.02
+                & --
+                & \citet{Bueler.Pelt.2015} \\
+
+        $\phi$  & Till friction angle
+                & 30
+                & \degree
+                & \citet[p.~268]{Cuffey.Paterson.2010} \\
+
+        $W_{\text{max}}$ & Maximum till water thickness
+                & 2
+                & m
+                & \citet{Bueler.Pelt.2015} \\
+
+        \middlehline
+        \multicolumn{2}{l}{{Bedrock and lithosphere}} \\
+        \middlehline
+
+        $\rho_{\mathrm{b}}$& Bedrock density
+                & 3300
+                & \unit{kg\,m^{-3}}
+                & -- \\
+
+        $c_{\mathrm{b}}$   & Bedrock specific heat capacity
+                & 1000
+                & \unit{J\,kg^{-1}\,K^{-1}}
+                & -- \\
+
+        $k_{\mathrm{b}}$   & Bedrock thermal conductivity
+                & 3
+                & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
+                & -- \\
+
+        $\nu_{\mathrm{m}}$ & Astenosphere viscosity
+                & $2.2 \times 10^{20}$
+                & \unit{Pa\,s}
+                & \citet{Mey.etal.2016} \\
+
+        $\rho_{\mathrm{m}}$& Astenosphere density
+                & 3300
+                & \unit{kg\,m^{-3}}
+                & \citet{Mey.etal.2016} \\
+
+        $D$     & Lithosphere flexural rigidity
+                & $1.389 \times 10^{24}$
+                & \unit{N\,m}
+                & \citet{Mey.etal.2016} \\
+
+        \middlehline
+        \multicolumn{2}{l}{{Surface and atmosphere}} \\
+        \middlehline
+
+        $T_{\mathrm{s}}$   & Temperature of snow precipitation
+                & 273.15
+                & K
+                & -- \\
+
+        $T_{\mathrm{r}}$   & Temperature of rain precipitation
+                & 275.15
+                & K
+                & -- \\
+
+        $F_{\mathrm{s}}$   & Degree-day factor for snow
+                & $3.297 \times 10^{-3}$
+                & \unit{m\,K^{-1}\,day^{-1}}
+                & \citet{Huybrechts.1998} \\
+
+        $F_{\mathrm{i}}$   & Degree-day factor for ice
+                & $8.791 \times 10^{-3}$
+                & \unit{m\,K^{-1}\,day^{-1}}
+                & \citet{Huybrechts.1998} \\
+
+        $R$     & Refreezing fraction
+                & 0.0
+                & --
+                & -- \\
+
+        $\gamma$& Air temperature lapse rate
+                & $6 \times 10^{-3}$
+                & \unit{K\,m{-1}}
+                & -- \\
+
+        $\psi$  & Precipitation factor
+                & 0.0704
+                & --
+                & \citet{Huybrechts.2002} \\
+
+        \bottomhline
+      \end{tabular}}
+    \end{table*}
+
 % ----------------------------------------------------------------------
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{table*}[t]
+      \caption{%
+        Palaeo-temperature proxy records and scaling factors yielding
+        temperature offset time-series used to force the ice sheet model
+        through the last glacial cycle (Fig.~\ref{fig:timeseries}). $f$
+        corresponds to the scaling factor adopted to yield Last Glacial Maximum
+        ice limits in the vicinity of mapped end moraines
+        (Fig.~\ref{fig:footprints}a), and $[{\Delta}T_{\textrm{TS}}]_{32}^{22}$
+        refers to the resulting mean temperature anomaly during the period 32
+        to~22\,ka after scaling.}
+      \label{tab:records}
+      {\begin{tabular}{lccccccl}
+        \tophline
+
+        Forcing   & Latitude & Longitude & Elev. (m~a.s.l.)
+                  & Proxy & $f$ & $[{\Delta}\text{TS}]_{32}^{22}$ (K)
+                  & Reference\\
+
+        \middlehline
+
+        GRIP      & \multirow{2}{*}{ 72{\degree}35$^{\prime}$\,N}   % 72.58 (decimal)
+                  & \multirow{2}{*}{ 37{\degree}38$^{\prime}$\,W}   % 37.64 (decimal)
+                  & \multirow{2}{*}{3238}
+                  & \multirow{2}{*}{\chem{\delta^{18}O}}
+                  & 0.50 & $-$8.2  % -16.4126 (before scaling)
+                  & \multirow{2}{*}{\citet{Dansgaard.etal.1993}} \\
+
+        GRIP, PP &&&&& 0.63 & $-$10.4 \\
+
+        EPICA     & \multirow{2}{*}{ 75{\degree}06$^{\prime}$\,S}   % 75.1
+                  & \multirow{2}{*}{123{\degree}21$^{\prime}$\,E}   % 123.35
+                  & \multirow{2}{*}{3233}
+                  & \multirow{2}{*}{\chem{\delta^{18}O}}
+                  & 1.05 & $-$9.7  % -9.2055
+                  & \multirow{2}{*}{\citet{Jouzel.etal.2007}} \\
+
+        EPICA, PP &&&&& 1.33 & $-$12.2 \\
+
+        MD01-2444 & \multirow{2}{*}{ 37{\degree}34$^{\prime}$\,N}   % 37.561
+                  & \multirow{2}{*}{ 10{\degree}04$^{\prime}$\,W}   % -10.142
+                  & \multirow{2}{*}{$-$2637}
+                  & \multirow{2}{*}{\chem{U^{K'}_{37}}}
+                  & 1.84 & $-$8.0  % -4.345625
+                  & \multirow{2}{*}{\citet{Martrat.etal.2007}} \\
+
+        MD01-2444, PP &&&&& 2.44 & $-$10.6 \\
+
+        \bottomhline
+      \end{tabular}}
+    \end{table*}
 
@@
 
+    \begin{figure*}[t]
+      \centerline{\includegraphics{alpcyc_lr_timeseries}}
+      \caption{%
+        \textbf{(a)} Temperature offset time-series from ice core and ocean
+        records (Table~\ref{tab:records}) used as palaeo-climate forcing for
+        the ice sheet model.
+        \textbf{(b)} Modelled total ice volume through the last 120~thousand
+        years (ka) expressed in meters of sea-level equivalent (m~s.l.e.). Gray
+        fields indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2
+        and MIS~4 according to a~global compilation of benthic
+        \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}. The simulation
+        driven by the EPICA temperature yields smaller ice volume variability
+        and a more realistic timing of deglaciation.}
+      \label{fig:timeseries}
+    \end{figure*}
+
     For the three palaeo-temperature records and the two palaeo-precipitation
@@
 
+    \begin{figure*}[t]
+      \centerline{\includegraphics{alpcyc_lr_footprints}}
+      \caption{%
+        \textbf{(a--c)} Cumulative extent of modelled ice cover during MIS~2
+        (29--14\,ka), without (dark colours) and with (light colours)
+        palaeo-precipitation corrections, and using temperature time-series
+        scaling factors (Table~\ref{tab:records}) adjusted to model the area
+        glaciated by the Rhine glacier piedmont lobe (black rectangle) in
+        agreement with the Last Glacial Maximum (LGM) geomorphological
+        reconstruction \citep[solid red line,][]{Ehlers.etal.2011}.
+        \textbf{(d--f)} Cumulative extent of modelled ice cover during MIS~4
+        (71--57\,ka). Only the simulation driven by the EPICA temperature
+        time-series yields reasonable MIS~4 ice cover.}
+      \label{fig:footprints}
+    \end{figure*}
+
     During MIS~2, all six simulations yield comparative cumulative glaciated
@@
     \citep{Wu.etal.2007}. Finally, the LGM extent of Alpine glaciers might have
-    been affected by east-west gradient in variables not accounted for
+    been affected by an east-west gradient in variables not accounted for
     by the PDD model, such as cloud cover or dust deposition.
@@
 
+    \begin{figure*}[t]
+      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
+      \caption{%
+        \textbf{(a)} Modelled bedrock topography (grey), ice surface topography
+        (200\,m contours), and ice surface velocity (blue) in the Alps
+        24.57\,ka before present, corresponding to the maximum modelled ice
+        cover. Modelled self-sustained ice domes (triangles) and some major
+        transfluences (crosses) are shown. Modelled LGM
+        ice extent (dashed orange line) and geomorphological
+        reconstruction \citep[solid red line,][]{Ehlers.etal.2011}. The
+        background map consists of depressed SRTM \citep{Jarvis.etal.2008}
+        topography and Natural Earth Data \citep{Patterson.Kelso.2017}.
+        \textbf{(b)} Temperature offset time-series from the EPICA ice core
+        used as palaeo-climate forcing for the ice flow model \citep[black
+        curve,][]{Jouzel.etal.2007}, and modelled total ice volume through the
+        last glacial cycle (120--0\,ka), expressed in meters of sea level
+        equivalent (m~s.l.e., blue curve). Gray fields indicate Marine
+        Oxygen Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
+        a~global compilation of benthic \chem{\delta^{18}O} records
+        \citep{Lisiecki.Raymo.2005}.}
+      \label{fig:lgmvel}
+    \end{figure*}
+
     The LGM extent of Alpine glaciers has been mapped with varying level of
@@
     Simplon Pass \citep{Kelly.etal.2004} and from the distribution of glacial
-    erratics across Brünig Pass \citep{Jackli.1962}.
+    erratics across Brünig Pass \citep{Jackli.1962}. On the other hand, the
+    simplified climate forcing used in this sumulation could not reproduced
+    the transport of glacial erratics from southern Valais to observed
+    deposition sites in the Solothurn region \citep[cf.~][]{Jouvet.etal.2017a}.
 
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{figure*}[t]
+      \centerline{\includegraphics{alpcyc_hr_timing}}
+      \caption{%
+        \textbf{(a)} Timing of the LGM given by the modelled age of maximum ice
+        thickness throughout the entire simulation (colour mapping) and
+        corresponding, ice surface elevation (200\,m contours).
+        \textbf{(b)} Temperature offset time-series from the EPICA ice core
+        used as palaeo-climate forcing for the ice flow model (black curve),
+        and modelled glacierized area during the LGM (coloured curve). The LGM
+        is here modelled as a time-transgressive event.}
+      \label{fig:timing}
+    \end{figure*}
 
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{figure}[t]
+      \centerline{\includegraphics{alpcyc_hr_trimlines}}
+      \caption{%
+        \textbf{(a)} Comparison of modelled ice surface elevation at the LGM
+        (time-transgressive, corresponding to maximum ice thickness,
+        Fig.~\ref{fig:timing}), compensated for bedrock deformation, against
+        observed trimline elevations for the upper Rhone Glacier
+        \citep[Table~1]{Kelly.etal.2004}. Colours
+        show the age of maximum ice thickness (as in Fig.~\ref{fig:timing}).
+        Model variables were bilinearly interpolated to the trimline locations.
+        \textbf{(b)} Histogram of differences between modelled LGM ice surface
+        elevation and trimline elevations (50\,m bands). The average
+        difference is 861\,m.
+        \textbf{(c)} Observed upper Rhone Valley trimline locations
+        \citep[Table~1]{Kelly.etal.2004}, modelled age of maximum ice thickness
+        (colour) and LGM ice surface elevation (200\,m contours). Hatches
+        mark areas modelled to have been glaciated but to have experienced
+        temperate ice for less than 1\,ka during MIS~2 (29--14\,ka).}
+      \label{fig:trimlines}
+    \end{figure}
 
@@
     difference of 861\,m (Fig.~\ref{fig:trimlines}b), wich is significant in
-    respect to the modelled regional ice thickness. This result is
+    respect to the modelled regional ice thickness. The result is
     somewhat dependent on uncertain basal sliding and ice rheological parameters
@@
     with regard to ice rheological parameters and 14\,\unit{\%} regarding basal
-    sliding parameters \citep[Fig.~7]{Seguinot.etal.2016}.
+    sliding parameters \citep[Fig.~7]{Seguinot.etal.2016}. In addition,
+    regional sensitivity tests in the Rhone Valley show that a modelled ice
+    surface compatible with the trimline elevations is incompatible with the
+    documented overspilling flow to the south across the Simplon pass
+    \citep{Becker.etal.2017}.
 
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{figure*}[t]
+      \centerline{\includegraphics{alpcyc_hr_profiles}}
+      \caption{%
+        \textbf{(a, c, e, g, i, k)} Profile lines roughly following valley
+        centerlines for the Rhine, Rhone, Dora Baltea and Isère, Inn and
+        Tagliamento Glaciers.
+        \textbf{(b, d, f, h, j, l)} Evolution of modelled glacier extent in
+        time, bilinearly interpolated along the corresponding profiles, showing
+        numerous cycles of advance and retreat over the last glacial cycle
+        modulated by subglacial topography and catchment geometry.
+        Isolated patches indicate periodic surges from tributary glaciers.}
+      \label{fig:profiles}
+    \end{figure*}
 
@@
 
-% ----------------------------------------------------------------------
-%TC: break Figures
-% Figures
-\clearpage
-% ----------------------------------------------------------------------
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpcyc_hr_inputs}}
-      \caption{%
-        \textbf{(a)} Geothermal heat flow from applying the similarity method
-        to multiple geophysical proxies \citep{Goutorbe.etal.2011} used as a
-        boundary condition to the bedrock thermal model 3\,km below the
-        ice-bedrock interface.
-        \textbf{(b)} Initial basal topography from SRTM
-        \citep{Jarvis.etal.2008} and geomorphological reconstruction of Last
-        Glacial Maximum Alpine glacier extent \citep[solid red
-        line,][]{Ehlers.etal.2011}.
-        \textbf{(c)} Extract from the estimated of present-day ice thickness
-        \citep{Huss.Farinotti.2012} substracted from the SRTM topography.
-        \textbf{(d)} Modern January and \textbf{(e)} July standard deviation
-        \citep{Seguinot.2013} of daily mean temperature from the ERA-Interim
-        \citep[1979--2012;][]{Dee.etal.2011} from the reference monthly
-        climatology used to force the surface mass balance (PDD) component of
-        the ice sheet model.
-        \textbf{(f)} Modern January and \textbf{(g)} July mean near-surface air
-        temperature, and \textbf{(h, i)} precipitation from WorldClim
-        \citep[1960--1990;][]{Hijmans.etal.2005}.}
-      \label{fig:inputs}
-    \end{figure*}
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpcyc_lr_timeseries}}
-      \caption{%
-        \textbf{(a)} Temperature offset time-series from ice core and ocean
-        records (Table~\ref{tab:records}) used as palaeo-climate forcing for
-        the ice sheet model.
-        \textbf{(b)} Modelled total ice volume through the last 120~thousand
-        years (ka) expressed in meters of sea-level equivalent (m~s.l.e.). Gray
-        fields indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2
-        and MIS~4 according to a~global compilation of benthic
-        \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}. The simulation
-        driven by the EPICA temperature yields smaller ice volume variability
-        and a more realistic timing of deglaciation.}
-      \label{fig:timeseries}
-    \end{figure*}
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpcyc_lr_footprints}}
-      \caption{%
-        \textbf{(a--c)} Cumulative extent of modelled ice cover during MIS~2
-        (29--14\,ka), without (dark colours) and with (light colours)
-        palaeo-precipitation corrections, and using temperature time-series
-        scaling factors (Table~\ref{tab:records}) adjusted to model the area
-        glaciated by the Rhine glacier piedmont lobe (black rectangle) in
-        agreement with the Last Glacial Maximum (LGM) geomorphological
-        reconstruction \citep[solid red line,][]{Ehlers.etal.2011}.
-        \textbf{(d--f)} Cumulative extent of modelled ice cover during MIS~4
-        (71--57\,ka). Only the simulation driven by the EPICA temperature
-        time-series yields reasonable MIS~4 ice cover.}
-      \label{fig:footprints}
-    \end{figure*}
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpcyc_hr_lgmvel}}
-      \caption{%
-        \textbf{(a)} Modelled bedrock topography (grey), ice surface topography
-        (200\,m contours), and ice surface velocity (blue) in the Alps
-        24.57\,ka before present, corresponding to the maximum modelled ice
-        cover. Modelled self-sustained ice domes (triangles) and some major
-        transfluences (crosses) are shown. Modelled LGM
-        ice extent (dashed orange line) and geomorphological
-        reconstruction \citep[solid red line,][]{Ehlers.etal.2011}. The
-        background map consists of depressed SRTM \citep{Jarvis.etal.2008}
-        topography and Natural Earth Data \citep{Patterson.Kelso.2017}.
-        \textbf{(b)} Temperature offset time-series from the EPICA ice core
-        used as palaeo-climate forcing for the ice flow model \citep[black
-        curve,][]{Jouzel.etal.2007}, and modelled total ice volume through the
-        last glacial cycle (120--0\,ka), expressed in meters of sea level
-        equivalent (m~s.l.e., blue curve). Gray fields indicate Marine
-        Oxygen Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
-        a~global compilation of benthic \chem{\delta^{18}O} records
-        \citep{Lisiecki.Raymo.2005}.}
-      \label{fig:lgmvel}
-    \end{figure*}
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpcyc_hr_timing}}
-      \caption{%
-        \textbf{(a)} Timing of the LGM given by the modelled age of maximum ice
-        thickness throughout the entire simulation (colour mapping) and
-        corresponding, ice surface elevation (200\,m contours).
-        \textbf{(b)} Temperature offset time-series from the EPICA ice core
-        used as palaeo-climate forcing for the ice flow model (black curve),
-        and modelled glacierized area during the LGM (coloured curve). The LGM
-        is here modelled as a time-transgressive event.}
-      \label{fig:timing}
-    \end{figure*}
-
-    \begin{figure}
-      \centerline{\includegraphics{alpcyc_hr_trimlines}}
-      \caption{%
-        \textbf{(a)} Comparison of modelled ice surface elevation at the LGM
-        (time-transgressive, corresponding to maximum ice thickness,
-        Fig.~\ref{fig:timing}), compensated for bedrock deformation, against
-        observed trimline elevations for the upper Rhone Glacier
-        \citep[Table~1]{Kelly.etal.2004}. Colours
-        show the age of maximum ice thickness (as in Fig.~\ref{fig:timing}).
-        Model variables were bilinearly interpolated to the trimline locations.
-        \textbf{(b)} Histogram of differences between modelled LGM ice surface
-        elevation and trimline elevations (500\,m bands). The average
-        difference is 861\,m.
-        \textbf{(c)} Observed upper Rhone Valley trimline locations
-        \citep[Table~1]{Kelly.etal.2004}, modelled age of maximum ice thickness
-        (colour) and LGM ice surface elevation (200\,m contours). Hatches
-        mark areas modelled to have been glaciated but to have experienced
-        temperate ice for less than 1\,ka during MIS~2 (29--14\,ka).}
-      \label{fig:trimlines}
-    \end{figure}
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpcyc_hr_profiles}}
-      \caption{%
-        \textbf{(a, c, e, g, i, k)} Profile lines roughly following valley
-        centerlines for the Rhine, Rhone, Dora Baltea and Isère, Inn and
-        Tagliamento Glaciers.
-        \textbf{(b, d, f, h, j, l)} Evolution of modelled glacier extent in
-        time, bilinearly interpolated along the corresponding profiles, showing
-        numerous cycles of advance and retreat over the last glacial cycle
-        modulated by subglacial topography and catchment geometry.
-        Isolated patches indicate periodic surges from tributary glaciers.}
-      \label{fig:profiles}
-    \end{figure*}
-
-% ----------------------------------------------------------------------
-% Tables
-\clearpage
-% ----------------------------------------------------------------------
-
-    \begin{table*}
-      \caption{%
-        Parameter values used in the ice sheet model. Symbols refer to
-        equations used by \citet{Seguinot.2014, Seguinot.etal.2016}.}
-      \label{tab:params}
-      \scalebox{0.75}
-      {\begin{tabular}{llrll}
-        \tophline
-
-        Not.    & Name & Value & Unit & Source \\
-
-        \middlehline
-        \multicolumn{2}{l}{{Ice rheology}} \\
-        \middlehline
-
-        $\rho$  & Ice density
-                & 910
-                & \unit{kg\,m^{-3}}
-                & \citet{Aschwanden.etal.2012} \\
-
-        $g$     & Standard gravity
-                & 9.81
-                & \unit{m\,s^{-2}}
-                & \citet{Aschwanden.etal.2012} \\
-
-        $n$     & Glen exponent
-                & 3
-                & --
-                & \citet[p.~55--57]{Cuffey.Paterson.2010} \\
-
-        $A_{\mathrm{c}}$   & Ice hardness coefficient cold
-                & $2.847 \times 10^{-13}$
-                & \unit{Pa^{-3}\,s^{-1}}
-                & \citet[p.~72]{Cuffey.Paterson.2010} \\
-
-        $A_{\mathrm{w}}$   & Ice hardness coefficient warm
-                & $2.356 \times 10^{-2}$
-                & \unit{Pa^{-3}\,s^{-1}}
-                & \citet[p.~72]{Cuffey.Paterson.2010} \\
-
-        $Q_{\mathrm{c}}$   & Flow law activation energy cold
-                & $6.0 \times 10^4$
-                & \unit{J\,mol^{-1}}
-                & \citet[p.~72]{Cuffey.Paterson.2010} \\
-
-        $Q_{\mathrm{w}}$   & Flow law activation energy warm
-                & $11.5 \times 10^4$
-                & \unit{J\,mol^{-1}}
-                & \citet[p.~72]{Cuffey.Paterson.2010} \\
-
-        $E_{\text{SIA}}$   & SIA enhancement factor
-                & 2
-                & --
-                & \citet[p.~77]{Cuffey.Paterson.2010} \\
-
-        $E_{\text{SSA}}$   & SSA enhancement factor
-                & 1
-                & --
-                & \citet[p.~77]{Cuffey.Paterson.2010} \\
-
-        $T_{\mathrm{c}}$   & Flow law critical temperature
-                & 263.15
-                & K
-                & \citet{Paterson.Budd.1982} \\
-
-        $f$     & Flow law water fraction coeff.
-                & 181.25
-                & --
-                & \citet{Lliboutry.Duval.1985} \\
-
-        $R$     & Ideal gas constant
-                & 8.31441
-                & \unit{J\,mol^{-1}\,K^{-1}}
-                & -- \\
-
-        $\beta$ & Clapeyron constant
-                & $7.9 \times 10^{-8}$
-                & \unit{K\,Pa^{-1}}
-                & \citet{Luthi.etal.2002} \\
-
-        $c_{\mathrm{i}}$   & Ice specific heat capacity
-                & 2009
-                & \unit{J\,kg^{-1}\,K^{-1}}
-                & \citet{Aschwanden.etal.2012} \\
-
-        $c_{\mathrm{w}}$   & Water specific heat capacity
-                & 4170
-                & \unit{J\,kg^{-1}\,K^{-1}}
-                & \citet{Aschwanden.etal.2012} \\
-
-        $k$     & Ice thermal conductivity
-                & 2.10
-                & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
-                & \citet{Aschwanden.etal.2012} \\
-
-        $L$     & Water latent heat of fusion
-                & $3.34\times10^5$
-                & \unit{J\,kg^{-1}\,K^{-1}}
-                & \citet{Aschwanden.etal.2012} \\
-
-        \middlehline
-        \multicolumn{2}{l}{{Basal sliding}} \\
-        \middlehline
-
-        $q$     & Pseudo-plastic sliding exponent
-                & 0.25
-                & --
-                & \citet{Aschwanden.etal.2013} \\
-
-        $v_{\text{th}}$& Pseudo-plastic threshold velocity
-                & 100
-                & \unit{m\,a^{-1}}
-                & \citet{Aschwanden.etal.2013} \\
-
-        $c_0$   & Till cohesion
-                & 0
-                & Pa
-                & \citet{Tulaczyk.etal.2000} \\
-
-        $e_0$   & Till reference void ratio
-                & 0.69
-                & --
-                & \citet{Tulaczyk.etal.2000} \\
-
-        $C_{\mathrm{c}}$   & Till compressibility coefficient
-                & 0.12
-                & --
-                & \citet{Tulaczyk.etal.2000} \\
-
-        $\delta$& Minimum effective pressure ratio
-                & 0.02
-                & --
-                & \citet{Bueler.Pelt.2015} \\
-
-        $\phi$  & Till friction angle
-                & 30
-                & \degree
-                & \citet[p.~268]{Cuffey.Paterson.2010} \\
-
-        $W_{\text{max}}$ & Maximum till water thickness
-                & 2
-                & m
-                & \citet{Bueler.Pelt.2015} \\
-
-        \middlehline
-        \multicolumn{2}{l}{{Bedrock and lithosphere}} \\
-        \middlehline
-
-        $\rho_{\mathrm{b}}$& Bedrock density
-                & 3300
-                & \unit{kg\,m^{-3}}
-                & -- \\
-
-        $c_{\mathrm{b}}$   & Bedrock specific heat capacity
-                & 1000
-                & \unit{J\,kg^{-1}\,K^{-1}}
-                & -- \\
-
-        $k_{\mathrm{b}}$   & Bedrock thermal conductivity
-                & 3
-                & \unit{J\,m^{-1}\,K^{-1}\,s^{-1}}
-                & -- \\
-
-        $\nu_{\mathrm{m}}$ & Astenosphere viscosity
-                & $2.2 \times 10^{20}$
-                & \unit{Pa\,s}
-                & \citet{Mey.etal.2016} \\
-
-        $\rho_{\mathrm{m}}$& Astenosphere density
-                & 3300
-                & \unit{kg\,m^{-3}}
-                & \citet{Mey.etal.2016} \\
-
-        $D$     & Lithosphere flexural rigidity
-                & $1.389 \times 10^{24}$
-                & \unit{N\,m}
-                & \citet{Mey.etal.2016} \\
-
-        \middlehline
-        \multicolumn{2}{l}{{Surface and atmosphere}} \\
-        \middlehline
-
-        $T_{\mathrm{s}}$   & Temperature of snow precipitation
-                & 273.15
-                & K
-                & -- \\
-
-        $T_{\mathrm{r}}$   & Temperature of rain precipitation
-                & 275.15
-                & K
-                & -- \\
-
-        $F_{\mathrm{s}}$   & Degree-day factor for snow
-                & $3.297 \times 10^{-3}$
-                & \unit{m\,K^{-1}\,day^{-1}}
-                & \citet{Huybrechts.1998} \\
-
-        $F_{\mathrm{i}}$   & Degree-day factor for ice
-                & $8.791 \times 10^{-3}$
-                & \unit{m\,K^{-1}\,day^{-1}}
-                & \citet{Huybrechts.1998} \\
-
-        $R$     & Refreezing fraction
-                & 0.0
-                & --
-                & -- \\
-
-        $\gamma$& Air temperature lapse rate
-                & $6 \times 10^{-3}$
-                & \unit{K\,m{-1}}
-                & -- \\
-
-        $\psi$  & Precipitation factor
-                & 0.0704
-                & --
-                & \citet{Huybrechts.2002} \\
-
-        \bottomhline
-      \end{tabular}}
-    \end{table*}
-
-    \begin{table*}
-      \caption{%
-        Palaeo-temperature proxy records and scaling factors yielding
-        temperature offset time-series used to force the ice sheet model
-        through the last glacial cycle (Fig.~\ref{fig:timeseries}). $f$
-        corresponds to the scaling factor adopted to yield Last Glacial Maximum
-        ice limits in the vicinity of mapped end moraines
-        (Fig.~\ref{fig:footprints}a), and $[{\Delta}T_{\textrm{TS}}]_{32}^{22}$
-        refers to the resulting mean temperature anomaly during the period 32
-        to~22\,ka after scaling.}
-      \label{tab:records}
-      {\begin{tabular}{lccccccl}
-        \tophline
-
-        Forcing   & Latitude & Longitude & Elev. (m~a.s.l.)
-                  & Proxy & $f$ & $[{\Delta}\text{TS}]_{32}^{22}$ (K)
-                  & Reference\\
-
-        \middlehline
-
-        GRIP      & \multirow{2}{*}{ 72{\degree}35$^{\prime}$\,N}   % 72.58 (decimal)
-                  & \multirow{2}{*}{ 37{\degree}38$^{\prime}$\,W}   % 37.64 (decimal)
-                  & \multirow{2}{*}{3238}
-                  & \multirow{2}{*}{\chem{\delta^{18}O}}
-                  & 0.50 & $-$8.2  % -16.4126 (before scaling)
-                  & \multirow{2}{*}{\citet{Dansgaard.etal.1993}} \\
-
-        GRIP, PP &&&&& 0.63 & $-$10.4 \\
-
-        EPICA     & \multirow{2}{*}{ 75{\degree}06$^{\prime}$\,S}   % 75.1
-                  & \multirow{2}{*}{123{\degree}21$^{\prime}$\,E}   % 123.35
-                  & \multirow{2}{*}{3233}
-                  & \multirow{2}{*}{\chem{\delta^{18}O}}
-                  & 1.05 & $-$9.7  % -9.2055
-                  & \multirow{2}{*}{\citet{Jouzel.etal.2007}} \\
-
-        EPICA, PP &&&&& 1.33 & $-$12.2 \\
-
-        MD01-2444 & \multirow{2}{*}{ 37{\degree}34$^{\prime}$\,N}   % 37.561
-                  & \multirow{2}{*}{ 10{\degree}04$^{\prime}$\,W}   % -10.142
-                  & \multirow{2}{*}{$-$2637}
-                  & \multirow{2}{*}{\chem{U^{K'}_{37}}}
-                  & 1.84 & $-$8.0  % -4.345625
-                  & \multirow{2}{*}{\citet{Martrat.etal.2007}} \\
-
-        MD01-2444, PP &&&&& 2.44 & $-$10.6 \\
-
-        \bottomhline
-      \end{tabular}}
-
-      %\noindent\small\makebox[\textwidth]
-      %{\begin{tabular}{llrlrlr}
-      %  \tophline
-      %
-      %  Record    & \multicolumn{2}{c}{GRIP}
-      %            & \multicolumn{2}{c}{EPICA}
-      %            & \multicolumn{2}{c}{MD01-2444} \\
-      %
-      %  Precip.   & cst. & PP
-      %            & cst. & PP
-      %            & cst. & PP \\
-      %
-      %  \middlehline
-      %
-      %  Latitude  & \multicolumn{2}{c}{ 72{\degree}35$^{\prime}$\,N}   % 72.58 (decimal)
-      %            & \multicolumn{2}{c}{ 75{\degree}06$^{\prime}$\,S}   % 75.1
-      %            & \multicolumn{2}{c}{ 37{\degree}34$^{\prime}$\,N} \\  % 37.561
-      %
-      %  Longitude & \multicolumn{2}{c}{ 37{\degree}38$^{\prime}$\,W}   % 37.64 (decimal)
-      %            & \multicolumn{2}{c}{123{\degree}21$^{\prime}$\,E}   % 123.35
-      %            & \multicolumn{2}{c}{ 10{\degree}04$^{\prime}$\,W} \\  % -10.142
-      %
-      %  Elevation & \multicolumn{2}{c}{3238\,m~a.s.l.}
-      %            & \multicolumn{2}{c}{3233\,m~a.s.l.}
-      %            & \multicolumn{2}{c}{$-$2637\,m~a.s.l.} \\
-      %
-      %  Proxy     & \multicolumn{2}{c}{\chem{\delta^{18}O}}
-      %            & \multicolumn{2}{c}{\chem{\delta^{18}O}}
-      %            & \multicolumn{2}{c}{\chem{U^{K'}_{37}}} \\
-      %
-      %  $f$       & 0.50 & 0.63
-      %            & 1.05 & 1.33
-      %            & 1.84 & 2.44 \\
-      %
-      %  $[{\Delta}\text{TS}]_{32}^{22}$
-      %            & $-$8.2\,K & $-$10.4\,K     % -16.4126 (before scaling)
-      %            & $-$9.7\,K & $-$12.2\,K     % -9.2055
-      %            & $-$8.0\,K & $-$10.6\,K \\  % -4.345625
-      %
-      %  Reference & \multicolumn{2}{c}{\citet{Dansgaard.etal.1993}}
-      %            & \multicolumn{2}{c}{\citet{Jouzel.etal.2007}}
-      %            & \multicolumn{2}{c}{\citet{Martrat.etal.2007}} \\
-      %
-      %  \bottomhline
-      %\end{tabular}}
-
-    \end{table*}
-
 % ======================================================================

### 2018-06-15 articles/cycle/response-monegato.tex
Init response to Giovanni Monegato.

% response-editor.tex
% ----------------------------------------------------------------------
\input{response-header}

\title{Authors' response to Giovanni Monegato}
\author{J.~Seguinot et al.}
%\date{}

\begin{document}
\maketitle
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

\newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
\newcommand{\referee}[1]{\bigskip\noindent\textcolor{darkblue}{#1}}
\newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}

Dear Giovanni Monegato,

Thank you very much for your public comment on our manuscript.

    \referee{%
        The manuscript is very interesting as approach on the Late Pleistocene
        Alpine glaciation, for which good data are available for the LGM
        onwards, but few is known about pre-MIS2. The present article does not
        solve the problem of what how the glaciers behave before the LGM, but
        casts new light on this topic suggesting interesting details and field
        to investigate. I would like to post some general comments on the
        manuscript especially regarding the Italian side of the Alps.}

    \referee{\textbf{1 --}
        Reading the manuscript and watching the supplemental file, it is
        remarkable how few systems in the model result to have the same
        extension compared to the geomorphological/geological evidence. For the
        Italian side roughly the Riparia, Baltea, Ossola/Ticino and Tagliamento
        looks right. While other systems are much over-(western Alps) or
        under-estimated (from Adda to Piave). The overestimation of the western
        Alps can be related to the scarcity of updated chronological data; for
        example the glacier reconstruction in Gesso Valley of the Maritime Alps
        (Federici et al. 2017) shows a much larger extent than Ehlers and
        Gibbard (2004) compilation. Anyhow the sentence of line 30 page 14 is
        reliable and suggests that more data are needed about. Also the Eastern
        Alpine glaciers resulted very overestimated, for these and for the
        not-matching glaciers in the Italian side I think that paleoclimate
        forcing and especially precipitation models are one of the keys
        factors. The WorldClim model that is considered and showed in Figure 1
        seems to be inconsistent respect to other models focused on the Alps. I
        suggest to take into consideration Isotta et al. (Int. J. Climatol.
        2013) where distribution of the precipitations shows areas of high
        precipitation rates (both as annual mean and daily peak). This
        distribution would point to high precipitation rates also in the Piave,
        Adda and Oglio catchments, while the Tagliamento and Ticino systems fit
        well in the model as well. The knot of the Valais-Ticino-upper Rhine
        has also modern high precipitations, and this is one of the key areas
        for large ice accumulation and for the southerly component during the
        LGM according to Luetscher et al. (2015). Actually, considering the
        modern precipitation rates and comparing to your model, the
        underestimation for the Adige system sounds reasonable but not in
        agreement with the chronology and geomorphology found in the Garda. So
        other causes, and not only precipitation, have to be considered.}

    \referee{\textbf{2 --}
        Concerning the ice-transfluence. I agree that it has to be much more
        considered. I wonder if the Adige had a great contribution from the
        Austrian Tauern (Toblach area) this would have increased the ice in
        Adige and deplete the Drava, or in Winschgau valley where the catchment
        upstream Mustar saddle is more related to Adige than the Inn. The same
        could have happened for the Piave catchment, which is very
        underestimated in the model. Again, ice flowing to the south did not
        flow to the east. But I think that this may be not enough for
        justifying the overestimation of the Eastern Alps. Concerning the
        Straninger saddle I am a bit skeptic because it is not the lowermost
        saddle C2(Plockenpass and Nassfeldpass are lower in elevation). The
        central Adda glacier (and not the Ticino as at page 14 line 26) is much
        underestimated and here the contribution from transfluence in St.
        Moritz area could have been remarkable. I attach a figure separately
        with transfluences (yellow stars and red arrows).}

    \referee{\textbf{3 --}
        The basal sliding of glaciers and its overall velocity is another
        interesting factor to discuss. For Adige for example we know that at
        about 28 cal BP the Adige glacier was damming a tributary valley north
        of Trento around the same age (Avanzini et al 2009), while the same
        glacier arrived at Garda at 24.6 cal BP (Monegato et al., 2017).  This
        means that the glacier front advanced of about 100 km in around 3.5 ka
        so 280 m/y. The same could have been for the Adda and Oglio glaciers,
        even if a such robust chronology is lacking. Is it possible that
        overestimated velocities produced large eastern glaciers?}

    \referee{\textbf{4 --}
        About the self-sustained ice domes, their importance is for me unclear
        and not well explained in the text. Why they are only two? Why the
        Adamello or the Tauern massifs, as an example, were not considered as
        self-sustained ice domes?}

    \referee{\textbf{5 --}
        Global circulation models (e.g., Löfverström et al., 2014, Beghin et
        al., 2015) suggest that the Polar front was at different latitudes
        during each cold phases of the Late Pleistocene. This could have effect
        of different impact on the Alps. For example during MIS4 if the
        westerlies were dominant this could have driven more effective
        precipitations in the western and northern Alps in respect to the
        southern and the eastern Alps. If this is true, and can be applied to
        the major cold phases. How was the behavior of glaciers during MIS 5
        and 3?}

\end{document}

### 2018-06-15 articles/cycle/response-referee1.tex
Init response to anonymouse referee no. 1.

% response-editor.tex
% ----------------------------------------------------------------------
\input{response-header}

\title{Authors' response to Anonymous Referee \#1}
\author{J.~Seguinot et al.}
%\date{}

\begin{document}
\maketitle
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

\newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
\newcommand{\referee}[1]{\bigskip\noindent\textcolor{darkblue}{#1}}
\newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}

    Dear Anonymous Referee \#1,

    Thank you very much for your detailed review of our manuscript.

    \referee{%
        This paper is a landmark advance in modelling European Alpine ice
        cover, applying a high-resolution (1\,km) ice model to the entire Alps
        through the last glacial cycle, for the first time to my knowledge.
        Results are compared with diverse geological data, and several
        important findings are presented, including time-transgressive ice
        marginal extents at LGM. The climate forcing is simple, applying
        uniform perturbations to modern observed datasets, which leads to some
        uncertainty in the results, but does not detract from them too much
        given the advances made in the ice modelling alone.}

    \referee{%
        The introduction gives an elegant summary of Alpine glacial science
        since the 1700's, including many historical references. The paper is
        well organized, with well-chosen sensitivities described first that
        calibrate the climate forcing, followed by detailed analysis of one
        best-fit high-resolution (1\,km) simulation through the last 120\,kyrs.
        Detailed comparisons to a variety of geological data are made,
        constituting a thorough assessment of model performance. An impressive
        animation of the whole cycle is included as supplementary material.}

% ----------------------------------------------------------------------

\sechead{Specific comments}

    \referee{\textbf{p.~4, l.~9--10:}
        Can the physical basis of englacial water fraction and sensitivity of
        results be summarized briefly? This is not a usual component in
        ice-sheet models. Is the cap value ("capped at 0.01") well constrained,
        and does it have a significant effect on results?}

    \referee{\textbf{p.~4, l.~19--20:}
        The sub-glacial hydrologic component should be described more (even if
        exactly as in Bueler and van Pelt, 2015). Is basal water transported
        horizontally down the hydropotential gradient? This is usually a highly
        uncertain component of ice-sheet models, but can have a large effect on
        results through its influence on basal sliding, and basal frozen vs.
        thawed areas, which is relevant to section 4.4 regarding trimlines.}

    \referee{\textbf{Somewhat related:}
        Little information is given on the choices of basal sliding parameter
        values in Table 1. This could be discussed briefly. Presumably no
        inversion or optimization was performed for these values beforehand,
        and they do vary spatially. Are they appropriate for Alpine bedrock
        overall?}

    \referee{\textbf{p.~7, l.~7:}
        Are there any data to support this atmospheric lapse rate value
        (6\,K\,km$^{-1}$), and do other values have the potential to
        significantly affect ice temperatures? In particular, could they change
        the basal areas of frozen/unfrozen ice and so the comparisons with
        trimlines in section 4.4?}

% ----------------------------------------------------------------------

\sechead{Climate forcing}

    \referee{%
        The method of spatially uniform shifts to modern climate forcing is
        common in paleo-modeling of large ice sheets, and in my opinion is
        acceptable as a starting point in this work, with coupling to regional
        climate models (RCMs) left to follow-on work. There are good
        discussions on possible shortcomings of this method, for instance as a
        cause of anomalous east-west marginal ice extents at LGM (pg. 11, line
        6-8). However, I suggest changing the sentence on pg. 9, line 5, which
        mentions some RCMs applied to LGM Europe, but also says "...over the
        Alps during the last glacial cycle, of which little is known apart from
        the LGM". There are several other RCM modeling studies over Europe
        during the last 120\,kyrs, e.g., for MIS Stage 3, Kjellstrom et al.,
        Boreas, 2010; Barron and Pollard, Quat. Res., 2002; Alfano et al.,
        Quat. Res., 2002; and for 6\,ka, Strandberg et al., Clim. Past, 2014.
        Perhaps there is little useful material there for the Alps, but such
        papers exist.}

    \referee{%
        The past climate variations are prescribed following 3 quite distal
        core records, and the most distal (EPICA) is chosen as yielding the
        best fit to Alpine glacial evidence. The basis for preferring EPICA
        seems reasonable (matching some higher-frequency amplitudes of ice
        variability, section 3.3). However, this agreement is in a sense
        coincidental, in that there is no direct meteorological link between
        Antarctic and Alpine regional climate variations. Are there any
        proximal proxy records of Alpine climate at all, perhaps lacustrian
        varves, that could be used to assess the EPICA-based shifts in air
        temperatures and precipitation, even over limited periods of the last
        120\,kyrs?}

    \referee{%
        A PDD scheme is used based only on seasonal air temperatures. van de
        Berg et al. (Nature Geosc., 2011) showed that for long-term variations
        including the Eemian, orbital changes in insolation are important and
        should be considered explicitly. This could be particularly relevant
        here, because the EPICA core does not reflect changes in insolation
        over the Alps. In further work, an insolation-change term (summer,
        local) could be combined with EPICA in the climate temperature
        paleo-forcing.}

% ----------------------------------------------------------------------

\sechead{Trimlines}

    \referee{%
        The point is well taken that trimlines do not necessarily indicate past
        ice surface elevations, but the upper limit of temperate ice with cold
        ice above (pg. 18, line 5 to pg. 19, line 4). It is an important point,
        because model LGM ice surfaces are far above most trimline elevations
        (as noted and in Fig.~6a,b). The pertinent results are shown in
        Fig.~6c, and I agree there is good support for the cold vs. temperate
        (basal) ice hypothesis.}

    \referee{%
        It might help general readers to spell out the interpretation even more
        in the text. That is, as I understand it, the observed trimlines should
        coincide with the boundaries between model areas of frozen vs.
        temperate beds, so the dots in Fig.~6c should all lie on the borders
        between the hatched and white areas. The sentence on p.~19, l.~20-21,
        is confusing in this regard. Incidentally, it would also help to add
        the word "basal" to the last sentence of the Fig.~6c caption:
        "...experienced temperate basal ice for ...".}

    \referee{%
        One reason for the remaining discrepancies in Fig.~6c could be temporal
        variations in the model boundaries, that are aggregated in time by the
        "< 1\,kyr" criterion for the hatching and the grouping of all trimline
        data. To go into this in more detail, in principle Fig.~6c could be
        expanded to show the model basal frozen-temperate boundaries at
        particular times (21.5, 22.5, etc, ka), with only the dots for each
        time period superimposed. But that may not be worth it unless there are
        large temporal variations in the model boundaries.}

    \referee{%
        A slight concern is that the majority of the trimline data seems to be
        orange dots i.e., older that 27\,ka in the timescale of Fig.~6c. The
        period for the model hatching extends back only to 29\,ka. Hopefully,
        most of the orange-dotted data are within that period and are not older
        than 29\,ka(?).}

    \referee{%
        The text could briefly mention (and hopefully rule out) the issue of
        very fine-scale topographic features on which the trimlines are
        located, not resolved by the 1-km model topography. If data sites are
        on small-scale highs or lows significantly different from their
        ∼km-scale surroundings, that could contribute to the discrepancies in
        Fig.~6c.}

% ----------------------------------------------------------------------

\sechead{Technical comments}

    \referee{\textbf{p.~3, l.~2:}
        Perhaps change "lead" to "led", "to which" to "to what".}

    \referee{\textbf{Fig.~1 caption, l.~4:}
        Perhaps change "estimated" to estimate", or "estimated of" to
        "estimated".}

\end{document}

### 2018-06-15 articles/cycle/response-referee2.tex
Init response to anonymouse referee no. 2.

% response-editor.tex
% ----------------------------------------------------------------------
\input{response-header}

\title{Authors' response to Anonymous Referee \#2}
\author{J.~Seguinot et al.}
%\date{}

\begin{document}
\maketitle
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

\newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
\newcommand{\referee}[1]{\bigskip\noindent\textcolor{darkblue}{#1}}
\newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}

Dear Anonymous Referee \#2,

Thank you very much for your detailed review of our manuscript.

% ----------------------------------------------------------------------

\sechead{General comments}

    \referee{%
        This manuscript describes a model study with the Parallel Ice Sheet
        Model applied to the last glacier cycle in the Alps. The climate
        forcing is derived from present day climate of WorldClim and the
        ERA-Interim reanalysis and time-dependent temperature offsets derived
        from the Greenland Ice core (GRIP), from Antarctic ice core (EPICA) and
        Marine sediment core from the Iberian margin (MD01-2444). The study is
        split in two, in the first part analysis of six model simulations (with
        and without precipitation scaling) made on a 2 km resolution grid is
        presented and concluded that out of the three climate forcing records
        used, the EPICA record gives the most realistic ice volume history
        during MIS 4 and 2. The second part analyses simulation made with the
        EPICA forcing record on a 1km grid. The authors draw conclusions about
        the ice cover, ice flow pattern, ice thickness, LGM ice extent, which
        in their model is a transient stage with varying timing across their
        model domain due to glacier dynamics. The manuscript is well organized
        and clearly written, the missing thing in this study is a discussion of
        and preferably a sensitivity study of the ice dynamic-model assumptions
        made. Is the sliding of the ice age ice sheet realistically modelled
        with pseudo-plastic assumption using the Shallow Shelf approximation?
        How sensitive are the results to the selected model parameters? It is
        briefly mentioned once on page 19, line 8, but a thorough analysis of
        the model sensitivity would strengthen the paper.}

% ----------------------------------------------------------------------

\sechead{Specific comments}

    \referee{%
        I find missing something that indicates that the times are before
        present, as in line 5, line 8 and line 16 on page 1 -- and elsewhere in
        the paper it is written (120-0 ka) should you add before present or BP
        to indicate the time interval?}

    \referee{%
        Could the reason for too large ice volumes when using the GRIP record,
        as mentioned in lines 13-15 on page 9 be due to Arctic Amplification?
        Could that have an effect then as it has now? This is also mentioned in
        line 11 on page 10.}

% ----------------------------------------------------------------------

\sechead{Minor comments}

    \referee{\textbf{p.~1, l.~24:}
        suggest: ``have extended well outside their current margins''.}

    \referee{\textbf{p.~2, l.~22:}
        could you add a reference and a timing for LGM?}

    \referee{\textbf{p.~3, l.~1:}
        suggest to replace ``thus'' with ``still''.}

    \referee{\textbf{p.~3, l.~3:}
        add s to responses.}

    \referee{\textbf{p.~3, l.~18:}
        something missing in the sentence, suggest ``formulation'' after
        ``creep''.}

    \referee{\textbf{p.~4, l.~4:}
        suggest to replace ``field'' with ``sheet''.}

    \referee{\textbf{p.~5, Fig.~1:}
        c) can you add a scale and maybe indication with a box in b) where this
        extract is from?  ``from the estimate'' (not plural in line 4 of
        caption), Line 6 (PDD) is not acronym for surface mass balance, some
        more explanation is needed here, indicate also, like in the other
        figures that h) is January and i) is July figuresPage 6 line 6 suggest
        to replace ``of'' with ``with''.}

    \referee{\textbf{p.~6, l.~14:}
        suggest ``The climate forcing driving the ice sheet simulations consist
        spatially of a present-day monthly mean climatology...``}

    \referee{\textbf{p.~6, l.~15:}
        suggest to delete ``s'' on corrections.}

    \referee{\textbf{p.~6, l.~19:}
        add ``mean'' after monthly.}

    \referee{\textbf{p.~6, l.~22:}
        note, if true (clarify in Figure caption, see comment above) then the
        reference should be to Fig. 1 i) for summer precipitation.}

    \referee{\textbf{p.~6, l.~27:}
        replace ``is'' with ``are''.}

    \referee{\textbf{p.~6, footnote:}
        add a reference for the correct formula for the rigidity and clarify
        (add something like, the consequence of this error is that the
        simulations effectively use....) Delete ``in'' before ``a small'' and
        add a quantification of the small change in the length scale – how
        small, is it a few percentage?}

    \referee{\textbf{p.~7, l.~2:}
        ``shipped with WorldClim'' is not clear, please edit, also suggest not
        to use b for topography, s surf and s bed , or h would be better.}

    \referee{\textbf{p.~7, l.~19:}
        add ``the'' before ``oxygen''.}

    \referee{\textbf{p.~7, l.~24:}
        ``and within a rectangular region ..'' is not clear, edit this text.}

    \referee{\textbf{p.~8:}
        add ``acceleration of'' before gravity, add a reference for the ideal
        gas constant?}

    \referee{\textbf{p.~9, l.~14--15:}
        see comment above, could this be an example of Arctic amplification?}

    \referee{\textbf{p.~9, l.~19:}
        add ``a'' before ``very'' and suggest to turn the sentence around, the
        EPICA simulations are in a good agreement with the data.}

    \referee{\textbf{p.~9, l.~20:}
        followed by first a retreat and then a standstill? replace ``blue''
        with ``red''.}

    \referee{\textbf{p.~9, l.~21:}
        suggest : The simulations forced by the GRIP palaeo-temperature forcing
        yield ... (blue curves)''.}

    \referee{\textbf{p.~10, l.~2:}
        suggest to add ``followed by a rapid retreat''.}

    \referee{\textbf{p.~10, l.~3:}
        why state ``two or three'' in intro the suggestion is either 4 major or
        15, is there a reference for 2 or 3?}

    \referee{\textbf{p.~10, l.~10:}
        suggest to replace ``lower'' with ``smaller''.}

    \referee{\textbf{p.~10, l.~11:}
        is this due to Arctic Amplification?}

    \referee{\textbf{p.~10, l.~12:}
        suggest to replace ``least'' with ``the smallest''.}

    \referee{\textbf{p.~11, Fig.~3 caption:}
        what does ``cumulative'' mean here? Do you mean maximum in each
        location?  clarify what the black line indicates. No solid red line is
        visible in figures. What is meant with ``reasonable'', maybe replace
        with ``realistic''? suggest to replace ``cover'' with ``extent''.}

    \referee{\textbf{p.~11, l.~2:}
        suggest to replace ``cumulative'' with ``maximum extent in each area''
        or something similar.}

    \referee{\textbf{p.~11, l.~1--3:}
        how sensitive is the model to different parameters in the applied
        sliding formulation?}

    \referee{\textbf{p.~11, l.~4--5:}
        ``outside this benchmark'' clarify if you mean spatially or temporally.}

    \referee{\textbf{p.~11, l.~6--11:}
        is this text better fitted in a discussion section?}

    \referee{\textbf{p.~12, l.~2:}
        suggest to replace ``higher'' with ``larger''.}

    \referee{\textbf{p.~12, l.~4:}
        do you mean to refer to Fig. 3?}

    \referee{\textbf{p.~12, l.~29--32:}
        does this text fit better in method section?}

    \referee{\textbf{p.~13, Fig.~4 caption:}
        the 200 m surface contours are not clearly visible in the figure, can
        they be made sharper, or just skipped? Line 3, something like ``are
        shown'' is missing. Suggest to replace``background'' with ``bedrock''.
        This is the first time ``Natural Earth Data'' is mentioned, should that
        be in the section on the data? Suggest to replace ``Gray fields'' with
        ``shaded gray area'' and ``boundaries'' with ``timing''.}

    \referee{\textbf{p.~15, l.~8:}
        replace ``was'' with ``is''.}

    \referee{\textbf{p.~15, l.~25:}
        ``occurred''.}

    \referee{\textbf{p.~16, Fig.~5:}
        in figure the color bar is written to indicated maximum surface
        elevation, but in figure caption the maximum ice thickness, which is
        correct? The surface elevation contours are not clearly visible. The
        dark orange color in the figure, that covers the central part of the
        ice sheet is not (?) in the bar on the left (or is it before 27 ka
        BP?).}

    \referee{\textbf{p.~17, l.~30:}
        suggest to replace ``have'' with ``could''.}

    \referee{\textbf{p.~18, Fig.~6:}
        the 200 m contours are not clearly visible.}

    \referee{\textbf{p.~19, l.~7:}
        could you add what the modelled regional ice thickness is, for
        comparison?}

    \referee{\textbf{p.~20, Fig.~7:}
        add info about the gray areas indicating MIS 4 and 2 ``Isolated patches
        indicate periodic surges from tributary glaciers'' needs more
        explanation and it is not clear what is meant. Does the model simulate
        periodic surges?}

    \referee{\textbf{p.~21, l.~21:}
        here is a reference for 2 or 3 glaciations (see comment above) but in
        the intro is only mentioned 4 or 15, suggest to change text to
        harmonize.}

    \referee{\textbf{p.~21, l.~27:}
        suggest to edit ``the study consists of'' or start for example like
        ``In this study the model has been applied...''}

    \referee{\textbf{p.~21, l.~28:}
        how important is it that the model has been validated for the
        Cordilleran ice sheet? Will that support the choices of the sliding
        model applied in the Alps? I think that it would be valuable for this
        study to do a sensitivity runs for at least some of the model parameter
        choices.}

    \referee{\textbf{p.~22, l.~6:}
        add ``records'' after forcing.}

    \referee{\textbf{p.~22, l.~19:}
        why do you add ``potentially'' here? Isn't this a firm conclusion from
        you study?}

    \referee{\textbf{p.~22, l.~20:}
        suggest to replace ``higher'' with ``larger''.}

    \referee{\textbf{p.~22, l.~22:}
        same as above, why ``potentially'' here?}

    \referee{\textbf{p.~22, l.~24:}
        suggest to replace ``nevertheless'' with ``however'' or edit the
        sentence.}

    \referee{\textbf{p.~22, l.~26--28:}
        the paper would be able to give stronger conclusions with sensitivity
        study, suggest to edit the sentence by replace ``statements'' with
        ``limitations'' or ``drawbacks'' and ``last glacier cycle ice dynamics
        in the Alps'' is not easy to read.}

    \referee{\textbf{p.~23, l.~2:}
        replace ``mode'' with ``more''.}

    \referee{%
        Figures are generally clear and well set up.  The surface contours in
        Figs 4,5 and 6 is not clearly visible in my printout and could maybe
        become clearer?}

\end{document}

### 2018-06-15 articles/cycle/response-referee3.tex
Init response to anonymouse referee no. 3.

% response-editor.tex
% ----------------------------------------------------------------------
\input{response-header}

\title{Authors' response to Anonymous Referee \#3}
\author{J.~Seguinot et al.}
%\date{}

\begin{document}
\maketitle
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

\newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
\newcommand{\referee}[1]{\bigskip\noindent\textcolor{darkblue}{#1}}
\newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}

Dear Anonymous Referee \#3,

Thank you very much for your detailed review of our manuscript.

    \referee{%
        This is an excellent manuscript and I highly recommend publication
        after a few minor changes. The authors present a well thought out
        modelling experiment which they combine (albeit in a qualitative
        manner) with extensive palaeo-glaciological data and cumulative work. I
        can see this work being extended into more extensive and rigorous work
        (RCM forcing, ice dynamic sensitivity, quantitative fitting to
        geomorphological record etc), but this is an important leap forward.}

% ----------------------------------------------------------------------

\sechead{Abstract}

    \referee{\textbf{p.1, l.~2:}
        ``pioneer'' should be ``pioneering'' P1, L16: I think the finding that
        you get asynchronous glaciation extents with a uniform climate offset
        is due to glacier hypsometry and setting should be stated here. i.e.
        that the timing of maximum glaciation and recession isn't purely a
        function of climate. This finding needs to be highlighted better in the
        abstract.}

% ----------------------------------------------------------------------

\sechead{Introduction}

    \referee{\textbf{p.2, l.~28:}
        Ballantyne and Stone (2015) should be added to this list.}

    \referee{\textbf{p.2, l.~34:}
        It should be stated that it could be a consequence of both
        glaioclimatic interactions and uncertainties in dating methods.}

    \referee{\textbf{p.3, l.~2:}
        These points serve the literature well to highlight gaps for future
        research. However, I would argue that you do not get very far here on 1
        and 5 and do not completely solve the other 3 points. Your text
        reflects these shortcomings very well, for which you should be
        applauded. Though I think at this stage of the manuscript, your
        statement of intent, you should state that you do not claim to solve
        these questions, but rather push forward on all of them using your new
        approach of ice sheet modelling.}

% ----------------------------------------------------------------------

\sechead{Section 2.6}

    \referee{\textbf{p.7:}
        The spatial distribution of your modern climate variables (precip,
        temp) will be massively influenced by elevation. Though there is a
        lapse rate, does this pattern of high precip and low temp over
        mountains remain throughout the simulation despite ice surface
        topography, and if so, how does this influence your results?}

% ----------------------------------------------------------------------

\sechead{Section 3}

    \referee{\textbf{p.7, l.~14:}
        You keep mentioning the number of processors. I find this infor- mation
        slightly irrelevant, and it will soon become outdated as processing
        speed and models increase (GPUs for example). The only way it will
        serve the community is if there is a full description of the computer
        set up. For example, it could be that the simulations took 4 days on
        144 processors, but the processors were slow. Suggest removing these
        references.}

    \referee{\textbf{p.10, l.~14:}
        This is a more general point. You eventually choose the EPICA record,
        and for justifiable reasons based on comparison to reconstructed ice
        extent and timing.  However, this is likely coincidence. EPICA is
        likely a complex record containing global and local antarctic
        influences upon climate. The ``real'' climate over the alps during
        glaciation is like decoupled from that of Antarctica to an extent.
        Therefore, the match you find is not an inference about climate, as
        different combinations of offsets may have made the same result
        (smoothed GRIP to remove some of the D-O scale noise?). You should make
        this explicit somewhere in the manuscript.}

    \referee{\textbf{p.11, l.~2:}
        First sentence needs reconsidering as it is slightly broken in its
        current form. Perhaps ``Figure 3a shows the cumulative extent of
        glaciated area during MIS2''.}

    \referee{\textbf{p.11, l.~10:}
        Is it possible some ice is missing from the geological reconstruction
        in some instances? I guess some outlets are well constrained, whilst
        others areas could be ``filled in'' by this modelling experiment.}

% ----------------------------------------------------------------------

\sechead{Section 4}

    \referee{\textbf{p.14, l.~15--31:}
        I find this description of sites and timing of glacier extent compared
        to dates difficult to follow. I suggest a new figure to convey this
        important comparison: Have the reconstructed and modelled ice extents
        at key times for each of the mentioned glaciers on several smaller
        maps, including geochronological constraints.}

    \referee{\textbf{p.17, l.~17:}
        This finding is important and should be highlighted better in abstract
        and conclusion.}

    \referee{\textbf{p.17, l.~22:}
        Is the model recreating possible surging? Or would this be an
        over-interpretation given the uncertainty in climate and physics. Seems
        to fit with the enthalpy model of Benn and others for surging. As
        reads, it suggests that these areas were possible palaeo-surges, please
        clarify.}

    \referee{\textbf{p.19, l.~9--12:}
        On trimlines: I think you justify well why you haven't yet modelled the
        sensitivity to your trimline result - previous work backs this up. But,
        I think you should consider the following: Adding a plus/minus to your
        results to reflect the uncertainty.  The importance of resolution -
        many trimlines will be below the resolution of your model, so perhaps
        aren't resolved enough for model-data comparison. They may have acted
        to deflect ice flow around mountain peaks for example. This resolution
        caveat should be mentioned. I would be surprised if all trimlines are
        subglacial transitions as this paper suggests - perhaps you need to
        directly challenge the geochronological community to find better
        trimline constraints (sub/supra) as a statement in this paper. Your
        mean value of 861 m is unrepresentative of your sample. Your sample is
        highly skewed, so a modal value (1050 m ish from Fig 6) is more
        appropriate. A similar finding with a similar approach has already been
        found for the British Isles, with the added constraint of GIA
        observations. I suggest referencing Kuchar et al. 2011 for this
        reason.}

    \referee{%
        A philosophical but important point is that your discussion throughout
        is written from the standpoint of the geochronological/geomorphological
        data and reconstructions as being ``truth''. It should consider
        somewhere that perhaps data is missing as it is hard-won in the places
        it exists and interpretations may be slightly wrong. The model is also
        not the ``truth'' and there is probably a blurred line inbetween upon
        which we can proceed.}

    \referee{\textbf{p.21, l.~11--19:}
        Be really clear here that these are modelled, and perhaps not
        geologically recorded, advances of the ice sheet. If there is no data,
        it might be correct, might be just a modelled result.}

    \referee{%
        Additional references: Ballantyne, C.K. and Stone, J.O., 2015.
        Trimlines, blockfields and the vertical extent of the last ice sheet in
        southern Ireland. Boreas, 44(2), pp.277- 287. Kuchar, J., Milne, G.,
        Hubbard, A., Patton, H., Bradley, S., Shennan, I. and Edwards, R.,
        2012. Evaluation of a numerical model of the British–Irish ice sheet
        using relative sea-level data: implications for the interpretation of
        trimline observations.  Journal of Quaternary Science, 27(6),
        pp.597-605.}

\end{document}

### 2018-06-18 articles/cycle/response-referee1.tex
Fix technical comments from referee 1.

@@
     \referee{\textbf{p.~3, l.~2:}
-        Perhaps change "lead" to "led", "to which" to "to what".}
+        Perhaps change ``lead'' to ``led'', ``to which'' to ``to what''.}
+
+        Done.
 
     \referee{\textbf{Fig.~1 caption, l.~4:}
-        Perhaps change "estimated" to estimate", or "estimated of" to
-        "estimated".}
+        Perhaps change ``estimated'' to ``estimate'', or ``estimated of'' to
+        ``estimated''.}
+
+        Done.
 

### 2018-06-18 articles/cycle/alps-cycle.tex
Fix technical comments from referee 1.

@@
     nearly three hundred years, it thus remains incompletely known
-      (1) what climate evolution lead to the known maximum ice limits,
-      (2) to which extent ice flow was controlled by subglacial topography,
+      (1) what climate evolution led to the known maximum ice limits,
+      (2) to what extent ice flow was controlled by subglacial topography,
       (3) what drove the different response of the individual lobes,
@@
         line,][]{Ehlers.etal.2011}.
-        \textbf{(c)} Extract from the estimated of present-day ice thickness
+        \textbf{(c)} Extract from the estimated present-day ice thickness
         \citep{Huss.Farinotti.2012} substracted from the SRTM topography and

### 2018-07-15 articles/cycle/response-referee2.tex
Fix technical comments from referee 2.

@@
 
+        Done.
+
     \referee{\textbf{p.~2, l.~22:}
         could you add a reference and a timing for LGM?}
+
+        Done.
 
@@
 
+        Done.
+
     \referee{\textbf{p.~3, l.~3:}
         add s to responses.}
+
+        Done.
 
@@
 
+        Done.
+
     \referee{\textbf{p.~4, l.~4:}
         suggest to replace ``field'' with ``sheet''.}
+
+        Done.
 
@@
         more explanation is needed here, indicate also, like in the other
-        figures that h) is January and i) is July figuresPage 6 line 6 suggest
-        to replace ``of'' with ``with''.}
+        figures that h) is January and i) is July figures.}
+
+    We have added a scale, marked the inset location, and clarified the text.
+
+    \referee{\textbf{p.~6, l.~6:}
+        suggest to replace ``of'' with ``with''.}
+
+        Done.
 
@@
 
+        Done.
+
     \referee{\textbf{p.~6, l.~15:}
@@
 
+        Done.
+
     \referee{\textbf{p.~6, l.~19:}
         add ``mean'' after monthly.}
+
+        Done.
 
@@
 
+        This is true, done.
+
     \referee{\textbf{p.~6, l.~27:}
         replace ``is'' with ``are''.}
+
+        Done.
 
@@
 
+    The correct formula can be found in \citep[p.~443]{Love.1906}. The lenght
+    scale of bedrock deformation can be computed as
+    ${\alpha=\sqrt[4]{\frac{4D}{(\rho_m-\rho_l)g}}}$ \citep{Walcott.1970}, thus
+    the error in lenght scale can be expressed as $\sqrt[4]{1+\nu}-1 = 5.7\%$.
+    The sentence was clarified and references were added.
+
     \referee{\textbf{p.~7, l.~2:}
@@
 
+    We clarified the sentence. However, we would like to keep the current
+    notation to remain consistent with previous papers \citep{Seguinot.2014,
+    Seguinot.etal.2014}.
+
     \referee{\textbf{p.~7, l.~19:}
@@
 
+    Done.
+
     \referee{\textbf{p.~7, l.~24:}
         ``and within a rectangular region ..'' is not clear, edit this text.}
+
+    The sentence was reorganised.
 
@@
 
+    Done.
+
     \referee{\textbf{p.~9, l.~14--15:}
         see comment above, could this be an example of Arctic amplification?}
+
+    %FIXME
 
@@
 
+    Done.
+
     \referee{\textbf{p.~9, l.~20:}
@@
         with ``red''.}
+
+    Done.
 
@@
 
+    Done.
+
     \referee{\textbf{p.~10, l.~2:}
         suggest to add ``followed by a rapid retreat''.}
+
+    Done.
 
@@
 
+    The numbers 4 and 15 relate to the total (minimum) number of Pleistocene
+    glaciations. However, evidence for pre-LGM glaciations during the last
+    glacial cycle is debated and very limited (see later part of the
+    introduction). We added the word ``Pleistocene'' in the introduction and
+    clarified the concerned sentence about model results.
+
     \referee{\textbf{p.~10, l.~10:}
         suggest to replace ``lower'' with ``smaller''.}
+
+    Done.
 
@@
 
+    %FIXME
+
     \referee{\textbf{p.~10, l.~12:}
         suggest to replace ``least'' with ``the smallest''.}
+
+    Done.
 
@@
 
+    We replaced ``cumulative extent of modelled ice cover'' with ``modelled
+    maximum extent'', ``reasonable'' with ``realistic'', and ``red line'' with
+    ``black line'', which is what we meant.
+
     \referee{\textbf{p.~11, l.~2:}
@@
         or something similar.}
+
+    As for the caption, we replaced with ``modelled maximum extent''.
 
@@
 
+    %FIXME
+
     \referee{\textbf{p.~11, l.~4--5:}
         ``outside this benchmark'' clarify if you mean spatially or temporally.}
+
+    Done. We mean spatially.
 
@@
 
+    We have included this discussion here as we think that understanding the
+    simiarities and differences in the spatial patterns of glaciation between
+    the different palaeo-temperature records used is necessary to justifying
+    the choice of the EPICA forcing used in the next section. We think that
+    this non-standard outline is the best way to present all results in a
+    single manuscript.
+
     \referee{\textbf{p.~12, l.~2:}
         suggest to replace ``higher'' with ``larger''.}
+
+    Done.
 
@@
 
+    No, but the reference to Fig.~2 was misplaced. This has been corrected.
+
     \referee{\textbf{p.~12, l.~29--32:}
         does this text fit better in method section?}
+
+    We agree that this text does not really fit in a results and discussion
+    section. However, this information is specific to Sect.~4. We feel that
+    moving these and the corresponding sentences in Sect.~3 to the methods
+    section would increase the complexity of the manuscript.
 
@@
 
+    The 200\,m contours were thickened, the caption was reworked as suggested,
+    and the reference to Natural Earth Data was moved to the caption of Fig.~1.
+
     \referee{\textbf{p.~15, l.~8:}
@@
 
+    Done.
+
     \referee{\textbf{p.~15, l.~25:}
         ``occurred''.}
+
+    Done.
 
@@
 
+    The map shows maximum thickness. The colourbar label was corrected.
+    Surface elevation contours were thickened. Indeed, in much of the mountain
+    area maximum thickness is reached before 27\,ka. This is now clarified in
+    the figure caption.
+
     \referee{\textbf{p.~17, l.~30:}
         suggest to replace ``have'' with ``could''.}
+
+    Done.
 
@@
 
+    The 200\,m contours were thickened.
+
     \referee{\textbf{p.~19, l.~7:}
@@
         comparison?}
+
+    Done.
 
@@
 
+    We added the info about MIS. The wording ``periodic surges'' was badly
+    chosen and replaced by a more cautious ``episodic advances''. These
+    advances are likely due to thermodynamical feedbacks modulated by
+    fluctuations in the climate forcing, but they were not explored in detail
+    and perhaps depend on model resolution.
+
     \referee{\textbf{p.~21, l.~21:}
@@
 
+    See my previous comment about last glacial cycle and older glaciations.
+
     \referee{\textbf{p.~21, l.~27:}
@@
         ``In this study the model has been applied...''}
+
+    Done.
 
@@
 
+    The reference to the Cordilleran ice sheet was removed here. As mentioned
+    in the later part of the conclusions, it is clear to us that additional
+    sensitivity studies are needed. With this study, we hope to have
+    highlighted areas of research were data-model comparison could be pursued
+    further. However, this will require not only more model runs, but also a
+    more systematic classication of geological data in digital databases.
+
     \referee{\textbf{p.~22, l.~6:}
         add ``records'' after forcing.}
+
+    We replaced ``forcing'' by ``records''.
 
@@
 
+    The timing of the LGM extent depends on the actual climate history of the
+    European Alps, which was almost certainly different and more complex than
+    our climate forcing scenarii.  Although our simulations show that even a
+    spatially homogeneous forcing produce spatial variations in the timing of
+    maximum extent, on can not exclude that a different climate history might
+    counteract these variations.
+
     \referee{\textbf{p.~22, l.~20:}
         suggest to replace ``higher'' with ``larger''.}
+
+    Done.
 
@@
 
+    We removed the first ``potentially''. Nevertheless, the actual number and
+    timing of glacier advances onto the foreland depends on Alpine climate
+    history, which was may have been very different from the forcing used in
+    our simulations.
+
     \referee{\textbf{p.~22, l.~24:}
@@
         sentence.}
+
+    Done.
 
@@
 
+    Done. The complicated sentence was rephrased.
+
     \referee{\textbf{p.~23, l.~2:}
         replace ``mode'' with ``more''.}
+
+    Done.
 
@@
 
+    Thank you. The 200\,m contour levels were thickened on all figures.
+
+% ----------------------------------------------------------------------
+
+\bibliographystyle{copernicus}
+\bibliography{../../references/references}
+
 \end{document}

### 2018-07-15 articles/cycle/alps-cycle.tex
Fix technical comments from referee 2.

@@
     century, more systematic studies of glacial features showed that Alpine
-    glaciers extended well below their current margins \citep{Venetz.1821}
+    glaciers extended well outside their current margins \citep{Venetz.1821}
     and even onto the Alpine foreland \citep{Charpentier.1841}, yielding the
@@
     of glaciofluvial stratigraphy in the foreland indicate at least 15
-    glaciations \citep{Schluchter.1988, Ivy-Ochs.etal.2008,
+    Pleistocene glaciations \citep{Schluchter.1988, Ivy-Ochs.etal.2008,
     Preusser.etal.2011}.
@@
     present a record of the last major glaciation of the Alps, dating from the
-    \emph{Last Glacial Maximum} (LGM).
+    \emph{Last Glacial Maximum} \citep[LGM;][]{Ivy-Ochs.2015,
+    Wirsig.etal.2016, Monegato.etal.2017}.
 
@@
     Although the glacial history of the European Alps has been studied for
-    nearly three hundred years, it thus remains incompletely known
+    nearly three hundred years, it still remains incompletely known
       (1) what climate evolution led to the known maximum ice limits,
       (2) to what extent ice flow was controlled by subglacial topography,
-      (3) what drove the different response of the individual lobes,
+      (3) what drove the different responses of the individual lobes,
       (4) how far above the trimline was the ice surface located, and
@@
 
-    Ice deformation follows temperature and water-content dependent creep
+    Ice deformation follows a temperature and water-content dependent creep
+    formulation
     (Sect.~\ref{sec:icedyn}). Basal sliding follows a~pseudo-plastic law where
@@
     where both vertical shear and longitudinal stretch are significant,
-    it was shown to effectively approximate complex ice field flow
+    it was shown to effectively approximate complex ice sheet flow
     dynamics over mountainous topographies similar to that of the European Alps
@@
         was computed using the erroneous formula, ${D=YE^3/12(1-\nu)}$, in
-    \citet{Mey.etal.2016}. The correct formula is ${D=YE^3/12(1-\nu^2)}$.
-    Using the Young's modulus, $Y=100$\,GPa, and the Poisson ratio, $\nu=0.25$
-    \citep{Mey.etal.2016}, our simulations effectively use an elastic thickness
-    of the lithosphere of $E=52.4$\,km, instead of the 50\,km value used by
-    \citet{Mey.etal.2016}, introducing in a small change in the lenght scale of
-    bedrock deformation well within uncertainties \citep{Mey.etal.2016}.}
+        \citet{Mey.etal.2016}. The correct formula is ${D=YE^3/12(1-\nu^2)}$
+        \citep[p.~443]{Love.1906}.  Using the Young's modulus, $Y=100$\,GPa,
+        and the Poisson ratio, $\nu=0.25$ \citep{Mey.etal.2016}, the
+        consequence of this error is that the simulations effectively use an
+        elastic thickness of the lithosphere of $E=53.8$\,km instead of 50\,km
+        which is well within uncertainties \citep{Mey.etal.2016}, and thus
+        introduces a ${\sqrt[4]{1+\nu}-1=5.7\,\%}$ change in the lenght scale
+        of bedrock deformation \citep{Walcott.1970}.}
 
@@
         \citep{Seguinot.2013} of daily mean temperature from the ERA-Interim
-        \citep[1979--2012;][]{Dee.etal.2011} from the reference monthly
-        climatology used to force the surface mass balance (PDD) component of
-        the ice sheet model.
+        \citep[1979--2012;][]{Dee.etal.2011} monthly climatology approximating
+        temperature variability.
         \textbf{(f)} Modern January and \textbf{(g)} July mean near-surface air
-        temperature, and \textbf{(h, i)} precipitation from WorldClim
-        \citep[1960--1990;][]{Hijmans.etal.2005}.}
+        temperature, and \textbf{(h)} January and \textbf{(i)} July
+        precipitation from WorldClim \citep[1960--1990;][]{Hijmans.etal.2005}
+        used to compute surface mass balance. The background maps contain
+        Natural Earth Data \citep{Patterson.Kelso.2017}.
+        \todo{(R2) add scale and mark inset.}}
       \label{fig:inputs}
@@
     The PDD computation accounts for stochastic temperature variations by
-    assuming a~normal temperature distribution of standard deviation, $\sigma$,
-    around the expected value, $T_{\mathrm{m}}$. It is expressed by an
-    error-function formulation \citep{Calov.Greve.2005}
+    assuming a~normal temperature distribution with standard deviation,
+    $\sigma$, around the expected value, $T_{\mathrm{m}}$. It is expressed by
+    an error-function formulation \citep{Calov.Greve.2005}
     which is numerically approximated using week-long sub-intervals. In
@@
 
-    Climate forcing driving ice sheet simulations consists of a~present-day
-    monthly climatology, $\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, modified by
-    temperature lapse-rate corrections, ${\Delta}T_{\text{LR}}$, temperature
+    The climate forcing driving the ice sheet simulations consist spatially of
+    a~present-day monthly mean climatology,
+    $\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, modified by temperature lapse-rate
+    correction, ${\Delta}T_{\text{LR}}$, temperature
     offset time series, ${\Delta}T_{\text{TS}}$, and time-dependent
@@
     %
-    The present-day monthly climatology was bilinearly interpolated from
+    The present-day monthly mean climatology was bilinearly interpolated from
     near-surface air temperature and precipitation rate fields from
@@
     precipitation (Fig.~\ref{fig:inputs}h) which is reversed in summer
-    (Fig.~\ref{fig:inputs}h). WorldClim data
+    (Fig.~\ref{fig:inputs}i). WorldClim data
     were selected as an input to the ice sheet model because they incorporate
@@
     density is lower than in the Alps. Finally, the last glacial cycle Alpine
-    glaciers did not extend over marine areas where WorldClim data is missing.
+    glaciers did not extend over marine areas where WorldClim data are missing.
 
     The temperature lapse-rate corrections, ${\Delta}T_{\text{LR}}$, are
-    computed as a~function of ice surface elevation, $s$, using the SRTM
-    topography shipped with WorldClim as a~reference, $b_{\text{ref}}$:
+    computed as a~function of ice surface elevation, $s$, using the SRTM-based
+    topography data provided with WorldClim as a~reference, $b_{\text{ref}}$:
     %
@@
 
-        $g$     & Standard gravity
+        $g$     & Standard acceleration due to gravity
                 & 9.81
@@
                 & \unit{J\,mol^{-1}\,K^{-1}}
-                & -- \\
+                & \citet[p.~72]{Cuffey.Paterson.2010} \\
 
@@
     core from the Iberian margin \citep[MD01-2444;][]{Martrat.etal.2007}.
-    Palaeo-temperature anomalies from the GRIP record are calculated from
+    Palaeo-temperature anomalies from the GRIP record are calculated from the
     oxygen isotope (\chem{\delta^{18}O}) measurements using a~quadratic
@@
     scaled linearly (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}a) so
-    that the modelled cumulative glaciated area during MIS~2 (29--14\,ka)
-    and within a rectangular region covering the Rhine
-    glacier piedmont lobe (Fig.~\ref{fig:footprints}a, black rectangle) is
+    that, within a $150\times100$\,km rectangular domain covering the Rhine
+    glacier piedmont lobe (Fig.~\ref{fig:footprints}a, black rectangle), the
+    modelled cumulative glaciated area during MIS~2 (29--14\,ka) is
     consistent with the glaciated area of the geological reconstruction
@@
         \textbf{(b)} Modelled total ice volume through the last 120~thousand
-        years (ka) expressed in meters of sea-level equivalent (m~s.l.e.). Gray
-        fields indicate Marine Oxygen Isotope Stage (MIS) boundaries for MIS~2
-        and MIS~4 according to a~global compilation of benthic
-        \chem{\delta^{18}O} records \citep{Lisiecki.Raymo.2005}. The simulation
+        years (ka) expressed in meters of sea-level equivalent (m~s.l.e.).
+        Shaded gray areas indicate the timing for MIS~2 and MIS~4
+        \citep{Lisiecki.Raymo.2005}. The simulation
         driven by the EPICA temperature yields smaller ice volume variability
@@
     as late as 22 to 17\,ka \citep[Fig.~5]{Ivy-Ochs.2015, Wirsig.etal.2016}.
-    This is in very
-    good agreement with the result of the EPICA simulations, which yield an
+    The EPICA simulations yield an
     early maximum ice volume at 25.2/24.6 (without/with palaeo-precipitation
-    reductions) followed by a standstill until 17.3\,ka
-    (Fig.~\ref{fig:timeseries}b, blue curves). The GRIP palaeo-temperature
-    forcing, on the other hand, yields two distinct total ice volume maxima
+    reductions) followed by a retreat and then a standstill until 17.3\,ka
+    (Fig.~\ref{fig:timeseries}b, red curves) in a very good agreement with
+    these geological data. On the other hand, the simulations forced by the
+    GRIP palaeo-temperature record yield two distinct total ice volume maxima
     at 27.3/27.0 and 21.8/21.7\,ka followed by an early deglaciation of the
     foreland by 21.4\,ka. The MD01-2444 palaeo-temperature forcing yields
-    a late LGM peak ice volume at 16.5/15.7\,ka several thousand years after
-    the dated geological evidence.
-
-    Finally, all simulations yield very strong ice volume
-    variability, indicative of many more than two or three cycles of glacier
-    advance and retreats onto the foreland (Fig.~\ref{fig:timeseries}b).
+    a late LGM peak ice volume at 16.5/15.7\,ka followed by a rapid retreat
+    several thousand years after the dated geological evidence.
+
+    Finally, all simulations yield very strong ice volume variability over the
+    last 120\,ka (Fig.~\ref{fig:timeseries}b), indicative of many more than the
+    two or three documented cycles of glacier advance and retreats onto the
+    foreland for this time period \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
     Palaeo-precipitation reductions dampen some of the small-scale variability,
@@
     \citep{Spotl.Mangini.2002, Luetscher.etal.2015}. Our results indicate that
-    their magnitude in Central Europe was likely lower than that recorded in
+    their magnitude in Central Europe was likely smaller than that recorded in
     GRIP \chem{\delta^{18}O}, which either implies smaller associated
@@
     temperature-independent controls on GRIP \chem{\delta^{18}O}. In contrast,
-    the EPICA palaeo-temperature record has least variability among the records
+    the EPICA palaeo-temperature record has the smallest variability among the
+    records
     used (Fig.~\ref{fig:timeseries}a, red curves). This results in smaller
@@
       \caption{%
-        \textbf{(a--c)} Cumulative extent of modelled ice cover during MIS~2
+        \textbf{(a--c)} Modelled maximum ice extent during MIS~2
         (29--14\,ka), without (dark colours) and with (light colours)
@@
         agreement with the Last Glacial Maximum (LGM) geomorphological
-        reconstruction \citep[solid red line,][]{Ehlers.etal.2011}.
-        \textbf{(d--f)} Cumulative extent of modelled ice cover during MIS~4
+        reconstruction \citep[solid black line,][]{Ehlers.etal.2011}.
+        \textbf{(d--f)} Modelled maximum ice extent during MIS~4
         (71--57\,ka). Only the simulation driven by the EPICA temperature
-        time-series yields reasonable MIS~4 ice cover.}
+        time-series yields realistic MIS~4 ice cover.}
       \label{fig:footprints}
@@
 
-    During MIS~2, all six simulations yield comparative cumulative glaciated
-    areas (Fig.~\ref{fig:footprints}a). This result is inherent to the
+    During MIS~2, all six simulations yield comparative modelled maximum
+    extents (Fig.~\ref{fig:footprints}a). This result is inherent to the
     palaeo-climate forcing approach used, which involves a linear scaling of
@@
     (Sect.~\ref{sec:paltemp}; Fig.~\ref{fig:footprints}a, black rectangle).
-    Outside this benchmark, all simulations underestimate glaciated area in the
-    western part of the model domain, and overestimate it in the eastern
+    Outside this benchmark area, all simulations underestimate ice extent in
+    the western part of the model domain, and overestimate it in the eastern
     part, relative to the geomorphological ice limits
@@
     precipitation corrections are applied (bright colours). In fact, both
-    palaeo-temperature records have a higher temperature variability and
-    contain brief spells of cold temperatures, which are too short to develop a
+    palaeo-temperature records have a larger temperature variability and
+    contain brief spells of cold temperatures (Fig.~\ref{fig:timeseries}b,
+    blue and green curves), which are too short to develop a
     fully grown Alpine ice sheet, but long enough to build up ice cover on a
-    smaller scale on these peripheral ranges (Fig.~\ref{fig:timeseries}). On
+    smaller scale on these peripheral ranges. On
     the other hand, peripheral glaciation modelled using the EPICA record
@@
         24.57\,ka before present, corresponding to the maximum modelled ice
-        cover. Modelled self-sustained ice domes (triangles) and some major
-        transfluences (crosses) are shown. Modelled LGM
-        ice extent (dashed orange line) and geomorphological
-        reconstruction \citep[solid red line,][]{Ehlers.etal.2011}. The
-        background map consists of depressed SRTM \citep{Jarvis.etal.2008}
-        topography and Natural Earth Data \citep{Patterson.Kelso.2017}.
+        cover. The modelled LGM ice extent (dashed orange line) and
+        geomorphological reconstruction \citep[solid red
+        line,][]{Ehlers.etal.2011}, modelled self-sustained ice domes
+        (triangles), and some major transfluences (crosses) are shown.
         \textbf{(b)} Temperature offset time-series from the EPICA ice core
@@
         last glacial cycle (120--0\,ka), expressed in meters of sea level
-        equivalent (m~s.l.e., blue curve). Gray fields indicate Marine
-        Oxygen Isotope Stage (MIS) boundaries for MIS~2 and MIS~4 according to
+        equivalent (m~s.l.e., blue curve). Shaded gray areas indicate the
+        timing for Marine Oxygen Isotope Stage (MIS)~2 and MIS~4 according to
         a~global compilation of benthic \chem{\delta^{18}O} records
-        \citep{Lisiecki.Raymo.2005}.}
+        \citep{Lisiecki.Raymo.2005}.
+        \todo{Thicken the 200\,m contours on all figures.}}
       \label{fig:lgmvel}
@@
 
-    The LGM Alpine ice flow pattern was traditionally described as that of a
+    The LGM Alpine ice flow pattern is traditionally described as that of a
     network of interconnected valley glaciers, primarily controlled by
@@
 
-    In the Eastern Alps, major transfluences are modelled to have occur for
+    In the Eastern Alps, major transfluences are modelled to have occurred for
     instance across Fern Pass, the Seefeld Saddle, the Gailberg Saddle, the
@@
         and modelled glacierized area during the LGM (coloured curve). The LGM
-        is here modelled as a time-transgressive event.}
+        is here modelled as a time-transgressive event. In much of the mountain
+        area, maximum thickness is reached before 27\,ka.
+        \todo{fix colourbar label}}
       \label{fig:timing}
@@
     Heterogeneous climatic anomalies, such as temperature or precipitation
-    anomaly gradients, not included in our model set-up, have perhaps induced
+    anomaly gradients, not included in our model set-up, could perhaps induce
     further spatial differences in the timing of the LGM Alpine ice sheet,
-    which may have counterbalanced or enhanced those modelled here.
+    which may counterbalance or enhance those modelled here.
 
@@
     difference of 861\,m (Fig.~\ref{fig:trimlines}b), wich is significant in
-    respect to the modelled regional ice thickness. The result is
+    respect to the modelled regional ice thickness \todo{add a number}. The result is
     somewhat dependent on uncertain basal sliding and ice rheological parameters
@@
         numerous cycles of advance and retreat over the last glacial cycle
-        modulated by subglacial topography and catchment geometry.
-        Isolated patches indicate periodic surges from tributary glaciers.}
+        modulated by subglacial topography and catchment geometry. Shaded gray
+        areas indicate the timing for MIS~2 and MIS~4
+        \citep{Lisiecki.Raymo.2005}. Isolated patches indicate episodic
+        advances from tributary glaciers.}
       \label{fig:profiles}
@@
 
-    This study consists in the application of the numerical ice sheet model
-    PISM to ice dynamics of the last glacial cycle in the Alps, using a model
-    set-up based on that previously developed and validated for the Cordilleran
-    ice sheet (Sect.~\ref{sec:model}).
+    In this study the numerical ice sheet model PISM (Sect.~\ref{sec:model})
+    has been applied to ice dynamics of the last glacial cycle in the Alps.
     Using three different palaeo-temperature forcing records (GRIP, EPICA, and
@@
             compatible with geological evidence during MIS~4 and 2, whereas
-            both GRIP and MD01-2444 forcing result in MIS~4 glaciation well
+            both GRIP and MD01-2444 records result in MIS~4 glaciation well
             beyond the mapped LGM ice limits (Sect.~\ref{sec:footprints}).
@@
             range due to inherent glacier dynamics (Sect.~\ref{sec:timing}).
-      \item Ice thickness during the LGM is modelled to be much higher than in
+      \item Ice thickness during the LGM is modelled to be much larger than in
             reconstructions. In average, modelled surface elevation is 861\,m
@@
             englacial thermal boundary (Sect.~\ref{sec:thickness}).
-      \item Alpine glaciers were potentially very dynamic. They quickly
+      \item Alpine glaciers were very dynamic. They quickly
             responded to climate fluctuations and some potentially advanced
@@
 
-    These results are nevertheless limited by the simplicity of the climate
+    However, these results are limited by the simplicity of the climate
     forcing and surface mass balance model applied, and uncertainties in ice
@@
     the European Alps may have caused more glaciations than modelled here.
-    Through more detailed sensitivity studies, targeted to
-    specific aspects of the last glacial cycle ice dynamics in the Alps, future
+    Using more specifically targeted sensitivity runs, future
     modelling studies will certainly be able to quantify uncertainties
-    associated with some of the above statements. Nevertheless, we hope that
+    associated with some of the above limitations. Nevertheless, we hope that
     these conclusions will also serve as a basis for future studies of glacial
@@
     publication, we will be happy to make aggregated model output available
-    through The Cryosphere. Please contact the corresponding author for mode
+    through The Cryosphere. Please contact the corresponding author for more
     detailed model output data.}

### 2018-07-15 articles/cycle/response-referee1.tex
Add Makefile rules for responsesAdd Makefile rules for responses..

@@
         on small-scale highs or lows significantly different from their
-        ∼km-scale surroundings, that could contribute to the discrepancies in
+        ~km-scale surroundings, that could contribute to the discrepancies in
         Fig.~6c.}

### 2018-07-15 figures/anim_alps_9k.sh
Add 9k animation tiling script.

@@ -1,0 +1,22 @@
+#!/bin/bash
+
+# animation name
+name=$(basename "${0%.sh}")
+orig="$HOME/anim/$name"  # folder containing animation frames
+dest="$name.tms"  # destination folder for tiles
+
+# find gdal2tiles command
+gdal2tiles="$(command -v gdal2tiles.py || command -v gdal2tiles2.py)"
+
+# loop on frames
+for year in {005..120..5}000
+do
+
+    # locally link current frame
+    ln -sf $orig/$year.png $name.png
+
+    # prepare tiles
+    [ -d $dest/$year ] || $gdal2tiles -r antialias -s "EPSG:32632" \
+        -w none -z 6-10 $name.png $dest/$year
+
+done

### 2018-07-17 articles/cycle/response-referee3.tex
Fix technical comments from referee 3.

@@
 
-    \referee{\textbf{p.1, l.~2:}
-        ``pioneer'' should be ``pioneering'' P1, L16: I think the finding that
+    \referee{\textbf{p.~1, l.~2:}
+        ``pioneer'' should be ``pioneering''.}
+
+        Corrected.
+
+    \referee{\textbf{p.~1, l.~16:}
+        I think the finding that
         you get asynchronous glaciation extents with a uniform climate offset
@@
 
+        Thank you for pointing this out. We have reworked the last sentence in
+        the abstract:
+
+        \msquote{%
+            Finally, despite the uniform climate forcing, differences in
+            glacier catchment hypsometry cause the Last Glacial Maximum advance
+            to be modelled as a time-transgressive event, with some glaciers
+            reaching their maximum as early as 26\,ka, and others as late as
+            20\,ka.}
+
 % ----------------------------------------------------------------------
@@
 
-    \referee{\textbf{p.2, l.~28:}
+    \referee{\textbf{p.~2, l.~28:}
         Ballantyne and Stone (2015) should be added to this list.}
 
-    \referee{\textbf{p.2, l.~34:}
+        This reference is relevant to the discussion on trimlines. We have
+        added the reference in the introduction and in the corresponding
+        section in the discussion.
+
+    \referee{\textbf{p.~2, l.~34:}
         It should be stated that it could be a consequence of both
@@
 
-    \referee{\textbf{p.3, l.~2:}
+        We have added ``or both''.
+
+    \referee{\textbf{p.~3, l.~2:}
         These points serve the literature well to highlight gaps for future
@@
 
+        To clarify our statement of intent, the following text was added:
+
+        \msquote{%
+            Additional geological research will be needed to complete our
+            knowledge. But here, we intend to explore these open questions from
+            a new angle and [use the Parallel Ice Sheet Model ...].}
+
 % ----------------------------------------------------------------------
@@
 
-    \referee{\textbf{p.7:}
+    \referee{\textbf{p.~7:}
         The spatial distribution of your modern climate variables (precip,
@@
 
-    \referee{\textbf{p.7, l.~14:}
-        You keep mentioning the number of processors. I find this infor- mation
+    \referee{\textbf{p.~7, l.~14:}
+        You keep mentioning the number of processors. I find this information
         slightly irrelevant, and it will soon become outdated as processing
@@
 
-    \referee{\textbf{p.10, l.~14:}
+    We removed mentions of computing times and numbers of processors.
+
+    \referee{\textbf{p.~10, l.~14:}
         This is a more general point. You eventually choose the EPICA record,
@@
 
-    \referee{\textbf{p.11, l.~2:}
+    \referee{\textbf{p.~11, l.~2:}
         First sentence needs reconsidering as it is slightly broken in its
@@
 
-    \referee{\textbf{p.11, l.~10:}
+    We reworked the first sentence. It now reads:
+
+    \msquote{%
+        During MIS 2, all six simulations yield comparative cumulative
+        glaciated areas modelled maximum extents (Fig.~3a).}
+
+    \referee{\textbf{p.~11, l.~10:}
         Is it possible some ice is missing from the geological reconstruction
@@
 
-    \referee{\textbf{p.14, l.~15--31:}
+    \referee{\textbf{p.~14, l.~15--31:}
         I find this description of sites and timing of glacier extent compared
@@
 
-    \referee{\textbf{p.17, l.~17:}
+    \referee{\textbf{p.~17, l.~17:}
         This finding is important and should be highlighted better in abstract
@@
 
-    \referee{\textbf{p.17, l.~22:}
+        This finding was better highlighted in the abstract (see our previous
+        comment). However, we feel it is already well expressed by the
+        following bullet point in the conclusion:
+
+        \msquote{%
+            The LGM (maximum) extent was a transient stage in which glaciers
+            were out of balance with the contemporary climate. Its timing
+            potentially varied across the range due to inherent glacier
+            dynamics (Sect. 4.3).}
+
+    \referee{\textbf{p.~17, l.~22:}
         Is the model recreating possible surging? Or would this be an
@@
 
-    \referee{\textbf{p.19, l.~9--12:}
+    \referee{\textbf{p.~19, l.~9--12:}
         On trimlines: I think you justify well why you haven't yet modelled the
@@
         it exists and interpretations may be slightly wrong. The model is also
-        not the ``truth'' and there is probably a blurred line inbetween upon
+        not the ``truth'' and there is probably a blurred line in-between upon
         which we can proceed.}
 
-    \referee{\textbf{p.21, l.~11--19:}
+    \referee{\textbf{p.~21, l.~11--19:}
         Be really clear here that these are modelled, and perhaps not
@@
         it might be correct, might be just a modelled result.}
+
+        Actually we mean geologically recorded asymmetry. This clearly needed
+        clarification. The corresponding sentence now reads:
+
+        \msquote{%
+            The observed asymmetric extent of ice north and south of the Alps
+            can be explained by the modelled transient nature of the LGM extent
+            without involving north-south gradients in temperature and
+            precipitation change (Sect.~4.1).}
 

### 2018-07-17 articles/cycle/alps-cycle.tex
Fix technical comments from referee 3.

@@
 
-    The European Alps, cradle of pioneer glacial studies, are one of the
+    The European Alps, cradle of pioneering glacial studies, are one of the
     regions where geological markers of past glaciations are most abundant and
@@
     glacial cycle (120--0\,ka) in the Alps, using horizontal resolutions of 2
-    and 1\,km and up to 576~processors.
-    Climate forcing is derived using present-day climate data from
+    and 1\,km. Climate forcing is derived using present-day climate data from
     WorldClim and the ERA-Interim reanalysis, and time-dependent temperature
@@
     our simulation predicts little erosion at high elevation due to cold-based
-    ice. Finally, the Last
-    Glacial Maximum advance, often considered synchronous, is here modelled as
-    a time-transgressive event, with some glacier lobes reaching their maximum
-    as early as 27\,ka, and some as late as 21\,ka.
+    ice. Finally, despite the uniform climate forcing, differences in glacier
+    catchment hypsometry cause the Last Glacial Maximum advance to be modelled
+    as a time-transgressive event, with some glaciers reaching their maximum as
+    early as 26\,ka, and others as late as 20\,ka.
 
@@
     been repeatedly invalidated in other glaciated regions of the globe
-    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012}. Although
+    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012,
+    Ballantyne.Stone.2015}. Although
     no evidence for cold-based glaciation has been reported in the Alps to
@@
     lobes exists \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear whether
-    this relates to climate or glacier dynamics \citep{Monegato.etal.2017} or
-    uncertainties in the dating methods.
+    this relates to climate or glacier dynamics \citep{Monegato.etal.2017},
+    uncertainties in the dating methods, or both.
 
@@
       (5) how many advances occurred during the last glacial cycle.
-
-    Here, we use the Parallel Ice Sheet Model
+    Additional geological research will be needed to complete our knowledge.
+
+    But here, we intend to explore these open questions from a new angle and
+    use the Parallel Ice Sheet Model
     \citep[PISM,][]{PISM-authors.2017}, a numerical ice sheet model that
@@
     layers in the ice, corresponding to vertical resolutions of 100 and
-    40\,m, respectively. Simulations ran for two to four days
-    using 144~processors.
+    40\,m, respectively.
 
@@
     251~enthalpy layers in the ice, corresponding to vertical resolutions of 50
-    and 20\,m, respectively. This simulation ran for 33~days using
-    576~processors.
+    and 20\,m, respectively.
 
@@
 
-    On a more regional level, the maximum extent of the Ticino, Oglio,
+    On a more regional level, the maximum extent of the Adda, Oglio,
     Adige and Piave Glaciers in the central southern Alps is modelled
@@
     nuclide dating evidence from
-    Scandinavia \citep[e.g.,][]{Kleman.1994, Kleman.Borgstrom.1994},
-    the British Isles \citep[e.g.,][]{Fabel.etal.2012}, and
+    Scandinavia \citep[e.g.,][]{Kleman.1994, Kleman.Borgstrom.1994}, the
+    British Isles \citep[e.g.,][]{Fabel.etal.2012, Ballantyne.Stone.2015}, and
     North America \citep[e.g.,][]{Kleman.etal.2010}, that pre-glacial landscapes
@@
             model forcing, probably controlled the LGM extent of ice cover in
-            the Alps. The asymmetric extent north and south of the Alps
-            can be explained by the transient nature of the LGM extent without
-            involving north-south gradients in temperature and precipitation
-            change (Sect.~\ref{sec:extent}).
+            the Alps. The observed asymmetric extent of ice north and south of
+            the Alps can be explained by the modelled transient nature of the
+            LGM extent without involving north-south gradients in temperature
+            and precipitation change (Sect.~\ref{sec:extent}).
       \item The LGM ice flow pattern in the Alps was largely controlled by

### 2018-07-19 articles/cycle/response-referee1.tex
Assess referee 1 comment on water content.

@@
 
+    Thank you very much for this elogious summary.
+
 % ----------------------------------------------------------------------
@@
         and does it have a significant effect on results?}
+
+    Laboratory experiments have demonstrated that the rheology of temperate,
+    polycrystalline ice depends on its content in liquid water
+    \citep[p.~65--66]{Cuffey.Paterson.2010}. Unfortunately, the only
+    measurements available to date \citep{Duval.1977}, used to quantify the
+    effect of liquid water on ice softness, the creep parameter $A$ in Glen's
+    flow law \citep{Lliboutry.Duval.1985}, only extend to fractions of liquid
+    water content between 0 and 0.8\%. They show a three-fold increase of ice
+    softness over this range \citep[Fig.~1]{Duval.1977}.
+
+    Although ice sheet models have previously ignored this effect, it has now
+    become a typical component of polythermal models such as SICOPOLIS
+    \citep{Greve.1997}, COMICE \citep{Ruckamp.etal.2010}, PISM
+    \citep{Aschwanden.etal.2012}, ISSM \citep{Seroussi.etal.2013}, and
+    TIM-FD$\unit{^{3}}$ \citep{Kleiner.Humbert.2014}.
+
+    Water fractions between 1 and 5\,\% have repreatedly been observed in
+    temperate glaciers \citep{Murray.etal.2000, Murray.etal.2007,
+    Bradford.Harper.2005, Bradford.etal.2009}, and also occur in model results
+    \citep[e.g.,][]{Blatter.Greve.2015}, but it is not known whether ice
+    viscosity continues to decrease substancially for values above 0.8\,\%.
+    Previous modelling studies have commonly assumed constant ice viscosity
+    above 1\,\%. This arbitrary threshold is not constrained at all, and the
+    urgent need for new ice deformation experiments has been pointed out
+    \citep{Kleiner.etal.2015}.
+
+    The sensitivity of our results to the 1\,\% threshold was not tested.
+    However, in our model results, liquid water fractions above 1\,\% typically
+    only occur within the basal temperate layer of the fastest-moving glaciers
+    where ice movement is dominated by basal sliding. We suspect that a local
+    increase of ice deformation there is negligible in comparaison to
+    uncertainties caused by basal sliding and more importantly climate forcing.
+
+    We prefer to avoid including the above discussion in the manuscript, but we
+    have reworked the sentence on water content:
+
+    \msquote{%
+        [Ice softness] increases with liquid water fractions up to 0.01
+        \citep[p.~65--66]{Duval.1977, Lliboutry.Duval.1985,
+        Cuffey.Paterson.2010}, an arbitrary threshold above which new
+        measurements are critically needed \citep{Kleiner.etal.2015}.}
+
+    The uncertainty to unknown rheology of water-rich temperate ice was also
+    mentioned in the conclusions.
 
@@
 
+% ----------------------------------------------------------------------
+
+\sechead{References}
+
+\bibliographystyle{copernicus}
+\bibliography{../../references/references}
+
 \end{document}

### 2018-07-19 articles/cycle/alps-cycle.tex
Assess referee 1 comment on water content.

@@
     pressure, and water content, through an enthalpy scheme
-    \citep[Table~\ref{tab:params};][]{Aschwanden.etal.2012} and a piece-wise
-    Arrhenius-type rheology representative of Holocene polar ice
-    \citep[Table~\ref{tab:params};][p.~77]{Cuffey.Paterson.2010}.
-    The englacial water fraction is
-    capped at a~maximum value of 0.01, above which no measurements are
-    available \citep[Eq.~5.7]{Lliboutry.Duval.1985, Greve.1997}.
+    \citep[Table~\ref{tab:params};][]{Aschwanden.etal.2012}. It follows a
+    piece-wise Arrhenius-type rheology representative of Holocene polar ice
+    \citep[Table~\ref{tab:params};][p.~77]{Cuffey.Paterson.2010}, and increases
+    with liquid water fractions up to 0.01 \citep[p.~65--66]{Duval.1977,
+    Lliboutry.Duval.1985, Cuffey.Paterson.2010}, an arbitrary threshold above
+    which new measurements are critically needed \citep{Kleiner.etal.2015}.
 
@@
 
-    However, these results are limited by the simplicity of the climate
-    forcing and surface mass balance model applied, and uncertainties in ice
-    flow physics. In particular, additional palaeo-climate variability over
+    However, these results are limited by uncertainties on basal sliding and
+    on the unknown rheology of water-rich temperate ice. The potential
+    variability in surface mass balance parameters, as well as the simplicity
+    of the climate forcing, are even more important.
+    In particular, additional palaeo-climate variability over
     the European Alps may have caused more glaciations than modelled here.

### 2018-07-19 articles/cycle/response-referee1.tex
Assess referee 1 comment on subglacial hydrology.

@@
         thawed areas, which is relevant to section 4.4 regarding trimlines.}
+
+    We refer to \citet{Bueler.Pelt.2015} as their paper contain the most
+    up-to-date description of PISM till effective pressure physics used in our
+    simulations \citet[Eqs.~18, 23, and 24]{Bueler.Pelt.2015}. However,
+    subglacial water is not routed down the hydropotential gradient. We have
+    clarified this:
+
+    \msquote{%
+        Effective pressure is related to the ice overburden stress and the
+        modelled amount of subglacial water, using a~formula derived from
+        laboratory experiments with till extracted from the base of Ice Stream
+        B in West Antarctica \citep[Table~1;][Eqs~23 and
+        24]{Tulaczyk.etal.2000, Bueler.Pelt.2015}. Basal meltwater is
+        accumulated locally without transportation. When the till becomes
+        saturated, additional meltwater assumed to drain off instantaneously
+        outside the glacier margins, i.e. it is removed from the system in an
+        accountable way.}
 

### 2018-07-19 articles/cycle/alps-cycle.tex
Assess referee 1 comment on subglacial hydrology.

@@
     sliding velocity. The yield stress is modelled using the Mohr--Coulomb
-    criterion using a constant basal friction angle (Table~\ref{tab:params}).
+    criterion using a constant basal friction angle
+    \citep[Table~\ref{tab:params};][Eq.~18]{Tulaczyk.etal.2000}.
     Effective pressure is related to the ice overburden stress and the modelled
@@
     experiments with till extracted from the base of Ice Stream B in West
-    Antarctica \citep[Table~\ref{tab:params};][]{Tulaczyk.etal.2000,
-    Bueler.Pelt.2015}. When the till becomes saturated, additional melt water
-    is assumed to drain off instantaneously.
+    Antarctica \citep[Table~\ref{tab:params};][Eqs~23 and
+    24]{Tulaczyk.etal.2000, Bueler.Pelt.2015}. Basal meltwater is accumulated
+    locally without transportation. When the till becomes saturated, additional
+    meltwater assumed to drain off instantaneously outside the glacier margins,
+    i.e. it is removed from the system in an accountable way.
 

### 2018-07-19 articles/cycle/response-referee1.tex
Assess referee 1 comment on basal sliding.

@@
     The uncertainty to unknown rheology of water-rich temperate ice was also
-    mentioned in the conclusions.
+    mentioned in the conclusions:
+
+    \msquote{%
+        A constant rheology was used for temperate ice containing more than
+        1\,\% of liquid water.}
 
@@
         overall?}
+
+    In general, we chosed to not repeat parameter values from Table~1 in the
+    main text. The following text was added in the methods:
+
+    \msquote{%
+        [a constant basal friction angle] corresponding to the average of
+        available measurements \citep[p.~268]{Cuffey.Paterson.2010}. [...]
+        Other parameters (Table~1) follow simulations of the Greenland ice
+        sheet \citep{Aschwanden.etal.2013}, or benchmarks when other data is
+        missing \citep{Bueler.Pelt.2015}.}
+
+    Inversion of specific basal sliding parameters for past Alpine glaciers
+    is difficult because the altitude and age of maximum ice surface elevation
+    is discussed (cf. introduction and discussion on trimlines), and also
+    depends on the even more uncertain regional climate history.  Therefore,
+    basal sliding was also mentioned as one of the major sources of
+    uncertainty in the conclusions.
+
+    \msquote{%
+        The till deformation model used here does not hold for sliding over
+        bedrock surfaces. On the other hand, the constant friction angle used
+        is representative of wet till but weaker basal conditions may have
+        applied over saturated lake sediments where they occured.}
 

### 2018-07-19 articles/cycle/alps-cycle.tex
Assess referee 1 comment on basal sliding.

@@
     sliding velocity. The yield stress is modelled using the Mohr--Coulomb
-    criterion using a constant basal friction angle
-    \citep[Table~\ref{tab:params};][Eq.~18]{Tulaczyk.etal.2000}.
+    criterion \citep[Table~\ref{tab:params};][Eq.~18]{Tulaczyk.etal.2000}.
+    using a constant basal friction angle corresponding to the average of
+    available measurements \citep[p.~268]{Cuffey.Paterson.2010}.
     Effective pressure is related to the ice overburden stress and the modelled
@@
     meltwater assumed to drain off instantaneously outside the glacier margins,
-    i.e. it is removed from the system in an accountable way.
+    i.e. it is removed from the system in an accountable way. Other parameters
+    (Table~\ref{tab:params}) follow simulations of the Greenland ice sheet
+    \citep{Aschwanden.etal.2013}, or benchmarks when other data is missing
+    \citep{Bueler.Pelt.2015}.
 
@@
 
-    However, these results are limited by uncertainties on basal sliding and
-    on the unknown rheology of water-rich temperate ice. The potential
-    variability in surface mass balance parameters, as well as the simplicity
-    of the climate forcing, are even more important.
+    However, these results are limited by uncertainties on ice physics. The
+    the till deformation model used here does not hold for sliding over
+    bedrock surfaces. On the other hand, the constant friction angle used is
+    representative of wet till but weaker basal conditions may have applied
+    over saturated lake sediments where they occured. A constant rheology was
+    used for temperature ice containing more than 1\,\% of liquid water.
+
+    More importantly, these results are limited by the simplicity of the
+    surface mass balance parameters and climate forcing used.
     In particular, additional palaeo-climate variability over

### 2018-07-19 articles/cycle/response-referee1.tex
Assess referee 1 comment on lapse rate.

@@
 
+    In the European Alps, monthly temperature lapse rates vary between
+    approximately 4\,K\,km$^{-1}$ in winter and 7\,K\,km$^{-1}$ in summer, and
+    annual temperature lapse rates vary spatially between 5.4 and
+    5.8\,K\,km$^{-1}$ \citep{Rolland.2003}. The reference to
+    \citet{Rolland.2003} was added to Table~1, and as we have now made
+    explained in the manuscript, our constant value of 6\,K\,km$^{-1}$ is thus:
+
+    \msquote{%
+        slightly above average but more representative of summer months when
+        surface melt occurs \citep[Fig.~3]{Rolland.2003}.}
+
+    Although no sensitivity tests were conducted, we argue here that the effect
+    of atmospheric temperature lapse rate variations on ice temperature is
+    negligible. First, seasonal variations do not penetrate ice or bedrock
+    below a few metres. Second, even above 2\,km of ice, spatial variations of
+    0.4\,K\,km$^{-1}$ would translate into surface temperature variations of
+    only 0.8\,K. But during the Last Glacial Maximum, ice surface temperatures
+    in the trimline region are typically between 15 and 20\,K below freezing.
+    In addition, this effect would nearly disappear near the base where the
+    temperature gradient is much steeper and primarily controlled by geothermal
+    heat flux and shear heating.
+
 % ----------------------------------------------------------------------

### 2018-07-19 articles/cycle/response-referee1.tex
Assess referee 1 comment on MIS~3 precipitation.

@@
 
+    We did not know about references on MIS~3 and have developped our sentence
+    to better reflect the current state of knowledge on palaeo-precipitation:
+
+    \msquote{%
+        Palaeoclimate proxies indicate slightly reduced LGM precipitation in
+        western Europe with anomalies diminishing eastwards
+        \citep{Wu.etal.2007}. Regional circulation models indicate generally
+        dryer conditions during MIS~3 \citep{Barron.Pollard.2002,
+        Kjellstrom.etal.2010} but more precipitation south of the Alps during
+        MIS~2 LGM \citep{Strandberg.etal.2011, Ludwig.etal.2016}.}
+
     \referee{%
@@
 % ----------------------------------------------------------------------
-
-\sechead{References}
+% References
 

### 2018-07-19 articles/cycle/alps-cycle.tex
Assess referee 1 comment on lapse rate.
Assess referee 1 comment on MIS~3 precipitation.

@@
     the other hand. All simulations use an annual temperature lapse rate of
-    $\gamma=6\,\unit{K\,km^{-1}}$.
+    $\gamma=6\,\unit{K\,km^{-1}}$, which is slightly above average but more
+    representative of summer months when surface melt occurs
+    \citep[Fig.~3]{Rolland.2003}.
 
@@
                 & \unit{K\,m{-1}}
-                & -- \\
+                & \citep{Rolland.2003} \\
 
@@
     circulation changes that governed moisture availability over the Alps
-    during the last glacial cycle, of which little is known apart from the
-    LGM \citep[cf.][]{Wu.etal.2007, Strandberg.etal.2011, Ludwig.etal.2016}.
+    during the last glacial cycle. Palaeoclimate proxies indicate slightly
+    reduced LGM precipitation in western Europe with anomalies diminishing
+    eastwards \citep{Wu.etal.2007}. Regional circulation models indicate
+    generally dryer conditions during MIS~3 \citep{Barron.Pollard.2002,
+    Kjellstrom.etal.2010} but more precipitation south of the Alps during MIS~2
+    LGM \citep{Strandberg.etal.2011, Ludwig.etal.2016}.
     Thus, another set of simulations presented here use constant

### 2018-07-20 articles/cycle/response-referee1.tex
Assess referee 1 comment on regional proxies.

@@
 
+    A short review of available proximal proxy records was added here:
+
+    \msquote{%
+        Only few regional proxy records exist that extend over periods when the
+        Alps were glaciated \citep{Heiri.etal.2014}. These include lake
+        sediment records in north and west of the Alps
+        \citep[e.g.,]{Beaulieu.Reille.1992, Wohlfarth.etal.2008,
+        Duprat-Oualid.etal.2017}, and cave speleothem in the Eastern
+        \citep[e.g.,][]{Spotl.Mangini.2002, Boch.etal.2011} and Western
+        \citep{Luetscher.etal.2015} Alps. Due to the scarcity of vegetation
+        north of the Alps during glacial periods, varying sources for moisture
+        advection, and the limited duration of the records, quantitative
+        palaeoclimatic interpretation will require combining multiple proxies
+        in space and time, and comparing them against regional circulation
+        model output \citep{Heiri.etal.2014}.}
+
+    In the review paper by \citet{Heiri.etal.2014}, the latter has been
+    specifically identified as a neeeded future development, but to our
+    knowledge, no quantitative reconstruction has been made available since
+    then. The following word of caution was also added in the conclusions:
+
+    \msquote{%
+        This suprising result is likely coincidental as there is no direct link
+        between European and Antarctic climate. This highlights the need for
+        more quantitative reconstructions of European palaeoclimate.}
+
     \referee{%

### 2018-07-20 articles/cycle/alps-cycle.tex
Assess referee 1 comment on regional proxies.

@@
 
-    Temperature offset time-series, ${\Delta}T_{\text{TS}}$, are derived from
+    Only few regional proxy records exist that extend over periods when the Alps
+    were glaciated \citep{Heiri.etal.2014}. These include lake sediment records
+    in north and west of the Alps \citep[e.g.,]{Beaulieu.Reille.1992,
+    Wohlfarth.etal.2008, Duprat-Oualid.etal.2017}, and cave speleothem in the
+    Eastern \citep[e.g.,][]{Spotl.Mangini.2002, Boch.etal.2011} and Western
+    \citep{Luetscher.etal.2015} Alps. Due to the scarcity of vegetation north
+    of the Alps during glacial periods, varying sources for moisture advection,
+    and the limited duration of the records, quantitative palaeoclimatic
+    interpretation will require combining multiple proxies in space and time,
+    and comparing them against regional circulation model output
+    \citep{Heiri.etal.2014}.
+
+    In our simulations, temperature offset time-series,
+    ${\Delta}T_{\text{TS}}$, are derived from distal
     palaeo-temperature proxy records from the Greenland Ice Core Project
@@
     \end{itemize}
+
+    This suprising result is likely coincidental as there is no direct link
+    between European and Antarctic climate. This highlights the need for more
+    quantitative reconstructions of European palaeoclimate.
 
@@
     However, these results are limited by uncertainties on ice physics. The
-    the till deformation model used here does not hold for sliding over
+    till deformation model used here does not hold for sliding over
     bedrock surfaces. On the other hand, the constant friction angle used is

### 2018-07-20 articles/cycle/response-referee1.tex
Assess referee 1 comment on insulation.

@@
 
+    We agree. This is one of many possible improvements that could be tested
+    upon the climate forcing used here. In the conclusions we now advocate for:
+
+    \msquote{%
+        a more realistic climate forcing based on regional circulation model
+        output or including the effect of long-term term changes in incoming
+        solar radiation.}
+
 % ----------------------------------------------------------------------

### 2018-07-20 articles/cycle/alps-cycle.tex
Assess referee 1 comment on insulation.

@@
     the European Alps may have caused more glaciations than modelled here.
-    Using more specifically targeted sensitivity runs, future
+    Using more specifically targeted sensitivity runs, and a more realistic
+    climate forcing based on regional circulation model output or including
+    the effect of long-term term changes in incoming solar radiation, future
     modelling studies will certainly be able to quantify uncertainties

### 2018-07-20 data/postprocess_alpcyc_bestrun.py
Postprocess max thickness basal temperature.

@@ -26,7 +26,7 @@ y = nc.variables['y'][:]
 age = -nc.variables['time'][:]/(365.0*24*60*60)
 thk = da.from_array(nc.variables['thk'], chunks=chunks)
 srf = da.from_array(nc.variables['usurf'], chunks=chunks)
-tpa = da.from_array(nc.variables['temppabase'], chunks=chunks)
+tba = da.from_array(nc.variables['temppabase'], chunks=chunks)
 uba = da.from_array(nc.variables['uvelbase'], chunks=chunks)
 vba = da.from_array(nc.variables['vvelbase'], chunks=chunks)
 cba = da.from_array(nc.variables['velbase_mag'], chunks=chunks)
@@ -79,12 +79,14 @@ print "computing maximum ice thickness..."
 maxicethk = thk.max(axis=0).compute()  # could use nd fancy indexing
 print "computing max thickness surface elevation..."
 maxthksrf = srf.compute()[argmaxthk, jj, kk]  # no dask, very slow
+print "computing max thickness basal temperature..."
+maxthktba = tba.compute()[argmaxthk, jj, kk]  # no dask, very slow
 print "computing total basal sliding..."
 totalslip = (icy*cba).sum(axis=0).compute(get=dask.get)*dt
 print "computing MIS2 warm-based ice cover..."
 mis2 = (age < 29e3) * (age >= 14e3)
 mis2print = icy[mis2].any(axis=0).compute()
-warmbased = (icy*(tpa >= -1e-3))[mis2].sum(axis=0).compute(get=dask.get)*dt
+warmbased = (icy*(tba >= -1e-3))[mis2].sum(axis=0).compute(get=dask.get)*dt
 
 # compute derived variables
 print "computing footprint..."
@@ -116,6 +118,7 @@ maxicethk = np.ma.masked_where(~footprint, maxicethk)
 maxsrfage = np.ma.masked_where(~footprint, maxsrfage)
 maxthkage = np.ma.masked_where(~footprint, maxthkage)
 maxthksrf = np.ma.masked_where(~footprint, maxthksrf)
+maxthktba = np.ma.masked_where(~footprint, maxthktba)
 totalslip = np.ma.masked_where(~footprint, totalslip)
 warmbased = np.ma.masked_where(~mis2print, warmbased)
 
[…]

### 2018-07-20 articles/cycle/response-referee1.tex
Adress referee 1 comments on trimlines.

@@
 
+    Thank you for emphasizing this point.
+
     \referee{%
@@
         "...experienced temperate basal ice for ...".}
+
+    We realise this is confusing. We had chosen this approach because temperate
+    basal ice tends to occur above the trimlines during periods of warmer
+    climate. To avoid confusion, Fig.~6c has been simplified.
 
@@
 
+    Fig.~6c has been simplified to show the basal thermal boundary at the age
+    of maximum ice thickness. Although this results in a time-transgressive
+    picture, it means that the frozen areas indicated on the new figure are
+    contemporaneous with the plotted ages and surface topography contours, and
+    data shown on panels~a and b.
+
+    Although the model output allows to study the migration of the basal
+    thermal boundary over time into more detail, the basal velocity, which
+    controls erosion, should probably be considered as well. We leave this for
+    future studies.
+
     \referee{%
@@
         than 29\,ka(?).}
+
+    Indeed, much of the mountainous areas reach maximum ice thickness during
+    early MIS~2 when ice is colder and thus stiffer. This is also visible on
+    Fig.~5. Some areas even reach maximum ice thickness during MIS~4 when the
+    bedrock is slightly less depressed. There is certainly very much scope to
+    discuss the age of trimline, which may differ significantly from that of
+    the maximum ice extent in the lowland, but we are not aware of any data
+    that could be used for validation here. However, frozen areas plotted on
+    Fig.~6c are now consistent with the modelled ages used in panels a and c.
 
@@
         Fig.~6c.}
+
+    This is a valid point which we can unfortunately not rule out. Despite
+    PISM's enthalpy scheme, which ensures a seamless transition between cold
+    and temperate ice physics, resolving the basal thermal boundary on the
+    valley sides is delicate, because it implies resolving a steep (vertical)
+    enthalpy gradient over a steep bedrock slope. This requires both high
+    vertical and horizontal grid resolutions. This limitation was mentioned
+    in the section on trimlines:
+
+    \msquote{%
+        This result is certainly affected by the limited ability of the 1\,km
+        horizontal resolution to resolve small-scale topographic features. It
+        calls for a more detailed comparison.}
 

### 2018-07-20 articles/cycle/alps-cycle.tex
Adress referee 1 comments on trimlines.

@@
         mark areas modelled to have been glaciated but to have experienced
-        temperate ice for less than 1\,ka during MIS~2 (29--14\,ka).}
+        temperate ice for less than 1\,ka during MIS~2 (29--14\,ka).
+        \todo{simplify basal thermal map, caption and main text.}}
       \label{fig:trimlines}
@@
     show a remarkable agreement with the areal extent of regions that are
-    modelled to have experienced warm-based ice for less than 1\,ka (7\,\%) of the
-    MIS~2 (Fig.~\ref{fig:trimlines}c). Although this result calls for more
-    detailed comparisons across the
-    entire Alpine range and specific sensitivity studies to relevant basal
-    sliding and ice rheological parameters, and to the uncertain subglacial
-    topography the presence of an upper layer of
+    modelled to have experienced warm-based ice for less than 1\,ka (7\,\%) of
+    the MIS~2 (Fig.~\ref{fig:trimlines}c). This result is certainly affected by
+    the limited ability of the 1\,km horizontal resolution to resolve
+    small-scale topographic features. It calls for a more detailed comparison
+    spanning the entire Alpine range and specific sensitivity studies to
+    relevant basal sliding and ice rheological parameters, and to the uncertain
+    subglacial topography. However, the presence of an upper layer of
     cold ice during the LGM, already found at high altitude in the much warmer

### 2018-07-21 articles/cycle/response-referee1.tex
Update responses on trimlines using new figure.

@@
     basal ice tends to occur above the trimlines during periods of warmer
-    climate. To avoid confusion, Fig.~6c has been simplified.
+    climate. To avoid confusion, Fig.~6c has been simplified to show the basal
+    thermal boundary at the age of maximum ice thickness. The last sentence in
+    the caption now read:
+
+    \msquote{%
+        Hatches mark the LGM cold-based ares (basal temperature above $1 \times
+        10^{-3}$\,K below freezing at the age of maximum ice thickness).}
+
+    Unfortunately this also results in a worse fit to trimline locations, as
+    was noted in the main text:
+
+    \msquote{%
+        In the upper Rhone Valley, observed trimlines are often located near
+        the LGM cold-temperate basal thermal transition, or in cold-based areas
+        (Fig.~6c).}
 
@@
     controls erosion, should probably be considered as well. We leave this for
-    future studies.
+    future studies. The following sentence was added in the main text:
+
+    \msquote{%
+        The remaining discrepancies may relate to temporal migrations of the
+        basal thermal boundary, an absence of sliding in warm-based areas}
 
@@
     \msquote{%
-        This result is certainly affected by the limited ability of the 1\,km
-        horizontal resolution to resolve small-scale topographic features. It
-        calls for a more detailed comparison.}
+        [The remaining discrepancies may relate to] levelling of small-scale
+        topographic features in the 1\,km horizontal grid. They call for more
+        detailed comparisons [...]}
 

### 2018-07-21 articles/cycle/alps-cycle.tex
Update responses on trimlines using new figure.

@@
         \citep[Table~1]{Kelly.etal.2004}, modelled age of maximum ice thickness
-        (colour) and LGM ice surface elevation (200\,m contours). Hatches
-        mark areas modelled to have been glaciated but to have experienced
-        temperate ice for less than 1\,ka during MIS~2 (29--14\,ka).
-        \todo{simplify basal thermal map, caption and main text.}}
+        (colour) and LGM ice surface elevation (200\,m contours). Hatches mark
+        the LGM cold-based ares (basal temperature above $1 \times 10^{-3}$\,K
+        below freezing at the age of maximum ice thickness).}
       \label{fig:trimlines}
@@
 
-    In the upper Rhone Valley, the geographical location of observed trimlines
-    show a remarkable agreement with the areal extent of regions that are
-    modelled to have experienced warm-based ice for less than 1\,ka (7\,\%) of
-    the MIS~2 (Fig.~\ref{fig:trimlines}c). This result is certainly affected by
-    the limited ability of the 1\,km horizontal resolution to resolve
-    small-scale topographic features. It calls for a more detailed comparison
+    In the upper Rhone Valley, observed trimlines are often located near the
+    LGM cold-temperate basal thermal transition, or in cold-based areas
+    (Fig.~\ref{fig:trimlines}c). The remaining discrepancies may relate to
+    temporal migrations of the basal thermal boundary, an absence of sliding in
+    warm-based areas, and levelling of small-scale topographic features in the
+    1\,km horizontal grid. They call for more detailed comparisons
     spanning the entire Alpine range and specific sensitivity studies to

### 2018-07-21 articles/cycle/response-referee2.tex
Adress referee 2 request for sliding sensitivity.

@@
         in their model is a transient stage with varying timing across their
-        model domain due to glacier dynamics. The manuscript is well organized
+        model domain due to glacier dynamics.}
+
+    Thank you very much for this accurate summary.
+
+    \referee{%
+        The manuscript is well organized
         and clearly written, the missing thing in this study is a discussion of
@@
         the model sensitivity would strengthen the paper.}
+
+    We agree that a sensitivity study to ice physics, in particular basal
+    sliding, is one of many improvements needed upon our results. However, for
+    the kind of exercise presented here, the sensitivity of to palaeo-climate
+    history, which is virtually unknown and has many degrees of freedom, far
+    outweight the model sensitivity to ice physics. For this manuscript we have
+    chosen to focus on the primary source of sensitivity, which is climate.
+
+    Because the kind of simulations presented here depends on computing
+    resources only available through a peer-review application process, we are
+    currently unable to perform an additional sensitivity study on basal
+    sliding for the entire Alps.  Regional sentivity studies are more
+    accessible. A regional sensitivity analysis on basal sliding is precisely
+    the topic of a follow-up study currently under preparation for the Rhine
+    Glacier \citep{Imhof.etal.2017}, which has a relatively isolated catchment
+    for its size as compared to other Alpine Glaciers.
+
+    Meanwhile, the following text was added in the section on trimlines, which
+    is certainly the mostly sensitive aspect of our results to basal sliding:
+
+    \msquote{%
+        The remaining discrepancies may relate to temporal migrations of the
+        basal thermal boundary, an absence of sliding in warm-based areas, and
+        levelling of small-scale topographic features in the 1\,km horizontal
+        grid. They call for more detailed comparisons spanning the entire
+        Alpine range and specific sensitivity studies to relevant basal sliding
+        and ice rheological parameters, and to the uncertain subglacial
+        topography.}
+
+    Basal sliding and other sources of uncertainties in ice physics where also
+    mentioned in the conclusions:
+
+    \msquote{%
+        However, these results are limited by uncertainties on ice physics. The
+        till deformation model used here does not hold for sliding over bedrock
+        surfaces. On the other hand, the constant friction angle used is
+        representative of wet till but weaker basal conditions may have applied
+        over saturated lake sediments where they occured. A constant rheology
+        was used for temperature ice containing more than 1\,\% of liquid
+        water.}
 

### 2018-07-21 articles/cycle/alps-cycle.tex
Address referee 2 comment on defining age scale.

@@
     indicating several tens of glacial and interglacial periods. During the
-    last 800\,ka before present, these \emph{glacial cycles}
+    last 800\,ka (thousand years before present), these \emph{glacial cycles}
     have succeeded each other with a periodicity about 100\,ka \citep{Hays.etal.1976,

### 2018-07-21 articles/cycle/response-referee2.tex
Address referee 2 comment on defining age scale.
Address referee 2 comment on Arctic amplification.

@@
 
+    We have now defined this better in the introduction:
+
+    \msquote{
+        During the last 800\,ka (thousand years before present)}
+
     \referee{%
@@
         line 11 on page 10.}
+
+    These passages refer to the Younger Dryas and Dansgaard-Oeschger events.
+    One must keep in mind that a linear scaling was applied to the GRIP record,
+    so that it is already ``de-amplified'' before being used as temperature
+    forcing.  Arctic amplification could be an explanation if it was stronger
+    then than for changes that lead the LGM climate. Although we lack expertise
+    on this topic, we could not find literature supporting this idea. Instead,
+    current theories for the Younger Dryas and Dansgaard-Oeschger events mostly
+    involve large amounts of freshwater discharge into the North Atlantic
+    implying a more regional mechanism than arctic amplification. Therefore, we
+    decided not mention arctic amplification in the text.
 
@@
 
-    %FIXME
+    Please refer to my comment above.
 

### 2018-07-22 articles/cycle/response-referee3.tex
Fix formatting in the response to referee 3.

@@
 
+Thank you very much for this supportive comment.
+
 % ----------------------------------------------------------------------
@@
     \referee{\textbf{p.~1, l.~16:}
-        I think the finding that
-        you get asynchronous glaciation extents with a uniform climate offset
-        is due to glacier hypsometry and setting should be stated here. i.e.
-        that the timing of maximum glaciation and recession isn't purely a
-        function of climate. This finding needs to be highlighted better in the
-        abstract.}
+        I think the finding that you get asynchronous glaciation extents with a
+        uniform climate offset is due to glacier hypsometry and setting should
+        be stated here. i.e.  that the timing of maximum glaciation and
+        recession isn't purely a function of climate. This finding needs to be
+        highlighted better in the abstract.}
 
@@
     \msquote{%
-            Finally, despite the uniform climate forcing, differences in
-            glacier catchment hypsometry cause the Last Glacial Maximum advance
-            to be modelled as a time-transgressive event, with some glaciers
-            reaching their maximum as early as 26\,ka, and others as late as
-            20\,ka.}
+        Finally, despite the uniform climate forcing, differences in glacier
+        catchment hypsometry cause the Last Glacial Maximum advance to be
+        modelled as a time-transgressive event, with some glaciers reaching
+        their maximum as early as 26\,ka, and others as late as 20\,ka.}
 
@@
 
-        This reference is relevant to the discussion on trimlines. We have
-        added the reference in the introduction and in the corresponding
-        section in the discussion.
+    This reference is indeed relevant to the discussion on trimlines. We have
+    added the reference in the introduction and in the corresponding section in
+    the discussion.
 
@@
 
-        To clarify our statement of intent, the following text was added:
+    In order to clarify our statement of intent, the following text was added:
 
@@
         Additional geological research will be needed to complete our
-            knowledge. But here, we intend to explore these open questions from
-            a new angle and [use the Parallel Ice Sheet Model ...].}
+        knowledge. But here, we intend to explore these open questions from a
+        new angle and [use the Parallel Ice Sheet Model ...].}
 
@@
     This finding was better highlighted in the abstract (see our previous
-        comment). However, we feel it is already well expressed by the
-        following bullet point in the conclusion:
+    comment). However, we feel it is already well expressed by the following
+    bullet point in the conclusion:
 
     \msquote{%
-            The LGM (maximum) extent was a transient stage in which glaciers
-            were out of balance with the contemporary climate. Its timing
-            potentially varied across the range due to inherent glacier
-            dynamics (Sect. 4.3).}
+        The LGM (maximum) extent was a transient stage in which glaciers were
+        out of balance with the contemporary climate. Its timing potentially
+        varied across the range due to inherent glacier dynamics (Sect. 4.3).}
 
@@
     \msquote{%
-            The observed asymmetric extent of ice north and south of the Alps
-            can be explained by the modelled transient nature of the LGM extent
-            without involving north-south gradients in temperature and
-            precipitation change (Sect.~4.1).}
+        The observed asymmetric extent of ice north and south of the Alps can
+        be explained by the modelled transient nature of the LGM extent without
+        involving north-south gradients in temperature and precipitation change
+        (Sect.~4.1).}
 

### 2018-07-22 articles/cycle/response-referee3.tex
Address referee 3 comment on precipitation change.

@@
         topography, and if so, how does this influence your results?}
+
+    The temperature lapse rate corrections dampen local high temperatures in
+    ice-filled valleys. The precipitation pattern, on the other hand, remains
+    the same throughout the simulation, and this results in higher accumulation
+    above the location of present-day mountains even if they are buried by ice
+    in the model.
+
+    Although we expect that small-scale variations in surface mass balance are
+    smoothed out by ice flow, the presence of an ice sheet may have
+    redistributed precipitation on a larger scale, for instance from the inner
+    ranges to the foothills, or inequally between north and south of the Alps,
+    etc. However, global precipitation reductions appear to have little effect
+    on the average ice thickness (Fig.~2b, compare dark and light curves).
+    Nevertheless, the new text shortly mentions this caveat:
+
+    \msquote{%
+        Besides [circulation changes], changes in the ice surface topography
+        may have redistributed orographic precipitation.}
+
+    Our conclusions also highlight the need for a more realistic forcing.
+
+    \msquote{%
+        Using [...] a more realistic climate forcing based on regional
+        circulation model output [...] future modelling studies will certainly
+        be able to quantify uncertainties associated with some of the above
+        limitations.}
 

### 2018-07-22 articles/cycle/alps-cycle.tex
Address referee 3 comment on precipitation change.

@@
     %
-    with $\psi=0.169/2.4=0.0704$ \citep{Huybrechts.2002}. This simple
-    relationship likely does not reflect the complexity of atmospheric
+    with $\psi=0.169/2.4=0.0704$ \citep{Huybrechts.2002}. The control set of
+    simulations use constant precipitation, corresponding to $\psi=0$.
+    This simple relationship does not reflect the complexity of atmospheric
     circulation changes that governed moisture availability over the Alps
@@
     Kjellstrom.etal.2010} but more precipitation south of the Alps during MIS~2
-    LGM \citep{Strandberg.etal.2011, Ludwig.etal.2016}.
-    Thus, another set of simulations presented here use constant
-    precipitation, corresponding to $\psi=0$.
+    LGM \citep{Strandberg.etal.2011, Ludwig.etal.2016}. Besides, changes in the
+    ice surface topography may have redistributed orographic precipitation.
 

### 2018-07-23 articles/cycle/response-referee3.tex
Address referee 3 comment on EPICA.
Address referee 3 comment on surge.

@@
 
+    We agree. The following text was added in the conclusions:
+
+    \msquote{%
+        This suprising result is likely coincidental as there is no direct link
+        between European and Antarctic climate. This highlights the need for
+        more quantitative reconstructions of European palaeoclimate.}
+
     \referee{\textbf{p.~11, l.~2:}
@@
         clarify.}
+
+    Our simulations reproduce a thermodynamical instability akin to a surge.
+    The internal energy (so-called entropy) conservation scheme used in PISM
+    allows for variations in ice temperature and liquid water content
+    \citep{Aschwanden.etal.2012}, which may result in thermodynamical
+    oscillations \citep{Pelt.Oerlemans.2012, Feldmann.Levermann.2017}.
+
+    Unlike periodic surges \citep[as described by, e.g.,][]{Sevestre.Benn.2015},
+    we interpret the model results as triggered by an initial perturbation of
+    mass (and energy) caused by rapid climate cooling and inefficient mass
+    transfer to lower elevations. We have rephrase the following sentence:
+
+    \msquote{%
+        Several Alpine lobes surge and overshoot their balanced extent before
+        thinning and receding towards the mountains as they warm towards
+        thermodynamical equilibrium.}
 
@@
 
+% ----------------------------------------------------------------------
+% References
+
+\bibliographystyle{copernicus}
+\bibliography{../../references/references}
+
 \end{document}

### 2018-07-23 articles/cycle/alps-cycle.tex
Address referee 3 comment on surge.

@@
     For larger glaciers, this thermodynamical equilibrium is typically not yet
-    attained around 25\,ka. This is why several Alpine lobes overshoot their
+    attained around 25\,ka. Several Alpine lobes surge and overshoot their
     balanced extent before thinning and receding towards the mountains as they

### 2018-07-23 figures/alpcyc_hr_trimlines.py
Fix bug in trimline figure histogram.

@@ -30,13 +30,6 @@ maxthkage, extent = ut.io.load_postproc_gtif(ut.alpcyc_bestrun, 'maxthkage')
 maxthksrf, extent = ut.io.load_postproc_gtif(ut.alpcyc_bestrun, 'maxthksrf')
 maxthktba, extent = ut.io.load_postproc_gtif(ut.alpcyc_bestrun, 'maxthktba')
 
-# get coordinates  # FIXME not efficient
-filepath = ut.alpcyc_bestrun + 'y0120000-extra.nc'
-nc = ut.io.load(filepath)
-x = nc.variables['x'][:]
-y = nc.variables['y'][:]
-nc.close()
-
 # load boot topography
 nc = ut.io.load('input/boot/alps-srtm+thk+gou11simi-1km.nc')
 x = nc.variables['x'][:]
@@ -78,13 +71,13 @@ step = 100.0
 bmin = dt.min() - dt.min() % step
 bmax = dt.max() - dt.max() % step + step
 bins = np.arange(bmin, bmax+step, step)
-hsax.hist(ht, bins=bins, orientation='horizontal', alpha=0.75)
+hsax.hist(dt, bins=bins, orientation='horizontal', alpha=0.75)
 hsax.set_xlabel('frequency')
 hsax.set_ylabel('difference (m)')
 hsax.yaxis.set_label_position("right")
 hsax.yaxis.tick_right()
 
-# highlight mean thickness
+# highlight mean difference
 zavg = zt.mean()
 davg = dt.mean()
 dstd = dt.std()

### 2018-07-23 articles/cycle/response-referee3.tex
Address referee 3 comment on trimlines.

@@
 
+    This is a valid point. However we found an error in our figure: the data
+    plotted on the histogram (Fig.~3b) were different from that in the scatter
+    plot (Fig.~3a). These data corresponded to raw interpolated ice
+    thicknesses, i.e. they missed a correction for altitude shifts between our
+    initial basal topography and the observed trimline elevations.
+
+    The corrected histogram is closer to symmetry. A slight skewness remains,
+    but this is not really a result of our simulations. In fact the
+    distribution of observations itself is skewed, with more trimlines observed
+    at higher elevations, and the differences between the modelled ice surface
+    and trimlines are slightly higher at higher elevations (not plotted because
+    correlation is poor).
+
+    The modal value (now ca.~950\,m) corresponds to a cluster of trimline
+    locations around 2850\,m in the Aletsch Glacier (south of Jungfrau) region
+    where trimlines are well visible in the landscape. We prefer not to
+    use it because the number of trimlines mapped is not necessarily
+    proportional to the lenght, but may be instead depend on other factors such
+    as bedrock lithology and post-glacial erosion.
+
+    We now use the standard deviation (197\,m) as a measure of variance, but
+    not uncertainty, which would require more sensitivity tests. We also
+    mention resolution as one of several caveats in the text:
+
+    \msquote{%
+        The remaining discrepancies may relate to temporal migrations of the
+        basal thermal boundary, an absence of sliding in warm-based areas, and
+        levelling of small-scale topographic features in the 1\,km horizontal
+        grid.}
+
+    We added a reference to the study by \citet{Kuchar.etal.2012} although this
+    approach is not doable for the Alps.
+
+    \msquote{%
+        Unfortunately, validation through the bedrock uplift rates
+        \citep[cf.][]{Kuchar.etal.2012} is not doable in the Alps due its lower
+        values, active tectonics and uncertainties on geological properties
+        \citep[cf.][]{Mey.etal.2016}.}
+
     \referee{%

### 2018-07-23 articles/cycle/alps-cycle.tex
Address referee 3 comment on trimlines.

@@
     observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
-    difference of 861\,m (Fig.~\ref{fig:trimlines}b), wich is significant in
-    respect to the modelled regional ice thickness \todo{add a number}. The result is
-    somewhat dependent on uncertain basal sliding and ice rheological parameters
+    difference of $861\pm197$\,m (one standard deviation,
+    Fig.~\ref{fig:trimlines}b), wich is significant in respect to the modelled
+    regional ice thickness \todo{add a number}. The result is somewhat
+    dependent on uncertain basal sliding and ice rheological parameters
     to which the model sensitivity was not tested here. However, a previous
@@
     relevant basal sliding and ice rheological parameters, and to the uncertain
-    subglacial topography. However, the presence of an upper layer of
+    subglacial topography. Unfortunately, validation through the bedrock uplift
+    rates \citep[cf.][]{Kuchar.etal.2012} is not doable in the Alps due its
+    lower values, active tectonics and uncertainties on geological properties
+    \citep[cf.][]{Mey.etal.2016}. However, the presence of an upper layer of
     cold ice during the LGM, already found at high altitude in the much warmer

### 2018-07-23 articles/cycle/response-referee3.tex
Address referee 3 comment on geomorph uncertainty.

@@
 
+    Yes, parts of the reconstructed LGM ice limits are uncertain due to the
+    lack of moraines or dated evidence. However, the discrepancies in ice
+    extent discussed here are beyond uncertainties.
+
 % ----------------------------------------------------------------------
@@
         maps, including geochronological constraints.}
+
+    This would be very useful but to our knowledge, no geological data have
+    been made available that would allow such a comparison. The review by
+    \citep[Fig.~5]{Wirsig.etal.2016} which we refer to is the most up-to date
+    compilation of LGM ages in and around the Alps, and deglacial outlines like
+    there exists for the Eurasian and Laurentide ice sheets are not available
+    for the Alps.
 
@@
 
+    Thank you. Indeed the quality and availability of geomorphological and
+    geochronological data is very variable across the Alps, and long-standing
+    interpretations are sometimes proven wrong
+    \citep[e.g.,][]{Monegato.etal.2007}. We appreciate your opinion, but the
+    parts of the Alpine community sees our challenging of the trimline
+    assumption as quite provocative given that it has been around for at least
+    a century \citep{Penck.Bruckner.1909}.
+
     \referee{\textbf{p.~21, l.~11--19:}
@@
 
-    \referee{%
-        Additional references: Ballantyne, C.K. and Stone, J.O., 2015.
-        Trimlines, blockfields and the vertical extent of the last ice sheet in
-        southern Ireland. Boreas, 44(2), pp.277- 287. Kuchar, J., Milne, G.,
-        Hubbard, A., Patton, H., Bradley, S., Shennan, I. and Edwards, R.,
-        2012. Evaluation of a numerical model of the British–Irish ice sheet
-        using relative sea-level data: implications for the interpretation of
-        trimline observations.  Journal of Quaternary Science, 27(6),
-        pp.597-605.}
-
 % ----------------------------------------------------------------------

### 2018-07-23 figures/alpcyc_lr_footprints.py

@@ -45,7 +45,7 @@ for i, rec in enumerate(ut.alpcyc_records):
         mask = (thk[b0:b1] < 1.0).prod(axis=0)
         cs = ax.contourf(x, y, mask, levels=[-0.5, 0.5], colors=[c], alpha=0.75)
         ut.pl.add_corner_tag('MIS %d' % (2+2*j), ax=ax, va='bottom')
-        ut.pl.add_subfig_label('(%s)' % list('abcdef')[i/2+j], ax=ax)
+        #ut.pl.add_subfig_label('(%s)' % list('abcdef')[i/2+j], ax=ax)
         ut.pl.draw_boot_topo(ax, res='2km')
         ut.pl.draw_natural_earth(ax)
 

### 2018-07-23 articles/cycle/response-referee1.tex
Reread the responses to referees 1, 2, and 3.

@@
 \title{Authors' response to Anonymous Referee \#1}
-\author{J.~Seguinot et al.}
+\author{J.~Seguinot, on behalf of all authors.}
 %\date{}
@@
 
-    Thank you very much for this elogious summary.
+    Thank you very much for these elogious and supportive comments!
 
@@
 
+    Thank you for bringing this up.
     Laboratory experiments have demonstrated that the rheology of temperate,
@@
 
-    Although ice sheet models have previously ignored this effect, it has now
+    Ice sheet models have previously ignored this effect, but it has now
     become a typical component of polythermal models such as SICOPOLIS
@@
 
-    Water fractions between 1 and 5\,\% have repreatedly been observed in
-    temperate glaciers \citep{Murray.etal.2000, Murray.etal.2007,
+    However water fractions between 1 and 5\,\% have repeatedly been observed
+    in temperate glaciers \citep{Murray.etal.2000, Murray.etal.2007,
     Bradford.Harper.2005, Bradford.etal.2009}, and also occur in model results
@@
     above 1\,\%. This arbitrary threshold is not constrained at all, and the
-    urgent need for new ice deformation experiments has been pointed out
-    \citep{Kleiner.etal.2015}.
+    urgent need for new ice deformation experiments has already been pointed
+    out \citep{Kleiner.etal.2015}.
 
@@
     only occur within the basal temperate layer of the fastest-moving glaciers
-    where ice movement is dominated by basal sliding. We suspect that a local
-    increase of ice deformation there is negligible in comparaison to
-    uncertainties caused by basal sliding and more importantly climate forcing.
+    where ice movement is largely dominated by basal sliding. We suspect that
+    increased ice deformation in these regions is negligible in comparaison to
+    uncertainties related to basal sliding and, more importantly, to climate 
+    forcing.
 
-    We prefer to avoid including the above discussion in the manuscript, but we
-    have reworked the sentence on water content:
+    Thus, we prefer to avoid including the above discussion in the manuscript,
+    but have reworked the sentence on water content:
 
@@
         Cuffey.Paterson.2010}, an arbitrary threshold above which new
-        measurements are critically needed \citep{Kleiner.etal.2015}.}
+        ice deformation measurements are critically needed
+        \citep{Kleiner.etal.2015}.}
 
@@
     \msquote{%
-        A constant rheology was used for temperate ice containing more than
-        1\,\% of liquid water.}
+        In the absence of ice deformation measurements, a constant rheology was
+        used for temperate ice containing more than 1\,\% of liquid water.}
 
@@
     up-to-date description of PISM till effective pressure physics used in our
-    simulations \citet[Eqs.~18, 23, and 24]{Bueler.Pelt.2015}. However,
+    simulations \citep[Eqs.~18, 23, and 24]{Bueler.Pelt.2015}. However,
     subglacial water is not routed down the hydropotential gradient. We have
@@
 
-    In general, we chosed to not repeat parameter values from Table~1 in the
-    main text. The following text was added in the methods:
+    Although we refrain from repeating parameter values given in Table~1 in the
+    main text, the following text was added in the methods:
 
@@
     annual temperature lapse rates vary spatially between 5.4 and
-    5.8\,K\,km$^{-1}$ \citep{Rolland.2003}. The reference to
-    \citet{Rolland.2003} was added to Table~1, and as we have now made
-    explained in the manuscript, our constant value of 6\,K\,km$^{-1}$ is thus:
+    5.8\,K\,km$^{-1}$ \citep{Rolland.2003}. A reference to the study by
+    \citet{Rolland.2003} was added in Table~1, and as we have now explicited,
+    our constant value of 6\,K\,km$^{-1}$ is thus:
 
@@
     in the trimline region are typically between 15 and 20\,K below freezing.
-    In addition, this effect would nearly disappear near the base where the
+    Thus the effect on the temperature gradient would be small.  Actually, the
+    effect would even nearly disappear near the glacier base, where the
     temperature gradient is much steeper and primarily controlled by geothermal
@@
 
+    Thank you.
     We did not know about references on MIS~3 and have developped our sentence
@@
         Kjellstrom.etal.2010} but more precipitation south of the Alps during
-        MIS~2 LGM \citep{Strandberg.etal.2011, Ludwig.etal.2016}.}
+        MIS~2 \citep{Strandberg.etal.2011, Ludwig.etal.2016}.}
 
@@
         \citep[e.g.,]{Beaulieu.Reille.1992, Wohlfarth.etal.2008,
-        Duprat-Oualid.etal.2017}, and cave speleothem in the Eastern
+        Duprat-Oualid.etal.2017}, and cave speleothems in the Eastern
         \citep[e.g.,][]{Spotl.Mangini.2002, Boch.etal.2011} and Western
@@
 
-    We realise this is confusing. We had chosen this approach because temperate
-    basal ice tends to occur above the trimlines during periods of warmer
+    We realise that Fig.~6 is confusing. We had chosen the 1\,ka limit because temperate
+    basal ice tends to occur above the trimlines during short periods of warmer
     climate. To avoid confusion, Fig.~6c has been simplified to show the basal
     thermal boundary at the age of maximum ice thickness. The last sentence in
-    the caption now read:
+    the caption now reads:
 
@@
     Although the model output allows to study the migration of the basal
-    thermal boundary over time into more detail, the basal velocity, which
+    thermal boundary over time into more detail, the basal velocity, which also
     controls erosion, should probably be considered as well. We leave this for
@@
         The remaining discrepancies may relate to temporal migrations of the
-        basal thermal boundary, an absence of sliding in warm-based areas}
+        basal thermal boundary, an absence of sliding in warm-based areas [...]}
 
@@
 
-    Indeed, much of the mountainous areas reach maximum ice thickness during
-    early MIS~2 when ice is colder and thus stiffer. This is also visible on
+    In fact, much of the mountainous areas reach maximum ice thickness during
+    early MIS~2 when ice is colder and stiffer. This is also visible on
     Fig.~5. Some areas even reach maximum ice thickness during MIS~4 when the
     bedrock is slightly less depressed. There is certainly very much scope to
-    discuss the age of trimline, which may differ significantly from that of
-    the maximum ice extent in the lowland, but we are not aware of any data
+    discuss the age of the trimlines, which may differ significantly from that of
+    the maximum ice extent on the lowland, but we are not aware of any data
     that could be used for validation here. However, frozen areas plotted on
@@
         on small-scale highs or lows significantly different from their
-        ~km-scale surroundings, that could contribute to the discrepancies in
+        \~km-scale surroundings, that could contribute to the discrepancies in
         Fig.~6c.}
@@
     enthalpy gradient over a steep bedrock slope. This requires both high
-    vertical and horizontal grid resolutions. This limitation was mentioned
+    vertical and high horizontal grid resolutions. This limitation was mentioned
     in the section on trimlines:
@@
 
-        Done.
+        Done. We thank you very much again for the time and effort you put into
+        our manuscript.
 

### 2018-07-23 articles/cycle/response-referee2.tex
Address referee 3 comment on surge.
Reread the responses to referees 1, 2, and 3.

@@
 \title{Authors' response to Anonymous Referee \#2}
-\author{J.~Seguinot et al.}
+\author{J.~Seguinot, on behalf of all authors.}
 %\date{}
@@
 
-    Thank you very much for this accurate summary.
+    Thank you very much for this fully accurate summary.
 
@@
 
+    Thank you.
     We agree that a sensitivity study to ice physics, in particular basal
     sliding, is one of many improvements needed upon our results. However, for
-    the kind of exercise presented here, the sensitivity of to palaeo-climate
+    the kind of exercise presented here, the sensitivity to palaeo-climate
     history, which is virtually unknown and has many degrees of freedom, far
-    outweight the model sensitivity to ice physics. For this manuscript we have
-    chosen to focus on the primary source of sensitivity, which is climate.
+    outweight the model sensitivity to ice physics. From the perspective of
+    glacial geology, a 5\,K change in air temperature is more influential than
+    an order-of-magnitude increase in ice velocities (perhaps with the
+    exception of our conclusions on trimlines, see below). In its current
+    state, our manuscript studies the model sensitivity to the primary source
+    of uncertainty, which is climate.
 
-    Because the kind of simulations presented here depends on computing
+    Because the simulations presented here depend on computing
     resources only available through a peer-review application process, we are
     currently unable to perform an additional sensitivity study on basal
-    sliding for the entire Alps.  Regional sentivity studies are more
-    accessible. A regional sensitivity analysis on basal sliding is precisely
-    the topic of a follow-up study currently under preparation for the Rhine
+    sliding for the entire Alps. However, a regional study on the model
+    sentivity to basal sliding is actually
+    the topic for a follow-up study currently under preparation for the Rhine
     Glacier \citep{Imhof.etal.2017}, which has a relatively isolated catchment
@@
     Meanwhile, the following text was added in the section on trimlines, which
-    is certainly the mostly sensitive aspect of our results to basal sliding:
+    is certainly the most sensitive aspect of our results to basal sliding:
 
@@
 
-    Basal sliding and other sources of uncertainties in ice physics where also
-    mentioned in the conclusions:
+    In addition, basal sliding and other sources of uncertainties in ice
+    physics where also mentioned in the conclusions:
 
@@
         representative of wet till but weaker basal conditions may have applied
-        over saturated lake sediments where they occured. A constant rheology
-        was used for temperature ice containing more than 1\,\% of liquid
-        water.}
+        over saturated lake sediments where they occured. In the absence of ice
+        deformation measuremets, a constant rheology was used for temperature
+        ice containing more than 1\,\% of liquid water.}
 
@@
     so that it is already ``de-amplified'' before being used as temperature
-    forcing.  Arctic amplification could be an explanation if it was stronger
+    forcing. But still, Arctic amplification could be an explanation if it was stronger
     then than for changes that lead the LGM climate. Although we lack expertise
@@
     involve large amounts of freshwater discharge into the North Atlantic
-    implying a more regional mechanism than arctic amplification. Therefore, we
-    decided not mention arctic amplification in the text.
+    implying a more regional mechanism than Arctic amplification. To avoid
+    restrictive interpretations, we thus decided not to mention Arctic
+    amplification as a possible explanation.
 
@@
 
-    Done.
+    Thank you. Done.
 
@@
 
-    The correct formula can be found in \citep[p.~443]{Love.1906}. The lenght
-    scale of bedrock deformation can be computed as
+    The correct formula can be found in the study by \citet[p.~443]{Love.1906}.
+    The lenght scale of bedrock deformation can be computed as
     ${\alpha=\sqrt[4]{\frac{4D}{(\rho_m-\rho_l)g}}}$ \citep{Walcott.1970}, thus
     the error in lenght scale can be expressed as $\sqrt[4]{1+\nu}-1 = 5.7\%$.
-    The sentence was clarified and references were added.
+    The footnote was clarified and references were added.
 
@@
     our climate forcing scenarii.  Although our simulations show that even a
-    spatially homogeneous forcing produce spatial variations in the timing of
+    spatially homogeneous forcing produces spatial variations in the timing of
     maximum extent, on can not exclude that a different climate history might
@@
     timing of glacier advances onto the foreland depends on Alpine climate
-    history, which was may have been very different from the forcing used in
+    history, which may have been very different from the forcing used in
     our simulations.
@@
 
-    Thank you. The 200\,m contour levels were thickened on all figures.
+    Thank you. The 200\,m contour levels were thickened on all figures. We
+    thank you again very much for the time and effort you put into this very
+    meticulous review.
 
 % ----------------------------------------------------------------------
+% References
 

### 2018-07-23 articles/cycle/response-referee3.tex
Reread the responses to referees 1, 2, and 3.

@@
 \title{Authors' response to Anonymous Referee \#3}
-\author{J.~Seguinot et al.}
+\author{J.~Seguinot, on behalf of all authors.}
 %\date{}
@@
 
-Thank you very much for this supportive comment.
+    Thank you very much for these supportive words!
 
@@
 
-        Corrected.
+    Thank you. Corrected.
 
@@
         modelled as a time-transgressive event, with some glaciers reaching
-        their maximum as early as 26\,ka, and others as late as 20\,ka.}
+        their maximum as early as 27\,ka, and others as late as 21\,ka.}
 
@@
 
-    This reference is indeed relevant to the discussion on trimlines. We have
-    added the reference in the introduction and in the corresponding section in
-    the discussion.
+    Thank you. This is relevant indeed. We have added the reference in the
+    introduction and in the corresponding section of the discussion.
 
@@
 
-    In order to clarify our statement of intent, the following text was added:
+    Thank you.
+    In order to clarify our statement of intent, we added the following text:
 
@@
     ice-filled valleys. The precipitation pattern, on the other hand, remains
-    the same throughout the simulation, and this results in higher accumulation
-    above the location of present-day mountains even if they are buried by ice
+    the same throughout the simulation, and this results in a higher accumulation
+    above the location of present-day mountains even as they are buried by ice
     in the model.
@@
     \msquote{%
-        During MIS 2, all six simulations yield comparative cumulative
-        glaciated areas modelled maximum extents (Fig.~3a).}
+        During MIS 2, all six simulations yield comparative modelled maximum
+        extents (Fig.~3a).}
 
@@
 
-    Yes, parts of the reconstructed LGM ice limits are uncertain due to the
-    lack of moraines or dated evidence. However, the discrepancies in ice
-    extent discussed here are beyond uncertainties.
+    Yes, this is possible. Parts of the reconstructed LGM ice limits are
+    uncertain due to the lack of moraines or dated evidence. However, without
+    very detailed knowledge about palaeoclimate, and tight constraints on ice
+    physics, the model won't replace observations. Actually, we are very
+    confident that the model-data discrepancies in ice extent discussed in this
+    passage are beyond uncertainties.
 
@@
     been made available that would allow such a comparison. The review by
-    \citep[Fig.~5]{Wirsig.etal.2016} which we refer to is the most up-to date
+    \citet[Fig.~5]{Wirsig.etal.2016} which we refer to is the most up-to date
     compilation of LGM ages in and around the Alps, and deglacial outlines like
-    there exists for the Eurasian and Laurentide ice sheets are not available
-    for the Alps.
+    there exists for the Eurasian and Laurentide ice sheets are unfortunately
+    not available.
 
@@
     Our simulations reproduce a thermodynamical instability akin to a surge.
-    The internal energy (so-called entropy) conservation scheme used in PISM
+    The internal energy (so-called enthalpy) conservation scheme used in PISM
     allows for variations in ice temperature and liquid water content
@@
 
+    Thank you.
     This is a valid point. However we found an error in our figure: the data
-    plotted on the histogram (Fig.~3b) were different from that in the scatter
-    plot (Fig.~3a). These data corresponded to raw interpolated ice
+    plotted on the histogram (Fig.~3b) were different from those in the scatter
+    plot (Fig.~3a). These data were raw interpolated ice
     thicknesses, i.e. they missed a correction for altitude shifts between our
@@
     distribution of observations itself is skewed, with more trimlines observed
-    at higher elevations, and the differences between the modelled ice surface
-    and trimlines are slightly higher at higher elevations (not plotted because
-    correlation is poor).
+    at higher elevations, and there is a tendency for the differences between
+    the modelled ice surface and trimlines to increase with altitude.
 
@@
     locations around 2850\,m in the Aletsch Glacier (south of Jungfrau) region
-    where trimlines are well visible in the landscape. We prefer not to
-    use it because the number of trimlines mapped is not necessarily
-    proportional to the lenght, but may be instead depend on other factors such
+    where trimlines are well visible in the landscape. We prefer to stick with
+    the mean because the number of trimlines mapped is not necessarily
+    representative of their lenght, and may depend on other factors such
     as bedrock lithology and post-glacial erosion.
@@
     \msquote{%
-        Unfortunately, validation through the bedrock uplift rates
+        Unfortunately, validation through the bedrock uplift rate
         \citep[cf.][]{Kuchar.etal.2012} is not doable in the Alps due its lower
@@
     interpretations are sometimes proven wrong
-    \citep[e.g.,][]{Monegato.etal.2007}. We appreciate your opinion, but the
-    parts of the Alpine community sees our challenging of the trimline
-    assumption as quite provocative given that it has been around for at least
-    a century \citep{Penck.Bruckner.1909}.
+    \citep[e.g.,][]{Monegato.etal.2007}. Also, parts of the Alpine glacial
+    geology community will certainly disagree with your comment and find our
+    challenging of the trimline assumption quite provocative, given that it
+    has been around for at least a century \citep{Penck.Bruckner.1909}.
 
@@
 
-    Actually we mean geologically recorded asymmetry. This clearly needed
-    clarification. The corresponding sentence now reads:
+    Actually we meant geologically recorded asymmetry. Obviously, this needed
+    clarification. This sentence now reads:
 
@@
         (Sect.~4.1).}
+
+    Finally, we thank you again very much for the time and effort you put into
+    our manuscript.
 

### 2018-07-23 articles/cycle/alps-cycle.tex
Reread the responses to referees 1, 2, and 3.

@@
     as a time-transgressive event, with some glaciers reaching their maximum as
-    early as 26\,ka, and others as late as 20\,ka.
+    early as 27\,ka, and others as late as 21\,ka.
 
@@
     Lliboutry.Duval.1985, Cuffey.Paterson.2010}, an arbitrary threshold above
-    which new measurements are critically needed \citep{Kleiner.etal.2015}.
+    which new ice deformation measurements are critically needed
+    \citep{Kleiner.etal.2015}.
 
@@
     Kjellstrom.etal.2010} but more precipitation south of the Alps during MIS~2
-    LGM \citep{Strandberg.etal.2011, Ludwig.etal.2016}. Besides, changes in the
+    \citep{Strandberg.etal.2011, Ludwig.etal.2016}. Besides, changes in the
     ice surface topography may have redistributed orographic precipitation.
@@
     subglacial topography. Unfortunately, validation through the bedrock uplift
-    rates \citep[cf.][]{Kuchar.etal.2012} is not doable in the Alps due its
+    rate \citep[cf.][]{Kuchar.etal.2012} is not doable in the Alps due its
     lower values, active tectonics and uncertainties on geological properties
@@
     representative of wet till but weaker basal conditions may have applied
-    over saturated lake sediments where they occured. A constant rheology was
-    used for temperature ice containing more than 1\,\% of liquid water.
+    over saturated lake sediments where they occured. In the absence of ice
+    deformation measurements, a constant rheology was used for temperature ice
+    containing more than 1\,\% of liquid water.
 

### 2018-07-23 articles/cycle/response-monegato.tex
Reread the responses to referees 1, 2, and 3.
Address Giovanni's comment on precipitation.

@@
 \title{Authors' response to Giovanni Monegato}
-\author{J.~Seguinot et al.}
+\author{J.~Seguinot, on behalf of all authors.}
 %\date{}
@@
         manuscript especially regarding the Italian side of the Alps.}
+
+    Thank you very much for your supportive comments!
 
@@
         other causes, and not only precipitation, have to be considered.}
+
+    Thank you for this detailed analysis. We want to point out that WorldClim
+    is not a model, but is also based on observations. However we agree, that
+    the regional dataset by \citep{Isotta.etal.2013}, tailored to the dense
+    station network and steep topography of the Alps, may be a more robust
+    reference precitation forcing that needs to be considered in future
+    studies. A reference was added:
+
+    \msquote{%
+        Finally, modern precipitation data from WorldClim also bear
+        uncertainties and exhibit local disagreement with other regional data
+        \citep{Isotta.etal.2013}.}
+
+    Besides, the following sentence was added to the discussion of glacier
+    extent in the south-western Alps,
+
+    \msquote{%
+    However, geochronological data from the south-western Alps are sparse, and
+    the LGM extent compilation \citep{Ehlers.Gibbard.2004} is inconsistent with
+    more recent regional reconstructions \citep{Federici.etal.2017}.}
 

### 2018-07-23 articles/cycle/alps-cycle.tex
Address Giovanni's comment on precipitation.

@@
     perhaps enhanced during the LGM, as indicated by pollen reconstructions
-    \citep{Wu.etal.2007}. Finally, the LGM extent of Alpine glaciers might have
+    \citep{Wu.etal.2007}. The LGM extent of Alpine glaciers might have
     been affected by an east-west gradient in variables not accounted for
-    by the PDD model, such as cloud cover or dust deposition.
+    by the PDD model, such as cloud cover or dust deposition. Finally, modern
+    precipitation data from WorldClim also bear uncertainties and exhibit local
+    disagreement with other regional data \citep{Isotta.etal.2013}.
 
@@
     dampened by Mediterranean climate in this part of the model domain.
+    However, geochronological data from the south-western Alps are sparse, and
+    the LGM extent compilation \citep{Ehlers.Gibbard.2004} is inconsistent with
+    more recent regional reconstructions \citep{Federici.etal.2017}.
 

### 2018-07-23 articles/cycle/response-monegato.tex
Address Giovanni's comment on transfluences.

@@
         Straninger saddle I am a bit skeptic because it is not the lowermost
-        saddle C2(Plockenpass and Nassfeldpass are lower in elevation). The
+        saddle (Plockenpass and Nassfeldpass are lower in elevation). The
         central Adda glacier (and not the Ticino as at page 14 line 26) is much
@@
         with transfluences (yellow stars and red arrows).}
+
+    The amount of ice flowing through transfluences is certainly dependent on
+    model resolution and basal sliding parameters. However, we agree that this
+    will have little effect on the ice volume overestimation in the eastern
+    Alps in comparison to potential shortcomings in the climate forcing.
+
+    The tranfluence at the Straniger Saddle may be a byproduct of bedrock
+    topography aggregation to the 1\,km grid.  As a follow-up study, we have
+    begun to use higher model resolution to study the location of transfluences
+    in more detail \citep{Seguinot.etal.2018a}. In these results, the
+    transfluence at the Straniger Saddle looses importance relative to other
+    transfluences in the region including, in fact, Plöckenpass and
+    Nassfeldpass. Because the choice of transfluences plotted on Fig.~4 is
+    somewhat arbitrary, we have decided to update it.
 

### 2018-07-23 articles/cycle/response-monegato.tex
Address Giovanni's comment on basal sliding.

@@
 
+    Thank you. In fact basal sliding is one of the major uncertainties in our
+    results which is now better acknowledged for in our manuscript following
+    other reviewer's comments. However, basal sliding velocity are somewhat
+    decoupled from glacier front advance velocities and glacier front maximum
+    extent. Although this is not shown in our manuscript, increased basal
+    velocities tend to result in thinner ice tongues. Due to the
+    mass-balance-elevation feedback, this causes more surface melt and, in
+    turn, lesser extent. However this effect is small in comparison to changes
+    due to climate forcing.
+
     \referee{\textbf{4 --}

### 2018-07-23 articles/cycle/response-monegato.tex
Address Giovanni's comment on ice domes.

@@
 
+    By self-sustained ice domes, we mean ice domes that are not sitting on top
+    of basal topographic highs, but instead shift away from the modern
+    topographic divides and ``sustain'' themselves by reaching higher
+    elevation than the local mountains. We indicate these as a sign that ice
+    begins to behave as in an ice sheet, and flow in directions independent
+    from the local basal topography.
+
+    However, we agree that this is somewhat arbitrary, and again, as we found
+    out, dependent on model resolution \citep{Seguinot.etal.2018a}. Therefore,
+    we decided to remove the ``ice domes'' from Fig.~4a and the main text, and
+    save this discussion for future studies.
+
     \referee{\textbf{5 --}

### 2018-07-23 articles/cycle/response-monegato.tex
Address Giovanni's comment on polar front.

@@
 
+    We agree. In fact, our simulations are limited by the constant temperature
+    and precipitation patterns applied throughout the glacial cycle. Although
+    we can't answer this question without new climate and glacier simulations,
+    we have raised this important point in the very last paragraph:
+
+    \msquote{%
+        Shifts in the North Atlantic storm track and polar front may have
+        caused varied patterns of glaciations through different cold phases.}
+
+    With this, we would like to thank you again very much for the time and
+    effort you put into our manuscript.
+
 % ----------------------------------------------------------------------

### 2018-07-23 articles/cycle/response-monegato.tex
Reread response to Giovanni Monegato.

@@
     is not a model, but is also based on observations. However we agree, that
-    the regional dataset by \citep{Isotta.etal.2013}, tailored to the dense
-    station network and steep topography of the Alps, may be a more robust
-    reference precitation forcing that needs to be considered in future
+    the regional dataset by \citet{Isotta.etal.2013}, tailored to the dense
+    station network and steep topography of the Alps, is certainly a more
+    robust precitation forcing that needs to be considered in future
     studies. A reference was added:
@@
     Besides, the following sentence was added to the discussion of glacier
-    extent in the south-western Alps,
+    extent regarding the south-western Alps,
 
@@
     With this, we would like to thank you again very much for the time and
-    effort you put into our manuscript.
+    effort you have put into our manuscript.
 

### 2018-07-24 articles/cycle/alps-cycle.tex
Address Giovanni's comment on polar front.
Replace Straniger Alm transfluence in the text.

@@
         \citep{Lisiecki.Raymo.2005}.
-        \todo{Thicken the 200\,m contours on all figures.}}
+        \todo{Thicken the 200\,m contours on all figures. Remove ice domes and
+              replace Straniger.}}
       \label{fig:lgmvel}
@@
     instance across Fern Pass, the Seefeld Saddle, the Gailberg Saddle, the
-    Kreuxberg Saddle, the Straniger Alp, and Pyhrn Pass
+    Kreuzberg Saddle, Kronhof Pass, Studenbodenalm, and Pyhrn Pass
     (Fig.~\ref{fig:lgmvel}a). Transfluences across Fern Pass and the Seefeld
@@
     and Pyhrn Pass \citep[Fig.~2.5]{Husen.2011}.
-    On the other hand, no transfluence has previously been documented at the
-    Straniger Alp. The Tagliamento catchment is usually assumed to not have
+    On the other hand, no transfluence has previously been documented over the
+    Kronhof Pass, Studenbodenalm or elsewhere over the Karnic Alps.
+    The Tagliamento catchment is usually assumed to not have
     received tranfluences from the Drava catchment \citep{Monegato.etal.2007},
@@
     the European Alps may have caused more glaciations than modelled here.
+    Shifts in the North Atlantic storm track and polar front may have caused
+    varied patterns of glaciations through different cold phases.
     Using more specifically targeted sensitivity runs, and a more realistic

### 2018-07-24 articles/cycle/alps-cycle.tex
Remove text on self-sustained ice domes.

@@
     cycle. Ice flow patterns during peak glaciation are largely governed by
-    subglacial topography but include occasional transfluences and
-    self-sustained ice domes. Modelled maximum ice surface is 861\,m
+    subglacial topography but include occasional transfluences through the
+    mountain passes. Modelled maximum ice surface is 861\,m
     higher than observed trimline elevations in the upper Rhone Valley, yet
@@
         geomorphological reconstruction \citep[solid red
-        line,][]{Ehlers.etal.2011}, modelled self-sustained ice domes
-        (triangles), and some major transfluences (crosses) are shown.
+        line,][]{Ehlers.etal.2011}, and some major transfluences across
+        mountain passes (crosses) are shown.
         \textbf{(b)} Temperature offset time-series from the EPICA ice core
@@
     elevations in this area \citep{Husen.1987}.
-
-    Finally, self-sustained ice domes characteristic of ice caps are modelled
-    in two locations over Flüela Pass and Ötztal (Fig.~\ref{fig:lgmvel}a).
-    These modelled ice domes exhibit radial ice surface slopes and radial flow
-    velocities crosscutting the underlying topographic troughs and ridges.
-    On the one hand, ice-dispersal centres previously labelled as ice domes
-    in the upper Rhone and Rhine Valley \citep{Florineth.Schluchter.1998,
-    Kelly.etal.2004, Bini.etal.2009} feature as mere ice divides in our
-    simulation. On the other hand, the modelled ice dome over Flüela Pass
-    overlaps with the previously mapped Engadine ice dome
-    \citep{Florineth.1998}, although its centre is modelled ca.~20\,km north of
-    that indicated by geomorphology \citep[Fig.~3]{Florineth.1998}.
 
@@
       \item The LGM ice flow pattern in the Alps was largely controlled by
-            subglacial topography, but several transfluences and two
-            self-sustained ice domes may have occurred (Sect.~\ref{sec:flow}).
+            subglacial topography, but transfluences across several mountain
+            passes may have occurred (Sect.~\ref{sec:flow}).
       \item The LGM (maximum) extent was a transient stage in which glaciers

### 2018-07-24 articles/cycle/alps-cycle.tex
Add maximum ice thickness and remove todo notes.

@@
         used to compute surface mass balance. The background maps contain
-        Natural Earth Data \citep{Patterson.Kelso.2017}.
-        \todo{(R2) add scale and mark inset.}}
+        Natural Earth Data \citep{Patterson.Kelso.2017}.}
       \label{fig:inputs}
@@
         a~global compilation of benthic \chem{\delta^{18}O} records
-        \citep{Lisiecki.Raymo.2005}.
-        \todo{Thicken the 200\,m contours on all figures. Remove ice domes and
-              replace Straniger.}}
+        \citep{Lisiecki.Raymo.2005}.}
       \label{fig:lgmvel}
@@
         is here modelled as a time-transgressive event. In much of the mountain
-        area, maximum thickness is reached before 27\,ka.
-        \todo{fix colourbar label}}
+        area, maximum thickness is reached before 27\,ka.}
       \label{fig:timing}
@@
     difference of $861\pm197$\,m (one standard deviation,
-    Fig.~\ref{fig:trimlines}b), wich is significant in respect to the modelled
-    regional ice thickness \todo{add a number}. The result is somewhat
-    dependent on uncertain basal sliding and ice rheological parameters
+    Fig.~\ref{fig:trimlines}b), which is significant in respect to the modelled
+    maximum ice thickness in the Rhone Valley (and the Alps) of 2586\,m. This
+    result depends on uncertain basal sliding and ice rheological parameters
     to which the model sensitivity was not tested here. However, a previous

### 2018-07-24 articles/cycle/alps-cycle.tex
Apply minor fixes before sending to co-authors.

@@
 
-    The climate forcing driving the ice sheet simulations consist spatially of
+    The climate forcing driving the ice sheet simulations consists spatially of
     a~present-day monthly mean climatology,
-    $\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, modified by temperature lapse-rate
+    $\{T_{\mathrm{m}0}, P_{\mathrm{m}0}\}$, modified by a temperature lapse-rate
     correction, ${\Delta}T_{\text{LR}}$, temperature
@@
     %
-    with $\psi=0.169/2.4=0.0704$ \citep{Huybrechts.2002}. The control set of
+    with $\psi=0.169/2.4=0.0704$ \citep{Huybrechts.2002}. The control
     simulations use constant precipitation, corresponding to $\psi=0$.
@@
     In the upper Rhone Valley, observed trimlines are often located near the
-    LGM cold-temperate basal thermal transition, or in cold-based areas
-    (Fig.~\ref{fig:trimlines}c). The remaining discrepancies may relate to
+    LGM cold-temperate basal thermal transition or within cold-based areas
+    (Fig.~\ref{fig:trimlines}c). The remaining discrepancies between the
+    trimlines locations and the modelled basal thermal boundary may relate to
     temporal migrations of the basal thermal boundary, an absence of sliding in

### 2018-07-27 articles/cycle/alps-cycle.tex
Implement comments from Guillaume.

@@
     subglacial topography but include occasional transfluences through the
-    mountain passes. Modelled maximum ice surface is 861\,m
+    mountain passes. Modelled maximum ice surface is in average 861\,m
     higher than observed trimline elevations in the upper Rhone Valley, yet
@@
     date, it is supported by numerical modelling
-    \citep[Fig.~3]{Haeberli.Schluchter.1987}. Alpine ice flow
+    \citep[Fig.~3]{Haeberli.Schluchter.1987, Cohen.2017}. Alpine ice flow
     patterns were primarily controlled by subglacial topography, but there is
@@
     Seguinot.etal.2016} and subsequently adapted for steady climate
-    \citep{Becker.etal.2016} and regional \citep{Jouvet.etal.2017a}
-    applications to the European Alps.
+    \citep{Becker.etal.2016} and regional \citep{Jouvet.etal.2017a,
+    Becker.Becker.etal.2017} applications to the European Alps.
 
@@
     dynamics over mountainous topographies similar to that of the European Alps
-    \citep{Golledge.etal.2012, Ziemen.etal.2016}. This hybrid scheme is
-    computationally much more efficient than a fully three-dimensional model
-    \citep[e.g.,][]{Cohen.etal.2017}
-    with which simulations of the scale presented here would not be feasible.
+    \citep{Golledge.etal.2012, Ziemen.etal.2016}. In the SIA, topographic
+    roughness using a bed smoother range of 5\,km \citep{Schoof.2003}.
 
@@
         (colour) and LGM ice surface elevation (200\,m contours). Hatches mark
-        the LGM cold-based ares (basal temperature above $1 \times 10^{-3}$\,K
+        the LGM cold-based areas (basal temperature above $1 \times 10^{-3}$\,K
         below freezing at the age of maximum ice thickness).}
@@
     observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
-    difference of $861\pm197$\,m (one standard deviation,
+    difference of 861\,m (and a standard deviation 197\,m,
     Fig.~\ref{fig:trimlines}b), which is significant in respect to the modelled
@@
     \citep{Becker.etal.2017}.
+    \todo{Add \citep{Cohen.etal.2018}, and more?}
 

### 2018-07-30 articles/cycle/alps-cycle.tex
Address private comments from Wilfried Haeberli.

@@
     no evidence for cold-based glaciation has been reported in the Alps to
-    date, it is supported by numerical modelling
-    \citep[Fig.~3]{Haeberli.Schluchter.1987, Cohen.2017}. Alpine ice flow
+    date, it is supported by numerical modelling \citep{Blatter.Haeberli.1984,
+    Haeberli.Schluchter.1987, Cohen.etal.2017}. Alpine ice flow
     patterns were primarily controlled by subglacial topography, but there is
@@
     \citep{Becker.etal.2016} and regional \citep{Jouvet.etal.2017a,
-    Becker.Becker.etal.2017} applications to the European Alps.
+    Becker.etal.2017} applications to the European Alps.
 
@@
     result depends on uncertain basal sliding and ice rheological parameters
-    to which the model sensitivity was not tested here. However, a previous
-    sensitivity study conducted with PISM and a very similar model set-up on
-    the Cordilleran ice sheet show that varying ice rheological parameters
-    resulted in relative errors in modelled total ice volume of 25\,\unit{\%}
-    with regard to ice rheological parameters and 14\,\unit{\%} regarding basal
-    sliding parameters \citep[Fig.~7]{Seguinot.etal.2016}. In addition,
-    regional sensitivity tests in the Rhone Valley show that a modelled ice
-    surface compatible with the trimline elevations is incompatible with the
-    documented overspilling flow to the south across the Simplon pass
-    \citep{Becker.etal.2017}.
-    \todo{Add \citep{Cohen.etal.2018}, and more?}
+    to which the model sensitivity was not tested here. Unfortunately,
+    validation through the bedrock uplift rate \citep[cf.][]{Kuchar.etal.2012}
+    is not doable in the Alps due its lower values, active tectonics and
+    uncertainties on geological properties \citep[cf.][]{Mey.etal.2016}.
+    However, regional sensitivity tests show that a modelled surface compatible
+    with the trimline elevations is incompatible with the documented
+    overspilling flow across the Simplon Pass \citep{Becker.etal.2017}, and LGM
+    palaeo-climate proxy records \citep{Cohen.etal.2017}.
 
@@
     relevant basal sliding and ice rheological parameters, and to the uncertain
-    subglacial topography. Unfortunately, validation through the bedrock uplift
-    rate \citep[cf.][]{Kuchar.etal.2012} is not doable in the Alps due its
-    lower values, active tectonics and uncertainties on geological properties
-    \citep[cf.][]{Mey.etal.2016}. However, the presence of an upper layer of
+    subglacial topography. However, the presence of an upper layer of
     cold ice during the LGM, already found at high altitude in the much warmer
     climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
-    inevitable \citep[Fig.~3]{Haeberli.Schluchter.1987}.
+    inevitable \citep{Blatter.Haeberli.1984, Haeberli.Schluchter.1987,
+    Cohen.etal.2017}. The bedrock themal model yield significant volumes of
+    frozen ground periglacially north of the Alps and subglacially under the
+    highest mountains, which is also in agreement with previous studies
+    \citep{Haeberli.etal.1984, Lindgren.etal.2016, Cohen.etal.2017}.
+
     In this context, our results challenge the assumption that

### 2018-07-30 articles/cycle/alps-cycle.tex
Implement comments from Susan.

@@
     in north and west of the Alps \citep[e.g.,]{Beaulieu.Reille.1992,
-    Wohlfarth.etal.2008, Duprat-Oualid.etal.2017}, and cave speleothem in the
+    Wohlfarth.etal.2008, Duprat-Oualid.etal.2017}, and cave speleothems in the
     Eastern \citep[e.g.,][]{Spotl.Mangini.2002, Boch.etal.2011} and Western
@@
     foreland by 21.4\,ka. The MD01-2444 palaeo-temperature forcing yields
-    a late LGM peak ice volume at 16.5/15.7\,ka followed by a rapid retreat
-    several thousand years after the dated geological evidence.
+    a late LGM peak ice volume at 16.5/15.7\,ka followed by a rapid retreat,
+    in contrast to dated records of large-scale ice retreat by 17 to
+    16\,ka\citep{Ivy-Ochs.etal.2006, Ivy-Ochs.etal.2008}.
 
@@
     by geologic evidence \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
-    Dansgaard-Oeschger events, however, have been recorded in Central European
+    Dansgaard-Oeschger events, however, have been recorded in European
     lake sediments \citep{Wohlfarth.etal.2008} and cave speleothems
     \citep{Spotl.Mangini.2002, Luetscher.etal.2015}. Our results indicate that
-    their magnitude in Central Europe was likely smaller than that recorded in
+    their magnitude in Europe was likely smaller than that recorded in
     GRIP \chem{\delta^{18}O}, which either implies smaller associated
-    palaeo-temperature fluctuations in Central Europe than Greenland, or
+    palaeo-temperature fluctuations in Europe than Greenland, or
     temperature-independent controls on GRIP \chem{\delta^{18}O}. In contrast,
@@
     The LGM extent of Alpine glaciers has been mapped with varying level of
-    detail across the Western \citep{Coutterand.2010, Buoncristiani.Campy.2011,
-    Jackli.1962, Bini.etal.2009, Hantke.2011}, and Eastern Alps
+    detail across the Western \citep{Jackli.1962, Bini.etal.2009,
+    Coutterand.2010, Buoncristiani.Campy.2011, Hantke.2011}, and Eastern Alps
     \citep{Penck.Bruckner.1909, Castiglioni.1940, Husen.1987, BGR.2007,
@@
     On the other hand, the model yields excessive ice
-    cover in the eastern Alps, where the Mur, Drava and Sava Glaciers are
+    cover in the Eastern Alps, where the Mur, Drava and Sava Glaciers are
     modelled to extend tens of km beyond the mapped ice limits
@@
     erratics across Brünig Pass \citep{Jackli.1962}. On the other hand, the
-    simplified climate forcing used in this sumulation could not reproduced
+    simplified climate forcing used in this sumulation could not reproduce
     the transport of glacial erratics from southern Valais to observed
@@
     glacier dynamics controlled by spatial variations in catchment size and
-    hypsometry. Importantly, Dansgaard-Oeschger events, recorded in Central
+    hypsometry. Importantly, Dansgaard-Oeschger events, recorded in
     Europe \citep{Spotl.Mangini.2002, Wohlfarth.etal.2008,
@@
     different palaeo-precipitation parametrizations (with and without
+    ca.~-68\,\unit{mm\,{\degree}C^{-1}}
     precipitation reductions; Sect.~\ref{sec:palprec}), we find that only the
@@
 
-    This suprising result is likely coincidental as there is no direct link
+    This interesting result may be coincidental as there is no direct link
     between European and Antarctic climate. This highlights the need for more

### 2018-07-31 articles/cycle/alps-cycle.tex
Update author list and affiliations.

@@
 
-\Author[1]{Julien}{Seguinot}
+\Author[1,2]{Julien}{Seguinot}
+\Author[3]{Susan}{Ivy-Ochs}
 \Author[1]{Guillaume}{Jouvet}
@@
 \Author[1]{Martin}{Funk}
-\Author[2]{Susan}{Ivy-Ochs}
-\Author[3]{Frank}{Preusser}
+\Author[4]{Frank}{Preusser}
 
@@
           ETH Zürich, Switzerland}
-\affil[2]{Laboratory of Ion Beam Physics, ETH Zürich, Switzerland}
-\affil[3]{Institute of Earth and Environmental Sciences,
+\affil[2]{Arctic Research Center, Hokkaido University, Sapporo, Japan}
+\affil[3]{Laboratory of Ion Beam Physics, ETH Zürich, Switzerland}
+\affil[4]{Institute of Earth and Environmental Sciences,
           University of Freiburg, Germany}

### 2018-07-31 articles/cycle/alps-cycle.tex
Update acknowledgements and data availability.

@@
 \codedataavailability{%
-    PISM is available as open-source software at http://pism-docs.org/. Upon
-    publication, we will be happy to make aggregated model output available
-    through The Cryosphere. Please contact the corresponding author for more
-    detailed model output data.}
+    PISM is available as open-source software at http://pism-docs.org/.
+    Aggregated model output will be made available upon publication. Please
+    contact the corresponding author for more detailed model output data.}
 
@@
     J.~Seguinot designed the study, ran the simulations, and wrote most of
-    the manuscript. M.~Huss provided modern ice thickness data. All authors
-    contributed to interpreting results and improving the text. The idea for
-    this study stems in part from an excursion organised by F.~Preusser in the
-    central Alps in 2012.}
+    the manuscript. All authors contributed to interpreting the results and
+    improving the text. M.~Huss provided modern ice thickness data. The idea
+    for this study stems in part from an excursion organised by F.~Preusser in
+    the central Alps in 2012.}
 
 \competinginterests{%
-    We have no conflict of interest in publishing these results.}
+    The authors declare that they have no conflict of interest.}
 
@@
 
-    This manuscript is partly based on a previous study on the Cordilleran Ice
-    Sheet which is part of J.~Seguinot's PhD thesis. The experimental design
-    used here, as well as the general layout of this study, are largely
-    inspired by Irina Rogozhina and Arjen~P. Stroeven. We are very thankful to
-    Constantine Khroulev, Ed Bueler, and Andy Aschwanden for providing constant
-    help and development with PISM, in particular with recent issues with the
-    computation of ice temperature (Github issue no.~371) and with the
-    computation of bedrock deformation (Github issues no.~370 and 377). We are
-    equally thankful to Giovanni Monegato and Marc Luetscher for very
-    insightful discussions of preliminary model results during the European
-    Geoscience Union General Assembly~2017. This work was supported by the
-    Swiss National Foundation grant no.~200021-153179/1 to M.~Funk. Computer
-    resources were provided by the Swiss National Supercomputing Centre (CSCS)
-    allocation no.~s573 to J.~Seguinot.
+    We are very thankful to Constantine Khroulev, Ed Bueler, and Andy
+    Aschwanden for providing constant help and development with PISM, in
+    particular with recent issues with the computation of ice temperature
+    (Github issue no.~371) and with the computation of bedrock deformation
+    (Github issues no.~370 and 377). We are equally thankful to three
+    anonymous reviewers and Giovanni Monagato for very constructive inputs
+    during peer-review, and to Marc Luetscher for insightful discussions of
+    preliminary results during the EGU~General Assembly~2017. The experimental
+    design and article layout used here are based on previous work which
+    received much guidance from Irina Rogozhina and Arjen~P. Stroeven.
+    The current work was supported by the Swiss National Science Foundation
+    grants no.~200020-169558 and 200021-153179/1 to M.~Funk. Computer resources
+    were provided by the Swiss National Supercomputing Centre (CSCS)
+    allocations no.~s573 and sm13 to J.~Seguinot.
 

### 2018-07-31 articles/cycle/response-editor.tex
Add reponse to the editor.

% response-editor.tex
% ----------------------------------------------------------------------
\input{response-header}

\title{Authors' response to the Editor}
\author{J. Seguinot, on behalf of all authors.}
%\date{}

\begin{document}
\maketitle
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

    Dear Andreas Vieli,

    We apologize for delays accumulated during the peer review of our
    manuscript. We believe that we have addressed all points raised by the
    reviews and hereby submit our revised version to \emph{The Cryosphere}.
    Please find hereafter a short summary and a marked-up file listing
    changes made to the manuscript since its first publication in \emph{The
    Cryosphere Discussion}. Please refer to our public reponses for more
    detail explanations.

    \begin{itemize}

        \item \textbf{Sect.~2 (Ice sheet model set-up):}
        We have made explicit the lack of measurements on ice softness for
        water contents above 1\,\%, clarified that subglacial water is not
        transported, explained the effect of erroneous elastic thickness, and
        justified our value for the air temperature lapse-rate.

        \item \textbf{Sect.~3 (Palaeo-climate forcing):}
        We have provided a short review on available palaeo-climate proxy
        records in or near the Alps, summarized the state of knowledge on past
        precipitation changes, and pointed out inconsistencies between our
        reference climate forcing and other data.

        \item \textbf{Sect.~4 (Results and discussion):}
        We have pointed out inconsistent LGM ice extent reconstructions in the
        south-western Alps, removed the discussion on self-sustained ice domes,
        and discussed uncertainties regarding our conclusion on trimlines.
        \emph{New references were added following private comments from
              Christopher~Carcaillet and Wilfried~Haeberli}.

        \item \textbf{Sect.~5 (Conclusions):}
        We have added a word of caution regarding our conclusion on EPICA being
        the optimal forcing, listed major sources of uncertainties, and
        emphasized the need for more realistic climate forcing in future
        studies.

        \item \textbf{Fig.~1:}
        We have removed self-sustained ice domes and updated our somewhat
        arbitrary choice of transfluence locations.

        \item \textbf{Fig.~6:}
        We have corrected histogram data and replaced areas modelled to have
        experienced temperate ice for less than 1\,ka by frozen-based areas
        at the time of maximum ice thicknes.

        \item \textbf{Various:}
        \emph{We found that thicker 200\,m contour lines were disturbing other
              figure elements and eventually decided to keep original thickness
              inconsistently with our response to referee~\#2.}
        We have implemented numerous suggestions for text improvements.

    \end{itemize}

    We thank you very much again for your editorial work for our manuscript.

\end{document}

### 2018-07-31 articles/cycle/alps-cycle.tex
Add refs on Alpine tree refugia.
Add reponse to the editor.

@@
     uncertainties on geological properties \citep[cf.][]{Mey.etal.2016}.
+    Interestingly, tree and fire remains evidence nunataks in the south-western
+    Alps by 21\,ka \citep{Carcaillet.Blarquez.2017, Carcaillet.etal.2018}.
     However, regional sensitivity tests show that a modelled surface compatible
@@
     (Github issues no.~370 and 377). We are equally thankful to three
-    anonymous reviewers and Giovanni Monagato for very constructive inputs
-    during peer-review, and to Marc Luetscher for insightful discussions of
+    anonymous reviewers, Giovanni Monagato, and the editor Andreas Vieli for
+    their careful read and constructive inputs during peer-review.
+    We thank Marc Luetscher for an insightful discussion of
     preliminary results during the EGU~General Assembly~2017. The experimental

### 2018-07-31 articles/cycle/alps-cycle.tex
Make last minor changes before resubmission.

@@
     in contrast to dated records of large-scale ice retreat by 17 to
-    16\,ka\citep{Ivy-Ochs.etal.2006, Ivy-Ochs.etal.2008}.
+    16\,ka \citep{Ivy-Ochs.etal.2006, Ivy-Ochs.etal.2008}.
 
@@
     late-glacial ice surface elevation \citep[Fig.~1, p.~403]{Coutterand.2010}.
-    However, no such evidence has been reported in the Alps.
+    However, no such evidence has been reported in the Alps. Instead, tree and
+    fire remains record the presence of nunataks in the south-western Alps by
+    21\,ka \citep{Carcaillet.Blarquez.2017, Carcaillet.etal.2018}.
 
@@
     result depends on uncertain basal sliding and ice rheological parameters
-    to which the model sensitivity was not tested here. Unfortunately,
+    to which the model sensitivity was not tested here.
+    However, regional model sensitivity tests show that a modelled surface
+    compatible with the trimline elevations is incompatible with the documented
+    overspilling flow across the Simplon Pass \citep{Becker.etal.2017}, and LGM
+    palaeo-climate proxy records \citep{Cohen.etal.2017}. Unfortunately,
     validation through the bedrock uplift rate \citep[cf.][]{Kuchar.etal.2012}
@@
     uncertainties on geological properties \citep[cf.][]{Mey.etal.2016}.
-    Interestingly, tree and fire remains evidence nunataks in the south-western
-    Alps by 21\,ka \citep{Carcaillet.Blarquez.2017, Carcaillet.etal.2018}.
-    However, regional sensitivity tests show that a modelled surface compatible
-    with the trimline elevations is incompatible with the documented
-    overspilling flow across the Simplon Pass \citep{Becker.etal.2017}, and LGM
-    palaeo-climate proxy records \citep{Cohen.etal.2017}.
 
@@
 
-    \begin{figure*}[t]
+    \begin{figure*}[p]
       \centerline{\includegraphics{alpcyc_hr_profiles}}
@@
     anonymous reviewers, Giovanni Monagato, and the editor Andreas Vieli for
-    their careful read and constructive inputs during peer-review.
+    their meticulous readings and constructive inputs during peer-review.
     We thank Marc Luetscher for an insightful discussion of

### 2018-08-16 articles/cycle/response-editor1.tex
Rename editor responses with numbers.

% response-editor.tex
% ----------------------------------------------------------------------
\input{response-header}

\title{Authors' response to the Editor}
\author{J.~Seguinot et al.}
\date{February 15, 2018}

\begin{document}
\maketitle
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

\newcommand{\referee}[2][]{\bigskip\noindent\textcolor{darkblue}{\textbf{#1:} #2}}
\newcommand{\msquote}[1]{\begin{quote}\textit{#1}\end{quote}}
\newcommand{\doi}[1]{doi:\allowbreak\href{http://dx.doi.org/#1}{#1}}

Dear Andreas Vieli,

Thank you very much for your feedback and support for our manuscript.

    \referee[p.~1, l.~6]{%
        to be correct it should probably say: '...resolution of 1 to 2 km'.}

    We rephrased to ``using horizontal resolutions of 2 and 1\,km and up to
    576~processors''.

    \referee[p.~2, l.~12]{%
        '...indicating SEVERAL tens of ...'. And change 800 000 years to 800 ka
        (and delete the bracket (800ka) on next line.}

    Done.

    \referee[p.~5, Fig.~1c caption]{%
        maybe say that this is at the 1km resolution used in the model.}

    Good idea, we've added this precision.

    \referee[p.~11, l.~6]{%
        '...by a east-west...'}

    Thanks, we've corrected.

    \referee[p.~15]{%
        I am just wondering and maybe this is to be considered in a later stage
        after a first round of reviews: in the section 4.2 i wonder if you
        should not briefly consider the discrepancy of the erratics from the
        southern Valais that reached the solothurner lobe... (work of Jouvet et
        al). it seems with your modelling you do not get Valais ice to flow to
        Solothurn (only aare valley)}

    Although not entirely obvious from the figures this is a correct
    observation. We've added a sentence about this.

    \referee[p.~18, Fig.~6 caption l.~4]{%
        should it not say 50m bands (instead of 500m bands)?}

    That is right, we've corrected.

    \referee[p.~18, l.~5--7]{%
        should you not mention the work by Becker here (his Phd work), as he
        also did preliminary sensitivity studies exploring a range of factors
        parameters (including sliding/rheology) for a part of the alpine ice
        sheet. Or am I wrong here?}

    We've added a sentence about that.

\end{document}

### 2018-08-16 articles/cycle/response-editor3.tex
Paper accepted! Initiate third editor response.

% response-editor.tex
% ----------------------------------------------------------------------
\input{response-header}

\title{Authors' response to the Editor}
\author{J. Seguinot, on behalf of all authors.}
%\date{}

\begin{document}
\thispagestyle{empty}
\maketitle
\bigskip

% ----------------------------------------------------------------------
% Interactive comment text begins
% ----------------------------------------------------------------------

    Dear Andreas Vieli,

    \newcommand{\referee}[1]{\bigskip\noindent\textcolor{journalname}{#1}}

    \referee{\textbf{p.~1, l.~14-17:}
        somewhat odd phrasing: `...cause ...to be modelled...', maybe
        `...produces in the model...' or similar would be better.}

    \referee{\textbf{p.~3, l.~4:}
        `it still remains incompletely known' is awkward phrasing, maybe better
        say: `...knowledge remains incomplete...'}

    \referee{\textbf{p.~3, l.~7-8:}
        the added sentence starting with `Additional...' seems slightly out of
        place here (or the flow is someaht disrupted). Maybe better move it to
        end of nextz paragraph (p.~3, l.~14).}

    \referee{\textbf{p.~3, l.~9:}
        starting a paragraph with a `But...' is awkward, maybe change back to
        `Here we...' in particular when you move the sentence before.}

    \referee{\textbf{p.~3, l.~32:}
        would `different' instead of `distinct' not be more appropriate.}

    \referee{\textbf{p.~4, l.~8:}
        `In the SIA, topographic roughness using a bed smoother range of 5 km
        (Schoof, 2003).' I really do not understand this sentence (what is
        message), nor its connection to the sentence before.}

    \referee{\textbf{p.~4, l.~24:}
        `...without horizontal transport...'?}

    \referee{\textbf{p.~4, l.~25:}
        maybe simplify to `...is assumed to drain instantaneously and is
        removed....'}

    \referee{\textbf{p.~7, l.~3:}
        `above average' of what? What is reference?}

    \referee{\textbf{p.~14, fig.~4:}
        are the surface contours really in 200m intervals?, that would make the
        ice surface only a bit over 1200m high. Or do you leave some out?  You
        should add some labels to the surface contour lines so one can judge
        the surface elevation better.}

    \referee{\textbf{p.~13, l.~30 -- p.~14, l.~3:}
        but jouvet (TC) claims (based on tracking erratics) that north south
        contrast in in precip is needed. Is this still consistent with this
        work? Maybe should be included. This is mentioned on next page but
        would be relevant here as well.}

    \referee{\textbf{p.~14, fig.~4 caption:}
        for `(b)' it should say what the vertical black line (time of max
        extent in (a) at 24.57 ka???) refers to.}

    \referee{\textbf{p.~15, l.~31:}
        `simulation' (not sumulation)}

    \referee{\textbf{p.~16, l.~12:}
        `Fig. 5' has nothing to do with `reviews', what do you mean with `Fig.
        5 for reviews'???}

    \referee{\textbf{p.~19, l.~15-19:}
        in this context of sensitivity to basal sliding parameter (and lacking
        exploration) maybe some reference to the Phd-thesis of Becker which
        includes at least some preliminary (mostly steady state) sliding
        investigation could be made.}

    \referee{\textbf{p.~21, l.~19:}
        `last major Alpine reliefs' ...? Do you mean the `last foothills' of the
        alps? Clarify.}

    \referee{\textbf{p.~23, l.~1:}
        is it not rather `glacier physics' as the till is not ice (or ice and
        glacier physics).}

    We thank you again very much for your editorial work on our manuscript.

\end{document}

### 2018-08-16 articles/cycle/response-editor3.tex
Adress last comments from the editor.

@@
 
+\newcommand{\sechead}[1]{\bigskip\noindent\textbf{#1}}
+\newcommand{\referee}[1]{\bigskip\noindent\textcolor{journalname}{#1}}
+
     Dear Andreas Vieli,
 
-    \newcommand{\referee}[1]{\bigskip\noindent\textcolor{journalname}{#1}}
+    Thank you very much for supporting our manuscript for publication in
+    \emph{The Cryosphere}. We have adressed your comments and performed a few
+    last changes on the manuscript as detailed below.
+
+% ----------------------------------------------------------------------
+
+\sechead{Editor comments}
 
@@
 
+    Thanks. Done.
+
     \referee{\textbf{p.~3, l.~4:}
@@
         say: `...knowledge remains incomplete...'}
+
+    We replaced with ``uncertainties remain on''.
 
@@
 
+    Done.
+
     \referee{\textbf{p.~3, l.~9:}
@@
 
+    Done.
+
     \referee{\textbf{p.~3, l.~32:}
         would `different' instead of `distinct' not be more appropriate.}
+
+    We replaced with ``different''.
 
@@
 
+    Sorry, two words were missing here: ``topographic roughness is parametrized
+    using a bed smoother range of 5\,km.'' We have moved this sentence directly
+    after the previous mention of the SIA.
+
     \referee{\textbf{p.~4, l.~24:}
         `...without horizontal transport...'?}
+
+    Done.
 
@@
 
+    Done.
+
     \referee{\textbf{p.~7, l.~3:}
         `above average' of what? What is reference?}
+
+    We have changed to ``above annual lapse rates measured in the Alps''. The
+    reference describes lapse rate measurements.
 
@@
 
+    An incorrect version of the figure missing 200m contours was uploaded. This
+    has been corrected. Thank you very much for noticing.
+
     \referee{\textbf{p.~13, l.~30 -- p.~14, l.~3:}
@@
 
+    The reference was added. However, to avoid redundancy with the next section
+    on flow patterns, we would prefer to limit the discussion in this section
+    to modelled ice marginal positions.
+
     \referee{\textbf{p.~14, fig.~4 caption:}
@@
 
+    We have made explicit that: ``The black vertical line indicates the
+    modelled age of maximum ice cover at 24.57\,ka.''
+
     \referee{\textbf{p.~15, l.~31:}
         `simulation' (not sumulation)}
+
+    Done.
 
@@
         5 for reviews'???}
+
+    With ``cf.~Ivy-Ochs, 2015; Wirsig et al., 2016, Fig. 5 for reviews'', we
+    meant to refer to Fig.~5 in the paper by Wirsig et al. (2016). We have
+    removed ``Fig.~5'' in this particular instance.
 
@@
 
+    Done.
+
     \referee{\textbf{p.~21, l.~19:}
@@
         alps? Clarify.}
+
+    We have replaced this formulation with ``the outermost limestone reliefs''.
 
@@
 
+    Correct. We have changed to ``glacier physics''.
+
+% ----------------------------------------------------------------------
+
+%\sechead{New references}
+
     We thank you again very much for your editorial work on our manuscript.

### 2018-08-16 articles/cycle/alps-cycle.tex
Adress last comments from the editor.

@@
     ice. Finally, despite the uniform climate forcing, differences in glacier
-    catchment hypsometry cause the Last Glacial Maximum advance to be modelled
-    as a time-transgressive event, with some glaciers reaching their maximum as
+    catchment hypsometry produce a time-transgressive Last Glacial Maximum
+    advance, with some glaciers reaching their modelled maximum extent as
     early as 27\,ka, and others as late as 21\,ka.
@@
     Although the glacial history of the European Alps has been studied for
-    nearly three hundred years, it still remains incompletely known
+    nearly three hundred years, uncertainties remain on
       (1) what climate evolution led to the known maximum ice limits,
@@
       (5) how many advances occurred during the last glacial cycle.
-    Additional geological research will be needed to complete our knowledge.
-
-    But here, we intend to explore these open questions from a new angle and
+
+    Here, we intend to explore these open questions from a new angle and
     use the Parallel Ice Sheet Model
@@
     at high resolution for the optimal forcing (Sect.~\ref{sec:results}).
+    Additional geological research will be needed to complete our knowledge.
 
@@
     entire Alpine range (Fig.~\ref{fig:inputs}). The simulations were run on
-    two distinct grids, using horizontal resolutions of 2 and 1\,km,
+    two different grids, using horizontal resolutions of 2 and 1\,km,
     respectively.
@@
     the two approximations \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}.
-    Although this heuristic approach is inaccurate in the transition zone
+    In the SIA, topographic roughness is parametrized using a bed smoother
+    range of 5\,km \citep{Schoof.2003}.
+    Although the SSA-SIA heuristic is inaccurate in the transition zone
     where both vertical shear and longitudinal stretch are significant,
@@
     dynamics over mountainous topographies similar to that of the European Alps
-    \citep{Golledge.etal.2012, Ziemen.etal.2016}. In the SIA, topographic
-    roughness using a bed smoother range of 5\,km \citep{Schoof.2003}.
+    \citep{Golledge.etal.2012, Ziemen.etal.2016}.
 
@@
     24]{Tulaczyk.etal.2000, Bueler.Pelt.2015}. Basal meltwater is accumulated
-    locally without transportation. When the till becomes saturated, additional
-    meltwater assumed to drain off instantaneously outside the glacier margins,
-    i.e. it is removed from the system in an accountable way. Other parameters
+    locally without horizontal transport. When the till becomes saturated,
+    additional meltwater assumed to drain instantaneously and is removed from
+    the system in an accountable way. Other parameters
     (Table~\ref{tab:params}) follow simulations of the Greenland ice sheet
@@
     the other hand. All simulations use an annual temperature lapse rate of
-    $\gamma=6\,\unit{K\,km^{-1}}$, which is slightly above average but more
-    representative of summer months when surface melt occurs
-    \citep[Fig.~3]{Rolland.2003}.
+    $\gamma=6\,\unit{K\,km^{-1}}$, which is slightly above annual lapse rates
+    measured in the Alps but more representative of summer months when surface
+    melt occurs \citep[Fig.~3]{Rolland.2003}.
 
@@
         (200\,m contours), and ice surface velocity (blue) in the Alps
-        24.57\,ka before present, corresponding to the maximum modelled ice
+        24.57\,ka before present, corresponding to the modelled maximum ice
         cover. The modelled LGM ice extent (dashed orange line) and
@@
         a~global compilation of benthic \chem{\delta^{18}O} records
-        \citep{Lisiecki.Raymo.2005}.}
+        \citep{Lisiecki.Raymo.2005}. The black vertical line indicates the
+        modelled age of maximum ice cover at 24.57\,ka.}
       \label{fig:lgmvel}
@@
     present. Although this may appear as a contradiction with previous,
-    constant-climate modelling results \citep{Becker.etal.2016}, it follows
+    constant-climate modelling results \citep{Becker.etal.2016,
+    Jouvet.etal.2017}, it follows
     from introducing a time-dependent palaeo-temperature forcing. Without
@@
     erratics across Brünig Pass \citep{Jackli.1962}. On the other hand, the
-    simplified climate forcing used in this sumulation could not reproduce
+    simplified climate forcing used in this simulation could not reproduce
     the transport of glacial erratics from southern Valais to observed
@@
     \citealp{Ivy-Ochs.2015};
-    \citealp[Fig.~5]{Wirsig.etal.2016} for reviews; and the more recent
+    \citealp{Wirsig.etal.2016} for reviews; and the more recent
     publications by \citealp{Monegato.etal.2017, Federici.etal.2017}). These
@@
     when they experienced rapid retreat synchronously to lowering of the ice
-    surface in the mountains (\citealp{Ivy-Ochs.2015}; \citealp[Fig.~5]{Wirsig.etal.2016};
-    \citealp[Fig.~3]{Monegato.etal.2017}).
+    surface in the mountains (\citealp{Ivy-Ochs.2015};
+    \citealp[Fig.~5]{Wirsig.etal.2016}; \citealp[Fig.~3]{Monegato.etal.2017}).
 
@@
     result depends on uncertain basal sliding and ice rheological parameters
-    to which the model sensitivity was not tested here.
-    However, regional model sensitivity tests show that a modelled surface
-    compatible with the trimline elevations is incompatible with the documented
-    overspilling flow across the Simplon Pass \citep{Becker.etal.2017}, and LGM
+    to which the model sensitivity was not tested here.  However, regional
+    model sensitivity tests show that a modelled surface compatible with the
+    trimline elevations is incompatible with the documented overspilling flow
+    across the Simplon Pass \citep{Becker.2017, Becker.etal.2017}, and LGM
     palaeo-climate proxy records \citep{Cohen.etal.2017}. Unfortunately,
@@
 
-    For instance the Rhine Glacier (Fig.~\ref{fig:profiles}a) reaches a length
-    exceeding 130\,km, roughly corresponding to the location of the last major
-    Alpine reliefs and the beginning of the foothills, six times during the
+    For instance the Rhine Glacier (Fig.~\ref{fig:profiles}a) extends beyond
+    the outermost limestone reliefs six times during the
     simulation, and sometimes retreats almost completely between two advances
     (Fig.~\ref{fig:profiles}b). The Rhone Glacier, fed by several high-altitude
-    accumulation areas, advanced eight times onto modern Lake Geneva
+    accumulation areas, advances eight times onto modern Lake Geneva
     (Fig.~\ref{fig:profiles}c and d). The Dora Baltea Glacier, characterised
@@
 
-    However, these results are limited by uncertainties on ice physics. The
+    However, these results are limited by uncertainties on glacier physics. The
     till deformation model used here does not hold for sliding over

### 2018-08-17 articles/cycle/alps-cycle.tex
Update reference to Cohen et al. (2018).

@@
     date, it is supported by numerical modelling \citep{Blatter.Haeberli.1984,
-    Haeberli.Schluchter.1987, Cohen.etal.2017}. Alpine ice flow
+    Haeberli.Schluchter.1987, Cohen.etal.2018}. Alpine ice flow
     patterns were primarily controlled by subglacial topography, but there is
@@
     across the Simplon Pass \citep{Becker.2017, Becker.etal.2017}, and LGM
-    palaeo-climate proxy records \citep{Cohen.etal.2017}. Unfortunately,
+    palaeo-climate proxy records \citep{Cohen.etal.2018}. Unfortunately,
     validation through the bedrock uplift rate \citep[cf.][]{Kuchar.etal.2012}
@@
     inevitable \citep{Blatter.Haeberli.1984, Haeberli.Schluchter.1987,
-    Cohen.etal.2017}. The bedrock themal model yield significant volumes of
+    Cohen.etal.2018}. The bedrock themal model yield significant volumes of
     frozen ground periglacially north of the Alps and subglacially under the
     highest mountains, which is also in agreement with previous studies
-    \citep{Haeberli.etal.1984, Lindgren.etal.2016, Cohen.etal.2017}.
+    \citep{Haeberli.etal.1984, Lindgren.etal.2016, Cohen.etal.2018}.
 

### 2018-08-17 articles/cycle/response-editor3.tex
Add newly published references from 2018.

@@
 
-    The reference was added. However, to avoid redundancy with the next section
-    on flow patterns, we would prefer to limit the discussion in this section
-    to modelled ice marginal positions.
+    A reference to \citep{Jouvet.etal.2017a} was added. However, to avoid
+    redundancy with the next section on flow patterns, we would prefer to limit
+    the discussion in this section to modelled ice marginal positions.
 
@@
 
-    With ``cf.~Ivy-Ochs, 2015; Wirsig et al., 2016, Fig. 5 for reviews'', we
-    meant to refer to Fig.~5 in the paper by Wirsig et al. (2016). We have
-    removed ``Fig.~5'' in this particular instance.
+    With ``cf.~\citealp[Fig.~5 for reviews]{Ivy-Ochs.2015, Wirsig.etal.2016}'',
+    we meant to refer to Fig.~5 in the paper by \citep{Wirsig.etal.2016}. We
+    have removed ``Fig.~5'' in this particular instance.
 
@@
 
-%\sechead{New references}
+\sechead{New references}
+
+    References to several new papers were added \citep{Cohen.etal.2018,
+    Gild.etal.2018, Ivy-Ochs.etal.2018, Barrett.etal.2018, Spotl.etal.2018}.
 

### 2018-08-17 articles/cycle/alps-cycle.tex
Add newly published references from 2018.

@@
     isotope dating techniques at multiple locations around the Alps (cf.
-    \citealp{Ivy-Ochs.2015};
-    \citealp{Wirsig.etal.2016} for reviews; and the more recent
-    publications by \citealp{Monegato.etal.2017, Federici.etal.2017}). These
+    reviews by \citealp{Ivy-Ochs.2015, Wirsig.etal.2016}; and the more recent
+    publications by \citealp{Federici.etal.2017, Monegato.etal.2017,
+    Gild.etal.2018, Ivy-Ochs.etal.2018}). These
     data indicate that Alpine glaciers reached their maximum extent between 26
@@
     Barrett.etal.2017, Haldimann.etal.2017}, and its timing remains uncertain
-    \citep[e.g.,][]{Link.Preusser.2006, Preusser.etal.2007}.
+    \citep[e.g.,][]{Link.Preusser.2006, Preusser.etal.2007}. During MIS~3,
+    major Eastern Alpine valleys hosted wolly mammoths \citep{Spotl.etal.2018}
+    and open vegetation \citep{Barrett.etal.2018}, indicating more restricted
+    glaciation.
 

### 2018-08-18 articles/cycle/alps-cycle.tex
Update data availability statement. DOIs pending.

@@
 \codedataavailability{%
-    PISM is available as open-source software at http://pism-docs.org/.
-    Aggregated model output will be made available upon publication. Please
-    contact the corresponding author for more detailed model output data.}
+    PISM is available as open-source software at http://pism-docs.org.
+    Aggregated model variables including those used to draw figures in this
+    manuscript are available as geotiff (DOI application pending). Selected
+    model output is available as compressed netCDF (DOI application pending).
+    The supplementary animation can be found at \doi{10.5446/35164}.}
 

### 2018-08-18 articles/cycle/movefigs.sh

@@ -19,5 +19,5 @@ do
 done
 
 # put them in zip archive
-zip figures.zip fig??.pdf
+zip $paper-figs.zip fig??.pdf
 rm fig??.pdf

### 2019-10-25 articles/erosion/alps-erosion.tex
Init Alps erosion paper (alpero).

\documentclass[utf8]{article}

%\usepackage{doi}
\usepackage{authblk}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
%\usepackage[pdftex]{xcolor}
%\usepackage[pdftex]{graphicx}
%\usepackage[authoryear,round]{natbib}

% review mode
\usepackage{geometry}
\usepackage{lineno}
\linenumbers
\linespread{1.5}

%\graphicspath{{../../figures/}}

\title{Last glacial cycle glacier erosion potential in the Alps}

\author[1]{Julien Seguinot}
\author[2]{Susan Ivy-Ochs}

\affil[1]{Laboratory of Hydraulics, Hydrology and Glaciology,
          ETH Zürich, Switzerland}
\affil[2]{Laboratory of Ion Beam Physics, ETH Zürich, Switzerland}

% ======================================================================
\begin{document}
% ======================================================================

\maketitle

\begin{abstract}
\end{abstract}

% ----------------------------------------------------------------------
%\section{Introduction}
% ----------------------------------------------------------------------

% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
%\subsection{Erosion}
% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -

% ----------------------------------------------------------------------
% Acknowledgements
% ----------------------------------------------------------------------

%\paragraph{Acknowledgements}
%\paragraph{Author contributions}
%\paragraph{Conflict of interest}
%\paragraph{Contribution to the field}
%\paragraph{Data availability}

% ----------------------------------------------------------------------
% References
% ----------------------------------------------------------------------

%\bibliographystyle{frontiersinSCNS_ENG_HUMS}
%\bibliography{../../references/references}

% ----------------------------------------------------------------------
% Figures
%\clearpage
% ----------------------------------------------------------------------

%    \begin{figure}
%      \centerline{\includegraphics{alpero_glerosion}}
%      \caption{%
%          Modelled glacial erosion over the last glacial cycle.}
%      \label{fig:glerosion}
%    \end{figure}

% ----------------------------------------------------------------------
% Tables
%\clearpage
% ----------------------------------------------------------------------

% ======================================================================
\end{document}
% ======================================================================

### 2019-10-25 articles/erosion/alps-erosion.tex
Add erosion paper preliminary abstract.

@@
 \begin{abstract}
+
+    The glacial landscape of the Alps has fascinated generations of explorers,
+    artists, mountaineers and scientists with its diversity, including
+    erosional features of all scales from high-mountain cirques, to steep
+    glacial valleys and large overdeepenings. Using previous glacier modelling
+    results, and modern observations of bedrock erosion under glaciers, we
+    infer a distribution of potential glacier erosion in the Alps over the last
+    glacial cycle from 120\,000 years ago to the present.
+    %
+    Depsite large uncertainties related to the climate history of the Alps and
+    glacier erosion processes, the resulting modelled patterns of glacier
+    erosion shows persistent features (hopefully). The cumulative imprint of
+    the last glacial cycle shows peak erosion at the mouth of the large Alpine
+    valleys where glaciers are modelled to have flown with the highest
+    velocity. The modelled erosion pattern varies significantly through the
+    glacial cycle, but surprisingly the total rate of glacier erosion is
+    modelled to be relatively stable during the entire glacial cycle.  While
+    glacial maxima lead to high erosion rates at low elevations, the
+    high-elevation areas are typically preserved under cold-based ice during
+    these periods, but are modelled to have experience mode intense erosion
+    during periods of intermediate glaciation extent.
+    %
+    This result indicate that different landscapes of the same mountain range
+    most likely correspond to different tmie periods, and explains the
+    diversity of glacial landscapes in the Alps.
+
 \end{abstract}

### 2019-11-15 data/postprocess_alpcyc_inp.sh
Update paths in input postprocessing script.

@@ -27,7 +27,7 @@ do
     echo "preparing $ofile..."
 
     # compress with shuffling
-    nccopy -sd1 "$HOME/pism/input/dt/${rec}3222cool$dt.nc" $ofile
+    nccopy -sd1 "$HOME/pism/input/dt/${rec}.3222.$dt.nc" $ofile
 
     # add global attributes
     [ "$pp" == "pp" ] && wo="with" || wo="without"
@@ -53,9 +53,9 @@ do
     echo "preparing $ofile..."
 
     # merge all input files
-    cp "$HOME/pism/input/boot/alps-srtm+thk+gou11simi-$res.nc" $ofile
-    ncks -A "$HOME/pism/input/atm/alps-wcnn-$res.nc" $ofile
-    ncks -A "$HOME/pism/input/sd/alps-erai-$res.nc" $ofile
+    cp "$HOME/pism/input/boot/alps.srtm.hus12.gou11simi.$res.nc" $ofile
+    ncks -A "$HOME/pism/input/atm/alps.wcnn.$res.nc" $ofile
+    ncks -A "$HOME/pism/input/sd/alps.erai.$res.nc" $ofile
     ncpdq -O -a y,x $ofile $ofile
 
     # remove (mostly) duplicate history and add global attributes

### 2019-11-15 data/postprocess_alpero_agg.py

@@ -116,6 +116,10 @@ def main():
     if not os.path.exists('processed'):
         os.makedirs('processed')
 
+    # activate dask client http://localhost:8787/status
+    # from dask.distributed import Client
+    # print(Client().scheduler_info()['services'])
+
     # postprocess selected runs
     for run in PROC_RUNS:
         postprocess_extra(os.environ['HOME'] + '/pism/output/e9d2d1f/' + run)

### 2020-12-14 articles/erosion/alps-erosion.tex
Add Alps erosion paper tentative outline.
Add uncommited ideas from old spark prop draft.

@@
 % ----------------------------------------------------------------------
-%\section{Introduction}
+\section{Introduction}
+% ----------------------------------------------------------------------
+
+    The glacial landscape of the Alps has fascinated generations of explorers,
+    artists, mountaineers and scientists for centuries.
+
+    In fact, the global obsession for the Alps has been so large that in
+    English, a non-native language foreign to the Alps, the word “alpine” is
+    now used to describe any glacially modified mountain landscape worldwide.
+
+    Understanding the beauty of Alpine glacial landscape through the rigour of
+    scientific research is yet another endeavour.
+
+    While it has long been understood that glaciers leave a specific imprint on
+    the landscape they lie on, observing and quantifying the long-term erosion
+    and sedimentation processes has been a challenge. While some regions are
+    specifically characterised by cirque glaciation (), or glacial scouring (),
+    the European Alps present a variety of glacial landforms with a density
+    seen in few other places on Earth.
+
+    It is typically assumed that the degree of glacial modification is a proxy
+    for the duration of past glacier cover. However glacier sliding velocities
+    vary by several orders of magnitude, and erosion laws can be non-linear.
+    Besides, cold-based glaciers have been shown to preserve landscape.
+
+    Fig.~\ref{fig:photos} -- Alpine glacial erosion landscape photos.
+        (a) piedmont overdeepening (ca. 100 km), Lake Geneva
+        (b) glacial valley (ca. 10 km), Lauterbrunnen
+        (c) mountain cirque (ca. 1 km), Chüebodengletscher
+
+% ----------------------------------------------------------------------
+\section{Methods}
 % ----------------------------------------------------------------------
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-%\subsection{Erosion}
+\subsection{Ice sheet modelling}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Erosion law}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+% ----------------------------------------------------------------------
+\section{Results}
+% ----------------------------------------------------------------------
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Cumulative erosion}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    Fig.~\ref{fig:cumulative} -- Cumulative erosion.
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Temporal evolution}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    Fig.~\ref{fig:evolution} -- Erosion rate snapshots or scatter plots.
+
+% ----------------------------------------------------------------------
+\section{Discussion}
+% ----------------------------------------------------------------------
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Erosion uncertainties}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    Fig.~\ref{fig:uncertainty} -- Erosion uncertainties
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Sensitivity to climate}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    Fig.~\ref{fig:sensitivity} -- Sensitivity to climate.
+
+% ----------------------------------------------------------------------
+\section{Conclusions}
+% ----------------------------------------------------------------------
 
@@
 %    \begin{figure}
-%      \centerline{\includegraphics{alpero_glerosion}}
+%      \centerline{\includegraphics{alpero_cumulative}}
 %      \caption{%

### 2020-12-15 figures/alpero_rhineglacier.py
Add old Rhine glacier Spark proposal figure.

@@ -1,0 +1,109 @@
+#!/usr/bin/env python
+# Copyright (c) 2020, Julien Seguinot <seguinot@vaw.baug.ethz.ch>
+# Creative Commons Attribution-ShareAlike 4.0 International License
+# (CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/)
+
+"""Plot Rhine glacier velocity and erosion potential."""
+
+import matplotlib as mpl
+import cartopy.crs as ccrs
+import cartowik.profiletools as cpf
+import absplots as apl
+import pismx.open
+import util
+
+
+def main():
+    """Main program called during execution."""
+
+    # initialize figure
+    fig, grid = apl.subplots_mm(
+        figsize=(177, 60), ncols=3, subplot_kw=dict(projection=ccrs.UTM(32)),
+        gridspec_kw=dict(left=15, right=15, bottom=22, top=6, wspace=1.5))
+    cax1 = fig.add_axes_mm([10.5, 22, 3, 32])
+    cax2 = fig.add_axes_mm([163.5, 22, 3, 32])
+    pfgrid = fig.subplots_mm(ncols=3, sharex=True, gridspec_kw=dict(
+        left=15, right=15, bottom=9, top=39.5, wspace=1.5))
+
+    # set extent and subfig labels
+    for ax, label in zip(grid, 'abc'):
+        util.fig.prepare_map_axes(ax)
+        util.fig.add_subfig_label('(%s)' % label, ax=ax)
+        ax.set_extent([410e3, 620e3, 5160e3, 5300e3], crs=ax.projection)
+
+    # Map axes
+    # --------
+
+    # open postprocessed output
+    with pismx.open.dataset(
+                '../data/processed/alpcyc.1km.epic.pp.ex.1ka.nc') as ds:
+        ds = ds.sel(age=25)
+
+        # compute sliding magnitude and erosion rates
+        sliding = (ds.uvelbase**2+ds.vvelbase**2)**0.5
+        her15 = 2.7e-4*sliding**2.02  # (mm/a, Herman et al., 2015)
+        coo20 = 1.665e-1*sliding**0.6459  # (mm/a, Cook et al., 2020)
+
+    # plot background topo
+    for ax in grid:
+        ds.topg.plot.imshow(
+            ax=ax, add_colorbar=False, cmap='Greys', vmin=0, vmax=3e3)
+
[…]

### 2020-12-15 articles/erosion/alps-erosion.tex
Update erosion paper authors and figure outline.

@@
 \title{Last glacial cycle glacier erosion potential in the Alps}
-
-\author[1]{Julien Seguinot}
-\author[2]{Susan Ivy-Ochs}
-
-\affil[1]{Laboratory of Hydraulics, Hydrology and Glaciology,
-          ETH Zürich, Switzerland}
-\affil[2]{Laboratory of Ion Beam Physics, ETH Zürich, Switzerland}
+\author{Julien Seguinot}
+\affil{Anafi, Greece}
 
@@
 
-    Fig.~\ref{fig:photos} -- Alpine glacial erosion landscape photos.
+    Fig.~\ref{fig:landscape} -- Alpine glacial erosion landscape photos.
         (a) piedmont overdeepening (ca. 100 km), Lake Geneva
@@
 
-    Fig.~\ref{fig:evolution} -- Erosion rate snapshots or scatter plots.
+    Fig.~\ref{fig:transects} -- Erosion rate along transects.
+    Fig.~\ref{fig:evolution} -- Erosion rate and ice volume.
+    Fig.~\ref{fig:hypsogram} -- Erosion rate and elevation.
 

### 2020-12-24 figures/alpero_hypsogram.py
Add boot hypsometry on hypsogram figure.

@@ -15,11 +15,16 @@ def main():
     """Main program called during execution."""
 
     # initialize figure
-    fig, (ax, tsax) = apl.subplots_mm(
-        figsize=(177, 80), nrows=2, sharex=True, gridspec_kw=dict(
-            left=12, right=18, bottom=9, top=1.5,
-            hspace=1.5, height_ratios=(48, 20)))
-    cax = fig.add_axes_mm([177-18+1.5, 9+20+1.5, 3, 48])
+    fig = apl.figure_mm(figsize=(177, 80))
+    spec = fig.add_gridspec_mm(
+        ncols=2, nrows=2, left=12, right=1.5, bottom=9, top=1.5, hspace=1.5,
+        wspace=1.5, height_ratios=(48, 20), width_ratios=(142, 20))
+    ax = fig.add_subplot(spec[0, 0])
+    hax = fig.add_subplot(spec[0, 1], sharey=ax)
+    tsax = fig.add_subplot(spec[1, 0], sharex=ax)
+    cax = fig.add_axes_mm([177-18+1.5, 9+20+1.5+24, 3, 24])
+
+    # add subfigure labels
     util.fig.add_subfig_label('(a)', ax=ax)
     util.fig.add_subfig_label('(b)', ax=tsax)
 
@@ -31,19 +36,37 @@ def main():
         np.log10(ds.her2015_hyps).plot.imshow(
             ax=ax, alpha=0.75, cmap='YlOrBr', vmin=-9, vmax=0, x='age',
             cbar_ax=cax, cbar_kwargs=dict(
-                label='log10 geometric mean\n'+r'erosion rate ($m\,a^{-1}$)'))
+                label='log10 geometric mean\n'+r'erosion rate ($m\,a^{-1}$)',
+                ticks=range(-9, 1, 3)))
         # this should work in matplotlib 3.3.2 (PR #18458)
         # ds.her2015_hyps.plot.imshow(
         #    ax=ax, alpha=0.75, cmap='YlOrBr', norm=mcolors.LogNorm(1e-9, 1e0),
         #    x='age', cbar_ax=cax, cbar_kwargs=dict(
         #        label=r'erosion rate ($m\,a^{-1}$)'))
         ax.set_xlabel('')
+        ax.set_ylabel('')
         ax.set_ylabel('elevation (m)')
+        ax.tick_params(labelbottom=False)
+
+    # plot boot hypsometry
[…]

### 2020-12-24 articles/erosion/alps-erosion.tex
Add figures and captions in erosion manuscript.

@@
 \usepackage[utf8]{inputenc}
-%\usepackage[pdftex]{xcolor}
-%\usepackage[pdftex]{graphicx}
-%\usepackage[authoryear,round]{natbib}
+\usepackage[pdftex]{xcolor}
+\usepackage[pdftex]{graphicx}
+\usepackage[authoryear,round]{natbib}
 
@@
 
-%\graphicspath{{../../figures/}}
+% custom colours
+\definecolor{darkblue}{cmyk}{0.9,0.3,0.0,0.0}
+\definecolor{darkgreen}{cmyk}{0.8,0.0,1.0,0.0}
+\definecolor{darkred}{cmyk}{0.1,0.9,0.8,0.0}
+\definecolor{darkorange}{cmyk}{0.0,0.5,1.0,0.0}
+\definecolor{darkpurple}{cmyk}{0.6,0.7,0.0,0.0}
+\definecolor{darkbrown}{cmyk}{0.23,0.73,0.98,0.12}
+
+% custom commands
+\newcommand{\idea}[1]{\textcolor{darkgreen}{\emph{[\textbf{IDEA:} #1]}}}
+\newcommand{\note}[1]{\textcolor{darkblue}{\emph{[\textbf{NOTE:} #1]}}}
+\newcommand{\todo}[1]{\textcolor{darkred}{\emph{[\textbf{TODO:} #1]}}}
+\newcommand{\aref}[0]{\textcolor{darkblue}{\textbf{[REF.]}}}
+
+\graphicspath{{../../figures/}}
 
@@
     Fig.~\ref{fig:landscape} -- Alpine glacial erosion landscape photos.
-        (a) piedmont overdeepening (ca. 100 km), Lake Geneva
-        (b) glacial valley (ca. 10 km), Lauterbrunnen
-        (c) mountain cirque (ca. 1 km), Chüebodengletscher
 
@@
     Fig.~\ref{fig:transects} -- Erosion rate along transects.
+
     Fig.~\ref{fig:evolution} -- Erosion rate and ice volume.
+
     Fig.~\ref{fig:hypsogram} -- Erosion rate and elevation.
@@
 
-%\bibliographystyle{frontiersinSCNS_ENG_HUMS}
-%\bibliography{../../references/references}
+\bibliographystyle{abbrvnat}
+\bibliography{../../../references/references}
 
@@
 % Figures
-%\clearpage
+\clearpage
 % ----------------------------------------------------------------------
 
-%    \begin{figure}
-%      \centerline{\includegraphics{alpero_cumulative}}
-%      \caption{%
-%          Modelled glacial erosion over the last glacial cycle.}
-%      \label{fig:glerosion}
-%    \end{figure}
+    \begin{figure*}
+      \centerline{\includegraphics{alpero_landscape}}
+      \caption{%
+        Alpine glacial erosion landscape diversity.
+        \textbf{(a)} Piedmont overdeepening of Lake Constance, ca.~10x50\,km.
+        \textbf{(b)} Glacial trough of Lauterbrunnental, ca.~1x10\,km.
+        \textbf{(c)} Mountain cirque of Chüebodengletscher, ca.~1x1\,km.}
+      \label{fig:landscape}
+    \end{figure*}
+
+    \begin{figure*}
+      \centerline{\includegraphics{alpero_cumulative}}
+      \caption{%
+        \textbf{(a)} Modelled cumulative (time-integrated) glacial erosion
+          potential over the last glacial cycle.
+        \textbf{(b)} Modelled total ice volume in centimetres of sea-level
+          equivalent (cm~s.l.e., black), volumic (domain-integrated) erosion
+          rate (light brown) and 100-a running mean (dark brown). Shaded gray
+          areas indicate the timing for MIS~2 and~4
+          \citep{Lisiecki.Raymo.2005}.} \label{fig:cumulative}
+    \end{figure*}
+
+    \begin{figure}
+      \centerline{\includegraphics{alpero_evolution}}
+      \caption{%
+        Modelled volumic (domain-integrated) erosion rate (light brown) and 100-a
+        running mean (dark brown) in relation to modelled total ice volume in
+        centimetres of sea-level equivalent (black).}
+      \label{fig:evolution}
+    \end{figure}
+
+    \begin{figure*}
+      \centerline{\includegraphics{alpero_transects}}
+      \caption{%
+        \textbf{(a, b, c)} Modelled instantaneous erosion rate of the Rhine
+          Glacier for selected post-Last Glacial Maximum ages.
+        \textbf{(d)} Interpolated instantaneous erosion rate along a Rhine
+          Glacier transect for the entire last glacial cycle (upper panels
+          dashed line).}
+      \label{fig:transects}
+    \end{figure*}
+
+    \begin{figure*}
+      \centerline{\includegraphics{alpero_hypsogram}}
+      \caption{%
+        \textbf{(a)} Modelled erosion rate ``hypsogram'', showing the geometric
+          mean of (non-zero) modelled erosion rates in 10-m elevation bands
+          across the entire model domain and its time evolution.
+        \textbf{(b)} Same as Fig.~\ref{fig:cumulative}b.}
+      \label{fig:hypsogram}
+    \end{figure*}
+
+    \begin{figure*}
+      \centerline{\includegraphics{alpero_sensitivity}}
+      \caption{%
+        Modelled cumulative glacial erosion potential over the last glacial
+        cycle without \textbf{(a, b, c)} and with \textbf{(d, e, f)}
+        paleo-precipitation corrections, and using three different
+        palaeo-temperature histories \citep[see][]{Seguinot.etal.2018}.}
+      \label{fig:sensitivity}
+    \end{figure*}
+
+    \begin{figure*}
+      \centerline{\includegraphics{alpero_uncertainty}}
+      \caption{%
+        Modelled cumulative (time-integrated) glacial erosion potential for
+        three different erosion laws published by
+        \textbf{(a)} \citet{Cook.etal.2020},
+          $1.665 \times 10^{-1} \textbf{u}_\mathrm{b} ^{0.6459}$,
+        \textbf{(b)} \citet{Herman.etal.2015},
+          $2.7 \times 10^{-7} \textbf{u}_\mathrm{b} ^{2.02}$
+          (same as Fig.~\ref{fig:cumulative}a), amd
+        \textbf{(c)} \citet{Koppes.etal.2015},
+          $5.2 \times 10^{-8} \textbf{u}_\mathrm{b} ^{2.34}$.
+        \textbf{(d)} Corresponding erosion power laws, and
+        \textbf{(e)} modelled cumulative (time-integrated) glacial erosion
+          potential along a Rhine Glacier transect (upper panels dashed line).}
+      \label{fig:uncertainty}
+    \end{figure*}
 

### 2020-12-28 articles/erosion/alps-erosion.tex
One sentence per paragraph in erosion paper.

@@
 
-    The glacial landscape of the Alps has fascinated generations of explorers,
-    artists, mountaineers and scientists for centuries.
+    The glacial erosion landscape of the Alps has fascinated generations of
+    explorers, artists, and scientists for centuries. In English, a non-Alpine
+    language, the word “alpine” is now used to describe any glacially modified
+    mountain landscape worldwide.
 
-    In fact, the global obsession for the Alps has been so large that in
-    English, a non-native language foreign to the Alps, the word “alpine” is
-    now used to describe any glacially modified mountain landscape worldwide.
-
-    Understanding the beauty of Alpine glacial landscape through the rigour of
-    scientific research is yet another endeavour.
+    Some regions are characterised by cirque glaciation, glacial valleys,
+    or overland scouring. Other regions, including the Alps, present a variety
+    of glacial erosional landforms. It is often assumed that the degree of
+    glacial modification is a proxy for the duration of past glacier cover.
 
@@
     the landscape they lie on, observing and quantifying the long-term erosion
-    and sedimentation processes has been a challenge. While some regions are
-    specifically characterised by cirque glaciation (), or glacial scouring (),
-    the European Alps present a variety of glacial landforms with a density
-    seen in few other places on Earth.
+    and sedimentation processes has been a challenge. Glacier sliding
+    velocities vary by several orders of magnitude, and erosion laws can be
+    non-linear. Besides, cold-based glaciers have been shown to preserve
+    landscape, but it remains unclear how sharp the erosional transition to
+    cold-base ice is, and why some glaciated regions show traces of cold-base
+    glaciations and others not.
 
-    It is typically assumed that the degree of glacial modification is a proxy
-    for the duration of past glacier cover. However glacier sliding velocities
-    vary by several orders of magnitude, and erosion laws can be non-linear.
-    Besides, cold-based glaciers have been shown to preserve landscape.
+    Here we apply a non-linear erosion law to previously published model
+    results and analyse the patterns of modelled erosion rate and cumulative
+    last glacial cycle erosion potential.
 
@@
 
+    Summary of PISM physics and boundary conditions.
+
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    We follow the near-quadratic erosion law of \citep{Herman.etal.2015}.
+
+    We refer to erosion potential as the time-integrated erosion rate.
 
@@
 
+    Modelled cumulative erosion potential vary by several orders of magnitude.
+    There is a very strong localization in valleys with a large catchment and
+    valley heads, where the results are limited by model resolution.
+
     Fig.~\ref{fig:cumulative} -- Cumulative erosion.
@@
 
+    Above a certain level of glaciation, erosion rates appear uncorrelated with
+    the ice volume. Volumetric erosion rates during MIS~4 and MIS~2 are not
+    higher than other parts of the last glacial cycle. However, there is a
+    tendency for higher erosion rate during periods of ice retreat.
+
+    Fig.~\ref{fig:evolution} -- Erosion rate and ice volume.
+
+    Along the Rhine Glacier transect, stages of extensive glaciation see
+    erosion rates localized in the lower part of the catchment. Erosion rates
+    propagate inward during periods of ice retreat.
+
     Fig.~\ref{fig:transects} -- Erosion rate along transects.
 
-    Fig.~\ref{fig:evolution} -- Erosion rate and ice volume.
+    These results can be generalized to the entire model domain by using
+    elevation as a proxy for along-flow distance. During periods of larger
+    extent, erosion rates shift to lower elevation, leaving upper grounds
+    untouched. During deglaciation, erosion propagates inland.
 
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Erosion uncertainties}
+\subsection{Results interpretation}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 
-    Fig.~\ref{fig:uncertainty} -- Erosion uncertainties
+    It is often believed that glaciers are more erosive during deglaciations
+    because of higher water availability. However, surface meltwater is absent
+    from the model. Higher erosion rates can be explained by enhanced ice
+    dynamics after longer-term ice cover.
+
+    The results also indicate that if a square-rule erosion law is valid, much
+    of the intra-Alpine erosional landscape do not date from stages of major
+    glaciation.
 
@@
 
+    The modelled erosion potential depends significantly on the choice of
+    pleoclimate forcing applied, but the general patterns remain similar.
+    Higher precipitation yields higher erosion. The results also appear to be
+    sensitive to model resolution, primarily in the mountains interior, where
+    discretized slopes depend on the grid size.
+
     Fig.~\ref{fig:sensitivity} -- Sensitivity to climate.
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Erosion uncertainties}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    The cumulative erosion potential modelled by the Franz-Joseph Glacier
+    erosion law \citep{Herman.etal.2015} appear overestimated. However, the
+    error bars are large.
+
+    The erosion law by \citet{Cook.etal.2020} results in similarly high values
+    but a flatter (less localized) pattern of cumulative erosion potential,
+    with lower values in the troughs and more erosion in the mountains. The two
+    erosion laws may appear incompatible, and there would be a lot of reasons
+    for them to be incompatible. But as pointed by \citet{Cook.etal.2020}, they
+    are not necessarily incompatible. The erosion's feedback onto ice dynamics
+    may yield to a delocalization of high-velocity flow in the long run.
+
+    The erosion law by \citet{Koppes.etal.2015} yields a similar pattern but
+    lower values. This law is for tidewater glaciers, which flow much faster
+    than mountain glaciers. However, we can imagine that Alpine paleoglaciers
+    flowing on a thick sediment bed (and sometimes into proglacial lakes, not
+    included), behave closer to tidewater glaciers than the Franz-Josef steep
+    mountain glacier.
+
+    Fig.~\ref{fig:uncertainty} -- Erosion uncertainties
 
@@
 % ----------------------------------------------------------------------
+
+    The modelled erosion rates are very highly uncertain. The results are
+    limited as there is no feedback of erosion onto ice dynamics, and
+    erosion by rivers is not included. But if a square-law is used:
+    \begin{itemize}
+      \item Cumulative erosion potential is strongly localized in regions
+        of fast past glacier flow.
+      \item The total erosion rate is not correlated with the ice volume.
+      \item During major glaciations, erosion rates are high at low elevation,
+        but the mountain's interior is preserved.
+      \item Higher-elevation glacial erosional landforms were formed during
+        stages of intermediate glaciation.
+      \item Erosion increases during deglaciations, regardless of surface
+        meltwater availability.
+    \end{itemize}
 

### 2020-12-30 articles/erosion/alps-erosion.tex
Draft Alps erosion intro, refs missing.

@@
     This result indicate that different landscapes of the same mountain range
-    most likely correspond to different tmie periods, and explains the
+    most likely correspond to different time periods, and explains the
     diversity of glacial landscapes in the Alps.
@@
     The glacial erosion landscape of the Alps has fascinated generations of
-    explorers, artists, and scientists for centuries. In English, a non-Alpine
-    language, the word “alpine” is now used to describe any glacially modified
-    mountain landscape worldwide.
+    explorers, artists, and scientists for centuries. Indeed, its cultural
+    impact is so far-reaching that in English, a non-Alpine language, the
+    adjective ``alpine'' with non-capital ``a'' is commonly used to describe
+    an Alpine-like, glacially modified, mountain landscape outside the Alps,
+    while the proper noun ``Alps'' has been reused to nick-name glacier eroded
+    mountain ranges of New Zealand, Japan, and others.
 
-    Some regions are characterised by cirque glaciation, glacial valleys,
-    or overland scouring. Other regions, including the Alps, present a variety
-    of glacial erosional landforms. It is often assumed that the degree of
-    glacial modification is a proxy for the duration of past glacier cover.
+    Some mountain ranges are predominantly characterised by cirque glaciation
+    (e.g. Uinta Mountains), glacial valleys (e.g. Putorana Plateau), or
+    large-scale overdeepenings (e.g. Patagonia). But other regions, including
+    the Alps, present a higher variety glacial erosional landforms, thereby
+    bearing a different history of glacial erosion (Fig.~\ref{fig:landscape}).
+    [The degree of glacial modification of a landscape has sometimes been used
+    as a proxy for the duration of past glacier cover \aref, including for
+    instance in geomorphological mapping \aref.] This paper challenges this
+    assumption.
 
-    While it has long been understood that glaciers leave a specific imprint on
-    the landscape they lie on, observing and quantifying the long-term erosion
-    and sedimentation processes has been a challenge. Glacier sliding
-    velocities vary by several orders of magnitude, and erosion laws can be
-    non-linear. Besides, cold-based glaciers have been shown to preserve
+    While it has long been understood that glaciers leave a specific erosional
+    imprint on the landscape, observing and quantifying the long-term erosion
+    and sedimentation processes has been a challenge, not least due to the
+    inaccessibility of glacier beds and the slow and stochastic nature of
+    glacial erosion processes. More recent studies have moved away from process
+    understanding in attempts to determine empirical erosion laws for an entire
+    glacier or set of glaciers. \citet{Herman.etal.2015} collected sediment
+    samples for \todo{X} months at the snout of Franz-Josef Glacier and mapped
+    their chemical composition to different geological zones, and hence
+    different glacier velocity, to construct a relationship between the
+    reconstructed glacier basal velocity and the measured erosion rate.
+    \citet{Koppes.etal.2015} \todo{check what they did}. \citet{Cook.etal.2020}
+    \todo{check what they did}.
+
+    These studies show that, while glacier and ice sheet velocities vary by
+    several orders of magnitude, their relationship to erosion can be
+    non-linear, yielding an even larger spectrum of variation for glacier
+    erosion rates. Besides, cold-based glaciers have been shown to preserve
     landscape, but it remains unclear how sharp the erosional transition to
@@
 
-    Here we apply a non-linear erosion law to previously published model
-    results and analyse the patterns of modelled erosion rate and cumulative
-    last glacial cycle erosion potential.
+    Here, a non-linear erosion law is applied to previously published model
+    results and the patterns of modelled erosion rate and cumulative
+    last glacial cycle erosion potential are analysed. The resulting
+    uncertainties are expectantly enormous, yet the results are hoped to
+    provide insights on the diversity of the Alpine glacial erosion landscape.
 
-    Fig.~\ref{fig:landscape} -- Alpine glacial erosion landscape photos.
+    \note{OK, maybe I need to do a literature review...}
 

### 2020-12-31 articles/erosion/alps-erosion.tex
Draft Alps erosion methods. That was easy.

@@
 
-    Summary of PISM physics and boundary conditions.
+    The ice-sheet model set-up was presented, and the results discussed, in an
+    earlier publication \citep{Seguinot.etal.2018}. The simulations use the
+    Parallel Ice Sheet Model (PISM, development version~e9d2d1f), an open
+    source, finite difference, shallow ice sheet model
+    \citep{PISM-authors.2017}. The model includes temperature and water-content
+    dependent creep, pseudo-plastic basal sliding accounting for till
+    dilatation under high water pressure, bedrock deformation under the ice
+    load, and a~positive degree-day (PDD) surface mass balance model. The model
+    is initialized with assumed present-day ice thickness and equilibrium
+    ice and bedrock temperature at 120\,ka, and ran to the present.
+    Climate forcing is provided by a~monthly climatology from interpolated
+    observational data \citep[WorldClim;][]{Hijmans.etal.2005} and the European
+    Centre for Medium-Range Weather Forecasts Reanalysis Interim
+    \citep[ERA-Interim;][]{Dee.etal.2011}, amended with temperature lapse-rate
+    corrections from the European Project for Ice Coring in Antarctica
+    \citep[EPICA;][] {Jouzel.etal.2007}, and in some cases, time-dependent
+    palaeo-precipitation reductions. Lower-resolution, alternative climate
+    scenarios are analyzed in the discussion part.
+
+    Physical parametres are listed in the earlier open-access publication
+    \citep{Seguinot.etal.2018} and complete PISM set-up is stored in long-term
+    archived model output metadata for this \todo{DOI}, and for the earlier
+    \todo{DOIs} publications.
 
@@
 
-    We follow the near-quadratic erosion law of \citep{Herman.etal.2015}.
+    Modelled erosion rates, $\dot{e}$, are calculated from the modelled basal
+    velocities, $u_\mathrm{b}$, using a near-quadratic empirical erosion law,
+    %
+    \begin{equation}
+        \dot{e} = K_\mathrm{g} u_\mathrm{b}^l ,
+    \end{equation}
+    %
+    where $K_g = 2.7\times 10^{-7}\,m^{1-l}\,a^{l-1}$ and $l = 2.02$ are
+    empirical constants fitted to measured suspended sediment loads and
+    reconstructed basal velocities of Franz-Joseph Glacier
+    \citep{Herman.etal.2015}. It is important to note that there are both large
+    uncertainties around these numbers concerning Franz-Joseph Glacier
+    itself \todo{quantify}, and little insurance that this erosion law applies
+    elsewhere.
 
-    We refer to erosion potential as the time-integrated erosion rate.
+    Unlike landscape evolution model studies \aref, this study does not account
+    for the feedback of glacier erosion onto bedrock topography and ice dynamics.
+    Neither are other forms of erosion accounted for. Instead, the
+    time-integrated erosion rate,
+    %
+    \begin{equation}
+        e =  K_\mathrm{g} \int u_\mathrm{b}^l dt,
+    \end{equation}
+    %
+    is later referred to as the cumulative ``erosion potential''. This erosion
+    potential is numerically approximated using a time-step of 10\,a in the
+    main simulation, and 100\,a in the alternative climate scenarios runs. It
+    can already be noticed from the above formula, that with an exponent, $l$,
+    higher than one, the cumulative erosion potential has a greater dependency
+    on the basal ice dynamics, $u_\mathrm{b}$, than the timing of glacier
+    cover, $dt$.
 

### 2021-01-06 articles/erosion/alps-erosion.tex
Draft the results part. Unsure about this outline.

@@
 
-    Modelled cumulative erosion potential vary by several orders of magnitude.
-    There is a very strong localization in valleys with a large catchment and
-    valley heads, where the results are limited by model resolution.
-
-    Fig.~\ref{fig:cumulative} -- Cumulative erosion.
+    The modelled cumulative (time-integrated) glacial erosion potential
+    (Fig.~\ref{fig:cumulative}a) varies by several orders of magnitude
+    from insignificant to (unrealistic, see below) kilometre-scale erosion
+    potential. Its spatial patterns of modelled erosion potential show a very
+    strong localization along the Alpine valleys, with local maxima at
+    the Alpine gates where ice flow transited from valley to piedmont glaciers,
+    and sometimes further up-valley.
+    There is a genereral tendency for higher cumulative erosion in the
+    north-western Alps where the input winter precipitation
+    \citep[WorldClim,][Fig.~1h]{Seguinot.etal.2018} is higher and the
+    preexisting glacial topography more pronounced.
 
@@
 
-    Above a certain level of glaciation, erosion rates appear uncorrelated with
-    the ice volume. Volumetric erosion rates during MIS~4 and MIS~2 are not
-    higher than other parts of the last glacial cycle. However, there is a
-    tendency for higher erosion rate during periods of ice retreat.
+    The modelled volumic (domain-integrated) glacial erosion rates
+    (Fig.~\ref{fig:cumulative}b) do not correlate with the modelled total ice
+    volume. A comparison between the modelled ice volume and the modelled
+    volumic erosion rate show no relation between these two quantities, except
+    for the onset and termination of the glacial cycle were ice volume is low
+    and sliding processes unlikely to be captured by the model resolution.
+    Instead, erosion is modelled to be higher by a factor 3 to 10 during
+    periods of glacier retreat, than during periods of glacier advance
+    (Fig.~\ref{fig:evolution}).
 
-    Fig.~\ref{fig:evolution} -- Erosion rate and ice volume.
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Spatial migration}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 
-    Along the Rhine Glacier transect, stages of extensive glaciation see
-    erosion rates localized in the lower part of the catchment. Erosion rates
-    propagate inward during periods of ice retreat.
-
-    Fig.~\ref{fig:transects} -- Erosion rate along transects.
+    A closer look at the Rhine Glacier, one of the paleo-ice sheet's largest
+    outlets, reveals a spatial migration of the modelled high erosion rates.
+    During glacier advance and stages of extensive glaciation, the modelled
+    erosion rates are moderate and restricted to low elevations, while the
+    most interior parts of the Alps \citep[modelled to be
+    largely-cold-based][Fig.~6c]{Seguinot.etal.2018}, experience insignificant
+    erosion rates (Fig.~\ref{fig:transects}a and~d). The modelled erosion
+    rates both increase and propagate inwards during periods of glacier
+    retreat (Fig.~\ref{fig:transects}b, c and~d).
 
     These results can be generalized to the entire model domain by using
-    elevation as a proxy for along-flow distance. During periods of larger
-    extent, erosion rates shift to lower elevation, leaving upper grounds
-    untouched. During deglaciation, erosion propagates inland.
+    (present-day) bedrock altitude as a proxy for along-flow distance
+    (Fig.~\ref{fig:hypsogram}). Periods of modelled increasing and maximum ice
+    volume corresponds to local modelled erosion rate minima, with significant
+    erosion rates restricted to lower elevations. On the other hand, periods of
+    modelled decreased ice volume (and area) correspond to higher local
+    modelled erosion rates and a shift of significant erosion rates to
+    higher-elevation areas (Fig.~\ref{fig:hypsogram}).
 
-    Fig.~\ref{fig:hypsogram} -- Erosion rate and elevation.
+    \note{I suddenly realize it may be better to highlight results from the
+          Koppes et al. erosion law for tidewater glaciers instead, and maybe
+          pull the sensitivity study up before the main results. To dedice
+          about this I will need to get back to the Koppes and Cook paper to
+          really understand what these laws are about.}
 

### 2021-01-18 articles/erosion/alps-erosion.tex
Introduce the three empirical studies in intro.

@@
     understanding in attempts to determine empirical erosion laws for an entire
-    glacier or set of glaciers. \citet{Herman.etal.2015} collected sediment
-    samples for \todo{X} months at the snout of Franz-Josef Glacier and mapped
-    their chemical composition to different geological zones, and hence
-    different glacier velocity, to construct a relationship between the
-    reconstructed glacier basal velocity and the measured erosion rate.
-    \citet{Koppes.etal.2015} \todo{check what they did}. \citet{Cook.etal.2020}
-    \todo{check what they did}.
+    glacier or set of glaciers. \citet{Koppes.etal.2015} quantified sediment
+    yields in 15~Patagonian and Antarctic Peninsula fjords, concluding at a
+    predominant control of surface air temperature on rapid glacier erosion and
+    calibrating a near-square-law from glacier velocity to erosion rate.
+    \citet{Herman.etal.2015} collected suspended sediment samples for 5~months
+    in the outlet stream of Franz-Josef Glacier, mapped their chemical
+    composition to geologic zones of distinctive glacier speed, and also
+    concluded to a nearly-square relationship between from basal sliding to
+    erosion rate, albeit much higher erosion rates. On the other hand,
+    \citet{Cook.etal.2020} assembled a global compilation of erosion rate for
+    38~glaciers, showing a predominant role of glacier sliding velocity over
+    climate variables, yet concluding at a sub-linear erosion law.
 

### 2021-01-24 movies/anim_alps_4k.py
Fix a warning in movies erosion timeseries.

@@ -363,8 +363,9 @@ def plot_tagline(ax, data, time, text='  {: .0f}', **kwargs):
     """Plot progress line with moving text time tag."""
     data = data[data.age >= -time/1e3]
     ax.plot(data.age*1e3, data, **kwargs)
-    ax.text(-time, data[-1], '  '+text.format(float(data[-1])),
-            ha='left', va='center', clip_on=True, **kwargs)
+    if data[-1].notnull():  # as happens with rolling means
+        ax.text(-time, data[-1], '  '+text.format(float(data[-1])),
+                ha='left', va='center', clip_on=True, **kwargs)
 
 
 def plot_rolling(ax, data, time, text='  {: .0f}', **kwargs):
@@ -478,7 +479,7 @@ def main():
     if args.crop in ('lu', 'ma'):
         start, end, step = -45000, -15000, 10
     else:
-        start, end, step = -120000, -0, 4000
+        start, end, step = -120000, -0, 40
 
     # plot colorbar separately
     prefix = '/run/media/julien/coldroom/anim/anim_alps_4k'

### 2021-01-24 movies/anim_alps_4k.sh
Update erosion movie metadata and bash script.

@@ -15,7 +15,7 @@ esac
 
 # color bar overlay stream
 case $mode in
-    er | ul ) cbar_args="-i anim_alps_4k_cbar_${mode}_${lang}.png"
+    er | ul ) cbar_args="-i $HOME/anim/anim_alps_4k_cbar_${mode}_${lang}.png"
               cbar_filt="[geog][7]overlay[geog];"
 esac
 
@@ -77,5 +77,5 @@ ffmpeg \
 # accelerate for social media
 ffmpeg \
     -i $HOME/anim/anim_alps_${size}_${crop}_${mode}_${over}_${lang}.mp4 \
-    -s 1920x1080 -vf "trim=5:125,setpts=PTS/30" \
-    $HOME/anim/anim_alps_${size}_${crop}_${mode}_${over}_${lang}_x30.mp4
+    -s 1920x1080 -vf "trim=5:125,setpts=PTS/10" \
+    $HOME/anim/anim_alps_${size}_${crop}_${mode}_${over}_${lang}_x10.mp4

### 2021-01-24 articles/erosion/alps-erosion.tex
Rework methods and results for new erosion law.

@@
     palaeo-precipitation reductions. Lower-resolution, alternative climate
-    scenarios are analyzed in the discussion part.
+    scenarios are analyzed in the discussion part
+    (Sect.~\ref{sec:sensitivity}).
 
@@
     Modelled erosion rates, $\dot{e}$, are calculated from the modelled basal
-    velocities, $u_\mathrm{b}$, using a near-quadratic empirical erosion law,
+    velocities, $u_\mathrm{b}$, using a empirical erosion power-law,
     %
@@
     %
-    where $K_g = 2.7\times 10^{-7}\,m^{1-l}\,a^{l-1}$ and $l = 2.02$ are
-    empirical constants fitted to measured suspended sediment loads and
-    reconstructed basal velocities of Franz-Joseph Glacier
-    \citep{Herman.etal.2015}. It is important to note that there are both large
-    uncertainties around these numbers concerning Franz-Joseph Glacier
-    itself \todo{quantify}, and little insurance that this erosion law applies
-    elsewhere.
+    where $K_g = 5.2\times 10^{-11}\,m^{1-l}\,a^{l-1}$ and $l = 2.34$ are
+    empirical constants calibrated on quantified sediment yields and velocities
+    from 15~Patagonian and Antarctic Peninsula tidewater glaciers
+    \citep{Koppes.etal.2015}. It is important to note that there are both large
+    uncertainties surrounding these numbers on the aforementioned glaciers
+    \todo{quantify}, and little insurance that this erosion law applies
+    elsewhere. Alternative erosion laws are analyzed in the discussion part
+    (Sect.~\ref{sec:uncertainty}).
+
+    \note{I initially built the paper around the erosion law by
+          \citep{Herman.etal.2015}, that yielded similar erosion patterns but
+          up to several km of cumulative erosion in the valleys.}
 
@@
     (Fig.~\ref{fig:cumulative}a) varies by several orders of magnitude
-    from insignificant to (unrealistic, see below) kilometre-scale erosion
-    potential. Its spatial patterns of modelled erosion potential show a very
-    strong localization along the Alpine valleys, with local maxima at
+    from insignificant to hundred-metres scale erosion
+    potential. Its spatial patterns show a very
+    strong localization along the Alpine valleys, with local maxima occuring both at
     the Alpine gates where ice flow transited from valley to piedmont glaciers,
-    and sometimes further up-valley.
+    and further up-valley where valley slopes increase.
     There is a genereral tendency for higher cumulative erosion in the
@@
     volume. A comparison between the modelled ice volume and the modelled
-    volumic erosion rate show no relation between these two quantities, except
+    volumic erosion rate shows no single relation between these two quantities, except
     for the onset and termination of the glacial cycle were ice volume is low
@@
 
-    \note{I suddenly realize it may be better to highlight results from the
-          Koppes et al. erosion law for tidewater glaciers instead, and maybe
-          pull the sensitivity study up before the main results. To dedice
-          about this I will need to get back to the Koppes and Cook paper to
-          really understand what these laws are about.}
-
 % ----------------------------------------------------------------------
@@
 \subsection{Sensitivity to climate}
+\label{sec:sensitivity}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
@@
 \subsection{Erosion uncertainties}
+\label{sec:uncertainty}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -

### 2021-01-24 articles/erosion/alps-erosion.tex
Draft results interpretation. On the move again.

@@
 
-    It is often believed that glaciers are more erosive during deglaciations
-    because of higher water availability. However, surface meltwater is absent
-    from the model. Higher erosion rates can be explained by enhanced ice
-    dynamics after longer-term ice cover.
+    Paradoxically, the model results do not show an increase of modelled
+    glacier erosion rates with increasing ice volume.  On the opposite, periods
+    of ice advance and maximum glaciation correspond to comparatively slow
+    erosion. While this result may seem paradoxical, it is in fact
+    compatible with previous field-based studies that have evidence an inverse
+    relationship between air temperatures and glacier erosion rates
+    \citep{Koppes.etal.2015, Cook.etal.2020}.
 
-    The results also indicate that if a square-rule erosion law is valid, much
-    of the intra-Alpine erosional landscape do not date from stages of major
-    glaciation.
+    Deglaciation periods, on the other hand, which are characterised by warmer
+    climate and above-balance glacier extent, yield peak domain-integrated
+    erosion rates. Field-based studies have associated deglaciation periods to
+    higher rates of glacial erosion and proglacial sedimentation, but the increase
+    has often been attributed to higher meltwater availability, promoting river
+    erosion, enhanced glacier sliding, and hydrofracturing (plucking) \aref. It
+    is important to note, however, that such processed are not accounted for in
+    the present study. In particular, deglaciation surface meltwater is assumed
+    to instantly exit the glacier with no effect on glacier dynamics.
+
+    The enhanced glacier dynamics during periods of ice retreat is here
+    modelled to result from changes of air temperature alone, reported through
+    diffusion and advection through the ice, thereby facilitating shear strain
+    and eventually activating basal sliding. The observed feedbacks between
+    surface meltwater and basal sliding would presumably enhance erosion peaks
+    modelled to occur during deglaciation. Enhanced glacier erosion can be
+    expected to accompany the ongoing 21st-century deglaciation phase.
+
+    The results also indicate that during stages of extensive glaciation, much
+    of the intra-Alpine region experiences insignificant erosion rates. For
+    instance, from 35 to 18\,ka BP, terrain with elevations above 1000\,m are
+    modelled to experiences virtually no glacier erosion. This would imply that
+    the spectacular glacial valleys found throughout the northwestern Alps do
+    not date from the LGM or similar periods, but from periods of intermediate,
+    and likely shrinking, ice cover.
 

### 2021-01-25 articles/erosion/alps-erosion.tex
Rename uncertainty figure as powerlaws figure.
Reshuffle erosion discussion part. Seeing clearer.
Draft choice of erosion law. A mess. Call it 0.4.

@@
     elsewhere. Alternative erosion laws are analyzed in the discussion part
-    (Sect.~\ref{sec:uncertainty}).
+    (Sect.~\ref{sec:powerlaws}).
 
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Results interpretation}
+\subsection{Climate sensitivity}
+\label{sec:sensitivity}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \note{I feel there is not much to learn from this section. I am thinking
+          about just leaving it out.}
+
+    The modelled erosion potential depends significantly on the choice of
+    paleoclimate forcing applied, but the general patterns remain similar.
+    Higher precipitation yields higher erosion. The results also appear to be
+    sensitive to model resolution, primarily in the mountains interior, where
+    bedrock slopes and sling speed depend more strongly on the grid size
+    (Fig.~\ref{fig:sensitivity}).
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Choice of erosion law}
+\label{sec:powerlaws}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    The choice of erosion law significantly impacts results presented in this
+    study (Fig.~\ref{fig:powerlaws}). The erosion law based on quantified
+    sediment yields from Patagonian and Antarctic Peninsula tidewater glaciers
+    \citet[$\dot{e} = 5.2 \times 10^{-8} u_\mathrm{b}^{2.34}$]{Koppes.etal.2015},
+    our default choice, yields moderate yet strongly localized cumulative erosion
+    potential (Figs~\ref{fig:cululative} and Fig.~\ref{fig:powerlaws}a).
+
+    The erosion law based on measured suspended sediment load and composition
+    in the outlet creek of Franz-Joseph Glacier
+    \citep[$\dot{e} = 2.7 \times 10^{-7} u_\mathrm{b}^{2.02}$]{Herman.etal.2015},
+    also yields strongly localized erosion potential but much higher values
+    (Fig.~\ref{fig:powerlaws}b).
+
+    The erosion law based on a global compilation of glacier velocity and erosion
+    rates
+    \citet[$\dot{e} = 1.665 \times 10^{-1} u_\mathrm{b}^{0.6459}$]{Cook.etal.2020},
+    results in similarly high values, but a flatter (less localized) pattern of
+    cumulative erosion potential (Fig.~\ref{fig:powerlaws}ac), with lower
+    values in the troughs and more erosion in the mountains.
+
+    These three erosion laws appear incompatible. Given the scarcity of measured
+    glacier erosion rates, and the variety in glacier systems, climate and
+    geologic setting, this is not a surprise.  But as more extensively
+    discussed by \citet{Cook.etal.2020}, they do not have to be entirely
+    incompatible. The erosion's feedback onto ice dynamics
+    may yield to a delocalization of high-velocity flow in the long run.
+
+    Alternative empirical erosion laws by
+    \citet{Herman.etal.2015, Cook.etal.2020} yield glacial cycle cumulative
+    erosion potential several orders of magnitude higher than modelled with the
+    law by \citet{Koppes.etal.2015}, with values in the kilometres or tens of
+    kilometres in places, which is of course well above the existing glacial
+    relief from the Alps resulting from multiple glacial cycles.
+
+    It is surprising that the erosion law by \citet{Koppes.etal.2015},
+    calibrated on tidewater glaciers, yields more realistic erosion rates for
+    the glacial Alps. However, we can imagine that Alpine paleoglaciers
+    flowing on a thick sediment bed (and sometimes into proglacial lakes, not
+    included), behave closer to tidewater glaciers than the Franz-Josef steep
+    mountain glacier. Moreover, the cold climate of the LGM Alps resembled more
+    that of modern Patagonia and Antarctica, than that of Franz-Joseph Glacier
+    or most glaciers in the global compilation by \citet{Cook.etal.2020}.
+
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Deglaciation rapid erosion}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
@@
 
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\subsection{Age of the glacial landscape}
+% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
     The results also indicate that during stages of extensive glaciation, much
@@
     and likely shrinking, ice cover.
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Sensitivity to climate}
-\label{sec:sensitivity}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    The modelled erosion potential depends significantly on the choice of
-    pleoclimate forcing applied, but the general patterns remain similar.
-    Higher precipitation yields higher erosion. The results also appear to be
-    sensitive to model resolution, primarily in the mountains interior, where
-    discretized slopes depend on the grid size.
-
-    Fig.~\ref{fig:sensitivity} -- Sensitivity to climate.
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Erosion uncertainties}
-\label{sec:uncertainty}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    The cumulative erosion potential modelled by the Franz-Joseph Glacier
-    erosion law \citep{Herman.etal.2015} appear overestimated. However, the
-    error bars are large.
-
-    The erosion law by \citet{Cook.etal.2020} results in similarly high values
-    but a flatter (less localized) pattern of cumulative erosion potential,
-    with lower values in the troughs and more erosion in the mountains. The two
-    erosion laws may appear incompatible, and there would be a lot of reasons
-    for them to be incompatible. But as pointed by \citet{Cook.etal.2020}, they
-    are not necessarily incompatible. The erosion's feedback onto ice dynamics
-    may yield to a delocalization of high-velocity flow in the long run.
-
-    The erosion law by \citet{Koppes.etal.2015} yields a similar pattern but
-    lower values. This law is for tidewater glaciers, which flow much faster
-    than mountain glaciers. However, we can imagine that Alpine paleoglaciers
-    flowing on a thick sediment bed (and sometimes into proglacial lakes, not
-    included), behave closer to tidewater glaciers than the Franz-Josef steep
-    mountain glacier.
-
-    Fig.~\ref{fig:uncertainty} -- Erosion uncertainties
 
@@
     \begin{figure*}
-      \centerline{\includegraphics{alpero_uncertainty}}
+      \centerline{\includegraphics{alpero_powerlaws}}
       \caption{%
@@
         three different erosion laws published by
-        \textbf{(a)} \citet{Cook.etal.2020},
-          $1.665 \times 10^{-1} \textbf{u}_\mathrm{b} ^{0.6459}$,
+        \textbf{(a)} \citet{Koppes.etal.2015},
+          $5.2 \times 10^{-8} u_\mathrm{b} ^{2.34}$ (same as
+          Fig.~\ref{fig:cumulative}a but with a different colour scale),
         \textbf{(b)} \citet{Herman.etal.2015},
-          $2.7 \times 10^{-7} \textbf{u}_\mathrm{b} ^{2.02}$
-          (same as Fig.~\ref{fig:cumulative}a), amd
-        \textbf{(c)} \citet{Koppes.etal.2015},
-          $5.2 \times 10^{-8} \textbf{u}_\mathrm{b} ^{2.34}$.
+          $2.7 \times 10^{-7} u_\mathrm{b} ^{2.02}$, and
+        \textbf{(c)} \citet{Cook.etal.2020},
+          $1.665 \times 10^{-1} u_\mathrm{b} ^{0.6459}$.
         \textbf{(d)} Corresponding erosion power laws, and
@@
           potential along a Rhine Glacier transect (upper panels dashed line).}
-      \label{fig:uncertainty}
+      \label{fig:powerlaws}
     \end{figure*}

### 2021-01-25 articles/erosion/alps-erosion.tex
Add Ian (hoping he accepts), rework the abstract.

@@
 \title{Last glacial cycle glacier erosion potential in the Alps}
-\author{Julien Seguinot}
-\affil{Anafi, Greece}
+\author[1]{Julien Seguinot}
+\author[2]{Ian Delaney}
+\affil[1]{Anafi, Greece} % {\small (juseg \emph{at} posteo.eu)}}
+\affil[2]{Institute of Earth Surface Dynamics, University of Lausanne, Switzerland}
 
@@
     erosional features of all scales from high-mountain cirques, to steep
-    glacial valleys and large overdeepenings. Using previous glacier modelling
-    results, and modern observations of bedrock erosion under glaciers, we
-    infer a distribution of potential glacier erosion in the Alps over the last
-    glacial cycle from 120\,000 years ago to the present.
+    glacial valleys and large over-deepened basins. Using previous glacier
+    modelling results, and empirical inferences bedrock erosion under modern
+    glaciers, we compute a distribution of potential glacier erosion in the Alps
+    over the last glacial cycle from 120\,000 years ago to the present.
     %
-    Depsite large uncertainties related to the climate history of the Alps and
-    glacier erosion processes, the resulting modelled patterns of glacier
-    erosion shows persistent features (hopefully). The cumulative imprint of
-    the last glacial cycle shows peak erosion at the mouth of the large Alpine
-    valleys where glaciers are modelled to have flown with the highest
-    velocity. The modelled erosion pattern varies significantly through the
-    glacial cycle, but surprisingly the total rate of glacier erosion is
-    modelled to be relatively stable during the entire glacial cycle.  While
-    glacial maxima lead to high erosion rates at low elevations, the
-    high-elevation areas are typically preserved under cold-based ice during
-    these periods, but are modelled to have experience mode intense erosion
-    during periods of intermediate glaciation extent.
+    Despite large uncertainties related to the climate history of the Alps and
+    unconstrained glacier erosion processes, the resulting modelled patterns of
+    glacier erosion shows persistent features. The cumulative imprint of
+    the last glacial cycle shows a very strong localization of glacier erosion
+    with local maxima at the mouths of major Alpine valleys and some other
+    upstream sections where glaciers are modelled to have flown with the
+    highest velocity. The modelled erosion rates vary significantly through the
+    glacial cycle, but show paradoxically little relation to the glaciation
+    extent. Phases of glacier advance and maximum extension, characterized by
+    low air and glacier temperatures, see a localization of rapid erosion rates
+    at low elevation, while intra-montane areas are modelled to date from phases
+    of intermediate glacier cover. The modelled erosion rates peak during
+    deglaciation phases, when higher air and glacier temperature enhance
+    glacier dynamics.
     %
-    This result indicate that different landscapes of the same mountain range
-    most likely correspond to different time periods, and explains the
-    diversity of glacial landscapes in the Alps.
+    Our results depict the Alpine glacier erosion landscape as a
+    time-transgressive patchwork, with different erosional landscapes of the
+    same mountain range corresponding to different glaciation stages and time
+    periods, thus offering insight into its diversity.
 

### 2021-01-25 articles/erosion/alps-erosion.tex
Rework the introduction, add note for Ian.

@@
 \newcommand{\aref}[0]{\textcolor{darkblue}{\textbf{[REF.]}}}
+\newcommand{\ian}[1]{\textcolor{darkbrown}{\emph{[\textbf{IAN:} #1]}}}
 
@@
     The glacial erosion landscape of the Alps has fascinated generations of
-    explorers, artists, and scientists for centuries. Indeed, its cultural
-    impact is so far-reaching that in English, a non-Alpine language, the
-    adjective ``alpine'' with non-capital ``a'' is commonly used to describe
-    an Alpine-like, glacially modified, mountain landscape outside the Alps,
-    while the proper noun ``Alps'' has been reused to nick-name glacier eroded
-    mountain ranges of New Zealand, Japan, and others.
+    explorers, artists, mountaineers and scientists for centuries. Its cultural
+    impact is indeed so far-reaching, that in English, a non-Alpine language,
+    the adjective ``alpine'' with non-capital ``a'' is now casually used to
+    describe an Alpine-like, glacially modified mountain landscape outside the
+    Alps, while the proper noun ``Alps'' has been applied to nick-name
+    Alpine-like, glacier eroded mountain ranges in Norway (\emph{Lyngsalpene}),
+    New Zealand (Southern Alps), Japan (\emph{Nihon Arupusu}), and elsewhere.
 
@@
     large-scale overdeepenings (e.g. Patagonia). But other regions, including
-    the Alps, present a higher variety glacial erosional landforms, thereby
-    bearing a different history of glacial erosion (Fig.~\ref{fig:landscape}).
-    [The degree of glacial modification of a landscape has sometimes been used
-    as a proxy for the duration of past glacier cover \aref, including for
-    instance in geomorphological mapping \aref.] This paper challenges this
-    assumption.
+    the Alps, present a higher variety glacial erosional landforms
+    (Fig.~\ref{fig:landscape}), whose implications on glacial history is yet to
+    be understood.
+
+    In absence of other glacial landforms, the degree of glacial modification
+    of a landscape has sometimes been used as a proxy for the duration of past
+    glacier cover, as for instance in geomorphological mapping \aref. However,
+    cold-based glaciers have been shown to preserve landforms as fragile as
+    sand beaches \aref. Yet it remains unclear how sharp the erosional
+    transition to cold-based ice is, and why some glaciated regions show
+    glacially preserved landscapes and others, including the Alps, do not
+    \aref.
 
@@
     inaccessibility of glacier beds and the slow and stochastic nature of
-    glacial erosion processes. More recent studies have moved away from process
-    understanding in attempts to determine empirical erosion laws for an entire
+    glacial erosion processes \aref. \ian{Perhaps this needs more detail?}
+
+    More recent studies have moved away from process
+    understanding towards attempts at calibrating empirical erosion laws for an entire
     glacier or set of glaciers. \citet{Koppes.etal.2015} quantified sediment
@@
     composition to geologic zones of distinctive glacier speed, and also
-    concluded to a nearly-square relationship between from basal sliding to
-    erosion rate, albeit much higher erosion rates. On the other hand,
-    \citet{Cook.etal.2020} assembled a global compilation of erosion rate for
+    concluded to a near-square relationship between basal sliding and glacier
+    erosion, but yielding much higher erosion rates.
+    \citet{Cook.etal.2020} assembled a global compilation of erosion rates for
     38~glaciers, showing a predominant role of glacier sliding velocity over
-    climate variables, yet concluding at a sub-linear erosion law.
+    climate variables, yet concluding at a sub-linear relationship to erosion.
 
-    These studies show that, while glacier and ice sheet velocities vary by
-    several orders of magnitude, their relationship to erosion can be
-    non-linear, yielding an even larger spectrum of variation for glacier
-    erosion rates. Besides, cold-based glaciers have been shown to preserve
-    landscape, but it remains unclear how sharp the erosional transition to
-    cold-base ice is, and why some glaciated regions show traces of cold-base
-    glaciations and others not.
-
-    Here, a non-linear erosion law is applied to previously published model
-    results and the patterns of modelled erosion rate and cumulative
-    last glacial cycle erosion potential are analysed. The resulting
-    uncertainties are expectantly enormous, yet the results are hoped to
-    provide insights on the diversity of the Alpine glacial erosion landscape.
-
-    \note{OK, maybe I need to do a literature review...}
+    Here, the non-linear erosion law by \citet{Koppes.etal.2015} is applied to
+    previously published model results \citep{Seguinot.etal.2018} and the
+    patterns of modelled erosion rate and cumulative last glacial cycle erosion
+    potential are analysed. Despite aggregated uncertainties on paleoclimate,
+    glacier flow and glacier erosion processes, out results provide insights
+    into the diversity of the Alpine glacial erosion landscape.
 

### 2021-01-25 articles/erosion/alps-erosion.tex
Rework the methods. Looking good.

@@
 
-    The ice-sheet model set-up was presented, and the results discussed, in an
-    earlier publication \citep{Seguinot.etal.2018}. The simulations use the
-    Parallel Ice Sheet Model (PISM, development version~e9d2d1f), an open
-    source, finite difference, shallow ice sheet model
+    The ice-sheet model set-up was presented in an earlier publication
+    \citep{Seguinot.etal.2018} and is briefly summarized here. The simulations
+    were conducted using the Parallel Ice Sheet Model (PISM, development
+    version~e9d2d1f), an open-source, finite-difference, shallow-ice and
+    shallow-shelf model
     \citep{PISM-authors.2017}. The model includes temperature and water-content
@@
     ice and bedrock temperature at 120\,ka, and ran to the present.
+    The ice-sheet model physical parameters are listed in the earlier
+    open-access publication \citep{Seguinot.etal.2018}, and the complete PISM
+    set-up is stored in long-term archived model output metadata for the
+    present study \todo{upload}, and for the earlier publication
+    \citep{Seguinot.2020, Seguinot.2020a}.
+
     Climate forcing is provided by a~monthly climatology from interpolated
@@
     corrections from the European Project for Ice Coring in Antarctica
-    \citep[EPICA;][] {Jouzel.etal.2007}, and in some cases, time-dependent
-    palaeo-precipitation reductions. Lower-resolution, alternative climate
-    scenarios are analyzed in the discussion part
-    (Sect.~\ref{sec:sensitivity}).
-
-    Physical parametres are listed in the earlier open-access publication
-    \citep{Seguinot.etal.2018} and complete PISM set-up is stored in long-term
-    archived model output metadata for this \todo{DOI}, and for the earlier
-    \todo{DOIs} publications.
+    \citep[EPICA;][] {Jouzel.etal.2007}, and time-dependent paleo-precipitation
+    reductions \citep{Huybrechts.2002}. Lower-resolution runs based on
+    alternative paleoclimate scenarios \citep{Seguinot.etal.2018} are analyzed
+    in the discussion part (Sect.~\ref{sec:sensitivity}).
 
@@
     where $K_g = 5.2\times 10^{-11}\,m^{1-l}\,a^{l-1}$ and $l = 2.34$ are
-    empirical constants calibrated on quantified sediment yields and velocities
-    from 15~Patagonian and Antarctic Peninsula tidewater glaciers
-    \citep{Koppes.etal.2015}. It is important to note that there are both large
-    uncertainties surrounding these numbers on the aforementioned glaciers
-    \todo{quantify}, and little insurance that this erosion law applies
-    elsewhere. Alternative erosion laws are analyzed in the discussion part
-    (Sect.~\ref{sec:powerlaws}).
+    empirical constants calibrated on quantified fjord sediment yields and
+    velocities from 15~Patagonian and Antarctic Peninsula tidewater glaciers
+    \citep{Koppes.etal.2015}. Alternative erosion laws are analyzed in the
+    discussion part (Sect.~\ref{sec:powerlaws}).
 
-    \note{I initially built the paper around the erosion law by
-          \citep{Herman.etal.2015}, that yielded similar erosion patterns but
-          up to several km of cumulative erosion in the valleys.}
-
-    Unlike landscape evolution model studies \aref, this study does not account
-    for the feedback of glacier erosion onto bedrock topography and ice dynamics.
-    Neither are other forms of erosion accounted for. Instead, the
-    time-integrated erosion rate,
+    Our approach does not account for the feedback of glacier erosion onto
+    bedrock topography and ice dynamics. Neither are other forms of erosion
+    accounted for. Instead, we compute the time-integrated erosion rate,
     %
@@
     %
-    is later referred to as the cumulative ``erosion potential''. This erosion
-    potential is numerically approximated using a time-step of 10\,a in the
+    and refer to it as the cumulative ``erosion potential''. The integration
+    is numerically approximated using a time-step of 10\,a in the
     main simulation, and 100\,a in the alternative climate scenarios runs. It
-    can already be noticed from the above formula, that with an exponent, $l$,
+    can be noted from the above formula, that with an exponent, $l$,
     higher than one, the cumulative erosion potential has a greater dependency
-    on the basal ice dynamics, $u_\mathrm{b}$, than the timing of glacier
-    cover, $dt$.
+    on the basal ice dynamics, $u_\mathrm{b}$, than the duration of glacier
+    cover.
 

### 2021-01-25 articles/erosion/alps-erosion.tex
Rework the results. Oups time for meeting.

@@
     (Fig.~\ref{fig:cumulative}a) varies by several orders of magnitude
-    from insignificant to hundred-metres scale erosion
-    potential. Its spatial patterns show a very
-    strong localization along the Alpine valleys, with local maxima occuring both at
+    from insignificant to hundred-metres-scale erosion potential. Its spatial
+    patterns show a very strong localization along the Alpine valleys, with
+    local maxima occurring both at
     the Alpine gates where ice flow transited from valley to piedmont glaciers,
-    and further up-valley where valley slopes increase.
-    There is a genereral tendency for higher cumulative erosion in the
+    and further upstream where the valley slopes increase. While mountain
+    cirques are poorly captured by the model horizontal resolution, high
+    cumulative erosion potential also occur near the valley heads.
+    There is a general tendency for higher cumulative erosion in the
     north-western Alps where the input winter precipitation
     \citep[WorldClim,][Fig.~1h]{Seguinot.etal.2018} is higher and the
-    preexisting glacial topography more pronounced.
+    glacial relief more pronounced in the input topography.
 
@@
 
-    The modelled volumic (domain-integrated) glacial erosion rates
-    (Fig.~\ref{fig:cumulative}b) do not correlate with the modelled total ice
-    volume. A comparison between the modelled ice volume and the modelled
-    volumic erosion rate shows no single relation between these two quantities, except
-    for the onset and termination of the glacial cycle were ice volume is low
-    and sliding processes unlikely to be captured by the model resolution.
-    Instead, erosion is modelled to be higher by a factor 3 to 10 during
-    periods of glacier retreat, than during periods of glacier advance
-    (Fig.~\ref{fig:evolution}).
+    The modelled volumetric (domain-integrated) glacial erosion rates
+    do no systematically correlate with the modelled total ice volume
+    (Fig.~\ref{fig:cumulative}b). Except for the onset and termination of the
+    glacial cycle were ice volume is low and sliding processes unlikely to be
+    captured by the model resolution, a comparison between the modelled ice
+    volume and the modelled volumetric erosion rate shows no single relation
+    between these two quantities (Fig.~\ref{fig:evolution}). After significant
+    ice volume has been reached, there is a slight tendency for slower erosion
+    during periods of extensive glaciation (Fig.~\ref{fig:evolution}). However,
+    the domain-integrated erosion rate is modelled to be consistently higher
+    (by a factor 2 to 10) during periods of decreasing ice volume, than during
+    periods of increasing ice volume (Fig.~\ref{fig:evolution}).
 
@@
     outlets, reveals a spatial migration of the modelled high erosion rates.
-    During glacier advance and stages of extensive glaciation, the modelled
-    erosion rates are moderate and restricted to low elevations, while the
-    most interior parts of the Alps \citep[modelled to be
-    largely-cold-based][Fig.~6c]{Seguinot.etal.2018}, experience insignificant
-    erosion rates (Fig.~\ref{fig:transects}a and~d). The modelled erosion
-    rates both increase and propagate inwards during periods of glacier
+    During stages of glacier advance and maximum extension, erosion is modelled
+    to be moderate and restricted the lower parts of the catchment, while much
+    the intra-montane Alps \citep[modelled to be largely
+    cold-based,][Fig.~6c]{Seguinot.etal.2018}, are modelled to experience
+    insignificant erosion (Fig.~\ref{fig:transects}a and~d). The modelled
+    erosion rates both increase and propagate inwards during periods of glacier
     retreat (Fig.~\ref{fig:transects}b, c and~d).
 
-    These results can be generalized to the entire model domain by using
-    (present-day) bedrock altitude as a proxy for along-flow distance
-    (Fig.~\ref{fig:hypsogram}). Periods of modelled increasing and maximum ice
-    volume corresponds to local modelled erosion rate minima, with significant
-    erosion rates restricted to lower elevations. On the other hand, periods of
-    modelled decreased ice volume (and area) correspond to higher local
-    modelled erosion rates and a shift of significant erosion rates to
-    higher-elevation areas (Fig.~\ref{fig:hypsogram}).
+    The results found on the Rhine Glacier catchment can be generalized to the
+    entire model domain by using (present-day) bedrock altitude as a proxy for
+    along-flow distance (Fig.~\ref{fig:hypsogram}). Thus more generally,
+    periods of modelled increasing and maximum ice volume corresponds to lower
+    values for elevation-aggregated erosion rates, with significant erosion
+    restricted to lower elevations. On the opposite, periods of modelled
+    decreased ice volume correspond to higher local modelled erosion rates and
+    more extensive rapid erosion into higher-elevation areas
+    (Fig.~\ref{fig:hypsogram}).
 
@@
 % ----------------------------------------------------------------------
+
+    \note{I have not yet reworked the discussion.}
 
@@
 % ----------------------------------------------------------------------
+
+    \note{I still need to develop the conclusions.}
 

### 2021-01-28 figures/alpero_sensitivity.py
Fix erosion climate sensitivity subfigure labels.

@@ -29,7 +29,7 @@ def main():
 
         # prepare axes and add labels
         util.fig.prepare_map_axes(ax)
-        util.fig.add_subfig_label('(%s) ' % 'acebdf'[i], ax=ax)
+        util.fig.add_subfig_label('(%s) ' % 'adbecf'[i], ax=ax)
         util.com.add_corner_tag(label, ax=ax, va='bottom', y=(i == 5)*1/3)
 
         # load aggregated data

### 2021-01-28 articles/erosion/alps-erosion.tex
Rework the climate sensitivity discussion.

@@
 
-    \note{I have not yet reworked the discussion.}
-
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
@@
 
-    \note{I feel there is not much to learn from this section. I am thinking
-          about just leaving it out.}
-
-    The modelled erosion potential depends significantly on the choice of
-    paleoclimate forcing applied, but the general patterns remain similar.
-    Higher precipitation yields higher erosion. The results also appear to be
-    sensitive to model resolution, primarily in the mountains interior, where
-    bedrock slopes and sling speed depend more strongly on the grid size
-    (Fig.~\ref{fig:sensitivity}).
+    Different paleoclimate forcing scenarios lead to different modelled
+    glaciation histories and thus different results in terms of erosion
+    potential (Fig.~\ref{fig:sensitivity}). Scenarios including higher
+    (constant) input precipitation rates yield generally higher modelled
+    erosion potential (Fig.~\ref{fig:sensitivity}a, b, and~c), while scenarios
+    including paleo-precipitation reductions (Fig.~\ref{fig:sensitivity}d, e,
+    and~f), including our default high-resolution run, yield a smaller erosion
+    potential. The spatial pattern of modelled erosion potential is
+    nevertheless generally consistent across paleoclimate scenarios tested,
+    despite variations in the glaciation maximum extent
+    (Fig.~\ref{fig:sensitivity}). Furthermore, these changes are minor
+    considering the large uncertainties affecting the choice of erosion law.
 

### 2021-01-28 articles/erosion/alps-erosion.tex
Rework the choice of erosion law discussion.

@@
 
-    The choice of erosion law significantly impacts results presented in this
-    study (Fig.~\ref{fig:powerlaws}). The erosion law based on quantified
+    The choice of erosion law significantly impacts the results
+    (Fig.~\ref{fig:powerlaws}). Our default erosion law, based on quantified
     sediment yields from Patagonian and Antarctic Peninsula tidewater glaciers
-    \citet[$\dot{e} = 5.2 \times 10^{-8} u_\mathrm{b}^{2.34}$]{Koppes.etal.2015},
-    our default choice, yields moderate yet strongly localized cumulative erosion
-    potential (Figs~\ref{fig:cululative} and Fig.~\ref{fig:powerlaws}a).
+    \citep[${\dot{e}=5.2\times10^{-8}u_\mathrm{b}^{2.34}}$,][]{Koppes.etal.2015}
+    yields a moderate and strongly localized cumulative erosion potential
+    (Figs~\ref{fig:cumulative} and~\ref{fig:powerlaws}a). The erosion law
+    based on measured suspended sediment load from Franz-Joseph Glacier
+    \citep[${\dot{e}=2.7\times10^{-7}u_\mathrm{b}^{2.02}}$,][]{Herman.etal.2015}
+    also yields a strongly localized, and yet much higher, modelled erosion
+    potential (Fig.~\ref{fig:powerlaws}b). The erosion law based on a global
+    compilation of glacier velocity and erosion rates
+    \citep[${\dot{e}=1.665\times10^{-1}u_\mathrm{b}^{0.6459}}$,][]{Cook.etal.2020}
+    results in a similarly high values, but a flatter (less localized) pattern
+    of cumulative erosion potential (Fig.~\ref{fig:powerlaws}c).
 
-    The erosion law based on measured suspended sediment load and composition
-    in the outlet creek of Franz-Joseph Glacier
-    \citep[$\dot{e} = 2.7 \times 10^{-7} u_\mathrm{b}^{2.02}$]{Herman.etal.2015},
-    also yields strongly localized erosion potential but much higher values
-    (Fig.~\ref{fig:powerlaws}b).
+    It may seem surprising that the erosion law calibrated on Patagonian and
+    Antarctic Peninsula tidewater glaciers \citep{Koppes.etal.2015} yields a
+    more realistic cumulative erosion potential than those based on terrestrial
+    glaciers \citep{Herman.etal.2015, Cook.etal.2020}, themselves resulting
+    in cumulative erosion potential values in the order of kilometres and thus
+    higher than the Alpine relief itself. However, during the Last Glacial
+    Maximum and much of the last glacial cycle, Alpine paleoglaciers were
+    closer in size and climatic context to the present-day glaciers of
+    Patagonia and the Antarctic Peninsula \citep{Koppes.etal.2015} than to
+    Franz-Joseph Glacier \citep{Herman.etal.2015} and many of the glaciers
+    included in the global compilation by \citet{Cook.etal.2020}. Atmospheric
+    temperatures have been found to be a strong driver of erosion in studies
+    involving several glaciers \citep{Koppes.etal.2015, Cook.etal.2020}.
 
-    The erosion law based on a global compilation of glacier velocity and erosion
-    rates
-    \citet[$\dot{e} = 1.665 \times 10^{-1} u_\mathrm{b}^{0.6459}$]{Cook.etal.2020},
-    results in similarly high values, but a flatter (less localized) pattern of
-    cumulative erosion potential (Fig.~\ref{fig:powerlaws}ac), with lower
-    values in the troughs and more erosion in the mountains.
-
-    These three erosion laws appear incompatible. Given the scarcity of measured
-    glacier erosion rates, and the variety in glacier systems, climate and
-    geologic setting, this is not a surprise.  But as more extensively
-    discussed by \citet{Cook.etal.2020}, they do not have to be entirely
-    incompatible. The erosion's feedback onto ice dynamics
-    may yield to a delocalization of high-velocity flow in the long run.
-
-    Alternative empirical erosion laws by
-    \citet{Herman.etal.2015, Cook.etal.2020} yield glacial cycle cumulative
-    erosion potential several orders of magnitude higher than modelled with the
-    law by \citet{Koppes.etal.2015}, with values in the kilometres or tens of
-    kilometres in places, which is of course well above the existing glacial
-    relief from the Alps resulting from multiple glacial cycles.
-
-    It is surprising that the erosion law by \citet{Koppes.etal.2015},
-    calibrated on tidewater glaciers, yields more realistic erosion rates for
-    the glacial Alps. However, we can imagine that Alpine paleoglaciers
-    flowing on a thick sediment bed (and sometimes into proglacial lakes, not
-    included), behave closer to tidewater glaciers than the Franz-Josef steep
-    mountain glacier. Moreover, the cold climate of the LGM Alps resembled more
-    that of modern Patagonia and Antarctica, than that of Franz-Joseph Glacier
-    or most glaciers in the global compilation by \citet{Cook.etal.2020}.
+    That said, the magnitude of the modelled erosion rates can hardly be used
+    as a criteria to rule out this or that erosion law for the glacial Alps, as
+    all these laws are affected with very large uncertainties. It can not be
+    excluded that a linear or sub-linear erosion-law
+    \citep[e.g.][]{Cook.etal.2020} would lead to a less localized pattern of
+    erosion (Fig.~\ref{fig:powerlaws}c), and a stronger correlation between
+    modelled ice volume and erosion \note{not shown, should it be?}. Even
+    then, the intrinsically large variations in glacier sliding velocities
+    yields modelled glacial-cycle erosion potential two to three orders of
+    magnitude higher in the Alpine valley than near the mountain tops.
 
@@
         \textbf{(a)} \citet{Koppes.etal.2015},
-          $5.2 \times 10^{-8} u_\mathrm{b} ^{2.34}$ (same as
+          ${5.2 \times 10^{-8} u_\mathrm{b}^{2.34}}$ (same as
           Fig.~\ref{fig:cumulative}a but with a different colour scale),
         \textbf{(b)} \citet{Herman.etal.2015},
-          $2.7 \times 10^{-7} u_\mathrm{b} ^{2.02}$, and
+          ${2.7 \times 10^{-7} u_\mathrm{b}^{2.02}}$, and
         \textbf{(c)} \citet{Cook.etal.2020},
-          $1.665 \times 10^{-1} u_\mathrm{b} ^{0.6459}$.
+          ${1.665 \times 10^{-1} u_\mathrm{b}^{0.6459}}$.
         \textbf{(d)} Corresponding erosion power laws, and

### 2021-01-28 articles/erosion/alps-erosion.tex
Rework discussion of deglacial erosion.

@@
 
-    Paradoxically, the model results do not show an increase of modelled
-    glacier erosion rates with increasing ice volume.  On the opposite, periods
-    of ice advance and maximum glaciation correspond to comparatively slow
-    erosion. While this result may seem paradoxical, it is in fact
-    compatible with previous field-based studies that have evidence an inverse
-    relationship between air temperatures and glacier erosion rates
-    \citep{Koppes.etal.2015, Cook.etal.2020}.
+    The modelled, domain-integrated glacier erosion rates do not increase
+    together with ice volume. On the opposite, periods of ice advance and
+    maximum glaciation correspond to comparatively slow erosion
+    (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}). While this result may
+    seem paradoxical, and is modulated by the exponent, $l$, in the erosion
+    power-law, it is in fact compatible with previous field-based studies that
+    have evidenced an inverse relationship between air temperatures and glacier
+    erosion rates \citep{Koppes.etal.2015, Cook.etal.2020}.
 
-    Deglaciation periods, on the other hand, which are characterised by warmer
-    climate and above-balance glacier extent, yield peak domain-integrated
-    erosion rates. Field-based studies have associated deglaciation periods to
-    higher rates of glacial erosion and proglacial sedimentation, but the increase
-    has often been attributed to higher meltwater availability, promoting river
-    erosion, enhanced glacier sliding, and hydrofracturing (plucking) \aref. It
-    is important to note, however, that such processed are not accounted for in
-    the present study. In particular, deglaciation surface meltwater is assumed
-    to instantly exit the glacier with no effect on glacier dynamics.
+    Deglaciation periods, characterised by a warmer climate and above-balance
+    glacier extent, yield the highest domain-integrated erosion rates of the
+    glacial cycle (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}). This
+    result is compatible with the field-based studies, which have evidenced
+    increased glacier erosion and pro-glacial sedimentation during deglaciation
+    periods \ref, including the ongoing glacier retreat \aref. Yet this
+    increase has often been attributed to higher meltwater availability,
+    promoting river erosion, enhanced glacier sliding, and hydrofracturing
+    (plucking) \aref.
 
-    The enhanced glacier dynamics during periods of ice retreat is here
-    modelled to result from changes of air temperature alone, reported through
-    diffusion and advection through the ice, thereby facilitating shear strain
-    and eventually activating basal sliding. The observed feedbacks between
-    surface meltwater and basal sliding would presumably enhance erosion peaks
-    modelled to occur during deglaciation. Enhanced glacier erosion can be
-    expected to accompany the ongoing 21st-century deglaciation phase.
+    But importantly, these hydrological processes are not included in our
+    model, so that deglaciation surface meltwater is assumed to instantly exit
+    the glacier with no effect on glacier dynamics. The enhanced glacier
+    velocity during periods of ice retreat is here modelled to result from
+    changes of air temperature alone, propagating through diffusion and
+    advection into the ice, increasing the glacier's internal energy through
+    temperature and englacial water content, thereby facilitating shear strain
+    and eventually activating basal sliding \citep[cf.][]{Aschwanden.etal.2012,
+    Seguinot.etal.2016}. Observed feedbacks between surface meltwater and basal
+    sliding, not included in the model, presumably further accelerated erosion
+    during deglaciation. Rapid glacier erosion can be expected to accompany the
+    ongoing 21st-century deglaciation phase.
 

### 2021-01-28 articles/erosion/alps-erosion.tex
Rework/draft discussion on age of the landscape.

@@
     increased glacier erosion and pro-glacial sedimentation during deglaciation
-    periods \ref, including the ongoing glacier retreat \aref. Yet this
+    periods \aref, including the ongoing glacier retreat \aref. Yet this
     increase has often been attributed to higher meltwater availability,
@@
 
-    The results also indicate that during stages of extensive glaciation, much
-    of the intra-Alpine region experiences insignificant erosion rates. For
-    instance, from 35 to 18\,ka BP, terrain with elevations above 1000\,m are
-    modelled to experiences virtually no glacier erosion. This would imply that
-    the spectacular glacial valleys found throughout the northwestern Alps do
-    not date from the LGM or similar periods, but from periods of intermediate,
-    and likely shrinking, ice cover.
+    In many places throughout the Alps, significant modelled cumulative erosion
+    occurs far inside the Alpine valleys (Fig.~\ref{fig:cumulative}). Yet
+    instantaneous erosion rates show a time-transgressive radial pattern
+    (Fig.~\ref{fig:transects}), so that high-altitude rapid erosion is not
+    modelled to date from expansive glaciation stages, but instead to stages of
+    intermediate glacier cover (Fig.~\ref{fig:hypsogram}).
+
+    For instance, low-elevation piedmont over-deepened troughs such as Lake
+    Constance (Fig.~\ref{fig:landscape}), Lake Geneva and Lake Maggiore, are
+    here modelled to be the product of extensive, Last Glacial Maximum-like
+    glaciations. Alpine terrain with elevations above 1000\,m are modelled to
+    experience virtually no erosion from 35 to 18\,ka BP
+    (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
+    intermediate-elevation glacial valley of the Alps such as Lauterbrunnental
+    (Fig.~\ref{fig:hypsogram}), Haslital (Upper Aare) or Bout du Monde (Upper
+    Giffre). While the validity of the model results at high elevation are
+    limited by the 1-km resolution and the shallow-ice physics
+    \citep{Imhof.etal.2019}, the lack of high-elevation rapid erosion for much
+    of the modelled glacial cycle implies that the numerous Alpine mountain
+    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would date from
+    temperate periods with limited ice-cover such as the present day.
+
+    Finally, the time-transgressive nature of the modelled glacial erosion
+    rates hint at a new explanation for the Alpine glacier trim-line and lack
+    of cold-based glaciation evidence. Due to their geographic location in the
+    mid-latitudes, and their steep topographic gradient, the Alps appear as a
+    mountain range that hosted glaciers of various sizes throughout the varying
+    climate of the Quaternary, resulting in a transgressive localization of
+    glacier erosion throughout their elevational range, from the piedmont
+    during glacial maxima, to the highest valleys during interglacial periods.
+    %The Alpine trim-line, in this case, would neither correspond to the upper
+    %reach of the Last Glacial Maximum glaciers \aref, nor to an englacial
+    %temperate-ice boundary \aref, but to an upper limit of erosion from
+    %retreating glaciers on steep terrain.
 

### 2021-01-29 articles/erosion/alps-erosion.tex
Add references I thought of in the intro.

@@
 
-%\usepackage{doi}
+\usepackage{doi}
 \usepackage{authblk}
@@
 \usepackage[authoryear,round]{natbib}
+%\usepackage{CJK}  % texlive-langjapanese
 
@@
     Alpine-like, glacier eroded mountain ranges in Norway (\emph{Lyngsalpene}),
-    New Zealand (Southern Alps), Japan (\emph{Nihon Arupusu}), and elsewhere.
+    New Zealand (Southern Alps), Japan
+    (\emph{%\begin{CJK}{UTF8}{min}日本アルプス\end{CJK},
+    Nihon Arupusu}), and elsewhere.
 
@@
     be understood.
-
     In absence of other glacial landforms, the degree of glacial modification
     of a landscape has sometimes been used as a proxy for the duration of past
-    glacier cover, as for instance in geomorphological mapping \aref. However,
-    cold-based glaciers have been shown to preserve landforms as fragile as
-    sand beaches \aref. Yet it remains unclear how sharp the erosional
+    glacier cover, as for instance in geomorphological mapping
+    \citep[e.g.][]{Margold.etal.2011, Fu.etal.2012}. Cold-based glaciers have
+    been shown to preserve landforms as fragile as sand beaches
+    \citep{Kleman.1994}. Yet it remains unclear how sharp the erosional
     transition to cold-based ice is, and why some glaciated regions show
     glacially preserved landscapes and others, including the Alps, do not
-    \aref.
+    \citep{Wirsig.etal.2016, Seguinot.etal.2018}.
 
@@
         \textbf{(b)} Glacial trough of Lauterbrunnental, ca.~1x10\,km.
-        \textbf{(c)} Mountain cirque of Chüebodengletscher, ca.~1x1\,km.}
+        \textbf{(c)} Mountain cirque of Ch\"uebodengletscher, ca.~1x1\,km.}
       \label{fig:landscape}

### 2021-01-29 articles/erosion/alps-erosion.tex
Rework the conclusions. Call it 0.5 second draft.

@@
 
-    \note{I still need to develop the conclusions.}
-
-    The modelled erosion rates are very highly uncertain. The results are
-    limited as there is no feedback of erosion onto ice dynamics, and
-    erosion by rivers is not included. But if a square-law is used:
+    Due to stacked-up uncertainties regarding paleoclimate, glacier sliding and
+    erosion processes, our quantitative results are to be considered with extreme
+    uncertainties. Yet to the best of our current knowledge in these respective
+    areas of science, we can draw the following qualitative conclusions:
     \begin{itemize}
-      \item Cumulative erosion potential is strongly localized in regions
-        of fast past glacier flow.
-      \item The total erosion rate is not correlated with the ice volume.
-      \item During major glaciations, erosion rates are high at low elevation,
-        but the mountain's interior is preserved.
-      \item Higher-elevation glacial erosional landforms were formed during
-        stages of intermediate glaciation.
-      \item Erosion increases during deglaciations, regardless of surface
-        meltwater availability.
+      \item The non-linear physics of glacier deformation and sliding, combined
+        with non-linear empirical erosion laws, results in a very strong
+        localization of rapid glacier erosion in regions of fast flow.
+      \item Temperature-dependent creep and basal sliding provide a mechanism
+        for accelerated erosion during deglaciation periods, irrespective of
+        surface meltwater penetration to the glacier bed.
+      \item If a non-linear glacier erosion law is used, the climate-induced
+        slowdown of erosion counterbalances glacier expansion, so that
+        Alpine-wide glacier erosion rates do not correlate with the ice volume.
+      \item Rapid glacier erosion is restricted to low elevation during stages
+        of glacier advance and maximum glaciation, but propagates up-valley to
+        higher elevations during periods of glacier retreat.
+      \item The diversity of the Alpine glacial erosion landscape appears to
+        be the time-transgressive signature of a variety of glacial stages
+        ranging from pan-Alpine ice sheets to modern-style mountain glaciers.
     \end{itemize}

### 2021-02-01 articles/erosion/alps-erosion.tex
Fancier Makefile, add stripped bib rule.
Add alps erosion bibtex file.
Add Add Humphrey and Raymond (1994) law in text.

@@
+% Copyright (c) 2019-2021, Julien Seguinot (juseg.github.io), Ian Delanay
+% Creative Commons Attribution-ShareAlike 4.0 International License
+% (CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/)
+
+% Alps erosion paper
+% ==================
+
 \documentclass[utf8]{article}
 
+% default packages
 \usepackage{doi}
@@
 
+% figures directory
 \graphicspath{{../../figures/}}
 
+% document properties
 \title{Last glacial cycle glacier erosion potential in the Alps}
@@
 \author[2]{Ian Delaney}
-\affil[1]{Anafi, Greece} % {\small (juseg \emph{at} posteo.eu)}}
+\affil[1]{Anafi, Greece}
 \affil[2]{Institute of Earth Surface Dynamics, University of Lausanne, Switzerland}
@@
     understanding towards attempts at calibrating empirical erosion laws for an entire
-    glacier or set of glaciers. \citet{Koppes.etal.2015} quantified sediment
+    glacier or set of glaciers. \citet{Humphrey.Raymond.1994} correlated
+    temporal variations in ice velocity and suspended sediment load during a
+    surge of Variegated Glacier to establish a linear erosion law.
+    \citet{Koppes.etal.2015} quantified sediment
     yields in 15~Patagonian and Antarctic Peninsula fjords, concluding at a
@@
     also yields a strongly localized, and yet much higher, modelled erosion
-    potential (Fig.~\ref{fig:powerlaws}b). The erosion law based on a global
-    compilation of glacier velocity and erosion rates
+    potential (Fig.~\ref{fig:powerlaws}b). The erosion law based on measured
+    suspended sediment load during a surge of Variegated Glacier
+    \citep[${\dot{e}=1\times10^{-4}u_\mathrm{b}^{1}}$,][]{Humphrey.Raymond.1994},
+    results in similarly high values, but a less localized pattern of cumulative
+    erosion potential (Fig.~\ref{fig:powerlaws}c). Finally, The erosion law
+    based on a global compilation of glacier velocity and erosion rates
     \citep[${\dot{e}=1.665\times10^{-1}u_\mathrm{b}^{0.6459}}$,][]{Cook.etal.2020}
-    results in a similarly high values, but a flatter (less localized) pattern
-    of cumulative erosion potential (Fig.~\ref{fig:powerlaws}c).
+    yields an even flatter pattern of cumulative erosion potential
+    (Fig.~\ref{fig:powerlaws}d).
 
@@
     more realistic cumulative erosion potential than those based on terrestrial
-    glaciers \citep{Herman.etal.2015, Cook.etal.2020}, themselves resulting
+    glaciers \citep{Humphrey.Raymond.1994, Herman.etal.2015, Cook.etal.2020},
+    themselves resulting
     in cumulative erosion potential values in the order of kilometres and thus
@@
     \citep[e.g.][]{Cook.etal.2020} would lead to a less localized pattern of
-    erosion (Fig.~\ref{fig:powerlaws}c), and a stronger correlation between
-    modelled ice volume and erosion \note{not shown, should it be?}. Even
+    erosion (Fig.~\ref{fig:powerlaws}d), and a stronger correlation between
+    modelled ice volume and erosion (Fig.~\ref{fig:powerlaws}l). Even
     then, the intrinsically large variations in glacier sliding velocities
     yields modelled glacial-cycle erosion potential two to three orders of
-    magnitude higher in the Alpine valley than near the mountain tops.
+    magnitude higher in the Alpine valley than near the mountain tops
+    (Fig.~\ref{fig:powerlaws}d).
 
@@
         running mean (dark brown) in relation to modelled total ice volume in
-        centimetres of sea-level equivalent (black).}
+        centimetres of sea-level equivalent.}
       \label{fig:evolution}
@@
         \textbf{(b)} \citet{Herman.etal.2015},
-          ${2.7 \times 10^{-7} u_\mathrm{b}^{2.02}}$, and
-        \textbf{(c)} \citet{Cook.etal.2020},
+          ${2.7 \times 10^{-7} u_\mathrm{b}^{2.02}}$,
+        \textbf{(c)} \citet{Humphrey.Raymond.1994},
+          ${1 \times 10^{-4} u_\mathrm{b}^{1}}$, and
+        \textbf{(d)} \citet{Cook.etal.2020},
           ${1.665 \times 10^{-1} u_\mathrm{b}^{0.6459}}$.
-        \textbf{(d)} Corresponding erosion power laws, and
-        \textbf{(e)} modelled cumulative (time-integrated) glacial erosion
-          potential along a Rhine Glacier transect (upper panels dashed line).}
+        \textbf{(e-h)} Corresponding modelled cumulative glacial erosion
+          potential along a Rhine Glacier transect (upper panels dashed line), and
+        \textbf{(i-l)} corresponding modelled evolution of volumic
+        (domain-integrated) erosion rate (light brown) and 100-a
+        running mean (dark brown) in relation to modelled total ice volume in
+        centimetres of sea-level equivalent (as in Fig.~\ref{fig:evolution}a).}
       \label{fig:powerlaws}

### 2021-02-03 articles/erosion/alps-erosion.tex
Change to Copernicus ESurf latex class.
Add author contrib, acknowledgements etc.

@@
 
-\documentclass[utf8]{article}
+\documentclass[esurf, manuscript]{copernicus}
 
-% default packages
-\usepackage{doi}
-\usepackage{authblk}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage[pdftex]{xcolor}
-\usepackage[pdftex]{graphicx}
-\usepackage[authoryear,round]{natbib}
-%\usepackage{CJK}  % texlive-langjapanese
+% custom colours (colorbrewer2.org)
+\definecolor{Bu}{cmyk}{1.00,0.45,0.00,0.07}  % Blues
+\definecolor{Gn}{cmyk}{1.00,0.20,1.00,0.00}  % Greens
+\definecolor{Rd}{cmyk}{0.35,0.95,0.85,0.00}  % Reds
+\definecolor{Or}{cmyk}{0.35,0.75,1.00,0.00}  % Oranges
+\definecolor{Pu}{cmyk}{0.70,0.80,0.00,0.00}  % Purples
+\definecolor{Br}{cmyk}{0.40,0.75,1.00,0.00}  % YlOrBr
 
-% review mode
-\usepackage{geometry}
-\usepackage{lineno}
-\linenumbers
-\linespread{1.5}
+% custom commands (remove before submission)
+\newcommand{\note}[1]{\textcolor{Or}{\emph{[\textbf{NOTE:} #1]}}}
+\newcommand{\todo}[1]{\textcolor{Rd}{\emph{[\textbf{TODO:} #1]}}}
+\newcommand{\aref}[0]{\textcolor{Rd}{\textbf{[REF.]}}}
+\newcommand{\ian}[1]{\textcolor{Or}{\emph{[\textbf{IAN:} #1]}}}
 
-% custom colours
-\definecolor{darkblue}{cmyk}{0.9,0.3,0.0,0.0}
-\definecolor{darkgreen}{cmyk}{0.8,0.0,1.0,0.0}
-\definecolor{darkred}{cmyk}{0.1,0.9,0.8,0.0}
-\definecolor{darkorange}{cmyk}{0.0,0.5,1.0,0.0}
-\definecolor{darkpurple}{cmyk}{0.6,0.7,0.0,0.0}
-\definecolor{darkbrown}{cmyk}{0.23,0.73,0.98,0.12}
-
-% custom commands
-\newcommand{\idea}[1]{\textcolor{darkgreen}{\emph{[\textbf{IDEA:} #1]}}}
-\newcommand{\note}[1]{\textcolor{darkblue}{\emph{[\textbf{NOTE:} #1]}}}
-\newcommand{\todo}[1]{\textcolor{darkred}{\emph{[\textbf{TODO:} #1]}}}
-\newcommand{\aref}[0]{\textcolor{darkblue}{\textbf{[REF.]}}}
-\newcommand{\ian}[1]{\textcolor{darkbrown}{\emph{[\textbf{IAN:} #1]}}}
+% color links (remove before submission)
+\hypersetup{colorlinks, citecolor=Bu, linkcolor=Bu, urlcolor=Bu}
 
@@
 \title{Last glacial cycle glacier erosion potential in the Alps}
-\author[1]{Julien Seguinot}
-\author[2]{Ian Delaney}
+\Author[1]{Julien}{Seguinot}
+\Author[2]{Ian}{Delaney}
 \affil[1]{Anafi, Greece}
 \affil[2]{Institute of Earth Surface Dynamics, University of Lausanne, Switzerland}
+\runningtitle{Last glacial cycle glacier erosion potential in the Alps}
+\runningauthor{J.~Seguinot and I.~Delanay}
 
@@
 % ----------------------------------------------------------------------
-\section{Introduction}
+\introduction
 % ----------------------------------------------------------------------
@@
 
-%\paragraph{Acknowledgements}
-%\paragraph{Author contributions}
-%\paragraph{Conflict of interest}
-%\paragraph{Contribution to the field}
-%\paragraph{Data availability}
+\codedataavailability{%
+    PISM is available as open-source software (http://pism-docs.org).
+    Time and domain-aggregated erosion variables such as used in the figures is
+    available for the four presented erosion power-laws
+    (\doi{10.5281/zenodo.4495419}). Other model output variables were
+    previously made available (\doi{10.5281/zenodo.1423160},
+    \doi{10.5281/zenodo.1423176}).}
+
+\videosupplement{
+    A supplementary animation is available
+    (\url{https://vimeo.com/503162771}).}
+
+\authorcontribution{%
+    J.~Seguinot performed the computations and initially drafted the
+    manuscript. I.~Delanay contributed to interpret the results in broader
+    context.}
+
+\competinginterests{%
+    The authors declare that they have no conflict of interest.}
+
+\begin{acknowledgements}
+    We are very thankful to Constantine Khroulev, Ed Bueler, and Andy
+    Aschwanden for their constant help and development with PISM, without whom
+    this work would not have been possible. J.~Seguinot would like to thank
+    John Jansen and Martin Margold for insightful early discussions, Susan
+    Ivy-Ochs for being a constant (droid) motivator, as well as Popi, Karsta
+    and Stefan for welcoming him on Anafi and providing agreeable conditions
+    for work in times of pandemic.
+    J.~Seguinot exclusively used gear transported on a bike for 5000\,km across
+    Europe and fourteen countries (except for new underwear).
+    The (previously published) ice flow simulations were made with computer
+    resources provided by the Swiss National Supercomputing Centre (CSCS)
+    allocations no.~s573 and sm13 to J.~Seguinot.
+\end{acknowledgements}
 
@@
 
-\bibliographystyle{abbrvnat}
+\bibliographystyle{copernicus}
 \bibliography{../../../references/references}

### 2021-02-03 articles/erosion/alps-erosion.tex
Merge additions from Ian (in brackets).

@@
 
-    While it has long been understood that glaciers leave a specific erosional
-    imprint on the landscape, observing and quantifying the long-term erosion
-    and sedimentation processes has been a challenge, not least due to the
-    inaccessibility of glacier beds and the slow and stochastic nature of
-    glacial erosion processes \aref. \ian{Perhaps this needs more detail?}
+    \ian{Furthermore, glacier erosion likely  governs, at least in part, the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through the ``glacier buzz-saw''.
+      In turn, evidence suggests that increase erosion rates coincided with global cooling and  Pliestocene glaciations from roughly 2.6\unit{Ma} Bp \citep{Herman.Champagnac.2016}.
+      Yet, other work suggests that global weather rates remained constant over this time period \citep{Willenbring.vonBlanckenburg2010}.
+      Examination of contemporary erosion rates across different climates suggest that increased temperatures lead to higher erosion rates \citep{Koppes.etal.2015,Fernandez.etal.2016}.
+      However, understanding the link between climate and glacier erosion in the sedimentary record remains difficult due to time scale biases \citep{Ganti.etal.2016} and the non-linear relationship
+      between temperature and erosion \citep[e.g.][]{Anderson.etal.2012,Mariotti.etal.2021}}
 
-    More recent studies have moved away from process
-    understanding towards attempts at calibrating empirical erosion laws for an entire
-    glacier or set of glaciers. \citet{Humphrey.Raymond.1994} correlated
-    temporal variations in ice velocity and suspended sediment load during a
-    surge of Variegated Glacier to establish a linear erosion law.
+    \ian{
+      Observing and quantifying the long-term erosion
+      and sedimentation processes has been a challenge, and two general methods have been adopted to quantify erosion.
+      Physically-based models describe quarrying process and abrasion of bedrock by debris laden ice \citep[e.g.][]{Alley.etal.1997,Iverson.2012,Beaud.etal.2014}.
+      Describing the processes physically provides a basis for understanding erosion processes \citep{Hallet.1979, Ugelvig.etal.2018} and yields insight into the formation of glacial landforms, such as tunnel valleys and eskers \citep{Beaud.etal.2018,Hewitt.Creyts.2019}.
+      Yet, these models prove difficult to implement in many cases due to the large number of poorly constrained parameters and process, i.e.
+water pressure flucuations and subglacial debris concentration \citep[e.g.][]{Hallet.1979,Ugelvig.etal.2018}.}
+
+    Instead, empirical relationships between glacier sliding and erosion can represent erosional quantities  well \citep[e.g.][]{Humphrey.Raymond.1994, Herman.etal.2015, Koppes.etal.2015} and can recreate important glacial landforms \citep[e.g.][]{Harbor.etal.1988,Macgregor.etal.2000}.
     \citet{Koppes.etal.2015} quantified sediment
@@
     glacier flow and glacier erosion processes, out results provide insights
-    into the diversity of the Alpine glacial erosion landscape.
+    into the diversity of the Alpine glacial erosion landscape. \ian{Furthermore,
+      we examine the glaciologcial and conditions that cause erosion and discuss
+      the implications of these conditions on understanding the relationship between climate and glacier erosion.  }
 
@@
 
+    \ian{It might be worth mentioning/discussing the role of ``till dilatation''
+      in the sliding relationship. There are a lot of interpretations, but
+      the erosion rule is thought to be a proxy for glacier abrasion of a glacier flowing
+      over a hard bed. Thus if there is a layer of till present it is arguable
+      if such abrasion can occur. Given the uncertainties, I doubt this factor
+      changes the story very much. However, it might be worth mentioning
+      something (somewhere) about how multiple processes are lumped together in
+      the sliding relationship or that till processes are uncertain due to the
+      unknown distribution of till. This could ease some reviewer concerns.  }
+
     Climate forcing is provided by a~monthly climatology from interpolated
@@
     velocities from 15~Patagonian and Antarctic Peninsula tidewater glaciers
-    \citep{Koppes.etal.2015}. Alternative erosion laws are analyzed in the
-    discussion part (Sect.~\ref{sec:powerlaws}).
+    \citep{Koppes.etal.2015}. \ian{We discuss the role of alternative values of
+    $K_\mathrm{g}$ and $l$ in Sect.~\ref{sec:powerlaws}.}
 
-    Our approach does not account for the feedback of glacier erosion onto
-    bedrock topography and ice dynamics. Neither are other forms of erosion
-    accounted for. Instead, we compute the time-integrated erosion rate,
+    \ian{Our approach does not account for the feedback of glacier erosion onto
+    bedrock topography and, then onto, ice dynamics \citep[e.g.][]{Anderson.etal.2012}.
+    Neither does the approach account for downstream advection of debris \citep{Anderson.Anderson.2016}
+    or fluvial transport of subglacial sediment \citep{Delaney.etal.2019}. \textbf{this last
+    sentence might not be that important}}
+
+    We compute the time-integrated erosion rate,
     %
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+\ian{Instead of saying ``volumetric erosion rate (m$^3$ a$^{-1}$)'', is it wiser to use
+  ``Annual erosion volume (m$^3$)'',
+  Also, in many papers there are plots of erosion rate over time. For one plot,
+  would it be good to plot erosion rate instead of volume (divide by area)?
+  This way some comparison can be easier made with other literature.
+  For instance, \citep{Stutenbecker.etal.2017} mentions a current erosion rate of $1.24$ mm a$^{-1}$.
+  Such a scaling might be good for comparison... or not if things are way off
+  and the message is much more about the relative change.
+}
 
@@
  \subsection{Deglaciation rapid erosion}
+ \label{sect:deglac_ero}
  % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+ \ian{This section title strange now, maybe \textbf{Rapid erosion during
+     deglaciation}. However, what about this?  \textbf{Implications for
+     understanding the relationship between erosion and climate}... I think
+   it is not only the deglaciation that is interesting, but also the periods before.}
 
@@
     power-law, it is in fact compatible with previous field-based studies that
-    have evidenced an inverse relationship between air temperatures and glacier
-    erosion rates \citep{Koppes.etal.2015, Cook.etal.2020}.
+    have evidenced an \ian{\textbf{do you mean inverse or direct? inverse}}
+    relationship between air temperatures and glacier erosion rates
+    \ian{in some cases} \citep{Koppes.etal.2015, Cook.etal.2020}.
+
+    \ian{Cold temperatures generally reduce sliding \citep{Koppes.etal.2015} and
+      if the glacier becomes cold-based then sliding, and thus erosion, will cease altogether.
+      Furthermore, in cases of advance, ice surfaces gradients are lower and cold ice is
+      advected downglacier, thus limiting erosion.}
 
@@
     glacier extent, yield the highest domain-integrated erosion rates of the
-    glacial cycle (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}). This
+    glacial cycle (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).
+    \ian{\textbf{I think it would be good to have a sentence about the
+        differences in glacier dynamics that cause the increase.}
+      Recent work, examining a denudation records, highlights that a non-linear forcing between
+      climate and glacier erosion persists \citep{Mariotti.etal.2021}. This paper
+      noted an increase in glacier erosion in the southern Alps, in the Var Valley,
+      roughly at $25$\,\unit{Ka} Bp following an extended period of low dunudation rates.
+      The increase in erosion here, predates the increase observed in our study
+      ($\sim 15$\,\unit{Ka}; Figs.~\ref{fig:evolution}). However, similar processes caused by climate warming
+      likely persist. This sedimentary record \citep{Mariotti.etal.2021} and the model results
+      here demonstrate that climatic conditions over-which glacier erosion occurs
+      are relatively specific and likely transient. }
+
+    \ian{I added some citations here. Increases in erosion following glacier retreat
+      seems to happen. But there are good critisims of the sedimentary record, and the mechanism
+      seems a little bit uncertain. This is one of the cool things about this paper, but also makes this paragarph
+    a bit delicate.}
+    This
     result is compatible with the field-based studies, which have evidenced
     increased glacier erosion and pro-glacial sedimentation during deglaciation
-    periods \aref, including the ongoing glacier retreat \aref. Yet this
-    increase has often been attributed to higher meltwater availability,
-    promoting river erosion, enhanced glacier sliding, and hydrofracturing
-    (plucking) \aref.
+    periods \citep[e.g.][]{Koppes.Montgomery.2009,Micheletti.Lane.2016,Lane.etal.2017}, including the ongoing glacier retreat
+    \citep[e.g.][]{Bendixen.etal.2017}. Yet this increase has often been attributed
+    to higher meltwater availability, promoting sediment transport \citep{Delaney.Adhikari.2020},
+    enhanced glacier sliding \citep[][This citation might a bit strange]{Herman.etal.2011}, and hydrofracturing
+    (plucking) \citep[][This citation is more about monthly to diurnal variability]{Ugelvig.etal.2018}.
 
@@
     model, so that deglaciation surface meltwater is assumed to instantly exit
-    the glacier with no effect on glacier dynamics. The enhanced glacier
+    the glacier with no effect on glacier dynamics \ian{ \citep[e.g][]{Werder.etal.2013}
+      or glacier erosion  \citep[][]{Iverson.etal.2012,Ugelvig.etal.2018}}.
+    The enhanced glacier
     velocity during periods of ice retreat is here modelled to result from
@@
     sliding, not included in the model, presumably further accelerated erosion
-    during deglaciation. Rapid glacier erosion can be expected to accompany the
-    ongoing 21st-century deglaciation phase.
+    during deglaciation \citep{Herman.etal.2011}. \ian{Furthermore, we omit the
+      fluvial transport of sediment, which could well increase sediment discharge
+      and erosion with increased melt, so long as additional sediment remains
+      available \citep{Delaney.Adhikari.2020}. Rapid glacier erosion can be expected
+    to accompany the  ongoing 21st-century deglaciation phase. This trend might be
+    especially strong on the ice sheets of Greenland and Antarctica where
+    polythermal and cold-based glaciers are warming and sliding is increasing
+    \citep[e.g.][]{Moon.etal.2012,Overeem.etal.2017}.}
 
@@
     modelled to date from expansive glaciation stages, but instead to stages of
-    intermediate glacier cover (Fig.~\ref{fig:hypsogram}).
+    intermediate glacier cover \citep[Fig.~\ref{fig:hypsogram}; ][]{Barr.etal.2019}.
 
@@
     here modelled to be the product of extensive, Last Glacial Maximum-like
-    glaciations. Alpine terrain with elevations above 1000\,m are modelled to
+    glaciations. \ian{Other modeling work suggests that features such as these
+      likely form when water can access the glacier bed and encourage glacier
+      sliding and, thus, erosion \citep{Herman.etal.2011}. The model herein does not
+      include subglacial hydrology, the role of climate warming to concentrating
+      glacier erosion to the lower elevations and alpine remains evident.}
+
+    Alpine terrain with elevations above 1000\,m are modelled to
     experience virtually no erosion from 35 to 18\,ka BP
@@
     cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would date from
-    temperate periods with limited ice-cover such as the present day.
+    temperate periods with limited ice-cover such as the present day \ian{
+      or over interglacial periods, when glaciers are confined to their cirques
+      \citep{Barr.etal.2017,Barr.etal.2019}}.
 
@@
         ranging from pan-Alpine ice sheets to modern-style mountain glaciers.
+      \item \ian{The high spatialtemporal variability  in erosion suggests that
+          the relationship between climate and glacier erosion remains greatly complex.
+          As a result, a highly variable response of glacier erosion to climate forcings should be expected.}
     \end{itemize}

### 2021-02-03 articles/erosion/alps-erosion.tex
some last changes by Ian

@@
     glacial cycle, but show paradoxically little relation to the glaciation
-    extent. Phases of glacier advance and maximum extension, characterized by
+    extent \ian{Not to fine of a point, but in the text, it is about ice volume,
+      and not so much ice extent}. Phases of glacier advance and maximum extension, characterized by
     low air and glacier temperatures, see a localization of rapid erosion rates
@@
     deglaciation phases, when higher air and glacier temperature enhance
-    glacier dynamics.
+    glacier dynamics. \ian{These result that the climatic conditions that
+    result in glacier erosion might be quite specific. \textbf{or something like that}}
     %
@@
       In turn, evidence suggests that increase erosion rates coincided with global cooling and  Pliestocene glaciations from roughly 2.6\unit{Ma} Bp \citep{Herman.Champagnac.2016}.
-      Yet, other work suggests that global weather rates remained constant over this time period \citep{Willenbring.vonBlanckenburg2010}.
-      Examination of contemporary erosion rates across different climates suggest that increased temperatures lead to higher erosion rates \citep{Koppes.etal.2015,Fernandez.etal.2016}.
-      However, understanding the link between climate and glacier erosion in the sedimentary record remains difficult due to time scale biases \citep{Ganti.etal.2016} and the non-linear relationship
-      between temperature and erosion \citep[e.g.][]{Anderson.etal.2012,Mariotti.etal.2021}}
+      Yet, other work suggests that global weathering rates remained constant over this time period \citep{Willenbring.vonBlanckenburg2010}.
+      Examination of contemporary erosion rates across different climates suggest that increased temperatures in glaciated regions lead to higher erosion rates and that erosion rates may have increased towards present \citep{Koppes.Montgomery2009,Koppes.etal.2015,Fernandez.etal.2016}.
+      However, understanding the link between climate and glacier erosion in the sedimentary record remains difficult due to time scale biases \citep{Ganti.etal.2016} and the non-linear relationship  between temperature and erosion \citep[e.g.][]{Anderson.etal.2012,Mariotti.etal.2021}}
 
@@
       Observing and quantifying the long-term erosion
-      and sedimentation processes has been a challenge, and two general methods have been adopted to quantify erosion.
+      and sedimentation processes has been a challenge, and two general methods have been adopted to quantify erosion \citep{Alley.etal.2019}.
       Physically-based models describe quarrying process and abrasion of bedrock by debris laden ice \citep[e.g.][]{Alley.etal.1997,Iverson.2012,Beaud.etal.2014}.
@@
     into the diversity of the Alpine glacial erosion landscape. \ian{Furthermore,
-      we examine the glaciologcial and conditions that cause erosion and discuss
+      we examine the glaciologcial conditions that cause erosion and discuss
       the implications of these conditions on understanding the relationship between climate and glacier erosion.  }
@@
     bedrock topography and, then onto, ice dynamics \citep[e.g.][]{Anderson.etal.2012}.
-    Neither does the approach account for downstream advection of debris \citep{Anderson.Anderson.2016}
+    Neither does the approach account for downstream advection of debris \citep[e.g.][]{Anderson.Anderson.2016}
     or fluvial transport of subglacial sediment \citep{Delaney.etal.2019}. \textbf{this last
@@
 \ian{Instead of saying ``volumetric erosion rate (m$^3$ a$^{-1}$)'', is it wiser to use
-  ``Annual erosion volume (m$^3$)'',
+  ``Annual erosion volume (m$^3$)''?,
   Also, in many papers there are plots of erosion rate over time. For one plot,
@@
       available \citep{Delaney.Adhikari.2020}. Rapid glacier erosion can be expected
-    to accompany the  ongoing 21st-century deglaciation phase. This trend might be
+    to accompany the  ongoing 21st-century deglaciation phase. This process might be
     especially strong on the ice sheets of Greenland and Antarctica where
@@
       or over interglacial periods, when glaciers are confined to their cirques
-      \citep{Barr.etal.2017,Barr.etal.2019}}.
+      \citep{Barr.etal.2017,Barr.etal.2019}}. Similar processes also occur on
+    Himalayan glaciers and may limit erosion for the high elevation portion of
+    the mountain range during cold periods as well, failing to offset uplift
+    rates \citep{Harper.Humphrey.2003}.
 
@@
           the relationship between climate and glacier erosion remains greatly complex.
-          As a result, a highly variable response of glacier erosion to climate forcings should be expected.}
+          As a result, a highly variable response of glacier erosion to climate forcings should be expected.
+        \textbf{this last point could be somewhat coupled to the first point.}}
     \end{itemize}
@@
     J.~Seguinot performed the computations and initially drafted the
-    manuscript. I.~Delanay contributed to interpret the results in broader
+    manuscript. I.~Delaney contributed to interpret the results in broader
     context.}

### 2021-02-03 articles/erosion/alps-erosion.tex
Fix a few reference keys in manuscript.

@@
       In turn, evidence suggests that increase erosion rates coincided with global cooling and  Pliestocene glaciations from roughly 2.6\unit{Ma} Bp \citep{Herman.Champagnac.2016}.
-      Yet, other work suggests that global weathering rates remained constant over this time period \citep{Willenbring.vonBlanckenburg2010}.
-      Examination of contemporary erosion rates across different climates suggest that increased temperatures in glaciated regions lead to higher erosion rates and that erosion rates may have increased towards present \citep{Koppes.Montgomery2009,Koppes.etal.2015,Fernandez.etal.2016}.
+      Yet, other work suggests that global weathering rates remained constant over this time period \citep{Willenbring.Von-Blanckenburg.2010}.
+      Examination of contemporary erosion rates across different climates suggest that increased temperatures in glaciated regions lead to higher erosion rates and that erosion rates may have increased towards present \citep{Koppes.Montgomery.2009,Koppes.etal.2015,Fernandez.etal.2016}.
       However, understanding the link between climate and glacier erosion in the sedimentary record remains difficult due to time scale biases \citep{Ganti.etal.2016} and the non-linear relationship  between temperature and erosion \citep[e.g.][]{Anderson.etal.2012,Mariotti.etal.2021}}
@@
     the glacier with no effect on glacier dynamics \ian{ \citep[e.g][]{Werder.etal.2013}
-      or glacier erosion  \citep[][]{Iverson.etal.2012,Ugelvig.etal.2018}}.
+      or glacier erosion  \citep[][]{Iverson.2012,Ugelvig.etal.2018}}.
     The enhanced glacier

### 2021-02-03 articles/erosion/alps-erosion.tex
Replace volumetric rate with annual volume.

@@
 
-    The modelled volumetric (domain-integrated) glacial erosion rates
-    do no systematically correlate with the modelled total ice volume
+    The modelled annual (domain-integrated) glacial erosion volume
+    does no systematically correlate with the modelled total ice volume
     (Fig.~\ref{fig:cumulative}b). Except for the onset and termination of the
@@
     captured by the model resolution, a comparison between the modelled ice
-    volume and the modelled volumetric erosion rate shows no single relation
+    volume and the modelled annual erosion volume shows no single relation
     between these two quantities (Fig.~\ref{fig:evolution}). After significant
@@
         \textbf{(b)} Modelled total ice volume in centimetres of sea-level
-          equivalent (cm~s.l.e., black), volumic (domain-integrated) erosion
-          rate (light brown) and 100-a running mean (dark brown). Shaded gray
+          equivalent (cm~s.l.e., black), annual (domain-integrated) erosion
+          volume (light brown) and 100-a running mean (dark brown). Shaded gray
           areas indicate the timing for MIS~2 and~4
@@
       \caption{%
-        Modelled volumic (domain-integrated) erosion rate (light brown) and 100-a
+        Modelled annual (domain-integrated) erosion volume (light brown) and 100-a
         running mean (dark brown) in relation to modelled total ice volume in
@@
           potential along a Rhine Glacier transect (upper panels dashed line), and
-        \textbf{(i-l)} corresponding modelled evolution of volumic
-        (domain-integrated) erosion rate (light brown) and 100-a
+        \textbf{(i-l)} corresponding modelled evolution of annual
+        (domain-integrated) erosion volume (light brown) and 100-a
         running mean (dark brown) in relation to modelled total ice volume in

### 2021-02-03 figures/alpero_evolution.py

@@ -18,6 +18,7 @@ def figure():
         left=12, right=1.5, bottom=9, top=1.5))
 
     # load postprocessed data
+    # FIXME there is a data gap after merge, add tolerance
     with pismx.open.dataset(
             '../data/processed/alpcyc.1km.epic.pp.ts.10a.nc') as ds:
         ds = ds[['slvol']]
@@ -37,7 +38,7 @@ def figure():
 
     # set axes properties
     ax.set_xlabel('ice volume (cm s.l.e.)')
-    ax.set_ylabel(r'volumic erosion rate ($m^3\,a^{-1}$)')
+    ax.set_ylabel(r'annual erosion volume ($m^3 a^{-1}$)')
     ax.set_yscale('log')
     ax.set_ylim(10**4.3, 10**7.7)
 

### 2021-02-03 figures/util/ero.py

@@ -37,5 +37,5 @@ def plot_series(ax=None, run='1km.epic.pp'):
         ax = ax.twinx()
         ax.plot(ds.age, ds.kop2015_rate*1e-6, c='C11', alpha=0.5)
         ax.plot(ds.age, ds.rolling_mean*1e-6, c='C11')
-        ax.set_ylabel('volumic erosion\n'+r'rate ($10^6\,m^3\,a^{-1}$)', color='C11')
+        ax.set_ylabel('annual erosion\n'+r'volume ($10^6\,m^3 a^{-1}$)', color='C11')
         ax.set_ylim(-1, 7)

### 2021-02-04 articles/erosion/alps-erosion.tex
Rework abstract and integrate Ian's comments.

@@
     glacial valleys and large over-deepened basins. Using previous glacier
-    modelling results, and empirical inferences bedrock erosion under modern
+    modelling results, and empirical inferences of bedrock erosion under modern
     glaciers, we compute a distribution of potential glacier erosion in the Alps
@@
     %
-    Despite large uncertainties related to the climate history of the Alps and
+    Despite large uncertainties pertaining to the climate history of the Alps and
     unconstrained glacier erosion processes, the resulting modelled patterns of
-    glacier erosion shows persistent features. The cumulative imprint of
+    glacier erosion includes persistent features. The cumulative imprint of
     the last glacial cycle shows a very strong localization of glacier erosion
@@
     highest velocity. The modelled erosion rates vary significantly through the
-    glacial cycle, but show paradoxically little relation to the glaciation
-    extent \ian{Not to fine of a point, but in the text, it is about ice volume,
-      and not so much ice extent}. Phases of glacier advance and maximum extension, characterized by
+    glacial cycle, but show paradoxically little relation to the total glacier
+    volume. Phases of glacier advance and maximum extension, characterized by
     low air and glacier temperatures, see a localization of rapid erosion rates
-    at low elevation, while intra-montane areas are modelled to date from phases
-    of intermediate glacier cover. The modelled erosion rates peak during
-    deglaciation phases, when higher air and glacier temperature enhance
-    glacier dynamics. \ian{These result that the climatic conditions that
-    result in glacier erosion might be quite specific. \textbf{or something like that}}
+    at low elevation, while the intra-montane glacial landscape is modelled to
+    date from phases of intermediate glacier cover. The modelled erosion rates
+    peak during deglaciation phases, when higher air and glacier temperature
+    enhance glacier dynamics, implying that climatic conditions that result in
+    rapid glacier erosion might be quite specific.
     %

### 2021-02-04 articles/erosion/alps-erosion.tex
Rework intro and integrate Ian's comments.

@@
 \documentclass[esurf, manuscript]{copernicus}
+\usepackage{CJK}  % Japanese text
 
@@
     New Zealand (Southern Alps), Japan
-    (\emph{%\begin{CJK}{UTF8}{min}日本アルプス\end{CJK},
+    (\emph{\begin{CJK}{UTF8}{min}日本アルプス\end{CJK},
     Nihon Arupusu}), and elsewhere.
@@
     be understood.
-    In absence of other glacial landforms, the degree of glacial modification
-    of a landscape has sometimes been used as a proxy for the duration of past
-    glacier cover, as for instance in geomorphological mapping
-    \citep[e.g.][]{Margold.etal.2011, Fu.etal.2012}. Cold-based glaciers have
-    been shown to preserve landforms as fragile as sand beaches
-    \citep{Kleman.1994}. Yet it remains unclear how sharp the erosional
+
+    In absence of other landforms, glacially eroded topography has sometimes
+    been used as a proxy for glacier extension, for instance in
+    geomorphological mapping \citep[e.g.,][]{Margold.etal.2011, Fu.etal.2012}.
+    Cold-based glaciers, however, have been observed to preserve landforms as
+    fragile as sand beaches \citep{Kleman.1994}, leading to the
+    reinterpretation of complex glacial landscapes within a binary conceptual
+    framework of cold and temperate basal thermal regimes
+    \citep{Kleman.etal.2008, Kleman.etal.2010, Fabel.etal.2012, Fu.etal.2013}.
+    Nevertheless, it remains unclear how sharp the erosional
     transition to cold-based ice is, and why some glaciated regions show
@@
 
-    \ian{Furthermore, glacier erosion likely  governs, at least in part, the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through the ``glacier buzz-saw''.
-      In turn, evidence suggests that increase erosion rates coincided with global cooling and  Pliestocene glaciations from roughly 2.6\unit{Ma} Bp \citep{Herman.Champagnac.2016}.
-      Yet, other work suggests that global weathering rates remained constant over this time period \citep{Willenbring.Von-Blanckenburg.2010}.
-      Examination of contemporary erosion rates across different climates suggest that increased temperatures in glaciated regions lead to higher erosion rates and that erosion rates may have increased towards present \citep{Koppes.Montgomery.2009,Koppes.etal.2015,Fernandez.etal.2016}.
-      However, understanding the link between climate and glacier erosion in the sedimentary record remains difficult due to time scale biases \citep{Ganti.etal.2016} and the non-linear relationship  between temperature and erosion \citep[e.g.][]{Anderson.etal.2012,Mariotti.etal.2021}}
+    On longer time scales, glacier erosion likely governs, at least in part,
+    the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through
+    the ``glacier buzz-saw''. In turn, evidence suggests that increased
+    mountain erosion rates coincided with global cooling and Pliestocene
+    glaciations from roughly 2.6\,Ma (million years before present)
+    \citep{Herman.Champagnac.2016}. Yet, other work suggests that global
+    weathering rates remained constant over this time period
+    \citep{Willenbring.Von-Blanckenburg.2010}. \note{Evidence suggests, other
+    work suggests, maybe this can be reworked.} Examination of contemporary
+    erosion rates across different climates suggest that increased temperatures
+    in glaciated regions lead to higher erosion rates and that erosion rates
+    may have increased towards present \citep{Koppes.Montgomery.2009,
+    Koppes.etal.2015, Fernandez.etal.2016}. However, understanding the link
+    between climate and glacier erosion in the sedimentary record remains
+    difficult due to time scale biases \citep{Ganti.etal.2016} and the
+    non-linear relationship  between temperature and erosion
+    \citep[e.g.,][]{Anderson.etal.2012,Mariotti.etal.2021}.
 
-    \ian{
-      Observing and quantifying the long-term erosion
-      and sedimentation processes has been a challenge, and two general methods have been adopted to quantify erosion \citep{Alley.etal.2019}.
-      Physically-based models describe quarrying process and abrasion of bedrock by debris laden ice \citep[e.g.][]{Alley.etal.1997,Iverson.2012,Beaud.etal.2014}.
-      Describing the processes physically provides a basis for understanding erosion processes \citep{Hallet.1979, Ugelvig.etal.2018} and yields insight into the formation of glacial landforms, such as tunnel valleys and eskers \citep{Beaud.etal.2018,Hewitt.Creyts.2019}.
-      Yet, these models prove difficult to implement in many cases due to the large number of poorly constrained parameters and process, i.e.
-water pressure flucuations and subglacial debris concentration \citep[e.g.][]{Hallet.1979,Ugelvig.etal.2018}.}
+    Observing and quantifying the long-term glacier erosion and sedimentation
+    processes has been a challenge, and two general methods have been adopted
+    to quantify erosion \citep{Alley.etal.2019}. Physically-based models
+    describe quarrying process and abrasion of bedrock by debris laden ice
+    \citep[e.g.,][]{Alley.etal.1997, Iverson.2012, Beaud.etal.2014}. Describing
+    the processes physically provides a basis for understanding erosion
+    processes \citep{Hallet.1979, Ugelvig.etal.2018} and yields insight into
+    the formation of glacial landforms, such as tunnel valleys and eskers
+    \citep{Beaud.etal.2018, Hewitt.Creyts.2019}. Yet, these models prove
+    difficult to implement in many cases due to the large number of poorly
+    constrained parameters and process, i.e. water pressure flucuations and
+    subglacial debris concentration \citep[e.g.,][]{Hallet.1979, Seguinot.2008,
+    Ugelvig.etal.2018}.
 
-    Instead, empirical relationships between glacier sliding and erosion can represent erosional quantities  well \citep[e.g.][]{Humphrey.Raymond.1994, Herman.etal.2015, Koppes.etal.2015} and can recreate important glacial landforms \citep[e.g.][]{Harbor.etal.1988,Macgregor.etal.2000}.
+    Instead, empirical relationships between glacier sliding and erosion can
+    represent erosional quantities well \citep[e.g.,][]{Humphrey.Raymond.1994,
+    Herman.etal.2015, Koppes.etal.2015} and can recreate important glacial
+    landforms \citep[e.g.,][]{Harbor.etal.1988, Macgregor.etal.2000}.
     \citet{Koppes.etal.2015} quantified sediment
@@
     patterns of modelled erosion rate and cumulative last glacial cycle erosion
-    potential are analysed. Despite aggregated uncertainties on paleoclimate,
-    glacier flow and glacier erosion processes, out results provide insights
-    into the diversity of the Alpine glacial erosion landscape. \ian{Furthermore,
-      we examine the glaciologcial conditions that cause erosion and discuss
-      the implications of these conditions on understanding the relationship between climate and glacier erosion.  }
+    potential are analysed. We examine the glaciologcial conditions that cause
+    erosion and discuss the implications of these conditions on understanding
+    the relationship between climate and glacier erosion, Despite aggregated
+    uncertainties on paleoclimate, glacier flow and glacier erosion processes,
+    out results provide insights into the diversity of the Alpine glacial
+    erosion landscape.
 

### 2021-02-04 articles/erosion/alps-erosion.tex
Rework the methods and integrate Ian's comments.

@@
     set-up is stored in long-term archived model output metadata for the
-    present study \todo{upload}, and for the earlier publication
+    present study \citep{Seguinot.2021}, and for the earlier publication
     \citep{Seguinot.2020, Seguinot.2020a}.
@@
     velocities from 15~Patagonian and Antarctic Peninsula tidewater glaciers
-    \citep{Koppes.etal.2015}. \ian{We discuss the role of alternative values of
-    $K_\mathrm{g}$ and $l$ in Sect.~\ref{sec:powerlaws}.}
+    \citep{Koppes.etal.2015}. We discuss the role of alternative values of
+    $K_\mathrm{g}$ and $l$ in the discussion part (Sect.~\ref{sec:powerlaws}).
 
-    \ian{Our approach does not account for the feedback of glacier erosion onto
-    bedrock topography and, then onto, ice dynamics \citep[e.g.][]{Anderson.etal.2012}.
-    Neither does the approach account for downstream advection of debris \citep[e.g.][]{Anderson.Anderson.2016}
-    or fluvial transport of subglacial sediment \citep{Delaney.etal.2019}. \textbf{this last
-    sentence might not be that important}}
-
-    We compute the time-integrated erosion rate,
+    Our approach does not account for the feedback of glacier erosion onto
+    bedrock topography and, then onto, ice dynamics
+    \citep[e.g.,][]{Anderson.etal.2012}. Neither does the approach account for
+    downstream advection of debris \citep[e.g.,][]{Anderson.Anderson.2016} that
+    may have shed the bedrock from erosion, or fluvial transport of subglacial
+    sediment \citep{Delaney.etal.2019}. Instead, we compute the time-integrated
+    glacier erosion rate,
     %

### 2021-02-04 articles/erosion/alps-erosion.tex
Rework the results and integrate Ian's comments.

@@
     cirques are poorly captured by the model horizontal resolution, high
-    cumulative erosion potential also occur near the valley heads.
+    cumulative erosion potential also occurs near the valley heads.
     There is a general tendency for higher cumulative erosion in the
-    north-western Alps where the input winter precipitation
-    \citep[WorldClim,][Fig.~1h]{Seguinot.etal.2018} is higher and the
-    glacial relief more pronounced in the input topography.
+    north-western Alps where the input winter precipitation is higher
+    \citep[WorldClim, Fig.~1h in][]{Seguinot.etal.2018} and the glacial relief
+    more pronounced in the input topography.
 
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\ian{Instead of saying ``volumetric erosion rate (m$^3$ a$^{-1}$)'', is it wiser to use
-  ``Annual erosion volume (m$^3$)''?,
-  Also, in many papers there are plots of erosion rate over time. For one plot,
-  would it be good to plot erosion rate instead of volume (divide by area)?
-  This way some comparison can be easier made with other literature.
-  For instance, \citep{Stutenbecker.etal.2017} mentions a current erosion rate of $1.24$ mm a$^{-1}$.
-  Such a scaling might be good for comparison... or not if things are way off
-  and the message is much more about the relative change.
-}
+
+    \note{I have replaced ``volumetric erosion rate'' with ``annual erosion
+          volume'' but kept the unit as ``m$^3$\,a$^{-1}$''. We agreed to stick
+          with erosion volume rather than average rates for now. Should the
+          erosion rate of 1.24\,mm\,a$^{-1}$ \citep{Stutenbecker.etal.2017} be
+          discussed?}
 
     The modelled annual (domain-integrated) glacial erosion volume
-    does no systematically correlate with the modelled total ice volume
+    does not systematically correlate with the modelled total ice volume
     (Fig.~\ref{fig:cumulative}b). Except for the onset and termination of the
@@
     during periods of extensive glaciation (Fig.~\ref{fig:evolution}). However,
-    the domain-integrated erosion rate is modelled to be consistently higher
+    the domain-integrated erosion volume is modelled to be consistently higher
     (by a factor 2 to 10) during periods of decreasing ice volume, than during
@@
     A closer look at the Rhine Glacier, one of the paleo-ice sheet's largest
-    outlets, reveals a spatial migration of the modelled high erosion rates.
+    outlets, reveals a spatial migration of the modelled rapid erosion areas.
     During stages of glacier advance and maximum extension, erosion is modelled
@@
     the intra-montane Alps \citep[modelled to be largely
-    cold-based,][Fig.~6c]{Seguinot.etal.2018}, are modelled to experience
+    cold-based, Fig.~6c in][]{Seguinot.etal.2018}, are modelled to experience
     insignificant erosion (Fig.~\ref{fig:transects}a and~d). The modelled

### 2021-02-04 articles/erosion/alps-erosion.tex
Rework sensitivity sections with Ian's comments.

@@
     along-flow distance (Fig.~\ref{fig:hypsogram}). Thus more generally,
-    periods of modelled increasing and maximum ice volume corresponds to lower
+    periods of modelled increasing and maximum ice volume correspond to lower
     values for elevation-aggregated erosion rates, with significant erosion
@@
     Different paleoclimate forcing scenarios lead to different modelled
-    glaciation histories and thus different results in terms of erosion
+    glaciation histories \citep[cf.][]{Seguinot.etal.2018} and thus different
+    results in terms of erosion
     potential (Fig.~\ref{fig:sensitivity}). Scenarios including higher
-    (constant) input precipitation rates yield generally higher modelled
+    (constant, present-day) input precipitation rates yield generally higher
+    modelled
     erosion potential (Fig.~\ref{fig:sensitivity}a, b, and~c), while scenarios
     including paleo-precipitation reductions (Fig.~\ref{fig:sensitivity}d, e,
-    and~f), including our default high-resolution run, yield a smaller erosion
-    potential. The spatial pattern of modelled erosion potential is
-    nevertheless generally consistent across paleoclimate scenarios tested,
+    and~f), including our default high-resolution run
+    (Fig.~\ref{fig:cumulative}), yield a smaller erosion potential. The spatial
+    pattern of modelled time-integrated erosion is nevertheless generally
+    consistent across the paleoclimate scenarios tested,
     despite variations in the glaciation maximum extent
-    (Fig.~\ref{fig:sensitivity}). Furthermore, these changes are minor
+    (Fig.~\ref{fig:sensitivity} and \citealp[Fig.~3 in][]{Seguinot.etal.2018}).
+    Furthermore, these changes are minor
     considering the large uncertainties affecting the choice of erosion law.
@@
     The choice of erosion law significantly impacts the results
-    (Fig.~\ref{fig:powerlaws}). Our default erosion law, based on quantified
+    (Fig.~\ref{fig:powerlaws}). Our default, based on quantified
     sediment yields from Patagonian and Antarctic Peninsula tidewater glaciers
@@
     suspended sediment load during a surge of Variegated Glacier
-    \citep[${\dot{e}=1\times10^{-4}u_\mathrm{b}^{1}}$,][]{Humphrey.Raymond.1994},
+    \citep[${\dot{e}=1\times10^{-4}u_\mathrm{b}^{1}}$,][]{Humphrey.Raymond.1994}
     results in similarly high values, but a less localized pattern of cumulative
     erosion potential (Fig.~\ref{fig:powerlaws}c). Finally, The erosion law
-    based on a global compilation of glacier velocity and erosion rates
+    based on a global compilation of glacier velocities and erosion rates
     \citep[${\dot{e}=1.665\times10^{-1}u_\mathrm{b}^{0.6459}}$,][]{Cook.etal.2020}
@@
     Antarctic Peninsula tidewater glaciers \citep{Koppes.etal.2015} yields a
-    more realistic cumulative erosion potential than those based on terrestrial
+    more realistic glacial cycle erosion potential in the Alps than those based
+    on terrestrial
     glaciers \citep{Humphrey.Raymond.1994, Herman.etal.2015, Cook.etal.2020},
-    themselves resulting
-    in cumulative erosion potential values in the order of kilometres and thus
+    themselves resulting in kilometre-scale integrated erosion even
     higher than the Alpine relief itself. However, during the Last Glacial
     Maximum and much of the last glacial cycle, Alpine paleoglaciers were
-    closer in size and climatic context to the present-day glaciers of
+    closer in size, slope, and climatic context to the present-day glaciers of
     Patagonia and the Antarctic Peninsula \citep{Koppes.etal.2015} than to
     Franz-Joseph Glacier \citep{Herman.etal.2015} and many of the glaciers
-    included in the global compilation by \citet{Cook.etal.2020}. Atmospheric
+    included in the global compilation by \citet{Cook.etal.2020}. Indeed,
+    atmospheric
     temperatures have been found to be a strong driver of erosion in studies
@@
     as a criteria to rule out this or that erosion law for the glacial Alps, as
-    all these laws are affected with very large uncertainties. It can not be
-    excluded that a linear or sub-linear erosion-law
-    \citep[e.g.][]{Cook.etal.2020} would lead to a less localized pattern of
-    erosion (Fig.~\ref{fig:powerlaws}d), and a stronger correlation between
-    modelled ice volume and erosion (Fig.~\ref{fig:powerlaws}l). Even
-    then, the intrinsically large variations in glacier sliding velocities
-    yields modelled glacial-cycle erosion potential two to three orders of
+    all these laws are affected with very large uncertainties. Thus it can not
+    be excluded that a linear or sub-linear erosion-law
+    \citep[e.g.,][]{Cook.etal.2020} would lead to a less localized pattern of
+    erosion (Fig.~\ref{fig:powerlaws}c and~d), and a stronger correlation
+    between modelled ice volume and erosion (Fig.~\ref{fig:powerlaws}k and~l).
+    Even then, the intrinsically large variations in modelled glacier sliding
+    velocities yield cumulative glacial-cycle erosion two to three orders of
     magnitude higher in the Alpine valley than near the mountain tops
     (Fig.~\ref{fig:powerlaws}d).
+
+    \todo{Insert discussion of sediments here.}
 

### 2021-02-04 articles/erosion/alps-erosion.tex
Rework climate discussion with Ian's comments.

@@
     enhance glacier dynamics, implying that climatic conditions that result in
-    rapid glacier erosion might be quite specific.
+    rapid glacier erosion might be quite transient and specific.
     %
@@
     is initialized with assumed present-day ice thickness and equilibrium
-    ice and bedrock temperature at 120\,ka, and ran to the present.
+    ice and bedrock temperature at 120\,ka (thousand years before the present),
+    and ran to the present.
     The ice-sheet model physical parameters are listed in the earlier
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
- \subsection{Deglaciation rapid erosion}
- \label{sect:deglac_ero}
+\subsection{Climate control on erosion}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
- \ian{This section title strange now, maybe \textbf{Rapid erosion during
-     deglaciation}. However, what about this?  \textbf{Implications for
-     understanding the relationship between erosion and climate}... I think
-   it is not only the deglaciation that is interesting, but also the periods before.}
 
-    The modelled, domain-integrated glacier erosion rates do not increase
+    \note{What about this new section title?}
+
+    The modelled, domain-integrated glacier erosion volume do not increase
     together with ice volume. On the opposite, periods of ice advance and
@@
     power-law, it is in fact compatible with previous field-based studies that
-    have evidenced an \ian{\textbf{do you mean inverse or direct? inverse}}
-    relationship between air temperatures and glacier erosion rates
-    \ian{in some cases} \citep{Koppes.etal.2015, Cook.etal.2020}.
+    have evidenced an increase of erosion rates with air temperatures for at
+    least some sets of glaciers \citep{Koppes.etal.2015, Cook.etal.2020}.
 
-    \ian{Cold temperatures generally reduce sliding \citep{Koppes.etal.2015} and
-      if the glacier becomes cold-based then sliding, and thus erosion, will cease altogether.
-      Furthermore, in cases of advance, ice surfaces gradients are lower and cold ice is
-      advected downglacier, thus limiting erosion.}
+    Cold temperatures generally reduce sliding \citep{Koppes.etal.2015} and if
+    the glacier becomes cold-based then sliding, and thus erosion, will cease
+    altogether. In cases of glacier advance, newly formed, atmosphere-cold ice
+    expands onto cooled bedrock hindering ice deformation and basal sliding.
+    \note{Ian: ``Furthermore, in cases of advance, ice surfaces
+          gradients are lower and cold ice is advected downglacier, thus
+          limiting erosion''. Is this also valid for relatively low-sloping
+          piedmont glaciers?}
+    On the other hand, deglaciation periods are characterised by a warmer
+    climate and above-balance glacier extent. Glacier ice formed an isolating
+    blanket from the atmosphere so that geothermal heat has had time to warm up
+    the lower layers and activate the double-feedback loop between ice
+    temperature, deformation heat dissipation and basal melt-induced sliding.
+    Warmer air temperatures contribute to enhance deformation and sliding even
+    more, yielding the highest domain-integrated erosion rates of the
+    glacial cycle (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).
 
-    Deglaciation periods, characterised by a warmer climate and above-balance
-    glacier extent, yield the highest domain-integrated erosion rates of the
-    glacial cycle (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).
-    \ian{\textbf{I think it would be good to have a sentence about the
-        differences in glacier dynamics that cause the increase.}
-      Recent work, examining a denudation records, highlights that a non-linear forcing between
-      climate and glacier erosion persists \citep{Mariotti.etal.2021}. This paper
-      noted an increase in glacier erosion in the southern Alps, in the Var Valley,
-      roughly at $25$\,\unit{Ka} Bp following an extended period of low dunudation rates.
-      The increase in erosion here, predates the increase observed in our study
-      ($\sim 15$\,\unit{Ka}; Figs.~\ref{fig:evolution}). However, similar processes caused by climate warming
-      likely persist. This sedimentary record \citep{Mariotti.etal.2021} and the model results
-      here demonstrate that climatic conditions over-which glacier erosion occurs
-      are relatively specific and likely transient. }
-
-    \ian{I added some citations here. Increases in erosion following glacier retreat
-      seems to happen. But there are good critisims of the sedimentary record, and the mechanism
-      seems a little bit uncertain. This is one of the cool things about this paper, but also makes this paragarph
-    a bit delicate.}
-    This
-    result is compatible with the field-based studies, which have evidenced
-    increased glacier erosion and pro-glacial sedimentation during deglaciation
-    periods \citep[e.g.][]{Koppes.Montgomery.2009,Micheletti.Lane.2016,Lane.etal.2017}, including the ongoing glacier retreat
-    \citep[e.g.][]{Bendixen.etal.2017}. Yet this increase has often been attributed
-    to higher meltwater availability, promoting sediment transport \citep{Delaney.Adhikari.2020},
-    enhanced glacier sliding \citep[][This citation might a bit strange]{Herman.etal.2011}, and hydrofracturing
-    (plucking) \citep[][This citation is more about monthly to diurnal variability]{Ugelvig.etal.2018}.
+    \note{I took the freedom to shorten the Var story quite a bit.}
+    This result is corroborated by a recent denudation record from the
+    Mediterranean Alps, which includes an increase in glacier erosion in the
+    Var Valley, roughly at 25\,ka following an extended period of low
+    dunudation rates \citep{Mariotti.etal.2021}. While that increase in
+    erosion predates the increase observed in our study ($\sim$ 15\,ka;
+    Fig.~\ref{fig:evolution}), it demonstrates that climatic conditions
+    over-which glacier erosion occurs are relatively specific and likely
+    transient. Other field-based studies have evidenced increased glacier
+    erosion and pro-glacial sedimentation during deglaciation periods
+    \citep[e.g.,][]{Koppes.Montgomery.2009, Micheletti.Lane.2016,
+    Lane.etal.2017}, including the ongoing glacier retreat
+    \citep[e.g.,][]{Bendixen.etal.2017}. Yet this increase has often been
+    attributed to higher meltwater availability, promoting sediment transport
+    \citep{Delaney.Adhikari.2020}, enhanced glacier sliding
+    \citep{Herman.etal.2011}, and hydrofracturing (plucking)
+    \citep{Hallet.1996, Ugelvig.etal.2018, Hildes.etal.2004}.
 
@@
     model, so that deglaciation surface meltwater is assumed to instantly exit
-    the glacier with no effect on glacier dynamics \ian{ \citep[e.g][]{Werder.etal.2013}
-      or glacier erosion  \citep[][]{Iverson.2012,Ugelvig.etal.2018}}.
-    The enhanced glacier
+    the glacier with no effect on glacier dynamics \citep[cf.][for comparison]
+    {Werder.etal.2013, Iverson.2012, Ugelvig.etal.2018}. The enhanced glacier
     velocity during periods of ice retreat is here modelled to result from
@@
     temperature and englacial water content, thereby facilitating shear strain
-    and eventually activating basal sliding \citep[cf.][]{Aschwanden.etal.2012,
-    Seguinot.etal.2016}. Observed feedbacks between surface meltwater and basal
-    sliding, not included in the model, presumably further accelerated erosion
-    during deglaciation \citep{Herman.etal.2011}. \ian{Furthermore, we omit the
-      fluvial transport of sediment, which could well increase sediment discharge
-      and erosion with increased melt, so long as additional sediment remains
-      available \citep{Delaney.Adhikari.2020}. Rapid glacier erosion can be expected
-    to accompany the  ongoing 21st-century deglaciation phase. This process might be
-    especially strong on the ice sheets of Greenland and Antarctica where
-    polythermal and cold-based glaciers are warming and sliding is increasing
-    \citep[e.g.][]{Moon.etal.2012,Overeem.etal.2017}.}
+    and eventually activating basal sliding \citep[cf.][for the physics]
+    {Aschwanden.etal.2012, Seguinot.etal.2016}. Observed feedbacks between
+    surface meltwater and basal sliding, not included in the model, presumably
+    further accelerated erosion during deglaciation \citep{Herman.etal.2011}.
+    Furthermore, we omit the fluvial transport of sediment, which could well
+    increase sediment discharge and erosion with increased melt, so long as
+    additional sediment remains available \citep{Delaney.Adhikari.2020}. Rapid
+    glacier erosion can be expected to accompany the ongoing 21st-century
+    deglaciation phase, especially for the ice sheets of Greenland and
+    Antarctica where polythermal and cold-based glaciers are warming and
+    sliding is increasing \citep[e.g.,][]{Moon.etal.2012,Overeem.etal.2017}.
 

### 2021-02-04 articles/erosion/alps-erosion.tex
Rework landscape age section with Ian's comments.

@@
     dunudation rates \citep{Mariotti.etal.2021}. While that increase in
-    erosion predates the increase observed in our study ($\sim$ 15\,ka;
+    erosion predates the increase observed in our study (ca.~15\,ka;
     Fig.~\ref{fig:evolution}), it demonstrates that climatic conditions
@@
     modelled to date from expansive glaciation stages, but instead to stages of
-    intermediate glacier cover \citep[Fig.~\ref{fig:hypsogram}; ][]{Barr.etal.2019}.
+    intermediate glacier cover
+    \citep[Fig.~\ref{fig:hypsogram};][]{Barr.etal.2019}.
 
-    For instance, low-elevation piedmont over-deepened troughs such as Lake
+    For instance, low-elevation piedmont over-deepened basins such as Lake
     Constance (Fig.~\ref{fig:landscape}), Lake Geneva and Lake Maggiore, are
     here modelled to be the product of extensive, Last Glacial Maximum-like
-    glaciations. \ian{Other modeling work suggests that features such as these
+    glaciations. Other modeling work suggests that features such as these
     likely form when water can access the glacier bed and encourage glacier
-      sliding and, thus, erosion \citep{Herman.etal.2011}. The model herein does not
-      include subglacial hydrology, the role of climate warming to concentrating
-      glacier erosion to the lower elevations and alpine remains evident.}
-
+    sliding and, thus, erosion \citep{Herman.etal.2011}. The model herein does
+    not include subglacial hydrology, yet the role of climate warming to
+    increasing glacier erosion to the lower elevations remains evident
+    (Fig.~\ref{fig:hypsogram}).
     Alpine terrain with elevations above 1000\,m are modelled to
@@
     (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
-    intermediate-elevation glacial valley of the Alps such as Lauterbrunnental
-    (Fig.~\ref{fig:hypsogram}), Haslital (Upper Aare) or Bout du Monde (Upper
-    Giffre). While the validity of the model results at high elevation are
+    intermediate-elevation glacial valley of the Alps as Lauterbrunnental
+    (Fig.~\ref{fig:landscape}), Haslital (Upper Aare) or Bout du Monde (Upper
+    Giffre) would date from periods of intermediate glacier extension.
+    While the validity of the model results at high elevation are
     limited by the 1-km resolution and the shallow-ice physics
@@
     of the modelled glacial cycle implies that the numerous Alpine mountain
-    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would date from
-    temperate periods with limited ice-cover such as the present day \ian{
-      or over interglacial periods, when glaciers are confined to their cirques
-      \citep{Barr.etal.2017,Barr.etal.2019}}. Similar processes also occur on
+    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would have formed
+    over interglacial periods, when glaciers are confined to their cirques
+    \citep{Barr.etal.2017, Barr.etal.2019}. Similar processes also occur on
     Himalayan glaciers and may limit erosion for the high elevation portion of
@@
     glacier erosion throughout their elevational range, from the piedmont
-    during glacial maxima, to the highest valleys during interglacial periods.
-    %The Alpine trim-line, in this case, would neither correspond to the upper
-    %reach of the Last Glacial Maximum glaciers \aref, nor to an englacial
-    %temperate-ice boundary \aref, but to an upper limit of erosion from
-    %retreating glaciers on steep terrain.
+    during glacial maxima, to thei highest cirques during interglacial periods.
+    The Alpine trim-line, in this case, would neither correspond to the upper
+    reach of the Last Glacial Maximum glaciers \citep[e.g.]{Kelly.etal.2004},
+    nor to an englacial temperate-ice boundary \citep{Coutterand.2010,
+    Seguinot.etal.2018}, but to a time-transgresive upper limit of erosion from
+    retreating glaciers on steep terrain.
 

### 2021-02-04 articles/erosion/alps-erosion.tex
Rework conclusions with ian's comments.

@@
 \Author[2]{Ian}{Delaney}
-\affil[1]{Anafi, Greece}
+\affil[1]{Independent scholar, Anafi, Greece}
 \affil[2]{Institute of Earth Surface Dynamics, University of Lausanne, Switzerland}
@@
         localization of rapid glacier erosion in regions of fast flow.
+      \item This high spatialtemporal variability hints at a complex
+        relationship between climate and glacier erosion, so that a highly
+        variable response of glacier erosion to climate should be expected.
       \item Temperature-dependent creep and basal sliding provide a mechanism
@@
         ranging from pan-Alpine ice sheets to modern-style mountain glaciers.
-      \item \ian{The high spatialtemporal variability  in erosion suggests that
-          the relationship between climate and glacier erosion remains greatly complex.
-          As a result, a highly variable response of glacier erosion to climate forcings should be expected.
-        \textbf{this last point could be somewhat coupled to the first point.}}
     \end{itemize}
@@
 \authorcontribution{%
-    J.~Seguinot performed the computations and initially drafted the
-    manuscript. I.~Delaney contributed to interpret the results in broader
-    context.}
+    J.~Seguinot performed the computations and prepared the figures.
+    I.~Delaney interpreted the results in a broader context. Both authors
+    contributed to the text.}
 

### 2021-02-04 articles/erosion/alps-erosion.tex
Add discussion of sediments preventing erosion.

@@
 
-    \ian{It might be worth mentioning/discussing the role of ``till dilatation''
-      in the sliding relationship. There are a lot of interpretations, but
-      the erosion rule is thought to be a proxy for glacier abrasion of a glacier flowing
-      over a hard bed. Thus if there is a layer of till present it is arguable
-      if such abrasion can occur. Given the uncertainties, I doubt this factor
-      changes the story very much. However, it might be worth mentioning
-      something (somewhere) about how multiple processes are lumped together in
-      the sliding relationship or that till processes are uncertain due to the
-      unknown distribution of till. This could ease some reviewer concerns.  }
-
     Climate forcing is provided by a~monthly climatology from interpolated
@@
     That said, the magnitude of the modelled erosion rates can hardly be used
-    as a criteria to rule out this or that erosion law for the glacial Alps, as
-    all these laws are affected with very large uncertainties. Thus it can not
+    as a criteria to rule out this or that erosion law for the glacial Alps.
+    Firstly, while sliding is here modelled by a pseudo-plastic law based on
+    till dilatation under high water pressure \citep{Tulaczyk.etal.2000},
+    empirical erosion rules are thought to be a proxy for glacier abrasion
+    \note{and plucking?} of a glacier flowing over a hard bed. Secondly, it is
+    arguable that a layer of till and proglacial sediments may have hindered
+    erosion for at least parts of the glacial cycle. Finally, all empirical
+    erosion laws are affected with very large uncertainties. Thus it can not
     be excluded that a linear or sub-linear erosion-law
@@
     (Fig.~\ref{fig:powerlaws}d).
-
-    \todo{Insert discussion of sediments here.}
 

### 2021-02-05 articles/erosion/alps-erosion.tex
Move figures into the text. Complete draft 1.0!
More comments from Ian, remove notes agreed on.

@@
 \newcommand{\todo}[1]{\textcolor{Rd}{\emph{[\textbf{TODO:} #1]}}}
-\newcommand{\aref}[0]{\textcolor{Rd}{\textbf{[REF.]}}}
-\newcommand{\ian}[1]{\textcolor{Or}{\emph{[\textbf{IAN:} #1]}}}
 
@@
 % ----------------------------------------------------------------------
+
+    \begin{figure*}[ht]
+      \centerline{\includegraphics{alpero_landscape}}
+      \caption{%
+        Alpine glacial erosion landscape diversity.
+        \textbf{(a)} Piedmont overdeepening of Lake Constance, ca.~10x50\,km.
+        \textbf{(b)} Glacial trough of Lauterbrunnental, ca.~1x10\,km.
+        \textbf{(c)} Mountain cirque of Ch\"uebodengletscher, ca.~1x1\,km.}
+      \label{fig:landscape}
+    \end{figure*}
 
@@
 
+    \begin{figure*}[ht]
+      \centerline{\includegraphics{alpero_cumulative}}
+      \caption{%
+        \textbf{(a)} Modelled cumulative (time-integrated) glacial erosion
+          potential over the last glacial cycle.
+        \textbf{(b)} Modelled total ice volume in centimetres of sea-level
+          equivalent (cm~s.l.e., black), annual (domain-integrated) erosion
+          volume (light brown) and 100-a running mean (dark brown). Shaded gray
+          areas indicate the timing for MIS~2 and~4
+          \citep{Lisiecki.Raymo.2005}.}
+        \label{fig:cumulative}
+    \end{figure*}
+
     The modelled cumulative (time-integrated) glacial erosion potential
@@
 
-    \note{I have replaced ``volumetric erosion rate'' with ``annual erosion
-          volume'' but kept the unit as ``m$^3$\,a$^{-1}$''. We agreed to stick
-          with erosion volume rather than average rates for now. Should the
-          erosion rate of 1.24\,mm\,a$^{-1}$ \citep{Stutenbecker.etal.2017} be
-          discussed?}
+    \begin{figure}[ht]
+      \centerline{\includegraphics{alpero_evolution}}
+      \caption{%
+        Modelled annual (domain-integrated) erosion volume (light brown) and 100-a
+        running mean (dark brown) in relation to modelled total ice volume in
+        centimetres of sea-level equivalent.}
+      \label{fig:evolution}
+    \end{figure}
 
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{figure*}[ht]
+      \centerline{\includegraphics{alpero_transects}}
+      \caption{%
+        \textbf{(a, b, c)} Modelled instantaneous erosion rate of the Rhine
+          Glacier for selected post-Last Glacial Maximum ages.
+        \textbf{(d)} Interpolated instantaneous erosion rate along a Rhine
+          Glacier transect for the entire last glacial cycle (upper panels
+          dashed line).}
+      \label{fig:transects}
+    \end{figure*}
+
+    \begin{figure*}[ht]
+      \centerline{\includegraphics{alpero_hypsogram}}
+      \caption{%
+        \textbf{(a)} Modelled erosion rate ``hypsogram'', showing the geometric
+          mean of (non-zero) modelled erosion rates in 10-m elevation bands
+          across the entire model domain and its time evolution.
+        \textbf{(b)} Same as Fig.~\ref{fig:cumulative}b.}
+      \label{fig:hypsogram}
+    \end{figure*}
 
@@
 
+    \begin{figure*}[ht]
+      \centerline{\includegraphics{alpero_sensitivity}}
+      \caption{%
+        Modelled cumulative glacial erosion potential over the last glacial
+        cycle without \textbf{(a, b, c)} and with \textbf{(d, e, f)}
+        paleo-precipitation corrections, and using three different
+        palaeo-temperature histories \citep[see][]{Seguinot.etal.2018}.}
+      \label{fig:sensitivity}
+    \end{figure*}
+
     Different paleoclimate forcing scenarios lead to different modelled
@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+
+    \begin{figure*}[ht]
+      \centerline{\includegraphics{alpero_powerlaws}}
+      \caption{%
+        Modelled cumulative (time-integrated) glacial erosion potential for
+        three different erosion laws published by
+        \textbf{(a)} \citet{Koppes.etal.2015},
+          ${5.2 \times 10^{-8} u_\mathrm{b}^{2.34}}$ (same as
+          Fig.~\ref{fig:cumulative}a but with a different colour scale),
+        \textbf{(b)} \citet{Herman.etal.2015},
+          ${2.7 \times 10^{-7} u_\mathrm{b}^{2.02}}$,
+        \textbf{(c)} \citet{Humphrey.Raymond.1994},
+          ${1 \times 10^{-4} u_\mathrm{b}^{1}}$, and
+        \textbf{(d)} \citet{Cook.etal.2020},
+          ${1.665 \times 10^{-1} u_\mathrm{b}^{0.6459}}$.
+        \textbf{(e-h)} Corresponding modelled cumulative glacial erosion
+          potential along a Rhine Glacier transect (upper panels dashed line), and
+        \textbf{(i-l)} corresponding modelled evolution of annual
+        (domain-integrated) erosion volume (light brown) and 100-a
+        running mean (dark brown) in relation to modelled total ice volume in
+        centimetres of sea-level equivalent (as in Fig.~\ref{fig:evolution}a).}
+      \label{fig:powerlaws}
+    \end{figure*}
 
@@
     empirical erosion rules are thought to be a proxy for glacier abrasion
-    \note{and plucking?} of a glacier flowing over a hard bed. Secondly, it is
+    \citep[cf.][]{Hallet.1979} (and possibly plucking) of a glacier flowing
+    over a hard bed. Secondly, it is
     arguable that a layer of till and proglacial sediments may have hindered
@@
 
-    \note{What about this new section title?}
-
     The modelled, domain-integrated glacier erosion volume do not increase
@@
     expands onto cooled bedrock hindering ice deformation and basal sliding.
-    \note{Ian: ``Furthermore, in cases of advance, ice surfaces
-          gradients are lower and cold ice is advected downglacier, thus
-          limiting erosion''. Is this also valid for relatively low-sloping
-          piedmont glaciers?}
     On the other hand, deglaciation periods are characterised by a warmer
@@
 
-    \note{I took the freedom to shorten the Var story quite a bit.}
     This result is corroborated by a recent denudation record from the
@@
     glacier erosion can be expected to accompany the ongoing 21st-century
-    deglaciation phase, especially for the ice sheets of Greenland and
+    deglaciation, especially for the ice sheets of Greenland and
     Antarctica where polythermal and cold-based glaciers are warming and
-    sliding is increasing \citep[e.g.,][]{Moon.etal.2012,Overeem.etal.2017}.
+    sliding may increase \citep[e.g.,][]{Moon.etal.2012, Overeem.etal.2017,
+    Mouginot.etal.2019}.
 
@@
     Europe and fourteen countries (except for new underwear).
+    I.~Delaney thanks Fr\'ed\'eric Herman for valuable support and
+    conversations while at University of Lausanne.
     The (previously published) ice flow simulations were made with computer
@@
 
-% ----------------------------------------------------------------------
-% Figures
-\clearpage
-% ----------------------------------------------------------------------
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpero_landscape}}
-      \caption{%
-        Alpine glacial erosion landscape diversity.
-        \textbf{(a)} Piedmont overdeepening of Lake Constance, ca.~10x50\,km.
-        \textbf{(b)} Glacial trough of Lauterbrunnental, ca.~1x10\,km.
-        \textbf{(c)} Mountain cirque of Ch\"uebodengletscher, ca.~1x1\,km.}
-      \label{fig:landscape}
-    \end{figure*}
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpero_cumulative}}
-      \caption{%
-        \textbf{(a)} Modelled cumulative (time-integrated) glacial erosion
-          potential over the last glacial cycle.
-        \textbf{(b)} Modelled total ice volume in centimetres of sea-level
-          equivalent (cm~s.l.e., black), annual (domain-integrated) erosion
-          volume (light brown) and 100-a running mean (dark brown). Shaded gray
-          areas indicate the timing for MIS~2 and~4
-          \citep{Lisiecki.Raymo.2005}.} \label{fig:cumulative}
-    \end{figure*}
-
-    \begin{figure}
-      \centerline{\includegraphics{alpero_evolution}}
-      \caption{%
-        Modelled annual (domain-integrated) erosion volume (light brown) and 100-a
-        running mean (dark brown) in relation to modelled total ice volume in
-        centimetres of sea-level equivalent.}
-      \label{fig:evolution}
-    \end{figure}
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpero_transects}}
-      \caption{%
-        \textbf{(a, b, c)} Modelled instantaneous erosion rate of the Rhine
-          Glacier for selected post-Last Glacial Maximum ages.
-        \textbf{(d)} Interpolated instantaneous erosion rate along a Rhine
-          Glacier transect for the entire last glacial cycle (upper panels
-          dashed line).}
-      \label{fig:transects}
-    \end{figure*}
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpero_hypsogram}}
-      \caption{%
-        \textbf{(a)} Modelled erosion rate ``hypsogram'', showing the geometric
-          mean of (non-zero) modelled erosion rates in 10-m elevation bands
-          across the entire model domain and its time evolution.
-        \textbf{(b)} Same as Fig.~\ref{fig:cumulative}b.}
-      \label{fig:hypsogram}
-    \end{figure*}
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpero_sensitivity}}
-      \caption{%
-        Modelled cumulative glacial erosion potential over the last glacial
-        cycle without \textbf{(a, b, c)} and with \textbf{(d, e, f)}
-        paleo-precipitation corrections, and using three different
-        palaeo-temperature histories \citep[see][]{Seguinot.etal.2018}.}
-      \label{fig:sensitivity}
-    \end{figure*}
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpero_powerlaws}}
-      \caption{%
-        Modelled cumulative (time-integrated) glacial erosion potential for
-        three different erosion laws published by
-        \textbf{(a)} \citet{Koppes.etal.2015},
-          ${5.2 \times 10^{-8} u_\mathrm{b}^{2.34}}$ (same as
-          Fig.~\ref{fig:cumulative}a but with a different colour scale),
-        \textbf{(b)} \citet{Herman.etal.2015},
-          ${2.7 \times 10^{-7} u_\mathrm{b}^{2.02}}$,
-        \textbf{(c)} \citet{Humphrey.Raymond.1994},
-          ${1 \times 10^{-4} u_\mathrm{b}^{1}}$, and
-        \textbf{(d)} \citet{Cook.etal.2020},
-          ${1.665 \times 10^{-1} u_\mathrm{b}^{0.6459}}$.
-        \textbf{(e-h)} Corresponding modelled cumulative glacial erosion
-          potential along a Rhine Glacier transect (upper panels dashed line), and
-        \textbf{(i-l)} corresponding modelled evolution of annual
-        (domain-integrated) erosion volume (light brown) and 100-a
-        running mean (dark brown) in relation to modelled total ice volume in
-        centimetres of sea-level equivalent (as in Fig.~\ref{fig:evolution}a).}
-      \label{fig:powerlaws}
-    \end{figure*}
-
-% ----------------------------------------------------------------------
-% Tables
-%\clearpage
-% ----------------------------------------------------------------------
-
 % ======================================================================

### 2021-02-08 articles/erosion/alps-erosion.tex
Read through abstract to methods, minor fixes.

@@
     unconstrained glacier erosion processes, the resulting modelled patterns of
-    glacier erosion includes persistent features. The cumulative imprint of
+    glacier erosion include persistent features. The cumulative imprint of
     the last glacial cycle shows a very strong localization of glacier erosion
@@
     Our results depict the Alpine glacier erosion landscape as a
-    time-transgressive patchwork, with different erosional landscapes of the
-    same mountain range corresponding to different glaciation stages and time
+    time-transgressive patchwork, with different parts of the
+    range corresponding to different glaciation stages and time
     periods, thus offering insight into its diversity.
@@
     Nihon Arupusu}), and elsewhere.
-
+    %
     Some mountain ranges are predominantly characterised by cirque glaciation
@@
     large-scale overdeepenings (e.g. Patagonia). But other regions, including
-    the Alps, present a higher variety glacial erosional landforms
+    the Alps, present a higher variety of glacier erosional landforms
     (Fig.~\ref{fig:landscape}), whose implications on glacial history is yet to
@@
     mountain erosion rates coincided with global cooling and Pliestocene
-    glaciations from roughly 2.6\,Ma (million years before present)
-    \citep{Herman.Champagnac.2016}. Yet, other work suggests that global
+    glaciations from roughly 2.6 million years before present
+    \citep{Herman.Champagnac.2016}. Yet, oceanic isotopic proxies for global
     weathering rates remained constant over this time period
-    \citep{Willenbring.Von-Blanckenburg.2010}. \note{Evidence suggests, other
-    work suggests, maybe this can be reworked.} Examination of contemporary
+    \citep{Willenbring.Von-Blanckenburg.2010}. Examination of contemporary
     erosion rates across different climates suggest that increased temperatures
@@
     difficult due to time scale biases \citep{Ganti.etal.2016} and the
-    non-linear relationship  between temperature and erosion
+    non-linear relationship  between atmospheric temperature and erosion
     \citep[e.g.,][]{Anderson.etal.2012,Mariotti.etal.2021}.
@@
     Instead, empirical relationships between glacier sliding and erosion can
-    represent erosional quantities well \citep[e.g.,][]{Humphrey.Raymond.1994,
-    Herman.etal.2015, Koppes.etal.2015} and can recreate important glacial
+    represent erosional quantities well and can recreate important glacial
     landforms \citep[e.g.,][]{Harbor.etal.1988, Macgregor.etal.2000}.
+    For instance, \citet{Humphrey.Raymond.1994} correlated temporal variations
+    in ice velocity and suspended sediment load during a surge of Variegated
+    Glacier to establish a linear erosion law.
     \citet{Koppes.etal.2015} quantified sediment
@@
     patterns of modelled erosion rate and cumulative last glacial cycle erosion
-    potential are analysed. We examine the glaciologcial conditions that cause
+    potential are analysed. We examine the glaciological conditions that cause
     erosion and discuss the implications of these conditions on understanding
-    the relationship between climate and glacier erosion, Despite aggregated
+    the relationship between climate and glacier erosion. Despite aggregated
     uncertainties on paleoclimate, glacier flow and glacier erosion processes,
@@
     shallow-shelf model
-    \citep{PISM-authors.2017}. The model includes temperature and water-content
+    \citep{PISM-authors.2017}. Our setup includes temperature and water-content
     dependent creep, pseudo-plastic basal sliding accounting for till
@@
     ice and bedrock temperature at 120\,ka (thousand years before the present),
-    and ran to the present.
-    The ice-sheet model physical parameters are listed in the earlier
-    open-access publication \citep{Seguinot.etal.2018}, and the complete PISM
-    set-up is stored in long-term archived model output metadata for the
+    and ran to the present. The ice-sheet model physical parameters are listed
+    in the earlier open-access publication \citep{Seguinot.etal.2018}, and the
+    complete list of PISM
+    parameters is stored in long-term archived model output metadata for the
     present study \citep{Seguinot.2021}, and for the earlier publication

### 2021-02-08 articles/erosion/alps-erosion.tex
Read through results to conclusions, minor fixes.

@@
     difficult to implement in many cases due to the large number of poorly
-    constrained parameters and process, i.e. water pressure flucuations and
+    constrained parameters and process, i.e. water pressure fluctuations and
     subglacial debris concentration \citep[e.g.,][]{Hallet.1979, Seguinot.2008,
@@
     \citep[WorldClim, Fig.~1h in][]{Seguinot.etal.2018} and the glacial relief
-    more pronounced in the input topography.
+    more pronounced in the topography.
 
@@
     the intra-montane Alps \citep[modelled to be largely
-    cold-based, Fig.~6c in][]{Seguinot.etal.2018}, are modelled to experience
+    cold-based, Fig.~6c in][]{Seguinot.etal.2018}, experience
     insignificant erosion (Fig.~\ref{fig:transects}a and~d). The modelled
@@
     despite variations in the glaciation maximum extent
-    (Fig.~\ref{fig:sensitivity} and \citealp[Fig.~3 in][]{Seguinot.etal.2018}).
+    (\citealp[Fig.~3 in][]{Seguinot.etal.2018}; Fig.~\ref{fig:sensitivity}).
     Furthermore, these changes are minor
@@
     empirical erosion rules are thought to be a proxy for glacier abrasion
-    \citep[cf.][]{Hallet.1979} (and possibly plucking) of a glacier flowing
+    (cf.~\citealp{Hallet.1979}; and possibly plucking) of a glacier flowing
     over a hard bed. Secondly, it is
@@
     velocities yield cumulative glacial-cycle erosion two to three orders of
-    magnitude higher in the Alpine valley than near the mountain tops
+    magnitude higher in the Alpine valleys than near the mountain tops
     (Fig.~\ref{fig:powerlaws}d).
@@
     the lower layers and activate the double-feedback loop between ice
-    temperature, deformation heat dissipation and basal melt-induced sliding.
+    temperature, deformational heat dissipation and basal melt-induced sliding.
     Warmer air temperatures contribute to enhance deformation and sliding even
@@
     Var Valley, roughly at 25\,ka following an extended period of low
-    dunudation rates \citep{Mariotti.etal.2021}. While that increase in
+    denudation rates \citep{Mariotti.etal.2021}. While that increase in
     erosion predates the increase observed in our study (ca.~15\,ka;
@@
     \citep{Delaney.Adhikari.2020}, enhanced glacier sliding
-    \citep{Herman.etal.2011}, and hydrofracturing (plucking)
+    \citep{Herman.etal.2011}, and hydrofracturing, i.e.~plucking
     \citep{Hallet.1996, Ugelvig.etal.2018, Hildes.etal.2004}.
@@
     (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
-    intermediate-elevation glacial valley of the Alps as Lauterbrunnental
+    intermediate-elevation glacial valleys of the Alps as Lauterbrunnental
     (Fig.~\ref{fig:landscape}), Haslital (Upper Aare) or Bout du Monde (Upper
@@
     Himalayan glaciers and may limit erosion for the high elevation portion of
-    the mountain range during cold periods as well, failing to offset uplift
+    that mountain range during cold periods as well, failing to offset uplift
     rates \citep{Harper.Humphrey.2003}.
@@
     glacier erosion throughout their elevational range, from the piedmont
-    during glacial maxima, to thei highest cirques during interglacial periods.
+    during glacial maxima, to the highest cirques during interglacial periods.
     The Alpine trim-line, in this case, would neither correspond to the upper
@@
     nor to an englacial temperate-ice boundary \citep{Coutterand.2010,
-    Seguinot.etal.2018}, but to a time-transgresive upper limit of erosion from
+    Seguinot.etal.2018}, but to a time-transgressive upper limit of erosion from
     retreating glaciers on steep terrain.
@@
     Due to stacked-up uncertainties regarding paleoclimate, glacier sliding and
-    erosion processes, our quantitative results are to be considered with extreme
-    uncertainties. Yet to the best of our current knowledge in these respective
-    areas of science, we can draw the following qualitative conclusions:
+    erosion processes, our quantitative results are very likely inaccurate.
+    Yet to the best of our current knowledge in these respective areas of
+    the geosciences, we can draw the following qualitative conclusions:
     \begin{itemize}
@@
         localization of rapid glacier erosion in regions of fast flow.
-      \item This high spatialtemporal variability hints at a complex
+      \item This high spatio-temporal variability hints at a complex
         relationship between climate and glacier erosion, so that a highly
@@
         slowdown of erosion counterbalances glacier expansion, so that
-        Alpine-wide glacier erosion rates do not correlate with the ice volume.
+        Alpine-wide glacier erosion volumes do not correlate with the ice volume.
       \item Rapid glacier erosion is restricted to low elevation during stages
@@
         higher elevations during periods of glacier retreat.
-      \item The diversity of the Alpine glacial erosion landscape appears to
+      \item The diversity of the Alpine glacier erosion landscape appears to
         be the time-transgressive signature of a variety of glacial stages

### 2021-02-08 articles/erosion/alps-erosion.tex
alright... looking good

@@
     low air and glacier temperatures, see a localization of rapid erosion rates
-    at low elevation, while the intra-montane glacial landscape is modelled to
-    date from phases of intermediate glacier cover. The modelled erosion rates
+    at low elevation, \ian{I am not sure I understand this phrase: \textit{while the intra-montane glacial landscape is modelled to
+    date from phases of intermediate glacier cover}}. The modelled erosion rates
     peak during deglaciation phases, when higher air and glacier temperature
@@
     range corresponding to different glaciation stages and time
-    periods, thus offering insight into its diversity.
+    periods, thus offering insight into its diversity \ian{Diversity of what? I think the
+    sentence should stop at the comma. }.
 
@@
     large-scale overdeepenings (e.g. Patagonia). But other regions, including
-    the Alps, present a higher variety of glacier erosional landforms
-    (Fig.~\ref{fig:landscape}), whose implications on glacial history is yet to
-    be understood.
+    the Alps, present a higher variety of glacier erosional landforms \ian{added
+      these old refs... but maybe you can read the one in French better than me 
+    \citep[][(Fig.~\ref{fig:landscape})]{Penck.1905,Agassiz.2012},}
+    whose implications on glacial history is yet to be understood.
 
     In absence of other landforms, glacially eroded topography has sometimes
-    been used as a proxy for glacier extension, for instance in
+    been used as a proxy for glacier extension \ian{extent}, for instance in
     geomorphological mapping \citep[e.g.,][]{Margold.etal.2011, Fu.etal.2012}.
@@
     Nevertheless, it remains unclear how sharp the erosional
-    transition to cold-based ice is, and why some glaciated regions show
+    transition to cold-based ice is, and why some glaciated regions show \ian{
+    characteristic glacial landforms}
     glacially preserved landscapes and others, including the Alps, do not
@@
 
-    On longer time scales, glacier erosion likely governs, at least in part,
+    On longer time scales \ian{I might remove ``on longer timescales''. it seems
+    that the processes likely occur concurrently}, glacier erosion \ian{also} likely governs, at least in part,
     the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through
@@
     concluded to a near-square relationship between basal sliding and glacier
-    erosion, but yielding much higher erosion rates.
+    erosion, but yielding much higher erosion rates \ian{I might be reading this
+    sentence wrong, but higher erosion rates compared to what?}.
     \citet{Cook.etal.2020} assembled a global compilation of erosion rates for
@@
     previously published model results \citep{Seguinot.etal.2018} and the
-    patterns of modelled erosion rate and cumulative last glacial cycle erosion
+    patterns of modelled erosion rate\ian{s} and cumulative last glacial cycle erosion
     potential are analysed. We examine the glaciological conditions that cause
@@
 
-    Our approach does not account for the feedback of glacier erosion onto
+    Our approach does not account for \ian{ feedbacks} of glacier erosion onto
     bedrock topography and, then onto, ice dynamics
-    \citep[e.g.,][]{Anderson.etal.2012}. Neither does the approach account for
-    downstream advection of debris \citep[e.g.,][]{Anderson.Anderson.2016} that
-    may have shed the bedrock from erosion, or fluvial transport of subglacial
-    sediment \citep{Delaney.etal.2019}. Instead, we compute the time-integrated
+    \citep[e.g.,][]{Anderson.etal.2012}. \ian{Furthermore, we assume that
+      once material is eroded, it is transported out of the system. In turn, we
+      do not account for the downstream advection of debris
+      \citep[e.g.,][]{Anderson.Anderson.2016}, or fluvial
+      transport of subglacial sediment \citep{Delaney.etal.2019}.}
+
+Instead, we compute the time-integrated
     glacier erosion rate,
@@
     is numerically approximated using a time-step of 10\,a in the
-    main simulation, and 100\,a in the alternative climate scenarios runs. It
-    can be noted from the above formula, that with an exponent, $l$,
+    main simulation, and 100\,a in the alternative climate scenarios runs.
+    \ian{Remove?: \textit{It
+    can be noted  from } We note that in} the above formula, that with an exponent, $l$,
     higher than one, the cumulative erosion potential has a greater dependency
@@
     and further upstream where the valley slopes increase. While mountain
-    cirques are poorly captured by the model horizontal resolution, high
+    cirques are poorly captured by the model\ian{'s} horizontal resolution, high
     cumulative erosion potential also occurs near the valley heads.
@@
           mean of (non-zero) modelled erosion rates in 10-m elevation bands
-          across the entire model domain and its time evolution.
+          across the entire model domain and its time evolution. \ian{ the black bit
+            is cumulative erosion through the time period, right? maybe worth a clarifying word. }
         \textbf{(b)} Same as Fig.~\ref{fig:cumulative}b.}
@@
     During stages of glacier advance and maximum extension, erosion is modelled
-    to be moderate and restricted the lower parts of the catchment, while much
+    to be moderate \ian{moderate? around 1 mm a$^{-1}$} and restricted the lower parts of the catchment, while much
     the intra-montane Alps \citep[modelled to be largely
@@
     arguable that a layer of till and proglacial sediments may have hindered
-    erosion for at least parts of the glacial cycle. Finally, all empirical
-    erosion laws are affected with very large uncertainties. Thus it can not
+    erosion \ian{armored the glacier bed from erosion?} for at least parts
+    of the glacial cycle. \ian{remove? \textbf{Finally, all empirical erosion laws are affected
+        with very large uncertainties.}}
+    \ian{Instead? \textit{Given the lack of data to constrain the erosion rule, }}it can not
     be excluded that a linear or sub-linear erosion-law
@@
     maximum glaciation correspond to comparatively slow erosion
-    (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}). While this result may
-    seem paradoxical, and is modulated by the exponent, $l$, in the erosion
+    (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).
+    \ian{This result may seem paradoxical, and is modulated in a limited extent
+      by the exponent, $l$, in the erosion rule.
+      However, it is in fact compatible with previous field-based studies that
+    have evidenced an increase of erosion rates with air temperatures for at
+    least some sets of glaciers \citep{Koppes.etal.2015, Cook.etal.2020}.}
+
+  While this may seem paradoxical, and is modulated by the exponent, $l$, in the erosion
     power-law, it is in fact compatible with previous field-based studies that
@@
 
-    Cold temperatures generally reduce sliding \citep{Koppes.etal.2015} and if
-    the glacier becomes cold-based then sliding, and thus erosion, will cease
-    altogether. In cases of glacier advance, newly formed, atmosphere-cold ice
+    Cold temperatures generally reduce sliding  and if  the glacier becomes
+    cold-based then sliding, and thus erosion, will cease altogether
+    \citep{Koppes.etal.2015}. In cases of glacier advance, newly formed, atmosphere-cold ice
     expands onto cooled bedrock hindering ice deformation and basal sliding.
@@
     glacial cycle (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).
+    \ian{Maybe your field, but are there some papers that can reference the
+      polythermal processes, or specifically reference that the model includes
+      these processes?}
 
@@
     transient. Other field-based studies have evidenced increased glacier
-    erosion and pro-glacial sedimentation during deglaciation periods
+    erosion and pro-glacial sedimentation during the \ian{current} deglaciation period
     \citep[e.g.,][]{Koppes.Montgomery.2009, Micheletti.Lane.2016,
-    Lane.etal.2017}, including the ongoing glacier retreat
-    \citep[e.g.,][]{Bendixen.etal.2017}. Yet this increase has often been
+    Lane.etal.2017, Bendixen.etal.2017},\ian{remove?: \textbf{including the ongoing glacier retreat
+    \citep[e.g.,][]{Bendixen.etal.2017}}}. Yet this increase has often been
     attributed to higher meltwater availability, promoting sediment transport
@@
     limited by the 1-km resolution and the shallow-ice physics
-    \citep{Imhof.etal.2019}, the lack of high-elevation rapid erosion for much
+    \citep{Imhof.etal.2019}, the lack \ian{of rapid erosion at } high-elevations  for much
     of the modelled glacial cycle implies that the numerous Alpine mountain
@@
 
-    Due to stacked-up uncertainties regarding paleoclimate, glacier sliding and
+    Due to stacked-up \ian{compounded?} uncertainties regarding paleoclimate, glacier sliding and
     erosion processes, our quantitative results are very likely inaccurate.
-    Yet to the best of our current knowledge in these respective areas of
-    the geosciences, we can draw the following qualitative conclusions:
+    \ian{remove? \textbf{    Yet to the best of our current knowledge in these respective areas of
+    the geosciences,} Yet, we can \ian{remove can?} draw the following qualitative conclusions:
     \begin{itemize}

### 2021-02-09 articles/erosion/alps-erosion.tex
Integrate comments from Ian, add old erosion refs.

--- 2021-02-08 articles/erosion/alps-erosion.tex
@@
     low air and glacier temperatures, see a localization of rapid erosion rates
-    at low elevation, while the intra-montane glacial landscape is modelled to
-    date from phases of intermediate glacier cover. The modelled erosion rates
+    at low elevation, while glacier erosion at higher elevation is modelled
+    date from phases of less extensive glaciation. The modelled erosion rates
     peak during deglaciation phases, when higher air and glacier temperature
@@
     range corresponding to different glaciation stages and time
-    periods, thus offering insight into its diversity.
+    periods.
 
@@
     the Alps, present a higher variety of glacier erosional landforms
-    (Fig.~\ref{fig:landscape}), whose implications on glacial history is yet to
-    be understood.
+    \citep[][Fig.~\ref{fig:landscape}]{Ramsay.1862, Ball.1863, Gastaldi.1873,
+    Penck.1905} whose implications on glacial history is yet to be understood.
 
     In absence of other landforms, glacially eroded topography has sometimes
-    been used as a proxy for glacier extension, for instance in
+    been used as a proxy for glacier extent, for instance in
     geomorphological mapping \citep[e.g.,][]{Margold.etal.2011, Fu.etal.2012}.
@@
     transition to cold-based ice is, and why some glaciated regions show
-    glacially preserved landscapes and others, including the Alps, do not
+    characteristic landform preservation and others, including the Alps, do not
     \citep{Wirsig.etal.2016, Seguinot.etal.2018}.
 
-    On longer time scales, glacier erosion likely governs, at least in part,
+    Glacier erosion also likely governs, at least in part,
     the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through
@@
     concluded to a near-square relationship between basal sliding and glacier
-    erosion, but yielding much higher erosion rates.
+    erosion.
     \citet{Cook.etal.2020} assembled a global compilation of erosion rates for
@@
     previously published model results \citep{Seguinot.etal.2018} and the
-    patterns of modelled erosion rate and cumulative last glacial cycle erosion
+    patterns of modelled erosion rates and cumulative last glacial cycle erosion
     potential are analysed. We examine the glaciological conditions that cause
@@
 
-    Our approach does not account for the feedback of glacier erosion onto
+    Our approach does not account for feedbacks of glacier erosion onto
     bedrock topography and, then onto, ice dynamics
-    \citep[e.g.,][]{Anderson.etal.2012}. Neither does the approach account for
-    downstream advection of debris \citep[e.g.,][]{Anderson.Anderson.2016} that
-    may have shed the bedrock from erosion, or fluvial transport of subglacial
-    sediment \citep{Delaney.etal.2019}. Instead, we compute the time-integrated
-    glacier erosion rate,
+    \citep[e.g.,][]{Anderson.etal.2012}. Furthermore, we assume that once
+    material is eroded, it is transported out of the system. In turn, we do not
+    account for the downstream advection of debris
+    \citep[e.g.,][]{Anderson.Anderson.2016}, or fluvial transport of subglacial
+    sediment \citep{Delaney.etal.2019}. Instead, we compute the
+    time-integrated glacier erosion rate,
     %
@@
     is numerically approximated using a time-step of 10\,a in the
-    main simulation, and 100\,a in the alternative climate scenarios runs. It
-    can be noted from the above formula, that with an exponent, $l$,
+    main simulation, and 100\,a in the alternative climate scenarios runs.
+    Note that in the above formula with an exponent, $l$,
     higher than one, the cumulative erosion potential has a greater dependency
@@
     and further upstream where the valley slopes increase. While mountain
-    cirques are poorly captured by the model horizontal resolution, high
+    cirques are poorly captured by the model's horizontal resolution, high
     cumulative erosion potential also occurs near the valley heads.
@@
           mean of (non-zero) modelled erosion rates in 10-m elevation bands
-          across the entire model domain and its time evolution.
+          across the entire model domain and its time evolution. The histogram
+          to the side shows the distribution of topography in the model domain.
         \textbf{(b)} Same as Fig.~\ref{fig:cumulative}b.}
@@
     During stages of glacier advance and maximum extension, erosion is modelled
-    to be moderate and restricted the lower parts of the catchment, while much
-    the intra-montane Alps \citep[modelled to be largely
+    to be under 1\,mm\,a$^{-1}$ and restricted the lower parts of the
+    catchment, while much the intra-montane Alps \citep[modelled to be largely
     cold-based, Fig.~6c in][]{Seguinot.etal.2018}, experience
@@
     over a hard bed. Secondly, it is
-    arguable that a layer of till and proglacial sediments may have hindered
-    erosion for at least parts of the glacial cycle. Finally, all empirical
-    erosion laws are affected with very large uncertainties. Thus it can not
+    arguable that a layer of till and proglacial sediments may have armored the
+    glacier bed from erosion for at least parts of the glacial cycle. Given the
+    lack of data to constrain the erosion rule, it can not
     be excluded that a linear or sub-linear erosion-law
@@
     maximum glaciation correspond to comparatively slow erosion
-    (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}). While this result may
-    seem paradoxical, and is modulated by the exponent, $l$, in the erosion
-    power-law, it is in fact compatible with previous field-based studies that
+    (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).
+    This result may seem paradoxical, and is modulated in a limited extent by
+    the exponent, $l$, in the erosion rule (Fig.~\ref{fig:powerlaws}).
+    However, it is in fact compatible with previous field-based studies that
     have evidenced an increase of erosion rates with air temperatures for at
@@
     the lower layers and activate the double-feedback loop between ice
-    temperature, deformational heat dissipation and basal melt-induced sliding.
+    temperature, deformational heat dissipation and basal melt-induced sliding
+    \citep[cf.][]{Aschwanden.etal.2012}.
     Warmer air temperatures contribute to enhance deformation and sliding even
@@
     transient. Other field-based studies have evidenced increased glacier
-    erosion and pro-glacial sedimentation during deglaciation periods
+    erosion and pro-glacial sedimentation during the current deglaciation period
     \citep[e.g.,][]{Koppes.Montgomery.2009, Micheletti.Lane.2016,
-    Lane.etal.2017}, including the ongoing glacier retreat
-    \citep[e.g.,][]{Bendixen.etal.2017}. Yet this increase has often been
+    Lane.etal.2017, Bendixen.etal.2017}. Yet this increase has often been
     attributed to higher meltwater availability, promoting sediment transport
@@
     limited by the 1-km resolution and the shallow-ice physics
-    \citep{Imhof.etal.2019}, the lack of high-elevation rapid erosion for much
+    \citep{Imhof.etal.2019}, the lack of rapid erosion at high-elevations for much
     of the modelled glacial cycle implies that the numerous Alpine mountain
@@
     The Alpine trim-line, in this case, would neither correspond to the upper
-    reach of the Last Glacial Maximum glaciers \citep[e.g.]{Kelly.etal.2004},
+    reach of the Last Glacial Maximum glaciers \citep[e.g.,][]{Kelly.etal.2004},
     nor to an englacial temperate-ice boundary \citep{Coutterand.2010,
@@
 
-    Due to stacked-up uncertainties regarding paleoclimate, glacier sliding and
+    Due to compounded uncertainties regarding paleoclimate, glacier sliding and
     erosion processes, our quantitative results are very likely inaccurate.
-    Yet to the best of our current knowledge in these respective areas of
-    the geosciences, we can draw the following qualitative conclusions:
+    Yet, we draw the following qualitative conclusions:
     \begin{itemize}

### 2021-02-09 articles/erosion/alps-erosion.tex
Add refs to Olivia's papers, Antarctic speed-up.

@@
     as a criteria to rule out this or that erosion law for the glacial Alps.
-    Firstly, while sliding is here modelled by a pseudo-plastic law based on
+    While sliding is here modelled by a pseudo-plastic law based on
     till dilatation under high water pressure \citep{Tulaczyk.etal.2000},
@@
     (cf.~\citealp{Hallet.1979}; and possibly plucking) of a glacier flowing
-    over a hard bed. Secondly, it is
-    arguable that a layer of till and proglacial sediments may have armored the
-    glacier bed from erosion for at least parts of the glacial cycle. Given the
-    lack of data to constrain the erosion rule, it can not
+    over a hard bed. Current erosion rates in the Alps strongly vary depending
+    on the substratum \citep{Steinemann.etal.2020, Steinemann.etal.2021}, and
+    it arguable that a layer of till and proglacial sediments may have armored
+    the glacier bed from erosion for at least parts of the glacial cycle.
+    Given the lack of data to constrain the erosion rule, it can not
     be excluded that a linear or sub-linear erosion-law
@@
     Antarctica where polythermal and cold-based glaciers are warming and
-    sliding may increase \citep[e.g.,][]{Moon.etal.2012, Overeem.etal.2017,
-    Mouginot.etal.2019}.
+    sliding may increase \citep[e.g.,][]{Moon.etal.2012, Mouginot.etal.2014,
+    Overeem.etal.2017}.
 

### 2021-02-11 data/postprocess_alpero_agg.py
Use regular profile coord in postprocessing.

@@ -112,7 +112,8 @@ def postprocess_extra(run_path):
         long_name='temperate-based ice cover area', units='m2')
 
     # compute erosion aggregated variables
-    x, y = cpf.read_shp_coords('../data/native/profile_rhine.shp')
+    x, y = cpf.read_shp_coords(
+        '../data/native/profile_rhine.shp', interval=1000)
     for law in ['coo2020', 'her2015', 'hum1994', 'kop2015']:
         pp[law+'_cumu'] = (dt*ex[law].sum(axis=0, min_count=1)).assign_attrs(
             long_name=ex[law].ref+' cumulative glacial erosion potential',

### 2021-02-11 figures/alpero_basaldrag.py
Add figure to show basal drag during retreat.

@@ -1,0 +1,74 @@
+#!/usr/bin/env python
+# Copyright (c) 2021, Julien Seguinot (juseg.github.io)
+# Creative Commons Attribution-ShareAlike 4.0 International License
+# (CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/)
+
+"""Plot Alps erosion change in glacier slope."""
+
+import cartowik.profiletools as cpf
+import cartowik.decorations as cde
+import absplots as apl
+import pismx.open
+
+
+def main():
+    """Main program called during execution."""
+
+    # initialize figure
+    fig, axes = apl.subplots_mm(
+        figsize=(85, 80), nrows=2, sharex=True, gridspec_kw=dict(
+            left=12, right=1.5, bottom=9, top=1.5, hspace=1.5))
+    for ax, label in zip(axes, 'ab'):
+        cde.add_subfig_label('(%s)' % label, ax=ax, loc='sw')
+
+    # read profile coordinates
+    x, y = cpf.read_shp_coords(
+        '../data/native/profile_rhine.shp', interval=1e3)
+
+    # each requested age
+    for age in [24, 20, 16]:
+        with pismx.open.subdataset(
+                ('~/pism/output/e9d2d1f/alpcyc4.1km.epica.1220.pp'
+                 '/ex.{:07.0f}.nc'), time=-age*1e3, shift=120000) as ds:
+
+            # interpolate along profile, use km
+            ds = ds.interp(x=x, y=y, method='linear')
+            ds = ds.assign_coords(d=ds.d/1e3)
+
+            # plot topographic profiles
+            ds['usurf'] = ds.topg + ds.thk.fillna(0)
+            ds.usurf.plot.line(ax=axes[0], x='d', label='{} ka'.format(age))
+            ds.topg.plot.line(ax=axes[0], x='d', color='k')
+
+            # plot normalized stresses
+            overburden = 910 * 9.81 * ds.thk.where(ds.thk > 1)
+            (ds.tauc/overburden*100).plot.line(
+                ax=axes[1], x='d', c='k', ls=':',
+                label='yield stress / overburden' if age == 24 else None)
+            ((ds.taub_x**2+ds.taub_y**2)**0.5/overburden*100).plot.line(
+                ax=axes[1], x='d',
+                label='basal drag / overburden' if age == 24 else None)
+
+            # to plot driving stress
[…]

### 2021-02-11 figures/alpero_temperature.py
Add Rhine temp fig. Don't loose this again.

@@ -1,0 +1,107 @@
+#!/usr/bin/env python
+# Copyright (c) 2021, Julien Seguinot (juseg.github.io)
+# Creative Commons Attribution-ShareAlike 4.0 International License
+# (CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/)
+
+"""Plot Alps erosion ice temperature profiles."""
+
+import matplotlib.pyplot as plt
+import cartowik.profiletools as cpf
+import absplots as apl
+import pismx.open
+
+
+def main():
+    """Main program called during execution."""
+
+    # initialize figure
+    fig, axes = apl.subplots_mm(
+        figsize=(177, 80), nrows=3, sharex=True, sharey=True, gridspec_kw=dict(
+            left=12, right=18, bottom=9, top=1.5, hspace=1.5))
+    cax = fig.add_axes_mm([177-18+1.5, 9, 3, 80-10.5])
+
+    # run and time for plot
+    run = '~/pism/output/e9d2d1f/alpcyc4.1km.epica.1220.pp'
+    ages = [24, 20, 16]
+
+    # read profile coordinates
+    x, y = cpf.read_shp_coords(
+        '../data/native/profile_rhine.shp', interval=1e3)
+
+    # loop on ages
+    for ax, age in zip(axes, ages):
+        ax.set_rasterization_zorder(2.5)
+
+        # open 3d output data
+        with pismx.open.dataset(
+                '{}/out.{:07.0f}.nc'.format(run, (120-age)*1e3)) as ds:
+
+            # interpolate along profile, use km, remove age
+            ds = ds[['topg', 'usurf', 'temp_pa']]
+            ds = ds.interp(x=x, y=y, method='linear')
+            ds = ds.assign_coords(d=ds.d/1e3).squeeze()
+
+            # plot topographic profiles
+            ds.topg.plot(ax=ax, x='d', color='k', lw=0.5)
+            ds.usurf.plot(ax=ax, x='d', color='k', lw=0.5)
+            #    ax.plot(dp, (bp+wp).T, 'k-', lw=0.5)
+
+            # mask temps above ice surface
+            ds['temp_pa'] = ds.temp_pa.where(ds.topg+ds.z < ds.usurf)
+
+            # plot temperature profile
+            cset = ax.contourf(
+                ds.d.expand_dims(z=ds.z), ds.z+ds.topg, ds.temp_pa, alpha=0.75,
+                levels=range(-24, 1, 3), cmap='RdBu_r', extend='both')
+
[…]

### 2021-02-11 articles/erosion/alps-erosion.tex
I managed to misspell your name again, sorry...
Add basal drag figure in manuscript. Update text.

@@
-% Copyright (c) 2019-2021, Julien Seguinot (juseg.github.io), Ian Delanay
+% Copyright (c) 2019-2021, Julien Seguinot (juseg.github.io), Ian Delaney
 % Creative Commons Attribution-ShareAlike 4.0 International License
@@
 \runningtitle{Last glacial cycle glacier erosion potential in the Alps}
-\runningauthor{J.~Seguinot and I.~Delanay}
+\runningauthor{J.~Seguinot and I.~Delaney}
 
@@
     glacial cycle, but show paradoxically little relation to the total glacier
-    volume. Phases of glacier advance and maximum extension, characterized by
-    low air and glacier temperatures, see a localization of rapid erosion rates
+    volume. Phases of glacier advance and maximum extension see a localization
+    of rapid erosion rates
     at low elevation, while glacier erosion at higher elevation is modelled
     date from phases of less extensive glaciation. The modelled erosion rates
-    peak during deglaciation phases, when higher air and glacier temperature
-    enhance glacier dynamics, implying that climatic conditions that result in
+    peak during deglaciation phases, when frontal retreat results in steeper
+    glacier surface slopes, implying that climatic conditions that result in
     rapid glacier erosion might be quite transient and specific.
@@
 
+    \begin{figure}[ht]
+      \centerline{\includegraphics{alpero_basaldrag}}
+      \caption{%
+        \textbf{(a)} Modelled basal and surface topographies interpolated along
+        Rhine Glacier (using the same transect and ages as in
+        Fig.~\ref{fig:basaldrag}).
+        \textbf{(b)} Interpolated yield stress in the pseudo-plastic sliding law
+        (dottel lines) and magnitude of the basal shear stress (basal drag,
+        solid lines), both normalized against the ice overburden pressure
+        \citep[cf.][for model physics]{Seguinot.2014, Seguinot.etal.2016}.}
+      \label{fig:basaldrag}
+    \end{figure}
+
     The modelled, domain-integrated glacier erosion volume do not increase
@@
     the exponent, $l$, in the erosion rule (Fig.~\ref{fig:powerlaws}).
-    However, it is in fact compatible with previous field-based studies that
-    have evidenced an increase of erosion rates with air temperatures for at
-    least some sets of glaciers \citep{Koppes.etal.2015, Cook.etal.2020}.
-
-    Cold temperatures generally reduce sliding  and if  the glacier becomes
-    cold-based then sliding, and thus erosion, will cease altogether
-    \citep{Koppes.etal.2015}. In cases of glacier advance, newly formed, atmosphere-cold ice
-    expands onto cooled bedrock hindering ice deformation and basal sliding.
-    On the other hand, deglaciation periods are characterised by a warmer
-    climate and above-balance glacier extent. Glacier ice formed an isolating
-    blanket from the atmosphere so that geothermal heat has had time to warm up
-    the lower layers and activate the double-feedback loop between ice
-    temperature, deformational heat dissipation and basal melt-induced sliding
-    \citep[cf.][]{Aschwanden.etal.2012}.
-    Warmer air temperatures contribute to enhance deformation and sliding even
-    more, yielding the highest domain-integrated erosion rates of the
-    glacial cycle (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).
+    During deglaciation periods, increased surface melt at low elevations
+    results in a steepened topographic profile (Figs.~\ref{fig:basaldrag}a).
+    The buttress formed by the glacier foot against gravitational forces is
+    reduced, causing an increase in the magnitude of the basal shear stress
+    (basal drag) relative to the ice overburden pressure, and an increase in
+    the area where sliding and thus, erosion, can occur
+    (Figs.~\ref{fig:basaldrag}b).
 
@@
     velocity during periods of ice retreat is here modelled to result from
-    changes of air temperature alone, propagating through diffusion and
-    advection into the ice, increasing the glacier's internal energy through
-    temperature and englacial water content, thereby facilitating shear strain
-    and eventually activating basal sliding \citep[cf.][for the physics]
-    {Aschwanden.etal.2012, Seguinot.etal.2016}. Observed feedbacks between
+    changes in glacier geometry alone (Figs.~\ref{fig:basaldrag}b).
+    Observed feedbacks between
     surface meltwater and basal sliding, not included in the model, presumably
@@
         variable response of glacier erosion to climate should be expected.
-      \item Temperature-dependent creep and basal sliding provide a mechanism
+      \item Increased gravitational drag due to surface profile steepening
+        provide a mechanism
         for accelerated erosion during deglaciation periods, irrespective of

### 2021-02-11 articles/erosion/alps-erosion.tex
Fix another embarassing mistake with running mean.

@@
           equivalent (cm~s.l.e., black), annual (domain-integrated) erosion
-          volume (light brown) and 100-a running mean (dark brown). Shaded gray
+          volume (light brown) and 1-ka running mean (dark brown). Shaded gray
           areas indicate the timing for MIS~2 and~4
@@
       \caption{%
-        Modelled annual (domain-integrated) erosion volume (light brown) and 100-a
+        Modelled annual (domain-integrated) erosion volume (light brown) and 1-ka
         running mean (dark brown) in relation to modelled total ice volume in
@@
         \textbf{(i-l)} corresponding modelled evolution of annual
-        (domain-integrated) erosion volume (light brown) and 100-a
+        (domain-integrated) erosion volume (light brown) and 1-ka
         running mean (dark brown) in relation to modelled total ice volume in
@@
         \textbf{(a)} Modelled basal and surface topographies interpolated along
-        Rhine Glacier (using the same transect and ages as in
+        the Rhine Glacier (using the same transect and ages as in
         Fig.~\ref{fig:basaldrag}).
         \textbf{(b)} Interpolated yield stress in the pseudo-plastic sliding law
-        (dottel lines) and magnitude of the basal shear stress (basal drag,
+        (dotted lines) and magnitude of the basal shear stress (basal drag,
         solid lines), both normalized against the ice overburden pressure

### 2021-02-11 figures/alpero_basaldrag.py
Use lighter black in basal drag figure.

@@ -38,12 +38,12 @@ def main():
             # plot topographic profiles
             ds['usurf'] = ds.topg + ds.thk.fillna(0)
             ds.usurf.plot.line(ax=axes[0], x='d', label='{} ka'.format(age))
-            ds.topg.plot.line(ax=axes[0], x='d', color='k')
+            ds.topg.plot.line(ax=axes[0], x='d', color='0.25')
 
             # plot normalized stresses
             overburden = 910 * 9.81 * ds.thk.where(ds.thk > 1)
             (ds.tauc/overburden*100).plot.line(
-                ax=axes[1], x='d', c='k', ls=':',
+                ax=axes[1], x='d', c='0.25', ls=':',
                 label='yield stress / overburden' if age == 24 else None)
             ((ds.taub_x**2+ds.taub_y**2)**0.5/overburden*100).plot.line(
                 ax=axes[1], x='d',

### 2021-02-11 figures/alpero_powerlaws.py
Fix ice volume units in power laws figure.

@@ -96,8 +96,8 @@ def main():
             pfax.set_yscale('log')
 
             # plot time series
-            tsax.plot(slvol, ds[ref+'_rate'], c='C11', alpha=0.5)
-            tsax.plot(slvol, ds[ref+'_rate'].rolling(
+            tsax.plot(slvol*100, ds[ref+'_rate'], c='C11', alpha=0.5)
+            tsax.plot(slvol*100, ds[ref+'_rate'].rolling(
                 age=100, center=True).mean(), c='C11')
 
             # set time series axes properties

### 2021-02-11 articles/erosion/alps-erosion.tex
Add corrections from Ian, two things to discuss.

@@
 
+    \todo{Ian: Given the importance of the sliding processes and dependence on
+          basal shear stress. Is it worth the thought that "experience with the
+          model shows that in the conditions applied here, till is mostly
+          saturated and the model behaves similar to Weertman"?}
+
     The ice-sheet model set-up was presented in an earlier publication
@@
     version~e9d2d1f), an open-source, finite-difference, shallow-ice and
-    shallow-shelf model
-    \citep{PISM-authors.2017}. Our setup includes temperature and water-content
-    dependent creep, pseudo-plastic basal sliding accounting for till
-    dilatation under high water pressure, bedrock deformation under the ice
+    shallow-shelf model \citep{PISM-authors.2017}. Our setup includes
+    temperature and water-content dependent creep, pseudo-plastic basal
+    sliding, bedrock deformation under the ice
     load, and a~positive degree-day (PDD) surface mass balance model. The model
@@
     Franz-Joseph Glacier \citep{Herman.etal.2015} and many of the glaciers
-    included in the global compilation by \citet{Cook.etal.2020}. Indeed,
-    atmospheric
-    temperatures have been found to be a strong driver of erosion in studies
-    involving several glaciers \citep{Koppes.etal.2015, Cook.etal.2020}.
+    included in the global compilation by \citet{Cook.etal.2020}.
 
@@
         solid lines), both normalized against the ice overburden pressure
-        \citep[cf.][for model physics]{Seguinot.2014, Seguinot.etal.2016}.}
+        \citep[cf.][for model physics]{Seguinot.2014, Seguinot.etal.2016}.
+        Basal drag values greater than yield stress result in glacier sliding.
+        \todo{Ian: make sure the legend matches the graph.}}
       \label{fig:basaldrag}

### 2021-02-11 movies/stills.py
Replace movie bumpers script by more generic one.

@@ -1,16 +1,22 @@
 #!/usr/bin/env python
-# coding: utf-8
+# Copyright (c) 2019-2021, Julien Seguinot (juseg.github.io)
+# Creative Commons Attribution-ShareAlike 4.0 International License
+# (CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/)
 
-import os
-import sys
-import yaml
+"""Assemble bumbers for Alpine ice sheet animation."""
+
+# NOTE: this is virtually the same script as for the Cordillera project.
+
+import os.path
+import argparse
 import subprocess
+import yaml
+import requests
 import matplotlib.pyplot as plt
 
 
 def bumper_init():
     """Init figure and axes for animation bumper."""
-    figw, figh = 192.0, 108.0
     fig = plt.figure(figsize=(192.0/25.4, 108.0/25.4))
     ax = fig.add_axes([0, 0, 1, 1])
     ax.set_facecolor('k')
@@ -42,18 +48,31 @@ def bumper_bysa(prefix, info):
     # initialize figure
     fig, ax = bumper_init()
 
-    # prepare cc icon bitmaps
+    # create icons directory if missing
+    if not os.path.isdir('icons'):
+        os.mkdir('icons')
+
+    # retrieve icons if missing
     for icon in ['cc', 'by', 'sa']:
-        pngpath = 'icons/{}.png'.format(icon)
-        svgpath = 'icons/{}.svg'.format(icon)
-        cmdline = 'inkscape {} -w 640 -h 640 --export-png={}'
+        url = 'https://mirrors.creativecommons.org/presskit/icons/'+icon+'.svg'
+        pngpath = 'icons/' + icon + '.png'
+        svgpath = 'icons/' + icon + '.svg'
+        if not os.path.isfile('icons/{}.svg'.format(icon)):
+            text = requests.get(url).text
+            text = text.replace('FFFFFF', '000000')
+            text = text.replace('path', 'path fill="#bfbfbf"')
+            with open(svgpath, 'w') as svgfile:
+                svgfile.write(text)
+
+        # prepare cc icon bitmaps
+        cmd = 'inkscape {} -w 640 -h 640 --export-filename={}'
         if not os.path.isfile(pngpath):
-            subprocess.call(cmdline.format(svgpath, pngpath).split(' '))
+            subprocess.call(cmd.format(svgpath, pngpath).split(' '))
 
     # add cc icons
[…]

### 2021-02-11 movies/render.sh
Add movie renderer to add stills and fade effects.

@@ -1,0 +1,44 @@
+#!/bin/bash
+# Copyright (c) 2021, Julien Seguinot (juseg.github.io)
+# Creative Commons Attribution-ShareAlike 4.0 International License
+# (CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/)
+
+# Add bumper frames and fading effects to any animation
+
+# prefix as command-line argument
+prefix="${1:-anim}"
+
+# prepare bumper frames
+python stills.py ${prefix}.yaml
+
+# assembling parametres
+imgs="??????.png"  # images to include, edit for quick tests
+fade=12  # number of frames for fade in and fade out effects
+hold=25  # number of frames to hold in the beginning and end
+secs=$((120+2*hold/25))  # duration of main scene in seconds
+
+# prepare filtergraph for main scene
+filt="[0]tpad=$hold:$hold:clone:clone"   # clone first and last frames
+
+# add title frame and bumpers
+filt+=",fade=in:0:$fade,fade=out:$((secs*25-fade)):$fade[main];"  # main scene
+filt+="[1]fade=in:0:$fade,fade=out:$((4*25-fade)):$fade[head];"  # title frame
+filt+="[2]fade=in:0:$fade,fade=out:$((3*25-fade)):$fade[refs];"  # references
+filt+="[3]fade=in:0:$fade,fade=out:$((3*25-fade)):$fade[disc];"  # disclaimer
+filt+="[4]fade=in:0:$fade,fade=out:$((3*25-fade)):$fade[bysa];"  # license
+filt+="[head][main][refs][disc][bysa]concat=5"
+
+# assemble video
+ffmpeg \
+    -i ${prefix}_main.mp4 \
+    -loop 1 -t 4 -i ${prefix}_head.png \
+    -loop 1 -t 3 -i ${prefix}_refs.png \
+    -loop 1 -t 3 -i ${prefix}_disc.png \
+    -loop 1 -t 3 -i ${prefix}_bysa.png \
+    -filter_complex $filt -pix_fmt yuv420p -c:v libx264 -r 25 -s 1920x1080 \
+    ${prefix}.mp4
+
+# accelerate for social media
+ffmpeg \
+    -i ${prefix}.mp4 -s 1920x1080 -vf "trim=5:125,setpts=PTS/10" \
+    ${prefix}_x10.mp4

### 2021-02-11 movies/alpero_volumes.py
Add Alps erosion spaghetti curves animation.

@@ -1,0 +1,154 @@
+#!/usr/bin/env python
+# Copyright (c) 2019-2021, Julien Seguinot (juseg.github.io)
+# Creative Commons Attribution-ShareAlike 4.0 International License
+# (CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/)
+
+"""Plot Alps erosion time evolution."""
+
+import numpy as np
+import matplotlib.animation as animation
+import absplots as apl
+import pismx.open
+
+
+def figure():
+    """Prepare initial animation figure."""
+
+    # initialize figure
+    fig = apl.figure_mm(figsize=(192, 108))
+    fig.subplots_mm(nrows=3, sharex=True, gridspec_kw=dict(
+        left=18, right=146, bottom=12, top=6, hspace=3))
+    fig.subplots_mm(gridspec_kw=dict(
+        left=61, right=6, bottom=12, top=6))
+    axes = fig.axes
+
+    # load postprocessed data (note: there seem to be currently no way to merge
+    # two dataset with slight errors using mfdataset)
+    with pismx.open.dataset(
+            '../data/processed/alpcyc.1km.epic.pp.ts.10a.nc') as ds:
+        with pismx.open.dataset(
+                '../data/processed/alpero.1km.epic.pp.agg.nc') as agg:
+            agg = agg.reindex_like(ds, method='nearest', tolerance=1e-9)
+            ds = ds.merge(agg, compat='override')
+
+        # plot lines on small panels
+        lines = []
+        thk = ds.volume_glacierized/ds.area_glacierized
+        for i, y in enumerate([
+                ds.kop2015_rate,
+                ds.kop2015_rate/ds.area_glacierized*1e3,
+                ds.hum1994_rate/ds.area_glacierized*1e4,
+                ]):
+
+            # variable-dependent properties
+            color = '0.25' if i == 2 else 'C11'
+            lines += axes[i].plot(
+                ds.area_glacierized/1e9, y, c=color, alpha=0.5)
+            lines += axes[i].plot(
+                ds.area_glacierized/1e9,
+                y.rolling(age=100, center=True).mean(), c=color)
+
+        # plot lines on main panel
+        lines += axes[3].plot(
[…]

### 2021-02-15 articles/erosion/alps-erosion.tex
Clear last notes, change acknowledgements.

@@
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    \todo{Ian: Given the importance of the sliding processes and dependence on
-          basal shear stress. Is it worth the thought that "experience with the
-          model shows that in the conditions applied here, till is mostly
-          saturated and the model behaves similar to Weertman"?}
 
@@
 
-    \begin{figure}[ht]
+    \begin{figure}
       \centerline{\includegraphics{alpero_basaldrag}}
@@
         \textbf{(a)} Modelled basal and surface topographies interpolated along
-        the Rhine Glacier (using the same transect and ages as in
+        the Rhine Glacier during glacier advance (36\,ka), maximum extent
+        (24\,ka) and retreat (16\,ka, using the same transect as in
         Fig.~\ref{fig:basaldrag}).
@@
         (dotted lines) and magnitude of the basal shear stress (basal drag,
-        solid lines), both normalized against the ice overburden pressure
-        \citep[cf.][for model physics]{Seguinot.2014, Seguinot.etal.2016}.
-        Basal drag values greater than yield stress result in glacier sliding.
-        \todo{Ian: make sure the legend matches the graph.}}
+        solid lines), both normalized against the ice overburden pressure.
+        Basal drag values greater than yield stress result in glacier sliding
+        \citep[cf.][for model physics]{Seguinot.2014, Seguinot.etal.2016}.}
       \label{fig:basaldrag}
@@
     During deglaciation periods, increased surface melt at low elevations
-    results in a steepened topographic profile (Figs.~\ref{fig:basaldrag}a).
+    results in a steepened topographic profile (Figs.~\ref{fig:basaldrag}a),
+    as observed and further expected on currently retreating glaciers
+    \citep{Huss.etal.2010, Zekollari.Huybrechts.2015}.
     The buttress formed by the glacier foot against gravitational forces is
@@
 
-\videosupplement{
-    A supplementary animation is available
-    (\url{https://vimeo.com/503162771}).}
-
 \authorcontribution{%
@@
     John Jansen and Martin Margold for insightful early discussions, Susan
-    Ivy-Ochs for being a constant (droid) motivator, as well as Popi, Karsta
-    and Stefan for welcoming him on Anafi and providing agreeable conditions
-    for work in times of pandemic.
-    J.~Seguinot exclusively used gear transported on a bike for 5000\,km across
-    Europe and fourteen countries (except for new underwear).
+    Ivy-Ochs for being a constant (droid) motivator, Popi, Karsta and Stefan
+    for welcoming him on Anafi and providing agreeable conditions for work in
+    times of pandemic, and the Long Haul Trucker for conveying his laptop and
+    data unbroken for 5000\,km across Europe and fourteen countries.
     I.~Delaney thanks Fr\'ed\'eric Herman for valuable support and

### 2021-02-15 articles/erosion/alps-erosion.tex
Improve figure placement in the text.
Fix data dois to newer versions, add video links.

@@
 
-    \begin{figure*}[ht]
+    \begin{figure*}
       \centerline{\includegraphics{alpero_landscape}}
@@
 
-    \begin{figure*}[ht]
+    \begin{figure*}
       \centerline{\includegraphics{alpero_cumulative}}
@@
 
-    \begin{figure}[ht]
+    \begin{figure}
       \centerline{\includegraphics{alpero_evolution}}
@@
 
-    \begin{figure*}[ht]
+    \begin{figure*}
       \centerline{\includegraphics{alpero_transects}}
@@
 
-    \begin{figure*}[ht]
+    \begin{figure*}
       \centerline{\includegraphics{alpero_hypsogram}}
@@
 
-    \begin{figure*}[ht]
+    \begin{figure*}
       \centerline{\includegraphics{alpero_sensitivity}}
@@
 
-    \begin{figure*}[ht]
+    \begin{figure*}[t]
       \centerline{\includegraphics{alpero_powerlaws}}
@@
     (\doi{10.5281/zenodo.4495419}). Other model output variables were
-    previously made available (\doi{10.5281/zenodo.1423160},
-    \doi{10.5281/zenodo.1423176}).}
+    previously made available (\doi{10.5281/zenodo.3604174},
+    \doi{10.5281/zenodo.3604142}).}
+
+\videosupplement{
+    Animations made available online show the modelled glacier erosion rates
+    (\url{https://vimeo.com/503162771}), and their evolving relationship with
+    ice volume (\url{https://vimeo.com/512478926}) and bedrock
+    altitude (\url{https://vimeo.com/512479008}).}
 

### 2021-02-15 figures/alpero_transects.py
Optimize dpi and figure size for submission.

@@ -91,6 +91,7 @@ def main():
                 ticks=levels[::3]))  # (mpl issue #11937)
 
     # set axes properties
+    tsax.set_rasterization_zorder(2.5)
     tsax.set_xlim(120, 0)
     tsax.set_xlabel('age (ka)')
     tsax.set_ylabel('distance along flow (km)')

### 2021-02-15 movies/render.sh
Update movie renderer to detect image series.

@@ -12,7 +12,6 @@ prefix="${1:-anim}"
 python stills.py ${prefix}.yaml
 
 # assembling parametres
-imgs="??????.png"  # images to include, edit for quick tests
 fade=12  # number of frames for fade in and fade out effects
 hold=25  # number of frames to hold in the beginning and end
 secs=$((120+2*hold/25))  # duration of main scene in seconds
@@ -28,9 +27,12 @@ filt+="[3]fade=in:0:$fade,fade=out:$((3*25-fade)):$fade[disc];"  # disclaimer
 filt+="[4]fade=in:0:$fade,fade=out:$((3*25-fade)):$fade[bysa];"  # license
 filt+="[head][main][refs][disc][bysa]concat=5"
 
+# look for input file(s)
+[ -f ${prefix}_main.mp4 ] && iargs="-i ${prefix}_main.mp4" ||
+    iargs="-f image2 -pattern_type glob -i $HOME/anim/$prefix/??????.png"
+
 # assemble video
-ffmpeg \
-    -i ${prefix}_main.mp4 \
+ffmpeg $iargs \
     -loop 1 -t 4 -i ${prefix}_head.png \
     -loop 1 -t 3 -i ${prefix}_refs.png \
     -loop 1 -t 3 -i ${prefix}_disc.png \

### 2021-02-23 articles/erosion/alps-erosion.tex
Add ref to Jansen et al. (2019), fix typo.

@@
     Penck.1905} whose implications on glacial history is yet to be understood.
+    The cosmogenic nuclide memory of bedrock erosion provides a more
+    quantitative, yet equally varied picture of glacier erosion effectiveness
+    both within and between glaciated regions \citep{Jansen.etal.2019,
+    Steinemann.etal.2020, Steinemann.etal.2021}.
 
@@
     the ``glacier buzz-saw''. In turn, evidence suggests that increased
-    mountain erosion rates coincided with global cooling and Pliestocene
+    mountain erosion rates coincided with global cooling and Pleistocene
     glaciations from roughly 2.6 million years before present
@@
     magnitude higher in the Alpine valleys than near the mountain tops
-    (Fig.~\ref{fig:powerlaws}d).
+    (Fig.~\ref{fig:powerlaws}d). Yet this result is compatible with the
+    order-of-magnitude variation of glacier erosion rates observed in the
+    cosmogenic nuclide record \citep{Jansen.etal.2019, Steinemann.etal.2020,
+    Steinemann.etal.2021}.
 

### 2021-05-16 articles/erosion/response-referee1.md
Add review from anonymous referee #1.

<!-- Copyright (c) 2021, Julien Seguinot (juseg.github.io), Ian Delaney
Creative Commons Attribution-ShareAlike 4.0 International License
(CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/) -->

# Authors' response to Anonymous Referee #1

Dear Anonymous Referee #1,

Thank you very much for your detailed review of our manuscript.

> The study of Seguinot and Delaney presents a time-integrated model on the
> glacial erosion potential over the last 120 ka in the Alps. Authors tie the
> calculations on a previous study by Seguinot et 2018 who modeled the glacial
> extend over Alps’ last glacial cycle by testing three different records on
> palaeo-temperature (GRIP, EPICA, and MD01-2444). Calculated basal velocities
> which base on the Parallel Ice Sheet Model PISM were used to test different
> glacial postulated erosional laws (Koppes et al., 2015; Herman et al., 2015,
> Humphrey and Raymond, 1994; Cook et al., 2020) focusing on the Rhine glacier
> area.

> The main achievement of the study is to combine a realistic, high-resolution
> model of glacial extension over a glacial cycle in a mountain range with
> assumptions on glacial erosion. Even though the utilized PISM code does not
> account for feedbacks of glacial erosion on bed topography, which makes it
> not directly comparable with common landscape codes on glacial erosion (e.g.
> Egholm et al. 2009; Sternai, et al., 2013), some relevant findings could be
> derived by this integrated high-resolution approach. This includes the
> observation of low erosion during glacier advance and maximum glacier
> extension, and the role of profile steeping during deglaciation and related
> increasing erosion rates.

## General comments

> Authors decided to keep the paper short, referring most information on
> glacial model set up and resulting ice cover to the Seguinot et 2018 paper.
> This makes it no always very easy to read and needs checking in the original
> manuscript (for example, ice cover deviations model – field bases LGM
> extension).

> The introduction gives a nice, relatively long (in relation to other
> chapters) overview on different aspects of glacial erosion and resulting
> features, and guides through problems to investigate them. Here, I don’t see
> very clear how different aspects or parts in the introduction are
> specifically addressed in this study, i.e. which problems are exactly aimed
> to be solved of this large portfolio of mentioned gaps in the understanding
> of glacial processes. I think this could be better formulated and balanced.

> I think the study could also well contribute to the discussion on the
> elevation distribution of the cumulative glacial erosion over several cycles
> in the Alps (high elevations vs low elevation, e.g. Valla et al., 2011). A
> plot showing the integrated glacial erosion potential over (e.g. 100 m)
> elevation bins would easy to do and eventually an enlightening supplement
> summing up Fig. 5a.

> I would also recommend to make some regional statements on the distribution
> of the glacial erosion potential (Fig. 2). It is quite obvious that some
> tectonic massifs can readily be discriminated, e.g. the Tauern Window, Oetzal
> Crystallin complex, the Aare Massif, Mt. Rosa - Gran Paradiso and Pelvoux
> Massif. Is it the steepness of these low erodible crystallin massifs making
> the erosional potential appearing strikingly high? I think this should/could
> be discussed..

## Specific comments

> Line 24 Maybe a bit odd to use 4 very old references from the Alps only (you
> hardly used more than 2 references throughout the MS and there are also high
> variety of glacial landforms in other mt ranges as you mentioned).

> Line 33 I think that’s not very easy to understand what landform you refer
> to? The “periglacial blockfields topped by glacial erratic boulders” (Wirsig
> et al., 2018)? Eventually be more precise on this would help readers.

> Line 101 Could be misleading as (glacial and periglacial) cirque erosion
> processes are not really covered by any glacial erosion law discussed here
> (or any other I am aware of, Sanders et al., 2012).

> Line 116  ..while much OF the intra-montane..

> Line 127 Higher precipitation increases ice flux and thus erosion, I guess?
> Would eventually helpful to mention (even though it might be referred in
> Seguinot et al., 2018).

> Line 140 Can you be more precise what you mean by realistic? You mean because
> of the localized pattern? Maybe I am wrong but shouldn’t the erosion
> potential at least at the lake Constance overdeepening (Fig. 7) not in the
> order of hundreds of meters (cf. Preusser et al., 2010) and the best fit
> rather (b) or (c) – at least from what I read from the transect (e-h)?
> Visually (a)-(d) seem not to correspond to (e)-(h) if (a)-(d) is also
> presented in meters (annotation at the bar is missing). For example, in (b)
> the maximum erosion potential is like >>1000 m (if in meters) while in
> corresponding (f) it seems clearly lower than 1000 m.

> Furthermore, isn’t the impression of the fit potentially very dependent on
> the initial model parametrization (Seguinot et al., 2018), i.e. the ice flux
> velocity? You should address these dependences!

> Line 163 Observing Alpine topography I find this result important, which can
> maybe also serve as explanation why e.g. (low erodible) areas away from the
> big troughs covered during glacial maxima only, do surprisingly often show
> no/very low degree in glacial modification (e.g. Ticino; Kelly et al., 2004).

> Line 187 “time-transgressive radial pattern”. I don’t understand what you
> mean..

> Line 195 Very much share this view!

> Line 197 Would recommend to be more precise. Guess you know that there are
> many, many cirques in the Alps as low as 1500 m (and even below) e.g. forming
> in areas outside the connected ice stream network in the SW and easternmost
> Alps. This can be revealed by a quick check at any higher resolution DEM or
> google earth.

> In the MS the word “yet” is very often used. Eventually consider reducing.
> The frequency is a bit irritating when reading.

## Figures

> Fig. 2 Please indicate the outline of the connected ice stream network during
> the LGM (like in Fig. 4 of Seguinot et al., 2018). Especially in the eastern
> and SW Alps coverage largely deviates from what has been suggested from field
> data compilation (e.g. Ehlers and Gibbard, 2004). Even though outcomes might
> not be changing much, it might be helpful to know how much the %overlap is -
> erosional potential might change as fluvial topography turns into glacial one
> (e.g. Harbor, 1988) and this is probably not what you want to mix, I guess.

> Fig. 5 I have to confess Fig. 5a surprises me, there are really glaciers as
> low as 500 m.a.s.l (and even below) from 110 – 40 ka in the Seguinot et al.,
> 2018 model?

## References not in the manuscript:

> Egholm, D.L., Nielsen, S.B., Pedersen, V.K., Lesemann, J.E., 2009. Glacial
> effects limiting mountain height: Nature, v. 460, p. 884–888.

> Ehlers, J., Gibbard, P.L., 2007. The extent and chronology of Cenozoic Global
> Glaciation. Quat. Int. 164–165, 6–20.

> Harbor, J.M., Hallet, B., and Raymond, C.F., 1988. A numerical model of
> landform development by glacial erosion: Nature, v. 333, p. 347–349.

> Kelly, M.A., Buoncristiani, J.F. and Schlüchter, C., 2004. A reconstruction
> of the last glacial maximum (LGM) ice-surface geometry in the western Swiss
> Alps and contiguous Alpine regions in Italy and France. Eclogae Geol. Helv.
> 97, 57–75.

> Mey, J., Scherler, D., Wickert, A.D., Egholm, D.L., Tesauro, M., Schildgen,
> T.F., and Strecker, M., R., 2016. Glacial isostatic uplift of the European
> Alps: Nature Communications, v. 7.

> Preusser, F., Reitner, J.M., and Schlüchter, C., 2010. Distribution,
> geometry, age and origin of overdeepened valleys and basins in the Alps and
> their foreland: Swiss Journal Geoscience, v. 103, p. 407–426.

> Sanders, J.W., Cuffey, K.M., Moore, J.R., MacGregor, K.R., and Kavanaugh,
> J.L., 2012. Periglacial weathering and headwall erosion in cirque glacier
> bergschrunds: Geology, v. 40, p. 779-782.

> Sternai, P., Herman, F., Valla, P.G., Champagnac, J.-D., 2013. Spatial and
> temporal variations of glacial erosion in the Rhône valley (Swiss Alps):
> Insights from numerical modeling. Earth and Planetary Science Letters, 368,
> p. 119-131.

> Valla, P.G., Shuster, D.L., van der Beek, P.A., 2011. Significant increase in
> relief of the European Alps during mid-Pleistocene glaciations. Nat. Geosci.
> 4, 688–692.

### 2021-05-16 articles/erosion/response-referee2.md
Add review from referee #2 Ian Evans.

<!-- Copyright (c) 2021, Julien Seguinot (juseg.github.io), Ian Delaney
Creative Commons Attribution-ShareAlike 4.0 International License
(CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/) -->

# Authors' response to Ian Evans

Dear Ian Evans,

Thank you very much for your detailed review of our manuscript.

## General Comments

> The authors provide large amounts of very useful results from simulations,
> concerning the effects of varying glacier geometry on erosion potential. It
> was quite an achievement to compute this at fairly high temporal and spatial
> resolution (1 km and 50 a throughout the Alps). The conclusions are
> reasonable, and I value the results. They provide support for the
> traditional view, that high-altitude features such as cirques are eroded by
> steep, local glaciers and not under extensive icefields well above ELA. Most
> erosion is expected along major valleys (glacial troughs): the potential
> erosion is greatest low down the glaciers, with a lesser maximum at high
> altitude and a minimum between. The paper is well written and the Figures
> are all relevant and interesting. The qualitative conclusions are well
> justified.

> An interesting result, not commented on, is that from Figs. 3 and 7 i & j,
> potential erosion volume in the Alps peaks when ice cover amounts to 1.6 cm
> sea level equivalent. I would love to see a map of the glaciers at that
> stage\*, which is well below maximum but considerably more than recent ice
> cover (0.3 ± 0.1 mm s.l.e., Farinotti et al. 2019). This peak applies to the
> ‘Koppes’ and ‘Herman’ erosion laws, but not to ‘Humphrey’ and ‘Cook’ laws
> which give very broad maxima for s.l.e. 10-30 cm (Fig. 7 k & l).

> As the comment (lines 221-222) on erosion distribution during advance,
> retreat and maxima is of great interest, it would be useful to illustrate
> this with three maps of potential erosion during each of these types of
> phase.

> I find no mention of lithology, rock type, geology, resistance or erodibility
> (excluding that citing Herman on line 56). This omission of half the erosion
> equation should be mentioned; it implies that the ‘potential’ nature of the
> erosion should be emphasised repeatedly. This adds to the admitted
> limitations of lack of feedback to changing topography, absence of hydrology
> and absence of subglacial sediment accumulation or movement. Using constant
> topography might be OK for the last glaciation, as here, but would become
> increasingly unrealistic for several glaciations. Destinations of eroded
> material are not considered: deposition in temporary storage may inhibit
> erosion in parts. The omission of hydrological processes must be a major
> drawback, as Herman et al. (2011), Egholm et al. (2012a, b) and Pedersen et
> al. (2014) showed that its inclusion of subglacial meltwater, plus horizontal
> stress gradients, gave more realistic simulations of erosion. For erosion
> rate as well as for cumulative erosion, it is potential erosion that is
> calculated, rather than predicted erosion.

> A further necessary geographical qualification concerns the misfit between
> simulated and actual glacier extent. The latter is well established from
> mapping of moraines. It is necessary to check Seguinot et al. (2018) to see
> that all the models predict too much ice in the southwest (south of 45° N)
> and far too much in the east (Sava, Drava, Mur, Enns and Traun valleys),
> while the EPICA models produce too little in the west (Rhône and Jura). I
> agree that this is probably because while the amount of precipitation is
> changed, the spatial pattern is not: thus the orographic effect of ice
> build-up increasing precipitation shadow, and the circulation changes due to
> southward shift of westerlies at LGM, are not included. To make this paper
> more self-contained, a brief mention of the 2018 evaluation is in order.
> Once a more realistic LGM precipitation pattern is accepted, future modelling
> might use gradations between that and the present pattern as a function of
> temperature.

> Overall, the paper reports on a very worthwhile computer-intensive exercise.
> It is densely packed with results and repays a careful read and a study of
> the videos. Clearly there can be interesting comparisons with Lai and Anders
> (2021) very recent submission to esurf !

>   [\*That stage can be approximated by pausing the first erosion
>    rates video around 99 ka and 13 ka. The three videos are strongly
>    recommended!]

> The text is short and pithy, so there is space for some extension. There are
> a few missed opportunities, and some incompleteness in the presentation. In
> detail I suggest the following improvements:

## Specific comments - Text:

> The 1 km resolution of the modelling is clearly important, so it is strange
> that the first mention is on p.11. Although it is very demanding of computer
> time, it is still (as admitted) not adequate to represent cirques (averaging
> of the order of 700 m across). Even stranger, there is no mention of the DEM
> used (it was SRTM in the 2018 paper). This information should be provided in
> section 2, p. 3-4.

> The lateral constraints on valley glaciers are usually covered by a ‘form
> factor’: how are they handled by this version of the PISM ice sheet model –
> how has lateral drag been included?

> It might be pointed out that the ‘Koppes’ relationship with an exponent of
> 2.34 is based on 13 data points that are very poorly distributed: in fact,
> Koppes et al. (2015) go on to drop two outliers and reach an exponent of
> 2.62.

> In lines 206-209, the point about trimlines, relating them to
> time-transgressive erosion, is well taken. But they represent the integrated
> effect of many phases of glacial erosion (as noted on 223-224), not
> necessarily just from recessional phases.

## - Figures:

> In Fig. 3 is almost impossible to follow the brown lines, unless advancing
> and retreating parts of lines are distinguished (e.g. by colour). (The ice
> volume video may help out.)  Also: keep to ‘potential erosion volume’. The
> pale brown bars show a decline in annual erosion volume with ice volumes over
> 1.8 cm s.l.e., stronger than is implied in line 110 of the text – it is more
> than ‘slight’.

> To aid interpretation of Figs 4 a-c and 7 a-d, a corresponding topographic
> map or hillshade of the Rhine area on the same scale should be provided.

> Fig. 1c illustrates a very unusual cirque, with a huge lake replacing a very
> unhealthy convex glacier. Fascinating, but surely a more representative
> cirque should be used. Actually these three photos are not well integrated
> with the text (pace lines 191 & 195-198). Lauterbrunnen is an extreme
> (vertical-sided) trough example, over-used by textbooks.

> Fig. 5: The elevation histogram in Fig. 5a could be misleading, as it seems
> to cover the whole study area, including parts never glacier-covered. It
> should be limited to areas glacier-covered at the maximum, and preferably
> supplemented by a histogram for those covered now or at the minimum. The
> accompanying main graph (5a) relates to areas glacier-covered at each time
> slice. Re another reviewer’s comment, yes LGM glaciers did reach below 500 m:
> at L. Garda (67 m?) and the Tagliamento Glacier lobe LGM ice reached below
> 100 m. An integrated plot of total-cycle potential erosion rate (and
> volume?) at each altitude, over glacier-covered areas, would be interesting
> (as an extra Figure): can rates at high altitude, where glacial duration is
> longer, catch up with those at low altitude?  Probably not, given the
> concentration of flow down major valleys. Striping above 3000 m, presumably
> related to low surface area, suggests a switch there to 20 m or 40 m bands
> would be advisable. Caption; ‘modelled potential erosion rates’?

> In Fig. 7, the quantitative contrasts between the ‘Koppes’ law (e) and the
> others are alarming. If the scales are taken literally, (e) gives potential
> erosion around 1 m, while (f) and (g) give hundreds of m, in 120 ka. As 120
> ka is only a fraction of the Quaternary, hundreds of m seems too much, while
> 1 m seems very low (0.008 mm a-1 overall). For me, there should be a correct
> answer between these extremes. The spatial pattern, however, is more
> important.

> Fig. 8a shows a flat dome at 3000 m for about the first 20 km of the Rhine.
> As there are mountains at 3630, 3583 and 3192 m around the sources, it is
> likely that   there would be steeper ice slopes up to these peaks, giving
> rather different stresses and erosion potentials. Also are the instabilities
> in Fig. 8b, near source and terminus, edge effects or related to the 1 km
> resolution?

>   The three videos are well worth watching  -in fact they clarify some
>   queries arising from the paper. Perhaps they could be captioned in the
>   ‘Video supplement’ paragraph. Some extra information on the third (bedrock
>   altitude) video could point out that there are numerous zero values not
>   shown: the zero geometric mean rates from 2300 to 3000 m around 24 ka
>   otherwise seem to conflict with the numerous positive rates plotted. Zero
>   rates presumably show where ice is frozen to the bed: their altitudinal
>   distribution might be worth discussion in the text. Above 3000 m means are
>   positive, which seems strange.

## Technical corrections:

> All map figures need km scale bars.

> Line 24  ‘are yet’

> 27 ‘in the absence’

> 38 ‘Examination … suggests’

> 61 ‘modelled erosion potential’ rather than ‘ modelled erosion rates’ ?

> 70 insert ‘isostatic’?

> 72 ‘and is run to…’

> 78 ‘temperature lapse-rate’ or is it ‘temperature change’ ?

> 116-7 ‘much of the’; and delete comma after bracket.

> 161 ‘does not increase’

> 190 delete both commas

## References not in original:

> Egholm, D.L., Pedersen, V.K., Knudsen, M.F., Larsen, N.K., 2012a. Coupling
> the flow of ice, water, and sediment in a glacial landscape evolution model.
> Geomorphology 141–142: 47–66.
> http://dx.doi.org/10.1016/j.geomorph.2011.12.019.

> Egholm, D.L., Pedersen, V.K., Knudsen, M.F., Larsen, N.K., 2012b. On the
> importance of higher-order ice dynamics for glacial landscape evolution.
> Geomorphology 141–142: 67-80.

> Farinotti, D., et al., 2019. A consensus estimate for the ice thickness
> distribution of all glaciers on Earth. Nat. Geosci. 12, 168–173.

> Lai, J., Anders, A.M., 2021. Climatic controls on mountain glacier basal
> thermal regimes dictate spatial patterns of glacial erosion.
> https://esurf.copernicus.org/preprints/esurf-2021-26/

> Pedersen, V.K., Huismans, R.S., Herman, F., Egholm, D.L., 2014. Controls of
> initial topography on temporal and spatial patterns of glacial erosion.
> Geomorphology 223, 96-116.

## Acknowledgements:

> I am grateful for helpful discussions with Iestyn Barr, Jeremy Ely, Matt
> Tomkins, Pippa Whitehouse, Cristina Balaban and Matt Wiecek.

> - Ian S. Evans, Durham University, U.K.

### 2021-05-17 articles/erosion/response-referee1.md
Address reviewer #1 specific comments, figs todo.

@@
 
+We agree the references are not directly relevant. The citation was reduced to
+the latest reference by Penck (1905) and moved within the sentence to clarify
+that it is specific to the Alps.
+
 > Line 33 I think that’s not very easy to understand what landform you refer
@@
 > et al., 2018)? Eventually be more precise on this would help readers.
+
+Agreed. "Characteristic landform preservation" was replaced with "preserved
+periglacial blockfields".
 
@@
 
+Thank you for pointing out this study. The lack of cirque erosion processes was
+acknowledged and the reference added in this sentence, as well as in the
+relevant part of the discussion on the age of the glacial landscape.
+
 > Line 116  ..while much OF the intra-montane..
+
+Corrected.
 
@@
 > Seguinot et al., 2018).
+
+Indeed. The increase in ice discharge was made explicit.
 
@@
 
+TODO: check the numbers here. Make sure that the maps (a-d) fit the profiles
+(e-h), and then think again how to address the comment.
+
 > Furthermore, isn’t the impression of the fit potentially very dependent on
@@
 > velocity? You should address these dependences!
+
+TODO: discuss uncertainty with sliding.
 
@@
 
+The following sentence was addedin this paragraph. "Nevertheless, it may
+explain why some areas covered during glacial maxima only, such as some valleys
+on the southern side of the Alps, appear to have experienced only little
+glacial modification of their topography."
+
 > Line 187 “time-transgressive radial pattern”. I don’t understand what you
@@
 
+This sentence was only meant to recall the aforementioned results. We rephrased
+to "the erosion patterns shift spatially through the glacial cycle".
+
 > Line 195 Very much share this view!
+
+Good to know!
 
@@
 
+The statement was reworked and restricted to "the higest mountain cirques".
+
 > In the MS the word “yet” is very often used. Eventually consider reducing.
 > The frequency is a bit irritating when reading.
+
+In several instances we replaced "yet" with "but" and "however".
 
@@
 
+We added the LGM outline.
+
+TODO: I don't understand the second part of the comment.
+
 > Fig. 5 I have to confess Fig. 5a surprises me, there are really glaciers as
@@
 > 2018 model?
+
+Yes. This primarily concerns the southern slope of the Alps where topography
+drops abruptedly from the highest peaks to deep valleys. It concerns only a few
+grid cells for a given time step and is certainly not a realistic model output.
+
+TODO: test cutting the hypsogram below a minimum number of grid cells.
 

### 2021-05-17 articles/erosion/alps-erosion.tex
Address reviewer #1 specific comments, figs todo.

--- 2021-02-15 articles/erosion/alps-erosion.tex
@@
     large-scale overdeepenings (e.g. Patagonia). But other regions, including
-    the Alps, present a higher variety of glacier erosional landforms
-    \citep[][Fig.~\ref{fig:landscape}]{Ramsay.1862, Ball.1863, Gastaldi.1873,
-    Penck.1905} whose implications on glacial history is yet to be understood.
+    the Alps \citep[e.g.][Fig.~\ref{fig:landscape}]{Penck.1905}, present a
+    higher variety of glacier erosional landforms, whose implications on
+    glacial history is yet to be understood.
+    The cosmogenic nuclide memory of bedrock erosion provides a more
+    quantitative, but equally varied picture of glacier erosion effectiveness
+    both within and between glaciated regions \citep{Jansen.etal.2019,
+    Steinemann.etal.2020, Steinemann.etal.2021}.
 
@@
     transition to cold-based ice is, and why some glaciated regions show
-    characteristic landform preservation and others, including the Alps, do not
+    preserved periglacial blockfields and others, including the Alps, do not
     \citep{Wirsig.etal.2016, Seguinot.etal.2018}.
@@
     the ``glacier buzz-saw''. In turn, evidence suggests that increased
-    mountain erosion rates coincided with global cooling and Pliestocene
+    mountain erosion rates coincided with global cooling and Pleistocene
     glaciations from roughly 2.6 million years before present
-    \citep{Herman.Champagnac.2016}. Yet, oceanic isotopic proxies for global
-    weathering rates remained constant over this time period
+    \citep{Herman.Champagnac.2016}. However, oceanic isotopic proxies for
+    global weathering rates remained constant over this time period
     \citep{Willenbring.Von-Blanckenburg.2010}. Examination of contemporary
@@
           areas indicate the timing for MIS~2 and~4
-          \citep{Lisiecki.Raymo.2005}.}
+          \citep{Lisiecki.Raymo.2005}.
+        \todo{add LGM outline}}
         \label{fig:cumulative}
@@
     and further upstream where the valley slopes increase. While mountain
-    cirques are poorly captured by the model's horizontal resolution, high
+    cirques erosion processes may not be captured by the model's physics
+    \citep[cf.][]{Sanders.etal.2012} and horizontal resolution, high
     cumulative erosion potential also occurs near the valley heads.
@@
           to the side shows the distribution of topography in the model domain.
-        \textbf{(b)} Same as Fig.~\ref{fig:cumulative}b.}
+        \textbf{(b)} Same as Fig.~\ref{fig:cumulative}b.
+      \todo{test cutting the hypsogram below a minimum number of grid cells}}
       \label{fig:hypsogram}
@@
     to be under 1\,mm\,a$^{-1}$ and restricted the lower parts of the
-    catchment, while much the intra-montane Alps \citep[modelled to be largely
-    cold-based, Fig.~6c in][]{Seguinot.etal.2018}, experience
+    catchment, while much of the intra-montane Alps \citep[modelled to be
+    largely cold-based, Fig.~6c in][]{Seguinot.etal.2018}, experience
     insignificant erosion (Fig.~\ref{fig:transects}a and~d). The modelled
@@
     (constant, present-day) input precipitation rates yield generally higher
-    modelled
+    modelled ice discharge and thus higher
     erosion potential (Fig.~\ref{fig:sensitivity}a, b, and~c), while scenarios
@@
         running mean (dark brown) in relation to modelled total ice volume in
-        centimetres of sea-level equivalent (as in Fig.~\ref{fig:evolution}a).}
+        centimetres of sea-level equivalent (as in Fig.~\ref{fig:evolution}a).
+        \todo{check the numbers here. Make sure that the maps (a-d) fit the
+          profiles (e-h), and then think again how to address the comment.}}
       \label{fig:powerlaws}
@@
     magnitude higher in the Alpine valleys than near the mountain tops
-    (Fig.~\ref{fig:powerlaws}d).
+    (Fig.~\ref{fig:powerlaws}d). This result is compatible with the
+    order-of-magnitude variation of glacier erosion rates observed in the
+    cosmogenic nuclide record \citep{Jansen.etal.2019, Steinemann.etal.2020,
+    Steinemann.etal.2021}.
 
@@
     the exponent, $l$, in the erosion rule (Fig.~\ref{fig:powerlaws}).
+    Nevertheless, it may explain why some areas covered during glacial maxima
+    only, such as some valleys on the southern side of the Alps, appear to have
+    experienced only little glacial modification of their topography.
     During deglaciation periods, increased surface melt at low elevations
@@
     \citep[e.g.,][]{Koppes.Montgomery.2009, Micheletti.Lane.2016,
-    Lane.etal.2017, Bendixen.etal.2017}. Yet this increase has often been
+    Lane.etal.2017, Bendixen.etal.2017}. This increase has often been
     attributed to higher meltwater availability, promoting sediment transport
@@
     In many places throughout the Alps, significant modelled cumulative erosion
-    occurs far inside the Alpine valleys (Fig.~\ref{fig:cumulative}). Yet
-    instantaneous erosion rates show a time-transgressive radial pattern
+    occurs far inside the Alpine valleys (Fig.~\ref{fig:cumulative}). However,
+    the erosion patterns experience spatial shifts through the glacial cycle
     (Fig.~\ref{fig:transects}), so that high-altitude rapid erosion is not
@@
     Giffre) would date from periods of intermediate glacier extension.
-    While the validity of the model results at high elevation are
-    limited by the 1-km resolution and the shallow-ice physics
-    \citep{Imhof.etal.2019}, the lack of rapid erosion at high-elevations for much
-    of the modelled glacial cycle implies that the numerous Alpine mountain
+    The validity of the model results at high elevation are limited by the 1-km
+    resolution, the absence of glacier cirque erosion processes
+    \citep[cf.][]{Sanders.etal.2012}, and the the shallow-ice physics
+    \citep{Imhof.etal.2019}.
+    the lack of rapid erosion at high elevations for much
+    of the modelled glacial cycle implies that the highest mountain
     cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would have formed
-    over interglacial periods, when glaciers are confined to their cirques
+    over interglacial periods, when glaciers are confined to high terrain
     \citep{Barr.etal.2017, Barr.etal.2019}. Similar processes also occur on
@@
     erosion processes, our quantitative results are very likely inaccurate.
-    Yet, we draw the following qualitative conclusions:
+    However, we draw the following qualitative conclusions:
     \begin{itemize}

### 2021-05-18 articles/erosion/response-referee2.md
Begin work on reviewer #2 generic comments.

@@
 
+Thank you for pointing this out. We are not confident that this result is
+significant. The situation you describe occurs, for instance, during much of
+MIS~5 (Fig.~2) and can be visualized in the animations. During such periods,
+glaciers are steep and only a few grid cells in width. Their velocities and
+thus computed erosion rates may be overestimated by the limiting horizontal
+resolution and the shallow-ice physics (Imhof et al., 2019).
+
+Increasing ice volume yields the build-up of less steep and thus slower-flowing
+valley glaciers. Decreasing ice volume yields to (roughly) equally steep but
+smaller glaciers. In the case of non-linear erosion laws ('Koppes' and 'Herman'
+laws), both yield a decrease of erosion volume, hence the local peak in erosion
+volume. We now emphasize this uncertainty in several of our plots by using
+hatches over areas of limited glacier cover, documented in the figure captions.
+
 > As the comment (lines 221-222) on erosion distribution during advance,
@@
 > phase.
+
+TODO: Isn't this current Fig. 4?
 
@@
 
+TODO: discuss lithology, go through entire text?
+
 > A further necessary geographical qualification concerns the misfit between
@@
 
+TODO: add LGM outline, discuss under "climate sensitivity".
+
 > Overall, the paper reports on a very worthwhile computer-intensive exercise.
@@
 > (2021) very recent submission to esurf !
+
+TODO: check that out.
 

### 2021-05-19 articles/erosion/response-referee1.md
Add notes about reviewer #1 generic comments.
Add LGM outline on cumulative erosion figure.

@@
 
+TODO: maybe remove some parts of the introduction? Ian?
+
 > I think the study could also well contribute to the discussion on the
@@
 > summing up Fig. 5a.
+
+TODO: try and compute cumulative erosion by elevation bins.
 
@@
 > be discussed..
+
+TODO: I think this will best fit in "age of the glacial landscape".
 
@@
 
-We added the LGM outline.
+The LGM outline was added on Fig. 2.
 
+TODO: discuss shortly the discrepancy in "climate sensitivity".
 TODO: I don't understand the second part of the comment.
@@
 > 4, 688–692.
+
+TODO: I am not sure what to do about this long list of papers. Some of them are
+already cited, some just added in.

### 2021-05-19 articles/erosion/alps-erosion.tex
Add notes about reviewer #1 generic comments.
Begin work on reviewer #2 generic comments.
Add LGM outline on cumulative erosion figure.

@@
         \textbf{(a)} Modelled cumulative (time-integrated) glacial erosion
-          potential over the last glacial cycle.
+          potential over the last glacial cycle and geomorphological
+          reconstruction of Last Glacial Maximum Alpine glacier extent for
+          comparison \citep[solid red line,][]{Ehlers.etal.2011}.
+          The background maps consists of the initial basal topography from
+          SRTM \citep{Jarvis.etal.2008} and Natural Earth Data
+          \citep{Patterson.Kelso.2017}.
         \textbf{(b)} Modelled total ice volume in centimetres of sea-level
@@
           \citep{Lisiecki.Raymo.2005}.
-        \todo{add LGM outline}}
+        \todo{hatches on all plots.}}
         \label{fig:cumulative}
@@
         \textbf{(b)} Same as Fig.~\ref{fig:cumulative}b.
-      \todo{test cutting the hypsogram below a minimum number of grid cells}}
+      \todo{%
+        test cutting the hypsogram below a minimum number of grid cells,
+        try and compute cumulative erosion by elevation bins.}}
       \label{fig:hypsogram}

### 2021-05-19 figures/alpero_cumulative.py

@@ -30,6 +30,7 @@ def main():
     # add map elements
     util.geo.draw_boot_topo(ax)
     util.geo.draw_natural_earth(ax)
+    util.geo.draw_lgm_outline(ax)
 
     # plot time series
     util.ero.plot_series(ax=tsax)

### 2021-05-19 movies/alpcyc_4k.py
Fix colors and main layer filename in movies.
Changes from yesterday. This is really conter-intuitive but extend='both' is
needed to ensure that filled contours are plotted the same colors across plots
with different data limits.

@@ -229,15 +229,16 @@ def figure_mainmap(time, args, background=True):
         ds = ds.hyoga.where_thicker()  # needs to come after interp
 
         # plot specific visual
+        kwargs = dict(ax=ax, add_colorbar=False, extend='both')
         if args.visual == 'bedrock':
             interp.hyoga.plot.bedrock_isostasy(
-                add_colorbar=False, levels=[-100, -50, -20, 0, 2, 5, 10])
+                levels=[-100, -50, -20, 0, 2, 5, 10], **kwargs)
         elif args.visual == 'erosion':
             interp.hyoga.plot.bedrock_erosion(
-                add_colorbar=False, levels=[10**i for i in range(-9, 1)])
+                levels=[10**i for i in range(-9, 1)], **kwargs)
         elif args.visual == 'velsurf':
             interp.hyoga.plot.surface_velocity(
-                add_colorbar=False, vmin=10, vmax=1000)
+                vmin=10, vmax=1000, **kwargs)
         elif args.visual == 'streams':
             interp.hyoga.plot.ice_margin(facecolor='w', alpha=0.75)
             ds.hyoga.plot.surface_velocity_streamplot(
@@ -452,7 +453,7 @@ def main():
     # frame output directories
     outdirs = dict(
         citymap='~/anim/alpcyc_4k_citymap_{0.region}_{0.lang}'.format(args),
-        mainmap='~/anim/alpcyc_4k_{0.visual}_{0.region}_{0.lang}'.format(args),
+        mainmap='~/anim/alpcyc_4k_{0.visual}_{0.region}'.format(args),
         timetag='~/anim/alpcyc_4k_timetag_{0.lang}'.format(args),
         timebar='~/anim/alpcyc_4k_{0.visual}_timebar_{0.lang}'.format(args))
 

### 2021-05-19 articles/erosion/response-referee2.md
Add LGM outline on cumulative erosion figure.
Use advance and retreat colors in evol figure.

@@
 
-TODO: add LGM outline, discuss under "climate sensitivity".
+The LGM outline was added on Fig. 2.
+
+TODO: discuss shortly the discrepancy in "climate sensitivity".
 
@@
 
+TODO: The input SRTM topography was mentioned in [...] and in the caption of
+Fig. 2.
+
 > The lateral constraints on valley glaciers are usually covered by a ‘form
@@
 > than ‘slight’.
+
+We applied different colours for periods of increasing and decreasing ice
+volume. This actually strikingly highlights the two different regimes. Thank
+you for the suggestion. The label was changed to 'potential annual erosion
+volume'.
+
+TODO: not so slight decrease.
 

### 2021-05-19 articles/erosion/alps-erosion.tex
Use advance and retreat colors in evol figure.

@@
       \caption{%
-        Modelled annual (domain-integrated) erosion volume (light brown) and 1-ka
-        running mean (dark brown) in relation to modelled total ice volume in
-        centimetres of sea-level equivalent.}
+        Modelled annual (domain-integrated) erosion volume (light curves) and 1-ka
+        running mean (dark curves) in relation to modelled total ice volume in
+        centimetres of sea-level equivalent. Blue indicates increasing ice
+        volume and brown decreasing ice volume.}
       \label{fig:evolution}

### 2021-05-19 figures/alpero_evolution.py

@@ -29,26 +29,32 @@ def figure():
     # unit conversion and rolling mean
     ds['slvol'] *= 100
     ds['rolling_mean'] = ds.kop2015_rate.rolling(age=100, center=True).mean()
+    ds['growing'] = ds.slvol.differentiate('age') < 0  # age coord decreasing
 
     # plot
-    line0, = ax.plot(ds.slvol, ds.kop2015_rate, c='C11', alpha=0.5)
-    line1, = ax.plot(ds.slvol, ds.rolling_mean, c='C11')
+    lines = (
+        ax.plot(ds.slvol, ds.kop2015_rate.where(ds.growing), c='C1',
+                alpha=0.25)[0],
+        ax.plot(ds.slvol, ds.kop2015_rate.where(~ds.growing), c='C11',
+                alpha=0.25)[0],
+        ax.plot(ds.slvol, ds.rolling_mean.where(ds.growing), c='C1')[0],
+        ax.plot(ds.slvol, ds.rolling_mean.where(~ds.growing), c='C11')[0])
     timetag = ax.text(0.95, 0.95, '', ha='right', va='top',
                       transform=ax.transAxes)
 
     # set axes properties
     ax.set_xlabel('ice volume (cm s.l.e.)')
-    ax.set_ylabel(r'annual erosion volume ($m^3 a^{-1}$)')
+    ax.set_ylabel(r'potential annual erosion volume ($m^3 a^{-1}$)')
     ax.set_yscale('log')
     ax.set_ylim(10**4.3, 10**7.7)
 
     # annotate advance and retreat
     ax.annotate('', xy=(5, 10**6.7), xytext=(25, 10**6.7), arrowprops=dict(
-        arrowstyle='->', color='0.25', lw=1, connectionstyle='arc3,rad=0.25'))
+        arrowstyle='->', color='C11', lw=1, connectionstyle='arc3,rad=0.25'))
     ax.annotate('', xy=(25, 10**5.1), xytext=(5, 10**5.1), arrowprops=dict(
-        arrowstyle='->', color='0.25', lw=1, connectionstyle='arc3,rad=0.25'))
-    ax.text(15, 10**4.6, 'advance', ha='center', va='center')
-    ax.text(15, 10**7.2, 'retreat', ha='center', va='center')
+        arrowstyle='->', color='C1', lw=1, connectionstyle='arc3,rad=0.25'))
+    ax.text(15, 10**4.6, 'advance', color='C1', ha='center', va='center')
[…]

### 2021-05-19 figures/alpero_evolution.py
Remove old evolution figure animation code.

@@ -10,8 +10,8 @@ import hyoga.open
 import util
 
 
-def figure():
-    """Prepare initial animation figure."""
+def main():
+    """Main program called during execution."""
 
     # initialize figure
     fig, ax = apl.subplots_mm(figsize=(85, 80), gridspec_kw=dict(
@@ -32,15 +32,11 @@ def figure():
     ds['growing'] = ds.slvol.differentiate('age') < 0  # age coord decreasing
 
     # plot
-    lines = (
-        ax.plot(ds.slvol, ds.kop2015_rate.where(ds.growing), c='C1',
-                alpha=0.25)[0],
-        ax.plot(ds.slvol, ds.kop2015_rate.where(~ds.growing), c='C11',
-                alpha=0.25)[0],
-        ax.plot(ds.slvol, ds.rolling_mean.where(ds.growing), c='C1')[0],
-        ax.plot(ds.slvol, ds.rolling_mean.where(~ds.growing), c='C11')[0])
-    timetag = ax.text(0.95, 0.95, '', ha='right', va='top',
-                      transform=ax.transAxes)
+    ax.plot(ds.slvol, ds.kop2015_rate.where(ds.growing), c='C1', alpha=0.25)
+    ax.plot(ds.slvol, ds.kop2015_rate.where(~ds.growing), c='C11', alpha=0.25)
+    ax.plot(ds.slvol, ds.rolling_mean.where(ds.growing), c='C1')
+    ax.plot(ds.slvol, ds.rolling_mean.where(~ds.growing), c='C11')
+    ax.text(0.95, 0.95, '', ha='right', va='top', transform=ax.transAxes)
 
     # set axes properties
     ax.set_xlabel('ice volume (cm s.l.e.)')
@@ -62,42 +58,8 @@ def figure():
     ax.text(21, 10**6.3, 'MIS 4', ha='center', va='center')
     ax.text(28, 10**6.2, 'MIS 2', ha='center', va='center')
 
-    # return figure, data and animated artists
-    return fig, ds, lines, timetag
-
-
-def animate(time, ds, lines, timetag):
-    """Update figure data."""
-
-    # replace line data
-    lines[0].set_ydata(ds.kop2015_rate.where(ds.growing))
-    lines[1].set_ydata(ds.kop2015_rate.where(~ds.growing))
-    lines[2].set_ydata(ds.rolling_mean.where(ds.growing))
-    lines[3].set_ydata(ds.rolling_mean.where(~ds.growing))
-
-    # replace text tag
[…]

### 2021-05-19 figures/alpero_evolution.py
Fix small data mismatch in evolution figure.

@@ -17,14 +17,14 @@ def main():
     fig, ax = apl.subplots_mm(figsize=(85, 80), gridspec_kw=dict(
         left=12, right=1.5, bottom=9, top=1.5))
 
-    # load postprocessed data
-    # FIXME there is a data gap after merge, add tolerance
+    # load postprocessed data (note: there seem to be currently no way to merge
+    # two dataset with slight errors using mfdataset)
     with hyoga.open.dataset(
             '../data/processed/alpcyc.1km.epic.pp.ts.10a.nc') as ds:
-        ds = ds[['slvol']]
-    with hyoga.open.dataset(
-            '../data/processed/alpero.1km.epic.pp.agg.nc') as agg:
-        ds['kop2015_rate'] = agg.kop2015_rate
+        with hyoga.open.dataset(
+                '../data/processed/alpero.1km.epic.pp.agg.nc') as agg:
+            agg = agg.reindex_like(ds, method='nearest', tolerance=1e-9)
+            ds = ds.merge(agg, compat='override')
 
     # unit conversion and rolling mean
     ds['slvol'] *= 100

### 2021-05-20 figures/alpero_transects.py
Add 0 ka Rhine topo map, rearrange panels.

@@ -18,20 +18,20 @@ def main():
 
     # initialize figure
     fig, grid = apl.subplots_mm(
-        figsize=(177, 80), ncols=3, sharex=True, sharey=True,
+        figsize=(177, 80), ncols=4, sharex=True, sharey=True,
         subplot_kw=dict(projection=ccrs.UTM(32)), gridspec_kw=dict(
-            left=1.5, right=1.5, bottom=40.5, top=1.5, wspace=1.5))
-    tsax = fig.add_axes_mm([15, 9, 177-30-3-1.5, 30])
-    cax = fig.add_axes_mm([177-18+1.5, 9, 3, 30])
+            left=1.5, right=19, bottom=40.5, top=1.5, wspace=1.5))
+    tsax = fig.add_axes_mm([1.5, 9, 177-1.5-19, 30])
+    cax = fig.add_axes_mm([177-19+1.5, 40.5, 3, 38])
 
     # set extent and subfig labels
-    for ax, label in zip(grid, 'abc'):
+    for ax, label in zip(grid, 'abcd'):
         util.fig.prepare_map_axes(ax)
         util.fig.add_subfig_label('(%s)' % label, ax=ax)
-        ax.set_extent([410e3, 620e3, 5160e3, 5300e3], crs=ax.projection)
+        ax.set_extent([440e3, 580e3, 5160e3, 5300e3], crs=ax.projection)
 
     # plot ages and levels for consistency
-    ages = [24, 20, 16]
+    ages = [36, 24, 16, 0]
     levels = [10**i for i in range(-9, 1)]
 
     # read profile coordinates
@@ -66,10 +66,13 @@ def main():
         ax.plot(x, y, color='0.25', dashes=(2, 1))
         ax.plot(x[0], y[0], color='0.25', marker='o')
 
-        # add age tag and vertical line
+        # add age tag
         ax.set_title('')
-        util.fig.add_subfig_label('%d ka' % age, ax=ax, ha='right', va='bottom')
+        util.fig.add_subfig_label('%d ka' % age, ax=ax, ha='right', va='top')
+
+        # add profile axes vertical line
         tsax.axvline(age, color='0.25', dashes=(2, 1))
+        tsax.plot(age, 0, color='0.25', marker='o')
 
     # Profiles
     # --------
@@ -86,7 +89,7 @@ def main():
         (ds.assign_coords(d=ds.d/1e3).kop2015_rhin*1e3).plot.contourf(
[…]

### 2021-05-20 articles/erosion/response-referee2.md
Document transects fig changes in ms and response.

@@
 significant. The situation you describe occurs, for instance, during much of
-MIS~5 (Fig.~2) and can be visualized in the animations. During such periods,
+MIS~5 (Fig. 2) and can be visualized in the animations. During such periods,
 glaciers are steep and only a few grid cells in width. Their velocities and
@@
 
-TODO: Isn't this current Fig. 4?
+Figure 4 was reworked to include a map of Rhine glacier modelled erosion rates
+at 36 ka, corresponding to the last marjor advance phase before the LGM and with
+somewhat similar extent to the deglacial state of 16 ka. This also corresponds
+to the topographic profiles shown on Fig. 8.
 
@@
 > map or hillshade of the Rhine area on the same scale should be provided.
+
+In addition to the aforementioned changes to Fig. 4, a new panel was added for
+the model final state at 0 ka, revealing the bedrock topography.
 

### 2021-05-20 articles/erosion/alps-erosion.tex
Document transects fig changes in ms and response.

@@
       \caption{%
-        \textbf{(a, b, c)} Modelled instantaneous erosion rate of the Rhine
-          Glacier for selected post-Last Glacial Maximum ages.
-        \textbf{(d)} Interpolated instantaneous erosion rate along a Rhine
+        \textbf{(a--d)} Modelled instantaneous erosion rate of the Rhine
+          Glacier for selected glacier advance and retreat ages, and the final
+          model state for topographic reference.
+        \textbf{(e)} Interpolated instantaneous erosion rate along a Rhine
           Glacier transect for the entire last glacial cycle (upper panels
@@
     largely cold-based, Fig.~6c in][]{Seguinot.etal.2018}, experience
-    insignificant erosion (Fig.~\ref{fig:transects}a and~d). The modelled
+    insignificant erosion (Fig.~\ref{fig:transects}a and~b). The modelled
     erosion rates both increase and propagate inwards during periods of glacier
-    retreat (Fig.~\ref{fig:transects}b, c and~d).
+    retreat (Fig.~\ref{fig:transects}c and~e).
 

### 2021-05-20 figures/alpero_transects.py

@@ -79,7 +79,7 @@ def main():
 
     # plot marine isotope stages
     util.fig.plot_mis(ax=tsax, y=0.9)
-    util.fig.add_subfig_label('(d)', ax=tsax)
+    util.fig.add_subfig_label('(e)', ax=tsax)
 
     # open aggregated output
     with hyoga.open.dataset(

### 2021-05-20 figures/alpero_powerlaws.py
Fewer levels, improve powerlaws fig readability.

@@ -38,11 +38,11 @@ def main():
     fig = apl.figure_mm(figsize=(177, 119))
     grid = fig.subplots_mm(
         ncols=4, subplot_kw=dict(projection=ccrs.UTM(32)), gridspec_kw=dict(
-            left=15, right=1.5, bottom=75.5, top=4.5, wspace=1.5))
+            left=1.5, right=19, bottom=75.5, top=4.5, wspace=1.5))
     tsgrid = fig.subplots_mm(
         ncols=4, nrows=2, sharex='row', sharey='row', gridspec_kw=dict(
-            left=15, right=1.5, bottom=9, top=45, hspace=9, wspace=1.5))
-    cax = fig.add_axes_mm([10.5, 75.5, 3, 39])
+            left=1.5, right=19, bottom=9, top=45, hspace=9, wspace=1.5))
+    cax = fig.add_axes_mm([177-19+1.5, 75.5, 3, 38])
 
     # set extent and subfig labels
     for ax, label in zip(list(grid)+list(tsgrid.flat), 'abcdefghijkl'):
@@ -72,7 +72,7 @@ def main():
             # plot cumulative erosion
             cset = ds[ref+'_cumu'].plot.contourf(
                 ax=ax, add_colorbar=False, alpha=0.75, cmap='YlOrBr',
-                extend='both', levels=[10**i for i in range(-2, 5)])
+                extend='both', levels=[10**i for i in range(-1, 4)])
 
             # plot background topo and ice margin
             util.geo.draw_boot_topo(ax)
@@ -94,6 +94,9 @@ def main():
             pfax.set_xlabel('distance along flow (km)')
             pfax.set_ylabel('')
             pfax.set_yscale('log')
+            pfax.set_yticks([1e-3, 1e-1, 1e1, 1e3])
+            pfax.yaxis.set_ticks_position('right')
+            pfax.yaxis.set_label_position('right')
 
             # plot time series
             tsax.plot(slvol*100, ds[ref+'_rate'], c='C11', alpha=0.5)
@@ -105,14 +108,17 @@ def main():
             tsax.set_yscale('log')
             tsax.set_ylabel('')
             tsax.set_ylim(10**4.5, 10**10.5)
+            tsax.set_yticks([1e5, 1e7, 1e9])
+            tsax.yaxis.set_ticks_position('right')
+            tsax.yaxis.set_label_position('right')
 
[…]

### 2021-05-20 articles/erosion/response-referee1.md
Document power laws fig changes in response.

@@
 
-TODO: check the numbers here. Make sure that the maps (a-d) fit the profiles
-(e-h), and then think again how to address the comment.
+TODO: "what you mean by realistic"
+
+There is no mismatch between panels (a-d) and (e-h), but to improve readability
+we have reduced the number of colour levels on the maps, and changed the
+positions of ticks on the transects.
 

### 2021-05-22 figures/alpero_cumulative.py
Add scale bar on all erosion maps.

@@ -6,6 +6,7 @@
 """Plot Alps erosion cumulative."""
 
 import hyoga.open
+import cartowik.decorations as cde
 import util
 
 
@@ -31,6 +32,7 @@ def main():
     util.geo.draw_boot_topo(ax)
     util.geo.draw_natural_earth(ax)
     util.geo.draw_lgm_outline(ax)
+    cde.add_scale_bar(ax, label='100 km', length=100e3, pad=220e3)
 
     # plot time series
     util.ero.plot_series(ax=tsax)

### 2021-05-22 figures/alpero_powerlaws.py

@@ -3,11 +3,12 @@
 # Creative Commons Attribution-ShareAlike 4.0 International License
 # (CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/)
 
-"""Plot Rhine glacier velocity and erosion potential."""
+"""Plot Alps erosion choice of power law."""
 
 import numpy as np
 import cartopy.crs as ccrs
 import cartowik.profiletools as cpf
+import cartowik.decorations as cde
 import absplots as apl
 import hyoga.open
 import util
@@ -111,6 +112,9 @@ def main():
             tsax.set_yticks([1e5, 1e7, 1e9])
             tsax.yaxis.set_ticks_position('right')
             tsax.yaxis.set_label_position('right')
+
+        # add scale bar on first panel
+        cde.add_scale_bar(grid.flat[0], label='50 km', length=50e3, pad=10e3)
 
     # add y-labels on rightmost axes
     tsgrid[0, -1].set_ylabel('erosion potential (m)')

### 2021-05-22 figures/alpero_sensitivity.py

@@ -7,6 +7,7 @@
 
 import matplotlib.pyplot as plt
 import cartopy.crs as ccrs
+import cartowik.decorations as cde
 import absplots as apl
 import hyoga.open
 import util
@@ -30,7 +31,8 @@ def main():
         # prepare axes and add labels
         util.fig.prepare_map_axes(ax)
         util.fig.add_subfig_label('(%s) ' % 'adbecf'[i], ax=ax)
-        util.com.add_corner_tag(label, ax=ax, va='bottom', y=(i == 5)*1/3)
+        util.com.add_corner_tag(
+            label, ax=ax, va='bottom', x=(i == 0)*0.45, y=(i == 5)*1/3)
 
         # load aggregated data
         with hyoga.open.dataset(
@@ -47,6 +49,9 @@ def main():
         # add map elements
         util.geo.draw_boot_topo(ax=ax, filename='alpcyc.2km.in.nc')
         util.geo.draw_natural_earth(ax)
+
+    # add scale bar on first panel
+    cde.add_scale_bar(grid.T.flat[0], label='200 km', length=200e3, pad=50e3)
 
     # cut out white rectangle
     ax.spines['geo'].set_visible(False)

### 2021-05-22 figures/alpero_transects.py

@@ -8,6 +8,7 @@
 import matplotlib as mpl
 import cartopy.crs as ccrs
 import cartowik.profiletools as cpf
+import cartowik.decorations as cde
 import absplots as apl
 import hyoga.open
 import util
@@ -73,6 +74,9 @@ def main():
         # add profile axes vertical line
         tsax.axvline(age, color='0.25', dashes=(2, 1))
         tsax.plot(age, 0, color='0.25', marker='o')
+
+    # add scale bar on second panel
+    cde.add_scale_bar(grid[0], label='50 km', length=50e3, pad=10e3)
 
     # Profiles
     # --------

### 2021-05-22 articles/erosion/alps-erosion.tex
Add technical corrections from reviewer #2.

@@
     higher variety of glacier erosional landforms, whose implications on
-    glacial history is yet to be understood.
+    glacial history are yet to be understood.
     The cosmogenic nuclide memory of bedrock erosion provides a more
@@
 
-    In absence of other landforms, glacially eroded topography has sometimes
-    been used as a proxy for glacier extent, for instance in
+    In the absence of other landforms, glacially eroded topography has
+    sometimes been used as a proxy for glacier extent, for instance in
     geomorphological mapping \citep[e.g.,][]{Margold.etal.2011, Fu.etal.2012}.
@@
     \citep{Willenbring.Von-Blanckenburg.2010}. Examination of contemporary
-    erosion rates across different climates suggest that increased temperatures
+    erosion rates across different climates suggests that increased temperatures
     in glaciated regions lead to higher erosion rates and that erosion rates
@@
     previously published model results \citep{Seguinot.etal.2018} and the
-    patterns of modelled erosion rates and cumulative last glacial cycle erosion
+    patterns of modelled erosion potential and cumulative last glacial cycle
+    erosion
     potential are analysed. We examine the glaciological conditions that cause
@@
     temperature and water-content dependent creep, pseudo-plastic basal
-    sliding, bedrock deformation under the ice
+    sliding, bedrock isostatic deformation under the ice
     load, and a~positive degree-day (PDD) surface mass balance model. The model
@@
     ice and bedrock temperature at 120\,ka (thousand years before the present),
-    and ran to the present. The ice-sheet model physical parameters are listed
+    and is run to the present. The ice-sheet model physical parameters are listed
     in the earlier open-access publication \citep{Seguinot.etal.2018}, and the
@@
     Centre for Medium-Range Weather Forecasts Reanalysis Interim
-    \citep[ERA-Interim;][]{Dee.etal.2011}, amended with temperature lapse-rate
-    corrections from the European Project for Ice Coring in Antarctica
+    \citep[ERA-Interim;][]{Dee.etal.2011}, amended with temperature changes
+    from the European Project for Ice Coring in Antarctica
     \citep[EPICA;][] {Jouzel.etal.2007}, and time-dependent paleo-precipitation
@@
     catchment, while much of the intra-montane Alps \citep[modelled to be
-    largely cold-based, Fig.~6c in][]{Seguinot.etal.2018}, experience
+    largely cold-based, Fig.~6c in][]{Seguinot.etal.2018} experience
     insignificant erosion (Fig.~\ref{fig:transects}a and~b). The modelled
@@
 
-    The modelled, domain-integrated glacier erosion volume do not increase
+    The modelled, domain-integrated glacier erosion volume does not increase
     together with ice volume. On the opposite, periods of ice advance and
@@
     For instance, low-elevation piedmont over-deepened basins such as Lake
-    Constance (Fig.~\ref{fig:landscape}), Lake Geneva and Lake Maggiore, are
-    here modelled to be the product of extensive, Last Glacial Maximum-like
+    Constance (Fig.~\ref{fig:landscape}), Lake Geneva and Lake Maggiore are
+    here modelled to be the product of extensive Last Glacial Maximum-like
     glaciations. Other modeling work suggests that features such as these

### 2021-05-23 articles/erosion/alps-erosion.tex
Update hypsogram fig caption for new histograms.

@@
       \caption{%
-        \textbf{(a)} Modelled erosion rate ``hypsogram'', showing the geometric
-          mean of (non-zero) modelled erosion rates in 10-m elevation bands
-          across the entire model domain and its time evolution. The histogram
-          to the side shows the distribution of topography in the model domain.
-        \textbf{(b)} Same as Fig.~\ref{fig:cumulative}b.
-      \todo{%
-        test cutting the hypsogram below a minimum number of grid cells,
-        try and compute cumulative erosion by elevation bins.}}
+        \textbf{(a)} Potential erosion rate ``hypsogram'', showing the
+          geometric mean of (non-zero) modelled rates in 10-m elevation bands
+          across the entire model domain and its time evolution.
+        \textbf{(b)} Distribution of model domain bedrock topography (grey
+          bars) and glaciated topography (blue bars) in 100-m elevation bands,
+          and the corresponding cumulative potential erosion volume (brown
+          line).
+        \textbf{(c)} Same as Fig.~\ref{fig:cumulative}b.
+        \todo{fix labels, test hatching below a minimum number of grid cells}}
       \label{fig:hypsogram}

### 2021-05-23 articles/erosion/alps-erosion.tex
Add refs to SRTM, paleotemperature records.

@@
     load, and a~positive degree-day (PDD) surface mass balance model. The model
-    is initialized with assumed present-day ice thickness and equilibrium
+    is initialized with assumed present-day ice thickness and bedrock topography
+    from SRTM \citep{Jarvis.etal.2008},
     ice and bedrock temperature at 120\,ka (thousand years before the present),
@@
     reductions \citep{Huybrechts.2002}. Lower-resolution runs based on
-    alternative paleoclimate scenarios \citep{Seguinot.etal.2018} are analyzed
+    alternative paleoclimate scenarios \citep{Seguinot.etal.2018}, forced by
+    temperature changes from the the Greenland Ice Core Project
+    \citep[GRIP;][]{Dansgaard.etal.1993} and and an oceanic sediment core from
+    the Iberian margin \citep[MD01-2444;][]{Martrat.etal.2007}, are analyzed
     in the discussion part (Sect.~\ref{sec:sensitivity}).
@@
         Modelled cumulative glacial erosion potential over the last glacial
-        cycle without \textbf{(a, b, c)} and with \textbf{(d, e, f)}
-        paleo-precipitation corrections, and using three different
-        palaeo-temperature histories \citep[see][]{Seguinot.etal.2018}.}
+        cycle using three different paleo-temperature histories from
+        \textbf{(a, d)} the Greenland Ice Core Project
+        \citep[GRIP;][]{Dansgaard.etal.1993}, \textbf{(b, e)} the European
+        Project for Ice Coring in Antarctica \citep[our default,
+        EPICA;][]{Jouzel.etal.2007}, and \textbf{(c, f)} an oceanic sediment
+        core from the Iberian margin \citep[MD01-2444;][]{Martrat.etal.2007},
+        and both \textbf{(a--c)} without and \textbf{(d--f)} with
+        paleo-precipitation reductions \citep[cf.][]{Seguinot.etal.2018}.}
       \label{fig:sensitivity}

### 2021-05-23 articles/erosion/alps-erosion.tex
Complete reread, write "potential erosion rate".

@@
     modelling results, and empirical inferences of bedrock erosion under modern
-    glaciers, we compute a distribution of potential glacier erosion in the Alps
-    over the last glacial cycle from 120\,000 years ago to the present.
+    glaciers, we compute a distribution of potential glacier erosion in the
+    Alps over the last glacial cycle from 120\,000 years ago to the present.
     %
-    Despite large uncertainties pertaining to the climate history of the Alps and
-    unconstrained glacier erosion processes, the resulting modelled patterns of
-    glacier erosion include persistent features. The cumulative imprint of
-    the last glacial cycle shows a very strong localization of glacier erosion
-    with local maxima at the mouths of major Alpine valleys and some other
-    upstream sections where glaciers are modelled to have flown with the
-    highest velocity. The modelled erosion rates vary significantly through the
-    glacial cycle, but show paradoxically little relation to the total glacier
-    volume. Phases of glacier advance and maximum extension see a localization
-    of rapid erosion rates
-    at low elevation, while glacier erosion at higher elevation is modelled
-    date from phases of less extensive glaciation. The modelled erosion rates
-    peak during deglaciation phases, when frontal retreat results in steeper
-    glacier surface slopes, implying that climatic conditions that result in
-    rapid glacier erosion might be quite transient and specific.
+    Despite large uncertainties pertaining to the climate history of the Alps
+    and unconstrained glacier erosion processes, the resulting modelled
+    patterns of glacier erosion include persistent features. The cumulative
+    imprint of the last glacial cycle shows a very strong localization of
+    erosion potential with local maxima at the mouths of major Alpine valleys
+    and some other upstream sections where glaciers are modelled to have flown
+    with the highest velocity. The potential erosion rates vary significantly
+    through the glacial cycle, but show paradoxically little relation to the
+    total glacier volume. Phases of glacier advance and maximum extension see a
+    localization of rapid potential erosion rates at low elevation, while
+    glacier erosion at higher elevation is modelled date from phases of less
+    extensive glaciation. The modelled erosion rates peak during deglaciation
+    phases, when frontal retreat results in steeper glacier surface slopes,
+    implying that climatic conditions that result in rapid glacier erosion
+    might be quite transient and specific.
     %
     Our results depict the Alpine glacier erosion landscape as a
-    time-transgressive patchwork, with different parts of the
-    range corresponding to different glaciation stages and time
-    periods.
+    time-transgressive patchwork, with different parts of the range
+    corresponding to different glaciation stages and time periods.
 
@@
     \citep[e.g.,][]{Anderson.etal.2012,Mariotti.etal.2021}.
+    \todo{shorten this paragraph.}
 
@@
     to quantify erosion \citep{Alley.etal.2019}. Physically-based models
-    describe quarrying process and abrasion of bedrock by debris laden ice
+    describe quarrying process and abrasion of bedrock by debris-laden ice
     \citep[e.g.,][]{Alley.etal.1997, Iverson.2012, Beaud.etal.2014}. Describing
@@
     difficult to implement in many cases due to the large number of poorly
-    constrained parameters and process, i.e. water pressure fluctuations and
+    constrained parameters and processes, i.e. water pressure fluctuations and
     subglacial debris concentration \citep[e.g.,][]{Hallet.1979, Seguinot.2008,
@@
     velocities from 15~Patagonian and Antarctic Peninsula tidewater glaciers
-    \citep{Koppes.etal.2015}. We discuss the role of alternative values of
+    \citep[using the full dataset from][including outliers]{Koppes.etal.2015}.
+    We discuss the role of alternative values of
     $K_\mathrm{g}$ and $l$ in the discussion part (Sect.~\ref{sec:powerlaws}).
@@
     bedrock topography and, then onto, ice dynamics
-    \citep[e.g.,][]{Anderson.etal.2012}. Furthermore, we assume that once
-    material is eroded, it is transported out of the system. In turn, we do not
-    account for the downstream advection of debris
-    \citep[e.g.,][]{Anderson.Anderson.2016}, or fluvial transport of subglacial
-    sediment \citep{Delaney.etal.2019}. Instead, we compute the
-    time-integrated glacier erosion rate,
+    \citep[e.g.,][]{Anderson.etal.2012}. Furthermore, we do not account for the
+    downstream advection of debris \citep[e.g.,][]{Anderson.Anderson.2016},
+    or fluvial transport of subglacial sediment \citep{Delaney.etal.2019}.
+    Instead we assume that eroded material is instantly transported out of the
+    system, thus neglegting its role in shielding the bedrock from glacier
+    erosion in zones of temporary storage \citep{Preusser.etal.2010}. Neither
+    do we account for differences in erosion effectiveness on different
+    lithologies or erosion from subglacial and interglacial hydrologic
+    processes. For the aforementioned reasons, we refer to the above computed
+    rates, $\dot{e}$, as ``potential erosion rates''.
+
+    Secondly, we compute the time-integrated glacier erosion rate,
     %
@@
     %
-    and refer to it as the cumulative ``erosion potential''. The integration
+    and refer to it as the ``cumulative erosion potential''. The integration
     is numerically approximated using a time-step of 10\,a in the
@@
         \textbf{(b)} Modelled total ice volume in centimetres of sea-level
-          equivalent (cm~s.l.e., black), annual (domain-integrated) erosion
-          volume (light brown) and 1-ka running mean (dark brown). Shaded gray
-          areas indicate the timing for MIS~2 and~4
+          equivalent (cm~s.l.e., black), annual (domain-integrated) potential
+          erosion volume (light brown) and its 1-ka running mean (dark brown).
+          Shaded gray areas indicate the timing for MIS~2 and~4
           \citep{Lisiecki.Raymo.2005}.
@@
       \caption{%
-        Modelled annual (domain-integrated) erosion volume (light curves) and 1-ka
-        running mean (dark curves) in relation to modelled total ice volume in
-        centimetres of sea-level equivalent. Blue indicates increasing ice
-        volume and brown decreasing ice volume.}
+        Modelled annual (domain-integrated) potential erosion volume (light
+        curves) and its 1-ka running mean (dark curves) in relation to the
+        modelled total ice volume in centimetres of sea-level equivalent. Blue
+        indicates increasing ice volume and brown decreasing ice volume.}
       \label{fig:evolution}
@@
 
-    The modelled annual (domain-integrated) glacial erosion volume
+    The modelled annual (domain-integrated) potential glacial erosion volume
     does not systematically correlate with the modelled total ice volume
@@
     between these two quantities (Fig.~\ref{fig:evolution}). After significant
-    ice volume has been reached, there is a slight tendency for slower erosion
+    ice volume has been reached, there is a general tendency for slower erosion
     during periods of extensive glaciation (Fig.~\ref{fig:evolution}). However,
@@
       \caption{%
-        \textbf{(a--d)} Modelled instantaneous erosion rate of the Rhine
-          Glacier for selected glacier advance and retreat ages, and the final
-          model state for topographic reference.
-        \textbf{(e)} Interpolated instantaneous erosion rate along a Rhine
-          Glacier transect for the entire last glacial cycle (upper panels
-          dashed line).}
+        \textbf{(a--d)} Modelled instantaneous potential erosion rate of the
+          Rhine Glacier for selected glacier advance and retreat ages, and the
+          final model state for topographic reference.
+        \textbf{(e)} Interpolated instantaneous potential erosion rate along a
+          Rhine Glacier transect (upper panels dashed line) for the entire last
+          glacial cycle.}
       \label{fig:transects}
@@
     largely cold-based, Fig.~6c in][]{Seguinot.etal.2018} experience
-    insignificant erosion (Fig.~\ref{fig:transects}a and~b). The modelled
-    erosion rates both increase and propagate inwards during periods of glacier
-    retreat (Fig.~\ref{fig:transects}c and~e).
+    insignificant erosion potential (Fig.~\ref{fig:transects}a and~b). The
+    potential erosion rates both increase and propagate inwards during periods
+    of glacier retreat (Fig.~\ref{fig:transects}c and~e).
 
@@
     periods of modelled increasing and maximum ice volume correspond to lower
-    values for elevation-aggregated erosion rates, with significant erosion
+    values for elevation-aggregated potential erosion rates, with significant
+    erosion potential
     restricted to lower elevations. On the opposite, periods of modelled
     decreased ice volume correspond to higher local modelled erosion rates and
-    more extensive rapid erosion into higher-elevation areas
+    more extensive rapid erosion potential into higher-elevation areas
     (Fig.~\ref{fig:hypsogram}).
@@
     glaciation histories \citep[cf.][]{Seguinot.etal.2018} and thus different
-    results in terms of erosion
+    results in terms of glacial cycle cumulative erosion
     potential (Fig.~\ref{fig:sensitivity}). Scenarios including higher
@@
     modelled ice discharge and thus higher
-    erosion potential (Fig.~\ref{fig:sensitivity}a, b, and~c), while scenarios
-    including paleo-precipitation reductions (Fig.~\ref{fig:sensitivity}d, e,
-    and~f), including our default high-resolution run
+    erosion potential (Fig.~\ref{fig:sensitivity}a--c), while scenarios
+    including paleo-precipitation reductions (Fig.~\ref{fig:sensitivity}d--f),
+    including our default high-resolution run
     (Fig.~\ref{fig:cumulative}), yield a smaller erosion potential. The spatial
@@
         running mean (dark brown) in relation to modelled total ice volume in
-        centimetres of sea-level equivalent (as in Fig.~\ref{fig:evolution}a).
-        \todo{check the numbers here. Make sure that the maps (a-d) fit the
-          profiles (e-h), and then think again how to address the comment.}}
+        centimetres of sea-level equivalent (as in Fig.~\ref{fig:evolution}a).}
       \label{fig:powerlaws}
@@
     included in the global compilation by \citet{Cook.etal.2020}.
+    \todo{rework this paragraph. The realistic laws lies in-between.}
 
@@
     on the substratum \citep{Steinemann.etal.2020, Steinemann.etal.2021}, and
-    it arguable that a layer of till and proglacial sediments may have armored
-    the glacier bed from erosion for at least parts of the glacial cycle.
-    Given the lack of data to constrain the erosion rule, it can not
-    be excluded that a linear or sub-linear erosion-law
+    it is arguable that a layer of till and proglacial sediments may have
+    armored the glacier bed from erosion for at least parts of the glacial
+    cycle. Thus, given the lack of data to constrain the erosion rule, it can
+    not be excluded that a linear or sub-linear erosion law
     \citep[e.g.,][]{Cook.etal.2020} would lead to a less localized pattern of
@@
     between modelled ice volume and erosion (Fig.~\ref{fig:powerlaws}k and~l).
+    \todo{discuss Koppes without outliers}
     Even then, the intrinsically large variations in modelled glacier sliding
@@
     magnitude higher in the Alpine valleys than near the mountain tops
-    (Fig.~\ref{fig:powerlaws}d). This result is compatible with the
+    (Fig.~\ref{fig:powerlaws}d).
+    \todo{discuss sliding.}
+    This result is compatible with the
     order-of-magnitude variation of glacier erosion rates observed in the
@@
     together with ice volume. On the opposite, periods of ice advance and
-    maximum glaciation correspond to comparatively slow erosion
+    maximum glaciation correspond to comparatively slow erosion potential
     (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).
@@
     Fig.~\ref{fig:evolution}), it demonstrates that climatic conditions
-    over-which glacier erosion occurs are relatively specific and likely
+    over which glacier erosion occurs are relatively specific and likely
     transient. Other field-based studies have evidenced increased glacier
@@
     occurs far inside the Alpine valleys (Fig.~\ref{fig:cumulative}). However,
-    the erosion patterns experience spatial shifts through the glacial cycle
-    (Fig.~\ref{fig:transects}), so that high-altitude rapid erosion is not
-    modelled to date from expansive glaciation stages, but instead to stages of
-    intermediate glacier cover
+    the potential erosion patterns experience spatial shifts through the
+    glacial cycle (Fig.~\ref{fig:transects}), so that high-altitude rapid
+    erosion is not modelled to date from expansive glaciation stages, but
+    instead from stages of intermediate glacier cover
     \citep[Fig.~\ref{fig:hypsogram};][]{Barr.etal.2019}.
@@
     here modelled to be the product of extensive Last Glacial Maximum-like
-    glaciations. Other modeling work suggests that features such as these
+    glaciations. Other modelling work suggests that features such as these
     likely form when water can access the glacier bed and encourage glacier
@@
     resolution, the absence of glacier cirque erosion processes
-    \citep[cf.][]{Sanders.etal.2012}, and the the shallow-ice physics
-    \citep{Imhof.etal.2019}.
+    \citep[cf.][]{Sanders.etal.2012}, and the shallow-ice glacier flow
+    physics \citep{Imhof.etal.2019}. However,
     the lack of rapid erosion at high elevations for much
@@
     Seguinot.etal.2018}, but to a time-transgressive upper limit of erosion from
-    retreating glaciers on steep terrain.
+    advancing and retreating glaciers on steep terrain.
 
@@
       \item Increased gravitational drag due to surface profile steepening
-        provide a mechanism
+        provides a mechanism
         for accelerated erosion during deglaciation periods, irrespective of

### 2021-05-23 articles/erosion/response-referee1.md
Discuss cristalline massif high erosion.

@@
 
-TODO: I think this will best fit in "age of the glacial landscape".
+More accurate modelling of small-scale glaciers would be needed to fully
+address this point. Steep topography though is likely to be part of the
+explanation, and a more substantial discussion of the high erosion potential
+values in these areas and the model limitations that hinder interpretation was
+added in the "age of the glacial landscape" subsection.
 

### 2021-05-23 articles/erosion/alps-erosion.tex
Discuss cristalline massif high erosion.

@@
     Giffre) would date from periods of intermediate glacier extension.
-    The validity of the model results at high elevation are limited by the 1-km
-    resolution, the absence of glacier cirque erosion processes
-    \citep[cf.][]{Sanders.etal.2012}, and the shallow-ice glacier flow
-    physics \citep{Imhof.etal.2019}. However,
+
+    The validity of the model results at high elevation is discussable.
+    Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
+    Ötztal and Tauern Massifs locally exhibit a strikingly high erosion
+    potential. However, the computation of glacier flow velocities on such
+    steep surfaces is strongly limited by the model horizontal resolution of
+    1\,km, the shallow-ice glacier flow physics \citep{Imhof.etal.2019}, and
+    PISM's current mass-conservation heuristics \citep{Imhof.2021}. Besides,
+    bergschrund (rimaye) processes likely to dominate interglacial cirque
+    erosion at such altitudes \citep{Sanders.etal.2012} are not captured by the
+    velocity-based glacier erosion power-laws. Nevertheless,
     the lack of rapid erosion at high elevations for much

### 2021-05-23 articles/erosion/alps-erosion.tex
Add a paragraph to discuss erosion and sliding.

@@
     \todo{discuss Koppes without outliers}
-    Even then, the intrinsically large variations in modelled glacier sliding
+
+    The modelled pattern of erosion potential depends on PISM glacier physics
+    and sliding model parameters. In particular, the pseudo-plastic sliding law
+    exponent \citep[$q=0.25$ in][]{Seguinot.etal.2018} controls the sharpness
+    of the transition between adherent and decoupled basal conditions. Recent
+    ensemble validation of Antarctic Ice Sheet glacial-cycle simulations
+    against geological and present-day observations \citep{Albrecht.etal.2020a,
+    Albrecht.etal.2020b} support a higher value of $q=0.75$, and thus a
+    sliding law closer to linear, which would perhaps result in a smoother
+    distribution of sliding velocity and erosion potential. Lateral stress
+    gradients missing from the shallow-shelf approximation stress balance could
+    also contribute to moderate sliding velocity in narrow troughs
+    \citep{Herman.etal.2011, Egholm.etal.2012a, Egholm.etal.2012b,
+    Pedersen.etal.2014}. \todo{check these references}.
+
+    Because of the intrinsically large variations in modelled glacier sliding
     velocities yield cumulative glacial-cycle erosion two to three orders of
     magnitude higher in the Alpine valleys than near the mountain tops
-    (Fig.~\ref{fig:powerlaws}d).
-    \todo{discuss sliding.}
-    This result is compatible with the
+    (Fig.~\ref{fig:powerlaws}d), we argue that the modelled highly localized
+    pattern of erosion potential is qualitatively robust. This pattern, and
+    the lithologic variations in bedrock erodability, may explain
     order-of-magnitude variation of glacier erosion rates observed in the

### 2021-05-24 articles/erosion/alps-erosion.tex
Fix new double-papers references.

@@
     ensemble validation of Antarctic Ice Sheet glacial-cycle simulations
-    against geological and present-day observations \citep{Albrecht.etal.2020a,
-    Albrecht.etal.2020b} support a higher value of $q=0.75$, and thus a
+    against geological and present-day observations \citep{Albrecht.etal.2020,
+    Albrecht.etal.2020a} support a higher value of $q=0.75$, and thus a
     sliding law closer to linear, which would perhaps result in a smoother
@@
     also contribute to moderate sliding velocity in narrow troughs
-    \citep{Herman.etal.2011, Egholm.etal.2012a, Egholm.etal.2012b,
+    \citep{Herman.etal.2011, Egholm.etal.2012, Egholm.etal.2012a,
     Pedersen.etal.2014}. \todo{check these references}.
@@
     Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
-    Ötztal and Tauern Massifs locally exhibit a strikingly high erosion
+    \"Otztal and Tauern Massifs locally exhibit a strikingly high erosion
     potential. However, the computation of glacier flow velocities on such

### 2021-05-24 articles/erosion/response-referee1.md
Add a paragraph to discuss erosion and sliding.
Polish response #1, some points still missing.

--- 2021-05-20 articles/erosion/response-referee1.md
@@
 
-Thank you very much for your detailed review of our manuscript.
+Thank you very much for your positive comments and for your time spent
+reviewing our manuscript.
 
@@
 
+Thank you for this accurate summary of our paper.
+
 ## General comments
@@
 > extension).
+
+Indeed, it is our deliberate choice to avoid paraphrasing or duplicating
+content from the 2018 paper, where the preparation of the text and figures also
+involved a different author team. We believe this is justified given that both
+papers will be published open-access. Nonetheless, small additions were made.
+Please see our responses to your technical comments below.
 
@@
 
-TODO: maybe remove some parts of the introduction? Ian?
+TODO: This remain a tricky points to me. Shoulw we remove parts of the
+introduction? Or add sentences tying our statements to upcoming results?
 
@@
 
-TODO: try and compute cumulative erosion by elevation bins.
+Thank you for the suggestion. Figure 5 was amended with (as Fig. 5b) the
+distribution of cumulative erosion potential over 100 m elevation bins.
+The resulting plot shows that most modelled erosion occurs below 2000 m,
+with the potential erosion volumes between 1000 and 2000 m elevation.
 
@@
 
-TODO: I think this will best fit in "age of the glacial landscape".
+Our calculations indeed produce locally high cumulative erosion potential in
+the aforementioned crystalline massifs. More accurate modelling of small-scale
+glaciers would be needed to answer your question. Steep topography, though, is
+likely to be part of the explanation. A more substantial discussion of the high
+erosion potential values in these areas and the model limitations that hinder
+interpretation was added in the "age of the glacial landscape" subsection.
+The new passage reads as follow:
+
+    The validity of the model results at high elevation is discussable.
+    Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
+    \"Otztal and Tauern Massifs locally exhibit a strikingly high erosion
+    potential. However, the computation of glacier flow velocities on such
+    steep surfaces is strongly limited by the model horizontal resolution of
+    1\,km, the shallow-ice glacier flow physics (Imhof et al., 2019), and
+    PISM's current mass-conservation heuristics (Imhof, 2021). Besides,
+    bergschrund (rimaye) processes likely to dominate interglacial cirque
+    erosion at such altitudes (Sanders et al., 2012) are not captured by the
+    velocity-based glacier erosion power-laws.
 
@@
 
-We agree the references are not directly relevant. The citation was reduced to
-the latest reference by Penck (1905) and moved within the sentence to clarify
-that it is specific to the Alps.
+We agree that the references are not directly relevant. The citation was
+reduced to the latest reference by Penck (1905) and moved within the sentence
+to clarify that it is specific to the Alps.
 
@@
 Agreed. "Characteristic landform preservation" was replaced with "preserved
-periglacial blockfields".
+periglacial blockfields topped by erratic boulders".
 
@@
 
-Thank you for pointing out this study. The lack of cirque erosion processes was
-acknowledged and the reference added in this sentence, as well as in the
-relevant part of the discussion on the age of the glacial landscape.
+Thank you for pointing out this study. This particular sentenced was reworked
+to acknowledged not only the limiting resolution but also the lack of relevant
+erosion erosion processes. The reference was additionally incorporated in the
+relevant part of the discussion (see general comments above).
 
@@
 
-Corrected.
+Thanks. Corrected.
 
@@
 
-TODO: "what you mean by realistic"
+TODO: Another tricky point. Here we need to discuss what is realistic. Probably
+no single law gives realistic results indeed.
 
@@
 
-TODO: discuss uncertainty with sliding.
+A paragraph was added under "choice of erosion law" to discuss the
+uncertainties on glacier sliding physics. Also including suggestions from
+reviewer #2, the new paragraph reads as follow:
+
+    The modelled pattern of erosion potential depends on PISM glacier physics
+    and sliding model parameters. In particular, the pseudo-plastic sliding law
+    exponent (q=0.25 in Seguinot et al., 2018) controls the sharpness of the
+    transition between adherent and decoupled basal conditions. Recent ensemble
+    validation of Antarctic Ice Sheet glacial-cycle simulations against
+    geological and present-day observations (Albrecht et al., 2020a, b) support
+    a higher value of q=0.75, and thus a sliding law closer to linear, which
+    would perhaps result in a smoother distribution of sliding velocity and
+    erosion potential. Lateral stress gradients missing from the shallow-shelf
+    approximation stress balance could also contribute to moderate sliding
+    velocity in narrow troughs (Herman et al., 2011; Egholm et al, 2012a, b;
+    Pedersen et al., 2014).
 
@@
 
-The following sentence was addedin this paragraph. "Nevertheless, it may
-explain why some areas covered during glacial maxima only, such as some valleys
-on the southern side of the Alps, appear to have experienced only little
-glacial modification of their topography."
+Thank you for pointing this out. The following sentence was added in this
+paragraph:
+
+    Nevertheless, it may explain why some areas covered during glacial maxima
+    only, such as some valleys on the southern side of the Alps, appear to have
+    experienced only little glacial modification of their topography.
 
@@
 This sentence was only meant to recall the aforementioned results. We rephrased
-to "the erosion patterns shift spatially through the glacial cycle".
+to "the potential erosion patterns experience spatial shifts through the
+glacial cycle".
 
@@
 
-TODO: discuss shortly the discrepancy in "climate sensitivity".
-TODO: I don't understand the second part of the comment.
+TODO: I still need to add a sentence or two discussing the discrepancy under
+"climate sensitivity". I don't understand the point about fluvial topography.
 

### 2021-05-24 articles/erosion/alps-erosion.tex
Polish response #1, some points still missing.

@@
     transition to cold-based ice is, and why some glaciated regions show
-    preserved periglacial blockfields and others, including the Alps, do not
-    \citep{Wirsig.etal.2016, Seguinot.etal.2018}.
+    preserved periglacial blockfields topped by erratic boulders and others,
+    including the Alps, do not \citep{Wirsig.etal.2016, Seguinot.etal.2018}.
 
@@
     and further upstream where the valley slopes increase. While mountain
-    cirques erosion processes may not be captured by the model's physics
-    \citep[cf.][]{Sanders.etal.2012} and horizontal resolution, high
+    cirque glaciers and relevant erosion processes may not be captured by the
+    model physics and horizontal resolution, high
     cumulative erosion potential also occurs near the valley heads.
@@
     velocity-based glacier erosion power-laws. Nevertheless,
-    the lack of rapid erosion at high elevations for much
-    of the modelled glacial cycle implies that the highest mountain
+    the lack of rapid erosion at high elevations for much of the modelled
+    glacial cycle (Fig.~\ref{fig:hypsogram}) implies that the highest mountain
     cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would have formed

### 2021-05-24 articles/erosion/response-referee1.md
Improve discussion of realistic erosion magnitude.

@@
 
-TODO: Another tricky point. Here we need to discuss what is realistic. Probably
-no single law gives realistic results indeed.
+We are not sure what the cumulative erosion potential of the last glacial cycle
+should be. This sentence was about the amplitude of the cumulative erosion
+potential, but we were probably a bit too fast deeming one erosion law "more
+realistic" than the others. Based on the estimates of total Pleistocene glacial
+relief from the suggested references, we now argue that no tested erosion law
+give results in the expectable range of last glacial cycle erosion.
 
-There is no mismatch between panels (a-d) and (e-h), but to improve readability
-we have reduced the number of colour levels on the maps, and changed the
-positions of ticks on the transects.
+    With a total Pleistocene glacial relief on the order of a kilometre
+    (Preusser et al., 2011; Valla et al., 2011), a cumulative glacial erosion
+    for the last glacial cycle in the order of 10-100 m can be expected.
+    However, none of the tested erosion power-laws fall within this range.
+    Instead, the erosion law calibrated on tidewater glaciers (Koppes et al.,
+    2015) yields cumulative erosion in the Rhine Valley in the orders of
+    metres, while the three erosion laws based on terrestrial glaciers
+    (Humphrey and Raymond, 1994; Herman et al., 2015; Cook et al., 2020},
+    result in kilometre-scale integrated erosion potential. During the Last
+    Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers
+    were closer in size, slope (an important parameter as we argue in the next
+    section), and climatic context to the present-day glaciers of Patagonia and
+    the Antarctic Peninsula (Koppes et al., 2015) than to Franz-Joseph Glacier
+    (Herman et al., 2015) and many of the glaciers included in the global
+    compilation by Cook et al., (2020). This may help to explain why the
+    reality appears to fall in-between the tested erosion laws.
+
+Regarding Fig. 7, there is no mismatch between panels (a-d) and (e-h), but to
+improve readability we have reduced the number of colour levels on the maps,
+and changed the positions of ticks on the transects.
 

### 2021-05-24 articles/erosion/alps-erosion.tex
Improve discussion of realistic erosion magnitude.

@@
 
-    It may seem surprising that the erosion law calibrated on Patagonian and
-    Antarctic Peninsula tidewater glaciers \citep{Koppes.etal.2015} yields a
-    more realistic glacial cycle erosion potential in the Alps than those based
-    on terrestrial
+    With a total Pleistocene glacial relief on the order of a kilometre
+    \citep{Preusser.etal.2011, Valla.etal.2011}, a cumulative glacial erosion
+    for the last glacial cycle in the order of 10--100\,m can be expected.
+    However, none of the tested erosion power-laws fall within this range.
+    Instead, the erosion law calibrated on tidewater glaciers
+    \citep{Koppes.etal.2015} yields cumulative erosion in the Rhine Valley in
+    the orders of metres, while the three erosion laws based on terrestrial
     glaciers \citep{Humphrey.Raymond.1994, Herman.etal.2015, Cook.etal.2020},
-    themselves resulting in kilometre-scale integrated erosion even
-    higher than the Alpine relief itself. However, during the Last Glacial
-    Maximum and much of the last glacial cycle, Alpine paleoglaciers were
-    closer in size, slope, and climatic context to the present-day glaciers of
+    result in kilometre-scale integrated erosion potential. During the Last
+    Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers
+    were closer in size, slope (an important parameter as we argue in the next
+    section), and climatic context to the present-day glaciers of
     Patagonia and the Antarctic Peninsula \citep{Koppes.etal.2015} than to
@@
     included in the global compilation by \citet{Cook.etal.2020}.
-    \todo{rework this paragraph. The realistic laws lies in-between.}
+    This may help to explain why the reality appears to fall in-between the
+    tested erosion laws.
 

### 2021-05-25 articles/erosion/response-referee1.md
Add discussion of eastward ice cover bias.

@@
 
-The LGM outline was added on Fig. 2.
+The field-based LGM outline was added on Fig. 2. Computing a percent overlap is
+not straightforward. As for Fig. 4 in the 2018 paper, we use a manually
+modified version of the Ehlers et al. (2011) data. The original data presents
+numerous, expansive nunataks, many of which are found in accumulation area and
+thus directly incompatible with the mass-balance scheme used in the 2018 paper.
+However, the general eastward bias was discussed in two new sentences:
 
-TODO: I still need to add a sentence or two discussing the discrepancy under
-"climate sensitivity". I don't understand the point about fluvial topography.
+    It should be noted, however, that all runs presented here show a systematic
+    bias with excessive glacier cover in the Eastern Alps and a diminished
+    glacier extent in the Western Alps (Fig. 2a; further discussed in Seguinot
+    et al., 2018). Thus the modelled patterns of erosion potential certainly
+    includes a similar bias.
 

### 2021-05-25 articles/erosion/alps-erosion.tex
Add discussion of eastward ice cover bias.

@@
     including our default high-resolution run
-    (Fig.~\ref{fig:cumulative}), yield a smaller erosion potential. The spatial
-    pattern of modelled time-integrated erosion is nevertheless generally
-    consistent across the paleoclimate scenarios tested,
+    (Fig.~\ref{fig:cumulative}), yield a smaller erosion potential.
+
+    The spatial pattern of modelled time-integrated erosion is nevertheless
+    generally consistent across the paleoclimate scenarios tested,
     despite variations in the glaciation maximum extent
     (\citealp[Fig.~3 in][]{Seguinot.etal.2018}; Fig.~\ref{fig:sensitivity}).
-    Furthermore, these changes are minor
-    considering the large uncertainties affecting the choice of erosion law.
+    It should be noted, however, that all runs presented here show a systematic
+    bias with excessive glacier cover in the Eastern Alps and a diminished
+    glacier extent in the Western Alps (Fig.~\ref{fig:cumulative}a; further
+    discussed in \citealp{Seguinot.etal.2018}). Thus the modelled patterns of
+    erosion potential certainly includes a similar bias. However, uncertainties
+    due to paleoclimate appear minor in regard to the much larger uncertainties
+    affecting the choice of erosion law.
 

### 2021-05-25 articles/erosion/response-referee1.md
Add reference to Sternai et al. (2013).

@@
 
-TODO: I am not sure what to do about this long list of papers. Some of them are
-already cited, some just added in.
+In addition to the aforementioned changes, the Sternai et al. (2013) reference
+was added in the discussion of the vertical distribution of erosion with this
+new sentence:
+
+    Over longer timescales, though, the vertical distribution of erosion rates
+    also depends on the erosional modification of topography (Sternai et
+    al., 2013).

### 2021-05-25 articles/erosion/response-referee1.md
Address reviewer #2 general comments, 1 left todo.

@@
     Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
-    \"Otztal and Tauern Massifs locally exhibit a strikingly high erosion
+    Ötztal and Tauern Massifs locally exhibit a strikingly high erosion
     potential. However, the computation of glacier flow velocities on such
@@
     Over longer timescales, though, the vertical distribution of erosion rates
-    also depends on the erosional modification of topography (Sternai et
-    al., 2013).
+    also depends on the erosional modification of topography (Sternai et al.,
+    2013).

### 2021-05-25 articles/erosion/response-referee2.md
Add scale bar on all erosion maps.
Add technical corrections from reviewer #2.
Address reviewer #2 general comments, 1 left todo.

@@
 
-Thank you very much for your detailed review of our manuscript.
+Thank you very much for your and your colleague's time in preparing this
+review of our manuscript.
 
@@
 > which give very broad maxima for s.l.e. 10-30 cm (Fig. 7 k & l).
+>
+>   [\*That stage can be approximated by pausing the first erosion
+>    rates video around 99 ka and 13 ka. The three videos are strongly
+>    recommended!]
 
-Thank you for pointing this out. We are not confident that this result is
-significant. The situation you describe occurs, for instance, during much of
-MIS~5 (Fig. 2) and can be visualized in the animations. During such periods,
-glaciers are steep and only a few grid cells in width. Their velocities and
-thus computed erosion rates may be overestimated by the limiting horizontal
-resolution and the shallow-ice physics (Imhof et al., 2019).
+Thank you for pointing this out. However, we are not confident that this result
+is significant. The situation you describe occurs, for instance, during much of
+MIS~5 (Fig. 2) and can be visualized in the animations (as you noted). During
+such periods, glaciers are steep and only a few grid cells in width.
 
@@
 laws), both yield a decrease of erosion volume, hence the local peak in erosion
-volume. We now emphasize this uncertainty in several of our plots by using
-hatches over areas of limited glacier cover, documented in the figure captions.
+volume. However, we would rather not include this as a result as the modelled
+glacier velocities for such situations are unreliable due to the limiting
+horizontal resolution and the shallow-ice physics. This latter fact was
+highlighted in a new passage in the section on the "age of the glacial
+landscape". Also including suggestions from reviewer #1, the new passage
+reads as follow:
+
+    The validity of the model results at high elevation is discussable.
+    Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
+    Ötztal and Tauern Massifs locally exhibit a strikingly high erosion
+    potential. However, the computation of glacier flow velocities on such
+    steep surfaces is strongly limited by the model horizontal resolution of
+    1\,km, the shallow-ice glacier flow physics (Imhof et al., 2019), and
+    PISM's current mass-conservation heuristics (Imhof, 2021). Besides,
+    bergschrund (rimaye) processes likely to dominate interglacial cirque
+    erosion at such altitudes (Sanders et al., 2012) are not captured by the
+    velocity-based glacier erosion power-laws.
+
+TODO: also emphasize this uncertainty in several of our plots by using
+hatches over areas of limited glacier cover, documented in the figure captions?
 
@@
 
-Figure 4 was reworked to include a map of Rhine glacier modelled erosion rates
-at 36 ka, corresponding to the last marjor advance phase before the LGM and with
-somewhat similar extent to the deglacial state of 16 ka. This also corresponds
-to the topographic profiles shown on Fig. 8.
+Thank you. Figure 4 was reworked to include a map of Rhine glacier modelled
+erosion rates at 36 ka, corresponding to the last major advance phase before
+the LGM and with a similar extent to the deglacial stage of 16 ka. This also
+corresponds to the topographic profiles shown on Fig. 8 and thus seems more
+appropriate in hindsight.
+
+TODO: also include complete maps as a supplement?
 
@@
 
-TODO: discuss lithology, go through entire text?
+The above limitateions were made explicit with the following added sentences in
+the methods subsection on the "erosion law":
+
+    Instead , we we assume that eroded material is instantly transported out of
+    the system, thus neglegting its role in shielding the bedrock from glacier
+    erosion in zones of temporary storage (Preusser et al., 2010). Neither do
+    we account for differences in erosion effectiveness on different
+    lithologies or erosion from subglacial and interglacial hydrologic
+    processes. For the aforementioned reasons, we refer to the above computed
+    rates, ė, as “potential erosion rates”.
+
+Second and as implied, the entire text and the figure labels were reworked to
+replace “erosion rates” with “potential erosion rates” nearly everywhere. To
+avoid clutter a few sentences still mention “modelled erosion rates” instead of
+“potential erosion rates”.
+
+TODO: rework figure labels
 
@@
 
-The LGM outline was added on Fig. 2.
+The field-based LGM outline was added on Fig. 2 (Alps-wide map of cumulative
+erosion potential). In addition, the following sentences were added in the
+"climate sensitivity" section of the discussion mentioning the general
+eastwards bias of our results with regard to field evidence.
 
-TODO: discuss shortly the discrepancy in "climate sensitivity".
+    It should be noted, however, that all runs presented here show a systematic
+    bias with excessive glacier cover in the Eastern Alps and a diminished
+    glacier extent in the Western Alps (Fig. 2a; further discussed in Seguinot
+    et al., 2018). Thus the modelled patterns of erosion potential certainly
+    includes a similar bias.
 
@@
 
->   [\*That stage can be approximated by pausing the first erosion
->    rates video around 99 ka and 13 ka. The three videos are strongly
->    recommended!]
-
 > The text is short and pithy, so there is space for some extension. There are
@@
 > detail I suggest the following improvements:
+
+The manuscript text has been extended in several places. Please see our
+responses to your specific comments below.
 
@@
 
+Scale bars were added on all map figures.
+
 > Line 24  ‘are yet’
@@
 > 190 delete both commas
+
+Thank you for spotting all the above errors! These were corrected.
 

### 2021-05-25 articles/erosion/alps-erosion.tex
Add reference to Sternai et al. (2013).
Address reviewer #2 general comments, 1 left todo.

@@
     load, and a~positive degree-day (PDD) surface mass balance model. The model
-    is initialized with assumed present-day ice thickness and bedrock topography
-    from SRTM \citep{Jarvis.etal.2008},
+    is initialized with assumed present-day ice thickness and 1-km resolution
+    bedrock topography from SRTM \citep{Jarvis.etal.2008},
     ice and bedrock temperature at 120\,ka (thousand years before the present),
@@
     \citep[EPICA;][] {Jouzel.etal.2007}, and time-dependent paleo-precipitation
-    reductions \citep{Huybrechts.2002}. Lower-resolution runs based on
+    reductions \citep{Huybrechts.2002}. Coarser, 2-km resolution runs based on
     alternative paleoclimate scenarios \citep{Seguinot.etal.2018}, forced by
@@
     that mountain range during cold periods as well, failing to offset uplift
-    rates \citep{Harper.Humphrey.2003}.
+    rates \citep{Harper.Humphrey.2003}. Over longer timescales, though, the
+    vertical distribution of erosion rates also depends on the erosional
+    modification of topography \citep{Sernai.etal.2013}.
 

### 2021-05-25 articles/erosion/response-referee2.md
Address reviewer #2 spec comments, todos left.

@@
 
-TODO: The input SRTM topography was mentioned in [...] and in the caption of
-Fig. 2.
+The model horizontal resolutions of 1 and (for climate sensitivity runs) 2 km
+is now mentioned in the methods. The input SRTM topography is also referenced
+in both the methods and the caption of Fig.~2 where it is displayed in the
+background. Following comments from reviewer #1, the discussion of model
+limitations regarding the representation of cirque glaciers has also been
+extended.
 
@@
 > how has lateral drag been included?
+
+PISM uses a combination of shallow-ice and shallow-shelf stress balances,
+neither of which includes lateral stresses. Glacier sliding velocities in
+troughs are goverened by (the sliding law and) longitudinal stresses, which is
+a limitation with respect to higher-order models. Following another comment
+from reviewer #1 an additional paragraph on basal sliding uncertainties was
+added, including the following sentence:
+
+    Lateral stress gradients missing from the shallow-shelf approximation
+    stress balance could also contribute to moderate sliding velocity in narrow
+    troughs (Herman et al., 2011; Egholm et al., 2012a, b; Pedersen et al.,
+    2014).
 
@@
 
+TODO: discuss Koppes without outliers.
+
 > In lines 206-209, the point about trimlines, relating them to
@@
 > necessarily just from recessional phases.
+
+Indeed. The sentence was corrected to include "advancing and retreating
+glaciers" instead of "retreating glaciers".
 
@@
 
-We applied different colours for periods of increasing and decreasing ice
-volume. This actually strikingly highlights the two different regimes. Thank
-you for the suggestion. The label was changed to 'potential annual erosion
-volume'.
-
-TODO: not so slight decrease.
+Figure 3 was reworked to use different colours for periods of increasing and
+decreasing ice volume. This actually strikingly highlights the two different
+regimes. Thank you for the suggestion. The label was changed to 'potential
+annual erosion volume'. In the text, 'slight' was replaced with a 'general'
+tendency for slower erosion during periods of extensive glaciation.
 
@@
 In addition to the aforementioned changes to Fig. 4, a new panel was added for
-the model final state at 0 ka, revealing the bedrock topography.
+the model final state at 0 ka, revealing the bedrock topography. So Fig. 4 now
+has one panel during advance (36 ka), one during a maximum stage (24 ka), one
+during retreat (16 ka), and one nearly ice-free (0 ka).
 
@@
 > (vertical-sided) trough example, over-used by textbooks.
+
+TODO: think about replacing the photos.
 
@@
 
+The figure now includes a histogram of model domain elevations, a histogram of
+glacier-covered elevations, and a plot of cumulative erosion potential volume
+per elevation bands of 100 m. The latter shows that, as you suspected, high
+altitude rapid erosion is largely offset by the limited number of grid cells
+it concerns. Instead, the bulk of the erosion potential occurs below 2000 m.
+
+TODO: hatch out regions with limited cell count?
+
 > In Fig. 7, the quantitative contrasts between the ‘Koppes’ law (e) and the
@@
 
+Discussion on the magnitude of modelled erosion rates remains somewhat
+speculative indeed due to the several sources of uncertainties, but the
+relevant discussion text was extended and reworked:
+
+    With a total Pleistocene glacial relief on the order of a kilometre
+    (Preusser et al., 2011; Valla et al., 2011), a cumulative glacial erosion
+    for the last glacial cycle in the order of 10-100 m can be expected.
+    However, none of the tested erosion power-laws fall within this range.
+    Instead, the erosion law calibrated on tidewater glaciers (Koppes et al.,
+    2015) yields cumulative erosion in the Rhine Valley in the orders of
+    metres, while the three erosion laws based on terrestrial glaciers
+    (Humphrey and Raymond, 1994; Herman et al., 2015; Cook et al., 2020},
+    result in kilometre-scale integrated erosion potential. During the Last
+    Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers
+    were closer in size, slope (an important parameter as we argue in the next
+    section), and climatic context to the present-day glaciers of Patagonia and
+    the Antarctic Peninsula (Koppes et al., 2015) than to Franz-Joseph Glacier
+    (Herman et al., 2015) and many of the glaciers included in the global
+    compilation by Cook et al., (2020). This may help to explain why the
+    reality appears to fall in-between the tested erosion laws.
+
 > Fig. 8a shows a flat dome at 3000 m for about the first 20 km of the Rhine.
@@
 > resolution?
+
+TODO: I think this is correct. Develop on basal shear stress patterns.
 
@@
 >   positive, which seems strange.
+
+TODO: tricky, I need to have a better look.
 

### 2021-05-26 figures/alpero_hypsogram.py
Hatch regions with few glacier cells on hypsogram.

@@ -6,6 +6,7 @@
 """Plot Alps erosion hypsogram."""
 
 import numpy as np
+import matplotlib.pyplot as plt
 import absplots as apl
 import xarray as xr
 import util
@@ -40,12 +41,20 @@ def main():
     with xr.open_mfdataset(
             '../data/processed/alpero.1km.epic.pp.agg.nc') as ds:
 
+        # maybe the only way to set hatch color
+        plt.rc('hatch', color='0.25')
+
         # plot hypsogram
         (np.log10(ds.kop2015_hyps)+3).plot.imshow(
             ax=ax, alpha=0.75, cmap='YlOrBr', vmin=-9, vmax=0, x='age',
             cbar_ax=cax, cbar_kwargs=dict(
                 label='geometric mean\n'+r'erosion rate ($mm\,a^{-1}$)',
                 ticks=range(-9, 1, 3)))
+        ds.glacier_hyps.plot.contourf(
+            ax=ax, add_colorbar=False, colors='none', extend='neither',
+            hatches=['//////'], levels=[0.5e6, 99.5e6], x='age')
+        ds.glacier_hyps.plot.contour(
+            ax=ax, colors='0.25', levels=[99.5e6], x='age')
         cax.yaxis.set_major_formatter(r'$10^{{{x:d}}}$')
         # this should work in matplotlib 3.3.2 (PR #18458)
         # (ds.kop2015_hyps*1e3).plot.imshow(
@@ -74,7 +83,6 @@ def main():
         hax = hax.twiny()
         (ds.kop2015_cumu.groupby_bins(boot, bins).sum()/1e3).plot.step(
             ax=hax, y='topg_bins', color='C11')
-        # hax.barh(bins[:-1], sums, height=100, align='edge', color='C11')
 
         # set twin axes properties
         hax.tick_params(axis='x', direction='in', pad=-15)

### 2021-05-26 articles/erosion/response-referee1.md
Update hypsogram caption and responses.

@@
 
-Yes. This primarily concerns the southern slope of the Alps where topography
-drops abruptedly from the highest peaks to deep valleys. It concerns only a few
-grid cells for a given time step and is certainly not a realistic model output.
-
-TODO: test cutting the hypsogram below a minimum number of grid cells.
+This is correct. It concerns only a small number of grid cells, primarily
+located on the southern slope of the Alps where topography drops abruptly from
+the highest peaks to deep valleys. This information was added to the plot by
+hatching the regions with fewer than a hundred ice-covered grid cells. This
+highlights that low-elevation values are not always representative, as can also
+be confirmed in the new subplot showing the distribution of cumulative erosion
+potential over 100 m elevation bins. The distribution of glacier cover per
+elevation band over time (from which the hundred-cell contour was extracted)
+will be included in a new version of the companion dataset.
 

### 2021-05-26 articles/erosion/response-referee2.md
Update hypsogram caption and responses.
Discuss basal drag fluctuations in upper Rhine.

@@
 
-TODO: discuss Koppes without outliers.
+We clarified in the methods that we use  Koppes et al. (2015) erosion law
+deriving from their full dataset. In addition, this sentence was added in the
+discussion of the "choice of erosion law":
+
+    On the other extreme, an even more non-linear erosion law, not tested here,
+    derived from tidewater glaciers but excluding two outliers (ė = 5.3 × 10−9
+    u_b^2.62, Koppes et al., 2015) would result in an even more localized
+    pattern of erosion potential.
 
@@
 it concerns. Instead, the bulk of the erosion potential occurs below 2000 m.
-
-TODO: hatch out regions with limited cell count?
+To highlight periods when low-elevation is relevant (and address the comment
+from reviewer #1), we also hatched elevation bands that contain fewer than a
+hundred ice-covered grid cells for a given time.
 
@@
 
-TODO: I think this is correct. Develop on basal shear stress patterns.
+This is correct. The "basal drag" quantity plotted on Fig. 8b is the magnitude
+of a two-component vector: the basal shear stress. The basal shear stress
+itself strongly depends on surface altitude gradients, but this includes both
+along-flow slopes visible on the topographic profile of Fig. 8a, and the
+invisible across-flow slopes. While the upper part of the topographic profile
+appears smooth, there are indeed steep (bedrock and surface) slopes on either
+side of the profile in this region. Because the upper part of the valley is
+narrow, the glacier center-line fluctuating over time, and the plot
+interpolated between 1-km grid cells, these lateral components enter the
+computation of the basal shear stress magnitude in the higher reaches.
+
+Second, the activation of sliding also increases basal drag, which explains
+that basal drag also fluctuates together with the yield stress. This is
+particularly affecting the lower part of the profiles. Finally, normalization
+by overburden pressure amplifies all stress variations where ice thickness is
+small. But due to changes in glacier thickness, it would not be possible to
+compare different profiles without such normalization.
+
+While we would rather not include such details in the paper, we tried our best
+to summarise the effect with this additional sentence in the figure caption:
+
+    Some of the observed basal drag fluctuations in the upper part of the
+    transect result from steep slopes on either side of the narrow valley.
 

### 2021-05-26 articles/erosion/alps-erosion.tex
Update hypsogram caption and responses.
Discuss basal drag fluctuations in upper Rhine.

@@
           geometric mean of (non-zero) modelled rates in 10-m elevation bands
-          across the entire model domain and its time evolution.
+          across the entire model domain and its time evolution. Hatches
+          indicate elevation bands with fewer than a hundred grid cells
+          (100\,$km^2$).
         \textbf{(b)} Distribution of model domain bedrock topography (grey
@@
     between modelled ice volume and erosion (Fig.~\ref{fig:powerlaws}k and~l).
-    \todo{discuss Koppes without outliers}
+    On the other extreme, an even more non-linear erosion law, not tested here,
+    derived from tidewater glaciers but excluding two outliers
+    \citep[${\dot{e}=5.3\times10^{-9}u_\mathrm{b}^{2.62}}$,][]{Koppes.etal.2015}
+    would result in an even more localized pattern of erosion potential.
 
@@
         Basal drag values greater than yield stress result in glacier sliding
-        \citep[cf.][for model physics]{Seguinot.2014, Seguinot.etal.2016}.}
+        \citep[cf.][for model physics]{Seguinot.2014, Seguinot.etal.2016}.
+        Some of the observed basal drag fluctuations in the upper part of the
+        transect result from steep slopes on either side of the narrow valley.}
       \label{fig:basaldrag}

### 2021-05-26 articles/erosion/response-referee2.md
Address reviewer comment on zero erosion values.

--- 2021-05-26 articles/erosion/response-referee2.md
@@
 
-TODO: tricky, I need to have a better look.
+Well, it seems that you have spend a lot of time looking at our figures and
+videos so let us thank you again for that. The caption of Fig. 5 mentioned "the
+geometric mean of (non-zero) modelled erosion rates". We found that the
+"non-zero" has no reason to be and it was removed. Sorry about the confusion
+and please read on for the long explanation.
+
+First, there are no zero values in the animation. As in the paper's Fig. 5, the
+animation shows geometric means (i.e. arithmetic means in log-domain), and
+geometric means can't contain zero values or they would result in a
+divide-by-zero error (or a log-of-zero error). Both animation panels have a
+log-scaled x-axis, hence they cannot contain zero values.
+
+On the other hand, both Fig. 5 and the animation depict very low values for
+erosion rates (especially around 24 ka). This is due to the glacier physics
+embedded in PISM. In particular, the pseudo-plastic sliding law implies that
+the glacier is sliding everywhere, but that sliding is infinitely small where
+the basal drag is much smaller than the yield stress, including for instance
+frozen-bed areas. These tiny sliding velocities yield even tinier erosion rates
+(due to the power 2.34 in the erosion law). Such values are not really relevant
+and they are off-the-colour-scheme on the Fig. 5 and off-the-chart in the
+animation (respectively lower than 10^-9 and 10^-10 mm a^-1).
+
+The only regions where PISM produces truly zero sliding are the regions of zero
+surface slopes or zero ice thickness, where the SSA can't be solved. In
+practice, this only occurs in the latter case: zero ice thickness. This is the
+kind of values we have previously filtered out. Your comment led us to re-check
+our calculations. We found that there is no zero-sliding values in the
+ice-covered area, and that an ice mask is sufficient to filter out undesirable
+values and compute the geometric means. Hence "non-zero" was removed from the
+caption.
+
+To be sure, the non-zero filter has been removed from the script generating
+companion data (which is used in Fig. 5). This change can be seen here:
+
+TODO: link Github commit.
 

### 2021-05-26 articles/erosion/alps-erosion.tex
Discuss Koppes et al. (2015) law without outliers.
Remove zero-sliding filter in postproc, caption.

@@
         \textbf{(a)} Potential erosion rate ``hypsogram'', showing the
-          geometric mean of (non-zero) modelled rates in 10-m elevation bands
+          geometric mean of modelled rates in 10-m elevation bands
           across the entire model domain and its time evolution. Hatches

### 2021-05-26 data/postprocess_alpero_agg.py

@@ -135,10 +135,9 @@ def postprocess_extra(run_path):
         pp[law+'_rate'] = (dx*dy*ex[law].sum(axis=(1, 2))).assign_attrs(
             long_name=ex[law].ref+' domain total volumic erosion rate',
             units='m3 year-1')
-        pp[law+'_hyps'] = np.exp(
-            np.log(ex[law].where(ex[law] > 0)).groupby_bins(
-                boot.topg, bins=range(0, 4501, 100)).mean(
-                    dim='stacked_y_x')).assign_attrs(
+        pp[law+'_hyps'] = np.exp(np.log(ex[law]).groupby_bins(
+            boot.topg, bins=range(0, 4501, 100)).mean(
+                dim='stacked_y_x')).assign_attrs(
             long_name=ex[law].ref+' erosion rate geometric mean in band',
             units='m year-1')
         pp[law+'_rhin'] = ex[law].where(ex.icy).interp(

### 2021-05-26 data/postprocess_alpero_agg.py
So 10m hypsogram finally works but I prefer 100m.
Using 10-m bands includes additional details (e.g. one can distinguish the
slightly higher-elevation north slope margin) but I think this is unnecessary.
The new 100-cell (or 10-cell for 10-m bands) contour and hatch pattern is
exceedingly detailed and that renders the figure very heavy on 10-m bands.
There are also unneeded striping instead of a clean contour around 3500 m
elevation. So I stick to 100-m bands on both panels (and in the movie).

Finally done with this! Time for some celebratory ouzo.

@@ -50,7 +50,7 @@ def postprocess_extra(run_path):
     boot = hyoga.open.dataset(
         '~/pism/input/boot/'+'alps.srtm.hus12.gou11simi.{}.nc'.format(res))
     print("* reading multi-file extra dataset...")
-    ex = hyoga.open.mfdataset(run_path+'/ex.???????.nc')
+    ex = hyoga.open.mfdataset(run_path+'/ex.???????.nc', parallel=True)
 
     # init postprocessed dataset with global attributes
     pp = xr.Dataset(attrs=ex.attrs, coords=dict(lon=ex.lon, lat=ex.lat))
@@ -96,10 +96,7 @@ def postprocess_extra(run_path):
     # trigger computation to avoid memory errors (18m, 70%, 8GiB)
     # NOTE: time and total mem for single-thread scheduler on altair, with
     # nothing else running. Trigger compute() more often to further reduce mem
-    # consumption (maybe at the cost of repeating some operations). I think
-    # the main trick is to separate time and space aggregations. Otherwise, it
-    # seems, there will be chunk-splitting, very long scheduling overhead and
-    # memory errors.
+    # consumption (maybe at the cost of repeating some operations).
     print("* computing time-integrated variables...")
     with dask.diagnostics.ProgressBar():
         pp = pp.compute()
@@ -129,6 +126,11 @@ def postprocess_extra(run_path):
         pp = pp.compute()
 
     # compute erosion space-aggregated variables
+    # there's still a divide warning here, maybe check what that is
+    # NOTE: for the 1-km-run hypsogram on 10-m bands, dask-2021.3.0 splits the
+    # chunks even when asked not to. The progress bar only appears after two
+    # hours (I nearly gave up), but the actual computations takes only 15 min!
+    # In total the script took 194 min on altair.
     x, y = cpf.read_shp_coords(
         '../data/native/profile_rhine.shp', interval=1000)
     for law in ['coo2020', 'her2015', 'hum1994', 'kop2015']:
@@ -198,7 +200,7 @@ def main():
     # client = dask.distributed.Client()
     # print(client.dashboard_link)
 
-    # split large chunks (and avoid warnings)
+    # ask dask to not split large chunks (but it still does)
     dask.config.set(**{'array.slicing.split_large_chunks': False})
 
     # postprocess selected runs

### 2021-05-26 figures/alpero_hypsogram.py
Fix hypsogram subfigure labels.

@@ -27,7 +27,8 @@ def main():
 
     # add subfigure labels
     util.fig.add_subfig_label('(a)', ax=ax)
-    util.fig.add_subfig_label('(b)', ax=tsax)
+    util.fig.add_subfig_label('(b)', ax=hax)
+    util.fig.add_subfig_label('(c)', ax=tsax)
 
     # load boot topo
     with xr.open_dataset('../data/processed/alpcyc.1km.in.nc') as ds:

### 2021-05-26 articles/erosion/response-referee2.md
Link last Github commit in the response.
Update hypsogram caption and responses.

@@
     On the other extreme, an even more non-linear erosion law, not tested here,
-    derived from tidewater glaciers but excluding two outliers (ė = 5.3 × 10−9
+    derived from tidewater glaciers but excluding two outliers (ė = 5.3×10^-9
     u_b^2.62, Koppes et al., 2015) would result in an even more localized
@@
 it concerns. Instead, the bulk of the erosion potential occurs below 2000 m.
+
 To highlight periods when low-elevation is relevant (and address the comment
 from reviewer #1), we also hatched elevation bands that contain fewer than a
-hundred ice-covered grid cells for a given time.
+hundred ice-covered grid cells for a given time. We also shifted to 100-m
+elevation bands on the main panel. Wihle some detail is lost, this avoids
+much of the striping (especially on the newly added contour and hatched
+pattern), and we find the new plot more readable.
 
@@
 
-TODO: link Github commit.
+    https://github.com/juseg/alps/commit/7d14ec0
 

### 2021-05-26 articles/erosion/alps-erosion.tex
Update hypsogram caption and responses.

@@
         \textbf{(a)} Potential erosion rate ``hypsogram'', showing the
-          geometric mean of modelled rates in 10-m elevation bands
+          geometric mean of modelled rates in 100-m elevation bands
           across the entire model domain and its time evolution. Hatches

### 2021-05-27 articles/erosion/response-referee1.md
Try to improve introduction somewhat.

@@
 
-TODO: This remain a tricky points to me. Shoulw we remove parts of the
-introduction? Or add sentences tying our statements to upcoming results?
+Our introduction is meant to revolve around two discussion points: the variable
+imprint of glaciers on mountain topography, and the link between climate and
+glacier erosion (somewhat adressed in the conclusions). This is followed by a
+short methodological bibliography of process-based and empirical glacier
+erosion laws, before we announce the scope of our work. A few key statements
+were added to try and better tie these different sections.
 

### 2021-05-27 articles/erosion/alps-erosion.tex
Try to improve introduction somewhat.

@@
     higher variety of glacier erosional landforms, whose implications on
-    glacial history are yet to be understood.
-    The cosmogenic nuclide memory of bedrock erosion provides a more
+    glacial history are yet to be understood. In parallel to such landscape
+    diversity, the cosmogenic nuclide memory of bedrock erosion provides a more
     quantitative, but equally varied picture of glacier erosion effectiveness
@@
 
-    In the absence of other landforms, glacially eroded topography has
-    sometimes been used as a proxy for glacier extent, for instance in
-    geomorphological mapping \citep[e.g.,][]{Margold.etal.2011, Fu.etal.2012}.
+    Glacially eroded topography has sometimes been used as a proxy for mapping
+    paleoglacier extent \citep[e.g.,][]{Margold.etal.2011, Fu.etal.2012}.
     Cold-based glaciers, however, have been observed to preserve landforms as
@@
     \citep{Kleman.etal.2008, Kleman.etal.2010, Fabel.etal.2012, Fu.etal.2013}.
-    Nevertheless, it remains unclear how sharp the erosional
-    transition to cold-based ice is, and why some glaciated regions show
+    Nevertheless, it remains elusive whether glacial topography is a proxy for
+    for temperate ice-cover duration, and why some glaciated regions show
     preserved periglacial blockfields topped by erratic boulders and others,
@@
 
-    Glacier erosion also likely governs, at least in part,
+    On a larger scale, glacier erosion also likely governs, at least in part,
     the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through
@@
     erosion rates across different climates suggests that increased temperatures
-    in glaciated regions lead to higher erosion rates and that erosion rates
-    may have increased towards present \citep{Koppes.Montgomery.2009,
-    Koppes.etal.2015, Fernandez.etal.2016}. However, understanding the link
+    within glaciated regions lead to higher erosion rates
+    \citep{Koppes.Montgomery.2009, Koppes.etal.2015, Fernandez.etal.2016}.
+    However, understanding the link
     between climate and glacier erosion in the sedimentary record remains
@@
     non-linear relationship between atmospheric temperature and erosion
-    \citep[e.g.,][]{Anderson.etal.2012,Mariotti.etal.2021}.
-    \todo{shorten this paragraph.}
+    \citep[e.g.,][]{Anderson.etal.2012, Mariotti.etal.2021}.
+    Numerical glacier modelling may provide insights into such complexity.
 
@@
     patterns of modelled erosion potential and cumulative last glacial cycle
-    erosion
-    potential are analysed. We examine the glaciological conditions that cause
-    erosion and discuss the implications of these conditions on understanding
+    erosion potential are analysed. We examine the glaciological conditions
+    that cause erosion and discuss the implications of these conditions on
+    understanding the mark of ancient glaciers on Alpine topography and
     the relationship between climate and glacier erosion. Despite aggregated

### 2021-05-27 figures/alpero_transects.py
Coarsen transect plot to reduce figure size.

@@ -89,7 +89,8 @@ def main():
     with hyoga.open.dataset(
             '../data/processed/alpero.1km.epic.pp.agg.nc') as ds:
 
-        # plot erosion profile
+        # plot erosion profile (coarsen to reduce figure size)
+        ds = ds.coarsen(age=5).mean()  # 3.7 -> 1.0 MiB
         (ds.assign_coords(d=ds.d/1e3).kop2015_rhin*1e3).plot.contourf(
             ax=tsax, alpha=0.75, cmap='YlOrBr', levels=levels, x='age', y='d',
             cbar_ax=cax, cbar_kwargs=dict(
@@ -98,7 +99,6 @@ def main():
                 ticks=levels[::3]))  # (mpl issue #11937)
 
     # set axes properties
-    tsax.set_rasterization_zorder(2.5)
     tsax.set_xlim(120, 0)
     tsax.set_xlabel('age (ka)')
     tsax.set_ylabel('distance along flow (km)')

### 2021-05-27 figures/alpero_cumulative.py
Always use potential erosion in fig labels.

@@ -23,7 +23,7 @@ def main():
         # plot map data
         ds.kop2015_cumu.plot.contourf(
             ax=ax, alpha=0.75, cmap='YlOrBr', cbar_ax=cax,
-            cbar_kwargs=dict(label='erosion potential (m)', format='%g'),
+            cbar_kwargs=dict(label='cumulative erosion potential (m)', format='%g'),
             levels=[10**i for i in range(-2, 3)])
         ds.kop2015_cumu.notnull().plot.contour(
             ax=ax, colors='k', linewidths=0.5, levels=[0.5])

### 2021-05-27 figures/alpero_powerlaws.py

@@ -117,12 +117,11 @@ def main():
         cde.add_scale_bar(grid.flat[0], label='50 km', length=50e3, pad=10e3)
 
     # add y-labels on rightmost axes
-    tsgrid[0, -1].set_ylabel('erosion potential (m)')
     tsgrid[1, -1].set_ylabel(r'annual erosion volume ($m^3 a^{-1}$)')
 
     # add colorbar
     cbar = fig.colorbar(cset, cax=cax, format='%g')
-    cbar.set_label('erosion potential (m)', labelpad=0)
+    cbar.set_label('cumulative erosion potential (m)', labelpad=0, y=0)
 
     # save
     util.com.savefig(fig)

### 2021-05-27 figures/alpero_sensitivity.py

@@ -63,8 +63,8 @@ def main():
 
     # add colorbar
     cax = fig.add_axes_mm([57*2+19+1.5*4, 38/3-3, 57*2/3-1.5, 3])
-    fig.colorbar(cset, cax=cax, label='erosion potential (m)', format='%g',
-                 orientation='horizontal')
+    fig.colorbar(cset, cax=cax, label='cumulative erosion potential (m)',
+                 format='%g', orientation='horizontal')
 
     # save
     util.com.savefig(fig)

### 2021-05-27 figures/alpero_transects.py

@@ -94,7 +94,7 @@ def main():
         (ds.assign_coords(d=ds.d/1e3).kop2015_rhin*1e3).plot.contourf(
             ax=tsax, alpha=0.75, cmap='YlOrBr', levels=levels, x='age', y='d',
             cbar_ax=cax, cbar_kwargs=dict(
-                label=r'erosion rate ($mm\,a^{-1}$)',
+                label=r'potential erosion rate ($mm\,a^{-1}$)',
                 format=mpl.ticker.LogFormatterMathtext(),
                 ticks=levels[::3]))  # (mpl issue #11937)
 

### 2021-05-27 figures/util/ero.py

@@ -37,5 +37,5 @@ def plot_series(ax=None, run='1km.epic.pp'):
         ax = ax.twinx()
         ax.plot(ds.age, ds.kop2015_rate*1e-6, c='C11', alpha=0.5)
         ax.plot(ds.age, ds.rolling_mean*1e-6, c='C11')
-        ax.set_ylabel('annual erosion\n'+r'volume ($10^6\,m^3 a^{-1}$)', color='C11')
+        ax.set_ylabel('potential annual erosion\n'+r'volume ($10^6\,m^3 a^{-1}$)', color='C11')
         ax.set_ylim(-1, 7)

### 2021-05-27 articles/erosion/alps-erosion.tex
Add reference to Lai.Anders.2021 in review.

@@
     changes in glacier geometry alone (Figs.~\ref{fig:basaldrag}b).
+    Coincidentally, the link between climate and erosion is also the topic of
+    another independent study using PISM \citep{Lai.Anders.2021}.
     Observed feedbacks between
@@
     vertical distribution of erosion rates also depends on the erosional
-    modification of topography \citep{Sernai.etal.2013}.
+    modification of topography \citep{Sternai.etal.2013}.
 

### 2021-05-27 figures/alpero_landscape.py
Replace Lauterbrunnen photo with Bout du Monde.

@@ -22,9 +22,9 @@ def main():
 
     # flickr photo identifiers
     photos = [  # use _b for 1024, _c for 800 px
-        '50743485202_405601c2d8_b.jpg',
-        '50743484067_800c6cfba8_b.jpg',
-        '50742646803_2bd4961b6a_b.jpg']
+        '50743485202_405601c2d8_b.jpg',  # Lake Constance
+        '50950153856_255009394c_b.jpg',  # Bout du monde
+        '50742646803_2bd4961b6a_b.jpg']  # Chüebodensee
 
     # loop on axes
     for i, ax in enumerate(grid):

### 2021-05-27 articles/erosion/alps-erosion.tex
Move photos to discussion, update caption, text.

@@
 
-    \begin{figure*}
-      \centerline{\includegraphics{alpero_landscape}}
-      \caption{%
-        Alpine glacial erosion landscape diversity.
-        \textbf{(a)} Piedmont overdeepening of Lake Constance, ca.~10x50\,km.
-        \textbf{(b)} Glacial trough of Lauterbrunnental, ca.~1x10\,km.
-        \textbf{(c)} Mountain cirque of Ch\"uebodengletscher, ca.~1x1\,km.}
-      \label{fig:landscape}
-    \end{figure*}
-
     The glacial erosion landscape of the Alps has fascinated generations of
@@
     large-scale overdeepenings (e.g. Patagonia). But other regions, including
-    the Alps \citep[e.g.][Fig.~\ref{fig:landscape}]{Penck.1905}, present a
+    the Alps \citep[e.g.]{Penck.1905}, present a
     higher variety of glacier erosional landforms, whose implications on
@@
 
+    \begin{figure*}
+      \centerline{\includegraphics{alpero_landscape}}
+      \caption{%
+        Alpine glacial erosion landscape diversity.
+        \textbf{(a)} Piedmont overdeepening of Lake Constance, ca.~10x50\,km.
+        \textbf{(b)} Glacial trough at Bout du Monde, ca.~1x10\,km.
+        \textbf{(c)} The unusually deep mountain cirque revealed by the current
+          demise of Ch\"uebodengletscher, ca.~1x1\,km.}
+      \label{fig:landscape}
+    \end{figure*}
+
     In many places throughout the Alps, significant modelled cumulative erosion
@@
     For instance, low-elevation piedmont over-deepened basins such as Lake
-    Constance (Fig.~\ref{fig:landscape}), Lake Geneva and Lake Maggiore are
+    Constance (Fig.~\ref{fig:landscape}a), Lake Geneva and Lake Maggiore are
     here modelled to be the product of extensive Last Glacial Maximum-like
@@
     (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
-    intermediate-elevation glacial valleys of the Alps as Lauterbrunnental
-    (Fig.~\ref{fig:landscape}), Haslital (Upper Aare) or Bout du Monde (Upper
-    Giffre) would date from periods of intermediate glacier extension.
+    intermediate-elevation glacial valleys of the Alps as Lauterbrunnental,
+    Haslital (Upper Aare) or Bout du Monde (Upper Giffre,
+    Fig.~\ref{fig:landscape}b)
+    would date from periods of intermediate glacier extension.
 
@@
     glacial cycle (Fig.~\ref{fig:hypsogram}) implies that the highest mountain
-    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would have formed
+    cirques (e.g. Fig.~\ref{fig:landscape}c) would have formed
     over interglacial periods, when glaciers are confined to high terrain

### 2021-05-27 articles/erosion/response-referee2.md
Respond to reviewer #2 on photos etc.

@@
 
-TODO: rework figure labels
-
 > A further necessary geographical qualification concerns the misfit between
@@
 
-TODO: check that out.
+Indeed, this is quite a coincidence! A reference was added to the paper by Lai
+and Anders (2021).
 
@@
 
-TODO: think about replacing the photos.
+The figure was moved into the discussion of the "age of the glacial landscape"
+to become Fig. 8, and is not longer referred to in the intro. The photo of
+Lauterbrunnental was replaced by one of Bout du Monde in the Giffre Massif, and
+the caption for the cirque photo now reads: "the unusually deep mountain cirque
+revealed by the current demise of Chüebodengletscher" (I don't have a much
+better photo for a high-elevation cirque, and glacial lakes from retreating
+cirque glaciers are a common occurence in the Alps these days).
 

### 2021-05-27 articles/erosion/response-referee1.md
Read through the responses once again.

@@
 Thank you very much for your positive comments and for your time spent
-reviewing our manuscript.
+reviewing our manuscript. We try to address your comments one by one below, and
+highlight relevant changes made to the manuscript.
 
@@
 
-Thank you for the suggestion. Figure 5 was amended with (as Fig. 5b) the
+Thank you for the suggestion. Figure 5 was amended with the
 distribution of cumulative erosion potential over 100 m elevation bins.
@@
 the aforementioned crystalline massifs. More accurate modelling of small-scale
-glaciers would be needed to answer your question. Steep topography, though, is
+glaciers would be needed to provide an accurate answer to your question. Steep
+topography, though, is
 likely to be part of the explanation. A more substantial discussion of the high
@@
 Thank you for pointing out this study. This particular sentenced was reworked
-to acknowledged not only the limiting resolution but also the lack of relevant
+to acknowledge not only the limiting resolution but also the lack of relevant
 erosion erosion processes. The reference was additionally incorporated in the
@@
 
-We are not sure what the cumulative erosion potential of the last glacial cycle
-should be. This sentence was about the amplitude of the cumulative erosion
-potential, but we were probably a bit too fast deeming one erosion law "more
-realistic" than the others. Based on the estimates of total Pleistocene glacial
+This sentence was about the amplitude of the cumulative erosion potential, but
+we were probably a bit too fast deeming one erosion law "more realistic" than
+the others. We are not sure what the cumulative erosion potential of the last
+glacial cycle should be. Based on the estimates of total Pleistocene glacial
 relief from the suggested references, we now argue that no tested erosion law
@@
 
-The statement was reworked and restricted to "the higest mountain cirques".
+The statement was reworked and restricted to "the highest mountain cirques".
 
@@
     2013).
+
+Thank you again for reading our work and giving us constructive feedback.

### 2021-05-27 articles/erosion/response-referee2.md
Read through the responses once again.

@@
 Thank you very much for your and your colleague's time in preparing this
-review of our manuscript.
+reviewing our manuscript. We try to address your comments one by one below, and
+highlight relevant changes made to the manuscript.
 
@@
 Increasing ice volume yields the build-up of less steep and thus slower-flowing
-valley glaciers. Decreasing ice volume yields to (roughly) equally steep but
+valley glaciers. Decreasing ice volume leads to (roughly) equally steep but
 smaller glaciers. In the case of non-linear erosion laws ('Koppes' and 'Herman'
@@
 
-The above limitateions were made explicit with the following added sentences in
+The above limitations were made explicit with the following added sentences in
 the methods subsection on the "erosion law":
@@
 "climate sensitivity" section of the discussion mentioning the general
-eastwards bias of our results with regard to field evidence.
+eastwards bias of our results with regard to field evidence:
 
@@
 decreasing ice volume. This actually strikingly highlights the two different
-regimes. Thank you for the suggestion. The label was changed to 'potential
+regimes, so thank you for the suggestion. The label was changed to 'potential
 annual erosion volume'. In the text, 'slight' was replaced with a 'general'
@@
 > - Ian S. Evans, Durham University, U.K.
+
+Thank you again to all of you for taking time to read our work and offer
+constructive feedback.

### 2021-05-27 articles/erosion/alps-erosion.tex
Read through the responses once again.
Add reviewers in acknowledgements, tag as 2.2.

@@
           Shaded gray areas indicate the timing for MIS~2 and~4
-          \citep{Lisiecki.Raymo.2005}.
-        \todo{hatches on all plots.}}
+          \citep{Lisiecki.Raymo.2005}.}
         \label{fig:cumulative}
@@
           line).
-        \textbf{(c)} Same as Fig.~\ref{fig:cumulative}b.
-        \todo{fix labels, test hatching below a minimum number of grid cells}}
+        \textbf{(c)} Same as Fig.~\ref{fig:cumulative}b.}
       \label{fig:hypsogram}
+
     \end{figure*}
@@
     John Jansen and Martin Margold for insightful early discussions, Susan
-    Ivy-Ochs for being a constant (droid) motivator, Popi, Karsta and Stefan
+    Ivy-Ochs for being a constant (droid) motivator, Popi, Mattheos, Karsta,
+    Stefan, Fran\c{c}oise and Dominique
     for welcoming him on Anafi and providing agreeable conditions for work in
@@
     conversations while at University of Lausanne.
+    An anonymous reviewer and Ian Evans (aided by Iestyn Barr, Jeremy Ely, Matt
+    Tomkins, Pippa Whitehouse, Cristina Balaban and Matt Wiecek) provided
+    constructive feedback which has been incorporated in the present paper.
     The (previously published) ice flow simulations were made with computer

### 2021-06-01 figures/alpero_evolution.py
Hatch low-confidence regions on several plots.

@@ -38,10 +38,15 @@ def main():
     ax.plot(ds.slvol, ds.rolling_mean.where(~ds.growing), c='C11')
     ax.text(0.95, 0.95, '', ha='right', va='top', transform=ax.transAxes)
 
+    # hatch regions of low ice volume
+    ax.fill_between(
+        [-3, 3], 1e4, 1e8, edgecolor='0.75', facecolor='none', hatch='//////')
+
     # set axes properties
     ax.set_xlabel('ice volume (cm s.l.e.)')
     ax.set_ylabel(r'potential annual erosion volume ($m^3 a^{-1}$)')
     ax.set_yscale('log')
+    ax.set_xlim(-1.5, 31.5)
     ax.set_ylim(10**4.3, 10**7.7)
 
     # annotate advance and retreat

### 2021-06-01 figures/alpero_hypsogram.py

@@ -43,7 +43,7 @@ def main():
             '../data/processed/alpero.1km.epic.pp.agg.nc') as ds:
 
         # maybe the only way to set hatch color
-        plt.rc('hatch', color='0.25')
+        plt.rc('hatch', color='0.75')
 
         # plot hypsogram
         (np.log10(ds.kop2015_hyps)+3).plot.imshow(

### 2021-06-01 figures/alpero_powerlaws.py

@@ -104,10 +104,16 @@ def main():
             tsax.plot(slvol*100, ds[ref+'_rate'].rolling(
                 age=100, center=True).mean(), c='C11')
 
+            # hatch regions of low ice volume
+            tsax.fill_between(
+                [-3, 3], 1e4, 1e11,
+                edgecolor='0.75', facecolor='none', hatch='//////')
+
             # set time series axes properties
             tsax.set_xlabel('ice volume (cm s.l.e.)')
             tsax.set_yscale('log')
             tsax.set_ylabel('')
+            tsax.set_xlim(-1.5, 31.5)
             tsax.set_ylim(10**4.5, 10**10.5)
             tsax.set_yticks([1e5, 1e7, 1e9])
             tsax.yaxis.set_ticks_position('right')

### 2021-06-01 figures/util/ero.py

@@ -28,14 +28,21 @@ def plot_series(ax=None, run='1km.epic.pp'):
         ax.set_ylim(-5, 35)
         ax.locator_params(axis='y', nbins=6)
 
+        # hatch regions of low ice volume
+        ax.fill_between(
+            ds.age, -10, 40, where=(ds.slvol < 0.03),
+            edgecolor='0.75', facecolor='none', hatch='//////')
+
     # load aggregated data
     with hyoga.open.dataset(
             '../data/processed/alpero.'+run+'.agg.nc') as ds:
 
         # plot erosion rate time series
-        ds['rolling_mean'] = ds.kop2015_rate.rolling(age=100, center=True).mean()
+        ds['rolling_mean'] = ds.kop2015_rate.rolling(
+            age=100, center=True).mean()
         ax = ax.twinx()
         ax.plot(ds.age, ds.kop2015_rate*1e-6, c='C11', alpha=0.5)
         ax.plot(ds.age, ds.rolling_mean*1e-6, c='C11')
-        ax.set_ylabel('potential annual erosion\n'+r'volume ($10^6\,m^3 a^{-1}$)', color='C11')
+        ax.set_ylabel('potential annual erosion\n'
+                      r'volume ($10^6\,m^3 a^{-1}$)', color='C11')
         ax.set_ylim(-1, 7)

### 2021-06-01 articles/erosion/response-referee2.md
Caption hatches, update response to reviewer.

@@
 
-TODO: also emphasize this uncertainty in several of our plots by using
-hatches over areas of limited glacier cover, documented in the figure captions?
+To avoid over-interpretation from future readers, this uncertainty was
+highlighted in several of our plots (including Figs. 3 and 7 ) using hatches
+over periods of limited glacier volume, documented in the figure captions.
 

### 2021-06-01 articles/erosion/alps-erosion.tex
Caption hatches, update response to reviewer.

@@
           Shaded gray areas indicate the timing for MIS~2 and~4
-          \citep{Lisiecki.Raymo.2005}.}
+          \citep{Lisiecki.Raymo.2005}. Hatches mark periods with ice volume
+          below 3\,cm~s.l.e. where glacier sliding may be affected by
+          stress-balance approximations and model horizontal resolution (see
+          Sect.~\ref{sec:landscape}).}
         \label{fig:cumulative}
@@
         modelled total ice volume in centimetres of sea-level equivalent. Blue
-        indicates increasing ice volume and brown decreasing ice volume.}
+        indicates increasing ice volume and brown decreasing ice volume.
+        Hatches indicate low confidence due to limited glacier volume.}
       \label{fig:evolution}
@@
 \subsection{Age of the glacial landscape}
+\label{sec:landscape}
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -

### 2021-06-01 articles/erosion/alps-erosion.tex
Aesthetic changes on figures, greyscale-friendly.

@@
           reconstruction of Last Glacial Maximum Alpine glacier extent for
-          comparison \citep[solid red line,][]{Ehlers.etal.2011}.
+          comparison \citep[solid blue line,][]{Ehlers.etal.2011}.
           The background maps consists of the initial basal topography from
@@
         modelled total ice volume in centimetres of sea-level equivalent. Blue
-        indicates increasing ice volume and brown decreasing ice volume.
-        Hatches indicate low confidence due to limited glacier volume.}
+        dashes indicates increasing ice volume and solid brown decreasing ice
+        volume. Hatches indicate low confidence due to limited glacier volume.}
       \label{fig:evolution}
@@
         \textbf{(a)} Modelled basal and surface topographies interpolated along
-        the Rhine Glacier during glacier advance (36\,ka), maximum extent
-        (24\,ka) and retreat (16\,ka, using the same transect as in
-        Fig.~\ref{fig:basaldrag}).
+        the Rhine Glacier during glacier advance (36\,ka, dashed blue), maximum
+        extent (24\,ka, solid grey) and retreat (16\,ka, solid brown, using the
+        same transect as in Fig.~\ref{fig:basaldrag}).
         \textbf{(b)} Interpolated yield stress in the pseudo-plastic sliding law

### 2021-06-01 figures/alpero_basaldrag.py

@@ -9,6 +9,7 @@ import cartowik.profiletools as cpf
 import cartowik.decorations as cde
 import absplots as apl
 import hyoga.open
+import util  # noqa color cycle
 
 
 def main():
@@ -26,7 +27,9 @@ def main():
         '../data/native/profile_rhine.shp', interval=1e3)
 
     # each requested age
-    for age in [36, 24, 16]:
+    # FIXME add stresses in postprocessing
+    for age, color, style in zip(
+            [36, 24, 16], ['C1', '0.75', 'C11'], ['--', '-', '-']):
         with hyoga.open.subdataset(
                 ('~/pism/output/e9d2d1f/alpcyc4.1km.epica.1220.pp'
                  '/ex.{:07.0f}.nc'), time=-age*1e3, shift=120000) as ds:
@@ -37,7 +40,8 @@ def main():
 
             # plot topographic profiles
             ds['usurf'] = ds.topg + ds.thk.fillna(0)
-            ds.usurf.plot.line(ax=axes[0], x='d', label='{} ka'.format(age))
+            ds.usurf.plot.line(ax=axes[0], x='d', color=color, ls=style,
+                               label='{} ka'.format(age))
             ds.topg.plot.line(ax=axes[0], x='d', color='0.25')
 
             # plot normalized stresses
@@ -46,7 +50,7 @@ def main():
                 ax=axes[1], x='d', c='0.25', ls=':',
                 label='yield stress / overburden' if age == 16 else ' ')
             ((ds.taub_x**2+ds.taub_y**2)**0.5/overburden*100).plot.line(
-                ax=axes[1], x='d',
+                ax=axes[1], x='d', color=color, ls=style,
                 label='basal drag / overburden' if age == 16 else ' ')
 
             # to plot ratio governing sliding

### 2021-06-01 figures/alpero_cumulative.py

@@ -31,7 +31,7 @@ def main():
     # add map elements
     util.geo.draw_boot_topo(ax)
     util.geo.draw_natural_earth(ax)
-    util.geo.draw_lgm_outline(ax)
+    util.geo.draw_lgm_outline(ax, edgecolor='C1')
     cde.add_scale_bar(ax, label='100 km', length=100e3, pad=220e3)
 
     # plot time series

### 2021-06-01 figures/alpero_evolution.py

@@ -34,7 +34,7 @@ def main():
     # plot
     ax.plot(ds.slvol, ds.kop2015_rate.where(ds.growing), c='C1', alpha=0.25)
     ax.plot(ds.slvol, ds.kop2015_rate.where(~ds.growing), c='C11', alpha=0.25)
-    ax.plot(ds.slvol, ds.rolling_mean.where(ds.growing), c='C1')
+    ax.plot(ds.slvol, ds.rolling_mean.where(ds.growing), c='C1', ls='--')
     ax.plot(ds.slvol, ds.rolling_mean.where(~ds.growing), c='C11')
     ax.text(0.95, 0.95, '', ha='right', va='top', transform=ax.transAxes)
 

### 2021-06-01 articles/erosion/response-referee1.md
Again read through the complete responses.

@@
 Thank you very much for your positive comments and for your time spent
-reviewing our manuscript. We try to address your comments one by one below, and
+reviewing our work. We try to address your comments one by one below, and
 highlight relevant changes made to the manuscript.
@@
 imprint of glaciers on mountain topography, and the link between climate and
-glacier erosion (somewhat adressed in the conclusions). This is followed by a
+glacier erosion (somewhat addressed in the conclusions). This is followed by a
 short methodological bibliography of process-based and empirical glacier
@@
 interpretation was added in the "age of the glacial landscape" subsection.
-The new passage reads as follow:
+This new passage reads as follow:
 
     The validity of the model results at high elevation is discussable.
-    Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
+    Crystalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
     Ötztal and Tauern Massifs locally exhibit a strikingly high erosion
@@
     (Preusser et al., 2011; Valla et al., 2011), a cumulative glacial erosion
-    for the last glacial cycle in the order of 10-100 m can be expected.
+    for the last glacial cycle in the order of 10 to 100 m can be expected.
     However, none of the tested erosion power-laws fall within this range.
@@
 the highest peaks to deep valleys. This information was added to the plot by
-hatching the regions with fewer than a hundred ice-covered grid cells. This
+hatching the regions with fewer than a hundred ice-covered grid cells per
+elevation band. This
 highlights that low-elevation values are not always representative, as can also
@@
 ## References not in the manuscript:
-
-> Egholm, D.L., Nielsen, S.B., Pedersen, V.K., Lesemann, J.E., 2009. Glacial
-> effects limiting mountain height: Nature, v. 460, p. 884–888.
-
-> Ehlers, J., Gibbard, P.L., 2007. The extent and chronology of Cenozoic Global
-> Glaciation. Quat. Int. 164–165, 6–20.
-
-> Harbor, J.M., Hallet, B., and Raymond, C.F., 1988. A numerical model of
-> landform development by glacial erosion: Nature, v. 333, p. 347–349.
-
-> Kelly, M.A., Buoncristiani, J.F. and Schlüchter, C., 2004. A reconstruction
-> of the last glacial maximum (LGM) ice-surface geometry in the western Swiss
-> Alps and contiguous Alpine regions in Italy and France. Eclogae Geol. Helv.
-> 97, 57–75.
-
-> Mey, J., Scherler, D., Wickert, A.D., Egholm, D.L., Tesauro, M., Schildgen,
-> T.F., and Strecker, M., R., 2016. Glacial isostatic uplift of the European
-> Alps: Nature Communications, v. 7.
-
-> Preusser, F., Reitner, J.M., and Schlüchter, C., 2010. Distribution,
-> geometry, age and origin of overdeepened valleys and basins in the Alps and
-> their foreland: Swiss Journal Geoscience, v. 103, p. 407–426.
-
-> Sanders, J.W., Cuffey, K.M., Moore, J.R., MacGregor, K.R., and Kavanaugh,
-> J.L., 2012. Periglacial weathering and headwall erosion in cirque glacier
-> bergschrunds: Geology, v. 40, p. 779-782.
-
-> Sternai, P., Herman, F., Valla, P.G., Champagnac, J.-D., 2013. Spatial and
-> temporal variations of glacial erosion in the Rhône valley (Swiss Alps):
-> Insights from numerical modeling. Earth and Planetary Science Letters, 368,
-> p. 119-131.
-
-> Valla, P.G., Shuster, D.L., van der Beek, P.A., 2011. Significant increase in
-> relief of the European Alps during mid-Pleistocene glaciations. Nat. Geosci.
-> 4, 688–692.
 
@@
 
-Thank you again for reading our work and giving us constructive feedback.
+Thank you again for taking the time to read our work and to give constructive
+feedback in these troubled times. We apologize for the delay and will soon be
+submitting a revised and, we believe, improved manuscript including the
+aforementioned suggested changes.

### 2021-06-01 articles/erosion/response-referee2.md
Again read through the complete responses.

--- 2021-05-27 articles/erosion/response-referee2.md
@@
 Thank you very much for your and your colleague's time in preparing this
-reviewing our manuscript. We try to address your comments one by one below, and
+review of our manuscript. We try to address your comments one by one below, and
 highlight relevant changes made to the manuscript.
@@
 laws), both yield a decrease of erosion volume, hence the local peak in erosion
-volume. However, we would rather not include this as a result as the modelled
+volume. However, we would rather not highlight this result. The modelled
 glacier velocities for such situations are unreliable due to the limiting
@@
 
-TODO: also emphasize this uncertainty in several of our plots by using
-hatches over areas of limited glacier cover, documented in the figure captions?
+To avoid over-interpretation from future readers, this uncertainty was also
+highlighted in several of our plots (including Figs. 3 and 7) using hatches
+over periods of limited glacier volume, and documented in the figure captions.
 
@@
 the LGM and with a similar extent to the deglacial stage of 16 ka. This also
-corresponds to the topographic profiles shown on Fig. 8 and thus seems more
+corresponds to the topographic profiles shown on Fig. 8. Thus it seems more
 appropriate in hindsight.
-
-TODO: also include complete maps as a supplement?
 
@@
 
-    Instead , we we assume that eroded material is instantly transported out of
-    the system, thus neglegting its role in shielding the bedrock from glacier
+    Instead, we assume that eroded material is instantly transported out of
+    the system, thus neglecting its role in shielding the bedrock from glacier
     erosion in zones of temporary storage (Preusser et al., 2010). Neither do
@@
 neither of which includes lateral stresses. Glacier sliding velocities in
-troughs are goverened by (the sliding law and) longitudinal stresses, which is
+troughs are governed by (the sliding law and) longitudinal stresses, which is
 a limitation with respect to higher-order models. Following another comment
@@
 the caption for the cirque photo now reads: "the unusually deep mountain cirque
-revealed by the current demise of Chüebodengletscher" (I don't have a much
+revealed by the current demise of Chüebodengletscher" (we don't have a much
 better photo for a high-elevation cirque, and glacial lakes from retreating
-cirque glaciers are a common occurence in the Alps these days).
+cirque glaciers are a common occurrence in the Alps these days).
 
@@
 hundred ice-covered grid cells for a given time. We also shifted to 100-m
-elevation bands on the main panel. Wihle some detail is lost, this avoids
+elevation bands on the main panel. While some detail is lost, this avoids
 much of the striping (especially on the newly added contour and hatched
@@
 side of the profile in this region. Because the upper part of the valley is
-narrow, the glacier center-line fluctuating over time, and the plot
-interpolated between 1-km grid cells, these lateral components enter the
+narrow, the glacier centre-line fluctuating over time, and the plot
+interpolated between 1-km grid points, these lateral components enter the
 computation of the basal shear stress magnitude in the higher reaches.
@@
 divide-by-zero error (or a log-of-zero error). Both animation panels have a
-log-scaled x-axis, hence they cannot contain zero values.
+log-scaled x-axis. While some values fall outside of the axis frame, they are
+very small but not zero.
 
@@
 (due to the power 2.34 in the erosion law). Such values are not really relevant
-and they are off-the-colour-scheme on the Fig. 5 and off-the-chart in the
+and they are off the colour range on Fig. 5 and off-chart in the
 animation (respectively lower than 10^-9 and 10^-10 mm a^-1).
@@
 kind of values we have previously filtered out. Your comment led us to re-check
-our calculations. We found that there is no zero-sliding values in the
-ice-covered area, and that an ice mask is sufficient to filter out undesirable
+our calculations. We found that there is no zero-sliding values within the
+ice-covered area, and thus an ice mask is sufficient to filter out undesirable
 values and compute the geometric means. Hence "non-zero" was removed from the
@@
 
-## References not in original:
-
-> Egholm, D.L., Pedersen, V.K., Knudsen, M.F., Larsen, N.K., 2012a. Coupling
-> the flow of ice, water, and sediment in a glacial landscape evolution model.
-> Geomorphology 141–142: 47–66.
-> http://dx.doi.org/10.1016/j.geomorph.2011.12.019.
-
-> Egholm, D.L., Pedersen, V.K., Knudsen, M.F., Larsen, N.K., 2012b. On the
-> importance of higher-order ice dynamics for glacial landscape evolution.
-> Geomorphology 141–142: 67-80.
-
-> Farinotti, D., et al., 2019. A consensus estimate for the ice thickness
-> distribution of all glaciers on Earth. Nat. Geosci. 12, 168–173.
-
-> Lai, J., Anders, A.M., 2021. Climatic controls on mountain glacier basal
-> thermal regimes dictate spatial patterns of glacial erosion.
-> https://esurf.copernicus.org/preprints/esurf-2021-26/
-
-> Pedersen, V.K., Huismans, R.S., Herman, F., Egholm, D.L., 2014. Controls of
-> initial topography on temporal and spatial patterns of glacial erosion.
-> Geomorphology 223, 96-116.
-
 ## Acknowledgements:
@@
 Thank you again to all of you for taking time to read our work and offer
-constructive feedback.
+constructive feedback in these troubled times. We hope that you will find our
+answers satisfactory, apologize for the delay and will soon be submitting a
+revised and, we believe, improved manuscript including the aforementioned
+suggested changes.

### 2021-06-01 articles/erosion/alps-erosion.tex
Again read through the complete responses.

@@
     or fluvial transport of subglacial sediment \citep{Delaney.etal.2019}.
-    Instead we assume that eroded material is instantly transported out of the
-    system, thus neglegting its role in shielding the bedrock from glacier
+    Instead, we assume that eroded material is instantly transported out of the
+    system, thus neglecting its role in shielding the bedrock from glacier
     erosion in zones of temporary storage \citep{Preusser.etal.2010}. Neither
@@
     \citep{Preusser.etal.2011, Valla.etal.2011}, a cumulative glacial erosion
-    for the last glacial cycle in the order of 10--100\,m can be expected.
+    for the last glacial cycle in the order of 10 to 100\,m can be expected.
     However, none of the tested erosion power-laws fall within this range.
@@
     \citep{Herman.etal.2011, Egholm.etal.2012, Egholm.etal.2012a,
-    Pedersen.etal.2014}. \todo{check these references}.
+    Pedersen.etal.2014}.
 
@@
     The validity of the model results at high elevation is discussable.
-    Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
+    Crystalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
     \"Otztal and Tauern Massifs locally exhibit a strikingly high erosion

### 2021-06-01 articles/erosion/response-referee1.md
Extend the video supplement paragraph.

@@
 Thank you again for taking the time to read our work and to give constructive
-feedback in these troubled times. We apologize for the delay and will soon be
-submitting a revised and, we believe, improved manuscript including the
-aforementioned suggested changes.
+feedback in these troubled times. We hope that you will find our answers
+satisfactory, apologize for the delay and will soon be submitting a revised
+and, we believe, improved manuscript including the aforementioned suggested
+changes.

### 2021-06-01 articles/erosion/alps-erosion.tex
Extend the video supplement paragraph.

@@
 \videosupplement{
-    Animations made available online show the modelled glacier erosion rates
-    (\url{https://vimeo.com/503162771}), and their evolving relationship with
-    ice volume (\url{https://vimeo.com/512478926}) and bedrock
-    altitude (\url{https://vimeo.com/512479008}).}
+    Animations made available online show the spatio-temporal distribution of
+    modelled glacier potential erosion rates
+    (\url{https://vimeo.com/503162771}, supplementing
+    Fig.~\ref{fig:cumulative}), the evolving relationship between ice volume
+    and potential annual erosion volume (\url{https://vimeo.com/512478926},
+    supplementing Fig.~\ref{fig:evolution}), and the distribution of potential
+    erosion rate against bedrock altitude (\url{https://vimeo.com/512479008},
+    supplementing Fig.~\ref{fig:hypsogram}).}
 

### 2021-06-01 articles/erosion/response-referee1.md
Reformat responses for web form, tag 2.3.

@@
 
-# Authors' response to Anonymous Referee #1
+# Authors' response on RC1
 
@@
 
-Thank you very much for your positive comments and for your time spent
-reviewing our work. We try to address your comments one by one below, and
-highlight relevant changes made to the manuscript.
+Thank you very much for your positive comments and for your time spent reviewing our work. We try to address your comments one by one below, and highlight relevant changes made to the manuscript.
 
-> The study of Seguinot and Delaney presents a time-integrated model on the
-> glacial erosion potential over the last 120 ka in the Alps. Authors tie the
-> calculations on a previous study by Seguinot et 2018 who modeled the glacial
-> extend over Alps’ last glacial cycle by testing three different records on
-> palaeo-temperature (GRIP, EPICA, and MD01-2444). Calculated basal velocities
-> which base on the Parallel Ice Sheet Model PISM were used to test different
-> glacial postulated erosional laws (Koppes et al., 2015; Herman et al., 2015,
-> Humphrey and Raymond, 1994; Cook et al., 2020) focusing on the Rhine glacier
-> area.
+> The study of Seguinot and Delaney presents a time-integrated model on the glacial erosion potential over the last 120 ka in the Alps.
+Authors tie the calculations on a previous study by Seguinot et 2018 who modeled the glacial extend over Alps’ last glacial cycle by testing three different records on palaeo-temperature (GRIP, EPICA, and MD01-2444).
+Calculated basal velocities which base on the Parallel Ice Sheet Model PISM were used to test different glacial postulated erosional laws (Koppes et al., 2015; Herman et al., 2015, Humphrey and Raymond, 1994; Cook et al., 2020) focusing on the Rhine glacier area.
 
-> The main achievement of the study is to combine a realistic, high-resolution
-> model of glacial extension over a glacial cycle in a mountain range with
-> assumptions on glacial erosion. Even though the utilized PISM code does not
-> account for feedbacks of glacial erosion on bed topography, which makes it
-> not directly comparable with common landscape codes on glacial erosion (e.g.
-> Egholm et al. 2009; Sternai, et al., 2013), some relevant findings could be
-> derived by this integrated high-resolution approach. This includes the
-> observation of low erosion during glacier advance and maximum glacier
-> extension, and the role of profile steeping during deglaciation and related
-> increasing erosion rates.
+> The main achievement of the study is to combine a realistic, high-resolution model of glacial extension over a glacial cycle in a mountain range with assumptions on glacial erosion.
+Even though the utilized PISM code does not account for feedbacks of glacial erosion on bed topography, which makes it not directly comparable with common landscape codes on glacial erosion (e.g.
+Egholm et al.
+2009; Sternai, et al., 2013), some relevant findings could be derived by this integrated high-resolution approach.
+This includes the observation of low erosion during glacier advance and maximum glacier extension, and the role of profile steeping during deglaciation and related increasing erosion rates.
 
@@
 
-> Authors decided to keep the paper short, referring most information on
-> glacial model set up and resulting ice cover to the Seguinot et 2018 paper.
-> This makes it no always very easy to read and needs checking in the original
-> manuscript (for example, ice cover deviations model – field bases LGM
-> extension).
+> Authors decided to keep the paper short, referring most information on glacial model set up and resulting ice cover to the Seguinot et 2018 paper.
+This makes it no always very easy to read and needs checking in the original manuscript (for example, ice cover deviations model – field bases LGM extension).
 
-Indeed, it is our deliberate choice to avoid paraphrasing or duplicating
-content from the 2018 paper, where the preparation of the text and figures also
-involved a different author team. We believe this is justified given that both
-papers will be published open-access. Nonetheless, small additions were made.
+Indeed, it is our deliberate choice to avoid paraphrasing or duplicating content from the 2018 paper, where the preparation of the text and figures also involved a different author team.
+We believe this is justified given that both papers will be published open-access.
+Nonetheless, small additions were made.
 Please see our responses to your technical comments below.
 
-> The introduction gives a nice, relatively long (in relation to other
-> chapters) overview on different aspects of glacial erosion and resulting
-> features, and guides through problems to investigate them. Here, I don’t see
-> very clear how different aspects or parts in the introduction are
-> specifically addressed in this study, i.e. which problems are exactly aimed
-> to be solved of this large portfolio of mentioned gaps in the understanding
-> of glacial processes. I think this could be better formulated and balanced.
+> The introduction gives a nice, relatively long (in relation to other chapters) overview on different aspects of glacial erosion and resulting features, and guides through problems to investigate them.
+Here, I don’t see very clear how different aspects or parts in the introduction are specifically addressed in this study, i.e.
+which problems are exactly aimed to be solved of this large portfolio of mentioned gaps in the understanding of glacial processes.
+I think this could be better formulated and balanced.
 
-Our introduction is meant to revolve around two discussion points: the variable
-imprint of glaciers on mountain topography, and the link between climate and
-glacier erosion (somewhat addressed in the conclusions). This is followed by a
-short methodological bibliography of process-based and empirical glacier
-erosion laws, before we announce the scope of our work. A few key statements
-were added to try and better tie these different sections.
+Our introduction is meant to revolve around two discussion points: the variable imprint of glaciers on mountain topography, and the link between climate and glacier erosion (somewhat addressed in the conclusions).
+This is followed by a short methodological bibliography of process-based and empirical glacier erosion laws, before we announce the scope of our work.
+A few key statements were added to try and better tie these different sections.
 
-> I think the study could also well contribute to the discussion on the
-> elevation distribution of the cumulative glacial erosion over several cycles
-> in the Alps (high elevations vs low elevation, e.g. Valla et al., 2011). A
-> plot showing the integrated glacial erosion potential over (e.g. 100 m)
-> elevation bins would easy to do and eventually an enlightening supplement
-> summing up Fig. 5a.
+> I think the study could also well contribute to the discussion on the elevation distribution of the cumulative glacial erosion over several cycles in the Alps (high elevations vs low elevation, e.g.
+Valla et al., 2011).
+A plot showing the integrated glacial erosion potential over (e.g.
+100 m) elevation bins would easy to do and eventually an enlightening supplement summing up Fig.
+5a.
 
-Thank you for the suggestion. Figure 5 was amended with the
-distribution of cumulative erosion potential over 100 m elevation bins.
-The resulting plot shows that most modelled erosion occurs below 2000 m,
-with the potential erosion volumes between 1000 and 2000 m elevation.
+Thank you for the suggestion.
+Figure 5 was amended with the distribution of cumulative erosion potential over 100 m elevation bins.
+The resulting plot shows that most modelled erosion occurs below 2000 m, with the potential erosion volumes between 1000 and 2000 m elevation.
 
-> I would also recommend to make some regional statements on the distribution
-> of the glacial erosion potential (Fig. 2). It is quite obvious that some
-> tectonic massifs can readily be discriminated, e.g. the Tauern Window, Oetzal
-> Crystallin complex, the Aare Massif, Mt. Rosa - Gran Paradiso and Pelvoux
-> Massif. Is it the steepness of these low erodible crystallin massifs making
-> the erosional potential appearing strikingly high? I think this should/could
-> be discussed..
+> I would also recommend to make some regional statements on the distribution of the glacial erosion potential (Fig.
+2).
+It is quite obvious that some tectonic massifs can readily be discriminated, e.g.
+the Tauern Window, Oetzal Crystallin complex, the Aare Massif, Mt.
+Rosa - Gran Paradiso and Pelvoux Massif.
+Is it the steepness of these low erodible crystallin massifs making the erosional potential appearing strikingly high?
+I think this should/could be discussed..
 
-Our calculations indeed produce locally high cumulative erosion potential in
-the aforementioned crystalline massifs. More accurate modelling of small-scale
-glaciers would be needed to provide an accurate answer to your question. Steep
-topography, though, is
-likely to be part of the explanation. A more substantial discussion of the high
-erosion potential values in these areas and the model limitations that hinder
-interpretation was added in the "age of the glacial landscape" subsection.
+Our calculations indeed produce locally high cumulative erosion potential in the aforementioned crystalline massifs.
+More accurate modelling of small-scale glaciers would be needed to provide an accurate answer to your question.
+Steep topography, though, is likely to be part of the explanation.
+A more substantial discussion of the high erosion potential values in these areas and the model limitations that hinder interpretation was added in the "age of the glacial landscape" subsection.
 This new passage reads as follow:
 
-    The validity of the model results at high elevation is discussable.
-    Crystalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
-    Ötztal and Tauern Massifs locally exhibit a strikingly high erosion
-    potential. However, the computation of glacier flow velocities on such
-    steep surfaces is strongly limited by the model horizontal resolution of
-    1\,km, the shallow-ice glacier flow physics (Imhof et al., 2019), and
-    PISM's current mass-conservation heuristics (Imhof, 2021). Besides,
-    bergschrund (rimaye) processes likely to dominate interglacial cirque
-    erosion at such altitudes (Sanders et al., 2012) are not captured by the
-    velocity-based glacier erosion power-laws.
+    "The validity of the model results at high elevation is discussable.
+Crystalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare, Ötztal and Tauern Massifs locally exhibit a strikingly high erosion potential.
+However, the computation of glacier flow velocities on such steep surfaces is strongly limited by the model horizontal resolution of 1\,km, the shallow-ice glacier flow physics (Imhof et al., 2019), and PISM's current mass-conservation heuristics (Imhof, 2021).
+Besides, bergschrund (rimaye) processes likely to dominate interglacial cirque erosion at such altitudes (Sanders et al., 2012) are not captured by the velocity-based glacier erosion power-laws."
 
@@
 
-> Line 24 Maybe a bit odd to use 4 very old references from the Alps only (you
-> hardly used more than 2 references throughout the MS and there are also high
-> variety of glacial landforms in other mt ranges as you mentioned).
+> Line 24 Maybe a bit odd to use 4 very old references from the Alps only (you hardly used more than 2 references throughout the MS and there are also high variety of glacial landforms in other mt ranges as you mentioned).
 
-We agree that the references are not directly relevant. The citation was
-reduced to the latest reference by Penck (1905) and moved within the sentence
-to clarify that it is specific to the Alps.
+We agree that the references are not directly relevant. The citation was reduced to the latest reference by Penck (1905) and moved within the sentence to clarify that it is specific to the Alps.
 
-> Line 33 I think that’s not very easy to understand what landform you refer
-> to? The “periglacial blockfields topped by glacial erratic boulders” (Wirsig
-> et al., 2018)? Eventually be more precise on this would help readers.
+> Line 33 I think that’s not very easy to understand what landform you refer to?
+The “periglacial blockfields topped by glacial erratic boulders” (Wirsig et al., 2018)?
+Eventually be more precise on this would help readers.
 
-Agreed. "Characteristic landform preservation" was replaced with "preserved
-periglacial blockfields topped by erratic boulders".
+Agreed. "Characteristic landform preservation" was replaced with "preserved periglacial blockfields topped by erratic boulders".
 
-> Line 101 Could be misleading as (glacial and periglacial) cirque erosion
-> processes are not really covered by any glacial erosion law discussed here
-> (or any other I am aware of, Sanders et al., 2012).
+> Line 101 Could be misleading as (glacial and periglacial) cirque erosion processes are not really covered by any glacial erosion law discussed here (or any other I am aware of, Sanders et al., 2012).
 
-Thank you for pointing out this study. This particular sentenced was reworked
-to acknowledge not only the limiting resolution but also the lack of relevant
-erosion erosion processes. The reference was additionally incorporated in the
-relevant part of the discussion (see general comments above).
+Thank you for pointing out this study.
+This particular sentenced was reworked to acknowledge not only the limiting resolution but also the lack of relevant erosion erosion processes.
+The reference was additionally incorporated in the relevant part of the discussion (see general comments above).
 
@@
 
-> Line 127 Higher precipitation increases ice flux and thus erosion, I guess?
-> Would eventually helpful to mention (even though it might be referred in
-> Seguinot et al., 2018).
+> Line 127 Higher precipitation increases ice flux and thus erosion, I guess?  Would eventually helpful to mention (even though it might be referred in Seguinot et al., 2018).
 
@@
 
-> Line 140 Can you be more precise what you mean by realistic? You mean because
-> of the localized pattern? Maybe I am wrong but shouldn’t the erosion
-> potential at least at the lake Constance overdeepening (Fig. 7) not in the
-> order of hundreds of meters (cf. Preusser et al., 2010) and the best fit
-> rather (b) or (c) – at least from what I read from the transect (e-h)?
-> Visually (a)-(d) seem not to correspond to (e)-(h) if (a)-(d) is also
-> presented in meters (annotation at the bar is missing). For example, in (b)
-> the maximum erosion potential is like >>1000 m (if in meters) while in
-> corresponding (f) it seems clearly lower than 1000 m.
+> Line 140 Can you be more precise what you mean by realistic?
+You mean because of the localized pattern?
+Maybe I am wrong but shouldn’t the erosion potential at least at the lake Constance overdeepening (Fig.
+7) not in the order of hundreds of meters (cf.
+Preusser et al., 2010) and the best fit rather (b) or (c) – at least from what I read from the transect (e-h)?
+Visually (a)-(d) seem not to correspond to (e)-(h) if (a)-(d) is also presented in meters (annotation at the bar is missing).
+For example, in (b) the maximum erosion potential is like >>1000 m (if in meters) while in corresponding (f) it seems clearly lower than 1000 m.
 
-This sentence was about the amplitude of the cumulative erosion potential, but
-we were probably a bit too fast deeming one erosion law "more realistic" than
-the others. We are not sure what the cumulative erosion potential of the last
-glacial cycle should be. Based on the estimates of total Pleistocene glacial
-relief from the suggested references, we now argue that no tested erosion law
-give results in the expectable range of last glacial cycle erosion.
+This sentence was about the amplitude of the cumulative erosion potential, but we were probably a bit too fast deeming one erosion law "more realistic" than the others.
+We are not sure what the cumulative erosion potential of the last glacial cycle should be.
+Based on the estimates of total Pleistocene glacial relief from the suggested references, we now argue that no tested erosion law give results in the expectable range of last glacial cycle erosion.
 
-    With a total Pleistocene glacial relief on the order of a kilometre
-    (Preusser et al., 2011; Valla et al., 2011), a cumulative glacial erosion
-    for the last glacial cycle in the order of 10 to 100 m can be expected.
+    "With a total Pleistocene glacial relief on the order of a kilometre (Preusser et al., 2011; Valla et al., 2011), a cumulative glacial erosion for the last glacial cycle in the order of 10 to 100 m can be expected.
 However, none of the tested erosion power-laws fall within this range.
-    Instead, the erosion law calibrated on tidewater glaciers (Koppes et al.,
-    2015) yields cumulative erosion in the Rhine Valley in the orders of
-    metres, while the three erosion laws based on terrestrial glaciers
-    (Humphrey and Raymond, 1994; Herman et al., 2015; Cook et al., 2020},
-    result in kilometre-scale integrated erosion potential. During the Last
-    Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers
-    were closer in size, slope (an important parameter as we argue in the next
-    section), and climatic context to the present-day glaciers of Patagonia and
-    the Antarctic Peninsula (Koppes et al., 2015) than to Franz-Joseph Glacier
-    (Herman et al., 2015) and many of the glaciers included in the global
-    compilation by Cook et al., (2020). This may help to explain why the
-    reality appears to fall in-between the tested erosion laws.
+Instead, the erosion law calibrated on tidewater glaciers (Koppes et al., 2015) yields cumulative erosion in the Rhine Valley in the orders of metres, while the three erosion laws based on terrestrial glaciers (Humphrey and Raymond, 1994; Herman et al., 2015; Cook et al., 2020}, result in kilometre-scale integrated erosion potential.
+During the Last Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers were closer in size, slope (an important parameter as we argue in the next section), and climatic context to the present-day glaciers of Patagonia and the Antarctic Peninsula (Koppes et al., 2015) than to Franz-Joseph Glacier (Herman et al., 2015) and many of the glaciers included in the global compilation by Cook et al., (2020).
+This may help to explain why the reality appears to fall in-between the tested erosion laws."
 
-Regarding Fig. 7, there is no mismatch between panels (a-d) and (e-h), but to
-improve readability we have reduced the number of colour levels on the maps,
-and changed the positions of ticks on the transects.
+Regarding Fig.
+7, there is no mismatch between panels (a-d) and (e-h), but to improve readability we have reduced the number of colour levels on the maps, and changed the positions of ticks on the transects.
 
-> Furthermore, isn’t the impression of the fit potentially very dependent on
-> the initial model parametrization (Seguinot et al., 2018), i.e. the ice flux
-> velocity? You should address these dependences!
+> Furthermore, isn’t the impression of the fit potentially very dependent on the initial model parametrization (Seguinot et al., 2018), i.e.
+the ice flux velocity?
+You should address these dependences!
 
-A paragraph was added under "choice of erosion law" to discuss the
-uncertainties on glacier sliding physics. Also including suggestions from
-reviewer #2, the new paragraph reads as follow:
+A paragraph was added under "choice of erosion law" to discuss the uncertainties on glacier sliding physics. Also including suggestions from reviewer #2, the new paragraph reads as follow:
 
-    The modelled pattern of erosion potential depends on PISM glacier physics
-    and sliding model parameters. In particular, the pseudo-plastic sliding law
-    exponent (q=0.25 in Seguinot et al., 2018) controls the sharpness of the
-    transition between adherent and decoupled basal conditions. Recent ensemble
-    validation of Antarctic Ice Sheet glacial-cycle simulations against
-    geological and present-day observations (Albrecht et al., 2020a, b) support
-    a higher value of q=0.75, and thus a sliding law closer to linear, which
-    would perhaps result in a smoother distribution of sliding velocity and
-    erosion potential. Lateral stress gradients missing from the shallow-shelf
-    approximation stress balance could also contribute to moderate sliding
-    velocity in narrow troughs (Herman et al., 2011; Egholm et al, 2012a, b;
-    Pedersen et al., 2014).
+    "The modelled pattern of erosion potential depends on PISM glacier physics and sliding model parameters.
+In particular, the pseudo-plastic sliding law exponent (q=0.25 in Seguinot et al., 2018) controls the sharpness of the transition between adherent and decoupled basal conditions.
+Recent ensemble validation of Antarctic Ice Sheet glacial-cycle simulations against geological and present-day observations (Albrecht et al., 2020a, b) support a higher value of q=0.75, and thus a sliding law closer to linear, which would perhaps result in a smoother distribution of sliding velocity and erosion potential.
+Lateral stress gradients missing from the shallow-shelf approximation stress balance could also contribute to moderate sliding velocity in narrow troughs (Herman et al., 2011; Egholm et al, 2012a, b; Pedersen et al., 2014)."
 
-> Line 163 Observing Alpine topography I find this result important, which can
-> maybe also serve as explanation why e.g. (low erodible) areas away from the
-> big troughs covered during glacial maxima only, do surprisingly often show
-> no/very low degree in glacial modification (e.g. Ticino; Kelly et al., 2004).
+> Line 163 Observing Alpine topography I find this result important, which can maybe also serve as explanation why e.g.
+(low erodible) areas away from the big troughs covered during glacial maxima only, do surprisingly often show no/very low degree in glacial modification (e.g.
+Ticino; Kelly et al., 2004).
 
-Thank you for pointing this out. The following sentence was added in this
-paragraph:
+Thank you for pointing this out. The following sentence was added in this paragraph:
 
-    Nevertheless, it may explain why some areas covered during glacial maxima
-    only, such as some valleys on the southern side of the Alps, appear to have
-    experienced only little glacial modification of their topography.
+    "Nevertheless, it may explain why some areas covered during glacial maxima only, such as some valleys on the southern side of the Alps, appear to have experienced only little glacial modification of their topography."
 
-> Line 187 “time-transgressive radial pattern”. I don’t understand what you
-> mean..
+> Line 187 “time-transgressive radial pattern”. I don’t understand what you mean..
 
-This sentence was only meant to recall the aforementioned results. We rephrased
-to "the potential erosion patterns experience spatial shifts through the
-glacial cycle".
+This sentence was only meant to recall the aforementioned results. We rephrased to "the potential erosion patterns experience spatial shifts through the glacial cycle".
 
@@
 
-> Line 197 Would recommend to be more precise. Guess you know that there are
-> many, many cirques in the Alps as low as 1500 m (and even below) e.g. forming
-> in areas outside the connected ice stream network in the SW and easternmost
-> Alps. This can be revealed by a quick check at any higher resolution DEM or
-> google earth.
+> Line 197 Would recommend to be more precise.
+Guess you know that there are many, many cirques in the Alps as low as 1500 m (and even below) e.g.
+forming in areas outside the connected ice stream network in the SW and easternmost Alps.
+This can be revealed by a quick check at any higher resolution DEM or google earth.
 
@@
 
-> In the MS the word “yet” is very often used. Eventually consider reducing.
-> The frequency is a bit irritating when reading.
+> In the MS the word “yet” is very often used. Eventually consider reducing.  The frequency is a bit irritating when reading.
 
@@
 
-> Fig. 2 Please indicate the outline of the connected ice stream network during
-> the LGM (like in Fig. 4 of Seguinot et al., 2018). Especially in the eastern
-> and SW Alps coverage largely deviates from what has been suggested from field
-> data compilation (e.g. Ehlers and Gibbard, 2004). Even though outcomes might
-> not be changing much, it might be helpful to know how much the %overlap is -
-> erosional potential might change as fluvial topography turns into glacial one
-> (e.g. Harbor, 1988) and this is probably not what you want to mix, I guess.
+> Fig.
+2 Please indicate the outline of the connected ice stream network during the LGM (like in Fig.
+4 of Seguinot et al., 2018).
+Especially in the eastern and SW Alps coverage largely deviates from what has been suggested from field data compilation (e.g.
+Ehlers and Gibbard, 2004).
+Even though outcomes might not be changing much, it might be helpful to know how much the %overlap is - erosional potential might change as fluvial topography turns into glacial one (e.g.
+Harbor, 1988) and this is probably not what you want to mix, I guess.
 
-The field-based LGM outline was added on Fig. 2. Computing a percent overlap is
-not straightforward. As for Fig. 4 in the 2018 paper, we use a manually
-modified version of the Ehlers et al. (2011) data. The original data presents
-numerous, expansive nunataks, many of which are found in accumulation area and
-thus directly incompatible with the mass-balance scheme used in the 2018 paper.
+The field-based LGM outline was added on Fig.
+2.
+Computing a percent overlap is not straightforward.
+As for Fig.
+4 in the 2018 paper, we use a manually modified version of the Ehlers et al.
+(2011) data.
+The original data presents numerous, expansive nunataks, many of which are found in accumulation area and thus directly incompatible with the mass-balance scheme used in the 2018 paper.
 However, the general eastward bias was discussed in two new sentences:
 
-    It should be noted, however, that all runs presented here show a systematic
-    bias with excessive glacier cover in the Eastern Alps and a diminished
-    glacier extent in the Western Alps (Fig. 2a; further discussed in Seguinot
-    et al., 2018). Thus the modelled patterns of erosion potential certainly
-    includes a similar bias.
+    "It should be noted, however, that all runs presented here show a systematic bias with excessive glacier cover in the Eastern Alps and a diminished glacier extent in the Western Alps (Fig.
+2a; further discussed in Seguinot et al., 2018).
+Thus the modelled patterns of erosion potential certainly includes a similar bias."
 
-> Fig. 5 I have to confess Fig. 5a surprises me, there are really glaciers as
-> low as 500 m.a.s.l (and even below) from 110 – 40 ka in the Seguinot et al.,
-> 2018 model?
+> Fig. 5 I have to confess Fig. 5a surprises me, there are really glaciers as low as 500 m.a.s.l (and even below) from 110 – 40 ka in the Seguinot et al., 2018 model?
 
-This is correct. It concerns only a small number of grid cells, primarily
-located on the southern slope of the Alps where topography drops abruptly from
-the highest peaks to deep valleys. This information was added to the plot by
-hatching the regions with fewer than a hundred ice-covered grid cells per
-elevation band. This
-highlights that low-elevation values are not always representative, as can also
-be confirmed in the new subplot showing the distribution of cumulative erosion
-potential over 100 m elevation bins. The distribution of glacier cover per
-elevation band over time (from which the hundred-cell contour was extracted)
-will be included in a new version of the companion dataset.
+This is correct.
+It concerns only a small number of grid cells, primarily located on the southern slope of the Alps where topography drops abruptly from the highest peaks to deep valleys.
+This information was added to the plot by hatching the regions with fewer than a hundred ice-covered grid cells per elevation band.
+This highlights that low-elevation values are not always representative, as can also be confirmed in the new subplot showing the distribution of cumulative erosion potential over 100 m elevation bins.
+The distribution of glacier cover per elevation band over time (from which the hundred-cell contour was extracted) will be included in a new version of the companion dataset.
 
@@
 
-In addition to the aforementioned changes, the Sternai et al. (2013) reference
-was added in the discussion of the vertical distribution of erosion with this
-new sentence:
+In addition to the aforementioned changes, the Sternai et al. (2013) reference was added in the discussion of the vertical distribution of erosion with this new sentence:
 
-    Over longer timescales, though, the vertical distribution of erosion rates
-    also depends on the erosional modification of topography (Sternai et al.,
-    2013).
+    "Over longer timescales, though, the vertical distribution of erosion rates also depends on the erosional modification of topography (Sternai et al., 2013)."
 
-Thank you again for taking the time to read our work and to give constructive
-feedback in these troubled times. We hope that you will find our answers
-satisfactory, apologize for the delay and will soon be submitting a revised
-and, we believe, improved manuscript including the aforementioned suggested
-changes.
+Thank you again for taking the time to read our work and to give constructive feedback in these troubled times.
+We hope that you will find our answers satisfactory, apologize for the delay and will soon be submitting a revised and, we believe, improved manuscript including the aforementioned suggested changes.

### 2021-06-01 articles/erosion/response-referee2.md
Extend the video supplement paragraph.
Reformat responses for web form, tag 2.3.

@@
 
-# Authors' response to Ian Evans
+# Authors' response on RC2
 
@@
 
-Thank you very much for your and your colleague's time in preparing this
-review of our manuscript. We try to address your comments one by one below, and
-highlight relevant changes made to the manuscript.
+Thank you very much for your and your colleague's time in preparing this review of our manuscript.
+We try to address your comments one by one below, and highlight relevant changes made to the manuscript.
 
@@
 
-> The authors provide large amounts of very useful results from simulations,
-> concerning the effects of varying glacier geometry on erosion potential. It
-> was quite an achievement to compute this at fairly high temporal and spatial
-> resolution (1 km and 50 a throughout the Alps). The conclusions are
-> reasonable, and I value the results. They provide support for the
-> traditional view, that high-altitude features such as cirques are eroded by
-> steep, local glaciers and not under extensive icefields well above ELA. Most
-> erosion is expected along major valleys (glacial troughs): the potential
-> erosion is greatest low down the glaciers, with a lesser maximum at high
-> altitude and a minimum between. The paper is well written and the Figures
-> are all relevant and interesting. The qualitative conclusions are well
-> justified.
+> The authors provide large amounts of very useful results from simulations, concerning the effects of varying glacier geometry on erosion potential.
+It was quite an achievement to compute this at fairly high temporal and spatial resolution (1 km and 50 a throughout the Alps).
+The conclusions are reasonable, and I value the results.
+They provide support for the traditional view, that high-altitude features such as cirques are eroded by steep, local glaciers and not under extensive icefields well above ELA.
+Most erosion is expected along major valleys (glacial troughs): the potential erosion is greatest low down the glaciers, with a lesser maximum at high altitude and a minimum between.
+The paper is well written and the Figures are all relevant and interesting.
+The qualitative conclusions are well justified.
 
-> An interesting result, not commented on, is that from Figs. 3 and 7 i & j,
-> potential erosion volume in the Alps peaks when ice cover amounts to 1.6 cm
-> sea level equivalent. I would love to see a map of the glaciers at that
-> stage\*, which is well below maximum but considerably more than recent ice
-> cover (0.3 ± 0.1 mm s.l.e., Farinotti et al. 2019). This peak applies to the
-> ‘Koppes’ and ‘Herman’ erosion laws, but not to ‘Humphrey’ and ‘Cook’ laws
-> which give very broad maxima for s.l.e. 10-30 cm (Fig. 7 k & l).
+> An interesting result, not commented on, is that from Figs.
+3 and 7 i & j, potential erosion volume in the Alps peaks when ice cover amounts to 1.6 cm sea level equivalent.
+I would love to see a map of the glaciers at that stage\*, which is well below maximum but considerably more than recent ice cover (0.3 ± 0.1 mm s.l.e., Farinotti et al.
+2019).
+This peak applies to the ‘Koppes’ and ‘Herman’ erosion laws, but not to ‘Humphrey’ and ‘Cook’ laws which give very broad maxima for s.l.e.
+10-30 cm (Fig.
+7 k & l).
 >
->   [\*That stage can be approximated by pausing the first erosion
->    rates video around 99 ka and 13 ka. The three videos are strongly
->    recommended!]
+>   [\*That stage can be approximated by pausing the first erosion rates video around 99 ka and 13 ka. The three videos are strongly recommended!]
 
-Thank you for pointing this out. However, we are not confident that this result
-is significant. The situation you describe occurs, for instance, during much of
-MIS~5 (Fig. 2) and can be visualized in the animations (as you noted). During
-such periods, glaciers are steep and only a few grid cells in width.
+Thank you for pointing this out.
+However, we are not confident that this result is significant.
+The situation you describe occurs, for instance, during much of MIS~5 (Fig.
+2) and can be visualized in the animations (as you noted).
+During such periods, glaciers are steep and only a few grid cells in width.
 
-Increasing ice volume yields the build-up of less steep and thus slower-flowing
-valley glaciers. Decreasing ice volume leads to (roughly) equally steep but
-smaller glaciers. In the case of non-linear erosion laws ('Koppes' and 'Herman'
-laws), both yield a decrease of erosion volume, hence the local peak in erosion
-volume. However, we would rather not highlight this result. The modelled
-glacier velocities for such situations are unreliable due to the limiting
-horizontal resolution and the shallow-ice physics. This latter fact was
-highlighted in a new passage in the section on the "age of the glacial
-landscape". Also including suggestions from reviewer #1, the new passage
-reads as follow:
+Increasing ice volume yields the build-up of less steep and thus slower-flowing valley glaciers.
+Decreasing ice volume leads to (roughly) equally steep but smaller glaciers.
+In the case of non-linear erosion laws ('Koppes' and 'Herman' laws), both yield a decrease of erosion volume, hence the local peak in erosion volume.
+However, we would rather not highlight this result.
+The modelled glacier velocities for such situations are unreliable due to the limiting horizontal resolution and the shallow-ice physics.
+This latter fact was highlighted in a new passage in the section on the "age of the glacial landscape".
+Also including suggestions from reviewer #1, the new passage reads as follow:
 
-    The validity of the model results at high elevation is discussable.
-    Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
-    Ötztal and Tauern Massifs locally exhibit a strikingly high erosion
-    potential. However, the computation of glacier flow velocities on such
-    steep surfaces is strongly limited by the model horizontal resolution of
-    1\,km, the shallow-ice glacier flow physics (Imhof et al., 2019), and
-    PISM's current mass-conservation heuristics (Imhof, 2021). Besides,
-    bergschrund (rimaye) processes likely to dominate interglacial cirque
-    erosion at such altitudes (Sanders et al., 2012) are not captured by the
-    velocity-based glacier erosion power-laws.
+    "The validity of the model results at high elevation is discussable.
+Cristalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare, Ötztal and Tauern Massifs locally exhibit a strikingly high erosion potential.
+However, the computation of glacier flow velocities on such steep surfaces is strongly limited by the model horizontal resolution of 1\,km, the shallow-ice glacier flow physics (Imhof et al., 2019), and PISM's current mass-conservation heuristics (Imhof, 2021).
+Besides, bergschrund (rimaye) processes likely to dominate interglacial cirque erosion at such altitudes (Sanders et al., 2012) are not captured by the velocity-based glacier erosion power-laws."
 
-To avoid over-interpretation from future readers, this uncertainty was also
-highlighted in several of our plots (including Figs. 3 and 7) using hatches
-over periods of limited glacier volume, and documented in the figure captions.
+To avoid over-interpretation from future readers, this uncertainty was also highlighted in several of our plots (including Figs.
+3 and 7) using hatches over periods of limited glacier volume, and documented in the figure captions.
 
-> As the comment (lines 221-222) on erosion distribution during advance,
-> retreat and maxima is of great interest, it would be useful to illustrate
-> this with three maps of potential erosion during each of these types of
-> phase.
+> As the comment (lines 221-222) on erosion distribution during advance, retreat and maxima is of great interest, it would be useful to illustrate this with three maps of potential erosion during each of these types of phase.
 
-Thank you. Figure 4 was reworked to include a map of Rhine glacier modelled
-erosion rates at 36 ka, corresponding to the last major advance phase before
-the LGM and with a similar extent to the deglacial stage of 16 ka. This also
-corresponds to the topographic profiles shown on Fig. 8. Thus it seems more
-appropriate in hindsight.
+Thank you.
+Figure 4 was reworked to include a map of Rhine glacier modelled erosion rates at 36 ka, corresponding to the last major advance phase before the LGM and with a similar extent to the deglacial stage of 16 ka.
+This also corresponds to the topographic profiles shown on Fig.
+8.
+Thus it seems more appropriate in hindsight.
 
-> I find no mention of lithology, rock type, geology, resistance or erodibility
-> (excluding that citing Herman on line 56). This omission of half the erosion
-> equation should be mentioned; it implies that the ‘potential’ nature of the
-> erosion should be emphasised repeatedly. This adds to the admitted
-> limitations of lack of feedback to changing topography, absence of hydrology
-> and absence of subglacial sediment accumulation or movement. Using constant
-> topography might be OK for the last glaciation, as here, but would become
-> increasingly unrealistic for several glaciations. Destinations of eroded
-> material are not considered: deposition in temporary storage may inhibit
-> erosion in parts. The omission of hydrological processes must be a major
-> drawback, as Herman et al. (2011), Egholm et al. (2012a, b) and Pedersen et
-> al. (2014) showed that its inclusion of subglacial meltwater, plus horizontal
-> stress gradients, gave more realistic simulations of erosion. For erosion
-> rate as well as for cumulative erosion, it is potential erosion that is
-> calculated, rather than predicted erosion.
+> I find no mention of lithology, rock type, geology, resistance or erodibility (excluding that citing Herman on line 56).
+This omission of half the erosion equation should be mentioned; it implies that the ‘potential’ nature of the erosion should be emphasised repeatedly.
+This adds to the admitted limitations of lack of feedback to changing topography, absence of hydrology and absence of subglacial sediment accumulation or movement.
+Using constant topography might be OK for the last glaciation, as here, but would become increasingly unrealistic for several glaciations.
+Destinations of eroded material are not considered: deposition in temporary storage may inhibit erosion in parts.
+The omission of hydrological processes must be a major drawback, as Herman et al.
+(2011), Egholm et al.
+(2012a, b) and Pedersen et al.
+(2014) showed that its inclusion of subglacial meltwater, plus horizontal stress gradients, gave more realistic simulations of erosion.
+For erosion rate as well as for cumulative erosion, it is potential erosion that is calculated, rather than predicted erosion.
 
-The above limitations were made explicit with the following added sentences in
-the methods subsection on the "erosion law":
+The above limitations were made explicit with the following added sentences in the methods subsection on the "erosion law":
 
-    Instead, we assume that eroded material is instantly transported out of
-    the system, thus neglecting its role in shielding the bedrock from glacier
-    erosion in zones of temporary storage (Preusser et al., 2010). Neither do
-    we account for differences in erosion effectiveness on different
-    lithologies or erosion from subglacial and interglacial hydrologic
-    processes. For the aforementioned reasons, we refer to the above computed
-    rates, ė, as “potential erosion rates”.
+    "Instead, we assume that eroded material is instantly transported out of the system, thus neglecting its role in shielding the bedrock from glacier erosion in zones of temporary storage (Preusser et al., 2010).
+Neither do we account for differences in erosion effectiveness on different lithologies or erosion from subglacial and interglacial hydrologic processes.
+For the aforementioned reasons, we refer to the above computed rates, ė, as “potential erosion rates”."
 
-Second and as implied, the entire text and the figure labels were reworked to
-replace “erosion rates” with “potential erosion rates” nearly everywhere. To
-avoid clutter a few sentences still mention “modelled erosion rates” instead of
-“potential erosion rates”.
+Second and as implied, the entire text and the figure labels were reworked to replace “erosion rates” with “potential erosion rates” nearly everywhere.
+To avoid clutter a few sentences still mention “modelled erosion rates” instead of “potential erosion rates”.
 
-> A further necessary geographical qualification concerns the misfit between
-> simulated and actual glacier extent. The latter is well established from
-> mapping of moraines. It is necessary to check Seguinot et al. (2018) to see
-> that all the models predict too much ice in the southwest (south of 45° N)
-> and far too much in the east (Sava, Drava, Mur, Enns and Traun valleys),
-> while the EPICA models produce too little in the west (Rhône and Jura). I
-> agree that this is probably because while the amount of precipitation is
-> changed, the spatial pattern is not: thus the orographic effect of ice
-> build-up increasing precipitation shadow, and the circulation changes due to
-> southward shift of westerlies at LGM, are not included. To make this paper
-> more self-contained, a brief mention of the 2018 evaluation is in order.
-> Once a more realistic LGM precipitation pattern is accepted, future modelling
-> might use gradations between that and the present pattern as a function of
-> temperature.
+> A further necessary geographical qualification concerns the misfit between simulated and actual glacier extent.
+The latter is well established from mapping of moraines.
+It is necessary to check Seguinot et al.
+(2018) to see that all the models predict too much ice in the southwest (south of 45° N) and far too much in the east (Sava, Drava, Mur, Enns and Traun valleys), while the EPICA models produce too little in the west (Rhône and Jura).
+I agree that this is probably because while the amount of precipitation is changed, the spatial pattern is not: thus the orographic effect of ice build-up increasing precipitation shadow, and the circulation changes due to southward shift of westerlies at LGM, are not included.
+To make this paper more self-contained, a brief mention of the 2018 evaluation is in order.
+Once a more realistic LGM precipitation pattern is accepted, future modelling might use gradations between that and the present pattern as a function of temperature.
 
-The field-based LGM outline was added on Fig. 2 (Alps-wide map of cumulative
-erosion potential). In addition, the following sentences were added in the
-"climate sensitivity" section of the discussion mentioning the general
-eastwards bias of our results with regard to field evidence:
+The field-based LGM outline was added on Fig.
+2 (Alps-wide map of cumulative erosion potential).
+In addition, the following sentences were added in the "climate sensitivity" section of the discussion mentioning the general eastwards bias of our results with regard to field evidence:
 
-    It should be noted, however, that all runs presented here show a systematic
-    bias with excessive glacier cover in the Eastern Alps and a diminished
-    glacier extent in the Western Alps (Fig. 2a; further discussed in Seguinot
-    et al., 2018). Thus the modelled patterns of erosion potential certainly
-    includes a similar bias.
+    "It should be noted, however, that all runs presented here show a systematic bias with excessive glacier cover in the Eastern Alps and a diminished glacier extent in the Western Alps (Fig.
+2a; further discussed in Seguinot et al., 2018).
+Thus the modelled patterns of erosion potential certainly includes a similar bias."
 
 > Overall, the paper reports on a very worthwhile computer-intensive exercise.
-> It is densely packed with results and repays a careful read and a study of
-> the videos. Clearly there can be interesting comparisons with Lai and Anders
-> (2021) very recent submission to esurf !
+It is densely packed with results and repays a careful read and a study of the videos.
+Clearly there can be interesting comparisons with Lai and Anders (2021) very recent submission to esurf !
 
-Indeed, this is quite a coincidence! A reference was added to the paper by Lai
-and Anders (2021).
+Indeed, this is quite a coincidence! A reference was added to the paper by Lai and Anders (2021).
 
-> The text is short and pithy, so there is space for some extension. There are
-> a few missed opportunities, and some incompleteness in the presentation. In
-> detail I suggest the following improvements:
+> The text is short and pithy, so there is space for some extension. There are a few missed opportunities, and some incompleteness in the presentation. In detail I suggest the following improvements:
 
-The manuscript text has been extended in several places. Please see our
-responses to your specific comments below.
+The manuscript text has been extended in several places. Please see our responses to your specific comments below.
 
@@
 
-> The 1 km resolution of the modelling is clearly important, so it is strange
-> that the first mention is on p.11. Although it is very demanding of computer
-> time, it is still (as admitted) not adequate to represent cirques (averaging
-> of the order of 700 m across). Even stranger, there is no mention of the DEM
-> used (it was SRTM in the 2018 paper). This information should be provided in
-> section 2, p. 3-4.
+> The 1 km resolution of the modelling is clearly important, so it is strange that the first mention is on p.11.
+Although it is very demanding of computer time, it is still (as admitted) not adequate to represent cirques (averaging of the order of 700 m across).
+Even stranger, there is no mention of the DEM used (it was SRTM in the 2018 paper).
+This information should be provided in section 2, p.
+3-4.
 
-The model horizontal resolutions of 1 and (for climate sensitivity runs) 2 km
-is now mentioned in the methods. The input SRTM topography is also referenced
-in both the methods and the caption of Fig.~2 where it is displayed in the
-background. Following comments from reviewer #1, the discussion of model
-limitations regarding the representation of cirque glaciers has also been
-extended.
+The model horizontal resolutions of 1 and (for climate sensitivity runs) 2 km is now mentioned in the methods.
+The input SRTM topography is also referenced in both the methods and the caption of Fig.~2 where it is displayed in the background.
+Following comments from reviewer #1, the discussion of model limitations regarding the representation of cirque glaciers has also been extended.
 
-> The lateral constraints on valley glaciers are usually covered by a ‘form
-> factor’: how are they handled by this version of the PISM ice sheet model –
-> how has lateral drag been included?
+> The lateral constraints on valley glaciers are usually covered by a ‘form factor’: how are they handled by this version of the PISM ice sheet model – how has lateral drag been included?
 
-PISM uses a combination of shallow-ice and shallow-shelf stress balances,
-neither of which includes lateral stresses. Glacier sliding velocities in
-troughs are governed by (the sliding law and) longitudinal stresses, which is
-a limitation with respect to higher-order models. Following another comment
-from reviewer #1 an additional paragraph on basal sliding uncertainties was
-added, including the following sentence:
+PISM uses a combination of shallow-ice and shallow-shelf stress balances, neither of which includes lateral stresses.
+Glacier sliding velocities in troughs are governed by (the sliding law and) longitudinal stresses, which is a limitation with respect to higher-order models.
+Following another comment from reviewer #1 an additional paragraph on basal sliding uncertainties was added, including the following sentence:
 
-    Lateral stress gradients missing from the shallow-shelf approximation
-    stress balance could also contribute to moderate sliding velocity in narrow
-    troughs (Herman et al., 2011; Egholm et al., 2012a, b; Pedersen et al.,
-    2014).
+    "Lateral stress gradients missing from the shallow-shelf approximation stress balance could also contribute to moderate sliding velocity in narrow troughs (Herman et al., 2011; Egholm et al., 2012a, b; Pedersen et al., 2014)."
 
-> It might be pointed out that the ‘Koppes’ relationship with an exponent of
-> 2.34 is based on 13 data points that are very poorly distributed: in fact,
-> Koppes et al. (2015) go on to drop two outliers and reach an exponent of
-> 2.62.
+> It might be pointed out that the ‘Koppes’ relationship with an exponent of 2.34 is based on 13 data points that are very poorly distributed: in fact, Koppes et al.
+(2015) go on to drop two outliers and reach an exponent of 2.62.
 
-We clarified in the methods that we use Koppes et al. (2015) erosion law
-deriving from their full dataset. In addition, this sentence was added in the
-discussion of the "choice of erosion law":
+We clarified in the methods that we use Koppes et al. (2015) erosion law deriving from their full dataset. In addition, this sentence was added in the discussion of the "choice of erosion law":
 
-    On the other extreme, an even more non-linear erosion law, not tested here,
-    derived from tidewater glaciers but excluding two outliers (ė = 5.3×10^-9
-    u_b^2.62, Koppes et al., 2015) would result in an even more localized
-    pattern of erosion potential.
+    "On the other extreme, an even more non-linear erosion law, not tested here, derived from tidewater glaciers but excluding two outliers (ė = 5.3×10^-9 u_b^2.62, Koppes et al., 2015) would result in an even more localized pattern of erosion potential."
 
-> In lines 206-209, the point about trimlines, relating them to
-> time-transgressive erosion, is well taken. But they represent the integrated
-> effect of many phases of glacial erosion (as noted on 223-224), not
-> necessarily just from recessional phases.
+> In lines 206-209, the point about trimlines, relating them to time-transgressive erosion, is well taken.
+But they represent the integrated effect of many phases of glacial erosion (as noted on 223-224), not necessarily just from recessional phases.
 
-Indeed. The sentence was corrected to include "advancing and retreating
-glaciers" instead of "retreating glaciers".
+Indeed. The sentence was corrected to include "advancing and retreating glaciers" instead of "retreating glaciers".
 
@@
 
-> In Fig. 3 is almost impossible to follow the brown lines, unless advancing
-> and retreating parts of lines are distinguished (e.g. by colour). (The ice
-> volume video may help out.)  Also: keep to ‘potential erosion volume’. The
-> pale brown bars show a decline in annual erosion volume with ice volumes over
-> 1.8 cm s.l.e., stronger than is implied in line 110 of the text – it is more
-> than ‘slight’.
+> In Fig.
+3 is almost impossible to follow the brown lines, unless advancing and retreating parts of lines are distinguished (e.g.
+by colour).
+(The ice volume video may help out.)  Also: keep to ‘potential erosion volume’.
+The pale brown bars show a decline in annual erosion volume with ice volumes over 1.8 cm s.l.e., stronger than is implied in line 110 of the text – it is more than ‘slight’.
 
-Figure 3 was reworked to use different colours for periods of increasing and
-decreasing ice volume. This actually strikingly highlights the two different
-regimes, so thank you for the suggestion. The label was changed to 'potential
-annual erosion volume'. In the text, 'slight' was replaced with a 'general'
-tendency for slower erosion during periods of extensive glaciation.
+Figure 3 was reworked to use different colours for periods of increasing and decreasing ice volume.
+This actually strikingly highlights the two different regimes, so thank you for the suggestion.
+The label was changed to 'potential annual erosion volume'.
+In the text, 'slight' was replaced with a 'general' tendency for slower erosion during periods of extensive glaciation.
 
-> To aid interpretation of Figs 4 a-c and 7 a-d, a corresponding topographic
-> map or hillshade of the Rhine area on the same scale should be provided.
+> To aid interpretation of Figs 4 a-c and 7 a-d, a corresponding topographic map or hillshade of the Rhine area on the same scale should be provided.
 
-In addition to the aforementioned changes to Fig. 4, a new panel was added for
-the model final state at 0 ka, revealing the bedrock topography. So Fig. 4 now
-has one panel during advance (36 ka), one during a maximum stage (24 ka), one
-during retreat (16 ka), and one nearly ice-free (0 ka).
+In addition to the aforementioned changes to Fig.
+4, a new panel was added for the model final state at 0 ka, revealing the bedrock topography.
+So Fig.
+4 now has one panel during advance (36 ka), one during a maximum stage (24 ka), one during retreat (16 ka), and one nearly ice-free (0 ka).
 
-> Fig. 1c illustrates a very unusual cirque, with a huge lake replacing a very
-> unhealthy convex glacier. Fascinating, but surely a more representative
-> cirque should be used. Actually these three photos are not well integrated
-> with the text (pace lines 191 & 195-198). Lauterbrunnen is an extreme
-> (vertical-sided) trough example, over-used by textbooks.
+> Fig.
+1c illustrates a very unusual cirque, with a huge lake replacing a very unhealthy convex glacier.
+Fascinating, but surely a more representative cirque should be used.
+Actually these three photos are not well integrated with the text (pace lines 191 & 195-198).
+Lauterbrunnen is an extreme (vertical-sided) trough example, over-used by textbooks.
 
-The figure was moved into the discussion of the "age of the glacial landscape"
-to become Fig. 8, and is not longer referred to in the intro. The photo of
-Lauterbrunnental was replaced by one of Bout du Monde in the Giffre Massif, and
-the caption for the cirque photo now reads: "the unusually deep mountain cirque
-revealed by the current demise of Chüebodengletscher" (we don't have a much
-better photo for a high-elevation cirque, and glacial lakes from retreating
-cirque glaciers are a common occurrence in the Alps these days).
+The figure was moved into the discussion of the "age of the glacial landscape" to become Fig.
+8, and is not longer referred to in the intro.
+The photo of Lauterbrunnental was replaced by one of Bout du Monde in the Giffre Massif, and the caption for the cirque photo now reads: "the unusually deep mountain cirque revealed by the current demise of Chüebodengletscher" (we don't have a much better photo for a high-elevation cirque, and glacial lakes from retreating cirque glaciers are a common occurrence in the Alps these days).
 
-> Fig. 5: The elevation histogram in Fig. 5a could be misleading, as it seems
-> to cover the whole study area, including parts never glacier-covered. It
-> should be limited to areas glacier-covered at the maximum, and preferably
-> supplemented by a histogram for those covered now or at the minimum. The
-> accompanying main graph (5a) relates to areas glacier-covered at each time
-> slice. Re another reviewer’s comment, yes LGM glaciers did reach below 500 m:
-> at L. Garda (67 m?) and the Tagliamento Glacier lobe LGM ice reached below
-> 100 m. An integrated plot of total-cycle potential erosion rate (and
-> volume?) at each altitude, over glacier-covered areas, would be interesting
-> (as an extra Figure): can rates at high altitude, where glacial duration is
-> longer, catch up with those at low altitude?  Probably not, given the
-> concentration of flow down major valleys. Striping above 3000 m, presumably
-> related to low surface area, suggests a switch there to 20 m or 40 m bands
-> would be advisable. Caption; ‘modelled potential erosion rates’?
+> Fig.
+5: The elevation histogram in Fig.
+5a could be misleading, as it seems to cover the whole study area, including parts never glacier-covered.
+It should be limited to areas glacier-covered at the maximum, and preferably supplemented by a histogram for those covered now or at the minimum.
+The accompanying main graph (5a) relates to areas glacier-covered at each time slice.
+Re another reviewer’s comment, yes LGM glaciers did reach below 500 m: at L.
+Garda (67 m?) and the Tagliamento Glacier lobe LGM ice reached below 100 m.
+An integrated plot of total-cycle potential erosion rate (and volume?) at each altitude, over glacier-covered areas, would be interesting (as an extra Figure): can rates at high altitude, where glacial duration is longer, catch up with those at low altitude?
+Probably not, given the concentration of flow down major valleys.
+Striping above 3000 m, presumably related to low surface area, suggests a switch there to 20 m or 40 m bands would be advisable.
+Caption; ‘modelled potential erosion rates’?
 
-The figure now includes a histogram of model domain elevations, a histogram of
-glacier-covered elevations, and a plot of cumulative erosion potential volume
-per elevation bands of 100 m. The latter shows that, as you suspected, high
-altitude rapid erosion is largely offset by the limited number of grid cells
-it concerns. Instead, the bulk of the erosion potential occurs below 2000 m.
+The figure now includes a histogram of model domain elevations, a histogram of glacier-covered elevations, and a plot of cumulative erosion potential volume per elevation bands of 100 m.
+The latter shows that, as you suspected, high altitude rapid erosion is largely offset by the limited number of grid cells it concerns.
+Instead, the bulk of the erosion potential occurs below 2000 m.
 
-To highlight periods when low-elevation is relevant (and address the comment
-from reviewer #1), we also hatched elevation bands that contain fewer than a
-hundred ice-covered grid cells for a given time. We also shifted to 100-m
-elevation bands on the main panel. While some detail is lost, this avoids
-much of the striping (especially on the newly added contour and hatched
-pattern), and we find the new plot more readable.
+To highlight periods when low-elevation is relevant (and address the comment from reviewer #1), we also hatched elevation bands that contain fewer than a hundred ice-covered grid cells for a given time.
+We also shifted to 100-m elevation bands on the main panel.
+While some detail is lost, this avoids much of the striping (especially on the newly added contour and hatched pattern), and we find the new plot more readable.
 
-> In Fig. 7, the quantitative contrasts between the ‘Koppes’ law (e) and the
-> others are alarming. If the scales are taken literally, (e) gives potential
-> erosion around 1 m, while (f) and (g) give hundreds of m, in 120 ka. As 120
-> ka is only a fraction of the Quaternary, hundreds of m seems too much, while
-> 1 m seems very low (0.008 mm a-1 overall). For me, there should be a correct
-> answer between these extremes. The spatial pattern, however, is more
-> important.
+> In Fig.
+7, the quantitative contrasts between the ‘Koppes’ law (e) and the others are alarming.
+If the scales are taken literally, (e) gives potential erosion around 1 m, while (f) and (g) give hundreds of m, in 120 ka.
+As 120 ka is only a fraction of the Quaternary, hundreds of m seems too much, while 1 m seems very low (0.008 mm a-1 overall).
+For me, there should be a correct answer between these extremes.
+The spatial pattern, however, is more important.
 
-Discussion on the magnitude of modelled erosion rates remains somewhat
-speculative indeed due to the several sources of uncertainties, but the
-relevant discussion text was extended and reworked:
+Discussion on the magnitude of modelled erosion rates remains somewhat speculative indeed due to the several sources of uncertainties, but the relevant discussion text was extended and reworked:
 
-    With a total Pleistocene glacial relief on the order of a kilometre
-    (Preusser et al., 2011; Valla et al., 2011), a cumulative glacial erosion
-    for the last glacial cycle in the order of 10 to 100 m can be expected.
+    "With a total Pleistocene glacial relief on the order of a kilometre (Preusser et al., 2011; Valla et al., 2011), a cumulative glacial erosion for the last glacial cycle in the order of 10 to 100 m can be expected.
 However, none of the tested erosion power-laws fall within this range.
-    Instead, the erosion law calibrated on tidewater glaciers (Koppes et al.,
-    2015) yields cumulative erosion in the Rhine Valley in the orders of
-    metres, while the three erosion laws based on terrestrial glaciers
-    (Humphrey and Raymond, 1994; Herman et al., 2015; Cook et al., 2020},
-    result in kilometre-scale integrated erosion potential. During the Last
-    Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers
-    were closer in size, slope (an important parameter as we argue in the next
-    section), and climatic context to the present-day glaciers of Patagonia and
-    the Antarctic Peninsula (Koppes et al., 2015) than to Franz-Joseph Glacier
-    (Herman et al., 2015) and many of the glaciers included in the global
-    compilation by Cook et al., (2020). This may help to explain why the
-    reality appears to fall in-between the tested erosion laws.
+Instead, the erosion law calibrated on tidewater glaciers (Koppes et al., 2015) yields cumulative erosion in the Rhine Valley in the orders of metres, while the three erosion laws based on terrestrial glaciers (Humphrey and Raymond, 1994; Herman et al., 2015; Cook et al., 2020}, result in kilometre-scale integrated erosion potential.
+During the Last Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers were closer in size, slope (an important parameter as we argue in the next section), and climatic context to the present-day glaciers of Patagonia and the Antarctic Peninsula (Koppes et al., 2015) than to Franz-Joseph Glacier (Herman et al., 2015) and many of the glaciers included in the global compilation by Cook et al., (2020).
+This may help to explain why the reality appears to fall in-between the tested erosion laws."
 
-> Fig. 8a shows a flat dome at 3000 m for about the first 20 km of the Rhine.
-> As there are mountains at 3630, 3583 and 3192 m around the sources, it is
-> likely that   there would be steeper ice slopes up to these peaks, giving
-> rather different stresses and erosion potentials. Also are the instabilities
-> in Fig. 8b, near source and terminus, edge effects or related to the 1 km
-> resolution?
+> Fig.
+8a shows a flat dome at 3000 m for about the first 20 km of the Rhine.
+As there are mountains at 3630, 3583 and 3192 m around the sources, it is likely that   there would be steeper ice slopes up to these peaks, giving rather different stresses and erosion potentials.
+Also are the instabilities in Fig.
+8b, near source and terminus, edge effects or related to the 1 km resolution?
 
-This is correct. The "basal drag" quantity plotted on Fig. 8b is the magnitude
-of a two-component vector: the basal shear stress. The basal shear stress
-itself strongly depends on surface altitude gradients, but this includes both
-along-flow slopes visible on the topographic profile of Fig. 8a, and the
-invisible across-flow slopes. While the upper part of the topographic profile
-appears smooth, there are indeed steep (bedrock and surface) slopes on either
-side of the profile in this region. Because the upper part of the valley is
-narrow, the glacier centre-line fluctuating over time, and the plot
-interpolated between 1-km grid points, these lateral components enter the
-computation of the basal shear stress magnitude in the higher reaches.
+This is correct.
+The "basal drag" quantity plotted on Fig.
+8b is the magnitude of a two-component vector: the basal shear stress.
+The basal shear stress itself strongly depends on surface altitude gradients, but this includes both along-flow slopes visible on the topographic profile of Fig.
+8a, and the invisible across-flow slopes.
+While the upper part of the topographic profile appears smooth, there are indeed steep (bedrock and surface) slopes on either side of the profile in this region.
+Because the upper part of the valley is narrow, the glacier centre-line fluctuating over time, and the plot interpolated between 1-km grid points, these lateral components enter the computation of the basal shear stress magnitude in the higher reaches.
 
-Second, the activation of sliding also increases basal drag, which explains
-that basal drag also fluctuates together with the yield stress. This is
-particularly affecting the lower part of the profiles. Finally, normalization
-by overburden pressure amplifies all stress variations where ice thickness is
-small. But due to changes in glacier thickness, it would not be possible to
-compare different profiles without such normalization.
+Second, the activation of sliding also increases basal drag, which explains that basal drag also fluctuates together with the yield stress.
+This is particularly affecting the lower part of the profiles.
+Finally, normalization by overburden pressure amplifies all stress variations where ice thickness is small.
+But due to changes in glacier thickness, it would not be possible to compare different profiles without such normalization.
 
-While we would rather not include such details in the paper, we tried our best
-to summarise the effect with this additional sentence in the figure caption:
+While we would rather not include such details in the paper, we tried our best to summarise the effect with this additional sentence in the figure caption:
 
-    Some of the observed basal drag fluctuations in the upper part of the
-    transect result from steep slopes on either side of the narrow valley.
+    "Some of the observed basal drag fluctuations in the upper part of the transect result from steep slopes on either side of the narrow valley."
 
->   The three videos are well worth watching  -in fact they clarify some
->   queries arising from the paper. Perhaps they could be captioned in the
->   ‘Video supplement’ paragraph. Some extra information on the third (bedrock
->   altitude) video could point out that there are numerous zero values not
->   shown: the zero geometric mean rates from 2300 to 3000 m around 24 ka
->   otherwise seem to conflict with the numerous positive rates plotted. Zero
->   rates presumably show where ice is frozen to the bed: their altitudinal
->   distribution might be worth discussion in the text. Above 3000 m means are
->   positive, which seems strange.
+> The three videos are well worth watching  -in fact they clarify some queries arising from the paper.
+Perhaps they could be captioned in the ‘Video supplement’ paragraph.
+Some extra information on the third (bedrock altitude) video could point out that there are numerous zero values not shown: the zero geometric mean rates from 2300 to 3000 m around 24 ka otherwise seem to conflict with the numerous positive rates plotted.
+Zero rates presumably show where ice is frozen to the bed: their altitudinal distribution might be worth discussion in the text.
+Above 3000 m means are positive, which seems strange.
 
-Well, it seems that you have spend a lot of time looking at our figures and
-videos so let us thank you again for that. The caption of Fig. 5 mentioned "the
-geometric mean of (non-zero) modelled erosion rates". We found that the
-"non-zero" has no reason to be and it was removed. Sorry about the confusion
-and please read on for the long explanation.
+Well, it seems that you have spend a lot of time looking at our figures and videos so let us thank you again for that.
+The 'video supplement' paragraph was somewhat extended.
+The caption of Fig.
+5 mentioned "the geometric mean of (non-zero) modelled erosion rates".
+We found that the "non-zero" has no reason to be and it was removed.
+Sorry about the confusion and please read on for the long explanation.
 
-First, there are no zero values in the animation. As in the paper's Fig. 5, the
-animation shows geometric means (i.e. arithmetic means in log-domain), and
-geometric means can't contain zero values or they would result in a
-divide-by-zero error (or a log-of-zero error). Both animation panels have a
-log-scaled x-axis. While some values fall outside of the axis frame, they are
-very small but not zero.
+First, there are no zero values in the animation.
+As in the paper's Fig.
+5, the animation shows geometric means (i.e.
+arithmetic means in log-domain), and geometric means can't contain zero values or they would result in a divide-by-zero error (or a log-of-zero error).
+Both animation panels have a log-scaled x-axis.
+While some values fall outside of the axis frame, they are very small but not zero.
 
-On the other hand, both Fig. 5 and the animation depict very low values for
-erosion rates (especially around 24 ka). This is due to the glacier physics
-embedded in PISM. In particular, the pseudo-plastic sliding law implies that
-the glacier is sliding everywhere, but that sliding is infinitely small where
-the basal drag is much smaller than the yield stress, including for instance
-frozen-bed areas. These tiny sliding velocities yield even tinier erosion rates
-(due to the power 2.34 in the erosion law). Such values are not really relevant
-and they are off the colour range on Fig. 5 and off-chart in the
-animation (respectively lower than 10^-9 and 10^-10 mm a^-1).
+On the other hand, both Fig.
+5 and the animation depict very low values for erosion rates (especially around 24 ka).
+This is due to the glacier physics embedded in PISM.
+In particular, the pseudo-plastic sliding law implies that the glacier is sliding everywhere, but that sliding is infinitely small where the basal drag is much smaller than the yield stress, including for instance frozen-bed areas.
+These tiny sliding velocities yield even tinier erosion rates (due to the power 2.34 in the erosion law).
+Such values are not really relevant and they are off the colour range on Fig.
+5 and off-chart in the animation (respectively lower than 10^-9 and 10^-10 mm a^-1).
 
-The only regions where PISM produces truly zero sliding are the regions of zero
-surface slopes or zero ice thickness, where the SSA can't be solved. In
-practice, this only occurs in the latter case: zero ice thickness. This is the
-kind of values we have previously filtered out. Your comment led us to re-check
-our calculations. We found that there is no zero-sliding values within the
-ice-covered area, and thus an ice mask is sufficient to filter out undesirable
-values and compute the geometric means. Hence "non-zero" was removed from the
-caption.
+The only regions where PISM produces truly zero sliding are the regions of zero surface slopes or zero ice thickness, where the SSA can't be solved.
+In practice, this only occurs in the latter case: zero ice thickness.
+This is the kind of values we have previously filtered out.
+Your comment led us to re-check our calculations.
+We found that there is no zero-sliding values within the ice-covered area, and thus an ice mask is sufficient to filter out undesirable values and compute the geometric means.
+Hence "non-zero" was removed from the caption.
 
-To be sure, the non-zero filter has been removed from the script generating
-companion data (which is used in Fig. 5). This change can be seen here:
+To be sure, the non-zero filter has been removed from the script generating companion data (which is used in Fig. 5). This change can be seen here:
 
@@
 
-> I am grateful for helpful discussions with Iestyn Barr, Jeremy Ely, Matt
-> Tomkins, Pippa Whitehouse, Cristina Balaban and Matt Wiecek.
+> I am grateful for helpful discussions with Iestyn Barr, Jeremy Ely, Matt Tomkins, Pippa Whitehouse, Cristina Balaban and Matt Wiecek.
 
@@
 
-Thank you again to all of you for taking time to read our work and offer
-constructive feedback in these troubled times. We hope that you will find our
-answers satisfactory, apologize for the delay and will soon be submitting a
-revised and, we believe, improved manuscript including the aforementioned
-suggested changes.
+Thank you again to all of you for taking time to read our work and offer constructive feedback in these troubled times.
+We hope that you will find our answers satisfactory, apologize for the delay and will soon be submitting a revised and, we believe, improved manuscript including the aforementioned suggested changes.

### 2021-06-07 articles/erosion/alps-erosion.tex
Reread entire alps erosion manuscript.

@@
     localization of rapid potential erosion rates at low elevation, while
-    glacier erosion at higher elevation is modelled date from phases of less
+    glacier erosion at higher elevation is modelled to date from phases of less
     extensive glaciation. The modelled erosion rates peak during deglaciation
@@
     large-scale overdeepenings (e.g. Patagonia). But other regions, including
-    the Alps \citep[e.g.]{Penck.1905}, present a
+    the Alps \citep[e.g.][]{Penck.1905}, present a
     higher variety of glacier erosional landforms, whose implications on
@@
     Nevertheless, it remains elusive whether glacial topography is a proxy for
-    for temperate ice-cover duration, and why some glaciated regions show
+    temperate ice-cover duration, and why some glaciated regions show
     preserved periglacial blockfields topped by erratic boulders and others,
@@
     Coincidentally, the link between climate and erosion is also the topic of
-    another independent study using PISM \citep{Lai.Anders.2021}.
+    a recent, independent study using PISM \citep{Lai.Anders.2021}.
     Observed feedbacks between
@@
     1\,km, the shallow-ice glacier flow physics \citep{Imhof.etal.2019}, and
-    PISM's current mass-conservation heuristics \citep{Imhof.2021}. Besides,
+    PISM's current mass-conservation heuristics \citep[appendix]{Imhof.2021}.
+    Besides,
     bergschrund (rimaye) processes likely to dominate interglacial cirque
@@
     I.~Delaney thanks Fr\'ed\'eric Herman for valuable support and
-    conversations while at University of Lausanne.
-    An anonymous reviewer and Ian Evans (aided by Iestyn Barr, Jeremy Ely, Matt
-    Tomkins, Pippa Whitehouse, Cristina Balaban and Matt Wiecek) provided
-    constructive feedback which has been incorporated in the present paper.
+    conversations while at University of Lausanne. Both authors thank
+    an anonymous reviewer and Ian Evans (aided by Iestyn Barr, Jeremy Ely, Matt
+    Tomkins, Pippa Whitehouse, Cristina Balaban and Matt Wiecek) for providing
+    constructive feedback which has been incorporated in the present paper, and
+    the associate editor Andreas Lang for handling the reviews.
     The (previously published) ice flow simulations were made with computer

### 2021-06-07 articles/erosion/response-editor.md
Add short summary of changes for the editor.

<!-- Copyright (c) 2021, Julien Seguinot (juseg.github.io), Ian Delaney
Creative Commons Attribution-ShareAlike 4.0 International License
(CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/) -->

# Authors' response to the Editor

Dear Andreas Lang,

We apologize for the delays during the peer review of our manuscript. We believe that we have addressed all points raised by the reviews and hereby submit our revised version to *Earth Surface Dynamics*.

Please find hereafter a short summary and a marked-up file listing changes made to the manuscript since its first publication as a preprint. If possible, we ask you to please refer to our public responses for more detailed point-by-point listing of changes.

* **Sect. 2 (Methods):** We made explicit the input topography and temperature records used in the parent study (Seguinot et al., 2018), clarified simplifications made when computing erosion rates from basal velocities, and, accordingly, refer to the resulting quantities as "erosion potential" in the entire manuscript.

* **Sect. 4 (Discussion):** We added sentences and whole paragraphs to discuss the systematic eastwards bias of ice cover (and thus erosion potential), the magnitude of our computed erosion potential versus expectable glacial cycle erosion, the uncertainties on basal sliding, and model limitations in high mountains due to resolution and approximated physics.

* **Figures:** We added scale bars on all map figures, and hatched regions on Figs. 1, 2, 4 and 6 to indicate low confidence in our results for small glaciers. We added the field-based LGM outline on Fig. 1 (ex-Fig. 2), used different line styles to distinguish glacier advance and retreat on Fig. 2 (ex-Fig. 3), added glacier-advance (36 ka) and ice-free (0 ka) erosion maps of the Rhine Glacier on Fig. 3 (ex-Fig. 4), plotted glaciated hypsometry and cumulative erosion per elevation band on Fig. 4 (ex-Fig. 5), made style changes to improve readability of Figs. 6 and 7 (ex-Figs. 7 and 8), moved Fig. 8 (ex-Fig. 1) into the discussion and replaced one of the photos.

* **Text improvements** were made throughout the text and figure captions following suggestions from the reviewers.

We thank you again very much for your editorial work on our manuscript.

### 2021-06-10 articles/erosion/alps-erosion-supplement.tex
Add supplementary figures latex file.

--- 2021-06-01 articles/erosion/alps-erosion.tex
@@
-% Copyright (c) 2019-2021, Julien Seguinot (juseg.github.io), Ian Delaney
+% Copyright (c) 2021, Julien Seguinot (juseg.github.io), Ian Delaney
 % Creative Commons Attribution-ShareAlike 4.0 International License
@@
 
-% Alps erosion paper
-% ==================
+% Alps erosion paper supplement
+% =============================
 
-\documentclass[esurf, manuscript]{copernicus}
-\usepackage{CJK}  % Japanese text
-
-% custom colours (colorbrewer2.org)
-\definecolor{Bu}{cmyk}{1.00,0.45,0.00,0.07}  % Blues
-\definecolor{Gn}{cmyk}{1.00,0.20,1.00,0.00}  % Greens
-\definecolor{Rd}{cmyk}{0.35,0.95,0.85,0.00}  % Reds
-\definecolor{Or}{cmyk}{0.35,0.75,1.00,0.00}  % Oranges
-\definecolor{Pu}{cmyk}{0.70,0.80,0.00,0.00}  % Purples
-\definecolor{Br}{cmyk}{0.40,0.75,1.00,0.00}  % YlOrBr
-
-% custom commands (remove before submission)
-\newcommand{\note}[1]{\textcolor{Or}{\emph{[\textbf{NOTE:} #1]}}}
-\newcommand{\todo}[1]{\textcolor{Rd}{\emph{[\textbf{TODO:} #1]}}}
-
-% color links (remove before submission)
-\hypersetup{colorlinks, citecolor=Bu, linkcolor=Bu, urlcolor=Bu}
+\documentclass[esurf]{copernicus}
 
@@
 % document properties
-\title{Last glacial cycle glacier erosion potential in the Alps}
+\title{Last glacial cycle glacier erosion potential in the Alps
+       \large \newline Supplementary figures}
 \Author[1]{Julien}{Seguinot}
@@
 
+\nolinenumbers
+\onecolumn
+
     \maketitle
 
-\begin{abstract}
+    This document contains copies of the main text Figs. 1--5, and their
+    equivalent using the other tested erosion laws.
 
-    The glacial landscape of the Alps has fascinated generations of explorers,
-    artists, mountaineers and scientists with its diversity, including
-    erosional features of all scales from high-mountain cirques, to steep
-    glacial valleys and large over-deepened basins. Using previous glacier
-    modelling results, and empirical inferences of bedrock erosion under modern
-    glaciers, we compute a distribution of potential glacier erosion in the
-    Alps over the last glacial cycle from 120\,000 years ago to the present.
-    %
-    Despite large uncertainties pertaining to the climate history of the Alps
-    and unconstrained glacier erosion processes, the resulting modelled
-    patterns of glacier erosion include persistent features. The cumulative
-    imprint of the last glacial cycle shows a very strong localization of
-    erosion potential with local maxima at the mouths of major Alpine valleys
-    and some other upstream sections where glaciers are modelled to have flown
-    with the highest velocity. The potential erosion rates vary significantly
-    through the glacial cycle, but show paradoxically little relation to the
-    total glacier volume. Phases of glacier advance and maximum extension see a
-    localization of rapid potential erosion rates at low elevation, while
-    glacier erosion at higher elevation is modelled date from phases of less
-    extensive glaciation. The modelled erosion rates peak during deglaciation
-    phases, when frontal retreat results in steeper glacier surface slopes,
-    implying that climatic conditions that result in rapid glacier erosion
-    might be quite transient and specific.
-    %
-    Our results depict the Alpine glacier erosion landscape as a
-    time-transgressive patchwork, with different parts of the range
-    corresponding to different glaciation stages and time periods.
-
-\end{abstract}
-
-% ----------------------------------------------------------------------
-\introduction
-% ----------------------------------------------------------------------
-
-    The glacial erosion landscape of the Alps has fascinated generations of
-    explorers, artists, mountaineers and scientists for centuries. Its cultural
-    impact is indeed so far-reaching, that in English, a non-Alpine language,
-    the adjective ``alpine'' with non-capital ``a'' is now casually used to
-    describe an Alpine-like, glacially modified mountain landscape outside the
-    Alps, while the proper noun ``Alps'' has been applied to nick-name
-    Alpine-like, glacier eroded mountain ranges in Norway (\emph{Lyngsalpene}),
-    New Zealand (Southern Alps), Japan
-    (\emph{\begin{CJK}{UTF8}{min}日本アルプス\end{CJK},
-    Nihon Arupusu}), and elsewhere.
-    %
-    Some mountain ranges are predominantly characterised by cirque glaciation
-    (e.g. Uinta Mountains), glacial valleys (e.g. Putorana Plateau), or
-    large-scale overdeepenings (e.g. Patagonia). But other regions, including
-    the Alps \citep[e.g.]{Penck.1905}, present a
-    higher variety of glacier erosional landforms, whose implications on
-    glacial history are yet to be understood. In parallel to such landscape
-    diversity, the cosmogenic nuclide memory of bedrock erosion provides a more
-    quantitative, but equally varied picture of glacier erosion effectiveness
-    both within and between glaciated regions \citep{Jansen.etal.2019,
-    Steinemann.etal.2020, Steinemann.etal.2021}.
-
-    Glacially eroded topography has sometimes been used as a proxy for mapping
-    paleoglacier extent \citep[e.g.,][]{Margold.etal.2011, Fu.etal.2012}.
-    Cold-based glaciers, however, have been observed to preserve landforms as
-    fragile as sand beaches \citep{Kleman.1994}, leading to the
-    reinterpretation of complex glacial landscapes within a binary conceptual
-    framework of cold and temperate basal thermal regimes
-    \citep{Kleman.etal.2008, Kleman.etal.2010, Fabel.etal.2012, Fu.etal.2013}.
-    Nevertheless, it remains elusive whether glacial topography is a proxy for
-    for temperate ice-cover duration, and why some glaciated regions show
-    preserved periglacial blockfields topped by erratic boulders and others,
-    including the Alps, do not \citep{Wirsig.etal.2016, Seguinot.etal.2018}.
-
-    On a larger scale, glacier erosion also likely governs, at least in part,
-    the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through
-    the ``glacier buzz-saw''. In turn, evidence suggests that increased
-    mountain erosion rates coincided with global cooling and Pleistocene
-    glaciations from roughly 2.6 million years before present
-    \citep{Herman.Champagnac.2016}. However, oceanic isotopic proxies for
-    global weathering rates remained constant over this time period
-    \citep{Willenbring.Von-Blanckenburg.2010}. Examination of contemporary
-    erosion rates across different climates suggests that increased temperatures
-    within glaciated regions lead to higher erosion rates
-    \citep{Koppes.Montgomery.2009, Koppes.etal.2015, Fernandez.etal.2016}.
-    However, understanding the link
-    between climate and glacier erosion in the sedimentary record remains
-    difficult due to time scale biases \citep{Ganti.etal.2016} and the
-    non-linear relationship between atmospheric temperature and erosion
-    \citep[e.g.,][]{Anderson.etal.2012, Mariotti.etal.2021}.
-    Numerical glacier modelling may provide insights into such complexity.
-
-    Observing and quantifying the long-term glacier erosion and sedimentation
-    processes has been a challenge, and two general methods have been adopted
-    to quantify erosion \citep{Alley.etal.2019}. Physically-based models
-    describe quarrying process and abrasion of bedrock by debris-laden ice
-    \citep[e.g.,][]{Alley.etal.1997, Iverson.2012, Beaud.etal.2014}. Describing
-    the processes physically provides a basis for understanding erosion
-    processes \citep{Hallet.1979, Ugelvig.etal.2018} and yields insight into
-    the formation of glacial landforms, such as tunnel valleys and eskers
-    \citep{Beaud.etal.2018, Hewitt.Creyts.2019}. Yet, these models prove
-    difficult to implement in many cases due to the large number of poorly
-    constrained parameters and processes, i.e. water pressure fluctuations and
-    subglacial debris concentration \citep[e.g.,][]{Hallet.1979, Seguinot.2008,
-    Ugelvig.etal.2018}.
-
-    Instead, empirical relationships between glacier sliding and erosion can
-    represent erosional quantities well and can recreate important glacial
-    landforms \citep[e.g.,][]{Harbor.etal.1988, Macgregor.etal.2000}.
-    For instance, \citet{Humphrey.Raymond.1994} correlated temporal variations
-    in ice velocity and suspended sediment load during a surge of Variegated
-    Glacier to establish a linear erosion law.
-    \citet{Koppes.etal.2015} quantified sediment
-    yields in 15~Patagonian and Antarctic Peninsula fjords, concluding at a
-    predominant control of surface air temperature on rapid glacier erosion and
-    calibrating a near-square-law from glacier velocity to erosion rate.
-    \citet{Herman.etal.2015} collected suspended sediment samples for 5~months
-    in the outlet stream of Franz-Josef Glacier, mapped their chemical
-    composition to geologic zones of distinctive glacier speed, and also
-    concluded to a near-square relationship between basal sliding and glacier
-    erosion.
-    \citet{Cook.etal.2020} assembled a global compilation of erosion rates for
-    38~glaciers, showing a predominant role of glacier sliding velocity over
-    climate variables, yet concluding at a sub-linear relationship to erosion.
-
-    Here, the non-linear erosion law by \citet{Koppes.etal.2015} is applied to
-    previously published model results \citep{Seguinot.etal.2018} and the
-    patterns of modelled erosion potential and cumulative last glacial cycle
-    erosion potential are analysed. We examine the glaciological conditions
-    that cause erosion and discuss the implications of these conditions on
-    understanding the mark of ancient glaciers on Alpine topography and
-    the relationship between climate and glacier erosion. Despite aggregated
-    uncertainties on paleoclimate, glacier flow and glacier erosion processes,
-    out results provide insights into the diversity of the Alpine glacial
-    erosion landscape.
-
-% ----------------------------------------------------------------------
-\section{Methods}
-% ----------------------------------------------------------------------
+    \listoffigures
 
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Ice sheet modelling}
+% Figures
 % -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
 
-    The ice-sheet model set-up was presented in an earlier publication
-    \citep{Seguinot.etal.2018} and is briefly summarized here. The simulations
-    were conducted using the Parallel Ice Sheet Model (PISM, development
-    version~e9d2d1f), an open-source, finite-difference, shallow-ice and
-    shallow-shelf model \citep{PISM-authors.2017}. Our setup includes
-    temperature and water-content dependent creep, pseudo-plastic basal
-    sliding, bedrock isostatic deformation under the ice
-    load, and a~positive degree-day (PDD) surface mass balance model. The model
-    is initialized with assumed present-day ice thickness and 1-km resolution
-    bedrock topography from SRTM \citep{Jarvis.etal.2008},
-    ice and bedrock temperature at 120\,ka (thousand years before the present),
-    and is run to the present. The ice-sheet model physical parameters are listed
-    in the earlier open-access publication \citep{Seguinot.etal.2018}, and the
-    complete list of PISM
-    parameters is stored in long-term archived model output metadata for the
-    present study \citep{Seguinot.2021}, and for the earlier publication
-    \citep{Seguinot.2020, Seguinot.2020a}.
-
-    Climate forcing is provided by a~monthly climatology from interpolated
-    observational data \citep[WorldClim;][]{Hijmans.etal.2005} and the European
-    Centre for Medium-Range Weather Forecasts Reanalysis Interim
-    \citep[ERA-Interim;][]{Dee.etal.2011}, amended with temperature changes
-    from the European Project for Ice Coring in Antarctica
-    \citep[EPICA;][] {Jouzel.etal.2007}, and time-dependent paleo-precipitation
-    reductions \citep{Huybrechts.2002}. Coarser, 2-km resolution runs based on
-    alternative paleoclimate scenarios \citep{Seguinot.etal.2018}, forced by
-    temperature changes from the the Greenland Ice Core Project
-    \citep[GRIP;][]{Dansgaard.etal.1993} and and an oceanic sediment core from
-    the Iberian margin \citep[MD01-2444;][]{Martrat.etal.2007}, are analyzed
-    in the discussion part (Sect.~\ref{sec:sensitivity}).
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Erosion law}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    Modelled erosion rates, $\dot{e}$, are calculated from the modelled basal
-    velocities, $u_\mathrm{b}$, using a empirical erosion power-law,
-    %
-    \begin{equation}
-        \dot{e} = K_\mathrm{g} u_\mathrm{b}^l ,
-    \end{equation}
-    %
-    where $K_g = 5.2\times 10^{-11}\,m^{1-l}\,a^{l-1}$ and $l = 2.34$ are
-    empirical constants calibrated on quantified fjord sediment yields and
-    velocities from 15~Patagonian and Antarctic Peninsula tidewater glaciers
-    \citep[using the full dataset from][including outliers]{Koppes.etal.2015}.
-    We discuss the role of alternative values of
-    $K_\mathrm{g}$ and $l$ in the discussion part (Sect.~\ref{sec:powerlaws}).
-
-    Our approach does not account for feedbacks of glacier erosion onto
-    bedrock topography and, then onto, ice dynamics
-    \citep[e.g.,][]{Anderson.etal.2012}. Furthermore, we do not account for the
-    downstream advection of debris \citep[e.g.,][]{Anderson.Anderson.2016},
-    or fluvial transport of subglacial sediment \citep{Delaney.etal.2019}.
-    Instead, we assume that eroded material is instantly transported out of the
-    system, thus neglecting its role in shielding the bedrock from glacier
-    erosion in zones of temporary storage \citep{Preusser.etal.2010}. Neither
-    do we account for differences in erosion effectiveness on different
-    lithologies or erosion from subglacial and interglacial hydrologic
-    processes. For the aforementioned reasons, we refer to the above computed
-    rates, $\dot{e}$, as ``potential erosion rates''.
-
-    Secondly, we compute the time-integrated glacier erosion rate,
-    %
-    \begin{equation}
-        e =  K_\mathrm{g} \int u_\mathrm{b}^l dt,
-    \end{equation}
-    %
-    and refer to it as the ``cumulative erosion potential''. The integration
-    is numerically approximated using a time-step of 10\,a in the
-    main simulation, and 100\,a in the alternative climate scenarios runs.
-    Note that in the above formula with an exponent, $l$,
-    higher than one, the cumulative erosion potential has a greater dependency
-    on the basal ice dynamics, $u_\mathrm{b}$, than the duration of glacier
-    cover.
-
-% ----------------------------------------------------------------------
-\section{Results}
-% ----------------------------------------------------------------------
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Cumulative erosion}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
+    \renewcommand\thefigure{S\arabic{figure}}
 
     \begin{figure*}
-      \centerline{\includegraphics{alpero_cumulative}}
-      \caption{%
+      \includegraphics{alpero_cumulative}
+      \caption[%
+        Cumulative erosion potential and annual erosion volume after
+        \citet{Koppes.etal.2015}, same as main text Fig. 1.
+        ]{%
         \textbf{(a)} Modelled cumulative (time-integrated) glacial erosion
@@
           below 3\,cm~s.l.e. where glacier sliding may be affected by
-          stress-balance approximations and model horizontal resolution (see
-          Sect.~\ref{sec:landscape}).}
+          stress-balance approximations and model horizontal resolution.}
         \label{fig:cumulative}
@@
 
-    The modelled cumulative (time-integrated) glacial erosion potential
-    (Fig.~\ref{fig:cumulative}a) varies by several orders of magnitude
-    from insignificant to hundred-metres-scale erosion potential. Its spatial
-    patterns show a very strong localization along the Alpine valleys, with
-    local maxima occurring both at
-    the Alpine gates where ice flow transited from valley to piedmont glaciers,
-    and further upstream where the valley slopes increase. While mountain
-    cirque glaciers and relevant erosion processes may not be captured by the
-    model physics and horizontal resolution, high
-    cumulative erosion potential also occurs near the valley heads.
-    There is a general tendency for higher cumulative erosion in the
-    north-western Alps where the input winter precipitation is higher
-    \citep[WorldClim, Fig.~1h in][]{Seguinot.etal.2018} and the glacial relief
-    more pronounced in the topography.
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Temporal evolution}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    \begin{figure}
-      \centerline{\includegraphics{alpero_evolution}}
+    \begin{figure*}
+      \includegraphics{alpero_cumulative_her2015}
       \caption{%
-        Modelled annual (domain-integrated) potential erosion volume (light
-        curves) and its 1-ka running mean (dark curves) in relation to the
-        modelled total ice volume in centimetres of sea-level equivalent. Blue
-        dashes indicates increasing ice volume and solid brown decreasing ice
-        volume. Hatches indicate low confidence due to limited glacier volume.}
-      \label{fig:evolution}
-    \end{figure}
-
-    The modelled annual (domain-integrated) potential glacial erosion volume
-    does not systematically correlate with the modelled total ice volume
-    (Fig.~\ref{fig:cumulative}b). Except for the onset and termination of the
-    glacial cycle were ice volume is low and sliding processes unlikely to be
-    captured by the model resolution, a comparison between the modelled ice
-    volume and the modelled annual erosion volume shows no single relation
-    between these two quantities (Fig.~\ref{fig:evolution}). After significant
-    ice volume has been reached, there is a general tendency for slower erosion
-    during periods of extensive glaciation (Fig.~\ref{fig:evolution}). However,
-    the domain-integrated erosion volume is modelled to be consistently higher
-    (by a factor 2 to 10) during periods of decreasing ice volume, than during
-    periods of increasing ice volume (Fig.~\ref{fig:evolution}).
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Spatial migration}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpero_transects}}
-      \caption{%
-        \textbf{(a--d)} Modelled instantaneous potential erosion rate of the
-          Rhine Glacier for selected glacier advance and retreat ages, and the
-          final model state for topographic reference.
-        \textbf{(e)} Interpolated instantaneous potential erosion rate along a
-          Rhine Glacier transect (upper panels dashed line) for the entire last
-          glacial cycle.}
-      \label{fig:transects}
+        Cumulative erosion potential and annual erosion volume after
+        \citet{Herman.etal.2015}.}
     \end{figure*}
@@
     \begin{figure*}
-      \centerline{\includegraphics{alpero_hypsogram}}
+      \includegraphics{alpero_cumulative_hum1994}
       \caption{%
-        \textbf{(a)} Potential erosion rate ``hypsogram'', showing the
-          geometric mean of modelled rates in 100-m elevation bands
-          across the entire model domain and its time evolution. Hatches
-          indicate elevation bands with fewer than a hundred grid cells
-          (100\,$km^2$).
-        \textbf{(b)} Distribution of model domain bedrock topography (grey
-          bars) and glaciated topography (blue bars) in 100-m elevation bands,
-          and the corresponding cumulative potential erosion volume (brown
-          line).
-        \textbf{(c)} Same as Fig.~\ref{fig:cumulative}b.}
-      \label{fig:hypsogram}
-
+        Cumulative erosion potential and annual erosion volume after
+        \citet{Humphrey.Raymond.1994}}
     \end{figure*}
 
-    A closer look at the Rhine Glacier, one of the paleo-ice sheet's largest
-    outlets, reveals a spatial migration of the modelled rapid erosion areas.
-    During stages of glacier advance and maximum extension, erosion is modelled
-    to be under 1\,mm\,a$^{-1}$ and restricted the lower parts of the
-    catchment, while much of the intra-montane Alps \citep[modelled to be
-    largely cold-based, Fig.~6c in][]{Seguinot.etal.2018} experience
-    insignificant erosion potential (Fig.~\ref{fig:transects}a and~b). The
-    potential erosion rates both increase and propagate inwards during periods
-    of glacier retreat (Fig.~\ref{fig:transects}c and~e).
-
-    The results found on the Rhine Glacier catchment can be generalized to the
-    entire model domain by using (present-day) bedrock altitude as a proxy for
-    along-flow distance (Fig.~\ref{fig:hypsogram}). Thus more generally,
-    periods of modelled increasing and maximum ice volume correspond to lower
-    values for elevation-aggregated potential erosion rates, with significant
-    erosion potential
-    restricted to lower elevations. On the opposite, periods of modelled
-    decreased ice volume correspond to higher local modelled erosion rates and
-    more extensive rapid erosion potential into higher-elevation areas
-    (Fig.~\ref{fig:hypsogram}).
-
-% ----------------------------------------------------------------------
-\section{Discussion}
-% ----------------------------------------------------------------------
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Climate sensitivity}
-\label{sec:sensitivity}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
     \begin{figure*}
-      \centerline{\includegraphics{alpero_sensitivity}}
+      \includegraphics{alpero_cumulative_coo2020}
       \caption{%
-        Modelled cumulative glacial erosion potential over the last glacial
-        cycle using three different paleo-temperature histories from
-        \textbf{(a, d)} the Greenland Ice Core Project
-        \citep[GRIP;][]{Dansgaard.etal.1993}, \textbf{(b, e)} the European
-        Project for Ice Coring in Antarctica \citep[our default,
-        EPICA;][]{Jouzel.etal.2007}, and \textbf{(c, f)} an oceanic sediment
-        core from the Iberian margin \citep[MD01-2444;][]{Martrat.etal.2007},
-        and both \textbf{(a--c)} without and \textbf{(d--f)} with
-        paleo-precipitation reductions \citep[cf.][]{Seguinot.etal.2018}.}
-      \label{fig:sensitivity}
+        Cumulative erosion potential and annual erosion volume after
+        \citet{Cook.etal.2020}}
     \end{figure*}
-
-    Different paleoclimate forcing scenarios lead to different modelled
-    glaciation histories \citep[cf.][]{Seguinot.etal.2018} and thus different
-    results in terms of glacial cycle cumulative erosion
-    potential (Fig.~\ref{fig:sensitivity}). Scenarios including higher
-    (constant, present-day) input precipitation rates yield generally higher
-    modelled ice discharge and thus higher
-    erosion potential (Fig.~\ref{fig:sensitivity}a--c), while scenarios
-    including paleo-precipitation reductions (Fig.~\ref{fig:sensitivity}d--f),
-    including our default high-resolution run
-    (Fig.~\ref{fig:cumulative}), yield a smaller erosion potential.
-
-    The spatial pattern of modelled time-integrated erosion is nevertheless
-    generally consistent across the paleoclimate scenarios tested,
-    despite variations in the glaciation maximum extent
-    (\citealp[Fig.~3 in][]{Seguinot.etal.2018}; Fig.~\ref{fig:sensitivity}).
-    It should be noted, however, that all runs presented here show a systematic
-    bias with excessive glacier cover in the Eastern Alps and a diminished
-    glacier extent in the Western Alps (Fig.~\ref{fig:cumulative}a; further
-    discussed in \citealp{Seguinot.etal.2018}). Thus the modelled patterns of
-    erosion potential certainly includes a similar bias. However, uncertainties
-    due to paleoclimate appear minor in regard to the much larger uncertainties
-    affecting the choice of erosion law.
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Choice of erosion law}
-\label{sec:powerlaws}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    \begin{figure*}[t]
-      \centerline{\includegraphics{alpero_powerlaws}}
-      \caption{%
-        Modelled cumulative (time-integrated) glacial erosion potential for
-        three different erosion laws published by
-        \textbf{(a)} \citet{Koppes.etal.2015},
-          ${5.2 \times 10^{-8} u_\mathrm{b}^{2.34}}$ (same as
-          Fig.~\ref{fig:cumulative}a but with a different colour scale),
-        \textbf{(b)} \citet{Herman.etal.2015},
-          ${2.7 \times 10^{-7} u_\mathrm{b}^{2.02}}$,
-        \textbf{(c)} \citet{Humphrey.Raymond.1994},
-          ${1 \times 10^{-4} u_\mathrm{b}^{1}}$, and
-        \textbf{(d)} \citet{Cook.etal.2020},
-          ${1.665 \times 10^{-1} u_\mathrm{b}^{0.6459}}$.
-        \textbf{(e-h)} Corresponding modelled cumulative glacial erosion
-          potential along a Rhine Glacier transect (upper panels dashed line), and
-        \textbf{(i-l)} corresponding modelled evolution of annual
-        (domain-integrated) erosion volume (light brown) and 1-ka
-        running mean (dark brown) in relation to modelled total ice volume in
-        centimetres of sea-level equivalent (as in Fig.~\ref{fig:evolution}a).}
-      \label{fig:powerlaws}
-    \end{figure*}
-
-    The choice of erosion law significantly impacts the results
-    (Fig.~\ref{fig:powerlaws}). Our default, based on quantified
-    sediment yields from Patagonian and Antarctic Peninsula tidewater glaciers
-    \citep[${\dot{e}=5.2\times10^{-8}u_\mathrm{b}^{2.34}}$,][]{Koppes.etal.2015}
-    yields a moderate and strongly localized cumulative erosion potential
-    (Figs~\ref{fig:cumulative} and~\ref{fig:powerlaws}a). The erosion law
-    based on measured suspended sediment load from Franz-Joseph Glacier
-    \citep[${\dot{e}=2.7\times10^{-7}u_\mathrm{b}^{2.02}}$,][]{Herman.etal.2015}
-    also yields a strongly localized, and yet much higher, modelled erosion
-    potential (Fig.~\ref{fig:powerlaws}b). The erosion law based on measured
-    suspended sediment load during a surge of Variegated Glacier
-    \citep[${\dot{e}=1\times10^{-4}u_\mathrm{b}^{1}}$,][]{Humphrey.Raymond.1994}
-    results in similarly high values, but a less localized pattern of cumulative
-    erosion potential (Fig.~\ref{fig:powerlaws}c). Finally, The erosion law
-    based on a global compilation of glacier velocities and erosion rates
-    \citep[${\dot{e}=1.665\times10^{-1}u_\mathrm{b}^{0.6459}}$,][]{Cook.etal.2020}
-    yields an even flatter pattern of cumulative erosion potential
-    (Fig.~\ref{fig:powerlaws}d).
-
-    With a total Pleistocene glacial relief on the order of a kilometre
-    \citep{Preusser.etal.2011, Valla.etal.2011}, a cumulative glacial erosion
-    for the last glacial cycle in the order of 10 to 100\,m can be expected.
-    However, none of the tested erosion power-laws fall within this range.
-    Instead, the erosion law calibrated on tidewater glaciers
-    \citep{Koppes.etal.2015} yields cumulative erosion in the Rhine Valley in
-    the orders of metres, while the three erosion laws based on terrestrial
-    glaciers \citep{Humphrey.Raymond.1994, Herman.etal.2015, Cook.etal.2020},
-    result in kilometre-scale integrated erosion potential. During the Last
-    Glacial Maximum and much of the last glacial cycle, Alpine paleoglaciers
-    were closer in size, slope (an important parameter as we argue in the next
-    section), and climatic context to the present-day glaciers of
-    Patagonia and the Antarctic Peninsula \citep{Koppes.etal.2015} than to
-    Franz-Joseph Glacier \citep{Herman.etal.2015} and many of the glaciers
-    included in the global compilation by \citet{Cook.etal.2020}.
-    This may help to explain why the reality appears to fall in-between the
-    tested erosion laws.
-
-    That said, the magnitude of the modelled erosion rates can hardly be used
-    as a criteria to rule out this or that erosion law for the glacial Alps.
-    While sliding is here modelled by a pseudo-plastic law based on
-    till dilatation under high water pressure \citep{Tulaczyk.etal.2000},
-    empirical erosion rules are thought to be a proxy for glacier abrasion
-    (cf.~\citealp{Hallet.1979}; and possibly plucking) of a glacier flowing
-    over a hard bed. Current erosion rates in the Alps strongly vary depending
-    on the substratum \citep{Steinemann.etal.2020, Steinemann.etal.2021}, and
-    it is arguable that a layer of till and proglacial sediments may have
-    armored the glacier bed from erosion for at least parts of the glacial
-    cycle. Thus, given the lack of data to constrain the erosion rule, it can
-    not be excluded that a linear or sub-linear erosion law
-    \citep[e.g.,][]{Cook.etal.2020} would lead to a less localized pattern of
-    erosion (Fig.~\ref{fig:powerlaws}c and~d), and a stronger correlation
-    between modelled ice volume and erosion (Fig.~\ref{fig:powerlaws}k and~l).
-    On the other extreme, an even more non-linear erosion law, not tested here,
-    derived from tidewater glaciers but excluding two outliers
-    \citep[${\dot{e}=5.3\times10^{-9}u_\mathrm{b}^{2.62}}$,][]{Koppes.etal.2015}
-    would result in an even more localized pattern of erosion potential.
-
-    The modelled pattern of erosion potential depends on PISM glacier physics
-    and sliding model parameters. In particular, the pseudo-plastic sliding law
-    exponent \citep[$q=0.25$ in][]{Seguinot.etal.2018} controls the sharpness
-    of the transition between adherent and decoupled basal conditions. Recent
-    ensemble validation of Antarctic Ice Sheet glacial-cycle simulations
-    against geological and present-day observations \citep{Albrecht.etal.2020,
-    Albrecht.etal.2020a} support a higher value of $q=0.75$, and thus a
-    sliding law closer to linear, which would perhaps result in a smoother
-    distribution of sliding velocity and erosion potential. Lateral stress
-    gradients missing from the shallow-shelf approximation stress balance could
-    also contribute to moderate sliding velocity in narrow troughs
-    \citep{Herman.etal.2011, Egholm.etal.2012, Egholm.etal.2012a,
-    Pedersen.etal.2014}.
-
-    Because of the intrinsically large variations in modelled glacier sliding
-    velocities yield cumulative glacial-cycle erosion two to three orders of
-    magnitude higher in the Alpine valleys than near the mountain tops
-    (Fig.~\ref{fig:powerlaws}d), we argue that the modelled highly localized
-    pattern of erosion potential is qualitatively robust. This pattern, and
-    the lithologic variations in bedrock erodability, may explain
-    order-of-magnitude variation of glacier erosion rates observed in the
-    cosmogenic nuclide record \citep{Jansen.etal.2019, Steinemann.etal.2020,
-    Steinemann.etal.2021}.
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Climate control on erosion}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    \begin{figure}
-      \centerline{\includegraphics{alpero_basaldrag}}
-      \caption{%
-        \textbf{(a)} Modelled basal and surface topographies interpolated along
-        the Rhine Glacier during glacier advance (36\,ka, dashed blue), maximum
-        extent (24\,ka, solid grey) and retreat (16\,ka, solid brown, using the
-        same transect as in Fig.~\ref{fig:basaldrag}).
-        \textbf{(b)} Interpolated yield stress in the pseudo-plastic sliding law
-        (dotted lines) and magnitude of the basal shear stress (basal drag,
-        solid lines), both normalized against the ice overburden pressure.
-        Basal drag values greater than yield stress result in glacier sliding
-        \citep[cf.][for model physics]{Seguinot.2014, Seguinot.etal.2016}.
-        Some of the observed basal drag fluctuations in the upper part of the
-        transect result from steep slopes on either side of the narrow valley.}
-      \label{fig:basaldrag}
-    \end{figure}
-
-    The modelled, domain-integrated glacier erosion volume does not increase
-    together with ice volume. On the opposite, periods of ice advance and
-    maximum glaciation correspond to comparatively slow erosion potential
-    (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}).
-    This result may seem paradoxical, and is modulated in a limited extent by
-    the exponent, $l$, in the erosion rule (Fig.~\ref{fig:powerlaws}).
-    Nevertheless, it may explain why some areas covered during glacial maxima
-    only, such as some valleys on the southern side of the Alps, appear to have
-    experienced only little glacial modification of their topography.
-    During deglaciation periods, increased surface melt at low elevations
-    results in a steepened topographic profile (Figs.~\ref{fig:basaldrag}a),
-    as observed and further expected on currently retreating glaciers
-    \citep{Huss.etal.2010, Zekollari.Huybrechts.2015}.
-    The buttress formed by the glacier foot against gravitational forces is
-    reduced, causing an increase in the magnitude of the basal shear stress
-    (basal drag) relative to the ice overburden pressure, and an increase in
-    the area where sliding and thus, erosion, can occur
-    (Figs.~\ref{fig:basaldrag}b).
-
-    This result is corroborated by a recent denudation record from the
-    Mediterranean Alps, which includes an increase in glacier erosion in the
-    Var Valley, roughly at 25\,ka following an extended period of low
-    denudation rates \citep{Mariotti.etal.2021}. While that increase in
-    erosion predates the increase observed in our study (ca.~15\,ka;
-    Fig.~\ref{fig:evolution}), it demonstrates that climatic conditions
-    over which glacier erosion occurs are relatively specific and likely
-    transient. Other field-based studies have evidenced increased glacier
-    erosion and pro-glacial sedimentation during the current deglaciation period
-    \citep[e.g.,][]{Koppes.Montgomery.2009, Micheletti.Lane.2016,
-    Lane.etal.2017, Bendixen.etal.2017}. This increase has often been
-    attributed to higher meltwater availability, promoting sediment transport
-    \citep{Delaney.Adhikari.2020}, enhanced glacier sliding
-    \citep{Herman.etal.2011}, and hydrofracturing, i.e.~plucking
-    \citep{Hallet.1996, Ugelvig.etal.2018, Hildes.etal.2004}.
-
-    But importantly, these hydrological processes are not included in our
-    model, so that deglaciation surface meltwater is assumed to instantly exit
-    the glacier with no effect on glacier dynamics \citep[cf.][for comparison]
-    {Werder.etal.2013, Iverson.2012, Ugelvig.etal.2018}. The enhanced glacier
-    velocity during periods of ice retreat is here modelled to result from
-    changes in glacier geometry alone (Figs.~\ref{fig:basaldrag}b).
-    Coincidentally, the link between climate and erosion is also the topic of
-    another independent study using PISM \citep{Lai.Anders.2021}.
-    Observed feedbacks between
-    surface meltwater and basal sliding, not included in the model, presumably
-    further accelerated erosion during deglaciation \citep{Herman.etal.2011}.
-    Furthermore, we omit the fluvial transport of sediment, which could well
-    increase sediment discharge and erosion with increased melt, so long as
-    additional sediment remains available \citep{Delaney.Adhikari.2020}. Rapid
-    glacier erosion can be expected to accompany the ongoing 21st-century
-    deglaciation, especially for the ice sheets of Greenland and
-    Antarctica where polythermal and cold-based glaciers are warming and
-    sliding may increase \citep[e.g.,][]{Moon.etal.2012, Mouginot.etal.2014,
-    Overeem.etal.2017}.
-
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-\subsection{Age of the glacial landscape}
-\label{sec:landscape}
-% -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -- -
-
-    \begin{figure*}
-      \centerline{\includegraphics{alpero_landscape}}
-      \caption{%
-        Alpine glacial erosion landscape diversity.
-        \textbf{(a)} Piedmont overdeepening of Lake Constance, ca.~10x50\,km.
-        \textbf{(b)} Glacial trough at Bout du Monde, ca.~1x10\,km.
-        \textbf{(c)} The unusually deep mountain cirque revealed by the current
-          demise of Ch\"uebodengletscher, ca.~1x1\,km.}
-      \label{fig:landscape}
-    \end{figure*}
-
-    In many places throughout the Alps, significant modelled cumulative erosion
-    occurs far inside the Alpine valleys (Fig.~\ref{fig:cumulative}). However,
-    the potential erosion patterns experience spatial shifts through the
-    glacial cycle (Fig.~\ref{fig:transects}), so that high-altitude rapid
-    erosion is not modelled to date from expansive glaciation stages, but
-    instead from stages of intermediate glacier cover
-    \citep[Fig.~\ref{fig:hypsogram};][]{Barr.etal.2019}.
-
-    For instance, low-elevation piedmont over-deepened basins such as Lake
-    Constance (Fig.~\ref{fig:landscape}a), Lake Geneva and Lake Maggiore are
-    here modelled to be the product of extensive Last Glacial Maximum-like
-    glaciations. Other modelling work suggests that features such as these
-    likely form when water can access the glacier bed and encourage glacier
-    sliding and, thus, erosion \citep{Herman.etal.2011}. The model herein does
-    not include subglacial hydrology, yet the role of climate warming to
-    increasing glacier erosion to the lower elevations remains evident
-    (Fig.~\ref{fig:hypsogram}).
-    Alpine terrain with elevations above 1000\,m are modelled to
-    experience virtually no erosion from 35 to 18\,ka BP
-    (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
-    intermediate-elevation glacial valleys of the Alps as Lauterbrunnental,
-    Haslital (Upper Aare) or Bout du Monde (Upper Giffre,
-    Fig.~\ref{fig:landscape}b)
-    would date from periods of intermediate glacier extension.
-
-    The validity of the model results at high elevation is discussable.
-    Crystalline massifs such as the Ecrins, Gran Paradiso, Monte Rosa, Aare,
-    \"Otztal and Tauern Massifs locally exhibit a strikingly high erosion
-    potential. However, the computation of glacier flow velocities on such
-    steep surfaces is strongly limited by the model horizontal resolution of
-    1\,km, the shallow-ice glacier flow physics \citep{Imhof.etal.2019}, and
-    PISM's current mass-conservation heuristics \citep{Imhof.2021}. Besides,
-    bergschrund (rimaye) processes likely to dominate interglacial cirque
-    erosion at such altitudes \citep{Sanders.etal.2012} are not captured by the
-    velocity-based glacier erosion power-laws. Nevertheless,
-    the lack of rapid erosion at high elevations for much of the modelled
-    glacial cycle (Fig.~\ref{fig:hypsogram}) implies that the highest mountain
-    cirques (e.g. Fig.~\ref{fig:landscape}c) would have formed
-    over interglacial periods, when glaciers are confined to high terrain
-    \citep{Barr.etal.2017, Barr.etal.2019}. Similar processes also occur on
-    Himalayan glaciers and may limit erosion for the high elevation portion of
-    that mountain range during cold periods as well, failing to offset uplift
-    rates \citep{Harper.Humphrey.2003}. Over longer timescales, though, the
-    vertical distribution of erosion rates also depends on the erosional
-    modification of topography \citep{Sternai.etal.2013}.
-
-    Finally, the time-transgressive nature of the modelled glacial erosion
-    rates hint at a new explanation for the Alpine glacier trim-line and lack
-    of cold-based glaciation evidence. Due to their geographic location in the
-    mid-latitudes, and their steep topographic gradient, the Alps appear as a
-    mountain range that hosted glaciers of various sizes throughout the varying
-    climate of the Quaternary, resulting in a transgressive localization of
-    glacier erosion throughout their elevational range, from the piedmont
-    during glacial maxima, to the highest cirques during interglacial periods.
-    The Alpine trim-line, in this case, would neither correspond to the upper
-    reach of the Last Glacial Maximum glaciers \citep[e.g.,][]{Kelly.etal.2004},
-    nor to an englacial temperate-ice boundary \citep{Coutterand.2010,
-    Seguinot.etal.2018}, but to a time-transgressive upper limit of erosion from
-    advancing and retreating glaciers on steep terrain.
-
-% ----------------------------------------------------------------------
-\conclusions
-% ----------------------------------------------------------------------
-
-    Due to compounded uncertainties regarding paleoclimate, glacier sliding and
-    erosion processes, our quantitative results are very likely inaccurate.
-    However, we draw the following qualitative conclusions:
-    \begin{itemize}
-      \item The non-linear physics of glacier deformation and sliding, combined
-        with non-linear empirical erosion laws, results in a very strong
-        localization of rapid glacier erosion in regions of fast flow.
-      \item This high spatio-temporal variability hints at a complex
-        relationship between climate and glacier erosion, so that a highly
-        variable response of glacier erosion to climate should be expected.
-      \item Increased gravitational drag due to surface profile steepening
-        provides a mechanism
-        for accelerated erosion during deglaciation periods, irrespective of
-        surface meltwater penetration to the glacier bed.
-      \item If a non-linear glacier erosion law is used, the climate-induced
-        slowdown of erosion counterbalances glacier expansion, so that
-        Alpine-wide glacier erosion volumes do not correlate with the ice volume.
-      \item Rapid glacier erosion is restricted to low elevation during stages
-        of glacier advance and maximum glaciation, but propagates up-valley to
-        higher elevations during periods of glacier retreat.
-      \item The diversity of the Alpine glacier erosion landscape appears to
-        be the time-transgressive signature of a variety of glacial stages
-        ranging from pan-Alpine ice sheets to modern-style mountain glaciers.
-    \end{itemize}
-
-% ----------------------------------------------------------------------
-% Acknowledgements
-% ----------------------------------------------------------------------
-
-\codedataavailability{%
-    PISM is available as open-source software (http://pism-docs.org).
-    Time and domain-aggregated erosion variables such as used in the figures is
-    available for the four presented erosion power-laws
-    (\doi{10.5281/zenodo.4495419}). Other model output variables were
-    previously made available (\doi{10.5281/zenodo.3604174},
-    \doi{10.5281/zenodo.3604142}).}
-
-\videosupplement{
-    Animations made available online show the modelled glacier erosion rates
-    (\url{https://vimeo.com/503162771}), and their evolving relationship with
-    ice volume (\url{https://vimeo.com/512478926}) and bedrock
-    altitude (\url{https://vimeo.com/512479008}).}
-
-\authorcontribution{%
-    J.~Seguinot performed the computations and prepared the figures.
-    I.~Delaney interpreted the results in a broader context. Both authors
-    contributed to the text.}
-
-\competinginterests{%
-    The authors declare that they have no conflict of interest.}
-
-\begin{acknowledgements}
-    We are very thankful to Constantine Khroulev, Ed Bueler, and Andy
-    Aschwanden for their constant help and development with PISM, without whom
-    this work would not have been possible. J.~Seguinot would like to thank
-    John Jansen and Martin Margold for insightful early discussions, Susan
-    Ivy-Ochs for being a constant (droid) motivator, Popi, Mattheos, Karsta,
-    Stefan, Fran\c{c}oise and Dominique
-    for welcoming him on Anafi and providing agreeable conditions for work in
-    times of pandemic, and the Long Haul Trucker for conveying his laptop and
-    data unbroken for 5000\,km across Europe and fourteen countries.
-    I.~Delaney thanks Fr\'ed\'eric Herman for valuable support and
-    conversations while at University of Lausanne.
-    An anonymous reviewer and Ian Evans (aided by Iestyn Barr, Jeremy Ely, Matt
-    Tomkins, Pippa Whitehouse, Cristina Balaban and Matt Wiecek) provided
-    constructive feedback which has been incorporated in the present paper.
-    The (previously published) ice flow simulations were made with computer
-    resources provided by the Swiss National Supercomputing Centre (CSCS)
-    allocations no.~s573 and sm13 to J.~Seguinot.
-\end{acknowledgements}
 
@@
 
+%\clearpage
 \bibliographystyle{copernicus}

### 2021-06-10 figures/alpero_evolution.py
Automate alternative laws for evolution plot.

@@ -5,13 +5,17 @@
 
 """Plot Alps erosion time evolution."""
 
+import os
 import absplots as apl
 import hyoga.open
-import util
+import util  # noqa color cycle
 
 
 def main():
     """Main program called during execution."""
+
+    # erosion law
+    law = os.getenv('ALPERO_LAW', 'kop2015')
 
     # initialize figure
     fig, ax = apl.subplots_mm(figsize=(85, 80), gridspec_kw=dict(
@@ -28,43 +32,51 @@ def main():
 
     # unit conversion and rolling mean
     ds['slvol'] *= 100
-    ds['rolling_mean'] = ds.kop2015_rate.rolling(age=100, center=True).mean()
+    erosion = ds[law+'_rate']
+    rolling = erosion.rolling(age=100, center=True).mean()
     ds['growing'] = ds.slvol.differentiate('age') < 0  # age coord decreasing
 
     # plot
-    ax.plot(ds.slvol, ds.kop2015_rate.where(ds.growing), c='C1', alpha=0.25)
-    ax.plot(ds.slvol, ds.kop2015_rate.where(~ds.growing), c='C11', alpha=0.25)
-    ax.plot(ds.slvol, ds.rolling_mean.where(ds.growing), c='C1', ls='--')
-    ax.plot(ds.slvol, ds.rolling_mean.where(~ds.growing), c='C11')
+    ax.plot(ds.slvol, erosion.where(ds.growing), c='C1', alpha=0.25)
+    ax.plot(ds.slvol, erosion.where(~ds.growing), c='C11', alpha=0.25)
+    ax.plot(ds.slvol, rolling.where(ds.growing), c='C1', ls='--')
+    ax.plot(ds.slvol, rolling.where(~ds.growing), c='C11')
     ax.text(0.95, 0.95, '', ha='right', va='top', transform=ax.transAxes)
 
     # hatch regions of low ice volume
     ax.fill_between(
-        [-3, 3], 1e4, 1e8, edgecolor='0.75', facecolor='none', hatch='//////')
+        [-3, 3], 1e4, 1e11, edgecolor='0.75', facecolor='none', hatch='//////')
 
     # set axes properties
+    ymid = erosion.where(ds.slvol > 3).mean()
     ax.set_xlabel('ice volume (cm s.l.e.)')
     ax.set_ylabel(r'potential annual erosion volume ($m^3 a^{-1}$)')
[…]

### 2021-06-10 articles/erosion/alps-erosion-supplement.tex
Add volume evolution plots in supplement.

@@
           below 3\,cm~s.l.e. where glacier sliding may be affected by
-          stress-balance approximations and model horizontal resolution.}
+          stress-balance approximations and model horizontal resolution.
+        After \citet{Koppes.etal.2015}, same as main text Fig. 1.}
         \label{fig:cumulative}
@@
 
+    \begin{figure*}
+      \begin{minipage}[t]{85mm}
+      \includegraphics{alpero_evolution}
+      \caption[%
+        Potential erosion volume in relation to ice volume after
+        \citet{Koppes.etal.2015}, same as main text Fig. 2.
+      ]{%
+        Modelled annual (domain-integrated) potential erosion volume (light
+        curves) and its 1-ka running mean (dark curves) in relation to the
+        modelled total ice volume in centimetres of sea-level equivalent. Blue
+        indicates increasing ice volume and brown decreasing ice volume.
+        After \citet{Koppes.etal.2015}, same as main text Fig. 2.}
+      \end{minipage}
+      \hfill
+      \begin{minipage}[t]{85mm}
+      \includegraphics{alpero_evolution_her2015}
+      \caption{%
+        Potential erosion volume in relation to ice volume after
+        \citet{Herman.etal.2015}.}
+      \end{minipage}
+    \end{figure*}
+
+    \begin{figure*}
+      \begin{minipage}[t]{85mm}
+      \includegraphics{alpero_evolution_hum1994}
+      \caption{%
+        Potential erosion volume in relation to ice volume after
+        \citet{Humphrey.Raymond.1994}.}
+      \end{minipage}
+      \hfill
+      \begin{minipage}[t]{85mm}
+      \includegraphics{alpero_evolution_coo2020}
+      \caption{%
+        Potential erosion volume in relation to ice volume after
+        \citet{Cook.etal.2020}.}
+      \end{minipage}
+    \end{figure*}
+
 % ----------------------------------------------------------------------

### 2021-06-11 figures/alpero_transects.py
Automate alternative laws for rhine transect plot.

@@ -5,6 +5,7 @@
 
 """Plot Alps erosion transect evolution."""
 
+import os
 import matplotlib as mpl
 import cartopy.crs as ccrs
 import cartowik.profiletools as cpf
@@ -16,6 +17,9 @@ import util
 
 def main():
     """Main program called during execution."""
+
+    # erosion law
+    law = os.getenv('ALPERO_LAW', 'kop2015')
 
     # initialize figure
     fig, grid = apl.subplots_mm(
@@ -33,7 +37,10 @@ def main():
 
     # plot ages and levels for consistency
     ages = [36, 24, 16, 0]
-    levels = [10**i for i in range(-9, 1)]
+    if law == 'kop2015':
+        levels = [10**i for i in range(-9, 1)]
+    else:
+        levels = [10**i for i in range(-6, 4)]
 
     # read profile coordinates
     x, y = cpf.read_shp_coords('../data/native/profile_rhine.shp')
@@ -52,7 +59,14 @@ def main():
         # compute erosion
         ds = extra.sel(age=age)
         sliding = (ds.uvelbase**2+ds.vvelbase**2)**0.5  # (m/a)
-        erosion = 5.2e-8*sliding**2.34  # (mm/a, Koppes et al., 2015)
+        if law == 'kop2015':
+            erosion = 5.2e-8*sliding**2.34  # (mm/a, Koppes et al., 2015)
+        elif law == 'her2015':
+            erosion = 2.7e-4*sliding**2.02  # (mm/a, Herman et al., 2015)
+        elif law == 'hum1994':
+            erosion = 1e-1*sliding  # (mm/a, Humphrey and Raymond, 1994)
+        elif law == 'coo2020':
+            erosion = 1.665e-1*sliding**0.6459  # (mm/a, Cook et al., 2020)
 
         # plot map data
         ds.topg.plot.imshow(
@@ -91,7 +105,7 @@ def main():
 
         # plot erosion profile (coarsen to reduce figure size)
         ds = ds.coarsen(age=5).mean()  # 3.7 -> 1.0 MiB
-        (ds.assign_coords(d=ds.d/1e3).kop2015_rhin*1e3).plot.contourf(
+        (ds.assign_coords(d=ds.d/1e3)[law+'_rhin']*1e3).plot.contourf(
             ax=tsax, alpha=0.75, cmap='YlOrBr', levels=levels, x='age', y='d',
             cbar_ax=cax, cbar_kwargs=dict(
[…]

### 2021-06-11 articles/erosion/alps-erosion-supplement.tex
Add Rhine transect plots in supplement.

@@
 
+    \begin{figure*}
+      \centerline{\includegraphics{alpero_transects}}
+      \caption[%
+        Rhine Glacier potential erosion rate over time after
+        \citet{Koppes.etal.2015}, same as main text Fig. 3.
+      ]{%
+        \textbf{(a--d)} Modelled instantaneous potential erosion rate of the
+          Rhine Glacier for selected glacier advance and retreat ages, and the
+          final model state for topographic reference.
+        \textbf{(e)} Interpolated instantaneous potential erosion rate along a
+          Rhine Glacier transect (upper panels dashed line) for the entire last
+          glacial cycle.
+        After \citet{Koppes.etal.2015}, same as main text Fig. 3.}
+    \end{figure*}
+
+    \begin{figure*}
+      \includegraphics{alpero_transects_her2015}
+      \caption{%
+        Rhine Glacier potential erosion rate over time after
+        \citet{Herman.etal.2015}.}
+    \end{figure*}
+
+    \begin{figure*}
+      \includegraphics{alpero_transects_hum1994}
+      \caption{%
+        Rhine Glacier potential erosion rate over time after
+        \citet{Humphrey.Raymond.1994}.}
+    \end{figure*}
+
+    \begin{figure*}
+      \includegraphics{alpero_transects_coo2020}
+      \caption{%
+        Rhine Glacier potential erosion rate over time after
+        \citet{Cook.etal.2020}.}
+    \end{figure*}
+
 % ----------------------------------------------------------------------

### 2021-06-11 figures/alpero_hypsogram.py
Automate alternative laws for hypsogram plot.

@@ -5,6 +5,7 @@
 
 """Plot Alps erosion hypsogram."""
 
+import os
 import numpy as np
 import matplotlib.pyplot as plt
 import absplots as apl
@@ -14,6 +15,9 @@ import util
 
 def main():
     """Main program called during execution."""
+
+    # erosion law
+    law = os.getenv('ALPERO_LAW', 'kop2015')
 
     # initialize figure
     fig = apl.figure_mm(figsize=(177, 80))
@@ -46,11 +50,12 @@ def main():
         plt.rc('hatch', color='0.75')
 
         # plot hypsogram
-        (np.log10(ds.kop2015_hyps)+3).plot.imshow(
-            ax=ax, alpha=0.75, cmap='YlOrBr', vmin=-9, vmax=0, x='age',
+        vmax = 0 if law == 'kop2015' else 3
+        (np.log10(ds[law+'_hyps'])+3).plot.imshow(
+            ax=ax, alpha=0.75, cmap='YlOrBr', vmin=vmax-9, vmax=vmax, x='age',
             cbar_ax=cax, cbar_kwargs=dict(
                 label='geometric mean\n'+r'erosion rate ($mm\,a^{-1}$)',
-                ticks=range(-9, 1, 3)))
+                ticks=range(vmax-9, vmax+1, 3)))
         ds.glacier_hyps.plot.contourf(
             ax=ax, add_colorbar=False, colors='none', extend='neither',
             hatches=['//////'], levels=[0.5e6, 99.5e6], x='age')
@@ -58,7 +63,7 @@ def main():
             ax=ax, colors='0.25', levels=[99.5e6], x='age')
         cax.yaxis.set_major_formatter(r'$10^{{{x:d}}}$')
         # this should work in matplotlib 3.3.2 (PR #18458)
-        # (ds.kop2015_hyps*1e3).plot.imshow(
+        # (ds[law+'_hyps']*1e3).plot.imshow(
         #    ax=ax, alpha=0.75, cmap='YlOrBr', norm=mcolors.LogNorm(1e-9, 1e0),
         #    x='age', cbar_ax=cax, cbar_kwargs=dict(
         #        label=r'erosion rate ($m\,a^{-1}$)'))
@@ -82,7 +87,7 @@ def main():
 
         # plot band cumulative erosion in km3
         hax = hax.twiny()
-        (ds.kop2015_cumu.groupby_bins(boot, bins).sum()/1e3).plot.step(
[…]

### 2021-06-11 articles/erosion/alps-erosion-supplement.tex
Add erosion hypsogram plots in supplement.

@@
 
+    \begin{figure*}
+      \centerline{\includegraphics{alpero_hypsogram}}
+      \caption[
+        Potential erosion rate per altitude band after
+        \citet{Koppes.etal.2015}, same as main text Fig. 4.
+      ]{%
+        \textbf{(a)} Potential erosion rate ``hypsogram'', showing the
+          geometric mean of modelled rates in 100-m elevation bands
+          across the entire model domain and its time evolution. Hatches
+          indicate elevation bands with fewer than a hundred grid cells
+          (100\,$km^2$).
+        \textbf{(b)} Distribution of model domain bedrock topography (grey
+          bars) and glaciated topography (blue bars) in 100-m elevation bands,
+          and the corresponding cumulative potential erosion volume (brown
+          line).
+        \textbf{(c)} Same as Fig.~\ref{fig:cumulative}b.
+        After \citet{Koppes.etal.2015}, same as main text Fig. 4.}
+    \end{figure*}
+
+    \begin{figure*}
+      \includegraphics{alpero_hypsogram_her2015}
+      \caption{%
+        Potential erosion rate per altitude band after
+        \citet{Herman.etal.2015}.}
+    \end{figure*}
+
+    \begin{figure*}
+      \includegraphics{alpero_hypsogram_hum1994}
+      \caption{%
+        Potential erosion rate per altitude band after
+        \citet{Humphrey.Raymond.1994}.}
+    \end{figure*}
+
+    \begin{figure*}
+      \includegraphics{alpero_hypsogram_coo2020}
+      \caption{%
+        Potential erosion rate per altitude band after
+        \citet{Cook.etal.2020}.}
+    \end{figure*}
+
 % ----------------------------------------------------------------------

### 2021-06-11 figures/alpero_sensitivity.py
Automate alternative laws for sensitivity plot.

@@ -5,6 +5,7 @@
 
 """Plot Alps erosion sensibility to climate scenario."""
 
+import os
 import matplotlib.pyplot as plt
 import cartopy.crs as ccrs
 import cartowik.decorations as cde
@@ -15,6 +16,9 @@ import util
 
 def main():
     """Main program called during execution."""
+
+    # erosion law
+    law = os.getenv('ALPERO_LAW', 'kop2015')
 
     # initialize figure
     fig, grid = apl.subplots_mm(
@@ -40,10 +44,12 @@ def main():
                     rec.lower()[:4], ('cp', 'pp')['pp' in conf])) as ds:
 
             # plot map data
-            cset = ds.kop2015_cumu.plot.contourf(
-                ax=ax, cmap='YlOrBr', levels=[10**i for i in range(-2, 3)],
+            midlev = 0 if law == 'kop2015' else 2
+            cset = ds[law+'_cumu'].plot.contourf(
+                ax=ax, cmap='YlOrBr',
+                levels=[10**i for i in range(midlev-2, midlev+3)],
                 alpha=0.75, add_colorbar=False, extend='both')
-            ds.kop2015_cumu.notnull().plot.contour(
+            ds[law+'_cumu'].notnull().plot.contour(
                 ax=ax, colors='k', linewidths=0.5, levels=[0.5])
 
         # add map elements
@@ -66,8 +72,8 @@ def main():
     fig.colorbar(cset, cax=cax, label='cumulative erosion potential (m)',
                  format='%g', orientation='horizontal')
 
-    # save
-    util.com.savefig(fig)
+    # save figure
+    fig.savefig(__file__[:-3] + ('_'+law if law != 'kop2015' else ''))
 
 
 if __name__ == '__main__':

### 2021-06-11 articles/erosion/alps-erosion-supplement.tex
Add climate sensitivity plots in supplement.

@@
     This document contains copies of the main text Figs. 1--5, and their
-    equivalent using the other tested erosion laws.
+    equivalent using the other tested erosion laws. Note that colour levels and
+    vertical axes were adapted to the magnitude of the results. Potential
+    erosion rates aggregated in time, space and elevation bands, and used to
+    plot these figures were made available in a long-term online archive
+    \citep[\doi{10.5281/zenodo.4495419}]{Seguinot.2021}.
 
@@
 
+    \begin{figure*}
+      \centerline{\includegraphics{alpero_sensitivity}}
+      \caption[
+        Climate sensitivity of cumulative erosion potential after
+        \citet{Koppes.etal.2015}, same as main text Fig. 5.
+      ]{%
+        Modelled cumulative glacial erosion potential over the last glacial
+        cycle using three different paleo-temperature histories from
+        \textbf{(a, d)} the Greenland Ice Core Project
+        \citep[GRIP;][]{Dansgaard.etal.1993}, \textbf{(b, e)} the European
+        Project for Ice Coring in Antarctica \citep[our default,
+        EPICA;][]{Jouzel.etal.2007}, and \textbf{(c, f)} an oceanic sediment
+        core from the Iberian margin \citep[MD01-2444;][]{Martrat.etal.2007},
+        and both \textbf{(a--c)} without and \textbf{(d--f)} with
+        paleo-precipitation reductions \citep[cf.][]{Seguinot.etal.2018}.
+        After \citet{Koppes.etal.2015}, same as main text Fig. 5.}
+    \end{figure*}
+
+    \begin{figure*}
+      \includegraphics{alpero_sensitivity_her2015}
+      \caption{%
+        Climate sensitivity of cumulative erosion potential after
+        \citet{Herman.etal.2015}.}
+    \end{figure*}
+
+    \begin{figure*}
+      \includegraphics{alpero_sensitivity_hum1994}
+      \caption{%
+        Climate sensitivity of cumulative erosion potential after
+        \citet{Humphrey.Raymond.1994}}
+    \end{figure*}
+
+    \begin{figure*}
+      \includegraphics{alpero_sensitivity_coo2020}
+      \caption{%
+        Climate sensitivity of cumulative erosion potential after
+        \citet{Cook.etal.2020}}
+    \end{figure*}
+
 % ----------------------------------------------------------------------
@@
 
-%\clearpage
+\clearpage
 \bibliographystyle{copernicus}

### 2021-06-14 articles/erosion/response-editor.tex
Convert response to the editor to latex.

% Copyright (c) 2021, Julien Seguinot (juseg.github.io), Ian Delaney
% Creative Commons Attribution-ShareAlike 4.0 International License
% (CC BY-SA 4.0, http://creativecommons.org/licenses/by-sa/4.0/) -->

\input{response-header}

\title{Authors' response to the Editor}
\author{J. Seguinot, on behalf of all authors.}
%\date{}

\begin{document}
\thispagestyle{empty}
\maketitle
\bigskip

    Dear Andreas Lang,

    We apologize for the delays during the peer review of our manuscript. We
    believe that we have addressed all points raised by the reviews and hereby
    submit our revised version to \emph{Earth Surface Dynamics}.

    Please find hereafter a short summary and a marked-up file listing changes
    made to the manuscript since its first publication as a preprint. If
    possible, we ask you to please refer to our public responses for more
    detailed point-by-point lists of changes.

    \begin{itemize}

        \item \textbf{Sect. 2 (Methods):}
        We made explicit the input topography and temperature records used in
        the parent study (Seguinot et~al., 2018), clarified simplifications
        made when computing erosion rates from basal velocities, and,
        accordingly, refer to the resulting quantities as ``erosion potential''
        in the entire manuscript.

        \item \textbf{Sect. 4 (Discussion):}
        We added sentences and whole paragraphs to discuss the systematic
        eastwards bias of ice cover (and thus erosion potential), the magnitude
        of our computed erosion potential versus expectable glacial cycle
        erosion, the uncertainties on basal sliding, and model limitations in
        high mountains due to resolution and approximated physics.

        \item \textbf{Figures:}
        We added scale bars on all map figures, and hatched regions on Figs.~1,
        2, 4 and~6 to indicate low confidence in our results for small
        glaciers. We added the field-based LGM outline on Fig.~1 (ex-Fig.~2),
        used different line styles to distinguish glacier advance and retreat
        on Fig.~2 (ex-Fig.~3), added glacier-advance (36 ka) and ice-free
        (0\,ka) erosion maps of the Rhine Glacier on Fig.~3 (ex-Fig.~4),
        plotted glaciated hypsometry and cumulative erosion per elevation band
        on Fig.~4 (ex-Fig.~5), made style changes to improve readability of
        Figs.~6 and 7 (ex-Figs.~7 and~8), moved Fig.~8 (ex-Fig.~1) into the
        discussion and replaced one of the photos.

        \item \textbf{Text improvements}
        were made throughout the text and figure captions following suggestions
        from the reviewers.

    \end{itemize}

    We thank you again very much for your editorial work on our manuscript.

\end{document}

### 2021-06-14 articles/erosion/response-editor.tex
Mention supplement in response, tag 3.0.

@@
 
-    Please find hereafter a short summary and a marked-up file listing changes
+    Please find hereafter a short summary (and attached, a marked-up file)
+    listing changes
     made to the manuscript since its first publication as a preprint. If
@@
 
+        \item \textbf{Supplementary figures:}
+        This was not mentioned in our responses. Following the reviews, and
+        if this does not result in additional page charges, we would like to
+        amend a supplementary document containing copies of Figs.~1--5, and
+        their equivalent as plotted with alternative erosion laws (more
+        briefly presented in the manuscript Fig.~6). Comments from both
+        reviewers highlight the difficulty of labelling any of the tested
+        erosion laws more ``realistic'' (this wording was removed), and the
+        supplementary figures may prove useful alongside future developments of
+        empirical glacier erosion laws.
+
         \item \textbf{Text improvements}

### 2021-07-09 articles/erosion/alps-erosion.tex
Fix agg data date, update erosion doi, tag 3.1.

@@
     present study \citep{Seguinot.2021}, and for the earlier publication
-    \citep{Seguinot.2020, Seguinot.2020a}.
+    \citep{Seguinot.2019, Seguinot.2020}.
 
@@
     available for the four presented erosion power-laws
-    (\doi{10.5281/zenodo.4495419}). Other model output variables were
+    (\doi{10.5281/zenodo.5084579}). Other model output variables were
     previously made available (\doi{10.5281/zenodo.3604174},
